---
title: "Using gpmapr"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gpmapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

```{r setup}
library(gpmapr)
```

# Introduction

The gpmapr package provides a set of functions to interact with the GPMap API.

# Getting started

The first port of entry is the `search_gpmap` function.  This function allows you to search for a trait, gene, or variant.

```{r search-trait}
haemoglobin <- search_gpmap("haemoglobin")
knitr::kable(haemoglobin[, c("name", "sample_size", "num_coloc_groups", "call")])
```
You can see information about the different traits that match your search, how to subsequently get more information (under `call`), and metadata about the trait (under `info`).

### Understanding `info`

* Extractions: Each trait has a number of extractions, related to the number of finemapped ld regions that had a p-value < 1.5e-4.
* Colocalisation Groups: The number of colocalisation groups associated with the trait / gene / variant.
* Colocalising Traits: The number of traits that are colocalising with the trait / gene / variant.
* Rare Result Groups: Each trait has a number of rare result groups, related to the number of groups of rare result groups that had a p-value < 1.5e-4.

You can also search for a gene (by ENSG ID or gene name) or variant (by CHR:BP or rsID).

```{r other-searches}
knitr::kable(search_gpmap("SLC44A")[, c("name", "sample_size", "num_study_extractions", "num_coloc_groups", "call")])
knitr::kable(search_gpmap("22:37042914"))
```

### Variant search returns proxy variants

The variant search will return the original variant and any proxy variants that are in LD with the original variant, which also have colocalisation results.  Each variant is marked as either `original_variant` or `proxy_variant`.  In this example, the original variant has no colocalisation results, but the proxy variant does.  The proxy variant is also marked as having a r^2 of 0.96

```{r proxy-variants}
proxy_variants <- search_gpmap("rs17078078")
knitr::kable(proxy_variants)
```

# Diving in

You can get more information about a trait by subsequently calling the `trait()`, `gene()`, or `variant()` functions.  This will provide you with the colocalisation results, rare disease results, and study extractions.

Each trait, gene, or variant call have optional parameters:

* `include_associations`: Whether to include associations (BETA, SE, P) of the variants in which it's colocalised in the output.
* `include_coloc_pairs`: Whether to include coloc pairs in the output (which include h3 and h4), as opposed to just the coloc groups.
* `h4_threshold`: The h4 threshold for coloc pairs.

## Trait example

```{r trait-example}
hemoglobin_trait <- trait(931)
names(hemoglobin_trait)
nrow(hemoglobin_trait$coloc_groups)
nrow(hemoglobin_trait$study_extractions)
```

This returns a list of elements, which you can access by name.  More detailed explanations of the below are described in the API documentation.

* `trait`: A list containing metadata about the trait, including common and rare studies associated with the trait
* `coloc_groups`: a dataframe containing information about which studies have coloc results for this trait.
* `study_extractions`: a list of dataframes containing the study extractions for this trait.
* `rare_results`: (if they exist) a list of dataframes containing the rare results for this trait
* `associations`: (optional) a dataframe containing the associations for the variants in which the trait is colocalised
* `coloc_pairs`: (optional) a dataframe containing all pairwise coloc results for this trait.

### How `study_extractions`, `coloc_groups`, and `rare_results` relate

All three layers refer to **study extractions** (GWAS hits in finemapped LD regions), keyed by `id` in `study_extractions` and by `study_extraction_id` in `coloc_groups` and `rare_results`.

Think of **`study_extractions`** as the catalogue of extractions linked to the trait or gene in the map: each row is a study signal at a locus. Not every listed extraction necessarily has a full colocalisation story or a rare-disease result row.

- **`coloc_groups`** holds extractions that sit inside a **colocalisation group** (shared coloc evidence with other studies/traits at that locus). If a study extraction appears here, it has coloc-structured results for that entity.
- **`rare_results`** holds extractions that belong to a **rare-result group** (rare-disease-style finemapping in the API). The same biological signal can, in principle, be reflected in both rare and common coloc paths, so overlap between this set of extraction ids and those in `coloc_groups` is possible.
- **Neither table:** Some extractions appear only under **`study_extractions`**. They are still tied to the trait or gene in the genotype–phenotype map (for example via **locus tagging, proximity, or other association logic** in the pipeline), but they do not currently carry a coloc-group row or a rare-result group row. Treat those as “present on the map at this locus” rather than “absent by mistake.”

The same idea applies when you call **`gene()`** (or other entity-specific endpoints): `study_extractions` is the wider catalogue, while `coloc_groups` and `rare_results` subdivide it where those analyses exist. The table below illustrates counts for the hemoglobin trait example (`trait(931)`).

```{r trait-extraction-membership}
extraction_ids <- function(x) {
  if (is.null(x) || !length(x)) {
    return(integer(0))
  }
  id_col <- if ("id" %in% names(x)) "id" else NULL
  if (!is.null(id_col)) {
    return(unique(stats::na.omit(x[[id_col]])))
  }
  unique(stats::na.omit(unlist(lapply(x, function(df) {
    if (!is.data.frame(df)) {
      return(NULL)
    }
    if ("id" %in% names(df)) {
      return(df[["id"]])
    }
    NULL
  }))))
}

cg_ids <- unique(stats::na.omit(hemoglobin_trait$coloc_groups$study_extraction_id))

rr <- hemoglobin_trait$rare_results
rr_ids <- if (is.null(rr) || !length(rr)) {
  integer(0)
} else {
  unique(stats::na.omit(unlist(lapply(rr, function(d) {
    if (!is.data.frame(d) || !"study_extraction_id" %in% names(d)) {
      return(NULL)
    }
    d[["study_extraction_id"]]
  }))))
}

se_ids <- extraction_ids(hemoglobin_trait$study_extractions)

only_study_extractions <- setdiff(se_ids, union(cg_ids, rr_ids))

knitr::kable(data.frame(
  source = c(
    "Distinct study_extraction_id in coloc_groups",
    "Distinct study_extraction_id in rare_results",
    "In study_extractions only (not in coloc_groups nor rare_results)",
    "Total distinct extraction id in study_extractions"
  ),
  n_extractions = c(
    length(cg_ids),
    length(rr_ids),
    length(only_study_extractions),
    length(se_ids)
  )
))
```

# Gene example

```{r gene-example}
slc44a1 <- gene("SLC44A1")
names(slc44a1)
nrow(slc44a1$coloc_groups)
knitr::kable(head(slc44a1$coloc_groups[, c("coloc_group_id", "trait_name", "data_type", "tissue")]))
```

Note that by default, the gene endpoint includes trans genetic effects. You can turn this off with `gene("SLC44A1", include_trans = FALSE)`, or filter `coloc_groups` after the fact (for example `cis_trans == "cis"` or a `min_p` cutoff).

After you subset rows, some `coloc_group_id` values may be left with only a single row. Those are no longer meaningful colocalisation **groups**, so drop them by keeping only groups with more than one row:

```{r gene-filter-example}
slc44a1_no_trans <- slc44a1
slc44a1_no_trans$coloc_groups <- dplyr::filter(
  slc44a1_no_trans$coloc_groups,
  cis_trans == "cis"
)
# or e.g. dplyr::filter(..., min_p <= 5e-8)

slc44a1_no_trans$coloc_groups <- slc44a1_no_trans$coloc_groups |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::filter(dplyr::n() > 1L) |>
  dplyr::ungroup()

nrow(slc44a1_no_trans$coloc_groups)
knitr::kable(head(slc44a1_no_trans$coloc_groups[, c("coloc_group_id", "trait_name", "data_type", "tissue")]))
```

## Coloc pairs example

Although coloc groups provide a helpful overview of the colocalisation results for a gene, you may want to look at the pairwise relationships between the studies.  You can do this by setting `include_coloc_pairs = TRUE` in the call.

As there may be many coloc pairs, the results are not as fleshed out as the coloc groups.  You may have to filter the results to get the information you need, or join with the study extractions to get the full picture.

```{r coloc-pairs-example}
slc44a1_coloc_pairs <- gene("SLC44A1", include_coloc_pairs = TRUE)
nrow(slc44a1_coloc_pairs$coloc_pairs)
knitr::kable(head(slc44a1_coloc_pairs$coloc_pairs[, c("study_extraction_a_id", "study_extraction_b_id", "h4")]))

study_extractions_tbl <- dplyr::bind_rows(slc44a1_no_trans$study_extractions)
study_extractions_subset <- dplyr::select(study_extractions_tbl, id, trait_name, min_p)

hydrated_coloc_pairs <- slc44a1_coloc_pairs$coloc_pairs |>
  dplyr::left_join(slc44a1_coloc_pairs$study_extractions, by = c("study_extraction_a_id" = "id")) |>
  dplyr::rename(trait_a = trait_name) |>
  dplyr::left_join(slc44a1_coloc_pairs$study_extractions, by = c("study_extraction_b_id" = "id")) |>
  dplyr::rename(trait_b = trait_name)

knitr::kable(head(hydrated_coloc_pairs[, c("trait_a", "trait_b", "h4")]))
```


# Using Pleiotropy Scores and Directionality to uncover genetic insights from colocalisation findings

For the gene, distinct_trait_categories and distinct_protein_coding_genes are available and automatically are returned when you call `gene()`.  These are the number of trait categories and protein coding genes that the gene is associated with via coloc groups.

```{r gene-pleiotropy-example}
trem2 <- gene("TREM2", include_associations = TRUE)
trem2$gene$distinct_trait_categories
trem2$gene$distinct_protein_coding_genes
```

We can see that TREM2 is associated with only one trait category, and two other protein coding genes, TREM1 and TREML2.

Furthermore, if we look at the colocalization group that is associated with Alzheimer's disease, we can see that the associations are such that any decrease in the expression of TREM2 is associated with an increase in Alzheimer's disease.

```{r trem2-alzheimer-example}

alzheimer_trait_ids <- search_gpmap("alzheimer")$type_id
trem2_alzheimer_coloc_group_ids <- unique(
  trem2$coloc_groups[trem2$coloc_groups$trait_id %in% alzheimer_trait_ids, ]$coloc_group_id
)
trem2_alzheimer_rare_result_group_ids <- unique(
  trem2$rare_results[trem2$rare_results$trait_id %in% alzheimer_trait_ids, ]$rare_result_group_id
)
trem2_alzheimer_groups <- trem2$coloc_groups[
  trem2$coloc_groups$coloc_group_id %in% trem2_alzheimer_coloc_group_ids,
]
trem2_alzheimer_rare_results <- trem2$rare_results[
  trem2$rare_results$rare_result_group_id %in% trem2_alzheimer_rare_result_group_ids,
]
```

We can see 2 colocalization groups and 2 rare result groups are associated with Alzheimer's disease, and TREM2 are not associated with any other trait categories.

Here we have made a forest plot of the association betas relating to Alzheimer's disease, combining **colocalisation groups** (one panel per `coloc_group_id`) and **rare disease result groups** (facet strip shows the **SNP** from `display_snp`, or `variant_id` if needed).

```{r trem2-alzheimer-forest, fig.width = 8, fig.height = 8}
forest_row_label <- function(trait_name, data_type, tissue) {
  tn <- if (missing(tissue) || is.null(tissue)) {
    rep(NA_character_, length(trait_name))
  } else {
    tissue
  }
  paste(
    trait_name,
    data_type,
    tidyr::replace_na(as.character(tn), ""),
    sep = " — "
  )
}

trem2_forest_coloc <- trem2_alzheimer_groups |>
  dplyr::filter(!is.na(.data$beta), !is.na(.data$se), .data$se > 0) |>
  dplyr::mutate(
    lo = .data$beta - 1.96 * .data$se,
    hi = .data$beta + 1.96 * .data$se,
    row_label = forest_row_label(.data$trait_name, .data$data_type, .data$tissue),
    panel_id = dplyr::coalesce(
      as.character(.data$display_snp),
      paste0("variant ", as.character(.data$variant_id))
    ),
    analysis = "Colocalisation"
  ) |>
  dplyr::group_by(.data$coloc_group_id) |>
  dplyr::arrange(.data$beta, .by_group = TRUE) |>
  dplyr::mutate(
    row_label = factor(.data$row_label, levels = unique(.data$row_label))
  ) |>
  dplyr::ungroup()

trem2_forest_rare <- if (is.null(trem2_alzheimer_rare_results) || nrow(trem2_alzheimer_rare_results) == 0L) {
  NULL
} else {
  rr <- trem2_alzheimer_rare_results
  if (!"tissue" %in% names(rr)) {
    rr$tissue <- NA_character_
  }
  rr |>
    dplyr::filter(!is.na(.data$beta), !is.na(.data$se), .data$se > 0) |>
    dplyr::mutate(
      lo = .data$beta - 1.96 * .data$se,
      hi = .data$beta + 1.96 * .data$se,
      row_label = forest_row_label(.data$trait_name, .data$data_type, .data$tissue),
      panel_id = dplyr::coalesce(
        as.character(.data$display_snp),
        paste0("variant ", as.character(.data$variant_id))
      ),
      analysis = "Rare result"
    ) |>
    dplyr::group_by(.data$panel_id) |>
    dplyr::arrange(.data$beta, .by_group = TRUE) |>
    dplyr::mutate(
      row_label = factor(.data$row_label, levels = unique(.data$row_label))
    ) |>
    dplyr::ungroup()
}

trem2_forest <- dplyr::bind_rows(
  trem2_forest_coloc,
  trem2_forest_rare
) |>
  dplyr::mutate(
    panel_id = factor(.data$panel_id, levels = unique(.data$panel_id))
  )

ggplot2::ggplot(trem2_forest, ggplot2::aes(x = beta, y = row_label, colour = analysis)) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, colour = "grey55") +
  ggplot2::geom_errorbarh(
    ggplot2::aes(xmin = lo, xmax = hi),
    height = 0,
    linewidth = 0.35
  ) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_wrap(~ panel_id, scales = "free_y", ncol = 1) +
  ggplot2::labs(
    x = "Effect (beta) with 95% CI",
    y = NULL,
    colour = "Analysis",
    title = "TREM2 — Alzheimer's"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    strip.text = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = "bottom"
  )
```

We now have a good overview of the use of TREM2 as a mechanism to investigate, and potentially target for Alzheimer's disease. We now know:

- TREM2 is associated with Alzheimer's disease
- TREM2 is associated with two other protein coding genes, TREM1 and TREML2, but there is no evidence of colocalisation with any other nearby genes.
- TREM2 is not associated with any other trait categories, or even any other traits.
- The direction of effect between TREM2 and Alzheimer's disease seems to be complex.  In 3 instances, an increase in the expression of TREM2 is associated with an increase in Alzheimer's disease, and in 1 instance, a decrease in the expression of TREM2 (across many gene expression, splice variants, and protein levels) is associated with an increase in Alzheimer's disease.

Existing literature has shown that TREM2 is associated with Alzheimer's disease, and that TREM2 is currently being investigated as a potential target for Alzheimer's disease.

Although the GPMap can not provide a definitive answer as a drug target for Alzheimer's disease, it has proven to helpful in illuminating the potential pleiotropy, and of the directionality of effect that perturbing specific variants associated with TREM2 may have.

# Accessing summary statistics

You can only access the summary statistics via `variant()`, by including `include_summary_stats = TRUE` in the call.  This is to limit the amount of data that can be returned from the API for cost reasons.  If you attempt to download too many summary statistics, you may be blocked from the API.

* `variant(1669064, include_summary_stats = TRUE)`

Instead, if you are interested in accessing all the summary statistics, or the whole database of results, you can download all databases and summary statistics the [Google Cloud Storage Bucket](https://console.cloud.google.com/storage/browser/genotype-phenotype-map), and can find an [explanation of the files here](https://github.com/MRCIEU/genotype-phenotype-map/wiki/Public-Data-Bucket). Note that the bucket is 'requester pays', so you will need to have a Google Cloud account and be logged in to download the files.
