---
title: "Using gpmapr"
output:
  rmarkdown::html_vignette:
    css: vignette-wide.css
vignette: >
  %\VignetteIndexEntry{gpmapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

```{r setup}
library(gpmapr)
```

# Introduction

The gpmapr package provides a set of functions to interact with the GPMap API.

# Getting started

The first port of entry is the `search_gpmap` function.  This function allows you to search for a trait, gene, or variant.

```{r search-trait}
hemoglobin <- search_gpmap("hemoglobin")
hemoglobin
```
You can see information about the different traits that match your search, how to subsequently get more information (under `call`), and metadata about the trait (under `info`).

### Understanding `info`

* Extractions: Each trait has a number of extractions, related to the number of finemapped ld regions that had a p-value < 1.5e-4.
* Colocalisation Groups: The number of colocalisation groups associated with the trait / gene / variant.
* Colocalising Traits: The number of traits that are colocalising with the trait / gene / variant.
* Rare Disease Results: Each trait has a number of rare disease results, related to the number of groups of rare disease results that had a p-value < 1.5e-4.

You can also search for a gene (by ENSG ID or gene name) or variant (by CHR:BP or rsID).

```{r other-searches}
search_gpmap("SLC44A")
search_gpmap("22:37042914")
```

### Variant search returns proxy variants

The variant search will return the original variant and any proxy variants that are in LD with the original variant, which also have colocalisation results.  Each variant is marked as either `original_variant` or `proxy_variant`.  In this example, the original variant has no colocalisation results, but the proxy variant does.  The proxy variant is also marked as having a r^2 of 0.96

```{r proxy-variants}
proxy_variants <- search_gpmap("rs17078078")
proxy_variants
```

# Diving in

You can get more information about a trait by subsequently calling the `trait()`, `gene()`, or `variant()` functions.  This will provide you with the colocalisation results, rare disease results, and study extractions.

Each trait, gene, or variant call have optional parameters:

* `include_associations`: Whether to include associations (BETA, SE, P) of the variants in which it's colocalised in the output.
* `include_coloc_pairs`: Whether to include coloc pairs in the output (which include h3 and h4), as opposed to just the coloc groups.
* `h4_threshold`: The h4 threshold for coloc pairs.

## Trait example

```{r trait-example}
hemoglobin_trait <- trait(931)
names(hemoglobin_trait)
head(hemoglobin_trait$coloc_groups)
```

This returns a list of elements, which you can access by name.  More detailed explanations of the below are described in the API documentation.

* `trait`: A list containing metadata about the trait, including common and rare studies associated with the trait
* `coloc_groups`: a dataframe containing information about which studies have coloc results for this trait.
* `study_extractions`: a list of dataframes containing the study extractions for this trait.
* `rare_results`: (if they exist) a list of dataframes containing the rare results for this trait
* `associations`: (optional) a dataframe containing the associations for the variants in which the trait is colocalised
* `coloc_pairs`: (optional) a dataframe containing all pairwise coloc results for this trait.

### How `study_extractions`, `coloc_groups`, and `rare_results` relate

All three layers refer to **study extractions** (GWAS hits in finemapped LD regions), keyed by `id` in `study_extractions` and by `study_extraction_id` in `coloc_groups` and `rare_results`.

Think of **`study_extractions`** as the catalogue of extractions linked to the trait or gene in the map: each row is a study signal at a locus. Not every listed extraction necessarily has a full colocalisation story or a rare-disease result row.

- **`coloc_groups`** holds extractions that sit inside a **colocalisation group** (shared coloc evidence with other studies/traits at that locus). If a study extraction appears here, it has coloc-structured results for that entity.
- **`rare_results`** holds extractions that belong to a **rare-result group** (rare-disease-style finemapping in the API). The same biological signal can, in principle, be reflected in both rare and common coloc paths, so overlap between this set of extraction ids and those in `coloc_groups` is possible.
- **Neither table:** Some extractions appear only under **`study_extractions`**. They are still tied to the trait or gene in the genotype–phenotype map (for example via **locus tagging, proximity, or other association logic** in the pipeline), but they do not currently carry a coloc-group row or a rare-result group row. Treat those as “present on the map at this locus” rather than “absent by mistake.”

The same idea applies when you call **`gene()`** (or other entity-specific endpoints): `study_extractions` is the wider catalogue, while `coloc_groups` and `rare_results` subdivide it where those analyses exist. The table below illustrates counts for the hemoglobin trait example (`trait(931)`).

```{r trait-extraction-membership, dependson = "trait-example"}
extraction_ids <- function(x) {
  if (is.null(x) || !length(x)) {
    return(integer(0))
  }
  id_col <- if ("id" %in% names(x)) "id" else NULL
  if (!is.null(id_col)) {
    return(unique(stats::na.omit(x[[id_col]])))
  }
  unique(stats::na.omit(unlist(lapply(x, function(df) {
    if (!is.data.frame(df)) {
      return(NULL)
    }
    if ("id" %in% names(df)) {
      return(df[["id"]])
    }
    NULL
  }))))
}

cg_ids <- unique(stats::na.omit(hemoglobin_trait$coloc_groups$study_extraction_id))

rr <- hemoglobin_trait$rare_results
rr_ids <- if (is.null(rr) || !length(rr)) {
  integer(0)
} else {
  unique(stats::na.omit(unlist(lapply(rr, function(d) {
    if (!is.data.frame(d) || !"study_extraction_id" %in% names(d)) {
      return(NULL)
    }
    d[["study_extraction_id"]]
  }))))
}

se_ids <- extraction_ids(hemoglobin_trait$study_extractions)

only_study_extractions <- setdiff(se_ids, union(cg_ids, rr_ids))

data.frame(
  source = c(
    "Distinct study_extraction_id in coloc_groups",
    "Distinct study_extraction_id in rare_results",
    "In study_extractions only (not in coloc_groups nor rare_results)",
    "Total distinct extraction id in study_extractions"
  ),
  n_extractions = c(
    length(cg_ids),
    length(rr_ids),
    length(only_study_extractions),
    length(se_ids)
  )
)
```

# Gene example

```{r gene-example}
slc44a1 <- gene("SLC44A1")
names(slc44a1)
nrow(slc44a1$coloc_groups)
head(slc44a1$coloc_groups[, c("coloc_group_id", "trait_name", "data_type", "tissue")])
```

Note that by default, the gene endpoint includes trans genetic effects. You can turn this off with `gene("SLC44A1", include_trans = FALSE)`, or filter `coloc_groups` after the fact (for example `cis_trans == "cis"` or a `min_p` cutoff).

After you subset rows, some `coloc_group_id` values may be left with only a single row. Those are no longer meaningful colocalisation **groups**, so drop them by keeping only groups with more than one row:

```{r gene-filter-example}
slc44a1_no_trans <- slc44a1
slc44a1_no_trans$coloc_groups <- dplyr::filter(
  slc44a1_no_trans$coloc_groups,
  cis_trans == "cis"
)
# or e.g. dplyr::filter(..., min_p <= 5e-8)

slc44a1_no_trans$coloc_groups <- slc44a1_no_trans$coloc_groups |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::filter(dplyr::n() > 1L) |>
  dplyr::ungroup()

nrow(slc44a1_no_trans$coloc_groups)
head(slc44a1_no_trans$coloc_groups[, c("coloc_group_id", "trait_name", "data_type", "tissue")])
```

## Coloc pairs example

Although coloc groups provide a helpful overview of the colocalisation results for a gene, you may want to look at the pairwise relationships between the studies.  You can do this by setting `include_coloc_pairs = TRUE` in the call.

As there may be many coloc pairs, the results are not as fleshed out as the coloc groups.  You may have to filter the results to get the information you need, or join with the study extractions to get the full picture.

```{r coloc-pairs-example}
slc44a1_coloc_pairs <- gene("SLC44A1", include_coloc_pairs = TRUE)
nrow(slc44a1_coloc_pairs$coloc_pairs)
head(slc44a1_coloc_pairs$coloc_pairs[, c("study_extraction_a_id", "study_extraction_b_id", "h4")])

study_extractions_tbl <- dplyr::bind_rows(slc44a1_no_trans$study_extractions)
study_extractions_subset <- dplyr::select(study_extractions_tbl, id, trait_name, min_p)
slc44a1_coloc_pairs$coloc_pairs |>
  dplyr::left_join(study_extractions_subset, by = c("study_extraction_a_id" = "id")) |>
  dplyr::rename(trait_a = trait_name, min_p_a = min_p) |>
  dplyr::left_join(study_extractions_subset, by = c("study_extraction_b_id" = "id")) |>
  dplyr::rename(trait_b = trait_name, min_p_b = min_p)
```

# Accessing summary statistics

You can only access the summary statistics via `variant()`, by including `include_summary_stats = TRUE` in the call.  This is to limit the amount of data that can be returned from the API for cost reasons.  If you attempt to download too many summary statistics, you may be blocked from the API.

* `variant(1669064, include_summary_stats = TRUE)`

Instead, if you are interested in accessing all the summary statistics, or the whole database of results, you can download all databases and summary statistics the [Google Cloud Storage Bucket](https://console.cloud.google.com/storage/browser/genotype-phenotype-map), and can find an [explanation of the files here](https://github.com/MRCIEU/genotype-phenotype-map/wiki/Public-Data-Bucket). Note that the bucket is 'requester pays', so you will need to have a Google Cloud account and be logged in to download the files.
