---
title: "Investigation of bivariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of bivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  association_source: "coloc"
  min_snp_signals: 5
  louvain_gamma: 2
  compress_method: "asinh"
  similarity_threshold: 0.5
  collapse_gene_tissue: FALSE
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
```

Bivariate analysis: SNPs are oriented per trait, cross-trait similarity `S^*` is
computed, and **signed Louvain** groups SNPs on each side. Molecular QTL rows can
be kept study-level or collapsed to **gene x tissue** features
(`collapse_gene_tissue = TRUE`), which merges cis/trans and QTL types while
preserving tissue labels.

Pipeline (verification aligned with SNP perturbation clustering):

1. Build / filter / orient trait x SNP pleiotropy matrices
2. Report phenotypic vs molecular contribution to profile mass
3. Cross-trait similarity → signed Louvain modules + bivariate blocks
4. Shared-SNP panel (orthogonal): \(S^*\) and concordance restricted to overlapping SNP IDs
5. **Verification:** module vs rest on the **same traits/features used for clustering**
6. Pathway enrichment (secondary) uses molecular QTL genes from `coloc_groups`

**Why module vs rest:** BMI pathway enrichment on the full gene set is often blank.
Asking whether a trait category is elevated in a module relative to all other SNPs
still gives a readable biological check. With gene x tissue collapse, elevated
tables can also surface gene x tissue drivers directly.

We use **BMI** (trait id `1992`) and **Waist-to-hip ratio** (trait id `1994`).

Controllable run parameters (YAML `params` / `rmarkdown::render(..., params = )`):

| Parameter | Values | Meaning |
|---|---|---|
| `api` | `production`, `local` | Public GPMap vs local QTL DB (`localhost:8000`) |
| `association_source` | `coloc`, `full` | Coloc-only associations vs denser `/associations-full` |
| `min_snp_signals` | `1`, `3`, `5` | Sparse-row hold-out threshold for \(S^*\) |
| `louvain_gamma` | `2`, `3` | Louvain resolution |
| `compress_method` | `none`, `asinh` | Soft z-score compression |
| `similarity_threshold` | `0`, `0.5` | Min \|SNP–SNP similarity\| kept as a Louvain edge (`0` = dense) |
| `collapse_gene_tissue` | `FALSE`, `TRUE` | Study-level QTLs vs gene x tissue Stouffer collapse |

```{r params}
trait_id_1 <- 1992
trait_id_2 <- 1994
min_module_size <- 3
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L

api <- params$api
association_source <- params$association_source
min_snp_signals <- as.integer(params$min_snp_signals)
louvain_gamma <- as.numeric(params$louvain_gamma)
compress_method <- params$compress_method
similarity_threshold <- as.numeric(params$similarity_threshold)
collapse_gene_tissue <- as.logical(params$collapse_gene_tissue)

select_api(api)

knitr::kable(
  data.frame(
    api = api,
    gpmap_url = getOption("gpmap_url"),
    association_source = association_source,
    min_snp_signals = min_snp_signals,
    louvain_gamma = louvain_gamma,
    compress_method = compress_method,
    similarity_threshold = similarity_threshold,
    collapse_gene_tissue = collapse_gene_tissue,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

traits_meta <- traits(c(trait_id_1, trait_id_2))
trait_1_name <- traits_meta$traits$trait_name[traits_meta$traits$id == trait_id_1]
trait_2_name <- traits_meta$traits$trait_name[traits_meta$traits$id == trait_id_2]
```

## Step 1: Pleiotropy matrices

`build_bivariate_pleiotropy_matrices()` builds aligned trait x SNP z-score matrices.
`association_source = "coloc"` uses coloc-merged associations only;
`"full"` fills values from `/associations-full` at the same coloc-defined SNP columns.
`collapse_gene_tissue = TRUE` replaces molecular QTL study rows with gene x tissue
features (cis/trans and QTL types combined by weighted Stouffer).

```{r step1-build}
bivariate <- build_bivariate_pleiotropy_matrices(
  trait_id_1 = trait_id_1,
  trait_id_2 = trait_id_2,
  snp_key = "variant_id",
  association_source = association_source,
  collapse_gene_tissue = collapse_gene_tissue
)

X1 <- bivariate$x1_matrix
X2 <- bivariate$x2_matrix
cg <- bivariate$coloc_groups

n_pheno <- sum(bivariate$trait_info$feature_type == "phenotypic", na.rm = TRUE)
n_mol <- sum(bivariate$trait_info$feature_type == "molecular", na.rm = TRUE)

knitr::kable(
  data.frame(
    trait_1 = trait_1_name,
    trait_2 = trait_2_name,
    association_source = bivariate$association_source,
    collapse_gene_tissue = bivariate$collapse_gene_tissue,
    n_rows = nrow(X1),
    n_phenotypic = n_pheno,
    n_molecular = n_mol,
    n_snps_trait_1 = ncol(X1),
    n_snps_trait_2 = ncol(X2)
  ),
  caption = "Pleiotropy matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous traits

Drop shared trait rows with fewer than `min_snp_signals` non-`NA` values in **both**
matrices, or with non-`NA` fraction above `max_snp_fraction` in **either** matrix
(target trait rows kept). The filtered matrices are used for \(S^*\), Louvain, and
trait-vs-rest verification. Molecular QTL genes in `coloc_groups` are still available
for pathway enrichment.

```{r step1-sparse-filter}
protected_ids <- c(as.character(trait_id_1), as.character(trait_id_2))
trait_snp_counts_1 <- rowSums(!is.na(X1))
trait_snp_counts_2 <- rowSums(!is.na(X2))
trait_snp_frac_1 <- trait_snp_counts_1 / ncol(X1)
trait_snp_frac_2 <- trait_snp_counts_2 / ncol(X2)

sparse_trait_ids <- names(trait_snp_counts_1)[
  (trait_snp_counts_1 < min_snp_signals) &
    (trait_snp_counts_2 < min_snp_signals)
]
ubiquitous_trait_ids <- names(trait_snp_frac_1)[
  (trait_snp_frac_1 > max_snp_fraction) |
    (trait_snp_frac_2 > max_snp_fraction)
]
drop_trait_ids <- setdiff(union(sparse_trait_ids, ubiquitous_trait_ids), protected_ids)

keep_rows <- !rownames(X1) %in% drop_trait_ids
X1 <- X1[keep_rows, , drop = FALSE]
X2 <- X2[keep_rows, , drop = FALSE]
bivariate$trait_info <- bivariate$trait_info |>
  dplyr::filter(!as.character(trait_id) %in% drop_trait_ids)

knitr::kable(
  data.frame(
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    n_sparse_traits_removed = length(setdiff(sparse_trait_ids, protected_ids)),
    n_ubiquitous_traits_removed = length(setdiff(ubiquitous_trait_ids, protected_ids)),
    n_traits_after = nrow(X1),
    n_phenotypic_after = sum(
      bivariate$trait_info$feature_type == "phenotypic",
      na.rm = TRUE
    ),
    n_molecular_after = sum(
      bivariate$trait_info$feature_type == "molecular",
      na.rm = TRUE
    ),
    n_snps_trait_1 = ncol(X1),
    n_snps_trait_2 = ncol(X2)
  ),
  caption = "After sparse / ubiquitous trait filter",
  align = "r"
)
```

## Step 1.6: Phenotypic vs molecular contribution

How much of each SNP profile is carried by phenotypic traits versus molecular QTL
rows? `frac_squared_mass` is the share of total \(\sum z^2\);
`mean_snp_frac_mass` averages that share within each SNP column.

```{r step1-contribution}
contrib_1 <- summarise_feature_type_contribution(X1, bivariate$trait_info)
contrib_2 <- summarise_feature_type_contribution(X2, bivariate$trait_info)

knitr::kable(
  contrib_1 |>
    dplyr::mutate(
      mean_abs_z = round(mean_abs_z, 3),
      squared_mass = signif(squared_mass, 4),
      frac_squared_mass = round(frac_squared_mass, 3),
      mean_snp_frac_mass = round(mean_snp_frac_mass, 3)
    ),
  caption = paste0(trait_1_name, ": phenotypic vs molecular contribution"),
  align = "r",
  row.names = FALSE
)

knitr::kable(
  contrib_2 |>
    dplyr::mutate(
      mean_abs_z = round(mean_abs_z, 3),
      squared_mass = signif(squared_mass, 4),
      frac_squared_mass = round(frac_squared_mass, 3),
      mean_snp_frac_mass = round(mean_snp_frac_mass, 3)
    ),
  caption = paste0(trait_2_name, ": phenotypic vs molecular contribution"),
  align = "r",
  row.names = FALSE
)
```

## Step 2: Orient and compress profiles (X₁\*, X₂\*)

Flip each SNP column by \(\mathrm{sign}(z_{\mathrm{target}})\) for that trait, then
soft-compress with \(\mathrm{asinh}(z)\) so extreme values do not dominate
similarity or trait rankings (same helper as SNP perturbation clustering).

```{r step2-orient}
X1_star <- orient_pleiotropy_matrix(
  X1,
  target_trait_id = trait_id_1
)$x_matrix

X2_star <- orient_pleiotropy_matrix(
  X2,
  target_trait_id = trait_id_2
)$x_matrix

X1_star <- compress_effect_matrix(X1_star, method = compress_method)
X2_star <- compress_effect_matrix(X2_star, method = compress_method)

# SNPs x traits for verification (same features used for clustering)
trait_matrix_1 <- t(X1_star)
trait_matrix_2 <- t(X2_star)
```

## Step 3: Cross-trait similarity (S\*)

```{r step3-similarity}
S_star <- cross_snp_similarity_matrix(X1_star, X2_star)$s_matrix

knitr::kable(
  data.frame(
    dim = paste(nrow(S_star), "x", ncol(S_star)),
    min_similarity = round(min(S_star, na.rm = TRUE), 2),
    max_similarity = round(max(S_star, na.rm = TRUE), 2),
    mean_similarity = round(mean(S_star, na.rm = TRUE), 2)
  ),
  caption = "Cross-trait SNP similarity summary",
  align = "r"
)
```

## Step 4: Signed Louvain modules

Louvain runs on cosine similarity of cross-trait **profiles** (rows and columns of `S^*`).
If `similarity_threshold > 0`, edges with `|similarity| < similarity_threshold` are
zeroed before clustering (`0` keeps the dense SNP–SNP graph).

```{r step4-cluster}
cross_modules <- cluster_cross_trait_snps(
  S_star,
  method = "louvain",
  similarity_threshold = similarity_threshold,
  gamma = louvain_gamma
)

knitr::kable(
  dplyr::bind_rows(
    data.frame(
      trait = trait_1_name,
      cluster = as.integer(names(table(cross_modules$clusters_rows))),
      n_snps = as.integer(table(cross_modules$clusters_rows)),
      stringsAsFactors = FALSE
    ),
    data.frame(
      trait = trait_2_name,
      cluster = as.integer(names(table(cross_modules$clusters_cols))),
      n_snps = as.integer(table(cross_modules$clusters_cols)),
      stringsAsFactors = FALSE
    )
  ) |>
    dplyr::filter(n_snps >= min_module_size) |>
    dplyr::arrange(trait, dplyr::desc(n_snps)),
  caption = "Signed Louvain module sizes",
  align = c("l", "r", "r")
)
```

Heatmap shows only SNPs in modules with more than `min_module_size` SNPs,
ordered by module id (tiny modules omitted).

```{r step4-heatmap, fig.height = 7, fig.width = 9}
row_sizes <- table(cross_modules$clusters_rows)
col_sizes <- table(cross_modules$clusters_cols)
large_row_ids <- names(row_sizes)[row_sizes > min_module_size]
large_col_ids <- names(col_sizes)[col_sizes > min_module_size]

keep_rows <- names(cross_modules$clusters_rows)[
  as.character(cross_modules$clusters_rows) %in% large_row_ids
]
keep_cols <- names(cross_modules$clusters_cols)[
  as.character(cross_modules$clusters_cols) %in% large_col_ids
]

if (length(keep_rows) >= 2 && length(keep_cols) >= 2) {
  row_ord <- keep_rows[order(cross_modules$clusters_rows[keep_rows])]
  col_ord <- keep_cols[order(cross_modules$clusters_cols[keep_cols])]
  S_plot <- S_star[row_ord, col_ord, drop = FALSE]

  knitr::kable(
    data.frame(
      n_row_snps_plotted = length(row_ord),
      n_col_snps_plotted = length(col_ord),
      n_large_row_modules = length(large_row_ids),
      n_large_col_modules = length(large_col_ids)
    ),
    caption = paste0(
      "Heatmap subset (modules > ", min_module_size, " SNPs)"
    ),
    align = "r"
  )

  heatmap(
    S_plot,
    Rowv = NA,
    Colv = NA,
    scale = "none",
    main = paste0(
      "Cross-trait S* (modules > ", min_module_size, " SNPs)"
    ),
    xlab = paste(trait_2_name, "SNPs"),
    ylab = paste(trait_1_name, "SNPs")
  )
} else {
  cat(
    "Fewer than 2 SNPs in large modules on one or both sides; ",
    "skipping heatmap.\n",
    sep = ""
  )
}
```

## Step 4.5: Shared-SNP concordance (orthogonal)

The full rectangular \(S^*\) asks whether **programs** of trait-1 SNPs align with
programs of trait-2 SNPs. A more direct answer to "how are these two traits
related?" uses only **overlapping SNP IDs** (present as columns on both sides):

1. Sign concordance of target-trait effects at those SNPs
2. Same-SNP phenome-profile cosine (diagonal of \(S^*_{\mathrm{shared}}\))
3. Full \(S^*_{\mathrm{shared}}\) among shared SNPs, with optional Louvain blocks

```{r step4-shared-snps}
shared_snp_ids <- intersect(colnames(X1_star), colnames(X2_star))

knitr::kable(
  data.frame(
    n_snps_trait_1 = ncol(X1_star),
    n_snps_trait_2 = ncol(X2_star),
    n_shared_snps = length(shared_snp_ids),
    frac_shared_of_trait_1 = round(length(shared_snp_ids) / ncol(X1_star), 3),
    frac_shared_of_trait_2 = round(length(shared_snp_ids) / ncol(X2_star), 3)
  ),
  caption = "Overlapping SNP IDs between trait sides",
  align = "r"
)

if (length(shared_snp_ids) == 0) {
  S_shared <- NULL
  shared_snp_table <- data.frame()
  cat("No overlapping SNP IDs; shared-SNP concordance skipped.\n")
} else {
  X1_shared <- X1_star[, shared_snp_ids, drop = FALSE]
  X2_shared <- X2_star[, shared_snp_ids, drop = FALSE]

  # Target-effect signs from pre-orientation matrices (orientation forces
  # target z >= 0, so oriented signs are not informative for concordance).
  z1_target <- as.numeric(X1[as.character(trait_id_1), shared_snp_ids])
  z2_target <- as.numeric(X2[as.character(trait_id_2), shared_snp_ids])
  names(z1_target) <- shared_snp_ids
  names(z2_target) <- shared_snp_ids

  both_obs <- is.finite(z1_target) & is.finite(z2_target) &
    z1_target != 0 & z2_target != 0
  sign_concordant <- sign(z1_target[both_obs]) == sign(z2_target[both_obs])

  S_shared <- cross_snp_similarity_matrix(X1_shared, X2_shared)$s_matrix
  # Same column order on both sides => diagonal = same-SNP profile cosine
  same_snp_profile_cos <- diag(S_shared[shared_snp_ids, shared_snp_ids, drop = FALSE])

  shared_snp_table <- data.frame(
    snp_id = shared_snp_ids,
    z_trait_1 = z1_target,
    z_trait_2 = z2_target,
    sign_concordant = dplyr::case_when(
      !is.finite(z1_target) | !is.finite(z2_target) ~ NA,
      z1_target == 0 | z2_target == 0 ~ NA,
      TRUE ~ sign(z1_target) == sign(z2_target)
    ),
    profile_cosine = as.numeric(same_snp_profile_cos),
    stringsAsFactors = FALSE
  )

  knitr::kable(
    data.frame(
      n_with_both_target_z = sum(both_obs),
      frac_sign_concordant = if (sum(both_obs) > 0) {
        round(mean(sign_concordant), 3)
      } else {
        NA_real_
      },
      mean_same_snp_profile_cos = round(mean(same_snp_profile_cos, na.rm = TRUE), 3),
      median_same_snp_profile_cos = round(
        stats::median(same_snp_profile_cos, na.rm = TRUE),
        3
      ),
      frac_profile_cos_gt_0.5 = round(
        mean(same_snp_profile_cos > 0.5, na.rm = TRUE),
        3
      ),
      frac_profile_cos_lt_m0.5 = round(
        mean(same_snp_profile_cos < -0.5, na.rm = TRUE),
        3
      )
    ),
    caption = "Shared-SNP concordance summary",
    align = "r"
  )

  knitr::kable(
    shared_snp_table |>
      dplyr::arrange(dplyr::desc(abs(profile_cosine))) |>
      dplyr::mutate(
        z_trait_1 = round(z_trait_1, 3),
        z_trait_2 = round(z_trait_2, 3),
        profile_cosine = round(profile_cosine, 3)
      ),
    caption = "Shared SNPs by |same-SNP profile cosine|",
    row.names = FALSE
  )
}
```

```{r step4-shared-heatmap, fig.height = 7, fig.width = 7}
if (!is.null(S_shared) && length(shared_snp_ids) >= 2) {
  shared_modules <- cluster_cross_trait_snps(
    S_shared,
    method = "louvain",
    similarity_threshold = similarity_threshold,
    gamma = louvain_gamma
  )
  heatmap(
    S_shared,
    Rowv = as.dendrogram(shared_modules$hclust_rows),
    Colv = as.dendrogram(shared_modules$hclust_cols),
    scale = "none",
    main = "Shared-SNP S* (Louvain ordering)",
    xlab = paste(trait_2_name, "shared SNPs"),
    ylab = paste(trait_1_name, "shared SNPs")
  )

  shared_blocks <- summarise_cross_trait_modules(
    S_shared,
    clusters_rows = shared_modules$clusters_rows,
    clusters_cols = shared_modules$clusters_cols
  ) |>
    dplyr::filter(
      n_snps_trait_1 > min_module_size,
      n_snps_trait_2 > min_module_size
    ) |>
    dplyr::mutate(
      block_score = abs(mean_similarity) * sqrt(n_snps_trait_1 * n_snps_trait_2)
    )

  knitr::kable(
    shared_blocks |>
      dplyr::count(block_type, name = "n_blocks") |>
      dplyr::arrange(dplyr::desc(n_blocks)),
    caption = paste0(
      "Shared-SNP block counts (both sides > ", min_module_size, " SNPs)"
    ),
    align = "lr"
  )

  if (nrow(shared_blocks) > 0) {
    knitr::kable(
      shared_blocks |>
        dplyr::group_by(block_type) |>
        dplyr::slice_max(block_score, n = 1, with_ties = FALSE) |>
        dplyr::ungroup() |>
        dplyr::mutate(
          mean_similarity = round(mean_similarity, 2),
          block_score = round(block_score, 2)
        ) |>
        dplyr::select(
          block_type, row_cluster, col_cluster,
          mean_similarity, block_score, n_snps_trait_1, n_snps_trait_2
        ),
      caption = "Top-scoring shared-SNP block of each type",
      align = "r",
      row.names = FALSE
    )
  }
} else {
  cat("Fewer than 2 shared SNPs; skipping shared S* heatmap / blocks.\n")
}
```

## Step 5: Bivariate blocks (all SNPs)

Each heatmap cell block pairs one Louvain module of trait-1 SNPs with one module of trait-2 SNPs. Mean block similarity classifies the relationship:

- **Concordant** — shared same-direction pleiotropy
- **Discordant** — opposing pleiotropy (trade-off)
- **Trait-specific** — weak cross-trait alignment

Block counts and “strongest” examples use only blocks where **both** sides have `> min_module_size` SNPs. Within each type, blocks are ranked by
`|mean_similarity| * sqrt(n1 * n2)` so tiny extreme cells do not dominate.

```{r step5-blocks}
module_summary <- summarise_cross_trait_modules(
  S_star,
  clusters_rows = cross_modules$clusters_rows,
  clusters_cols = cross_modules$clusters_cols
)

large_blocks <- module_summary |>
  dplyr::filter(
    n_snps_trait_1 > min_module_size,
    n_snps_trait_2 > min_module_size
  ) |>
  dplyr::mutate(
    block_score = abs(mean_similarity) * sqrt(n_snps_trait_1 * n_snps_trait_2)
  )

knitr::kable(
  large_blocks |>
    dplyr::count(block_type, name = "n_blocks") |>
    dplyr::arrange(dplyr::desc(n_blocks)),
  caption = paste0(
    "Block counts by type (both sides > ", min_module_size, " SNPs)"
  ),
  align = "lr"
)

blocks_of_interest <- large_blocks |>
  dplyr::filter(block_type %in% c("concordant", "discordant", "trait_specific")) |>
  dplyr::group_by(block_type) |>
  dplyr::slice_max(block_score, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

if (nrow(blocks_of_interest) == 0) {
  cat(
    "No blocks with both sides > ", min_module_size,
    " SNPs for concordant / discordant / trait_specific.\n",
    sep = ""
  )
} else {
  knitr::kable(
    blocks_of_interest |>
      dplyr::mutate(
        mean_similarity = round(mean_similarity, 2),
        block_score = round(block_score, 2)
      ) |>
      dplyr::select(
        block_type, row_cluster, col_cluster,
        mean_similarity, block_score, n_snps_trait_1, n_snps_trait_2
      ),
    caption = "Top-scoring large block of each type",
    align = "r",
    row.names = FALSE
  )
}
```

## Verification: module vs rest (clustering traits)

For each large Louvain module, ask whether traits / features present in the
**clustering matrix** are elevated relative to all other SNPs on that trait side
(`summarise_module_specific_traits()`). With `collapse_gene_tissue = TRUE`,
molecular rows appear as `gene | tissue` labels. Pathway enrichment from
`coloc_groups` remains a secondary check below.

```{r verify-traits-1}
group_traits_1 <- summarise_module_specific_traits(
  trait_matrix = trait_matrix_1,
  groups = cross_modules$clusters_rows,
  trait_info = bivariate$trait_info,
  coloc_groups = cg,
  exclude_trait_ids = c(trait_id_1, trait_id_2),
  min_group_size = min_group_size,
  top_n = top_n_traits,
  n_categories = 2L
)

knitr::kable(
  group_traits_1$summary |>
    dplyr::select(
      group, n_snps, top_trait, top_specificity, top_trait_categories
    ) |>
    dplyr::mutate(top_specificity = round(top_specificity, 2)),
  caption = paste0(
    trait_1_name, ": categories/traits elevated vs rest (> ",
    min_group_size, " SNPs)"
  ),
  na = "",
  row.names = FALSE
)
```

```{r verify-trait1-details, echo=FALSE, results='asis'}
for (x in group_traits_1$by_group) {
  cat("\n\n### ", trait_1_name, " module ", x$group, "\n\n", sep = "")
  cat("SNPs: ", x$n_snps, "\n\n", sep = "")
  if (nrow(x$categories) > 0) {
    print(knitr::kable(
      x$categories |>
        dplyr::mutate(
          score = round(score, 2),
          mean_specificity = round(mean_specificity, 2)
        ),
      caption = "Trait categories elevated vs rest",
      row.names = FALSE
    ))
  }
  if (nrow(x$traits) > 0) {
    print(knitr::kable(
      x$traits |>
        dplyr::select(
          trait_name, trait_category, abs_mean_z, abs_mean_z_rest,
          specificity, abs_mean_z_delta
        ) |>
        dplyr::mutate(
          abs_mean_z = round(abs_mean_z, 3),
          abs_mean_z_rest = round(abs_mean_z_rest, 3),
          specificity = round(specificity, 2),
          abs_mean_z_delta = round(abs_mean_z_delta, 3)
        ),
      caption = "Traits elevated vs rest",
      row.names = FALSE
    ))
  }
}
```

```{r verify-traits-2}
group_traits_2 <- summarise_module_specific_traits(
  trait_matrix = trait_matrix_2,
  groups = cross_modules$clusters_cols,
  trait_info = bivariate$trait_info,
  coloc_groups = cg,
  exclude_trait_ids = c(trait_id_1, trait_id_2),
  min_group_size = min_group_size,
  top_n = top_n_traits,
  n_categories = 2L
)

knitr::kable(
  group_traits_2$summary |>
    dplyr::select(
      group, n_snps, top_trait, top_specificity, top_trait_categories
    ) |>
    dplyr::mutate(top_specificity = round(top_specificity, 2)),
  caption = paste0(
    trait_2_name, ": categories/traits elevated vs rest (> ",
    min_group_size, " SNPs)"
  ),
  na = "",
  row.names = FALSE
)
```

```{r verify-trait2-details, echo=FALSE, results='asis'}
for (x in group_traits_2$by_group) {
  cat("\n\n### ", trait_2_name, " module ", x$group, "\n\n", sep = "")
  cat("SNPs: ", x$n_snps, "\n\n", sep = "")
  if (nrow(x$categories) > 0) {
    print(knitr::kable(
      x$categories |>
        dplyr::mutate(
          score = round(score, 2),
          mean_specificity = round(mean_specificity, 2)
        ),
      caption = "Trait categories elevated vs rest",
      row.names = FALSE
    ))
  }
  if (nrow(x$traits) > 0) {
    print(knitr::kable(
      x$traits |>
        dplyr::select(
          trait_name, trait_category, abs_mean_z, abs_mean_z_rest,
          specificity, abs_mean_z_delta
        ) |>
        dplyr::mutate(
          abs_mean_z = round(abs_mean_z, 3),
          abs_mean_z_rest = round(abs_mean_z_rest, 3),
          specificity = round(specificity, 2),
          abs_mean_z_delta = round(abs_mean_z_delta, 3)
        ),
      caption = "Traits elevated vs rest",
      row.names = FALSE
    ))
  }
}
```

## Secondary check: pathway enrichment (QTL genes)

Pathway enrichment uses molecular QTL genes linked to module SNPs via
`coloc_groups`. It is often empty for the full BMI gene set, so it is secondary.

```{r verify-pathways-secondary}
baseline_1 <- enrich_trait_pathways(
  trait_id = trait_id_1,
  coloc_groups = cg,
  sources = c("KEGG", "Reactome")
)
baseline_2 <- enrich_trait_pathways(
  trait_id = trait_id_2,
  coloc_groups = cg,
  sources = c("KEGG", "Reactome")
)
trait_1_enrichment <- enrich_snp_group_pathways(
  groups = cross_modules$clusters_rows,
  coloc_groups = cg,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)
trait_2_enrichment <- enrich_snp_group_pathways(
  groups = cross_modules$clusters_cols,
  coloc_groups = cg,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)

knitr::kable(
  dplyr::bind_rows(baseline_1$summary, baseline_2$summary),
  caption = "Baseline pathway enrichment (secondary)",
  na = "",
  row.names = FALSE
)

knitr::kable(
  trait_1_enrichment$summary,
  caption = paste0(trait_1_name, ": per-module pathway enrichment (secondary)"),
  na = "",
  row.names = FALSE
)

knitr::kable(
  trait_2_enrichment$summary,
  caption = paste0(trait_2_name, ": per-module pathway enrichment (secondary)"),
  na = "",
  row.names = FALSE
)
```

## Comparison notes

| Choice | Bivariate (this vignette) | EBMF |
|---|---|---|
| Input | Shared background traits x SNPs (both sides); optional gene x tissue QTL collapse | Phenotypic traits + gene-collapsed QTLs |
| Similarity / model | Rectangular cross-trait \(S^*\) (+ shared-SNP \(S^*\)) | Sparse matrix factorization |
| Clustering | Signed Louvain (each side) | Soft programs (`flashier`) |
| Extra | All-SNP blocks + shared-SNP concordance panel | Automated \(K\) via ELBO |
| Verification | Module vs rest on clustering traits; QTL pathways secondary | Program drivers + trait verification |
