---
title: "Simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  cluster_type: ebmf
  K: 3
  module_sizes: [5, 20, 50]
  n_traits: 600
  n_snps: 300
  p_molecular: 0.8
  p_active_molecular: 0.02
  min_snp_signals: 2
  min_module_size: 3
  compress_method: asinh
  compress_scale: 2
  similarity_threshold: 0.2
  ebmf_se_mode: unit
  ebmf_beta_scale: trait
  ebmf_lfsr_threshold: 0.05
  ebmf_magnitude_threshold: 0.5
  ebmf_drop_global: true
  ebmf_prior: point_normal
  ebmf_backfit: true
  ebmf_hard_assignment: false
  min_mean_internal: 0.3
  min_connectedness: 0.25
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 5
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  drivers_per_module: 12
  effect_size: 3
  p_structural_zero: 0.1
  p_spurious: 0.01
  p_active_background: 0.05
  driver_sharing: 0
  snp_driver_groups: 1
  n_reps: 5
vignette: >
  %\VignetteIndexEntry{Simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

low_noise_sd <- 0.5
moderate_noise_sd <- 0.8
high_noise_sd <- 1.2

n_reps <- as.integer(params$n_reps)
set.seed(1)

if (!identical(params$cluster_type, "ebmf")) {
  stop("This vignette is configured for cluster_type = 'ebmf'")
}
if (params$K != length(params$module_sizes)) {
  stop("K must equal length(module_sizes)")
}

clustering_settings <- list(
  cluster_type = params$cluster_type,
  min_snp_signals = params$min_snp_signals,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  similarity_threshold = params$similarity_threshold,
  ebmf_se_mode = params$ebmf_se_mode,
  ebmf_beta_scale = params$ebmf_beta_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  ebmf_hard_assignment = params$ebmf_hard_assignment,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  n_null = params$n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = params$n_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)

simulation_defaults <- list(
  module_sizes = params$module_sizes,
  drivers_per_module = params$drivers_per_module,
  n_traits = params$n_traits,
  p_molecular = params$p_molecular,
  p_active_molecular = params$p_active_molecular,
  effect_size = params$effect_size,
  p_structural_zero = params$p_structural_zero,
  p_spurious = params$p_spurious,
  p_active_background = params$p_active_background,
  driver_sharing = params$driver_sharing,
  snp_driver_groups = params$snp_driver_groups
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against synthetic data with known ground truth. It is organised around **input
versions** rather than generative parameter axes: the same simulation, EBMF,
and program-validation pipeline is run on four versions of the input that span
the difficulty range, from clean modules to a pure null.

Each version is generated by `simulate_trait()`, which produces a trait object
(the target trait plus driver and background traits) with known ground truth.
The total number of SNPs is set via `n_coloc_groups`, and the total number of
traits (rows) via `n_traits`; a fraction `p_molecular` of the non-target traits
is molecular (gene-annotated eQTL-like traits) and observed with per-SNP
probability `p_active_molecular`, which is much sparser than the phenotype
traits' density.

For every version, the identical battery is run: `run_univariate_clustering()`
fits EBMF and keeps overlapping soft programs, and
`summarise_ebmf_programs()` filters programs by lFSR, loading magnitude,
internal coherence, and trait-subsampling stability. The key output is a
per-program table in which **each validation check is its own column**, so it
is immediately visible which check failed for each program:

- `size` — at least `min_module_size` filtered SNPs (`n_snps_filtered`).
- `internal_sim` — mean internal SNP similarity `>= min_mean_internal`.
- `connectivity` — pair connectedness `>= min_connectedness`.
- `stability` — top-loading SNPs recovered with replication
  `>= stability_threshold` under trait subsampling.

A program is `valid` only when every check passes. The analysis settings below
are fixed across versions; only the generative input changes.

```{r settings-table}
settings_to_show <- c(
  "cluster_type", "min_snp_signals", "min_module_size", "module_sizes",
  "n_traits", "p_molecular", "p_active_molecular",
  "ebmf_lfsr_threshold", "ebmf_magnitude_threshold",
  "min_mean_internal", "min_connectedness", "n_rep",
  "frac_traits", "stability_top_n", "stability_threshold"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$module_sizes <- params$module_sizes
settings_values$n_traits <- params$n_traits
settings_values$p_molecular <- params$p_molecular
settings_values$p_active_molecular <- params$p_active_molecular
settings_values$stability_top_n <- validation_settings$top_n
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      if (nm == "module_sizes") {
        return(paste(settings_values[[nm]], collapse = ", "))
      }
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "Fixed EBMF analysis and validation settings"
)
```

```{r tier-helpers}
make_tier_spec <- function(...) {
  overrides <- list(...)
  spec <- utils::modifyList(simulation_defaults, overrides)
  if (is.null(spec$n_coloc_groups)) {
    spec$n_coloc_groups <- params$n_snps
  }
  if (is.null(spec$K)) {
    spec$K <- length(spec$module_sizes)
  }
  if (spec$K == 0) {
    spec$module_sizes <- NULL
  }
  return(spec)
}

run_tier_one <- function(spec, rep) {
  dots <- spec
  dots$seed <- sample.int(.Machine$integer.max, 1)
  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  smry <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), validation_settings)
  )
  valid_ids <- smry$programs$program[smry$programs$status == "valid"]
  valid_assigned <- smry$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    res,
    predicted_memberships = valid_assigned
  )
  programs <- smry$programs
  programs$rep <- rep(rep, nrow(programs))
  list(
    summary = data.frame(
      k_planted = ev$k_planted,
      k_hat = ev$k_hat,
      k_error = ev$k_error,
      ari_all = ev$ari_all,
      ari_structured = ev$ari_structured,
      module_recall = ev$mean_module_recall,
      background_absorbed = ev$background_absorbed,
      coverage = ev$coverage,
      n_programs = nrow(programs),
      n_valid_programs = length(valid_ids),
      stringsAsFactors = FALSE
    ),
    programs = programs
  )
}

run_tier <- function(name, label, spec, n_reps) {
  out <- lapply(seq_len(n_reps), function(i) run_tier_one(spec, i))
  list(
    summary = dplyr::bind_rows(lapply(out, `[[`, "summary")) |>
      dplyr::mutate(tier = label),
    programs = dplyr::bind_rows(lapply(out, `[[`, "programs")) |>
      dplyr::mutate(tier = label, tier_name = name)
  )
}
```

## Input versions

The four versions differ only in the generative parameters passed to
`simulate_trait()`. The **null** version keeps only `n_coloc_groups` (SNPs)
and `n_traits`; `K = 0` means no modules are planted and the driver/module
parameters are unused.

```{r tiers}
tier_specs <- list(
  easy = make_tier_spec(
    n_coloc_groups = params$n_snps,
    overlap = "disjoint",
    sign_pattern = "coherent",
    noise_sd = low_noise_sd,
    p_structural_zero = 0.05,
    p_spurious = 0.01,
    p_molecular = 0,
    p_active_background = 0.03,
    driver_sharing = 0,
    snp_driver_groups = 1
  ),
  moderate = make_tier_spec(
    n_coloc_groups = params$n_snps,
    overlap = "moderate",
    sign_pattern = "coherent",
    noise_sd = moderate_noise_sd,
    p_structural_zero = params$p_structural_zero,
    p_spurious = params$p_spurious,
    p_molecular = params$p_molecular,
    p_active_molecular = params$p_active_molecular,
    p_active_background = params$p_active_background,
    driver_sharing = 0,
    snp_driver_groups = 1
  ),
  noisy = make_tier_spec(
    n_coloc_groups = params$n_snps,
    overlap = "strong",
    sign_pattern = "random",
    noise_sd = high_noise_sd,
    p_structural_zero = 0.6,
    p_spurious = 0.1,
    p_molecular = 0.7,
    p_active_molecular = 0.005,
    p_active_background = 0.15,
    driver_sharing = 0.5,
    snp_driver_groups = 2
  ),
  null = make_tier_spec(
    n_coloc_groups = params$n_snps,
    K = 0,
    noise_sd = moderate_noise_sd,
    p_molecular = params$p_molecular,
    p_active_molecular = params$p_active_molecular,
    p_active_background = params$p_active_background
  )
)

tier_labels <- c(
  easy = "easy / clean modules",
  moderate = "moderate / slightly noisy",
  noisy = "very noisy",
  null = "null / no modules"
)

tier_results <- lapply(names(tier_specs), function(nm) {
  run_tier(nm, tier_labels[[nm]], tier_specs[[nm]], n_reps)
})
names(tier_results) <- names(tier_specs)

tier_summary <- dplyr::bind_rows(lapply(tier_results, `[[`, "summary"))
program_checks <- dplyr::bind_rows(lapply(tier_results, `[[`, "programs"))
```

```{r tier-params-table}
tier_param_table <- do.call(rbind, lapply(names(tier_specs), function(nm) {
  s <- tier_specs[[nm]]
  data.frame(
    version = tier_labels[[nm]],
    n_snps = s$n_coloc_groups,
    K = s$K,
    module_sizes = if (s$K > 0) paste(s$module_sizes, collapse = ", ") else "-",
    overlap = if (s$K > 0) s$overlap else "-",
    noise_sd = s$noise_sd,
    p_structural_zero = if (s$K > 0) s$p_structural_zero else "-",
    p_spurious = s$p_spurious,
    p_molecular = s$p_molecular,
    p_active_molecular = s$p_active_molecular,
    p_active_background = s$p_active_background,
    driver_sharing = s$driver_sharing,
    sign_pattern = if (s$K > 0) s$sign_pattern else "-",
    snp_driver_groups = if (s$K > 0) s$snp_driver_groups else "-",
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  tier_param_table,
  row.names = FALSE,
  caption = "Generative parameters for each input version"
)
```

## Aggregate results

Across all replicates and programs of each version, the table below reports
the recovery metrics and, for every validation check, the fraction of programs
that passed it. `valid` is the fraction of programs that passed all checks.

```{r aggregate}
check_rates <- program_checks |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    n_programs = dplyr::n(),
    pct_size = 100 * mean(size_pass),
    pct_internal_similarity = 100 * mean(internal_similarity_pass),
    pct_connectedness = 100 * mean(connectedness_pass),
    pct_stability = if (any(stability_checked)) {
      100 * mean(stability_pass[stability_checked])
    } else {
      NA_real_
    },
    pct_valid = 100 * mean(status == "valid"),
    .groups = "drop"
  )

aggregate_results <- tier_summary |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    mean_k_hat = mean(k_hat),
    mean_valid_programs = mean(n_valid_programs),
    mean_module_recall = mean(module_recall),
    mean_coverage = mean(coverage),
    mean_ari_structured = mean(ari_structured),
    mean_background_absorbed = mean(background_absorbed),
    .groups = "drop"
  ) |>
  dplyr::left_join(check_rates, by = "tier") |>
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric),
    ~ ifelse(is.nan(.x), NA_real_, .x)
  ))

aggregate_results |>
  dplyr::mutate(dplyr::across(
    dplyr::starts_with("pct_"),
    ~ ifelse(is.na(.x), "-", paste0(round(.x), "%"))
  )) |>
  knitr::kable(
    digits = 2,
    col.names = c(
      "version", "k_hat", "valid programs", "module recall", "coverage",
      "ARI (structured)", "background absorbed", "n programs",
      "size", "internal sim", "connectivity", "stability", "valid"
    ),
    caption = "Aggregate recovery and per-check pass rates by input version"
  )
```

The `stability` pass rate is computed only over programs that were actually
re-checked (those passing the size and internal-coherence checks).

## Program-level validation checks

For each version, every program found across the replicates is shown with the
four validation checks as separate columns. A check is marked `FAIL` when the
program does not meet that criterion; `n/a` under `stability` means the
program already failed the internal-coherence checks, so stability was not
re-tested.

```{r check-helper, include=FALSE}
check_table <- function(which_tier) {
  df <- program_checks |>
    dplyr::filter(.data$tier_name == which_tier)
  if (nrow(df) == 0) {
    return(NULL)
  }
  df |>
    dplyr::mutate(
      check_size = ifelse(size_pass, "pass", "FAIL"),
      check_internal_similarity = ifelse(internal_similarity_pass, "pass", "FAIL"),
      check_connectedness = ifelse(connectedness_pass, "pass", "FAIL"),
      check_stability = dplyr::case_when(
        !stability_checked ~ "n/a",
        stability_pass ~ "pass",
        TRUE ~ "FAIL"
      ),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      replication = round(replication, 3)
    ) |>
    dplyr::select(
      rep, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, replication,
      check_size, check_internal_similarity, check_connectedness,
      check_stability, status
    ) |>
    dplyr::arrange(rep, program) |>
    knitr::kable(
      row.names = FALSE,
      col.names = c(
        "rep", "program", "n_snps", "n_snps filtered",
        "mean internal similarity", "connectedness", "replication",
        "size", "internal sim", "connectivity", "stability", "status"
      )
    )
}
```

### Easy / clean modules

Clean generation: low observation noise, few structural zeros and spurious
effects, no molecular traits, coherent driver signs and disjoint modules. The
planted modules should be recovered with all checks passing.

```{r easy-checks}
easy_checks <- check_table("easy")
if (is.null(easy_checks)) {
  "No EBMF programs were produced for the easy version."
} else {
  easy_checks
}
```

### Moderate / slightly noisy

The default generation: moderate noise, structural zeros in one third of the
module support, a third of traits molecular and sparse, and moderate module
overlap. Some programs are expected to fail individual checks, revealing which
criterion is most sensitive to this level of noise.

```{r moderate-checks}
moderate_checks <- check_table("moderate")
if (is.null(moderate_checks)) {
  "No EBMF programs were produced for the moderate version."
} else {
  moderate_checks
}
```

### Very noisy

Heavy observation noise, most traits molecular and extremely sparse, strong
module overlap, shared drivers and random driver signs. Program recovery is
expected to break down, with most or all checks failing.

```{r noisy-checks}
noisy_checks <- check_table("noisy")
if (is.null(noisy_checks)) {
  "No EBMF programs were produced for the noisy version."
} else {
  noisy_checks
}
```

### Null / no modules

`K = 0`: only the number of SNPs (`n_coloc_groups`) and traits (`n_traits`)
are used; all traits are unstructured background. EBMF should not produce any
program that passes the checks, i.e. no false structure.

```{r null-checks}
null_checks <- check_table("null")
if (is.null(null_checks)) {
  "No EBMF programs were produced for the null version."
} else {
  null_checks
}
```

## Notes for interpreting runs

- `module_recall` and `coverage` preserve overlapping valid EBMF memberships.
- `ari_all` and `ari_structured` use each SNP's strongest assigned program as a
  hard projection, so they remain conventional partition metrics.
- `factor_strength` and `factor_strength_per_snp` are reported scores only;
  they do not currently filter programs.
- lFSR and magnitude thresholds determine `assigned` SNPs before coherence,
  factor-strength, and posterior-evidence calculations.
- Stability is run only for programs that pass the size and internal-coherence
  checks; the `stability` column shows `n/a` for programs that were not
  re-checked.
- Molecular traits are gene-annotated and much sparser than phenotype traits.
  Sparse molecular background traits frequently fall below `min_snp_signals`
  and are dropped before matrix construction, so `n_traits` is a nominal
  matrix size rather than the number of traits that survive to clustering.