---
title: "Simulation study of multi-trait program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # --- generative structure -------------------------------------------------
  n_loci_per_trait: 200
  n_shared_loci: 80
  K_specific: 2
  # 36 loci per program. Shared loci per overlap: 0.1 -> 4, 0.2 -> 7,
  # 0.33 -> 12, 0.66 -> 24, 1 -> 36. Linking needs 3 shared loci and
  # module_rg() needs 10 to estimate a direction, so 0.1 and 0.2 probe the
  # low end: linkable, but with no direction call.
  module_size: 36
  n_drivers_per_program: 8
  n_background_traits: 300
  effect_size: 6
  locus_effect_sd: 0.5
  noise_sd: 1
  p_structural_zero: 0
  # --- background realism ---------------------------------------------------
  p_active_background: 0.02
  background_sparsity_sd: 1.2
  background_effect_scale: 0.5
  background_corr: 0
  snp_pleiotropy_sd: 0
  # --- the design -------------------------------------------------------------
  # Crossed: every SNP overlap is run with a concordant and an antagonistic
  # shared program.
  snp_overlap: [0.1, 0.2, 0.33, 0.66, 1.0]
  directions: ["concordant", "antagonistic"]
  # --- stability gate, as in investigation-multitrait-programs.Rmd -----------
  stability_n_rep: 2
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  # --- cross-trait comparison ----------------------------------------------
  n_perm: 1000
  fdr_threshold: 0.05
  # --- planted vs resolved figure -------------------------------------------
  # The version (SNP overlap, concordant) whose first replicate is drawn in the
  # planted-vs-resolved loading heatmap.
  example_snp_overlap: 0.33
  # --- replication ----------------------------------------------------------
  n_sim: 50
  master_seed: 1
  # Replicates within a version run in parallel via parallel::mclapply
  # (fork-based; falls back to serial on Windows or when cores = 1).
  cores: 5
  # quick: fast test mode. Caps n_sim at 2 and lowers the permutation counts.
  # All input versions are always run. Set FALSE for the full study.
  quick: true
vignette: >
  %\VignetteIndexEntry{Simulation study of multi-trait program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, out.width = "100%", fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)

quick <- isTRUE(params$quick)
n_sim <- if (quick) min(as.integer(params$n_sim), 2L) else as.integer(params$n_sim)
n_perm <- if (quick) 300L else as.integer(params$n_perm)
similarity_perm <- if (quick) 300L else 1000L

# Shared EBMF settings, from the same file the univariate and real-data studies
# read, so the per-trait discovery step here is the same discovery step used
# everywhere else.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

# One seed vector for every version: common random numbers.
set.seed(params$master_seed)
sim_seeds <- sample.int(.Machine$integer.max, n_sim)

generative_base <- list(
  n_loci_per_trait = as.integer(params$n_loci_per_trait),
  n_shared_loci = as.integer(params$n_shared_loci),
  K_specific = as.integer(params$K_specific),
  module_size = as.integer(params$module_size),
  n_drivers_per_program = as.integer(params$n_drivers_per_program),
  # The same background GWAS sit at a shared locus in both traits, so the
  # shared program's driver studies are identical in both.
  profile_overlap = 1,
  n_background_traits = as.integer(params$n_background_traits),
  effect_size = params$effect_size,
  locus_effect_sd = params$locus_effect_sd,
  noise_sd = params$noise_sd,
  p_structural_zero = params$p_structural_zero,
  p_active_background = params$p_active_background,
  background_sparsity_sd = params$background_sparsity_sd,
  background_effect_scale = params$background_effect_scale,
  background_corr = params$background_corr,
  snp_pleiotropy_sd = params$snp_pleiotropy_sd
)

palette_direction <- c(concordant = "#2a78d6", antagonistic = "#eb6834")
```

## How to read this document

**The question.** gpmapr finds *programs* -- groups of loci that behave alike
across hundreds of background GWAS -- separately for each target trait. The
real-data study (`investigation-multitrait-programs.Rmd`) then links two
traits' programs by the **loci they share**, with
`compare_program_pairs_loadings()`, and takes the direction of each link from
`module_rg()`. This document checks that pipeline on simulated data where the
answer is known. Two traits share **one** program, and we ask three things:

1. Is the shared program **found** in each trait's own fit, and if it is thrown
   away, which validation gate threw it away?
2. Is it **linked** across the two traits, as the SNP overlap between the two
   copies falls from 1 to 0.1?
3. Is the link given the right **direction**: concordant (raises both traits)
   or antagonistic (raises one, lowers the other)?

Programs are linked on loci only. Background-study profiles and annotations
play no part in forming a link; on real data they are used afterwards, to
describe programs that are already linked.

**What happens in one simulated replicate.**

1. **Simulate** two target traits with `simulate_trait_pair()`. Each is a matrix
   of background studies x loci (z-scores). One program is planted in both
   traits, and `K_specific` programs in each trait alone.
2. **Fit each trait on its own** with EBMF (`run_univariate_clustering()`),
   exactly as the real-data analysis does. EBMF knows nothing about the other
   trait.
3. **Validate** every fitted program (`summarise_ebmf_programs()`), with the
   same gates used on real data, and record each program's `status`.
4. **Link** the programs across traits (`compare_program_pairs_loadings()`),
   which also runs `module_rg()` on every link. This is done twice: on every
   fitted program, and on the valid programs only, so the effect of the gates
   can be seen.
5. **Score** discovery, links and direction against what was planted.

The whole replicate is repeated `n_sim` times per setting, and each table
reports the mean with its Monte Carlo standard error in parentheses.

**Glossary.**

- **Program**: one EBMF factor: a set of loci (its SNP loadings) together with
  the background studies that load on it.
- **SNP overlap**: the fraction of the shared program's loci that are the same
  locus (coloc group) in both traits. At 1 the two copies act at exactly the
  same loci; at 0.1 they share 4 of their 36.
- **Driver study**: a background study planted as part of a program. The shared
  program has the same driver studies in both traits.
- **Common allele frame**: effects are generated as they are reported, for one
  fixed allele per locus. A driver study has the same sign at a shared locus in
  both traits, because it is the same GWAS. An antagonistic program has
  opposite-sign **target** effects at its loci.
- **Orientation**: before fitting, every locus is flipped to its own target
  trait's risk-increasing allele, as on real data. This is where antagonism
  reaches the fit: in the antagonistic trait the driver studies end up negative.
- **status**: the validation outcome of a fitted program. `valid`, or the
  gates it failed (`size`, `similarity`, `stability`, `redundancy`),
  joined by `; `.
- **Primary / candidate link**: a *primary* link is a mutual best match between
  two significant programs, the strict one-to-one claim. A *candidate* link is
  any other pair whose own FDR clears the threshold.
- **Recall**: the fraction of planted shared programs linked (here 0 or 1 per
  replicate).
- **Precision**: the fraction of reported links that are real.
- **Sign accuracy**: among real links, the fraction called in the planted
  direction.
- **SNP precision / recall**: for a resolved program, the share of its
  high-confidence SNPs that were planted, and the share of the planted SNPs it
  recovers.
- **MCSE**: Monte Carlo standard error. It measures how precisely `n_sim`
  replicates pin down the reported mean, not biological variability.

```{r settings-table}
knitr::kable(
  data.frame(
    parameter = c(
      names(generative_base), "snp_overlap", "directions", "stability_n_rep",
      "frac_traits", "stability_top_n", "stability_threshold", "n_sim",
      "n_perm", "fdr_threshold", "quick", "ebmf_settings_signature"
    ),
    value = c(
      vapply(generative_base, function(x) paste(x, collapse = ", "), character(1)),
      paste(params$snp_overlap, collapse = ", "),
      paste(params$directions, collapse = ", "),
      params$stability_n_rep, params$frac_traits, params$stability_top_n,
      params$stability_threshold, n_sim, n_perm, params$fdr_threshold, quick,
      settings_signature
    ),
    stringsAsFactors = FALSE
  ),
  row.names = FALSE,
  caption = paste0(
    "Active settings. ebmf_settings_signature identifies the shared analysis ",
    "settings this run resolved from _ebmf_settings.yml; two studies carrying ",
    "the same signature were run on the same discovery settings."
  )
)
```

## Simulation design (ADEMP)

Following Morris, White and Crowther (2019, *Statistics in Medicine* 38:2074-2102),
as the univariate studies do.

**Aims.** (i) Measure how often a program shared by two traits survives each
trait's own fit and validation. (ii) Measure how often it is linked across the
traits as a function of SNP overlap. (iii) Measure whether the direction of the
link is called correctly, by the concordance sign and by `module_rg()`. (iv)
Measure false links when nothing is shared.

**Data-generating mechanism.** `simulate_trait_pair()` with `K_shared = 1`. Each
trait carries `n_loci_per_trait` loci, of which `n_shared_loci` sit in a pool
both traits carry. The shared program has `module_size` loci in each trait; a
fraction `snp_overlap` of them are drawn from the shared pool (the rest are
private to each trait), and its `n_drivers_per_program` driver studies are the
same in both traits. Each program locus carries an effect `a_s`, drawn once and
reused in both traits at a shared locus. A driver study's effect at a locus is
proportional to `a_s`; the target's own effect is `+a_s` in trait A and
`+a_s` (concordant) or `-a_s` (antagonistic) in trait B. Each trait also
carries `K_specific` programs of its own, which must not be linked, and 300
unstructured background studies. The versions cross `snp_overlap` in
{`r paste(params$snp_overlap, collapse = ", ")`} with the two directions, plus
a null with nothing shared.

**Estimand / target.** For each replicate: whether the shared program is found
valid in each trait, whether it is linked, and the planted direction of the
link. This is a selection-plus-sign target, not a parameter.

**Methods.** The real-data pipeline, unchanged: `run_univariate_clustering()`
and `summarise_ebmf_programs()` per trait at the shared settings (with the
stability gate as in the real-data study), then
`compare_program_pairs_loadings()` across traits, at its defaults: each pair is
scored on the shared loci either program claims, and a pair needs at least 3
shared high-confidence loci to be linked. Linking is run twice, on every fitted
program and on the valid programs only; stability is computed for every
program (`stability_all_programs = TRUE`), so filtered programs carry the same
metrics as valid ones. Direction is read two ways: the sign
of `concordance_z`, and `rg_direction` from `module_rg()`.

**Performance measures.** Discovered programs are mapped onto planted programs
by SNP membership (`map_programs_to_planted()`, Jaccard at least 0.1).

- *Discovery*: per trait, whether a valid program maps to the shared program;
  if not, the `status` of the best-matching program, or "not found" when no
  program maps to it.
- *Match quality*: for the program representing each planted program, the
  Jaccard, SNP precision and SNP recall of its high-confidence SNPs against the
  planted SNPs, and the correlation of its loadings with the planted per-locus
  effects.
- *Linking*: `evaluate_multitrait_simulation()` reports `recall` (planted
  shared program linked) and `precision` (links that are real), at the primary
  tier and at primary + candidate.
- *Direction*: sign accuracy on true links, from the concordance sign and from
  `module_rg()`, split by planted direction. `module_rg()` needs at least 10
  usable shared loci; when it has fewer the direction is *undetermined*, which
  is reported rather than dropped.
- *Null*: under `K_shared = 0` every link is a false discovery.

**Monte Carlo uncertainty.** Every estimate is a mean over `n_sim` replicates
with a Monte Carlo standard error; the replicate is the MC unit. Common random
numbers are used -- one seed vector shared across all versions -- so
version-to-version differences are estimated more precisely than with
independent seeds.

```{r versions}
versions <- list()
for (dir in as.character(params$directions)) {
  for (ov in as.numeric(params$snp_overlap)) {
    versions[[length(versions) + 1L]] <- list(
      label = sprintf("snp_overlap = %.2f, %s", ov, dir),
      family = "shared",
      snp_overlap = ov,
      direction = dir,
      spec = list(
        K_shared = 1L, locus_overlap = ov,
        directions = if (dir == "antagonistic") -1L else 1L
      )
    )
  }
}
# The null: two traits with programs of their own and no correspondence. Any
# link is a false discovery.
versions[[length(versions) + 1L]] <- list(
  label = "null: nothing shared",
  family = "null",
  snp_overlap = NA_real_,
  direction = NA_character_,
  spec = list(K_shared = 0L, locus_overlap = 0)
)
version_order <- vapply(versions, function(v) v$label, character(1))
```

```{r run-helpers}
# Fit one trait and validate every program. Returns each program's status, its
# full set of validation metrics (stability is computed for every program, not
# only the coherent ones, so filtered programs can be compared with valid ones),
# and the extraction of every fitted program, valid or not.
discover_one_trait <- function(trait_object, trait_id) {
  res <- run_univariate_clustering(
    trait_object,
    target_trait_id = as.integer(trait_id),
    trait_subset = ebmf_settings$trait_subset,
    min_snp_signals = as.integer(ebmf_settings$min_snp_signals),
    compress_method = ebmf_settings$compress_method,
    compress_scale = as.numeric(ebmf_settings$compress_scale),
    ebmf_lfsr_threshold = as.numeric(ebmf_settings$ebmf_lfsr_threshold),
    ebmf_magnitude_threshold = as.numeric(ebmf_settings$ebmf_magnitude_threshold),
    ebmf_greedy_Kmax = as.integer(ebmf_settings$ebmf_greedy_Kmax),
    ebmf_prior = ebmf_settings$ebmf_prior,
    ebmf_backfit = isTRUE(ebmf_settings$ebmf_backfit),
    min_module_size = as.integer(ebmf_settings$min_module_size)
  )
  smry <- summarise_ebmf_programs(
    res, s_matrix = res$s_matrix,
    edge_threshold = as.numeric(ebmf_settings$similarity_threshold),
    min_module_size = as.integer(ebmf_settings$min_module_size),
    similarity_n_perm = similarity_perm,
    similarity_q_threshold = as.numeric(ebmf_settings$similarity_q),
    strength_q_threshold = as.numeric(ebmf_settings$strength_q),
    n_null = 2,
    n_rep = as.integer(params$stability_n_rep),
    frac_traits = as.numeric(params$frac_traits),
    top_n = as.integer(params$stability_top_n),
    stability_threshold = as.numeric(params$stability_threshold),
    stability_all_programs = TRUE,
    verbose = FALSE
  )
  status <- data.frame(
    program_id = paste0(trait_id, ":", smry$programs$program),
    status = as.character(smry$programs$status),
    stringsAsFactors = FALSE
  )
  metric_cols <- intersect(
    c(
      "program", "n_snps", "n_snps_filtered", "similarity_q",
      "similarity_lift", "mean_internal_similarity", "connectedness",
      "replication", "posterior_evidence", "factor_pve", "strength_q",
      "parent_program", "max_abs_factor_corr", "max_pair_redundancy",
      "size_pass", "similarity_pass", "strength_pass", "stability_pass",
      "status"
    ),
    names(smry$programs)
  )
  metrics <- smry$programs[, metric_cols, drop = FALSE]
  metrics$status <- as.character(metrics$status)
  metrics$program_id <- paste0(trait_id, ":", metrics$program)
  metrics$trait_id <- as.character(trait_id)
  return(list(
    all = extract_program_loadings(res, smry, valid_only = FALSE),
    status = status,
    metrics = metrics
  ))
}

# What each fitted program is, against the planted truth: part of the shared
# program, part of a trait-specific program, or no planted program at all.
truth_class <- function(planted_label) {
  return(dplyr::case_when(
    is.na(planted_label) ~ "unmatched",
    startsWith(planted_label, "shared") ~ "shared",
    TRUE ~ "trait-specific"
  ))
}

# Link two traits' programs and score the links against the planted truth.
# `programs` labels the program set ("all" fitted programs or "valid" only).
# Returns the per-tier scores, every reported link with the gate status and
# planted match of both its programs, and the module_rg readout of the true
# link.
link_and_score <- function(extraction, mapping, status, gt, tids, sp, programs) {
  n_programs <- vapply(extraction, function(x) {
    return(length(unique(x$loadings$program_id)))
  }, integer(1))
  # Linking needs a program in both traits; otherwise score an empty table so
  # the replicate counts as zero recall rather than going missing.
  cmp <- if (all(n_programs > 0)) {
    compare_program_pairs_loadings(
      extraction, n_perm = n_perm, fdr_threshold = params$fdr_threshold,
      seed = 1,
      coloc_groups = stats::setNames(
        lapply(tids, function(tid) sp$traits[[tid]]$coloc_groups), tids
      )
    )
  } else {
    NULL
  }
  pairs <- if (is.null(cmp)) NULL else cmp$pairs
  pairs_rg <- pairs
  if (!is.null(pairs) && nrow(pairs) > 0) {
    rg <- cmp$module_rg[, c(
      "program_id_a", "program_id_b", "rg", "rg_ci_lower", "rg_ci_upper",
      "n_loci_rg", "rg_direction"
    )]
    pairs_rg <- dplyr::left_join(
      pairs, rg, by = c("program_id_a", "program_id_b")
    )
    pairs_rg$direction <- pairs_rg$rg_direction
  }

  tier_sets <- list(primary = "primary", any = c("primary", "candidate"))
  links <- dplyr::bind_rows(lapply(names(tier_sets), function(ts) {
    by_sign <- evaluate_multitrait_simulation(
      pairs, mapping, gt, tiers = tier_sets[[ts]]
    )
    by_rg <- evaluate_multitrait_simulation(
      pairs_rg, mapping, gt, tiers = tier_sets[[ts]]
    )
    out <- by_sign
    out$tiers <- ts
    out$rg_sign_accuracy_concordant <- by_rg$sign_accuracy_concordant
    out$rg_sign_accuracy_antagonistic <- by_rg$sign_accuracy_antagonistic
    return(out)
  }))
  links$programs <- programs
  links$n_programs_a <- n_programs[[1]]
  links$n_programs_b <- n_programs[[2]]

  label_of <- stats::setNames(mapping$planted_label, mapping$program_id)
  status_of <- stats::setNames(status$status, status$program_id)
  link_details <- data.frame()
  true_link <- data.frame()
  if (!is.null(pairs_rg) && nrow(pairs_rg) > 0) {
    linked <- pairs_rg[!is.na(pairs_rg$link_tier), , drop = FALSE]
    if (nrow(linked) > 0) {
      label_a <- unname(label_of[as.character(linked$program_id_a)])
      label_b <- unname(label_of[as.character(linked$program_id_b)])
      link_details <- data.frame(
        programs = programs,
        program_id_a = as.character(linked$program_id_a),
        program_id_b = as.character(linked$program_id_b),
        link_tier = linked$link_tier,
        concordance_z = linked$concordance_z,
        n_loci_shared = linked$n_loci_shared,
        status_a = unname(status_of[as.character(linked$program_id_a)]),
        status_b = unname(status_of[as.character(linked$program_id_b)]),
        truth_a = truth_class(label_a),
        truth_b = truth_class(label_b),
        true_link = !is.na(label_a) & !is.na(label_b) & label_a == label_b &
          startsWith(label_a, "shared"),
        stringsAsFactors = FALSE
      )
    }
    # The true link itself (both programs map to the shared program), for the
    # module_rg readout: primary tier first, then the strongest candidate.
    if (nrow(gt$correspondence) > 0) {
      shared_label <- gt$correspondence$program_label[1]
      is_true <- !is.na(pairs_rg$link_tier) &
        label_of[as.character(pairs_rg$program_id_a)] %in% shared_label &
        label_of[as.character(pairs_rg$program_id_b)] %in% shared_label
      tl <- pairs_rg[is_true, , drop = FALSE]
      if (nrow(tl) > 0) {
        tl <- tl[order(tl$link_tier != "primary", -abs(tl$concordance_z)), ][1, ]
        true_link <- data.frame(
          programs = programs,
          link_tier = tl$link_tier, concordance_z = tl$concordance_z,
          rg = tl$rg,
          ci_excludes_zero = is.finite(tl$rg_ci_lower) &
            is.finite(tl$rg_ci_upper) &
            (tl$rg_ci_lower > 0 | tl$rg_ci_upper < 0),
          n_loci_rg = tl$n_loci_rg,
          rg_undetermined = !is.finite(tl$rg),
          stringsAsFactors = FALSE
        )
      }
    }
  }
  return(list(links = links, link_details = link_details, true_link = true_link))
}

# Keep only the valid programs of an extraction.
valid_extraction <- function(extraction, valid_ids) {
  keep <- function(tab) {
    return(tab[tab$program_id %in% valid_ids, , drop = FALSE])
  }
  return(list(
    loadings = keep(extraction$loadings),
    profiles = keep(extraction$profiles),
    programs = keep(extraction$programs)
  ))
}

# For each planted program in one trait: found valid, found but filtered (with
# the status of the best-matching program), or not found. For the program that
# represents it -- the best-matching valid program, else the best-matching
# program of any status -- also score how closely its SNPs match the planted
# ones: Jaccard, SNP precision (share of its high-confidence SNPs that were
# planted) and recall (share of planted SNPs it recovers), and the correlation
# of its loadings over every locus in the trait with the planted effect a_s
# (0 at loci outside the program).
planted_outcomes <- function(mapping, status, gt, trait_id, loadings) {
  labels <- gt$programs$program_label[gt$programs$trait_id == as.integer(trait_id)]
  m <- dplyr::left_join(mapping, status, by = "program_id")
  trait_loadings <- loadings[loadings$trait_id == as.character(trait_id), , drop = FALSE]
  rows <- lapply(labels, function(lab) {
    # Planted labels are not trait-specific ("shared1" is planted in both
    # traits), so a hit must also be one of this trait's programs.
    hits <- m[
      !is.na(m$planted_label) & m$planted_label == lab &
        m$trait_id == as.character(trait_id), ,
      drop = FALSE
    ]
    valid_hits <- hits[hits$status %in% "valid", , drop = FALSE]
    outcome <- if (nrow(hits) == 0) {
      "not found"
    } else if (nrow(valid_hits) > 0) {
      "valid"
    } else {
      best <- hits[which.max(hits$score), , drop = FALSE]
      paste0("filtered: ", dplyr::coalesce(best$status, "unsummarised"))
    }
    quality <- data.frame(
      jaccard = NA_real_, snp_precision = NA_real_, snp_recall = NA_real_,
      loading_cor = NA_real_
    )
    if (nrow(hits) > 0) {
      pool <- if (nrow(valid_hits) > 0) valid_hits else hits
      rep_id <- pool$program_id[which.max(pool$score)]
      prog <- trait_loadings[trait_loadings$program_id == rep_id, , drop = FALSE]
      planted <- gt$program_snps[[paste0(trait_id, ":", lab)]]
      resolved <- unique(prog$snp_id[prog$high_confidence %in% TRUE])
      shared <- length(intersect(resolved, planted))
      effect <- gt$program_effects[[paste0(trait_id, ":", lab)]]
      truth <- ifelse(prog$snp_id %in% names(effect), effect[prog$snp_id], 0)
      quality <- data.frame(
        jaccard = shared / length(union(resolved, planted)),
        snp_precision = if (length(resolved) > 0) shared / length(resolved) else NA_real_,
        snp_recall = shared / length(planted),
        loading_cor = if (nrow(prog) > 2 && stats::sd(prog$loading) > 0) {
          stats::cor(prog$loading, truth)
        } else {
          NA_real_
        }
      )
    }
    return(data.frame(
      trait_id = as.character(trait_id), planted_label = lab,
      is_shared = startsWith(lab, "shared"), outcome = outcome,
      n_programs_mapped = nrow(hits),
      quality,
      stringsAsFactors = FALSE
    ))
  })
  return(dplyr::bind_rows(rows))
}

# Planted vs resolved loadings for one replicate, for the heatmap: every fitted
# program's loading at every locus, with the planted program each locus belongs
# to (planted labels are made common to both traits so the traits line up) and
# the planted program each fitted program maps to, if any.
planted_vs_resolved <- function(loadings, status, gt, tids, mapping) {
  common_label <- function(lab) {
    out <- sub("^shared[0-9]+$", "shared", lab)
    return(sub("^specific([0-9]+)_t[0-9]+$", "specific \\1", out))
  }
  rows <- lapply(tids, function(tid) {
    tab <- loadings[loadings$trait_id == as.character(tid), , drop = FALSE]
    keys <- names(gt$program_snps)[startsWith(names(gt$program_snps), paste0(tid, ":"))]
    planted_of <- unlist(lapply(keys, function(k) {
      return(stats::setNames(
        rep(common_label(sub("^[^:]*:", "", k)), length(gt$program_snps[[k]])),
        gt$program_snps[[k]]
      ))
    }))
    effect_of <- unlist(unname(gt$program_effects[keys]))
    tab$planted <- dplyr::coalesce(unname(planted_of[tab$snp_id]), "unplanted")
    tab$planted_effect <- dplyr::coalesce(unname(effect_of[tab$snp_id]), 0)
    tab <- dplyr::left_join(tab, status, by = "program_id")
    tab <- dplyr::left_join(
      tab,
      mapping[, c("program_id", "planted_label")] |>
        dplyr::rename(maps_to = planted_label),
      by = "program_id"
    )
    return(tab[, c(
      "trait_id", "program_id", "program", "status", "maps_to", "snp_id",
      "loading", "high_confidence", "planted", "planted_effect"
    )])
  })
  return(dplyr::bind_rows(rows))
}

run_replicate <- function(spec, seed) {
  sp <- do.call(
    simulate_trait_pair,
    c(utils::modifyList(generative_base, spec), list(seed = seed))
  )
  gt <- sp$ground_truth
  tids <- names(sp$traits)
  fits <- lapply(tids, function(tid) discover_one_trait(sp$traits[[tid]], tid))
  names(fits) <- tids

  status <- dplyr::bind_rows(lapply(fits, function(x) x$status))
  valid_ids <- status$program_id[status$status == "valid"]
  loadings_all <- dplyr::bind_rows(lapply(fits, function(x) x$all$loadings))
  mapping_all <- map_programs_to_planted(loadings_all, gt, by = "snps")
  mapping_valid <- mapping_all[mapping_all$program_id %in% valid_ids, , drop = FALSE]

  discovery <- dplyr::bind_rows(lapply(tids, function(tid) {
    return(planted_outcomes(mapping_all, status, gt, tid, loadings_all))
  }))
  discovery$trait <- ifelse(discovery$trait_id == tids[1], "trait A", "trait B")

  # Every fitted program, with its validation metrics and what it is against
  # the planted truth.
  program_metrics <- dplyr::bind_rows(lapply(fits, function(x) x$metrics)) |>
    dplyr::left_join(
      mapping_all[, c("program_id", "planted_label", "score")] |>
        dplyr::rename(jaccard = score),
      by = "program_id"
    ) |>
    dplyr::mutate(
      truth = truth_class(planted_label),
      trait = ifelse(trait_id == tids[1], "trait A", "trait B")
    )

  # Link twice: on every fitted program (the primary analysis), and on the
  # valid programs only, as the gates would have it.
  extraction_all <- lapply(fits, function(x) x$all)
  extraction_valid <- lapply(fits, function(x) valid_extraction(x$all, valid_ids))
  scored <- list(
    link_and_score(extraction_all, mapping_all, status, gt, tids, sp, "all"),
    link_and_score(extraction_valid, mapping_valid, status, gt, tids, sp, "valid")
  )
  links <- dplyr::bind_rows(lapply(scored, function(x) x$links))
  link_details <- dplyr::bind_rows(lapply(scored, function(x) x$link_details))
  true_link <- dplyr::bind_rows(lapply(scored, function(x) x$true_link))

  # Keep the full loadings of the first replicate only, for the heatmap.
  example <- if (identical(seed, sim_seeds[[1]])) {
    planted_vs_resolved(loadings_all, status, gt, tids, mapping_all)
  } else {
    data.frame()
  }
  return(list(
    links = links, discovery = discovery, true_link = true_link,
    program_metrics = program_metrics, link_details = link_details,
    example = example
  ))
}

# Run every version. The runner returns a named list of tables; each is stacked
# across replicates and versions and tagged with both. Errors are returned
# rather than recorded with `<<-`, so they survive a forked mclapply worker (a
# child process's `<<-` never reaches the parent).
run_versions <- function(versions, runner) {
  n_jobs <- max(1L, min(as.integer(params$cores), n_sim))
  one <- function(spec, r) {
    err <- NULL
    out <- tryCatch(
      runner(spec, sim_seeds[[r]]),
      error = function(e) {
        err <<- conditionMessage(e)
        return(NULL)
      }
    )
    return(list(out = out, error = err, rep = r))
  }
  tables <- list()
  errors <- list()
  for (v in seq_along(versions)) {
    spec <- versions[[v]]$spec
    rep_out <- if (n_jobs > 1 && .Platform$OS.type != "windows") {
      parallel::mclapply(
        seq_len(n_sim), function(r) one(spec, r),
        mc.cores = n_jobs, mc.set.seed = FALSE
      )
    } else {
      lapply(seq_len(n_sim), function(r) one(spec, r))
    }
    for (x in rep_out) {
      if (!is.null(x$error)) {
        errors[[length(errors) + 1L]] <- data.frame(
          version = versions[[v]]$label, rep = x$rep, message = x$error,
          stringsAsFactors = FALSE
        )
      }
      for (nm in names(x$out)) {
        tab <- x$out[[nm]]
        if (is.null(tab) || nrow(tab) == 0) {
          next
        }
        tab$rep <- x$rep
        tab$version <- versions[[v]]$label
        tab$family <- versions[[v]]$family
        tab$snp_overlap <- versions[[v]]$snp_overlap
        tab$direction_planted <- versions[[v]]$direction
        if (is.null(tables[[nm]])) {
          tables[[nm]] <- list()
        }
        tables[[nm]][[length(tables[[nm]]) + 1L]] <- tab
      }
    }
  }
  out <- lapply(tables, dplyr::bind_rows)
  out$errors <- dplyr::bind_rows(errors)
  return(out)
}

mc_summary <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) {
    return(list(est = NA_real_, se = NA_real_, n = 0L))
  }
  return(list(
    est = mean(x),
    se = if (length(x) > 1) stats::sd(x) / sqrt(length(x)) else NA_real_,
    n = length(x)
  ))
}

fmt_mc <- function(x, digits = 2) {
  s <- mc_summary(x)
  if (!is.finite(s$est)) {
    return("--")
  }
  if (!is.finite(s$se)) {
    return(formatC(s$est, format = "f", digits = digits))
  }
  return(sprintf(
    "%s (%s)", formatC(s$est, format = "f", digits = digits),
    formatC(s$se, format = "f", digits = digits)
  ))
}

print_errors <- function(err) {
  if (is.null(err) || nrow(err) == 0) {
    return(invisible(NULL))
  }
  print(knitr::kable(
    err |>
      dplyr::group_by(version, message) |>
      dplyr::summarise(n_failed = dplyr::n(), .groups = "drop"),
    row.names = FALSE,
    caption = paste0(
      "Replicates that failed. Any row here means the version is reported on ",
      "fewer replicates than n_sim, or not at all."
    )
  ))
  return(invisible(NULL))
}
```

```{r run}
sim <- run_versions(versions, run_replicate)
links <- if (is.null(sim$links)) data.frame() else sim$links
discovery <- if (is.null(sim$discovery)) data.frame() else sim$discovery
true_links <- if (is.null(sim$true_link)) data.frame() else sim$true_link
examples <- if (is.null(sim$example)) data.frame() else sim$example
program_metrics <- if (is.null(sim$program_metrics)) data.frame() else sim$program_metrics
link_details <- if (is.null(sim$link_details)) data.frame() else sim$link_details
```

```{r errors, results='asis'}
print_errors(sim$errors)
```

## 1. Every fitted program, before filtering

**What this section shows.** EBMF returns many more programs than were
planted: the planted ones plus small factors fitted to noise. Here every
fitted program is shown with its validation metrics, before any gate is
applied, next to what it really is:

- **shared**: its high-confidence SNPs match the shared planted program;
- **trait-specific**: they match one of the trait's own planted programs;
- **unmatched**: they match no planted program (Jaccard below 0.1), so it is
  noise.

A gate is doing its job when it keeps the shared and trait-specific programs
and removes the unmatched ones. The figure shows where each metric separates
the three, which is the evidence for where a threshold should sit.

- **n_snps_filtered** (size gate): SNPs passing the lFSR/magnitude membership
  cut.
- **similarity_q** (similarity gate): BH q of the test that the member SNPs'
  mean similarity beats random SNP sets of the same size, shown as −log10.
- **strength_q** (reported, not a gate): BH q of the test that the factor
  explains more variance than factors fitted to row-permuted data, shown as
  −log10.
- **replication** (stability gate): mean Jaccard overlap between the program's
  top SNPs and the best-matching factor when 20% of the background studies are
  held out and EBMF is refit. Computed here for every program.
- **posterior_evidence**: median −log10(lFSR) of the program's members.
- **factor_pve**: share of the matrix variance the factor explains.
- **jaccard**: overlap with the best-matching planted program (0 when none).

```{r program-overview, results='asis'}
threshold_lines <- data.frame(
  metric = c("n_snps_filtered", "-log10(similarity_q)", "replication"),
  value = c(
    as.numeric(ebmf_settings$min_module_size),
    -log10(as.numeric(ebmf_settings$similarity_q)),
    as.numeric(params$stability_threshold)
  )
)
truth_levels <- c("shared", "trait-specific", "unmatched")
if (nrow(program_metrics) == 0) {
  cat("No replicates completed.\n\n")
} else {
  program_metrics <- program_metrics |>
    dplyr::mutate(
      truth = factor(truth, levels = truth_levels),
      valid = status == "valid",
      jaccard = dplyr::coalesce(jaccard, 0)
    )
  n_fits <- dplyr::n_distinct(
    paste(program_metrics$version, program_metrics$rep, program_metrics$trait)
  )
  med <- function(x) {
    return(if (all(!is.finite(x))) "--" else formatC(stats::median(x, na.rm = TRUE), format = "g", digits = 3))
  }
  pct <- function(x) {
    return(sprintf("%.0f%%", 100 * mean(x %in% TRUE)))
  }
  print(knitr::kable(
    program_metrics |>
      dplyr::group_by(truth) |>
      dplyr::summarise(
        per_trait_fit = round(dplyr::n() / n_fits, 1),
        pass_size = pct(size_pass),
        pass_similarity = pct(similarity_pass),
        pass_strength = pct(strength_pass),
        pass_stability = pct(stability_pass),
        valid = pct(valid),
        n_snps_filtered = med(n_snps_filtered),
        similarity_q = med(similarity_q),
        strength_q = med(strength_q),
        replication = med(replication),
        posterior_evidence = med(posterior_evidence),
        factor_pve = med(factor_pve),
        jaccard = med(jaccard),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "Every fitted program, pooled over versions, replicates and traits (",
      n_fits, " trait fits). per_trait_fit = programs of this kind per trait ",
      "fit, over all fits including the null versions, which have no shared ",
      "program; pass_* = share passing each gate; valid = share passing all; ",
      "the metric columns are medians."
    )
  ))
}
```

```{r program-metrics-plot, fig.width = 10, fig.height = 7}
if (nrow(program_metrics) > 0 && requireNamespace("ggplot2", quietly = TRUE)) {
  metric_long <- program_metrics |>
    dplyr::transmute(
      truth, valid,
      n_snps_filtered = as.numeric(n_snps_filtered),
      `-log10(similarity_q)` = -log10(pmax(similarity_q, 1e-10)),
      `-log10(strength_q)` = -log10(pmax(strength_q, 1e-10)),
      replication, posterior_evidence,
      factor_pve = 100 * factor_pve,
      jaccard
    ) |>
    tidyr::pivot_longer(-c(truth, valid), names_to = "metric", values_to = "value") |>
    dplyr::filter(is.finite(value)) |>
    dplyr::mutate(
      metric = factor(metric, levels = c(
        "n_snps_filtered", "-log10(similarity_q)", "-log10(strength_q)",
        "replication", "posterior_evidence", "factor_pve", "jaccard"
      )),
      gate = ifelse(valid, "valid", "filtered")
    )
  print(
    ggplot2::ggplot(metric_long, ggplot2::aes(x = truth, y = value)) +
      ggplot2::geom_boxplot(outlier.shape = NA, colour = "grey55", width = 0.6) +
      ggplot2::geom_jitter(
        ggplot2::aes(colour = gate), width = 0.2, height = 0, alpha = 0.6, size = 1.2
      ) +
      ggplot2::geom_hline(
        data = dplyr::mutate(
          threshold_lines,
          metric = factor(metric, levels = levels(metric_long$metric))
        ),
        ggplot2::aes(yintercept = value), linetype = "dashed", colour = "#b2182b"
      ) +
      ggplot2::facet_wrap(~metric, scales = "free_y", ncol = 4) +
      ggplot2::scale_colour_manual(values = c(valid = "#2a78d6", filtered = "grey40")) +
      ggplot2::labs(
        title = "Validation metrics of every fitted program, by what it really is",
        subtitle = "Dashed red lines: current gate thresholds. factor_pve in %.",
        x = NULL, y = NULL, colour = NULL
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(
        legend.position = "top",
        axis.text.x = ggplot2::element_text(angle = 20, hjust = 1)
      )
  )
}
```

```{r gate-by-truth, results='asis'}
if (nrow(program_metrics) > 0) {
  print(knitr::kable(
    program_metrics |>
      dplyr::count(status, truth) |>
      tidyr::pivot_wider(names_from = truth, values_from = n, values_fill = 0L) |>
      dplyr::arrange(status != "valid", status),
    row.names = FALSE,
    caption = paste0(
      "Programs by validation outcome (rows) and what they really are (columns), ",
      "pooled. A good gate puts shared and trait-specific programs in the ",
      "'valid' row and unmatched programs everywhere else."
    )
  ))

  metric_view <- function(tab) {
    return(tab |>
      dplyr::transmute(
        version, rep, trait, program, planted_label,
        jaccard = round(jaccard, 2),
        n_snps_filtered,
        similarity_q = signif(similarity_q, 2),
        strength_q = signif(strength_q, 2),
        replication = round(replication, 2),
        posterior_evidence = round(posterior_evidence, 1),
        factor_pve_pct = round(100 * factor_pve, 2),
        status
      ))
  }
  real_filtered <- program_metrics |>
    dplyr::filter(!valid, truth != "unmatched") |>
    dplyr::arrange(dplyr::desc(jaccard))
  if (nrow(real_filtered) > 0) {
    print(knitr::kable(
      metric_view(utils::head(real_filtered, 20)),
      row.names = FALSE,
      caption = paste0(
        "Real programs the gates removed: filtered programs that match a planted ",
        "program, best match first (up to 20 of ", nrow(real_filtered), "). ",
        "Many are fragments of a planted program that EBMF split off."
      )
    ))
  } else {
    cat("No filtered program matches a planted program.\n\n")
  }
  noise_valid <- program_metrics |>
    dplyr::filter(valid, truth == "unmatched") |>
    dplyr::arrange(dplyr::desc(n_snps_filtered))
  if (nrow(noise_valid) > 0) {
    print(knitr::kable(
      metric_view(utils::head(noise_valid, 20)),
      row.names = FALSE,
      caption = paste0(
        "Noise the gates let through: valid programs that match no planted ",
        "program (up to 20 of ", nrow(noise_valid), ")."
      )
    ))
  } else {
    cat("No valid program is unmatched: the gates let no noise program through.\n\n")
  }
}
```

**How to read it.**

- **Size and similarity** should remove most unmatched programs on their own.
  Check that no shared or trait-specific program sits below their lines in the
  figure.
- **Stability.** If shared and trait-specific programs cluster near 1 and
  unmatched ones spread lower, the stability threshold can be read off the gap.
  Where the two groups overlap, no threshold on that metric separates them.
- **The two lists** are the direct evidence. Real programs removed by a gate
  argue for loosening it; noise passing every gate argues for tightening one.
- **Quick mode floors similarity.** With `quick = TRUE` the similarity null
  uses 300 random SNP sets, so the smallest possible p is about 0.003. The full
  run uses 1000, which leaves more room.

## 2. Was the shared program found in each trait?

**What this table shows.** For the shared program, in each trait, the share of
replicates in which a **valid** program represents it, in which the best
program representing it was **filtered** out by validation (with the gates it
failed), and in which **no** fitted program represents it at all. The trait-
specific programs are shown last, pooled, for comparison.

```{r discovery-table, results='asis'}
if (nrow(discovery) == 0) {
  cat("No replicates completed.\n\n")
} else {
  shared_disc <- discovery[discovery$is_shared, , drop = FALSE]
  tbl <- shared_disc |>
    dplyr::count(version, trait, outcome) |>
    dplyr::group_by(version, trait) |>
    dplyr::mutate(share = sprintf("%.0f%%", 100 * n / sum(n))) |>
    dplyr::ungroup() |>
    dplyr::select(-n) |>
    tidyr::pivot_wider(names_from = outcome, values_from = share, values_fill = "0%") |>
    dplyr::mutate(version = factor(version, levels = version_order)) |>
    dplyr::arrange(version, trait)
  front <- intersect(c("version", "trait", "valid"), names(tbl))
  tbl <- tbl[, c(front, sort(setdiff(names(tbl), front)))]
  print(knitr::kable(
    tbl, row.names = FALSE,
    caption = paste0(
      "Discovery of the shared program, per trait. 'filtered: <status>' names ",
      "the validation gates the best-matching program failed; 'not found' ",
      "means no fitted program overlaps the planted loci (Jaccard >= 0.1)."
    )
  ))
  spec_tbl <- discovery[!discovery$is_shared, , drop = FALSE] |>
    dplyr::count(family, outcome) |>
    dplyr::group_by(family) |>
    dplyr::mutate(share = sprintf("%.0f%%", 100 * n / sum(n))) |>
    dplyr::ungroup() |>
    dplyr::select(-n) |>
    tidyr::pivot_wider(names_from = outcome, values_from = share, values_fill = "0%")
  print(knitr::kable(
    spec_tbl, row.names = FALSE,
    caption = "Discovery of the trait-specific programs, pooled over versions and traits."
  ))
}
```

**How to read it.** SNP overlap should not change discovery much: each trait's
fit sees only its own matrix, and the shared program is an ordinary program
there. If the `valid` column is well below 100%, the shared program is being
lost before linking, and the `filtered:` columns say which gate removed it.
A program that fails `stability` survives the full fit but not a refit on 80%
of the background studies; `similarity` means its member SNPs are no more
alike than random SNP sets of the same size. The redundancy gate is off by
default (`snp_redundancy_threshold = Inf`), so `redundancy` should not appear.

## 2b. How closely do the resolved programs match the planted ones?

**What this table shows.** Section 2 counts a planted program as found when a
fitted program overlaps it with Jaccard of at least 0.1, which is a low bar.
This table shows how close that match actually is. For each planted program,
the program that represents it is the best-matching valid program (or, if
none is valid, the best-matching program of any status). Its match is then
scored four ways:

- **jaccard**: the overlap between its high-confidence SNPs and the planted
  SNPs, divided by their union.
- **snp_precision**: the share of its high-confidence SNPs that were planted in
  the program. Below 1 means it has picked up SNPs from elsewhere.
- **snp_recall**: the share of the planted SNPs it recovers as high-confidence
  members. Below 1 means planted SNPs were missed or went to another program.
- **loading_cor**: the correlation of its SNP loadings, over every locus in the
  trait, with the planted per-locus effect `a_s` (0 outside the program). This
  scores the loadings themselves, not just the membership cut.

`n_programs` is the mean number of fitted programs mapped to the planted one.
Values above 1 mean EBMF split it.

```{r match-quality-table, results='asis'}
if (nrow(discovery) == 0) {
  cat("No replicates completed.\n\n")
} else {
  quality_summary <- function(tab) {
    return(tab |>
      dplyr::summarise(
        jaccard = fmt_mc(jaccard),
        snp_precision = fmt_mc(snp_precision),
        snp_recall = fmt_mc(snp_recall),
        loading_cor = fmt_mc(loading_cor),
        n_programs = fmt_mc(n_programs_mapped, 1),
        .groups = "drop"
      ))
  }
  shared_quality <- discovery[discovery$is_shared, , drop = FALSE] |>
    dplyr::group_by(version, trait) |>
    quality_summary() |>
    dplyr::mutate(version = factor(version, levels = version_order)) |>
    dplyr::arrange(version, trait)
  specific_quality <- discovery[!discovery$is_shared, , drop = FALSE] |>
    dplyr::mutate(
      version = dplyr::if_else(
        family == "null", "trait-specific, null version",
        "trait-specific, shared versions"
      ),
      trait = "both"
    ) |>
    dplyr::group_by(version, trait) |>
    quality_summary()
  print(knitr::kable(
    dplyr::bind_rows(
      dplyr::mutate(shared_quality, version = as.character(version)),
      specific_quality
    ),
    row.names = FALSE,
    caption = paste0(
      "Match between planted and resolved programs (mean over replicates, ",
      "Monte Carlo SE). The shared program is shown per version and trait; the ",
      "trait-specific programs are pooled in the last rows. Scores are over ",
      "replicates in which some fitted program mapped to the planted one."
    )
  ))
}
```

**How to read it.**

- **Precision and recall near 1** mean the resolved program is the planted
  program, locus for locus.
- **A high loading_cor** means the fitted loadings also track how strongly each
  locus acts in the program, not only which loci are in it.
- **SNP overlap should not matter here.** Each trait is fitted on its own, so
  the shared program's match should look the same at every overlap. A trend
  with overlap would point to a problem in the simulation, not the fit.
- **Recall well below precision** usually goes with `n_programs` above 1: the
  planted program was split, and the representative program holds only part of
  it.

```{r example-heatmap, fig.width = 9, fig.height = 8}
example_label <- sprintf("snp_overlap = %.2f, concordant", params$example_snp_overlap)
example <- if (nrow(examples) > 0) {
  examples[examples$version == example_label, , drop = FALSE]
} else {
  data.frame()
}
if (nrow(example) > 0 && requireNamespace("ggplot2", quietly = TRUE)) {
  # EBMF also returns many small factors that fail validation and match no
  # planted program; drawing them would bury the programs that matter.
  n_fitted <- dplyr::n_distinct(example$program_id)
  example <- example[example$status %in% "valid" | !is.na(example$maps_to), , drop = FALSE]
  n_omitted <- n_fitted - dplyr::n_distinct(example$program_id)
  planted_levels <- c("shared", sort(setdiff(unique(example$planted), c("shared", "unplanted"))), "unplanted")
  example <- example |>
    dplyr::mutate(
      trait = ifelse(trait_id == min(as.integer(trait_id)), "trait A", "trait B"),
      planted = factor(planted, levels = planted_levels),
      fitted = paste0(
        "P", program, ifelse(status == "valid", "", paste0("\n(", status, ")"))
      )
    )
  # Rank loci within each trait's planted block (strongest planted effect at
  # the top) so both traits' blocks line up row for row.
  snp_rank <- example |>
    dplyr::distinct(trait, snp_id, planted, planted_effect) |>
    dplyr::group_by(trait, planted) |>
    dplyr::arrange(dplyr::desc(planted_effect), snp_id, .by_group = TRUE) |>
    dplyr::mutate(y = -dplyr::row_number()) |>
    dplyr::ungroup() |>
    dplyr::select(trait, snp_id, y)
  example <- dplyr::left_join(example, snp_rank, by = c("trait", "snp_id"))
  limit <- max(abs(example$loading), na.rm = TRUE)
  print(
    ggplot2::ggplot(example, ggplot2::aes(x = fitted, y = y, fill = loading)) +
      ggplot2::geom_tile() +
      ggplot2::geom_point(
        data = example[example$high_confidence %in% TRUE, , drop = FALSE],
        size = 0.35, colour = "grey15"
      ) +
      ggplot2::facet_grid(planted ~ trait, scales = "free", space = "free") +
      ggplot2::scale_fill_gradient2(
        low = "#2166ac", mid = "white", high = "#b2182b",
        limits = c(-limit, limit), name = "SNP loading"
      ) +
      ggplot2::labs(
        title = "Planted vs resolved programs, one replicate",
        subtitle = paste0(
          example_label, ", first replicate. Dots: high-confidence members. ",
          n_omitted, " small unmatched, invalid programs not drawn."
        ),
        x = "fitted program", y = NULL
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(
        axis.text.y = ggplot2::element_blank(),
        panel.grid = ggplot2::element_blank(),
        strip.text.y = ggplot2::element_text(angle = 0),
        plot.subtitle = ggplot2::element_text(size = 8)
      )
  )
} else {
  cat("No example replicate for ", example_label, ".\n", sep = "")
}
```

**How to read it.** Each row block is one planted program, plus the loci that
belong to no program ("unplanted"). Columns are the fitted programs that are
valid or that map to a planted program. Small factors that fail validation and
match nothing are left out. A clean recovery is one strong column per
planted block, with dots, and pale everywhere else. Two columns sharing a block
show a split program. A column lit up across several blocks shows programs
merged into one. Dots outside a block's own column are high-confidence SNPs
assigned to the wrong program.

## 3. Was it linked?

**What this table shows.** Linking was run twice on every replicate:

- on **all** fitted programs, the primary analysis;
- on the **valid** programs only, as the gates would have it.

For each setting and program set, the table shows how many links were reported
per replicate, the fraction of replicates in which the shared program was
linked (`recall`), and the fraction of links that were real (`precision`). Both
are given at the strict primary tier and at primary + candidate (`any`).

```{r link-table, results='asis'}
shared_links <- if (nrow(links) > 0) {
  links[links$family == "shared", , drop = FALSE]
} else {
  data.frame()
}
if (nrow(shared_links) == 0) {
  cat("No replicates completed.\n\n")
} else {
  tbl <- shared_links |>
    dplyr::group_by(version, programs, tiers) |>
    dplyr::summarise(
      links = fmt_mc(n_links, 1),
      recall = fmt_mc(recall),
      precision = fmt_mc(precision),
      .groups = "drop"
    ) |>
    tidyr::pivot_wider(
      names_from = tiers, values_from = c(links, recall, precision)
    ) |>
    dplyr::mutate(
      version = factor(version, levels = version_order),
      programs = factor(programs, levels = c("all", "valid"))
    ) |>
    dplyr::arrange(version, programs)
  print(knitr::kable(
    tbl, row.names = FALSE,
    caption = paste0(
      "Link recovery on all fitted programs and on valid programs only (mean ",
      "over ", n_sim, " replicates, Monte Carlo SE). ",
      "recall is whether the shared program was linked; precision is over ",
      "reported links."
    )
  ))
}
```

**How to read it.** Links are scored on loci only, so a program shared at
entirely different variants (overlap 0) cannot be detected by design and is not
run. At `snp_overlap = 0.10` the two copies share 4 loci, just above the 3 a
link needs, so this is where recall should start to fall. For the valid-only
run, compare recall with section 2: it cannot exceed the rate at which the
shared program is valid in *both* traits. A recall below that rate is a failure
of the link, and a recall at that rate is a failure of discovery. Precision
below 1 means links other than the true one were reported. The next table says
whether they came from programs the gates would have removed.

```{r link-details, results='asis'}
if (nrow(link_details) > 0) {
  detail <- link_details[link_details$programs == "all", , drop = FALSE] |>
    dplyr::mutate(
      n_valid = (status_a %in% "valid") + (status_b %in% "valid"),
      pair_status = dplyr::case_when(
        n_valid == 2 ~ "both valid",
        n_valid == 1 ~ "one filtered",
        TRUE ~ "both filtered"
      ),
      pair_status = factor(pair_status, levels = c("both valid", "one filtered", "both filtered")),
      link = ifelse(true_link, "true link", "false link"),
      arm = ifelse(family == "null", "null version", "shared versions")
    )
  print(knitr::kable(
    detail |>
      dplyr::count(arm, link, pair_status) |>
      tidyr::pivot_wider(
        names_from = pair_status, values_from = n, values_fill = 0L,
        names_sort = TRUE
      ) |>
      dplyr::arrange(arm, link),
    row.names = FALSE,
    caption = paste0(
      "Links reported when every fitted program is linked, by whether they are ",
      "real and by the gate status of the two programs, pooled over replicates. ",
      "False links between filtered programs are what the gates prevent; true ",
      "links involving a filtered program are what the gates cost."
    )
  ))
  false_detail <- detail[!detail$true_link, , drop = FALSE]
  if (nrow(false_detail) > 0) {
    print(knitr::kable(
      false_detail |>
        dplyr::count(arm, pair_status, truth_a, truth_b, link_tier) |>
        dplyr::arrange(arm, pair_status, dplyr::desc(n)),
      row.names = FALSE,
      caption = "False links in the all-programs run, by what the two linked programs really are."
    ))
  }
} else {
  cat("No links were reported.\n\n")
}
```

```{r link-plot, fig.width=9, fig.height=6}
if (requireNamespace("ggplot2", quietly = TRUE) && nrow(shared_links) > 0) {
  plot_df <- shared_links |>
    dplyr::mutate(
      tiers = ifelse(tiers == "primary", "primary links", "primary + candidate links")
    ) |>
    dplyr::group_by(snp_overlap, direction_planted, programs, tiers) |>
    dplyr::summarise(recall = mean(recall, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(programs = paste(programs, "programs"))
  print(
    ggplot2::ggplot(
      plot_df,
      ggplot2::aes(
        snp_overlap, recall, colour = direction_planted, group = direction_planted
      )
    ) +
      ggplot2::geom_line(linewidth = 0.7) +
      ggplot2::geom_point(size = 2.5) +
      ggplot2::facet_grid(programs ~ tiers) +
      ggplot2::scale_colour_manual(values = palette_direction) +
      ggplot2::scale_y_continuous(limits = c(0, 1)) +
      ggplot2::labs(
        x = "SNP overlap (fraction of the shared program's loci common to both traits)",
        y = "recall of the shared program", colour = "planted direction"
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(legend.position = "top")
  )
}
```

## 4. Was the direction right?

**What this table shows.** For the all-programs run: among real links (primary +
candidate), the fraction called in the planted direction, from the sign of `concordance_z` and from
`module_rg()`. For the true link itself: the mean `rg`, how often its bootstrap
CI excludes 0, and how often `module_rg()` could not estimate it (fewer than 10
usable shared loci).

```{r sign-table, results='asis'}
if (nrow(shared_links) > 0) {
  sign_tbl <- shared_links[
    shared_links$tiers == "any" & shared_links$programs == "all", ,
    drop = FALSE
  ] |>
    dplyr::group_by(version, direction_planted) |>
    dplyr::summarise(
      concordance_sign = fmt_mc(dplyr::if_else(
        direction_planted == "concordant",
        sign_accuracy_concordant, sign_accuracy_antagonistic
      )),
      module_rg = fmt_mc(dplyr::if_else(
        direction_planted == "concordant",
        rg_sign_accuracy_concordant, rg_sign_accuracy_antagonistic
      )),
      .groups = "drop"
    )
  rg_tbl <- if (nrow(true_links) > 0) {
    true_links[true_links$programs == "all", , drop = FALSE] |>
      dplyr::group_by(version) |>
      dplyr::summarise(
        true_links = dplyr::n(),
        mean_rg = fmt_mc(rg),
        ci_excludes_zero = fmt_mc(as.numeric(ci_excludes_zero[!rg_undetermined])),
        undetermined = sprintf("%.0f%%", 100 * mean(rg_undetermined)),
        mean_n_loci_rg = fmt_mc(n_loci_rg, 1),
        .groups = "drop"
      )
  } else {
    data.frame(version = character(0))
  }
  tbl <- dplyr::left_join(sign_tbl, rg_tbl, by = "version") |>
    dplyr::mutate(version = factor(version, levels = version_order)) |>
    dplyr::arrange(version) |>
    dplyr::select(-direction_planted)
  print(knitr::kable(
    tbl, row.names = FALSE,
    caption = paste0(
      "Sign accuracy on real links by each direction call (mean, Monte Carlo ",
      "SE), and the module_rg readout for the true link. ci_excludes_zero and ",
      "mean_rg are over replicates where rg was estimated; undetermined is the ",
      "share where it was not."
    )
  ))
}
```

**How to read it.** The **concordance sign** is expected to read about 1 for
concordant and about 0 for antagonistic programs. That is not noise: each
trait's loci are oriented to its own risk allele before fitting, so a shared
locus loads positively in both traits' programs whatever the direction, and the
sign of `concordance_z` cannot express antagonism. **`module_rg()`** correlates
the two targets' own effects in a common allele frame, so it should read close
to 1 in both directions, with `mean_rg` positive for concordant and negative
for antagonistic. It is the direction call used on real data, and the one to
trust. `undetermined` should be 100% at overlaps 0.10 and 0.20, which leave 4
and 7 shared loci, below the 10 `module_rg()` needs. It should be 0% from 0.33
up (12 or more). On real data too, links resting on fewer than 10 shared loci
get no direction.

## 5. False links when nothing is shared

**What this table shows.** Under the null the two traits have programs of
their own and no correspondence, so every reported link is a false discovery.

```{r null-arm, results='asis'}
null_res <- if (nrow(links) > 0) {
  links[links$family == "null", , drop = FALSE]
} else {
  data.frame()
}
if (nrow(null_res) == 0) {
  cat("The null arm produced no completed replicates.\n\n")
} else {
  print(knitr::kable(
    null_res |>
      dplyr::group_by(programs, tiers) |>
      dplyr::summarise(
        false_links = fmt_mc(n_links, 2),
        reps_with_any_link = sprintf("%d of %d", sum(n_links > 0), dplyr::n()),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "False links per replicate with nothing shared, on all fitted programs ",
      "and on valid programs only (Monte Carlo SE in parentheses); any value ",
      "above 0 is a false discovery."
    )
  ))
}
```

**How to read it.** The trait-specific programs share no loci, so the locus
statistic has nothing to score between them, and false links should be
essentially 0 at both tiers. Treat this as optimistic: the background is
unstructured by default, and a false-discovery rate is only as realistic as the
noise it is measured in (see Notes).

## 6. Scale of each comparison

**What this table shows.** How many programs EBMF fitted per trait, and how many
of them survived validation, per setting.

```{r scale, results='asis'}
if (nrow(links) > 0) {
  print(knitr::kable(
    links[links$tiers == "primary", , drop = FALSE] |>
      dplyr::group_by(version) |>
      dplyr::summarise(
        fitted_trait_a = fmt_mc(n_programs_a[programs == "all"], 1),
        valid_trait_a = fmt_mc(n_programs_a[programs == "valid"], 1),
        fitted_trait_b = fmt_mc(n_programs_b[programs == "all"], 1),
        valid_trait_b = fmt_mc(n_programs_b[programs == "valid"], 1),
        .groups = "drop"
      ) |>
      dplyr::mutate(version = factor(version, levels = version_order)) |>
      dplyr::arrange(version),
    row.names = FALSE,
    caption = paste0(
      "Programs fitted and surviving per-trait validation, per version. Each trait carries ",
      "1 + K_specific planted programs in the shared versions and K_specific in ",
      "the null."
    )
  ))
}
```

**How to read it.** Each trait carries `1 + K_specific` =
`r 1 + params$K_specific` planted programs in the shared versions and
`K_specific` = `r params$K_specific` in the null. Counts close to those mean
EBMF recovered the planted programs almost one-to-one; counts well below mean
programs were lost to validation (section 2 says which). The gap between fitted
and valid is the noise the gates remove (section 1 shows what it is).

# Notes for interpreting runs

- **One shared program, isolated.** Each planted program has its own driver
  studies and its own loci, so the fits here are easy. This checks that the
  pipeline recovers and links a shared program when nothing else interferes;
  it does not test programs that share driver studies or loci with other
  programs in the same trait.
- **Recall is credited per planted program, not per link.** EBMF can split one
  planted program into several factors, so counting links would reward
  splitting.
- **Sign accuracy is conditional on a real link.** A high value alongside a low
  recall means the few links found were directionally right, not that direction
  is well recovered overall.
- **Driver overlap is fixed at 1.** At a shared locus the same background GWAS
  are present in both traits' matrices, so the shared program's driver studies
  are identical in both. Only SNP overlap varies.
- **Background realism is off by default.** `background_corr` and
  `snp_pleiotropy_sd` default to 0, the historical independent, random-signed
  background. The discovery and linking results are largely insensitive to
  that, but **the null is not**: run the realism arm of the grid script before
  quoting a false-link rate.
- **Sample overlap is not simulated.** Two real target traits from the same
  cohort share participants, which biases cross-trait statistics, and
  `module_rg()` in particular, away from the null.
- **Results are not comparable with earlier renders.** Earlier versions of this
  document planted antagonism by flipping driver signs and gave the target a
  flat effect at every locus, and compared a background-study profile axis
  alongside the locus axis. Effects are now generated in a common allele frame,
  and only the locus axis is used.
