---
title: "Investigation of collapsable univariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of collapsable univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(igraph)
```

This vignette combines the **univariate** SNP-clustering workflow with **pathway collapse** from the collapsable bivariate analysis. Gene-linked molecular QTL rows are merged into pathway/gene **features** first; singleton filtering and downstream similarity clustering then operate on the denser feature matrix.

We use **HDL cholesterol** (trait id `1630`) as the target trait.

## Step 1: Construct the feature pleiotropy matrix (X)

**Goal:** Build a matrix where each **column** is a target-trait SNP and each **row** is a **feature**: standard background traits plus pathway/gene features summarising molecular QTL signals.

```{r step1-fetch}
target_trait_id <- 1630

target_trait <- trait(
  target_trait_id,
  include_associations = TRUE
)
cg <- target_trait$coloc_groups
```

### Gene collapse and pathway mapping

1. **Background traits:** non-molecular colocalisation rows (no `gene_id`) are kept as individual trait rows.
2. **Molecular QTL:** merge gene-linked rows by `gene_id` (eQTL, pQTL, etc.) → one z-score per gene × SNP.
3. Run pathway enrichment on genes at target SNP loci.
4. Map genes to enriched pathways (`feature_id = source:term_id`); aggregate by mean within each pathway.
5. Genes outside enriched pathways remain as gene-level features (`gene:<gene_id>`).

```{r step1-feature-matrix}
pleiotropy <- build_feature_pleiotropy_matrix(
  trait_id = target_trait_id,
  coloc_groups = cg,
  snp_key = "variant_id",
  minimum_count_in_network = 2
)

X <- pleiotropy$x_matrix

list(
  target_trait = target_trait$trait$trait_name,
  n_features = nrow(X),
  n_trait_rows = sum(pleiotropy$feature_info$feature_type == "trait"),
  n_pathway_rows = sum(pleiotropy$feature_info$feature_type == "pathway"),
  n_gene_rows = sum(pleiotropy$feature_info$feature_type == "gene"),
  n_snps = ncol(X),
  fraction_non_na = mean(!is.na(X))
)
```

```{r step1-features}
head(pleiotropy$feature_info, 10)
```

```{r step1-dim}
dim(X)
```

Each column is one SNP's profile across features (traits + collapsed molecular networks). Entries are `NA` where a feature has no signal at that SNP's locus.

## Step 1.5: Remove singleton background features

After pathway collapse, many rows still colocalise at **a single SNP only**. Those singleton rows add sparsity without helping compare SNP profiles. We drop feature rows with a signal at exactly one SNP (the target trait row is kept regardless).

```{r step1-filter-singletons}
feature_snp_counts <- rowSums(!is.na(X))
singleton_feature_ids <- names(feature_snp_counts)[feature_snp_counts == 1]
singleton_feature_ids <- setdiff(singleton_feature_ids, as.character(target_trait_id))

list(
  n_features_before = nrow(X),
  n_singleton_features_removed = length(singleton_feature_ids),
  n_features_after = nrow(X) - length(singleton_feature_ids),
  fraction_non_na_before = mean(!is.na(X))
)
```

```{r step1-filter-apply}
X <- X[!rownames(X) %in% singleton_feature_ids, , drop = FALSE]

pleiotropy$feature_info <- pleiotropy$feature_info |>
  dplyr::filter(!feature_id %in% singleton_feature_ids)

list(
  n_features = nrow(X),
  n_trait_rows = sum(pleiotropy$feature_info$feature_type == "trait"),
  n_pathway_rows = sum(pleiotropy$feature_info$feature_type == "pathway"),
  n_gene_rows = sum(pleiotropy$feature_info$feature_type == "gene"),
  n_snps = ncol(X),
  fraction_non_na = mean(!is.na(X))
)
```

```{r step1-filter-dim}
dim(X)
```

## Step 2: Orient profiles relative to the target trait (X\*)

**Goal:** Calibrate each SNP column so profiles reflect what happens when the variant **increases liability** for the target trait.

\[
X^*_{fj} = \mathrm{sign}(z_j) \cdot X_{fj}
\]

where \(z_j\) is the target trait's GWAS z-score at SNP \(j\) (`pleiotropy$z_target`).

```{r step2-orient}
oriented <- orient_pleiotropy_matrix(
  X,
  target_trait_id = target_trait_id,
  z_target = pleiotropy$z_target
)

X_star <- oriented$x_matrix
z_target <- oriented$z_target
target_signs <- oriented$target_signs

table(target_signs, useNA = "ifany")
```

```{r step2-check}
target_row <- X_star[as.character(target_trait_id), ]

list(
  n_negative_target_z = sum(target_row < 0, na.rm = TRUE),
  n_positive_target_z = sum(target_row > 0, na.rm = TRUE),
  mean_abs_target_z = mean(abs(target_row), na.rm = TRUE)
)
```

## Step 3: Calculate the SNP-SNP similarity matrix (S)

**Goal:** Cosine similarity between SNP feature profiles.

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\| \|x^*_k\|}
\]

```{r step3-col-norms}
X_star_filled <- X_star
X_star_filled[is.na(X_star_filled)] <- 0

col_norms <- sqrt(colSums(X_star_filled^2))

summary(col_norms)
```

```{r step3-normalize}
zero_cols <- col_norms == 0
col_norms_div <- col_norms
col_norms_div[zero_cols] <- 1

X_star_norm <- sweep(X_star_filled, 2, col_norms_div, "/")
X_star_norm[, zero_cols] <- 0
```

```{r step3-similarity}
S <- crossprod(X_star_norm)

similarity <- snp_similarity_matrix(X_star)
stopifnot(all.equal(S, similarity$s_matrix, check.attributes = FALSE))
```

```{r step3-dim}
dim(S)
```

```{r step3-summary}
s_upper <- S[upper.tri(S)]
list(
  diag_all_one = all(abs(diag(S) - 1) < 1e-10),
  min_similarity = min(s_upper, na.rm = TRUE),
  max_similarity = max(s_upper, na.rm = TRUE),
  mean_similarity = mean(s_upper, na.rm = TRUE),
  matrix_density = mean(s_upper > 0)
)
```

`matrix_density` is the proportion of SNP pairs with strictly positive off-diagonal similarity (a sparsity summary of the positive similarity subgraph).

## Step 4: Cluster the SNPs

**Goal:** Group variants into functional classes from the geometry of their feature profiles.

We run several clustering methods and compare how many modules each finds. **Signed Louvain** (Gomez signed modularity on the full cosine similarity matrix, including anti-parallel edges) drives module interpretation in Step 5 (modules with more than three SNPs only). A Ward.D2 dendrogram is shown for visual exploration.

```{r step4-cluster}
k <- 3
D <- 1 - S
dist_snp <- as.dist(D)

hc <- hclust(dist_snp, method = "ward.D2")
clusters_hierarchical <- cutree(hc, k = k)

clusters_louvain <- cluster_snp_profiles(
  S,
  method = "community",
  community_algorithm = "louvain"
)
clusters_leading_eigen <- cluster_snp_profiles(
  S,
  k = k,
  method = "community",
  community_algorithm = "leading_eigen"
)

summarise_cluster_method <- function(method, cluster) {
  sizes <- as.integer(table(cluster))
  data.frame(
    method = method,
    n_clusters = length(sizes),
    sizes_gt_2 = paste(sort(sizes[sizes > 2], decreasing = TRUE), collapse = ", "),
    stringsAsFactors = FALSE
  )
}

comparison_summary <- dplyr::bind_rows(
  summarise_cluster_method("hierarchical (Ward.D2, k = 3)", clusters_hierarchical),
  summarise_cluster_method("leading_eigen", clusters_leading_eigen$cluster),
  summarise_cluster_method("signed_louvain", clusters_louvain$cluster)
)

comparison_summary

min_module_size <- 3
louvain_sizes <- table(clusters_louvain$cluster)
large_clusters <- as.integer(names(louvain_sizes)[louvain_sizes > min_module_size])
clusters_analysis <- clusters_louvain$cluster[
  clusters_louvain$cluster %in% large_clusters
]
```

```{r step4-dendrogram, fig.height = 6, fig.width = 10}
plot(hc, labels = names(clusters_hierarchical), cex = 0.4, main = "SNP hierarchical clustering (Ward.D2)")
rect.hclust(hc, k = k, border = 2:4)
```

`sizes_gt_2` lists cluster sizes strictly greater than two for each method. For interpretation below we use **signed Louvain** modules with **more than three SNPs** (`clusters_analysis`).

## Step 5: Interpret biological modules

**Goal:** Identify which background traits and collapsed pathway/gene features drive each signed Louvain module.

```{r step5-driver-info}
driver_info <- pleiotropy$feature_info |>
  dplyr::transmute(
    trait_id = feature_id,
    trait_name = feature_name,
    feature_type = feature_type
  )
```

```{r step5-summarise}
module_drivers <- summarise_snp_modules(
  x_matrix = X_star[, names(clusters_analysis), drop = FALSE],
  cluster = clusters_analysis,
  trait_info = driver_info,
  exclude_trait_id = target_trait_id,
  top_n = 10
)

module_drivers
```

```{r step5-by-cluster}
module_drivers |>
  dplyr::group_by(cluster, feature_type) |>
  dplyr::slice_head(n = 5) |>
  dplyr::select(cluster, feature_type, trait_name, mean_z, abs_mean_z, n_snps) |>
  dplyr::mutate(
    mean_z = round(mean_z, 2),
    abs_mean_z = round(abs_mean_z, 2)
  )
```

Pathway and gene rows in `module_drivers` reflect collapsed molecular QTL networks; trait rows are standard background phenotypes. Positive `mean_z` means the feature tends to increase when module SNPs increase target-trait liability.

### Heatmap of SNP similarity (signed Louvain modules > 3 SNPs)

```{r step5-heatmap, fig.height = 7, fig.width = 7}
snp_order <- names(clusters_analysis)[order(clusters_analysis)]
S_modules <- S[snp_order, snp_order, drop = FALSE]

heatmap(
  S_modules,
  scale = "none",
  main = paste0("Signed Louvain modules (n > ", min_module_size, " SNPs)"),
  xlab = "SNPs",
  ylab = "SNPs"
)
```

Collapsing molecular QTLs into pathway features before singleton filtering reduces row sparsity and lets signed Louvain modules reflect shared biology at both the phenotypic and pathway level.
