---
title: "Investigation of univariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
```

Non-collapsable univariate analysis: each row is a **background trait** (no pathway/gene collapse). Singleton trait rows are removed, SNP profiles are oriented to the target trait, and **signed Louvain** clusters SNPs on cosine similarity `S`. Compare with [Investigation of collapsable univariate analysis](investigation-collapsable-univariate-analysis.html), which uses the same traits and clustering parameters on collapsed feature rows.

We use **BMI** (trait id `1992`).

```{r params}
target_trait_id <- 1992
min_module_size <- 3
min_group_size <- 5
min_snp_signals <- 5L
max_snp_fraction <- 0.8
similarity_threshold <- 0.5
louvain_gamma <- 3
```

## Verification 1: baseline pathway enrichment

Before clustering, ask whether the trait's molecular QTL genes have recoverable pathway signal, and which SNPs those genes sit on.

```{r verify1-baseline}
coloc_groups <- trait(target_trait_id, include_associations = TRUE)$coloc_groups
baseline_pathways <- enrich_trait_pathways(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  sources = c("KEGG", "Reactome")
)

knitr::kable(
  baseline_pathways$summary,
  caption = "Baseline trait pathway enrichment (KEGG + Reactome)",
  na = "",
  align = c("r", "r", "r", "r", "l")
)

knitr::kable(
  baseline_pathways$genes |>
    dplyr::left_join(
      baseline_pathways$snp_genes |>
        dplyr::group_by(gene_id) |>
        dplyr::summarise(
          snps = paste(sort(unique(snp_id)), collapse = ", "),
          .groups = "drop"
        ),
      by = "gene_id"
    ),
  caption = "Genes at trait GWAS SNPs (with linked SNPs)",
  row.names = FALSE
)

if (nrow(baseline_pathways$pathways) > 0) {
  knitr::kable(
    baseline_pathways$pathways |>
      dplyr::mutate(
        p_value = signif(p_value, 3),
        fdr = signif(fdr, 3)
      ),
    caption = "Enriched pathways for the full trait gene set",
    row.names = FALSE
  )
}
```

## Step 1: Pleiotropy matrix

`build_pleiotropy_matrix()` uses colocalisation summary betas (`z = beta / se`) at each target SNP locus.

```{r step1-build}
pleiotropy <- build_pleiotropy_matrix(
  trait_id = target_trait_id,
  snp_key = "variant_id"
)

X <- pleiotropy$x_matrix

knitr::kable(
  data.frame(
    n_traits = nrow(X),
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3)
  ),
  caption = "Pleiotropy matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous traits

Drop trait rows with fewer than `min_snp_signals` non-`NA` values, or with non-`NA` fraction above `max_snp_fraction` (target trait row kept).

```{r step1-sparse-filter}
trait_snp_counts <- rowSums(!is.na(X))
trait_snp_frac <- trait_snp_counts / ncol(X)
sparse_trait_ids <- names(trait_snp_counts)[trait_snp_counts < min_snp_signals]
ubiquitous_trait_ids <- names(trait_snp_frac)[trait_snp_frac > max_snp_fraction]
drop_trait_ids <- setdiff(
  union(sparse_trait_ids, ubiquitous_trait_ids),
  as.character(target_trait_id)
)

X <- X[!rownames(X) %in% drop_trait_ids, , drop = FALSE]
pleiotropy$trait_info <- pleiotropy$trait_info |>
  dplyr::filter(!as.character(trait_id) %in% drop_trait_ids)

list(
  min_snp_signals = min_snp_signals,
  max_snp_fraction = max_snp_fraction,
  n_sparse_traits_removed = length(setdiff(sparse_trait_ids, as.character(target_trait_id))),
  n_ubiquitous_traits_removed = length(setdiff(ubiquitous_trait_ids, as.character(target_trait_id))),
  n_traits_after = nrow(X),
  n_snps = ncol(X)
)
```

## Step 2: Orient profiles (X\*)

Each SNP column is multiplied by `sign(z_j)` so profiles reflect increasing target-trait liability.

```{r step2-orient}
X_star <- orient_pleiotropy_matrix(
  X,
  target_trait_id = target_trait_id
)$x_matrix
```

## Step 3: SNP similarity (S)

```{r step3-similarity}
S <- snp_similarity_matrix(X_star)$s_matrix
s_upper <- S[upper.tri(S)]

list(
  n_snps = ncol(S),
  min_similarity = round(min(s_upper, na.rm = TRUE), 2),
  max_similarity = round(max(s_upper, na.rm = TRUE), 2),
  mean_similarity = round(mean(s_upper, na.rm = TRUE), 2)
)
```

## Step 4: Signed Louvain clustering

Weak edges with `|S| < similarity_threshold` are zeroed before Louvain. Modules with `> min_module_size` SNPs are analysed below.

```{r step4-cluster}
clusters_louvain <- cluster_snp_profiles(
  S,
  method = "community",
  community_algorithm = "louvain",
  similarity_threshold = similarity_threshold,
  gamma = louvain_gamma
)

louvain_sizes <- table(clusters_louvain$cluster)
large_clusters <- as.integer(names(louvain_sizes)[louvain_sizes > min_module_size])
clusters_analysis <- clusters_louvain$cluster[
  clusters_louvain$cluster %in% large_clusters
]

knitr::kable(
  data.frame(
    cluster = as.integer(names(louvain_sizes)),
    n_snps = as.integer(louvain_sizes),
    analysed = as.integer(names(louvain_sizes)) %in% large_clusters
  ),
  caption = "Signed Louvain module sizes",
  align = "rrl"
)
```

```{r step4-heatmap, fig.height = 7, fig.width = 7}
snp_order <- names(clusters_analysis)[order(clusters_analysis)]
heatmap(
  S[snp_order, snp_order, drop = FALSE],
  scale = "none",
  main = paste0("SNP similarity (Louvain modules > ", min_module_size, " SNPs)"),
  xlab = "SNPs",
  ylab = "SNPs"
)
```

## Step 5: Module drivers

Top background **traits** driving each analysed Louvain module (by mean oriented z-score).

```{r step5-drivers}
module_drivers <- summarise_snp_modules(
  x_matrix = X_star[, names(clusters_analysis), drop = FALSE],
  cluster = clusters_analysis,
  trait_info = pleiotropy$trait_info,
  exclude_trait_id = target_trait_id,
  top_n = 10
)

pivot_cluster_traits <- function(df) {
  df |>
    dplyr::group_by(cluster) |>
    dplyr::mutate(rank = dplyr::row_number()) |>
    dplyr::ungroup() |>
    dplyr::select(cluster, rank, trait_name) |>
    tidyr::pivot_wider(
      names_from = cluster,
      values_from = trait_name,
      names_prefix = "Cluster "
    ) |>
    dplyr::rename(`#` = rank)
}

unique_top10 <- module_drivers |>
  dplyr::group_by(trait_id) |>
  dplyr::filter(dplyr::n() == 1L) |>
  dplyr::ungroup() |>
  dplyr::group_by(cluster) |>
  dplyr::slice_head(n = 10) |>
  dplyr::ungroup()

knitr::kable(
  pivot_cluster_traits(module_drivers),
  caption = "Top 10 driver traits per module",
  na = "",
  align = c("r", rep("l", length(large_clusters)))
)

knitr::kable(
  pivot_cluster_traits(unique_top10),
  caption = "Top 10 unique driver traits per module (not in another module's top 10)",
  na = "",
  align = c("r", rep("l", length(large_clusters)))
)
```

## Verification 3: pathway enrichment per Louvain module

For modules with more than `min_group_size` SNPs, re-run the same KEGG + Reactome enrichment and compare to the baseline trait gene set: do the same pathways come back, and do they split cleanly across modules?

```{r verify3-enrichment}
module_enrichment <- enrich_snp_group_pathways(
  groups = clusters_louvain$cluster,
  coloc_groups = coloc_groups,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)
pathway_comparison <- compare_group_pathways(baseline_pathways, module_enrichment)

knitr::kable(
  module_enrichment$summary,
  caption = paste0(
    "Pathway enrichment per Louvain module (> ", min_group_size, " SNPs)"
  ),
  na = "",
  align = c("l", "r", "r", "r", "l")
)

knitr::kable(
  pathway_comparison$group_overlap,
  caption = "Per-module overlap with baseline trait pathways",
  na = "",
  align = c("l", "r", "r", "r")
)

knitr::kable(
  pathway_comparison$pathway_status,
  caption = "Baseline vs module pathway status",
  na = "",
  align = c("l", "l", "r", "r", "l", "l")
)
```

```{r verify3-details, echo=FALSE, results='asis'}
if (length(module_enrichment$by_group) == 0) {
  cat(
    "\n\nNo modules with more than ", min_group_size, " SNPs.\n\n",
    sep = ""
  )
} else {
  for (x in module_enrichment$by_group) {
    cat("\n\n### Module ", x$group, "\n\n", sep = "")
    cat("SNPs: ", x$n_snps, "; genes: ", nrow(x$genes), "\n\n", sep = "")
    if (nrow(x$genes) > 0) {
      print(knitr::kable(x$genes, caption = "Genes at module SNPs", row.names = FALSE))
    }
    if (nrow(x$pathways) > 0) {
      print(knitr::kable(
        x$pathways |>
          dplyr::mutate(
            p_value = signif(p_value, 3),
            fdr = signif(fdr, 3)
          ),
        caption = "Enriched pathways (KEGG + Reactome, FDR <= 0.05)",
        row.names = FALSE
      ))
    }
  }
}
```
