---
title: "Univariate SNP biological clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of SNP clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take a **single trait** and decompose its genetic architecture by grouping SNPs
according to their genome-wide pleiotropic profiles.

The central question is: **within one trait, what distinct biological programs
are represented by its associated SNPs?**

The output is a set of SNP modules, each annotated with background traits,
QTL studies, tissues, genes, and pathways. The runnable implementation is
[Investigation of univariate analysis](investigation-univariate-analysis.html).
[Investigation of EBMF analysis](investigation-ebmf-analysis.html) is the
overlapping-program alternative.

Background traits stay at **study level**. Molecular QTL studies are not
collapsed by gene, tissue, cis/trans, or QTL type.

### Core clustering preferences

Any method we adopt for SNP grouping should satisfy:

1. **Handles high sparsity** — Must natively process large, sparse matrices
   (thousands of singleton feature rows) without treating structural zeros as
   low-variance signal.  Alternatively, we can limit the sparsity by a hard
   threshold on the number of non-missing SNPs per trait.
2. **Signed and sign-driven clustering** — Must preserve the meaning of
   continuous positive/negative z-scores so variants group by **directional
   alignment**, not magnitude alone.
3. **Automated \(K\) discovery** — The engine should learn the number of
   clusters or latent factors from the data (modularity, ELBO), rather than
   relying on an arbitrary cut.  Filtering the results by a minimum module size
   can help with this.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Pleiotropy matrix\\n(study-level rows x SNPs)\"]
  S2 [label=\"2. Orient by sign(z_target)\\noptional asinh compression\"]
  S3 [label=\"3. SNP-SNP cosine similarity S\"]
  S4 [label=\"4. Cluster (signed Louvain default)\"]
  S5 [label=\"5. Reliability + magnitude annotations\"]
  S6 [label=\"6. Interpret reliable modules\\ntraits, tissues, pathways\"]
  S1 -> S2 -> S3 -> S4 -> S5 -> S6
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

### 1. Construct the pleiotropy matrix

Given:

- `nsnp` = independent SNPs associated with the target trait (coloc groups)
- `ntrait` = background studies in GPMap (phenotypic and molecular)

Construct a **traits x SNPs** matrix of signed z-scores (\(z = \beta / \mathrm{SE}\)).
Conceptually this is a SNP-by-trait profile matrix; the code stores traits as
rows and SNPs as columns.

```text
              SNP1   SNP2   SNP3   SNP4   ...
Trait A        z      z      z      z
Trait B        z      z      z      z
Trait C        z      z      z      z
...
```

Sparse rows (few non-missing SNPs) and ubiquitous rows (non-missing on almost
every SNP) are dropped; the target trait row is kept.

### 2. Orient SNP profiles relative to the target trait

Remove arbitrary allele coding so every column represents increasing target
liability. For SNP \(j\):

\[
X^*_{tj} = \mathrm{sign}(z_j)\, X_{tj}
\]

Optional \(\mathrm{asinh}(z)\) compression then limits the leverage of extreme
z-scores without a hard clip.

### 3. Calculate SNP-SNP similarity

Compare every SNP with every other SNP by its oriented trajectory across
background traits. Cosine similarity:

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\|\,\|x^*_k\|}
\]

The result is SNP x SNP:

- \(+1\) = concordant (parallel) profiles
- \(0\) = unrelated
- \(-1\) = discordant (anti-parallel) profiles

### 4. Cluster the SNPs

Group SNPs into modules from \(S\). The univariate vignette uses
`cluster_snp_profiles_louvain()`. Spectral clustering is available as
`cluster_snp_profiles_spectral()`. Soft overlapping programs live in the EBMF
vignette.

The recommended trial order is below. The first method provides the current
baseline; the following methods increasingly test different assumptions.

#### 1. Signed Louvain (current baseline)

Represent \(S\) as a signed weighted graph: SNP = node, similarity = edge.
Weak edges with \(|S| <\) `similarity_threshold` can be zeroed first.
Implemented as `cluster_snp_profiles_louvain()`.

Advantages:

- Returns a data-driven number of modules for fixed graph parameters
- Handles sparse networks
- Signed Louvain uses negative edges (anti-parallel profiles) rather than
  discarding them

Problems:

- Standard (unsigned) Louvain is directionally blind; we therefore use signed
  modularity
- Very sparse graphs can produce many tiny communities
- Resolution (`gamma`) and the similarity threshold both change community size
- Local optimisation can vary across runs or become trapped in suboptimal
  partitions

Use repeated runs across seeds, `gamma`, and `similarity_threshold` as the
reference against which other methods are compared.

#### 2. Signed Leiden (highest-priority addition)

Leiden is a community-detection refinement designed to produce better-connected
and more stable communities than Louvain.

Advantages:

- Closest methodological upgrade to the current pipeline
- Retains sparse graph processing and data-driven module discovery
- Reduces disconnected or poorly connected communities

Problems:

- Standard Leiden implementations generally assume non-negative weights
- A valid implementation needs signed modularity or coupled positive and
  negative graph layers; negative \(S\) values must not simply be discarded
- Resolution and edge-threshold choices still affect the number of modules

#### 3. Empirical Bayes matrix factorization

Treat the SNP x trait matrix as noisy observations of latent programs:

\[
Y = LF^{\top} + E
\]

- \(Y\) = SNP x trait matrix
- \(L\) = SNP loadings
- \(F\) = trait factors
- \(E\) = residual error

Advantages: SNPs can belong to multiple programs; \(K\) is inferred (ELBO);
overlapping biology is the default assumption.

Problems: factors are not necessarily mutually exclusive SNP modules and can
capture broad technical or global effects. Hard module labels require a
separate loading-to-membership rule.

Prefer this if the architecture is expected to be overlapping rather than
mutually exclusive SNP groups. See
[Investigation of EBMF analysis](investigation-ebmf-analysis.html).

#### 4. Signed degree-corrected stochastic block model

Model positive and negative graph edges as arising from latent SNP blocks.
A nested model with minimum-description-length selection can infer the number
of blocks while allowing SNPs to have different overall connectivity.

Advantages:

- Strong conceptual match for sparse signed networks
- Degree correction helps distinguish modules from highly connected hub SNPs
- Model selection can estimate the number of blocks

Problems:

- Weighted signed-edge support is more specialised than Louvain or Leiden
- Considerably more implementation and validation work
- Block structure is not identical to clustering by directional profile

#### 5. HDBSCAN on a sign-preserving embedding

Apply HDBSCAN to a low-dimensional representation of the oriented SNP profiles
or signed similarity graph.

Advantages:

- Learns a data-driven number of clusters
- Can label poorly supported SNPs as noise instead of forcing membership
- Handles clusters with unequal density

Problems:

- Does not operate directly on signed edges
- Results depend on the embedding, distance, and minimum cluster-size choices
- Sparse missing values must be handled before constructing the embedding

#### 6. Spectral clustering

Uses the graph Laplacian and its eigenvectors via
`cluster_snp_profiles_spectral()`.

Advantages: compresses many background traits into major axes of variation.

Problems: fragmented similarity graphs produce many disconnected components;
usually requires selecting \(K\).

### 5. Assessing module quality

Two complementary checks. Graph diagnostics ask whether a module is a real
blob on \(S\). Trait specificity asks whether that blob has a distinct
biological driver.

#### Module reliability scores

After clustering, go back to \(S\) with `summarise_snp_module_quality()`.
These diagnostics are not Louvain-specific: they need \(S\) and hard cluster
labels from any method.

A module is marked `reliable` only if all three hold:

1. Size \(\ge 3\) SNPs
2. Mean internal similarity \(\ge 0.3\)
3. Connectedness \(\ge 0.5\)

Size, internal similarity, and connectedness ask "is this a real blob?"
Downstream interpretation uses only SNPs in modules that pass all three.

**Internal similarity.** Mean signed cosine among member SNPs (diagonal
excluded). High values mean oriented profiles are roughly parallel. The
cutoff \(0.3\) is a weak "same direction, not noise" bar.

External similarity is the mean \(S\) from members to all other SNPs.
**Separation** is internal minus external. It is diagnostic, not a cutoff: a
module can be internally coherent but still sit inside a similar background.

**Connectedness.** Graph density on the thresholded unsigned graph
(\(|S| \ge\) `similarity_threshold`). Internal similarity can look decent if a
few pairs are very similar and the rest are near zero. Connectedness \(\ge 0.5\)
requires at least half of all pairs to be strong edges.

**Silhouette.** Rousseeuw silhouette with distance \(1 - S\) is still reported,
but is not part of the `reliable` rule. Because SNPs can participate in more
than one biological program, negative silhouettes are expected under hard
clustering and should not by themselves discard a coherent module.

The table also reports the number of connected components and the fraction of
SNPs in the largest component; those are not part of the `reliable` rule.

#### Enrichment of background traits

If no traits are specifically elevated in a module relative to the rest of the
SNPs, the module is harder to trust biologically.
`summarise_module_specific_traits()` ranks traits by module-vs-rest contrast:

| Metric | Meaning |
|---|---|
| `abs_mean_z` | Mean absolute z for the trait among SNPs in the module |
| `abs_mean_z_rest` | Mean absolute z among all other SNPs |
| `specificity` | Relative enrichment of the trait inside the module |
| `abs_mean_z_delta` | Absolute lift versus the rest |

\[
\mathrm{specificity} = \frac{\overline{|z|}_{\mathrm{module}}}{\overline{|z|}_{\mathrm{rest}} + 0.1}
\]

\[
\mathrm{abs\_mean\_z\_delta} = \overline{|z|}_{\mathrm{module}} - \overline{|z|}_{\mathrm{rest}}
\]

These describe what drives each cluster without using pathway enrichment as
the clustering criterion.

### 6. Interpret biological modules

The univariate vignette currently interprets **reliable** modules as:

- **Top traits by magnitude** (`abs_mean_z` within the module).
- **Trait-category composition** (`enrich_snp_group_trait_categories()`):
  observed vs expected share against the rest of the trait SNP baseline
  (coloc rows for SNPs attributed to the analysed trait, not the whole GPMap
  universe). Top elevated categories are shown with obs/exp/FE; FDR is noted
  when significant.
- **Tissue composition** (`enrich_snp_group_tissues()`, `background = "trait"`)
  on the same leave-one-out trait baseline.
- **Pathway enrichment** on QTL genes (KEGG and Reactome), treated as
  secondary because the full-trait gene set is often empty for broad traits
  such as BMI.

## Variables to consider

Exposed in the univariate vignette today:

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `api` | `production`, `local` | Public GPMap vs local QTL DB | Local data can add trans-QTL studies, increasing matrix density and changing the regulatory signals that connect SNPs. This can alter both module membership and tissue/gene interpretation. Production is more reproducible but may omit those signals. |
| `min_snp_signals` | `1`, `3`, `5` | Minimum SNPs with a value for a background trait | Raising it removes singleton or very sparse rows, reducing unstable edges and computation. It can also remove rare, module-specific biology. Lowering it preserves those signals but makes similarity more sensitive to sparse observations. |
| `similarity_threshold` | `0`, `0.5` | Minimum \(\|S\|\) retained as a Louvain edge | Raising it removes weak positive and negative edges, producing a sparser graph that may split into more or smaller modules. Lowering it preserves gradual structure but weak or noisy links can merge otherwise distinct modules. |
| `compress_method` | `none`, `asinh` | Transformation applied to z-scores | `asinh` is nearly linear around zero but compresses extreme values, so cosine similarity reflects patterns across several traits rather than one very large z-score. `none` preserves effect scale but allows outliers to dominate direction and norm. |
| `louvain_gamma` | `2`, `3` | Signed Louvain resolution | Higher `gamma` rewards smaller communities and generally increases the number of modules; lower values merge broader profiles. It directly affects module size, separation, connectedness, and the number passing reliability filters. |
| `similarity_missing_method` | `zero_fill`, `pairwise_complete` | How missing trait-SNP cells enter cosine similarity | `zero_fill` treats absence as a measured null effect. `pairwise_complete` compares only jointly observed traits, separating effect alignment from coverage but making low-overlap pairs unstable. This choice changes \(S\), every graph-based clustering result, and all reliability metrics calculated from \(S\). |
| `min_pairwise_overlap` | `3`, `5`, `10` | Shared traits required for pairwise-complete cosine | Raising it removes similarities supported by little common evidence, but also deletes more edges and may isolate sparse SNPs. Lowering it improves graph coverage while allowing extreme similarities based on only a few traits. |
| `overlap_shrinkage` | `0`, `5`, `10` | Strength of low-overlap similarity shrinkage | Higher values pull similarities based on few shared traits more strongly toward zero, reducing false strong edges. Similarities with substantial overlap change little. Excessive shrinkage may hide real but narrowly supported programs. |

Also relevant, but not currently params of the univariate vignette:

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| Association source | `coloc`, `full` | Coloc-merged z-scores vs denser `/associations-full` | `coloc` prioritises locus-level evidence and yields a sparse, specific matrix. `full` increases overlap and may stabilise similarity, but non-colocalising associations can add LD, power, or shared-polygenicity structure and produce broader modules. Not yet a `build_pleiotropy_matrix()` argument; dense associations are used by the perturbation / EBMF builders. |
| Trans QTLs | include / exclude | Whether distant molecular effects are retained | Including trans effects can reveal distributed regulatory programs and improve tissue/pathway interpretation, while increasing density and possible noise. Excluding them gives a more local cis-focused analysis but may miss system-level mechanisms. |
| Clustering method | Louvain, Leiden, EBMF, SBM, HDBSCAN, spectral | Model used to define modules or programs | Louvain/Leiden/SBM produce graph communities; EBMF allows overlapping programs; HDBSCAN can leave uncertain SNPs unassigned; spectral methods usually require \(K\). Different assumptions change module count and membership, so compare stability and interpretation rather than scores alone. |

Trans QTL include/exclude is currently a manual `cis_trans` filter or an API
choice (`local` vs `production`).

Reliability cutoffs used after clustering: `min_module_size = 3`,
`min_mean_internal = 0.3`, `min_connectedness = 0.5`.
