---
title: "Simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  cluster_type: ebmf
  K: 3
  module_sizes: [5, 20, 50]
  min_snp_signals: 5
  min_module_size: 3
  compress_method: asinh
  compress_scale: 2
  similarity_threshold: 0.2
  ebmf_se_mode: unit
  ebmf_beta_scale: trait
  ebmf_lfsr_threshold: 0.05
  ebmf_magnitude_threshold: 0.5
  ebmf_drop_global: true
  ebmf_prior: point_normal
  ebmf_backfit: true
  ebmf_hard_assignment: false
  min_mean_internal: 0.3
  min_connectedness: 0.25
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 5
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  drivers_per_module: 12
  n_background_traits: 40
  effect_size: 3
  noise_sd: 0.5
  p_structural_zero: 0.3
  p_spurious: 0.05
  p_active_background: 0.08
  driver_sharing: 0
  snp_driver_groups: 1
  n_reps: 5
vignette: >
  %\VignetteIndexEntry{Simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

n_reps <- as.integer(params$n_reps)
set.seed(1)

if (!identical(params$cluster_type, "ebmf")) {
  stop("This vignette is configured for cluster_type = 'ebmf'")
}
if (params$K != length(params$module_sizes)) {
  stop("K must equal length(module_sizes)")
}

clustering_settings <- list(
  cluster_type = params$cluster_type,
  min_snp_signals = params$min_snp_signals,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  similarity_threshold = params$similarity_threshold,
  ebmf_se_mode = params$ebmf_se_mode,
  ebmf_beta_scale = params$ebmf_beta_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  ebmf_hard_assignment = params$ebmf_hard_assignment,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  n_null = params$n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = params$n_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)

simulation_defaults <- list(
  module_sizes = params$module_sizes,
  drivers_per_module = params$drivers_per_module,
  n_background_traits = params$n_background_traits,
  effect_size = params$effect_size,
  noise_sd = params$noise_sd,
  p_structural_zero = params$p_structural_zero,
  p_spurious = params$p_spurious,
  p_active_background = params$p_active_background,
  driver_sharing = params$driver_sharing,
  snp_driver_groups = params$snp_driver_groups
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against synthetic data with known ground truth. `simulate_trait()` generates
latent SNP modules, overlapping module support, heterogeneous effects, and
unstructured background. The EBMF analysis keeps overlapping soft programs;
program membership is filtered by lFSR and loading magnitude in
`summarise_ebmf_programs()`, and replication is checked only after internal
coherence passes.

The important analysis parameters are kept separate from the generative
parameters:

- **Analysis** — lFSR threshold, magnitude threshold, minimum coherence,
  stability threshold, trait subsampling, and EBMF noise/prior settings.
- **Generation** — module size, effect size, driver sharing, driver subgroups,
  structural zeros, spurious effects, and background activation.

```{r helpers}
run_one <- function(...) {
  dots <- list(...)
  for (nm in names(simulation_defaults)) {
    if (nm != "module_sizes" && is.null(dots[[nm]])) {
      dots[[nm]] <- simulation_defaults[[nm]]
    }
  }
  if (is.null(dots$K)) {
    dots$K <- length(simulation_defaults$module_sizes)
  }
  if (dots$K == 0) {
    dots$module_sizes <- NULL
  } else if (is.null(dots$module_sizes)) {
    dots$module_sizes <- simulation_defaults$module_sizes
  }
  dots$seed <- sample.int(.Machine$integer.max, 1)

  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  summary <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), validation_settings)
  )
  valid_ids <- summary$programs$program[summary$programs$status == "valid"]
  valid_assigned <- summary$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    res,
    predicted_memberships = valid_assigned
  )
  programs <- summary$programs
  data.frame(
    k_planted = ev$k_planted,
    k_hat = ev$k_hat,
    k_error = ev$k_error,
    ari_all = ev$ari_all,
    ari_structured = ev$ari_structured,
    module_recall = ev$mean_module_recall,
    background_absorbed = ev$background_absorbed,
    coverage = ev$coverage,
    n_programs = nrow(programs),
    n_valid_programs = length(valid_ids),
    n_dropped_size = sum(!programs$size_pass),
    n_dropped_internal = sum(!programs$internal_pass),
    n_dropped_stability = sum(
      programs$stability_checked & !programs$stability_pass
    ),
    n_unassigned = length(sim$ground_truth$module_of_snp) -
      dplyr::n_distinct(valid_assigned$snp_id),
    mean_factor_strength = mean(programs$factor_strength, na.rm = TRUE),
    mean_factor_strength_per_snp = mean(
      programs$factor_strength_per_snp, na.rm = TRUE
    )
  )
}

run_grid <- function(n_reps, label, ...) {
  dplyr::bind_rows(lapply(seq_len(n_reps), function(i) run_one(...))) |>
    dplyr::mutate(scenario = label)
}
```

## Worked example

```{r example}
sim <- do.call(
  simulate_trait,
  c(
    list(
      n_coloc_groups = 120,
      K = params$K,
      overlap = "moderate",
      n_background_snps = 30,
      module_annotations = list(
        list(tissues = "Liver", trait_categories = "Metabolic"),
        NULL,
        NULL
      ),
      seed = 42
    ),
    simulation_defaults
  )
)
res <- do.call(
  run_univariate_clustering,
  c(list(trait_object = sim$trait_object), clustering_settings)
)
program_summary <- do.call(
  summarise_ebmf_programs,
  c(list(clustering_result = res, s_matrix = res$s_matrix), validation_settings)
)
valid_ids <- program_summary$programs$program[
  program_summary$programs$status == "valid"
]
valid_assigned <- program_summary$assigned |>
  dplyr::filter(program %in% valid_ids)
ev <- evaluate_univariate_simulation(
  sim,
  res,
  predicted_memberships = valid_assigned
)

settings_to_show <- c(
  "cluster_type", "min_snp_signals", "min_module_size", "module_sizes",
  "ebmf_lfsr_threshold", "ebmf_magnitude_threshold",
  "min_mean_internal", "min_connectedness", "n_rep",
  "frac_traits", "stability_top_n", "stability_threshold"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$module_sizes <- params$module_sizes
settings_values$stability_top_n <- validation_settings$top_n
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      if (nm == "module_sizes") {
        return(paste(settings_values[[nm]], collapse = ", "))
      }
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "EBMF analysis and validation settings"
)

knitr::kable(
  data.frame(
    metric = c(
      "planted K", "valid programs", "K error", "ARI (all SNPs)",
      "ARI (structured only)", "mean module recall", "background absorbed",
      "coverage"
    ),
    value = c(
      ev$k_planted, ev$k_hat, ev$k_error, round(ev$ari_all, 3),
      round(ev$ari_structured, 3), round(ev$mean_module_recall, 3),
      round(ev$background_absorbed, 3), round(ev$coverage, 3)
    )
  ),
  caption = "Single-run EBMF evaluation"
)
```

```{r example-programs}
knitr::kable(
  program_summary$programs |>
    dplyr::select(
      program, n_snps, mean_internal_similarity, connectedness,
      factor_strength, factor_strength_per_snp, replication,
      stability_checked, stability_pass, status
    ) |>
    dplyr::mutate(dplyr::across(
      c(mean_internal_similarity, connectedness, factor_strength,
        factor_strength_per_snp, replication),
      ~ round(.x, 3)
    )),
  row.names = FALSE,
  caption = "EBMF programs; factor-strength scores are reported but do not gate status"
)
```

```{r example-composition}
if (nrow(valid_assigned) > 0) {
  comp <- do.call(rbind, lapply(sort(unique(valid_assigned$program)), function(pg) {
    ids <- valid_assigned$snp_id[valid_assigned$program == pg]
    tv <- sim$ground_truth$module_of_snp[ids]
    tab <- table(ifelse(tv == 0, "background", paste0("mod", tv)))
    data.frame(
      program = pg,
      n_snps = length(ids),
      n_shared_boundary = sum(ids %in% sim$ground_truth$multi_module_snps),
      truth_mix = paste(names(tab), tab, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }))
  knitr::kable(
    comp,
    row.names = FALSE,
    caption = "Filtered SNP composition of valid EBMF programs"
  )
}
```

## Scenario 1: false structure under the null

Null simulations test whether EBMF creates valid programs from independent
sparse profiles with no planted structure.

```{r null-grid}
null_results <- dplyr::bind_rows(
  run_grid(n_reps, "n=15", n_coloc_groups = 20, K = 0),
  run_grid(n_reps, "n=150", n_coloc_groups = 200, K = 0),
  run_grid(n_reps, "n=1500", n_coloc_groups = 1500, K = 0)
)
null_results |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(
    mean_k_hat = mean(k_hat),
    mean_valid_programs = mean(n_valid_programs),
    mean_coverage = mean(coverage),
    mean_background_absorbed = mean(background_absorbed),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "Null simulations")
```

## Scenario 2: dataset size

The dataset-size grid tests recovery as the number of SNPs increases. The
smallest case is kept above the module-size and sparse-row requirements used
by the EBMF analysis.

```{r size-grid, fig.width=8, fig.height=4}
sizes <- c(100, 200, 500, 900)
size_results <- dplyr::bind_rows(lapply(sizes, function(n) {
  dplyr::bind_rows(
    run_grid(n_reps, "K=0", n_coloc_groups = n, K = 0),
    run_grid(n_reps, "K=", n_coloc_groups = n, K = params$K)
  ) |>
    dplyr::mutate(n_coloc_groups = n)
}))
ggplot(size_results, aes(factor(n_coloc_groups), k_hat, fill = scenario)) +
  geom_boxplot() +
  labs(x = "coloc groups", y = "valid EBMF programs", fill = NULL,
       title = "Valid EBMF programs versus dataset size") +
  theme_minimal()
```

## Scenario 3: signed effects

The signed-effects grid compares coherent, flipped, and random driver signs.
The generated sign pattern is evaluated using the same EBMF thresholds and
stability settings as the worked example.

```{r sign-grid, fig.width=6, fig.height=4}
sign_results <- dplyr::bind_rows(
  run_grid(n_reps, "coherent", n_coloc_groups = 150, K = params$K,
           sign_pattern = "coherent"),
  run_grid(n_reps, "flipped", n_coloc_groups = 150, K = params$K,
           sign_pattern = "flipped"),
  run_grid(n_reps, "random", n_coloc_groups = 150, K = params$K,
           sign_pattern = "random")
)
ggplot(sign_results, aes(scenario, module_recall, fill = scenario)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "driver sign regime", y = "mean module recall",
       title = "EBMF recovery under different driver-sign regimes") +
  theme_minimal()
```

## Scenario 4: module overlap

Shared SNPs can belong to multiple EBMF programs. The evaluator preserves this
overlap for coverage and module recall rather than forcing a one-to-one
partition.

```{r overlap-grid, fig.width=6, fig.height=4}
overlap_results <- dplyr::bind_rows(
  run_grid(n_reps, "disjoint", n_coloc_groups = 300, K = params$K, overlap = "disjoint"),
  run_grid(n_reps, "moderate", n_coloc_groups = 300, K = params$K, overlap = "moderate"),
  run_grid(n_reps, "strong", n_coloc_groups = 300, K = params$K, overlap = "strong")
)
ggplot(overlap_results, aes(scenario, module_recall, fill = scenario)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "overlap regime", y = "mean module recall",
       title = "EBMF recovery as modules increasingly overlap") +
  theme_minimal()
```

## Scenario 5: sparsity

Structural zeros remove true driver observations entirely. This tests whether
filtered EBMF programs retain their planted module signal as support becomes
sparser.

```{r sparsity-grid, fig.width=6, fig.height=4}
sparsity_results <- dplyr::bind_rows(
  run_grid(n_reps, "0.2", n_coloc_groups = 200, K = params$K, p_structural_zero = 0.2),
  run_grid(n_reps, "0.5", n_coloc_groups = 200, K = params$K, p_structural_zero = 0.5),
  run_grid(n_reps, "0.8", n_coloc_groups = 200, K = params$K, p_structural_zero = 0.8)
)
sparsity_results |>
  dplyr::mutate(p_zero = as.numeric(scenario)) |>
  ggplot(aes(p_zero, module_recall)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.05) +
  scale_x_continuous(breaks = c(0.2, 0.5, 0.8)) +
  labs(x = "P(structural zero in module support)", y = "mean module recall",
       title = "EBMF recovery versus within-module sparsity") +
  theme_minimal()
```

## Scenario 6: observation noise

`noise_sd` controls the random effect noise added to simulated trait-SNP
observations. This grid tests how recovery changes as the latent module signal
becomes less distinct from noise.

```{r noise-grid, fig.width=6, fig.height=4}
noise_results <- dplyr::bind_rows(
  run_grid(n_reps, "0.1", n_coloc_groups = 200, K = params$K, noise_sd = 0.1),
  run_grid(n_reps, "0.5", n_coloc_groups = 200, K = params$K, noise_sd = 0.5),
  run_grid(n_reps, "1.0", n_coloc_groups = 200, K = params$K, noise_sd = 1),
  run_grid(n_reps, "2.0", n_coloc_groups = 200, K = params$K, noise_sd = 2)
)
noise_results |>
  dplyr::mutate(noise = as.numeric(scenario)) |>
  ggplot(aes(noise, module_recall)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.05) +
  scale_x_continuous(breaks = c(0.1, 0.5, 1, 2)) +
  labs(x = "noise_sd", y = "mean module recall",
       title = "EBMF recovery versus simulated observation noise") +
  theme_minimal()
```

## Notes for interpreting runs

- `module_recall` and `coverage` preserve overlapping valid EBMF memberships.
- `ari_all` and `ari_structured` use each SNP's strongest assigned program as a
  hard projection, so they remain conventional partition metrics.
- `factor_strength` and `factor_strength_per_snp` are reported scores only;
  they do not currently filter programs.
- lFSR and magnitude thresholds determine `assigned` SNPs before coherence,
  factor-strength, and posterior-evidence calculations.
- Replication is run only for programs that pass internal coherence.
