---
title: "Investigation of SNP perturbation clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of SNP perturbation clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
```

This vignette implements the **SNP perturbation clustering** framework for comparison with the existing univariate / collapsable / EBMF workflows.

Pipeline:

1. Build / filter / orient SNP x trait and SNP x gene matrices
2. **Verification (baseline):** top associated traits + tissue composition
3. Weighted cosine similarity → hierarchical + Leiden clustering
4. **Verification (modules):** top traits and tissue enrichment per Leiden module
5. Pathway enrichment is retained as a secondary check (often empty for BMI)

**Why trait/tissue verification:** BMI pathway enrichment on the full gene set is often blank, so it cannot validate modules. Trait drivers and QTL tissue enrichment still give a readable before/after check.

We use **BMI** (trait id `1992`). Gene scores use weighted Stouffer with \(w_i = 1/\mathrm{SE}_i\).

```{r params}
target_trait_id <- 1992
min_group_size <- 5
min_feature_snps <- 5L
w_gene <- 0.3
w_trait <- 0.7
hierarchical_k <- 8L
leiden_resolution <- 0.5
similarity_threshold <- 0.2
use_column_scaling <- FALSE
compress_method <- "asinh"
top_n_traits <- 10L
```

## Step 1: Build feature matrices

- **Trait matrix:** SNPs x complex traits; values = signed \(z = \beta / \mathrm{SE}\) (best \(p\) per SNP-trait).
- **Gene matrix:** SNPs x genes; values = weighted Stouffer \(Z\) across molecular QTL evidence for each SNP-gene pair.
- After orientation, effects are soft-compressed with \(\mathrm{asinh}(z)\) so extreme /
  corrupt values do not dominate cosine similarity or mean-absolute trait rankings
  (no hard clip).

```{r step1-matrices}
coloc_groups <- trait(target_trait_id, include_associations = TRUE)$coloc_groups
pert <- build_perturbation_matrices(
  trait_id = target_trait_id,
  snp_key = "variant_id"
)

trait_matrix <- pert$trait_matrix
gene_matrix <- pert$gene_matrix

knitr::kable(
  data.frame(
    n_snps = nrow(trait_matrix),
    n_traits = ncol(trait_matrix),
    n_genes = ncol(gene_matrix),
    trait_non_na = round(mean(!is.na(trait_matrix)), 3),
    gene_non_na = round(mean(!is.na(gene_matrix)), 3)
  ),
  caption = "Raw perturbation feature matrix summary",
  align = "r"
)
```

## Step 1.5: Filter, orient, and compress

Drop trait/gene columns seen in fewer than `min_feature_snps` SNPs, flip each SNP
row by \(\mathrm{sign}(z_{\mathrm{target}})\), then apply `asinh` compression.

```{r step1-filter-orient}
filtered <- filter_perturbation_features(
  trait_matrix = trait_matrix,
  gene_matrix = gene_matrix,
  min_snps = min_feature_snps
)
oriented <- orient_perturbation_matrices(
  trait_matrix = filtered$trait_matrix,
  gene_matrix = filtered$gene_matrix,
  z_target = pert$z_target
)
compressed <- compress_perturbation_matrices(
  trait_matrix = oriented$trait_matrix,
  gene_matrix = oriented$gene_matrix,
  method = compress_method
)
trait_matrix <- compressed$trait_matrix
gene_matrix <- compressed$gene_matrix

knitr::kable(
  data.frame(
    n_traits_kept = filtered$n_traits_kept,
    n_traits_dropped = filtered$n_traits_dropped,
    n_genes_kept = filtered$n_genes_kept,
    n_genes_dropped = filtered$n_genes_dropped,
    compress_method = compress_method,
    max_abs_raw = round(
      max(c(abs(oriented$trait_matrix), abs(oriented$gene_matrix)), na.rm = TRUE),
      1
    ),
    max_abs_used = round(max(c(abs(trait_matrix), abs(gene_matrix)), na.rm = TRUE), 3),
    trait_non_na = round(mean(!is.na(trait_matrix)), 3),
    gene_non_na = round(mean(!is.na(gene_matrix)), 3)
  ),
  caption = "After sparse-feature filter, orientation, and asinh compression",
  align = "r"
)
```

## Verification 1: baseline traits and tissues

Before clustering, inspect (a) which complex traits dominate the oriented SNP profiles and (b) which QTL tissues appear at these SNPs.

```{r verify1-traits-tissues}
baseline_traits <- summarise_baseline_traits(
  trait_matrix = trait_matrix,
  trait_info = pert$trait_info,
  coloc_groups = coloc_groups,
  exclude_trait_ids = target_trait_id,
  top_n = 20L,
  n_categories = 2L
)
baseline_tissues <- summarise_baseline_tissues(
  snp_ids = rownames(trait_matrix),
  coloc_groups = coloc_groups,
  snp_key = "variant_id"
)

knitr::kable(
  data.frame(
    top_trait_categories = baseline_traits$top_categories,
    stringsAsFactors = FALSE
  ),
  caption = "Baseline leading trait categories",
  row.names = FALSE
)

knitr::kable(
  baseline_traits$categories |>
    dplyr::mutate(score = round(score, 2)),
  caption = "Baseline trait-category scores (summed abs mean z)",
  row.names = FALSE
)

knitr::kable(
  baseline_traits$traits |>
    dplyr::mutate(
      mean_z = round(mean_z, 3),
      abs_mean_z = round(abs_mean_z, 3)
    ),
  caption = "Baseline top associated traits (all SNPs)",
  row.names = FALSE
)

knitr::kable(
  utils::head(baseline_tissues, 20) |>
    dplyr::mutate(frac_links = round(frac_links, 3)),
  caption = "Baseline QTL tissue composition (top 20)",
  row.names = FALSE
)
```

## Step 2: Optional column-wise scaling

Cosine similarity already L2-normalises each SNP row, so extra column scaling is
optional. Mean-centring after `NA → 0` is harmful on this sparse matrix; if you
enable scaling, use `scale_feature_columns(..., center = FALSE)`.

```{r step2-normalize}
if (isTRUE(use_column_scaling)) {
  trait_for_sim <- scale_feature_columns(trait_matrix, center = FALSE)
  gene_for_sim <- scale_feature_columns(gene_matrix, center = FALSE)
} else {
  trait_for_sim <- trait_matrix
  gene_for_sim <- gene_matrix
  trait_for_sim[is.na(trait_for_sim)] <- 0
  gene_for_sim[is.na(gene_for_sim)] <- 0
}
```

## Step 3: Weighted SNP similarity

\[
S = w_{\mathrm{gene}} S_{\mathrm{gene}} + w_{\mathrm{trait}} S_{\mathrm{trait}}
\]

Trait weight is higher by default because the gene matrix remains much sparser even after filtering.

```{r step3-similarity}
sim <- weighted_snp_similarity(
  trait_matrix = trait_for_sim,
  gene_matrix = gene_for_sim,
  w_gene = w_gene,
  w_trait = w_trait
)
S <- sim$s_matrix
S_off <- S[upper.tri(S)]

knitr::kable(
  data.frame(
    dim = paste(nrow(S), "x", ncol(S)),
    w_gene = w_gene,
    w_trait = w_trait,
    mean_S = round(mean(S), 3),
    mean_S_trait = round(mean(sim$s_trait), 3),
    mean_S_gene = round(mean(sim$s_gene), 3),
    frac_S_gt_0.2 = round(mean(S_off > 0.2), 3),
    frac_S_gt_0.3 = round(mean(S_off > 0.3), 3)
  ),
  caption = "Weighted SNP similarity summary",
  align = "r"
)
```

## Step 4: Cluster SNPs

### Hierarchical clustering

```{r step4-hierarchical}
clusters_hclust <- cluster_perturbation_snps(
  S,
  method = "hierarchical",
  k = hierarchical_k
)

knitr::kable(
  data.frame(
    cluster = as.integer(names(table(clusters_hclust$cluster))),
    n_snps = as.integer(table(clusters_hclust$cluster))
  ) |>
    dplyr::arrange(dplyr::desc(n_snps)),
  caption = paste0("Hierarchical clusters (k = ", hierarchical_k, ")"),
  align = "r"
)
```

### Leiden community detection

Leiden keeps positive edges with similarity at least `similarity_threshold`
(`igraph::cluster_leiden()`). Resolution `0.5` is a practical default on this
weighted cosine graph (resolution `1` tended to shatter into singletons).

```{r step4-leiden}
clusters_leiden <- cluster_perturbation_snps(
  S,
  method = "leiden",
  similarity_threshold = similarity_threshold,
  resolution = leiden_resolution
)

leiden_sizes <- data.frame(
  cluster = as.integer(names(table(clusters_leiden$cluster))),
  n_snps = as.integer(table(clusters_leiden$cluster))
) |>
  dplyr::arrange(dplyr::desc(n_snps))

knitr::kable(
  data.frame(
    method = clusters_leiden$method,
    n_clusters = clusters_leiden$n_clusters,
    n_clusters_gt_min_group = sum(leiden_sizes$n_snps > min_group_size),
    max_cluster_size = max(leiden_sizes$n_snps)
  ),
  caption = "Leiden clustering summary",
  align = c("l", "r", "r", "r")
)

knitr::kable(
  leiden_sizes |>
    dplyr::filter(n_snps > 1),
  caption = "Leiden module sizes (excluding singletons)",
  align = "r"
)
```

```{r step4-heatmap, fig.height = 7, fig.width = 7}
ord <- clusters_hclust$details$hclust$order
heatmap(
  S[ord, ord],
  Rowv = NA,
  Colv = NA,
  scale = "none",
  main = "Weighted SNP similarity (hclust order)",
  xlab = "SNPs",
  ylab = "SNPs"
)
```

## Verification 3: module-specific traits and tissues

Ask whether modules recover drivers elevated vs other SNPs (trait/category
specificity) and tissues enriched vs leave-one-out rest.

```{r verify3-traits}
group_traits <- summarise_module_specific_traits(
  trait_matrix = trait_matrix,
  groups = clusters_leiden$cluster,
  trait_info = pert$trait_info,
  coloc_groups = coloc_groups,
  exclude_trait_ids = target_trait_id,
  min_group_size = min_group_size,
  top_n = top_n_traits,
  n_categories = 2L
)
trait_comparison <- compare_group_traits(
  baseline_traits = baseline_traits,
  group_traits = group_traits,
  top_n = top_n_traits
)

knitr::kable(
  group_traits$summary |>
    dplyr::mutate(
      top_specificity = round(top_specificity, 2),
      top_abs_mean_z = round(top_abs_mean_z, 3)
    ),
  caption = paste0(
    "Module-specific traits/categories per Leiden module (> ",
    min_group_size,
    " SNPs)"
  ),
  na = "",
  row.names = FALSE
)

knitr::kable(
  trait_comparison$group_overlap,
  caption = "Per-module overlap with baseline top traits",
  row.names = FALSE
)

knitr::kable(
  trait_comparison$trait_status |>
    dplyr::filter(status != "baseline_only" | n_groups > 0),
  caption = "Baseline vs module trait-driver status",
  na = "",
  row.names = FALSE
)
```

```{r verify3-trait-details, echo=FALSE, results='asis'}
for (x in group_traits$by_group) {
  cat("\n\n### Module ", x$group, " traits\n\n", sep = "")
  cat("SNPs: ", x$n_snps, "\n\n", sep = "")
  cat("Leading specific categories: ", x$top_categories, "\n\n", sep = "")
  if (nrow(x$categories) > 0) {
    print(knitr::kable(
      x$categories |>
        dplyr::mutate(
          score = round(score, 2),
          mean_specificity = round(mean_specificity, 2)
        ),
      caption = "Module-specific trait categories",
      row.names = FALSE
    ))
  }
  if (nrow(x$traits) > 0) {
    print(knitr::kable(
      x$traits |>
        dplyr::mutate(
          mean_z = round(mean_z, 3),
          abs_mean_z = round(abs_mean_z, 3),
          abs_mean_z_rest = round(abs_mean_z_rest, 3),
          specificity = round(specificity, 2),
          abs_mean_z_delta = round(abs_mean_z_delta, 3)
        ),
      caption = "Module-specific traits",
      row.names = FALSE
    ))
  }
}
```

```{r verify3-tissues}
group_tissues <- enrich_snp_group_tissues(
  groups = clusters_leiden$cluster,
  coloc_groups = coloc_groups,
  snp_key = "variant_id",
  min_group_size = min_group_size,
  background = "rest"
)

knitr::kable(
  group_tissues$summary,
  caption = paste0(
    "Module-specific tissue enrichment (> ", min_group_size, " SNPs; vs rest)"
  ),
  na = "",
  row.names = FALSE
)
```

```{r verify3-tissue-details, echo=FALSE, results='asis'}
for (x in group_tissues$by_group) {
  if (nrow(x$enriched) == 0) {
    next
  }
  cat("\n\n### Module ", x$group, " enriched tissues\n\n", sep = "")
  cat(
    "SNPs: ", x$n_snps,
    "; tissue links: ", x$n_tissue_links, "\n\n",
    sep = ""
  )
  print(knitr::kable(
    x$enriched |>
      dplyr::mutate(
        fold_enrichment = round(fold_enrichment, 2),
        p_value = signif(p_value, 3),
        fdr = signif(fdr, 3)
      ),
    caption = "Enriched tissues vs rest (hypergeometric FDR <= 0.05)",
    row.names = FALSE
  ))
}
```

## Secondary check: pathway enrichment

Pathway enrichment is often empty for the full BMI gene set, so it is not the primary verifier here. Results are summarised only.

```{r verify-pathways-secondary}
baseline_pathways <- enrich_trait_pathways(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  sources = c("KEGG", "Reactome")
)
module_enrichment <- enrich_snp_group_pathways(
  groups = clusters_leiden$cluster,
  coloc_groups = coloc_groups,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)

knitr::kable(
  baseline_pathways$summary,
  caption = "Baseline pathway enrichment (secondary)",
  na = "",
  row.names = FALSE
)

knitr::kable(
  module_enrichment$summary,
  caption = "Per-module pathway enrichment (secondary)",
  na = "",
  row.names = FALSE
)
```

## Comparison notes

| Choice | This framework (tuned) | First draft / naive reading |
|---|---|---|
| Sparse features | Drop columns with `< 5` SNPs | Keep all (gene matrix ~1% filled) |
| Orientation | Target-trait sign flip | None |
| Extreme values | Soft `asinh(z)` compression (no hard clip) | Raw z (GFR/fibrinogen etc. dominate) |
| Column normalisation | Off by default (optional SD scale, no centre) | Mean-centre after `NA → 0` |
| Weights | \(w_t=0.7\), \(w_g=0.3\) | Equal 0.5 / 0.5 |
| Leiden | `threshold=0.2`, `resolution=0.5` | `0.1` / `1` → singletons |
| Verification | Top traits + tissue enrichment (+ pathways secondary) | Pathways only |

Relative to existing gpmapr investigation pipelines, this still differs by keeping separate trait/gene similarities and using Leiden rather than signed Louvain.
