---
title: "Null simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # Simulation parameters (K is always 0: no modules are ever planted)
  n_snps: 300
  n_traits: 600
  noise_sd: 1
  effect_size: 6
  p_active_background: 0.1
  background_sparsity_sd: 1.2
  background_effect_scale: 0.5
  n_hub_traits: 0
  hub_snp_fraction: [0.3, 1.0]
  target_pattern: module
  p_negative: 0

  # EBMF / clustering parameters (inputs to run_univariate_clustering)
  cluster_type: ebmf
  trait_subset: "phenotypic"
  min_snp_signals: 2
  compress_method: asinh
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_se_mode: unit
  ebmf_beta_scale: trait
  ebmf_lfsr_threshold: 0.05
  ebmf_magnitude_threshold: 0.3
  ebmf_drop_global: true
  ebmf_prior: point_normal
  ebmf_backfit: true

  # Program validation parameters (inputs to summarise_ebmf_programs)
  min_module_size: 3
  min_mean_internal: 0.3
  min_connectedness: 0.25
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 5
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5

  # Replication settings
  n_reps: 5
vignette: >
  %\VignetteIndexEntry{Null simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

n_reps <- as.integer(params$n_reps)
set.seed(1)

if (!identical(params$cluster_type, "ebmf")) {
  stop("This vignette is configured for cluster_type = 'ebmf'")
}

clustering_settings <- list(
  cluster_type = params$cluster_type,
  min_snp_signals = params$min_snp_signals,
  trait_subset = params$trait_subset,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  similarity_threshold = params$similarity_threshold,
  ebmf_se_mode = params$ebmf_se_mode,
  ebmf_beta_scale = params$ebmf_beta_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  n_null = params$n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = params$n_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against **pure-null** synthetic data: `K = 0`, so no modules, driver SNPs, or
driver traits are ever planted. Each simulation is generated by
`simulate_trait()` with only the target-trait row plus unstructured background
traits. The pipeline — `run_univariate_clustering()` followed by
`summarise_ebmf_programs()` — is byte-for-byte the same battery used in the
module simulations
(`investigation-univariate-simulations.html`); the analysis settings are also
identical. The only difference is that EBMF is now trying to discover programs
in data where the only structure is randomness.

The four validation checks are exactly the same gates used on the real pipeline:

- `size` — at least `min_module_size` filtered SNPs (`n_snps_filtered`).
- `internal_sim` — mean internal SNP similarity `>= min_mean_internal`.
- `connectivity` — pair connectedness `>= min_connectedness`.
- `stability` — top-loading SNPs recovered with replication
  `>= stability_threshold` under trait subsampling.

A program is `valid` only when every check passes. Under a true null, **any**
program that passes all four checks is a false discovery, so the per-check pass
rates below are false-discovery rates: they show how often each gate (and the
pipeline as a whole) is fooled by random background. `background_absorbed` is
the SNP-level analogue — the fraction of background SNPs pulled into valid
programs.

Because `K = 0`, the module-only generative parameters (`module_sizes`,
`overlap`, `n_traits_per_module`, `trait_overlap`, `driver_sharing`,
`p_structural_zero`, `p_spurious`, `effect_tail`) are inert and are not exposed
here. The active axes of background realism are `p_active_background`
(how often a background trait is observed at a SNP), `background_sparsity_sd`
(heavy-tailed per-trait density variation), `background_effect_scale`
(background magnitude relative to `effect_size`), `noise_sd` (per-cell noise),
and `n_hub_traits` (dense pleiotropic rows that create a shared global factor).
Each version below holds everything else at its realistic baseline and varies
one axis.

```{r settings-table}
settings_to_show <- c(
  "cluster_type", "trait_subset", "min_snp_signals", "min_module_size",
  "compress_method", "compress_scale", "similarity_threshold",
  "ebmf_se_mode", "ebmf_beta_scale", "ebmf_lfsr_threshold",
  "ebmf_magnitude_threshold", "ebmf_drop_global", "ebmf_prior",
  "ebmf_backfit",
  "min_mean_internal", "min_connectedness",
  "n_null", "alpha_membership", "n_candidate_tier", "n_rep",
  "frac_traits", "stability_top_n", "stability_threshold",
  "posterior_stat", "posterior_evidence_cap", "redundancy_threshold", "cores"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$stability_top_n <- validation_settings$top_n
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "Fixed EBMF analysis and validation settings (identical across all null versions)"
)
```

```{r tier-helpers}
null_base <- list(
  n_coloc_groups = params$n_snps,
  K = 0L,
  n_traits = params$n_traits,
  noise_sd = params$noise_sd,
  effect_size = params$effect_size,
  p_active_background = params$p_active_background,
  background_sparsity_sd = params$background_sparsity_sd,
  background_effect_scale = params$background_effect_scale,
  n_hub_traits = params$n_hub_traits,
  hub_snp_fraction = params$hub_snp_fraction,
  target_pattern = params$target_pattern,
  p_negative = params$p_negative
)

# K is pinned to 0 no matter what: these are null simulations by construction.
make_tier_spec <- function(...) {
  overrides <- list(...)
  spec <- utils::modifyList(null_base, overrides)
  spec$K <- 0L
  spec$module_sizes <- NULL
  spec$n_coloc_groups <- params$n_snps
  return(spec)
}

run_tier_one <- function(spec, rep) {
  dots <- spec
  dots$seed <- sample.int(.Machine$integer.max, 1)
  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  smry <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), validation_settings)
  )
  valid_ids <- smry$programs$program[smry$programs$status == "valid"]
  valid_assigned <- smry$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    res,
    predicted_memberships = valid_assigned
  )
  programs <- smry$programs
  programs$rep <- rep(rep, nrow(programs))
  list(
    summary = data.frame(
      n_programs = nrow(programs),
      n_valid_programs = length(valid_ids),
      background_absorbed = ev$background_absorbed,
      stringsAsFactors = FALSE
    ),
    programs = programs
  )
}

run_tier <- function(name, label, spec, n_reps) {
  n_jobs <- 6L
  out <- if (n_reps > 1 && .Platform$OS.type != "windows") {
    parallel::mclapply(
      seq_len(n_reps),
      function(i) run_tier_one(spec, i),
      mc.cores = min(n_jobs, n_reps)
    )
  } else {
    lapply(seq_len(n_reps), function(i) run_tier_one(spec, i))
  }
  list(
    summary = dplyr::bind_rows(lapply(out, `[[`, "summary")) |>
      dplyr::mutate(tier = label),
    programs = dplyr::bind_rows(lapply(out, `[[`, "programs")) |>
      dplyr::mutate(tier = label, tier_name = name)
  )
}
```

## Null versions

Every version plants no modules (`K = 0`); the versions differ only in the
background realism parameter being stressed, with everything else at its
realistic baseline. The **hub_traits** version is deliberately not a pure null:
it plants dense pleiotropic rows so EBMF has a real shared global factor to
latch onto, testing whether such a mega-factor reads as false programs.

```{r tiers}
tier_specs <- list(
  baseline = make_tier_spec(),
  background_absent = make_tier_spec(p_active_background = 0),
  background_sparse = make_tier_spec(p_active_background = 0.02),
  background_dense = make_tier_spec(p_active_background = 0.3),
  sparsity_uniform = make_tier_spec(background_sparsity_sd = 0),
  sparsity_heavy = make_tier_spec(background_sparsity_sd = 1.8),
  weak_background_effects = make_tier_spec(background_effect_scale = 0.25),
  strong_background_effects = make_tier_spec(background_effect_scale = 1.5),
  low_noise = make_tier_spec(noise_sd = 0.25),
  high_noise = make_tier_spec(noise_sd = 4),
  hub_traits = make_tier_spec(n_hub_traits = 15)
)

tier_labels <- c(
  baseline = "baseline (default background)",
  background_absent = "no background associations",
  background_sparse = "sparse background (p_active_background = 0.02)",
  background_dense = "dense background (p_active_background = 0.3)",
  sparsity_uniform = "uniform background density (background_sparsity_sd = 0)",
  sparsity_heavy = "heavy-tailed background density (background_sparsity_sd = 1.8)",
  weak_background_effects = "weak background magnitudes (background_effect_scale = 0.25)",
  strong_background_effects = "strong background magnitudes (background_effect_scale = 1.5)",
  low_noise = "low noise (noise_sd = 0.25)",
  high_noise = "high noise (noise_sd = 4)",
  hub_traits = "pleiotropic hub traits (n_hub_traits = 15)"
)

tier_results <- lapply(names(tier_specs), function(nm) {
  run_tier(nm, tier_labels[[nm]], tier_specs[[nm]], n_reps)
})
names(tier_results) <- names(tier_specs)

tier_summary <- dplyr::bind_rows(lapply(tier_results, `[[`, "summary"))
program_checks <- dplyr::bind_rows(lapply(tier_results, `[[`, "programs"))
```

```{r tier-params-table}
tier_param_table <- do.call(rbind, lapply(names(tier_specs), function(nm) {
  s <- tier_specs[[nm]]
  data.frame(
    version = tier_labels[[nm]],
    K = 0L,
    n_snps = s$n_coloc_groups,
    n_traits = s$n_traits,
    noise_sd = s$noise_sd,
    effect_size = s$effect_size,
    p_active_background = s$p_active_background,
    background_sparsity_sd = s$background_sparsity_sd,
    background_effect_scale = s$background_effect_scale,
    n_hub_traits = s$n_hub_traits,
    hub_snp_fraction = if (s$n_hub_traits > 0) {
      paste(s$hub_snp_fraction, collapse = "-")
    } else {
      "-"
    },
    target_pattern = s$target_pattern,
    p_negative = s$p_negative,
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  tier_param_table,
  row.names = FALSE,
  caption = "Generative parameters for each null version (all versions plant no modules, K = 0)"
)
```

## Aggregate results

Across all replicates and programs of each version, the table below reports
the number of programs EBMF produced before validation (`n programs`) and how
many survived every check (`valid programs`) — under a null these are false
discoveries. `background absorbed` is the fraction of background SNPs pulled
into valid programs. The `pct_*` columns are, for each check, the fraction of
programs that passed it.

```{r aggregate}
check_rates <- program_checks |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    pct_size = 100 * mean(size_pass),
    pct_internal_similarity = 100 * mean(internal_similarity_pass),
    pct_connectedness = 100 * mean(connectedness_pass),
    pct_stability = if (any(stability_checked)) {
      100 * mean(stability_pass[stability_checked])
    } else {
      NA_real_
    },
    pct_valid = 100 * mean(status == "valid"),
    .groups = "drop"
  )

aggregate_results <- tier_summary |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    mean_programs = mean(n_programs),
    mean_valid_programs = mean(n_valid_programs),
    mean_background_absorbed = mean(background_absorbed),
    .groups = "drop"
  ) |>
  dplyr::left_join(check_rates, by = "tier") |>
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric),
    ~ ifelse(is.nan(.x), NA_real_, .x)
  ))

aggregate_results |>
  dplyr::mutate(dplyr::across(
    dplyr::starts_with("pct_"),
    ~ ifelse(is.na(.x), "-", paste0(round(.x), "%"))
  )) |>
  knitr::kable(
    digits = 2,
    col.names = c(
      "version", "n programs", "valid programs",
      "background absorbed",
      "size", "internal sim", "connectivity", "stability", "valid"
    ),
    caption = "Aggregate false-discovery output by null version: programs found, programs passing every validation check, background SNPs absorbed, and per-check pass rates"
  )
```

The `stability` pass rate is computed only over programs that were actually
re-checked (those passing the size and internal-coherence checks).

## Program-level validation checks

For each version, every program found across the replicates is shown with the
four validation checks as separate columns. A check is marked `FAIL` when the
program does not meet that criterion; `n/a` under `stability` means the
program already failed the internal-coherence checks, so stability was not
re-tested.

```{r check-helper, include=FALSE}
check_table <- function(which_tier) {
  df <- program_checks |>
    dplyr::filter(.data$tier_name == which_tier)
  if (nrow(df) == 0) {
    return(NULL)
  }
  df |>
    dplyr::mutate(
      check_size = ifelse(size_pass, "pass", "FAIL"),
      check_internal_similarity = ifelse(internal_similarity_pass, "pass", "FAIL"),
      check_connectedness = ifelse(connectedness_pass, "pass", "FAIL"),
      check_stability = dplyr::case_when(
        !stability_checked ~ "n/a",
        stability_pass ~ "pass",
        TRUE ~ "FAIL"
      ),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      replication = round(replication, 3)
    ) |>
    dplyr::select(
      rep, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, replication,
      check_size, check_internal_similarity, check_connectedness,
      check_stability, status
    ) |>
    dplyr::arrange(rep, program) |>
    knitr::kable(
      row.names = FALSE,
      col.names = c(
        "rep", "program", "n_snps", "n_snps filtered",
        "mean internal similarity", "connectedness", "replication",
        "size", "internal sim", "connectivity", "stability", "status"
      )
    )
}
```

```{r tier-checks, results='asis'}
for (nm in names(tier_specs)) {
  cat("\n###", tier_labels[[nm]], "\n\n")
  tbl <- check_table(nm)
  if (is.null(tbl)) {
    cat("No EBMF programs were produced for this version.\n\n")
  } else {
    cat(tbl, sep = "\n")
    cat("\n\n")
  }
}
```

## Notes for interpreting runs

- Under a true null (`K = 0`, no hub traits) any program passing all four
  validation checks is a **false discovery**; the `valid` column is the
  pipeline-level false-positive rate.
- `n programs` is the number of EBMF programs surviving the lFSR/magnitude
  membership gate, before any validation check is applied. `valid programs` is
  how many of those pass every check.
- `background_absorbed` is the fraction of background SNPs placed into valid
  programs. Because there is no planted structure, every background SNP is
  genuinely background, so this is the SNP-level false-discovery rate.
- `p_structural_zero`, `p_spurious`, `effect_tail`, and the module/driver
  parameters are inert when `K = 0` (they only act on driver rows), so they
  are not exposed here.
- With `p_active_background = 0` every background trait is dropped by the
  `min_snp_signals` filter and the EBMF input reduces to the oriented target
  row; any factor is then the global one and is dropped by `ebmf_drop_global`.
- The `hub_traits` version is not a pure null: dense pleiotropic rows create a
  real shared global factor. Hubs observed at more than `max_snp_fraction`
  (0.8) of SNPs are dropped as ubiquitous before clustering, so only the
  moderately dense hubs reach EBMF. This version tests whether a legitimate
  mega-factor reads as false programs under the validation gates.
- The four checks use the same thresholds as the module simulations
  (`investigation-univariate-simulations.html`), so pass rates here are directly
  comparable to how well the gates separate real modules from background there.