---
title: "Investigation of bivariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of bivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Consider changing this to the SCZ GWAS that I uploaded: c12cb4ba-7216-49ef-9332-b75592fac586

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
```

This vignette builds a bivariate pleiotropy analysis from GPMap colocalisation data. We use **LDL cholesterol** (trait id `1486`) and **HDL cholesterol** (trait id `1630`) as the target pair.

## Step 1: Construct the pleiotropy matrices (X₁ and X₂)

**Goal:** Build two baseline matrices. `X₁` contains pleiotropic profiles for the SNPs associated with Trait 1; `X₂` for Trait 2. Both must share the same background trait rows.

```{r step1-fetch}
trait_id_1 <- 1630
trait_id_2 <- 1486

traits_data <- traits(
  c(trait_id_1, trait_id_2),
  include_associations = TRUE
)
cg <- traits_data$coloc_groups
```

### Union of background traits

Background rows are the union of traits with a colocalisation signal at any SNP locus linked to either target.

```{r step1-build}
bivariate <- build_bivariate_pleiotropy_matrices(
  trait_id_1 = trait_id_1,
  trait_id_2 = trait_id_2,
  coloc_groups = cg,
  snp_key = "variant_id"
)

X1 <- bivariate$x1_matrix
X2 <- bivariate$x2_matrix

list(
  trait_1 = traits_data$traits$trait_name[traits_data$traits$id == trait_id_1],
  trait_2 = traits_data$traits$trait_name[traits_data$traits$id == trait_id_2],
  n_background_traits = nrow(X1),
  n_snps_trait_1 = ncol(X1),
  n_snps_trait_2 = ncol(X2),
  shared_rows = identical(rownames(X1), rownames(X2))
)
```

```{r step1-dims}
dim(X1)
dim(X2)
```

Each column is one target SNP's signed z-score profile (`beta / se`) across the shared background traits. Entries are `NA` where a trait has no colocalisation signal at that SNP's locus.

## Step 2: Orient profiles relative to each target trait (X₁\* and X₂\*)

**Goal:** Independently calibrate each matrix so columns reflect what happens when the variant **increases liability** for that matrix's target trait.

For Trait 1:

\[
X^*_{1,tj} = \mathrm{sign}(z_{1j}) \cdot X_{1,tj}
\]

And analogously for Trait 2 using \(z_2\).

```{r step2-z-targets}
z1 <- X1[as.character(trait_id_1), ]
z2 <- X2[as.character(trait_id_2), ]

list(
  trait_1_signs = table(sign(z1), useNA = "ifany"),
  trait_2_signs = table(sign(z2), useNA = "ifany")
)
```

```{r step2-orient}
X1_star <- sweep(X1, 2, sign(z1), `*`)
X2_star <- sweep(X2, 2, sign(z2), `*`)

oriented_1 <- orient_pleiotropy_matrix(X1, target_trait_id = trait_id_1)
oriented_2 <- orient_pleiotropy_matrix(X2, target_trait_id = trait_id_2)
stopifnot(
  all.equal(X1_star, oriented_1$x_matrix, check.attributes = FALSE),
  all.equal(X2_star, oriented_2$x_matrix, check.attributes = FALSE)
)
```

After orientation, each target trait's own row is non-negative within its matrix; background traits are flipped to align with that target's direction of effect.

```{r step2-check}
list(
  trait_1_negative = sum(X1_star[as.character(trait_id_1), ] < 0, na.rm = TRUE),
  trait_2_negative = sum(X2_star[as.character(trait_id_2), ] < 0, na.rm = TRUE)
)
```

## Step 3: Calculate cross-trait SNP similarity matrix (S\*)

**Goal:** Build a rectangular Trait-1 SNPs × Trait-2 SNPs matrix where each cell measures functional similarity between one variant from each disease.

\[
S_{jk} = \frac{{x^*_{1,j}}^{\top} x^*_{2,k}}{\|x^*_{1,j}\| \|x^*_{2,k}\|}
\]

```{r step3-normalize}
X1_filled <- X1_star
X1_filled[is.na(X1_filled)] <- 0
X2_filled <- X2_star
X2_filled[is.na(X2_filled)] <- 0

col_norms_1 <- sqrt(colSums(X1_filled^2))
col_norms_2 <- sqrt(colSums(X2_filled^2))
zero_cols_1 <- col_norms_1 == 0
zero_cols_2 <- col_norms_2 == 0
col_norms_1[zero_cols_1] <- 1
col_norms_2[zero_cols_2] <- 1

A <- sweep(X1_filled, 2, col_norms_1, "/")
B <- sweep(X2_filled, 2, col_norms_2, "/")
A[, zero_cols_1] <- 0
B[, zero_cols_2] <- 0
```

```{r step3-similarity}
S_star <- crossprod(A, B)

cross_similarity <- cross_snp_similarity_matrix(X1_star, X2_star)
stopifnot(all.equal(S_star, cross_similarity$s_matrix, check.attributes = FALSE))
```

```{r step3-summary}
list(
  dim = dim(S_star),
  min_similarity = min(S_star, na.rm = TRUE),
  max_similarity = max(S_star, na.rm = TRUE),
  mean_similarity = mean(S_star, na.rm = TRUE)
)
```

## Step 4: Identify shared biological modules

**Goal:** Cluster Trait-1 and Trait-2 SNPs independently to reveal block structure in the cross-similarity landscape. Because `S_star` is rectangular, we use dual hierarchical clustering on row and column distance spaces.

```{r step4-cluster}
dist_rows <- dist(S_star, method = "euclidean")
dist_cols <- dist(t(S_star), method = "euclidean")

hc1 <- hclust(dist_rows, method = "average")
hc2 <- hclust(dist_cols, method = "average")

cross_clusters <- cluster_cross_trait_snps(S_star, linkage = "average")
stopifnot(
  identical(hc1$labels, cross_clusters$hclust_rows$labels),
  identical(hc2$labels, cross_clusters$hclust_cols$labels)
)
```

```{r step4-heatmap, fig.height = 7, fig.width = 9}
heatmap(
  S_star,
  Rowv = as.dendrogram(hc1),
  Colv = as.dendrogram(hc2),
  scale = "none",
  main = "Clustered cross-trait SNP similarity blocks",
  xlab = "Trait 2 SNPs",
  ylab = "Trait 1 SNPs"
)
```

Bright blocks along the diagonal of the reordered heatmap indicate groups of SNPs from each trait with parallel pleiotropy footprints.

## Step 5: Interpret the bivariate blocks

Inspecting the clustered heatmap, SNP pairs segregate into structural blocks that describe how the two traits genetically intersect. The table below classifies each row-cluster × column-cluster panel. We then drill into representative blocks to see **which SNPs** fall in each group and **which trait categories** recur in their colocalisation footprints — the biological themes that make those SNP profiles similar.

### Block A: Shared concordant modules (cells ≈ +0.5 to +1.0)

Trait-1 and Trait-2 SNP groups with highly parallel pleiotropy profiles. These variants drive downstream traits in the same direction — a shared biological highway increasing risk for both diseases through the same mechanism.

### Block B: Shared discordant / antagonistic modules (cells ≈ −0.5 to −1.0)

Intensely negative cosine blocks. A biological toggle: variants that increase risk for one trait while protecting against the other — a genetic trade-off signature.

### Block C: Trait-specific modules (cells ≈ 0)

Flat, near-zero panels. Variants acting through isolated mechanisms unique to each trait, largely independent of the other disease network.

```{r step5-cut}
k1 <- 3
k2 <- 3

cross_modules <- cluster_cross_trait_snps(
  S_star,
  linkage = "average",
  k1 = k1,
  k2 = k2
)
```

```{r step5-summarise}
module_summary <- summarise_cross_trait_modules(
  S_star,
  clusters_rows = cross_modules$clusters_rows,
  clusters_cols = cross_modules$clusters_cols
)

module_summary |>
  dplyr::mutate(
    mean_similarity = round(mean_similarity, 2),
    min_similarity = round(min_similarity, 2),
    max_similarity = round(max_similarity, 2)
  )
```

```{r step5-by-type}
module_summary |>
  dplyr::count(block_type, name = "n_blocks") |>
  dplyr::arrange(dplyr::desc(n_blocks))
```

### SNP groups and shared colocalisation themes

For each block type, we take the panel with the strongest mean similarity and ask:

1. **Which SNPs** are in the Trait-1 and Trait-2 clusters?
2. **Which trait categories** appear in coloc groups at those SNP loci (excluding the target traits themselves)?
3. **Which categories overlap** between the two SNP sets — the phenotypic “vocabulary” shared by that block?

```{r step5-block-helpers}
blocks_of_interest <- module_summary |>
  dplyr::filter(block_type %in% c("concordant", "discordant", "trait_specific")) |>
  dplyr::group_by(block_type) |>
  dplyr::slice_max(abs(mean_similarity), n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

snp_category_profile <- function(snp_ids, target_trait_id, coloc_groups) {
  snp_ids <- as.character(snp_ids)
  coloc_groups |>
    dplyr::filter(as.character(variant_id) %in% snp_ids) |>
    dplyr::filter(trait_id != target_trait_id, !is.na(trait_category)) |>
    dplyr::mutate(
      category = dplyr::if_else(data_type != "Phenotype", "Molecular", trait_category)
    ) |>
    dplyr::count(category, name = "n_coloc_signals") |>
    dplyr::arrange(dplyr::desc(n_coloc_signals))
}

snp_examples <- function(snp_ids, snp_info) {
  snp_info |>
    dplyr::filter(as.character(variant_id) %in% as.character(snp_ids)) |>
    dplyr::distinct(display_snp, variant_id, chr, bp) |>
    dplyr::slice_head(n = 6)
}
```

```{r step5-block-details}
block_details <- lapply(seq_len(nrow(blocks_of_interest)), function(i) {
  block <- blocks_of_interest[i, , drop = FALSE]
  row_snps <- names(cross_modules$clusters_rows)[
    cross_modules$clusters_rows == block$row_cluster
  ]
  col_snps <- names(cross_modules$clusters_cols)[
    cross_modules$clusters_cols == block$col_cluster
  ]

  cat_trait_1 <- snp_category_profile(row_snps, trait_id_1, cg)
  cat_trait_2 <- snp_category_profile(col_snps, trait_id_2, cg)
  shared_cats <- dplyr::inner_join(
    cat_trait_1 |> dplyr::rename(trait_1_signals = n_coloc_signals),
    cat_trait_2 |> dplyr::rename(trait_2_signals = n_coloc_signals),
    by = "category"
  ) |>
    dplyr::mutate(total_signals = trait_1_signals + trait_2_signals) |>
    dplyr::arrange(dplyr::desc(total_signals))

  list(
    block_type = block$block_type,
    mean_similarity = block$mean_similarity,
    row_cluster = block$row_cluster,
    col_cluster = block$col_cluster,
    n_snps_trait_1 = length(row_snps),
    n_snps_trait_2 = length(col_snps),
    example_snps_trait_1 = snp_examples(row_snps, bivariate$snp_info_1),
    example_snps_trait_2 = snp_examples(col_snps, bivariate$snp_info_2),
    top_categories_trait_1 = dplyr::slice_head(cat_trait_1, n = 8),
    top_categories_trait_2 = dplyr::slice_head(cat_trait_2, n = 8),
    shared_categories = dplyr::slice_head(shared_cats, n = 10)
  )
})
names(block_details) <- blocks_of_interest$block_type
```

```{r step5-block-overview}
dplyr::bind_rows(lapply(block_details, function(b) {
  data.frame(
    block_type = b$block_type,
    mean_similarity = round(b$mean_similarity, 2),
    row_cluster = b$row_cluster,
    col_cluster = b$col_cluster,
    n_snps_trait_1 = b$n_snps_trait_1,
    n_snps_trait_2 = b$n_snps_trait_2,
    example_snps_trait_1 = paste(b$example_snps_trait_1$display_snp, collapse = ", "),
    example_snps_trait_2 = paste(b$example_snps_trait_2$display_snp, collapse = ", "),
    stringsAsFactors = FALSE
  )
}))
```

**Concordant block** — SNP clusters with parallel pleiotropy footprints. Shared categories below are the colocalisation themes that both SNP sets have in common (lipid/lipoprotein biology is typical for HDL/LDL):

```{r step5-concordant-shared, echo=FALSE, results='asis'}
b <- block_details$concordant
if (!is.null(b) && nrow(b$shared_categories) > 0) {
  knitr::kable(b$shared_categories, caption = "Shared trait categories in the strongest concordant block")
} else {
  cat("No concordant block with overlapping categories in this run.\n")
}
```

**Discordant block** — SNP groups whose profiles point in opposite directions. Shared categories here are phenotypes where both loci have colocalisation evidence but with opposing oriented effects in `S*`:

```{r step5-discordant-shared, echo=FALSE, results='asis'}
b <- block_details$discordant
if (!is.null(b) && nrow(b$shared_categories) > 0) {
  knitr::kable(b$shared_categories, caption = "Shared trait categories in the strongest discordant block")
} else {
  cat("No discordant block with overlapping categories in this run.\n")
}
```

**Trait-specific block** — weak cross-trait similarity; SNP groups may still share some categories locally, but their full pleiotropy profiles do not align across traits:

```{r step5-specific-top, echo=FALSE, results='asis'}
b <- block_details$trait_specific
if (!is.null(b)) {
  knitr::kable(
    dplyr::bind_rows(
      b$top_categories_trait_1 |> dplyr::mutate(trait = "HDL SNPs"),
      b$top_categories_trait_2 |> dplyr::mutate(trait = "LDL SNPs")
    ),
    caption = "Top colocalisation categories in the strongest trait-specific block (by trait)"
  )
}
```

Concordant blocks are candidates for **shared causal pathways** — look for the same trait categories (especially lipid and lipoprotein classes) on both sides. Discordant blocks highlight **opposing biology** on overlapping phenotypic axes. Trait-specific blocks surface mechanisms enriched at one trait's SNPs that do not translate to the other trait's cluster.
