---
title: "Investigation of univariate analysis"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "local"
  include_trans: true
  target_trait_id: 1992
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  compress_method: "asinh"
  compress_scale: 5
  include_situated_gene: false 
  similarity_threshold: 0.2
  ebmf_magnitude_threshold: 0.75
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  min_mean_internal: 0.3
  min_connectedness: 0.5
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 3
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r trait-metadata, include=FALSE}
api <- params$api
target_trait_id <- as.integer(params$target_trait_id)
select_api(api)
target_trait <- trait(target_trait_id, include_associations = TRUE)
target_trait_name <- target_trait$trait$trait_name
```

Single-trait SNP program discovery: each column is a GWAS SNP for the target
trait, each row is a background study (phenotypic or molecular; no collapsing).
Profiles are oriented, compressed, and factorised with empirical Bayes matrix
factorization (EBMF) into overlapping programs. Programs are then validated with
a filtering framework — internal coherence (mean internal similarity and connectedness),
factor strength normalised by the square root of the number of filtered SNPs,
and trait-subsampling stability —
plus descriptive evidence scores (aggregate posterior evidence and factor
distinctiveness) before interpretation.

We use **`r target_trait_name`** (trait id `r target_trait_id`).

```{r params}
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_specificity <- 1.25

# Shared analysis settings come from vignettes/_ebmf_settings.yml so the
# simulation, null-simulation, univariate and multi-trait studies cannot drift
# apart; anything supplied in params (e.g. by the HPC grid scripts) wins.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

trait_subset <- ebmf_settings$trait_subset
min_module_size <- as.integer(ebmf_settings$min_module_size)
min_snp_signals <- as.integer(ebmf_settings$min_snp_signals)
compress_method <- ebmf_settings$compress_method
compress_scale <- as.numeric(ebmf_settings$compress_scale)
include_trans <- isTRUE(ebmf_settings$include_trans)
include_situated_gene <- isTRUE(params$include_situated_gene)
similarity_threshold <- as.numeric(ebmf_settings$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(ebmf_settings$ebmf_lfsr_threshold)
ebmf_prior <- ebmf_settings$ebmf_prior
ebmf_backfit <- isTRUE(ebmf_settings$ebmf_backfit)
ebmf_magnitude_threshold <- as.numeric(
  ebmf_settings$ebmf_magnitude_threshold
)
coherence_n_perm <- as.integer(ebmf_settings$coherence_n_perm)
coherence_q_threshold <- as.numeric(ebmf_settings$coherence_q)
ebmf_greedy_Kmax <- as.integer(ebmf_settings$ebmf_greedy_Kmax)
associations <- params$associations
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
stability_threshold <- as.numeric(params$stability_threshold)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)

knitr::kable(
  data.frame(
    api = api,
    gpmap_url = getOption("gpmap_url"),
    min_snp_signals = min_snp_signals,
    trait_subset = trait_subset,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    include_situated_gene = include_situated_gene,
    min_specificity = min_specificity,
    similarity_threshold = similarity_threshold,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    stability_top_n = stability_top_n,
    stability_threshold = stability_threshold,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    coherence_n_perm = coherence_n_perm,
    coherence_q = coherence_q_threshold,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)
```

## Step 1: Pleiotropy matrix

`build_pleiotropy_matrix()` builds z-scores (`beta / se`) at each target SNP.
With `associations = "coloc"` it uses colocalisation rows only; with
`associations = "full"` it uses all full associations
(`/associations-full`). Molecular QTL studies stay as their own rows.
When `include_trans = FALSE`, trans target markers are removed from the
association rows before the matrix is built.

```{r step1-build}
coloc_groups <- target_trait$coloc_groups
if (!include_trans && "cis_trans" %in% names(coloc_groups)) {
  coloc_groups <- coloc_groups |>
    dplyr::filter(is.na(cis_trans) | tolower(cis_trans) != "trans")
}
pleiotropy <- build_pleiotropy_matrix(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  snp_key = "variant_id",
  associations = associations
)
X <- pleiotropy$x_matrix
cg <- coloc_groups

n_pheno <- sum(pleiotropy$trait_info$feature_type == "phenotypic", na.rm = TRUE)
n_mol <- sum(pleiotropy$trait_info$feature_type == "molecular", na.rm = TRUE)

knitr::kable(
  data.frame(
    trait = target_trait_name,
    n_rows = nrow(X),
    n_phenotypic = n_pheno,
    n_molecular = n_mol,
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3)
  ),
  caption = "Pleiotropy matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous traits

Drop rows with fewer than `min_snp_signals` non-`NA` values, or with non-`NA`
fraction above `max_snp_fraction` (target trait kept).

```{r step1-sparse-filter}
trait_snp_counts <- rowSums(!is.na(X))
trait_snp_frac <- trait_snp_counts / ncol(X)
sparse_trait_ids <- names(trait_snp_counts)[trait_snp_counts < min_snp_signals]
ubiquitous_trait_ids <- names(trait_snp_frac)[trait_snp_frac > max_snp_fraction]
drop_trait_ids <- setdiff(
  union(sparse_trait_ids, ubiquitous_trait_ids),
  as.character(target_trait_id)
)

X <- X[!rownames(X) %in% drop_trait_ids, , drop = FALSE]
data.table::fwrite(X, file = "pleiotropy_matrix_filtered.csv", row.names = TRUE)
pleiotropy$trait_info <- pleiotropy$trait_info |>
  dplyr::filter(!as.character(trait_id) %in% drop_trait_ids)

knitr::kable(
  data.frame(
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    n_sparse_traits_removed = length(setdiff(sparse_trait_ids, as.character(target_trait_id))),
    n_ubiquitous_traits_removed = length(setdiff(ubiquitous_trait_ids, as.character(target_trait_id))),
    n_traits_after = nrow(X),
    n_snps = ncol(X)
  ),
  caption = "After sparse / ubiquitous trait filter",
  align = "r"
)
```

## Step 2: Orient and compress (X\*)

```{r step2-orient}
X_star <- orient_pleiotropy_matrix(
  X,
  target_trait_id = target_trait_id
)$x_matrix
X_star <- compress_effect_matrix(
  X_star,
  method = compress_method,
  asinh_scale = compress_scale
)
trait_matrix <- t(X_star)
```

## Step 3: SNP similarity (S)

Missing trait-SNP cells are treated as zero effects, preserving the original
cosine similarity calculation.

```{r step3-similarity}
similarity <- snp_similarity_matrix(X_star)
S <- similarity$s_matrix
s_upper <- S[upper.tri(S)]
overlap_upper <- similarity$overlap_matrix[
  upper.tri(similarity$overlap_matrix)
]
eligible_upper <- similarity$eligible_matrix[
  upper.tri(similarity$eligible_matrix)
]

knitr::kable(
  data.frame(
    n_snps = ncol(S),
    median_pairwise_overlap = stats::median(overlap_upper),
    fraction_pairs_eligible = round(mean(eligible_upper), 3),
    min_similarity = round(min(s_upper, na.rm = TRUE), 2),
    max_similarity = round(max(s_upper, na.rm = TRUE), 2),
    mean_similarity = round(mean(s_upper, na.rm = TRUE), 2)
  ),
  caption = "SNP similarity summary",
  align = "r"
)
```

## Step 4: Clustering

The oriented compressed matrix `X_star` is factorised directly with EBMF
(features x SNPs), a global mega-factor is dropped if detected, and each SNP is
assigned to the factor with the largest absolute posterior loading among the
factors it significantly belongs to.

```{r step4-cluster}
ebmf_result <- run_univariate_clustering(
  list(
    trait = list(id = target_trait_id),
    coloc_groups = cg
  ),
  target_trait_id = target_trait_id,
  associations = associations,
  trait_subset = trait_subset,
  include_trans = include_trans,
  min_snp_signals = min_snp_signals,
  max_snp_fraction = max_snp_fraction,
  compress_method = compress_method,
  compress_scale = compress_scale,
  ebmf_lfsr_threshold = ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = ebmf_magnitude_threshold,
  ebmf_greedy_Kmax = ebmf_greedy_Kmax,
  ebmf_prior = ebmf_prior,
  ebmf_backfit = ebmf_backfit,
  min_module_size = min_module_size,
  min_mean_internal = min_mean_internal,
  min_connectedness = min_connectedness
)
```

## Step 4.5: Program validation

`summarise_ebmf_programs()` computes internal-coherence metrics on the SNP
similarity graph:

- **mean_internal_similarity** — mean off-diagonal `S` inside the module
- **separation** — internal minus external mean similarity
- **connectedness** — fraction of SNP pairs with `|S| >= similarity_threshold`

For EBMF, `summarise_ebmf_programs()` folds the graph-reliability metrics
together with trait-subsampling **stability** and the descriptive
**additional scores** into a single per-program table with pass/fail flags and
an overall `status`:

- **factor_strength / factor_strength_per_snp** — proportion of observed
  matrix signal attributable to the factor (from its fitted reconstruction),
  computed on the lFSR/magnitude-filtered subset of EBMF-supported SNPs;
  `factor_strength_per_snp = factor_strength / sqrt(n_snps_filtered)`. Reported for
  context only; it does not gate `status`.
- **replication** — trait-subsampling stability: how often the top
  `stability_top_n` loading SNPs are recovered when `1 - frac_traits` of the
  traits are held out and flashier is refit (`n_rep` reruns, run in parallel
  on `cores` cores). Because this is expensive, it is only computed for
  programs that pass the internal-coherence checks; those programs must reach
  `replication >= stability_threshold`. Stability should use the same filtered
  SNP reference sets as the other program-level calculations, and the
  refitted programs are filtered with the same lFSR/magnitude thresholds.
- **posterior_evidence** — aggregate (`posterior_stat`) of `-log10(lfsr)`
  across assigned SNPs. Distinguishes programs dominated by borderline
  `lfsr ~ 0.05` memberships from programs whose memberships carry much stronger
  posterior sign evidence.
- **max_abs_factor_corr / redundant** — maximum absolute correlation of the
  program's SNP loading vector with any other program; programs exceeding
  `redundancy_threshold` are flagged `redundant`.
- **max_pair_redundancy / most_redundant_program / redundancy_pass** —
  factor distinctiveness: for each program, the reciprocal SNP containment
  with every other program (`pair_redundancy = min(shared / n_snps_A,
  shared / n_snps_B)` on the assigned membership sets). The program fails
  (`redundancy_pass = FALSE`) when `max_pair_redundancy >= 0.9`, i.e. it
  captures essentially the same SNP set as another program.

A program is `valid` if it passes all four filters (`size`, `coherence`,
`stability`, `redundancy`); the failure reason is recorded in `status`.

`coherence` is the gate that decides whether a program is more than a random
set of SNPs. It weights **every** SNP by its squared posterior loading -- so it
does not depend on the lFSR/magnitude cutoff -- and permutes that weight vector
across SNPs to ask whether the program's loadings are aligned with the
similarity graph more than chance. Because permuting preserves the weight
distribution, the null is automatically matched on program size and shape.

This replaces comparing `mean_internal_similarity` with an absolute constant,
which is not informative on a real graph: in this trait's similarity matrix a
random set of 44 SNPs has mean internal similarity around 0.35 and clears a 0.3
threshold about 94% of the time, so the old gate sat below the graph's own
baseline. `mean_internal_similarity` and `connectedness` are still reported for
comparison but no longer decide `status`.

`n_eff` (the participation ratio of the loading weights) is the weighted
analogue of program size: read it alongside `coherence_q`, because a small
q-value on a very diffuse program is a weaker claim than the same q-value on a
tight one. `weighted_quorum` and its `quorum_q` measure how many SNP pairs
clear the edge threshold rather than the average alignment, and
`shape_disagreement` flags programs where the two readings fall on opposite
sides of the threshold.

```{r step4-quality, results='asis'}
program_summary <- summarise_ebmf_programs(
    ebmf_result,
    s_matrix = S,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    coherence_n_perm = coherence_n_perm,
    coherence_q_threshold = coherence_q_threshold,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    stability_threshold = stability_threshold,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = TRUE
  )

  valid_ids <- program_summary$programs$program[
    program_summary$programs$status == "valid"
  ]

  print(
    knitr::kable(
      program_summary$programs |>
        dplyr::select(
          program, n_snps, n_snps_filtered,
          mean_internal_similarity, connectedness,
          weighted_internal, weighted_quorum, n_eff, coherence_q, quorum_q,
          coherence_pass, shape_disagreement,
          factor_strength, factor_strength_per_snp, replication,
          posterior_evidence, max_abs_factor_corr, redundant,
          max_pair_redundancy, most_redundant_program, redundancy_pass,
          size_pass, internal_pass,
          stability_pass, stability_checked, status
        ) |>
        dplyr::mutate(
          factor_strength_pct = round(100 * factor_strength, 2),
          factor_strength_per_snp_pct = round(100 * factor_strength_per_snp, 3),
          mean_internal_similarity = round(mean_internal_similarity, 3),
          connectedness = round(connectedness, 3),
          weighted_internal = round(weighted_internal, 3),
          weighted_quorum = round(weighted_quorum, 3),
          n_eff = round(n_eff, 1),
          coherence_q = signif(coherence_q, 3),
          quorum_q = signif(quorum_q, 3),
          coherence_pass = dplyr::if_else(coherence_pass, "pass", "FAIL"),
          shape_disagreement = dplyr::if_else(
            shape_disagreement %in% TRUE, "yes", "no"
          ),
          replication = round(replication, 3),
          posterior_evidence = round(posterior_evidence, 2),
          max_abs_factor_corr = round(max_abs_factor_corr, 3),
          max_pair_redundancy = round(max_pair_redundancy, 3),
          replication = dplyr::if_else(
            is.na(replication), "--", as.character(replication)
          ),
          redundant = dplyr::if_else(redundant, "yes", "no"),
          most_redundant_program = dplyr::if_else(
            is.na(most_redundant_program), "--",
            as.character(most_redundant_program)
          ),
          size_pass = dplyr::if_else(size_pass, "pass", "fail"),
          internal_pass = dplyr::if_else(internal_pass, "pass", "fail"),
          redundancy_pass = dplyr::if_else(redundancy_pass, "pass", "fail"),
          stability_pass = dplyr::case_when(
            !stability_checked ~ "n/a (not checked)",
            stability_pass ~ "pass",
            TRUE ~ "fail"
          )
        ) |>
        dplyr::select(
          program, n_snps, n_snps_filtered,
          mean_internal_similarity, connectedness,
          weighted_internal, weighted_quorum, n_eff, coherence_q, quorum_q,
          coherence_pass, shape_disagreement,
          factor_strength_pct, factor_strength_per_snp_pct, replication, posterior_evidence,
          max_abs_factor_corr, redundant,
          max_pair_redundancy, most_redundant_program, redundancy_pass,
          size_pass, internal_pass, stability_pass, status
        ),
      row.names = FALSE,
      caption = paste0(
        "All EBMF programs returned by flash, with each filter check. n_snps ",
        "is the number of SNPs with nonzero EBMF support and n_snps_filtered is ",
        "the subset surviving the lFSR/magnitude gate; ",
        "factor_strength (as % of observed matrix signal) and ",
        "factor_strength/sqrt(n_snps_filtered) are reported for context but do not gate ",
        "status; replication is only checked for programs passing internal ",
        "coherence (otherwise --); max_pair_redundancy is the reciprocal SNP ",
        "containment with the most overlapping program (most_redundant_program), ",
        "and redundancy_pass fails at >= 0.9; status is 'valid' or lists the ",
        "failed checks (size; internal; stability; redundancy)."
      )
    )
  )

  n_snps_total <- ncol(X)
  n_unique_snps_all <- dplyr::n_distinct(program_summary$assigned$snp_id)
  n_unique_snps_valid <- dplyr::n_distinct(
    program_summary$assigned$snp_id[
      program_summary$assigned$program %in% valid_ids
    ]
  )

  print(
    knitr::kable(
      program_summary$programs |>
        dplyr::summarise(
          n_programs = dplyr::n(),
          n_valid = sum(status == "valid"),
          n_fail_size = sum(!size_pass),
          n_fail_internal = sum(!internal_pass),
          n_fail_stability = sum(!stability_pass),
          n_fail_redundancy = sum(!redundancy_pass, na.rm = TRUE),
          n_redundant = sum(redundant, na.rm = TRUE),
          n_unique_snps_all = n_unique_snps_all,
          n_unique_snps_valid = n_unique_snps_valid,
          pct_of_total_snps_all = round(100 * n_unique_snps_all / n_snps_total, 1),
          pct_of_total_snps_valid = round(100 * n_unique_snps_valid / n_snps_total, 1)
        ),
      row.names = FALSE,
      caption = paste0(
        "EBMF program validation summary (factor strength is reported but does not gate ",
        "status). n_unique_snps counts each SNP once even when it belongs to multiple ",
        "programs (all = across all programs, valid = across valid programs only); ",
        "n_fail_redundancy counts programs with reciprocal SNP containment >= 0.9; ",
        "pct_of_total_snps is relative to the trait's ", n_snps_total, " SNPs."
      )
    )
  )

  soft_memberships <- program_summary$assigned |>
    dplyr::filter(program %in% valid_ids)
  clusters_analysis <- soft_memberships |>
    dplyr::transmute(
      snp_id = snp_id,
      group = as.character(program)
    )
```

```{r step4-heatmap, fig.height = 7, fig.width = 7}
if (is.data.frame(clusters_analysis) && nrow(clusters_analysis) >= 2) {
  if (is.data.frame(clusters_analysis)) {
    heatmap_groups <- stats::setNames(
      clusters_analysis$group[!duplicated(clusters_analysis$snp_id)],
      clusters_analysis$snp_id[!duplicated(clusters_analysis$snp_id)]
    )
  } else {
    heatmap_groups <- clusters_analysis
  }
  snp_order <- names(heatmap_groups)[order(heatmap_groups)]
  snp_order <- intersect(snp_order, rownames(S))
  if (length(snp_order) < 2) {
    cat("Fewer than 2 reported SNPs are present in the similarity matrix; skipping heatmap.\n")
  } else {
    heatmap(
      S[snp_order, snp_order, drop = FALSE],
      Rowv = NA,
      Colv = NA,
      scale = "none",
      main = "SNP similarity (valid EBMF programs)",
      xlab = "SNPs",
      ylab = "SNPs"
    )
  }
} else {
  cat("Fewer than 2 SNPs in reliable modules; skipping heatmap.\n")
}
```

### Program overlap

Because EBMF programs are soft, the same SNP can belong to several programs.
The UpSet plot below shows all valid programs, and for every observed
intersection how many SNPs are shared by exactly that combination (a Venn
diagram generalised to any number of programs). The left-hand bar chart is each
program's size; the dot matrix marks which programs each intersection involves.

```{r step4-overlap, fig.height = 8, fig.width = 9}
if (is.data.frame(clusters_analysis) && nrow(clusters_analysis) >= 2) {
  members_df <- if (is.data.frame(clusters_analysis)) {
    clusters_analysis |>
      dplyr::select(snp_id, group) |>
      dplyr::distinct()
  } else {
    data.frame(
      snp_id = names(clusters_analysis),
      group = as.character(clusters_analysis),
      stringsAsFactors = FALSE
    )
  }
  n_programs <- dplyr::n_distinct(members_df$group)
  if (n_programs >= 2) {
    membership <- members_df |>
      dplyr::mutate(value = 1L) |>
      tidyr::pivot_wider(
        id_cols = "snp_id",
        names_from = "group",
        names_prefix = "P",
        values_from = "value",
        values_fill = 0L
      )
    set_order <- names(sort(colSums(membership[, -1, drop = FALSE]),
                            decreasing = TRUE))
    if (n_programs > 40) {
      cat("More than 40 programs; showing the 40 largest only.\n")
      set_order <- set_order[seq_len(40)]
    }
    plot_ok <- tryCatch({
      membership_plot <- as.data.frame(membership[, c("snp_id", set_order), drop = FALSE])
      rownames(membership_plot) <- membership_plot$snp_id
      membership_plot$snp_id <- NULL
      print(UpSetR::upset(
        membership_plot,
        sets = set_order,
        order.by = "freq",
        nsets = length(set_order),
        nintersects = 100,
        main.bar.color = "#1b9e77",
        sets.bar.color = "#7570b3",
        point.size = 2.6,
        line.size = 0.6
      ))
      TRUE
    }, error = function(e) {
      cat("UpSet plot failed (", conditionMessage(e),
          "); showing pairwise SNP-overlap counts instead.\n")
      FALSE
    })
    if (!plot_ok) {
      sizes <- colSums(membership[, set_order, drop = FALSE])
      overlaps <- outer(set_order, set_order, Vectorize(function(a, b) {
        sum(membership[[a]] & membership[[b]])
      }))
      dimnames(overlaps) <- list(set_order, set_order)
      diag(overlaps) <- sizes[set_order]
      print(knitr::kable(
        as.data.frame(overlaps),
        row.names = TRUE,
        caption = "Pairwise SNP overlap between programs (diagonal = program size)"
      ))
    }
  } else {
    cat("Only one program; skipping program-overlap plot.\n")
  }
} else {
  cat("Fewer than 2 SNPs in reliable modules; skipping program-overlap plot.\n")
}
```

## Step 5: Interpret reliable modules

For each reliable module, rank background traits by **module specificity vs the
rest of the trait** (`summarise_module_specific_traits()`), then summarise
**trait-category and tissue enrichments vs the trait baseline**, plus
pathway hits. `specificity = mean |z| in the module / (mean |z| in the rest +
0.1)` and `abs_mean_z_delta = module - rest` highlight traits that are
specifically elevated in this module rather than merely large everywhere (which
ranking by `abs_mean_z` alone would over-emphasise). Pathway enrichment is
queried across Reactome and KEGG with a separate call per source, and the hits
are reported side by side as **pathways**.

Trait baseline = all SNPs attributed to this trait in the analysis
(`colnames(S)`), using coloc rows for those SNPs only—not
the whole GPMap universe. Each module is compared to the **rest of the trait**
(leave-one-out): observed link share vs expected share, with fold enrichment.

Annotation tables show the top elevated categories/tissues even when FDR is not
significant; FDR is reported when it passes 0.05.

Each hit is scored with a BH-corrected significance summary, thresholded at
0.05:

- **FDR** — hypergeometric test of the module vs the rest of the trait.

The bootstrap `CI` for the fold enrichment shows how precisely the ratio is
estimated; because modules are small, a big ratio built on a few links carries
a wide interval that includes 1 (the null), so read the CI together with the
FDR rather than alone.

```{r step5-interpret}
baseline_snp_ids <- colnames(S)

has_analysis_groups <- if (is.data.frame(clusters_analysis)) {
  nrow(clusters_analysis) > 0
} else {
  length(clusters_analysis) > 0
}

if (!has_analysis_groups) {
  cat("No reliable modules to interpret.\n")
  group_traits <- NULL
  group_categories <- NULL
  group_tissues <- NULL
  module_enrichment <- NULL
} else {
    group_traits <- summarise_module_specific_traits(
    trait_matrix = trait_matrix,
    groups = clusters_analysis,
    trait_info = pleiotropy$trait_info,
    coloc_groups = cg,
    exclude_trait_ids = target_trait_id,
    min_group_size = min_module_size,
      top_n = top_n_traits,
      n_categories = 3L,
      min_specificity = min_specificity
  )
  group_categories <- enrich_snp_group_trait_categories(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    baseline_snp_ids = baseline_snp_ids
  )
  group_tissues <- enrich_snp_group_tissues(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    background = "trait",
    baseline_snp_ids = baseline_snp_ids
  )
  module_enrichment <- enrich_snp_group_pathways(
    groups = clusters_analysis,
    coloc_groups = cg,
    min_group_size = min_module_size,
    sources = c("KEGG", "Reactome"),
    include_situated_gene = include_situated_gene
  )
}

.format_enrichment_hits <- function(comparison, label_col, max_n = 3L) {
  if (is.null(comparison) || nrow(comparison) == 0) {
    return("none")
  }
  # Prefer elevated vs rest-of-trait; fall back to most abundant if none.
  elevated <- comparison[
    is.finite(comparison$fold_enrichment) & comparison$fold_enrichment >= 1,
    ,
    drop = FALSE
  ]
  if (nrow(elevated) == 0) {
    elevated <- comparison
  } else {
    elevated <- elevated[order(
      -elevated$fold_enrichment,
      elevated$fdr,
      elevated[[label_col]]
    ), , drop = FALSE]
  }
  top <- utils::head(elevated, max_n)
  labels <- vapply(seq_len(nrow(top)), function(i) {
    sig <- is.finite(top$fdr[i]) && top$fdr[i] <= 0.05
    fdr_txt <- if (sig) paste0(", FDR=", signif(top$fdr[i], 3)) else ""
    ci <- if (is.finite(top$fe_ci_lower[i]) && is.finite(top$fe_ci_upper[i])) {
      paste0(
        ", CI=[", signif(top$fe_ci_lower[i], 2),
        ",", signif(top$fe_ci_upper[i], 2), "]"
      )
    } else {
      ""
    }
    label <- top[[label_col]][i]
    if (sig) {
      label <- paste0("<b>", label, "</b>")
    }
    paste0(
      label,
      " (obs=", sprintf("%.0f%%", 100 * top$frac_group[i]),
      ", exp=", sprintf("%.0f%%", 100 * top$frac_universe[i]),
      ", FE=", signif(top$fold_enrichment[i], 2),
      ci,
      fdr_txt,
      ")"
    )
  }, character(1))
  paste(labels, collapse = "<br>")
}

.format_pathway_hits <- function(pathways, max_n = 3L, with_source = TRUE) {
  if (is.null(pathways) || nrow(pathways) == 0) {
    return("none")
  }
  top <- utils::head(pathways, max_n)
  labels <- if (with_source) {
    paste0(top$source, ": ", top$description)
  } else {
    top$description
  }
  if ("fdr" %in% names(top)) {
    labels <- paste0(labels, " (FDR=", signif(top$fdr, 3), ")")
  }
  paste(labels, collapse = "<br>")
}
```

```{r step5-module-overview, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  specific_traits <- dplyr::bind_rows(lapply(group_traits$by_group, function(x) {
    passing <- x$traits |>
      dplyr::filter(passes_min_specificity) |>
      utils::head(6L)
    data.frame(
      group = x$group,
      module_specific_traits = if (nrow(passing) == 0) {
        "none"
      } else {
        paste0(
          passing$trait_name,
          " (", round(passing$specificity, 2), ")",
          collapse = "<br>"
        )
      },
      stringsAsFactors = FALSE
    )
  }))

  # SNPs that belong to exactly one program: not shared with any other program.
  members_df <- if (is.data.frame(clusters_analysis)) {
    clusters_analysis |>
      dplyr::select(snp_id, group) |>
      dplyr::distinct()
  } else {
    data.frame(
      snp_id = names(clusters_analysis),
      group = as.character(clusters_analysis),
      stringsAsFactors = FALSE
    )
  }
  snp_group_counts <- members_df |>
    dplyr::count(snp_id, name = "n_groups")
  unique_by_group <- members_df |>
    dplyr::filter(
      snp_id %in% snp_group_counts$snp_id[snp_group_counts$n_groups == 1]
    ) |>
    dplyr::count(group, name = "n_unique_snps")

  # All significant hits per module, not just the top one.
  .format_enriched_all <- function(by_group_list, value_col) {
    vapply(by_group_list, function(x) {
      if (is.null(x$enriched) || nrow(x$enriched) == 0) {
        return("none")
      }
      hits <- x$enriched |>
        dplyr::filter(is.finite(fold_enrichment), is.finite(fdr)) |>
        dplyr::arrange(fdr, dplyr::desc(fold_enrichment))
      if (nrow(hits) == 0) {
        return("none")
      }
      paste0(
        hits[[value_col]],
        " (FE=", signif(hits$fold_enrichment, 2),
        ", FDR=", signif(hits$fdr, 3), ")",
        collapse = "<br>"
      )
    }, character(1))
  }
  .format_pathways_all <- function(by_group_list, source) {
    vapply(by_group_list, function(x) {
      if (is.null(x$pathways) || nrow(x$pathways) == 0) {
        return("none")
      }
      hits <- x$pathways[
        x$pathways$source == source &
          !is.na(x$pathways$fdr) & x$pathways$fdr <= 0.05,
        ,
        drop = FALSE
      ]
      if (nrow(hits) == 0) {
        return("none")
      }
      hits <- hits[order(hits$fdr), , drop = FALSE]
      paste0(
        hits$description, " (FDR=", signif(hits$fdr, 3), ")",
        collapse = "<br>"
      )
    }, character(1))
  }

  enrichment_overview <- data.frame(
    group = vapply(
      group_categories$by_group,
      function(x) as.character(x$group),
      character(1)
    ),
    top_enriched_category = .format_enriched_all(
      group_categories$by_group, "trait_category"
    ),
    top_enriched_tissue = .format_enriched_all(group_tissues$by_group, "tissue"),
    top_enriched_reactome = .format_pathways_all(module_enrichment$by_group, "Reactome"),
    top_enriched_kegg = .format_pathways_all(module_enrichment$by_group, "KEGG"),
    stringsAsFactors = FALSE
  )

  overview <- dplyr::full_join(
    group_traits$summary |>
      dplyr::select(group, n_snps),
    unique_by_group,
    by = "group"
  ) |>
    dplyr::full_join(enrichment_overview, by = "group") |>
    dplyr::left_join(specific_traits, by = "group") |>
    dplyr::mutate(
      n_unique_snps = as.integer(tidyr::replace_na(n_unique_snps, 0L)),
      top_enriched_category = tidyr::replace_na(top_enriched_category, "none"),
      top_enriched_tissue = tidyr::replace_na(top_enriched_tissue, "none"),
      top_enriched_reactome = tidyr::replace_na(top_enriched_reactome, "none"),
      top_enriched_kegg = tidyr::replace_na(top_enriched_kegg, "none")
    ) |>
    dplyr::arrange(dplyr::desc(n_snps), as.integer(group)) |>
    dplyr::select(
      group, n_snps, n_unique_snps, module_specific_traits, dplyr::everything()
    )

  cat("\n\n### Overview of significant annotations across modules\n\n")
  print(knitr::kable(
    overview,
    caption = "Top 6 module-specific traits and all significant category, tissue, and pathway (Reactome and KEGG separately) hits; n_snps = SNPs in the program, n_unique_snps = SNPs not present in any other program; trait values are specificity scores, and FDR is the BH-corrected hypergeometric enrichment (threshold 0.05)",
    row.names = FALSE,
    escape = FALSE
  ))
}
```

```{r step5-module-panels, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  categories_by_group <- stats::setNames(
    group_categories$by_group,
    vapply(
      group_categories$by_group,
      function(x) as.character(x$group),
      character(1)
    )
  )
  tissues_by_group <- stats::setNames(
    group_tissues$by_group,
    vapply(group_tissues$by_group, function(x) as.character(x$group), character(1))
  )
  pathways_by_group <- stats::setNames(
    module_enrichment$by_group,
    vapply(module_enrichment$by_group, function(x) as.character(x$group), character(1))
  )
}
```

## Step 5.5: Continuous loading-based enrichment (comparison)

Steps above gate SNPs into a module with a hard lFSR/magnitude cutoff
(`program_summary$assigned`) before testing enrichment. As an alternative,
`enrich_program_loadings_trait_categories()`, `enrich_program_loadings_tissues()`,
and `enrich_program_loadings_pathways()` skip that cutoff entirely: every SNP
with a finite EBMF loading contributes to every program's test, weighted by
its squared loading, regressed against each category/tissue/pathway's local
0/1 membership indicator (`lm(loading^2 ~ membership)`), following the
approach in `scripts/enrichment_via_loadings_example.r`. This section runs
them on the same `ebmf_result` and `cg` used above, for the same `valid_ids`
programs, so the two approaches can be compared side by side.


```{r step5-5-continuous-enrichment}
if (!has_analysis_groups) {
  cat("No reliable modules to interpret.\n")
  loading_categories <- NULL
  loading_tissues <- NULL
  loading_pathways <- NULL
} else {
  loading_categories <- enrich_program_loadings_trait_categories(
    clustering_result = ebmf_result,
    coloc_groups = cg,
    snp_key = "variant_id"
  )
  loading_tissues <- enrich_program_loadings_tissues(
    clustering_result = ebmf_result,
    coloc_groups = cg,
    snp_key = "variant_id"
  )
  pathway_mapping_universe <- tryCatch(
    pathway_mappings(),
    error = function(e) {
      cat("Pathway mappings endpoint unavailable (", conditionMessage(e), "); ",
          "skipping continuous pathway enrichment.\n", sep = "")
      NULL
    }
  )
  loading_pathways <- if (is.null(pathway_mapping_universe)) {
    NULL
  } else {
    enrich_program_loadings_pathways(
      clustering_result = ebmf_result,
      coloc_groups = cg,
      mappings = pathway_mapping_universe,
      snp_key = "variant_id",
      sources = c("KEGG", "Reactome"),
      include_situated_gene = include_situated_gene
    )
  }
}
```

```{r step5-5-continuous-enrichment-tables, echo=FALSE, results='asis'}
# Effect-size floor on top of FDR<=0.05: r_squared is the fraction of a
# program's squared-loading variance the category/tissue/pathway explains --
# bounded [0, 1] and comparable across programs (unlike the raw `enrichment`
# slope, whose scale depends on that program's own loadings). 0.05 sits
# between Cohen's (1988) "small" (0.01) and "medium" (0.09) benchmarks.
min_r_squared <- 0.05

# Same shape as .format_enriched_all()/.format_pathways_all() above (step5-module-overview),
# but reading the continuous comparison tables (enrichment effect instead of
# fold_enrichment) so the two overview tables line up column-for-column.
# Hits are filtered on FDR and r_squared together, then ordered by effect
# (the raw enrichment slope) rather than by significance.
.format_continuous_enriched_all <- function(by_program_list, value_col) {
  vapply(by_program_list, function(x) {
    hits <- x$comparison
    if (is.null(hits) || nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[
      is.finite(hits$fdr) & hits$fdr <= 0.05 &
        is.finite(hits$r_squared) & hits$r_squared >= min_r_squared,
      ,
      drop = FALSE
    ]
    if (nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[order(-hits$enrichment), , drop = FALSE]
    paste0(
      hits[[value_col]],
      " (effect=", signif(hits$enrichment, 2),
      ", R2=", signif(hits$r_squared, 2),
      ", FDR=", signif(hits$fdr, 3), ")",
      collapse = "<br>"
    )
  }, character(1))
}
.format_continuous_pathways_all <- function(by_program_list, source) {
  vapply(by_program_list, function(x) {
    hits <- x$comparison
    if (is.null(hits) || nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[
      hits$source == source & is.finite(hits$fdr) & hits$fdr <= 0.05 &
        is.finite(hits$r_squared) & hits$r_squared >= min_r_squared,
      ,
      drop = FALSE
    ]
    if (nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[order(-hits$enrichment), , drop = FALSE]
    paste0(
      hits$description, " (effect=", signif(hits$enrichment, 2),
      ", R2=", signif(hits$r_squared, 2),
      ", FDR=", signif(hits$fdr, 3), ")",
      collapse = "<br>"
    )
  }, character(1))
}

if (!is.null(loading_categories) && length(loading_categories$by_program) > 0) {
  valid_programs <- as.character(valid_ids)
  by_program_valid <- function(res) {
    if (is.null(res)) {
      return(list())
    }
    out <- Filter(
      function(x) as.character(x$program) %in% valid_programs,
      res$by_program
    )
    stats::setNames(out, vapply(out, function(x) as.character(x$program), character(1)))
  }

  categories_by_program <- by_program_valid(loading_categories)
  tissues_by_program <- by_program_valid(loading_tissues)
  pathways_by_program <- by_program_valid(loading_pathways)

  continuous_enrichment_overview <- data.frame(
    group = names(categories_by_program),
    top_enriched_category = .format_continuous_enriched_all(
      categories_by_program, "trait_category"
    ),
    top_enriched_tissue = .format_continuous_enriched_all(
      tissues_by_program[names(categories_by_program)], "tissue"
    ),
    top_enriched_reactome = if (length(pathways_by_program) == 0) {
      "not available"
    } else {
      .format_continuous_pathways_all(
        pathways_by_program[names(categories_by_program)], "Reactome"
      )
    },
    top_enriched_kegg = if (length(pathways_by_program) == 0) {
      "not available"
    } else {
      .format_continuous_pathways_all(
        pathways_by_program[names(categories_by_program)], "KEGG"
      )
    },
    stringsAsFactors = FALSE
  )

  # Same n_snps / n_unique_snps columns as the hard-cutoff overview above, so
  # the two tables describe the same modules and only the enrichment columns
  # differ.
  continuous_overview <- group_traits$summary |>
    dplyr::select(group, n_snps) |>
    dplyr::full_join(unique_by_group, by = "group") |>
    dplyr::full_join(continuous_enrichment_overview, by = "group") |>
    dplyr::mutate(
      n_unique_snps = as.integer(tidyr::replace_na(n_unique_snps, 0L)),
      top_enriched_category = tidyr::replace_na(top_enriched_category, "none"),
      top_enriched_tissue = tidyr::replace_na(top_enriched_tissue, "none"),
      top_enriched_reactome = tidyr::replace_na(top_enriched_reactome, "none"),
      top_enriched_kegg = tidyr::replace_na(top_enriched_kegg, "none")
    ) |>
    dplyr::arrange(dplyr::desc(n_snps), as.integer(group))

  cat("\n\n### Overview of significant annotations across modules (continuous, no SNP cutoff)\n\n")
  print(knitr::kable(
    continuous_overview,
    caption = paste0(
      "Continuous loading-based enrichment (lm(loading^2 ~ membership), no SNP ",
      "cutoff) for the same modules and n_snps/n_unique_snps as the hard-cutoff ",
      "overview above. FDR threshold at 0.05, R2 threshold at ", min_r_squared,
      "across every program tested for this trait, separately for each of the ",
      "three enrichment types (categories, tissues, pathways); pathway FDR also ",
      "pools KEGG and Reactome into that same family rather than correcting them ",
      "separately."
    ),
    row.names = FALSE,
    escape = FALSE
  ))
}
```

## Step 6: Pleiotropy of program-unique vs shared SNPs

Because EBMF programs are soft, a SNP can be assigned to exactly one valid
program (**unique**) or to several (**shared**). If programs capture distinct
biological programmes, SNPs that uniquely define a single program should be
more specific — i.e. carry a lower pleiotropy score (fewer distinct trait
categories and protein-coding genes via coloc groups) than SNPs whose signal is
shared across programs. We test that here with
`get_all_variant_pleiotropies()`.

```{r step6-unique-vs-shared-pleiotropy, fig.width=7, fig.height=5, out.width="100%"}
if (is.data.frame(clusters_analysis) && nrow(clusters_analysis) > 0) {
  members_df <- clusters_analysis |>
    dplyr::select(snp_id, group) |>
    dplyr::distinct()

  snp_class <- members_df |>
    dplyr::count(snp_id, name = "n_programs") |>
    dplyr::mutate(
      membership_class = dplyr::if_else(n_programs == 1, "unique", "shared")
    )

  variant_pleiotropy <- get_all_variant_pleiotropies() |>
    dplyr::mutate(variant_id = as.character(variant_id)) |>
    dplyr::select(
      variant_id, distinct_trait_categories, distinct_protein_coding_genes
    )

  pleio <- snp_class |>
    dplyr::left_join(
      variant_pleiotropy,
      by = c("snp_id" = "variant_id")
    ) |>
    tidyr::pivot_longer(
      c(distinct_trait_categories, distinct_protein_coding_genes),
      names_to = "metric",
      values_to = "pleiotropy"
    ) |>
    dplyr::mutate(
      metric = dplyr::if_else(
        metric == "distinct_trait_categories",
        "trait-category pleiotropy (distinct categories per SNP)",
        "gene pleiotropy (distinct genes per SNP)"
      ),
      membership_class = factor(membership_class, levels = c("unique", "shared"))
    )

  print(knitr::kable(
    pleio |>
      dplyr::group_by(membership_class, metric) |>
      dplyr::summarise(
        n_snps = sum(!is.na(pleiotropy)),
        mean_pleiotropy = round(mean(pleiotropy, na.rm = TRUE), 2),
        median_pleiotropy = round(stats::median(pleiotropy, na.rm = TRUE), 2),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "Pleiotropy of program-unique vs shared SNPs across valid programs (n = ",
      dplyr::n_distinct(snp_class$snp_id), " SNPs)"
    )
  ))

  test_tbl <- lapply(unique(pleio$metric), function(m) {
    sub <- pleio[pleio$metric == m, ]
    p <- tryCatch(
      stats::wilcox.test(pleiotropy ~ membership_class, data = sub)$p.value,
      error = function(e) NA_real_
    )
    data.frame(metric = m, wilcoxon_p = signif(p, 3), stringsAsFactors = FALSE)
  }) |>
    dplyr::bind_rows()
  print(knitr::kable(
    test_tbl,
    row.names = FALSE,
    caption = "Wilcoxon rank-sum test comparing pleiotropy of unique vs shared SNPs"
  ))

  if (nrow(pleio) > 0) {
    print(ggplot2::ggplot(
      pleio,
      ggplot2::aes(x = membership_class, y = pleiotropy, fill = membership_class)
    ) +
      ggplot2::geom_boxplot(alpha = 0.6, outlier.shape = NA) +
      ggplot2::geom_jitter(width = 0.15, alpha = 0.35, size = 0.8) +
      ggplot2::facet_wrap(~metric, scales = "free_y") +
      ggplot2::scale_fill_manual(values = c(unique = "#1b9e77", shared = "#7570b3")) +
      ggplot2::labs(
        title = "Pleiotropy of program-unique vs shared SNPs",
        subtitle = "unique = assigned to exactly one valid program; shared = assigned to >= 2",
        x = NULL,
        y = "pleiotropy score"
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(legend.position = "none"))
  }
} else {
  cat("No valid programs to classify SNPs.\n")
}
```
