---
title: "Investigation of multi-trait EBMF program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of multi-trait EBMF program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "prod"
  include_trans: false
  trait_ids: [1992, 114]
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  compress_method: "asinh"
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_se_mode: "unit"
  ebmf_beta_scale: "trait"
  ebmf_magnitude_threshold: 0.75
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  min_mean_internal: 0.3
  min_connectedness: 0.5
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 0
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  n_perm: 200
  fdr_threshold: 0.05
  zero_tol: 0.00000001
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

This vignette extends the single-trait investigation
(`investigation-univariate-analysis.html`) to **multiple traits**. EBMF programs
are discovered independently for each target trait, then cross-trait program
pairs are compared. The primary similarity is the Pearson correlation of the
programs' SNP loading vectors (`F_pm`), aligned on the union of their SNPs; a
supporting Jaccard overlap is taken on each program's high-confidence
(lFSR/magnitude-gated) SNP set. A matched permutation null turns every
correlation into an empirical p-value, which is BH-corrected across all
cross-trait pairs. Credibly corresponding programs are then joined into
**multi-trait program families** (weak connected components of the significant
graph), so no one-to-one matching is forced.

Traits are supplied as an array in `params$trait_ids`. We use the defaults:
BMI (`1992`) and Type 2 Diabetes (`114`).

```{r params}
trait_ids <- as.character(params$trait_ids)
if (length(trait_ids) < 2) {
  stop("At least two trait_ids are required for a multi-trait comparison")
}

api <- params$api
trait_subset <- params$trait_subset
min_module_size <- as.integer(params$min_module_size)
min_snp_signals <- as.integer(params$min_snp_signals)
compress_method <- params$compress_method
compress_scale <- as.numeric(params$compress_scale)
include_trans <- isTRUE(params$include_trans)
similarity_threshold <- as.numeric(params$similarity_threshold)
ebmf_se_mode <- params$ebmf_se_mode
ebmf_beta_scale <- params$ebmf_beta_scale
ebmf_lfsr_threshold <- as.numeric(params$ebmf_lfsr_threshold)
ebmf_prior <- params$ebmf_prior
ebmf_backfit <- isTRUE(params$ebmf_backfit)
ebmf_magnitude_threshold <- if (is.na(params$ebmf_magnitude_threshold)) {
  NA_real_
} else {
  as.numeric(params$ebmf_magnitude_threshold)
}
associations <- params$associations
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
stability_threshold <- as.numeric(params$stability_threshold)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)
n_perm <- as.integer(params$n_perm)
fdr_threshold <- as.numeric(params$fdr_threshold)
zero_tol <- as.numeric(params$zero_tol)

select_api(api)

knitr::kable(
  data.frame(
    api = api,
    trait_ids = paste(trait_ids, collapse = ", "),
    trait_subset = trait_subset,
    min_snp_signals = min_snp_signals,
    min_module_size = min_module_size,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    similarity_threshold = similarity_threshold,
    ebmf_se_mode = ebmf_se_mode,
    ebmf_beta_scale = ebmf_beta_scale,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    n_rep = n_rep,
    n_perm = n_perm,
    fdr_threshold = fdr_threshold,
    zero_tol = zero_tol,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

trait_meta <- traits(trait_ids)
trait_name_map <- stats::setNames(
  as.character(trait_meta$traits$trait_name),
  as.character(trait_meta$traits$id)
)
trait_name_map[is.na(trait_name_map)] <- names(trait_name_map)[is.na(trait_name_map)]

knitr::kable(
  data.frame(
    trait_id = trait_ids,
    trait_name = unname(trait_name_map[trait_ids]),
    stringsAsFactors = FALSE
  ),
  caption = "Target traits compared",
  row.names = FALSE
)
```

## Step 1: Univariate EBMF per trait

The existing single-trait pipeline is run independently for each target trait:
`run_univariate_clustering()` (EBMF) followed by `summarise_ebmf_programs()`.
Programs are kept when they pass the validation filters (`status == "valid"`).
Each program's posterior-mean SNP loadings (`F_pm`), lFSR, and high-confidence
SNP membership are then extracted, with loadings sign-oriented to the target
trait (see `extract_program_loadings()`).

```{r step1-run}
run_trait <- function(tid) {
  obj <- trait(tid, include_associations = TRUE)
  res <- run_univariate_clustering(
    obj,
    target_trait_id = tid,
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    max_snp_fraction = 0.8,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    ebmf_se_mode = ebmf_se_mode,
    ebmf_beta_scale = ebmf_beta_scale,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness
  )
  smry <- summarise_ebmf_programs(
    res,
    s_matrix = res$s_matrix,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    stability_threshold = stability_threshold,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = FALSE
  )
  extracted <- extract_program_loadings(res, smry, valid_only = TRUE)
  list(trait_id = tid, fit = res, summary = smry, extraction = extracted)
}

trait_results <- lapply(trait_ids, run_trait)
names(trait_results) <- trait_ids

for (tid in names(trait_results)) {
  x <- trait_results[[tid]]
  x$extraction$loadings$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_id <- tid
  trait_results[[tid]] <- x
}

program_meta <- dplyr::bind_rows(lapply(trait_results, function(x) {
  x$summary$programs |>
    dplyr::mutate(
      trait_id = x$trait_id,
      trait_name = trait_name_map[[x$trait_id]]
    )
}))
program_info <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$programs))
program_loadings <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$loadings))

knitr::kable(
  program_meta |>
    dplyr::group_by(trait_id, trait_name) |>
    dplyr::summarise(
      n_programs = dplyr::n(),
      n_valid = sum(status == "valid"),
      n_valid_snps = sum(n_snps_filtered[status == "valid"]),
      .groups = "drop"
    ),
  caption = paste0(
    "Per-trait EBMF programs (valid = passes size, internal coherence, ",
    "stability, and redundancy filters; stability is not checked when ",
    "n_rep = 0)"
  ),
  row.names = FALSE
)
```

```{r step1-program-table, results='asis'}
knitr::kable(
  program_meta |>
    dplyr::select(
      trait_name, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, factor_strength, posterior_evidence, status
    ) |>
    dplyr::mutate(
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      factor_strength = round(factor_strength, 3),
      posterior_evidence = round(posterior_evidence, 2)
    ),
  caption = "All EBMF programs discovered per trait, with validation status",
  row.names = FALSE
)
```

## Step 2: Common SNP universe

For every cross-trait program pair, the two loading vectors are aligned on the
**union of their `variant_id`s**; SNPs absent from one program are treated as
zero. Only entries that are effectively zero (`|loading| <= zero_tol`) for
**both** programs are dropped — SNPs that merely fail the lFSR/magnitude gate
are kept, since the correlation is meant to describe the full loading
architecture rather than only the confidently-called SNPs.

```{r step2-universe}
trait_snp_counts <- program_loadings |>
  dplyr::group_by(trait_id, trait_name) |>
  dplyr::summarise(
    n_programs = dplyr::n_distinct(program_id),
    n_snps_any = dplyr::n_distinct(snp_id),
    n_high_conf_snps = dplyr::n_distinct(snp_id[high_confidence]),
    .groups = "drop"
  )

knitr::kable(
  trait_snp_counts,
  caption = "SNP universe per trait (across valid programs)",
  row.names = FALSE
)
```

## Steps 3-4: Primary and supporting similarity

For every cross-trait program pair `compare_program_pairs()` computes:

- **primary** Pearson `r` on the aligned full loading vectors;
- **supporting** Jaccard overlap of the high-confidence SNP sets.

A matched null is generated by permuting one program's loadings across the
aligned SNP slots (preserving its size and loading distribution while breaking
the biological correspondence), giving a two-sided empirical p-value.

```{r step3-compare}
if (nrow(program_loadings) >= 2 && dplyr::n_distinct(program_loadings$trait_id) >= 2) {
  pairs_result <- compare_program_pairs(
    program_loadings,
    n_perm = n_perm,
    fdr_threshold = fdr_threshold,
    zero_tol = zero_tol,
    seed = 1
  )
} else {
  pairs_result <- list(pairs = compare_program_pairs(program_loadings[0, ])$pairs)
}
pairs_tbl <- pairs_result$pairs

n_pairs_total <- nrow(pairs_tbl)
n_pairs_testable <- sum(is.finite(pairs_tbl$r))
knitr::kable(
  data.frame(
    n_programs = dplyr::n_distinct(program_info$program_id),
    n_cross_trait_pairs = n_pairs_total,
    n_testable = n_pairs_testable,
    median_snps_used = if (n_pairs_testable > 0) {
      stats::median(pairs_tbl$n_snps_used[is.finite(pairs_tbl$r)])
    } else {
      NA_real_
    },
    n_significant = sum(pairs_tbl$shared, na.rm = TRUE),
    n_concordant = sum(pairs_tbl$shared & pairs_tbl$direction == "concordant", na.rm = TRUE),
    n_antagonistic = sum(pairs_tbl$shared & pairs_tbl$direction == "antagonistic", na.rm = TRUE),
    stringsAsFactors = FALSE
  ),
  caption = paste0(
    "Cross-trait program pairs compared. n_testable = pairs with a defined ",
    "correlation (at least 3 non-zero aligned SNPs and non-constant loadings). ",
    "Significant = BH FDR <= ", fdr_threshold, "."
  ),
  row.names = FALSE
)
```

## Step 5: Null calibration

The empirical p-values are BH-corrected across all testable cross-trait pairs.
The table below shows the strongest pairs by `|r|`.

```{r step5-null-table}
if (n_pairs_testable > 0) {
  suppressWarnings({
    pairs_tbl |>
      dplyr::filter(is.finite(r)) |>
      dplyr::arrange(dplyr::desc(abs(r))) |>
      utils::head(25) |>
      dplyr::mutate(
        r = round(r, 3),
        jaccard = round(jaccard, 3),
        p_emp = signif(p_emp, 3),
        fdr = signif(fdr, 3),
        shared = dplyr::if_else(shared, "yes", "no")
      ) |>
      knitr::kable(
        caption = "Top cross-trait program pairs by absolute correlation",
        row.names = FALSE
      )
  })
} else {
  cat("No testable cross-trait program pairs.\n")
}
```

```{r step5-null-distribution, fig.height = 4, fig.width = 7}
if (n_pairs_testable > 0) {
  print(
    ggplot2::ggplot(
      pairs_tbl |> dplyr::filter(is.finite(p_emp)),
      ggplot2::aes(x = p_emp)
    ) +
      ggplot2::geom_histogram(bins = 20, fill = "#7570b3", colour = "white") +
      ggplot2::geom_vline(xintercept = fdr_threshold, linetype = "dashed") +
      ggplot2::labs(
        title = "Empirical p-values across cross-trait program pairs",
        subtitle = paste0(
          "dashed line = FDR threshold (", fdr_threshold,
          "); excess near 0 indicates real correspondence"
        ),
        x = "empirical p-value",
        y = "program pairs"
      ) +
      ggplot2::theme_minimal()
  )
} else {
  cat("No empirical p-values to plot.\n")
}
```

## Steps 6-7: Correspondence and direction

A pair is called **shared** when its FDR-controlled null threshold is met; the
strength is kept continuous as signed `r`. `r > 0` is a concordant
(target-aligned) shared program and `r < 0` an antagonistic one. Significance
is based on `|r|`, so negative correlations are only reported when they are
extreme relative to the matched null.

```{r step6-correspondence}
shared_pairs <- pairs_tbl |>
  dplyr::filter(shared) |>
  dplyr::arrange(dplyr::desc(abs(r)))

if (nrow(shared_pairs) > 0) {
  knitr::kable(
    shared_pairs |>
      dplyr::select(
        trait_id_a, program_a, trait_id_b, program_b, r, jaccard,
        n_snps_used, n_high_conf_shared, p_emp, fdr, direction
      ) |>
      dplyr::mutate(
        r = round(r, 3),
        jaccard = round(jaccard, 3),
        p_emp = signif(p_emp, 3),
        fdr = signif(fdr, 3)
      ),
    caption = "Significant cross-trait program correspondences",
    row.names = FALSE
  )
} else {
  cat("No program pairs passed the FDR-controlled correspondence threshold.\n")
}
```

```{r step6-heatmap, fig.height = 8, fig.width = 8}
if (n_pairs_testable > 0) {
  program_order <- program_info |>
    dplyr::arrange(trait_id, program) |>
    dplyr::pull(program_id)
  program_order <- unique(c(program_order, pairs_tbl$program_id_a, pairs_tbl$program_id_b))

  pair_pos <- function(pairs_tbl, flip = FALSE) {
    out <- if (flip) {
      pairs_tbl |>
        dplyr::transmute(
          y = program_id_a, x = program_id_b, r = r, shared = shared
        )
    } else {
      pairs_tbl |>
        dplyr::transmute(
          y = program_id_b, x = program_id_a, r = r, shared = shared
        )
    }
    return(out)
  }
  heat_df <- dplyr::bind_rows(pair_pos(pairs_tbl), pair_pos(pairs_tbl, flip = TRUE))

  print(
    ggplot2::ggplot(
      heat_df,
      ggplot2::aes(
        x = factor(x, levels = program_order),
        y = factor(y, levels = rev(program_order)),
        fill = r
      )
    ) +
      ggplot2::geom_tile(colour = "white", linewidth = 0.2) +
      ggplot2::geom_text(
        data = heat_df |> dplyr::filter(shared),
        ggplot2::aes(label = "*"),
        colour = "black", size = 4, vjust = 0.75, na.rm = TRUE
      ) +
      ggplot2::scale_fill_gradient2(
        low = "#1b9e77", mid = "white", high = "#d95f02",
        midpoint = 0, limits = c(-1, 1), na.value = "grey90"
      ) +
      ggplot2::labs(
        title = "Cross-trait program correspondence",
        subtitle = "Pearson r on aligned F_pm vectors; * = FDR-significant",
        x = "program", y = "program", fill = "r"
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, size = 7),
        axis.text.y = ggplot2::element_text(size = 7),
        panel.grid = ggplot2::element_blank()
      )
  )
} else {
  cat("No cross-trait program pairs to plot.\n")
}
```

## Step 8: Program families

Programs are nodes and significant pairs are undirected, signed edges. The
**weak connected components** are the multi-trait program families: any chain
of significant correspondences links its members into one family, without
forcing one-to-one matching. Isolated programs form size-1 families.

```{r step8-families}
families_result <- build_program_families(
  pairs_tbl,
  program_info = program_info,
  fdr_threshold = fdr_threshold
)
family_nodes <- families_result$nodes
family_metrics <- families_result$families

knitr::kable(
  family_nodes |>
    dplyr::arrange(family, trait_id, program_id),
  caption = "Program family assignment",
  row.names = FALSE
)
```

```{r step8-network, fig.height = 7, fig.width = 9}
if (!is.null(families_result$graph)) {
  graph <- families_result$graph
  node_traits <- family_nodes$trait_id[
    match(igraph::V(graph)$name, family_nodes$program_id)
  ]
  trait_levels <- unique(node_traits)
  trait_pal <- stats::setNames(
    grDevices::hcl.colors(length(trait_levels), "Dark 3"),
    trait_levels
  )
  edge_sign <- igraph::E(graph)$r
  igraph::plot.igraph(
    graph,
    layout = igraph::layout_with_fr(graph),
    vertex.color = unname(trait_pal[node_traits]),
    vertex.size = 13,
    vertex.frame.color = "white",
    vertex.label = igraph::V(graph)$name,
    vertex.label.cex = 0.65,
    vertex.label.color = "black",
    edge.color = ifelse(edge_sign > 0, "#d95f02", "#1b9e77"),
    edge.width = 1 + 3 * abs(edge_sign),
    main = "Multi-trait program families (orange = concordant, green = antagonistic)"
  )
  graphics::legend(
    "topleft",
    legend = trait_levels,
    fill = unname(trait_pal[trait_levels]),
    bty = "n",
    title = "trait"
  )
} else {
  cat("No significant correspondence edges; every program is a singleton family.\n")
}
```

## Step 9: Family metrics

For each family we report how many traits and programs it spans, the number of
significant edges, the mean/median/minimum pairwise `r`, the direction split,
and the SNP-overlap (Jaccard) statistics.

```{r step9-metrics}
if (nrow(family_metrics) > 0) {
  knitr::kable(
    family_metrics |>
      dplyr::mutate(dplyr::across(
        c(mean_r, median_r, min_r, max_r, mean_jaccard, median_jaccard),
        ~ round(.x, 3)
      )),
    caption = "Multi-trait program family metrics",
    row.names = FALSE
  )

  multi_trait_families <- family_metrics |>
    dplyr::filter(n_programs > 1, n_traits > 1)
  knitr::kable(
    data.frame(
      n_families = nrow(family_metrics),
      n_singleton_families = sum(family_metrics$n_programs == 1),
      n_multi_trait_families = nrow(multi_trait_families),
      n_programs_in_multi_trait = sum(multi_trait_families$n_programs),
      n_traits = dplyr::n_distinct(program_info$trait_id),
      stringsAsFactors = FALSE
    ),
    caption = "Family summary",
    row.names = FALSE
  )
} else {
  cat("No program families with significant correspondence edges.\n")
}
```

## Interpretation notes

- **Direction is anchored, not absolute.** EBMF factor signs are not
  identifiable and are arbitrary across independent fits. `extract_program_loadings()`
  reflects each program so its target-trait loading is positive, making a
  positive cross-trait `r` a target-aligned ("concordant") architecture and a
  negative `r` an "antagonistic" one relative to each trait's own effect
  direction. Pairs that do not load on their target fall back to anchoring on
  the largest-|loading| SNP.
- **`variant_id`-level alignment.** Programs from different traits are aligned
  on exact shared variants only. Traits that tag distinct variants at the same
  locus will show low Jaccard overlap even when the underlying locus is shared;
  `r` on the union of loading vectors is the more robust signal in that case.
- **Restricted to valid programs.** Only programs passing the per-trait
  validation filters are compared. Set `n_rep > 0` to include the
  trait-subsampling stability gate (slower); with `n_rep = 0` stability is not
  checked.
- **Supporting, not decisive, Jaccard.** Jaccard measures literal SNP sharing
  and is reported alongside `r`; a pair can have high `r` with modest Jaccard
  when programs share an architecture but only some high-confidence SNPs.
- **FDR is across all cross-trait pairs**, as requested, so it does not account
  for the number of programs per trait beyond the pairs actually formed.
