---
title: "Simulation study of univariate clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  cluster_type: louvain
  louvain_gamma: 1.5
  similarity_threshold: 0.2
  min_module_size: 5
  drivers_per_module: 12
  p_structural_zero: 0.3
  n_null: 10
vignette: >
  %\VignetteIndexEntry{Simulation study of univariate clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

n_reps <- 5
set.seed(1)

clustering_settings <- list(
  cluster_type = params$cluster_type,
  louvain_gamma = params$louvain_gamma,
  similarity_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size
)
```

This vignette stress-tests the univariate SNP clustering pipeline
(`run_univariate_clustering()`) against synthetic data with known ground truth.
`simulate_trait()` plants latent module structure over SNPs (or none at all),
and `evaluate_univariate_simulation()` scores recovery via adjusted Rand index
(ARI), module-count error (`|K_hat - K|`), and how many unstructured background
SNPs get absorbed into modules ("hallucinated" structure).

Scenario coverage:

1. **Null** — no planted structure; does the method hallucinate modules?
2. **Dataset size** — 10 / 50 / 200 / 500 / 900 coloc groups.
3. **Signed effects** — coherent vs flipped vs random driver signs.
4. **Module overlap** — disjoint vs moderate vs strong.
5. **Sparsity** — structural-zero rate inside true module support.

```{r helpers}
run_one <- function(...) {
  dots <- list(...)
  if (is.null(dots$drivers_per_module)) {
    dots$drivers_per_module <- params$drivers_per_module
  }
  if (is.null(dots$p_structural_zero)) {
    dots$p_structural_zero <- params$p_structural_zero
  }
  sim <- do.call(
    simulate_trait,
    c(dots, list(seed = sample.int(.Machine$integer.max, 1)))
  )
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  ev <- evaluate_univariate_simulation(sim, res)
  quality <- res$module_quality
  passes_size <- quality$n_snps >= params$min_module_size
  passes_internal <- is.finite(quality$mean_internal_similarity) &
    quality$mean_internal_similarity >= 0.3
  passes_connectedness <- is.finite(quality$connectedness) &
    quality$connectedness >= 0.5
  data.frame(
    k_planted = ev$k_planted,
    k_hat = ev$k_hat,
    k_error = ev$k_error,
    ari_all = ev$ari_all,
    ari_structured = ev$ari_structured,
    module_recall = ev$mean_module_recall,
    background_absorbed = ev$background_absorbed,
    coverage = ev$coverage,
    n_louvain_modules = nrow(quality),
    n_dropped_size = sum(!passes_size),
    n_dropped_internal = sum(passes_size & !passes_internal),
    n_dropped_connectedness = sum(
      passes_size & passes_internal & !passes_connectedness
    ),
    n_unassigned = sum(!names(res$clusters) %in% names(res$clusters_reliable))
  )
}
run_grid <- function(n_reps, label, ...) {
  dplyr::bind_rows(lapply(seq_len(n_reps), function(i) run_one(...))) |>
    dplyr::mutate(scenario = label)
}
```

## Worked example

```{r example}
sim <- simulate_trait(
  n_coloc_groups = 120,
  K = 3,
  overlap = "moderate",
  n_background_snps = 30,
  drivers_per_module = params$drivers_per_module,
  p_structural_zero = params$p_structural_zero,
  module_annotations = list(
    list(tissues = "Liver", trait_categories = "Metabolic"),
    NULL,
    NULL
  ),
  seed = 42
)
res <- do.call(
  run_univariate_clustering,
  c(list(trait_object = sim$trait_object), clustering_settings)
)
ev <- evaluate_univariate_simulation(sim, res)

knitr::kable(
  data.frame(
    parameter = c(
      "cluster type",
      "Louvain gamma",
      "similarity threshold",
      "minimum module size",
      "drivers per module",
      "P(structural zero)"
    ),
    value = c(
      params$cluster_type,
      params$louvain_gamma,
      params$similarity_threshold,
      params$min_module_size,
      params$drivers_per_module,
      params$p_structural_zero
    )
  ),
  caption = "Clustering settings for this rendered vignette"
)

knitr::kable(
  data.frame(
    metric = c("planted K", "detected K-hat", "K error", "ARI (all SNPs)",
               "ARI (structured only)", "mean module recall",
               "background absorbed", "coverage"),
    value = c(ev$k_planted, ev$k_hat, ev$k_error,
              round(ev$ari_all, 3), round(ev$ari_structured, 3),
              round(ev$mean_module_recall, 3),
              round(ev$background_absorbed, 2), round(ev$coverage, 2))
  ),
  caption = paste(
    "Single-run evaluation (3 planted modules with moderate overlap,",
    "30 background SNPs). Module recall asks whether each planted module is",
    "contained in some predicted program, so it scores merged/overlapping",
    "programs (e.g. EBMF) fairly."
  )
)
```

```{r example-confusion}
knitr::kable(ev$confusion, caption = "Truth vs predicted partition")
```

```{r example-quality}
quality_display <- res$module_quality |>
  dplyr::mutate(
    passes_size = n_snps >= params$min_module_size,
    passes_internal = is.finite(mean_internal_similarity) &
      mean_internal_similarity >= 0.3,
    passes_connectedness = is.finite(connectedness) & connectedness >= 0.5,
    failure_reason = dplyr::case_when(
      reliable ~ "kept",
      !passes_size & !passes_internal & !passes_connectedness ~
        "size; internal; connectedness",
      !passes_size & !passes_internal ~ "size; internal",
      !passes_size & !passes_connectedness ~ "size; connectedness",
      !passes_internal & !passes_connectedness ~ "internal; connectedness",
      !passes_size ~ "size",
      !passes_internal ~ "internal similarity",
      !passes_connectedness ~ "connectedness",
      TRUE ~ "other"
    )
  ) |>
  dplyr::select(
    cluster, n_snps, mean_internal_similarity, connectedness,
    passes_size, passes_internal, passes_connectedness, reliable,
    failure_reason
  ) |>
  dplyr::mutate(
    dplyr::across(
      c(mean_internal_similarity, connectedness),
      ~ round(.x, 3)
    )
  )

knitr::kable(
  quality_display,
  caption = "Worked-example modules before and after reliability filtering"
)

knitr::kable(
  data.frame(
    stage = c(
      "Louvain output",
      "Dropped by minimum size",
      "Dropped by internal similarity after size",
      "Dropped by connectedness after size and internal similarity",
      "Reliable output"
    ),
    n_modules = c(
      nrow(res$module_quality),
      sum(!quality_display$passes_size),
      sum(quality_display$passes_size & !quality_display$passes_internal),
      sum(
        quality_display$passes_size & quality_display$passes_internal &
          !quality_display$passes_connectedness
      ),
      sum(quality_display$reliable)
    )
  ),
  caption = "Where worked-example modules are dropped"
)
```

## Calibrated EBMF reporting layer

The pipeline's hard gates (lFSR + magnitude) are bypassed here: the EBMF
posterior is calibrated against permutation nulls of the worked-example matrix,
and programs are reported only if they retain at least `min_module_size`
FDR-controlled core members. Core composition is scored against planted truth
*including* the shared-boundary SNPs, whose drivers legitimately belong to two
modules.

```{r example-calibration}
res_ebmf <- run_univariate_clustering(
  sim$trait_object,
  cluster_type = "ebmf",
  min_module_size = params$min_module_size
)
cal <- calibrate_ebmf_programs(
  res_ebmf,
  n_null = params$n_null,
  verbose = FALSE
)

knitr::kable(
  cal$programs[cal$programs$n_core > 0, ],
  row.names = FALSE,
  caption = sprintf(
    "EBMF programs after permutation calibration (%d nulls; detected = >= %d FDR-controlled members)",
    params$n_null, params$min_module_size
  )
)

core <- cal$memberships[cal$memberships$core, ]
shared <- sim$ground_truth$multi_module_snps
truth <- sim$ground_truth$module_of_snp
if (nrow(core) > 0) {
  comp <- do.call(rbind, lapply(sort(unique(core$program)), function(pg) {
    ids <- core$snp_id[core$program == pg]
    tv <- truth[ids]
    lab <- ifelse(tv == 0, "background", paste0("mod", tv))
    tab <- table(lab)
    data.frame(
      program = pg,
      n_core = length(ids),
      n_shared_boundary = sum(ids %in% shared),
      truth_mix = paste(names(tab), tab, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }))
  knitr::kable(
    comp,
    row.names = FALSE,
    caption = "Core-member composition per program (primary-label counts; shared-boundary SNPs are legitimate dual members)"
  )
}
```

## Scenario 1: false structure under the null

Independent sparse profiles with no shared latent structure. A well-behaved
method should return few reliable modules and leave most SNPs unassigned; a
large `k_hat` or high `coverage` indicates hallucinated structure.

```{r null-grid}
null_results <- dplyr::bind_rows(
  run_grid(n_reps, "n=50", n_coloc_groups = 50, K = 0, p_active_background = 0.05),
  run_grid(n_reps, "n=200", n_coloc_groups = 200, K = 0, p_active_background = 0.05)
)

null_results |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(
    mean_k_hat = mean(k_hat),
    sd_k_hat = sd(k_hat),
    mean_coverage = mean(coverage),
    mean_louvain_modules = mean(n_louvain_modules),
    mean_dropped_size = mean(n_dropped_size),
    mean_dropped_internal = mean(n_dropped_internal),
    mean_dropped_connectedness = mean(n_dropped_connectedness),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "Null simulations: detected K and coverage")
```

## Scenario 2: dataset size

Coloc-group counts of 10 / 50 / 200 / 500 / 900, crossed with planted K of 0
and 3. Small datasets are watched for hallucinated multi-module output;
large ones for degradation of recovery.

```{r size-grid, fig.width=8, fig.height=4}
sizes <- c(10, 50, 200, 500, 900)
size_results <- dplyr::bind_rows(
  lapply(sizes, function(n) {
    dplyr::bind_rows(
      run_grid(n_reps, sprintf("K=%d", 0), n_coloc_groups = n, K = 0,
               p_active_background = 0.05),
      run_grid(n_reps, sprintf("K=%d", 3), n_coloc_groups = n, K = 3,
               p_active_background = 0.05)
    ) |> dplyr::mutate(n_coloc_groups = n)
  })
)

ggplot(size_results, aes(factor(n_coloc_groups), k_hat, fill = scenario)) +
  geom_boxplot() +
  geom_hline(data = dplyr::filter(size_results, scenario == "K=3"),
             aes(yintercept = 3), linetype = "dashed") +
  labs(x = "coloc groups", y = expression(hat(K)), fill = NULL,
       title = "Detected number of modules vs dataset size") +
  theme_minimal()

ggplot(
  dplyr::filter(size_results, scenario == "K=3"),
  aes(factor(n_coloc_groups), ari_structured)
) +
  geom_boxplot(fill = "#c2571b", alpha = 0.7) +
  labs(x = "coloc groups", y = "ARI (structured SNPs)",
       title = "Module recovery vs dataset size (K = 3)") +
  theme_minimal()
```

## Scenario 3: signed clustering

Same effect magnitude, three sign regimes for drivers within each module:
coherent (all positive), flipped (alternating signs — anti-correlated profiles
share a module), random. If sign information is useful, coherent should beat
flipped/random at recovering the planted partition.

```{r sign-grid, fig.width=6, fig.height=4}
sign_results <- dplyr::bind_rows(
  run_grid(n_reps * 2, "coherent", n_coloc_groups = 150, K = 3,
           sign_pattern = "coherent", noise_sd = 1),
  run_grid(n_reps * 2, "flipped", n_coloc_groups = 150, K = 3,
           sign_pattern = "flipped", noise_sd = 1),
  run_grid(n_reps * 2, "random", n_coloc_groups = 150, K = 3,
           sign_pattern = "random", noise_sd = 1)
)

ggplot(sign_results, aes(scenario, ari_structured, fill = scenario)) +
  geom_boxplot(show.legend = FALSE) +
  scale_x_discrete(limits = c("coherent", "flipped", "random")) +
  labs(x = "driver sign regime", y = "ARI (structured SNPs)",
       title = "Recovery under different driver-sign regimes") +
  theme_minimal()
```

## Scenario 4: module overlap

Disjoint, moderate (~50 shared SNPs between adjacent modules) and strong
(~half the smaller module shared). Shared SNPs carry driver traits from both
modules; their primary truth assignment is their earlier module.

```{r overlap-grid, fig.width=6, fig.height=4}
overlap_results <- dplyr::bind_rows(
  run_grid(n_reps, "disjoint", n_coloc_groups = 300, K = 3, overlap = "disjoint"),
  run_grid(n_reps, "moderate", n_coloc_groups = 300, K = 3, overlap = "moderate"),
  run_grid(n_reps, "strong", n_coloc_groups = 300, K = 3, overlap = "strong")
)

ggplot(overlap_results, aes(scenario, ari_structured, fill = scenario)) +
  geom_boxplot(show.legend = FALSE) +
  scale_x_discrete(limits = c("disjoint", "moderate", "strong")) +
  labs(x = "overlap regime", y = "ARI (structured SNPs)",
       title = "Recovery as modules increasingly overlap") +
  theme_minimal()
```

## Scenario 5: sparsity

Structural zeros inside true module support (`p_structural_zero`) versus
density elsewhere (`p_spurious`, `p_active_background`). Structural zeros are
absent observations, not small effects.

```{r sparsity-grid, fig.width=6, fig.height=4}
sparsity_results <- dplyr::bind_rows(
  run_grid(n_reps, "0.2", n_coloc_groups = 200, K = 3, p_structural_zero = 0.2),
  run_grid(n_reps, "0.5", n_coloc_groups = 200, K = 3, p_structural_zero = 0.5),
  run_grid(n_reps, "0.8", n_coloc_groups = 200, K = 3, p_structural_zero = 0.8)
)

sparsity_results |>
  dplyr::mutate(p_zero = as.numeric(scenario)) |>
  ggplot(aes(p_zero, ari_structured)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.05) +
  scale_x_continuous(breaks = c(0.2, 0.5, 0.8)) +
  labs(x = "P(structural zero in module support)", y = "ARI (structured SNPs)",
       title = "Recovery vs within-module sparsity") +
  theme_minimal()
```

## Notes for interpreting runs

- `ari_structured` measures whether planted module members co-cluster; it is
  blind to background absorption. Read together with `background_absorbed`.
- Under the null (`K = 0`), `k_hat > 0` and nonzero `coverage` quantify
  hallucination; track both across replicates rather than single runs.
- The Louvain resolution (`gamma`) and edge threshold are fixed here; rerun any
  grid with different `run_univariate_clustering()` settings by editing
  `run_one()`.
