---
title: "Investigation of EBMF analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of EBMF analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(flashier)
library(ebnm)
```

This vignette clusters SNPs using **Empirical Bayes Matrix Factorization (EBMF)** via `flashier`. Rather than computing pairwise SNP similarity and then clustering, EBMF discovers latent "programs" directly from the trait x SNP z-score matrix. Each program captures a group of SNPs that share a common pattern of pleiotropic effects, and SNPs can belong to multiple programs (overlapping clusters).

We use **BMI** (trait id `1992`) as the target trait.

## Step 1: Construct the trait x SNP matrix (Y)

**Goal:** Build a flat matrix where each **row** is a background trait and each **column** is a target-trait SNP. Cell values are signed z-scores (`beta / se`). No rows are collapsed, pruned, or filtered based on pathway or gene boundaries.

```{r step1-fetch}
target_trait_id <- 1992

target_trait <- trait(
  target_trait_id,
  include_associations = TRUE
)
cg <- target_trait$coloc_groups
```

### Build the matrix

`build_ebmf_matrix()` constructs the pleiotropy matrix, removes rows and columns that are entirely missing or zero (required by `flashier`), and builds annotation vectors for rows (traits) and columns (SNPs).

```{r step1-ebmf-matrix}
ebmf_data <- build_ebmf_matrix(
  trait_id = target_trait_id,
  coloc_groups = cg,
  snp_key = "variant_id",
  minimum_count_in_network = 2
)

Y <- ebmf_data$x_matrix

list(
  target_trait = target_trait$trait$trait_name,
  n_traits = nrow(Y),
  n_snps = ncol(Y),
  fraction_non_na = round(mean(!is.na(Y)), 3)
)
```

```{r step1-dim}
dim(Y)
```

Each column of `Y` is one SNP's signed association profile across the colocalising phenome. `NA` entries indicate no colocalisation signal for that trait at that SNP's locus.

## Step 2: Inspect the annotation layer

**Goal:** Each row and column carries a concatenated metadata string that encodes its biological profile. These annotations are stored alongside the matrix for downstream interpretation.

### Trait (row) annotations

Format: `Tissue__Gene__EnrichedPathway__TraitCategory`. Traits without a specific annotation receive `"None"` or `"Unannotated"` tokens.

```{r step2-trait-annotations}
head(ebmf_data$trait_annotations, 10)
```

```{r step2-trait-annotation-summary}
annotation_parts <- strsplit(ebmf_data$trait_annotations, "__")
trait_tissues <- vapply(annotation_parts, `[`, character(1), 1)
trait_categories <- vapply(annotation_parts, `[`, character(1), 4)

list(
  n_with_tissue = sum(trait_tissues != "None"),
  n_without_tissue = sum(trait_tissues == "None"),
  top_categories = head(sort(table(trait_categories), decreasing = TRUE), 5)
)
```

### SNP (column) annotations

Format: `ColocGroup{ID}__VariantConsequence` where consequence is `"Genic"` or `"Intergenic"`.

```{r step2-snp-annotations}
head(ebmf_data$snp_annotations, 10)
```

```{r step2-snp-annotation-summary}
snp_parts <- strsplit(ebmf_data$snp_annotations, "__")
snp_consequences <- vapply(snp_parts, `[`, character(1), 2)

table(snp_consequences)
```

## Step 3: Run the EBMF factorization

**Goal:** Decompose the trait x SNP matrix \(Y = LF' + E\) into latent factors using `flashier`. The Kronecker (two-way) residual variance structure estimates per-row and per-column variance parameters, naturally down-weighting noisy singleton traits and intergenic SNPs.

- **L** (traits x K): how each trait loads on each discovered program
- **F** (SNPs x K): how each SNP loads on each program
- **E**: residual noise with Kronecker variance structure \(s_{ij} = a_i \cdot b_j\)

The algorithm greedily adds factors while they improve the variational lower bound (ELBO), then backfits all factors cyclically until convergence.

```{r step3-run-ebmf}
flash_fit <- run_ebmf(
  x_matrix = Y,
  greedy_Kmax = 20L,
  backfit = TRUE,
  var_type = 1L,
  verbose = 1L
)
```

```{r step3-fit-summary}
list(
  n_factors = flash_fit$n_factors,
  elbo = flash_fit$elbo,
  pve = round(flash_fit$pve, 4)
)
```

Each factor represents a discovered "program" -- a latent axis of shared pleiotropy among a subset of SNPs.

### Residual variance structure

The Kronecker variance estimates per-row (trait) and per-column (SNP) residual standard deviations. Traits/SNPs with high residual SD contribute less to the factorization.

```{r step3-residual-sd}
res_sd <- flash_fit$residuals_sd

if (is.matrix(res_sd)) {
  row_sd <- rowMeans(res_sd, na.rm = TRUE)
  col_sd <- colMeans(res_sd, na.rm = TRUE)
} else if (is.numeric(res_sd) && length(res_sd) == 1) {
  row_sd <- rep(res_sd, nrow(Y))
  col_sd <- rep(res_sd, ncol(Y))
} else {
  row_sd <- res_sd
  col_sd <- NULL
}

list(
  row_sd_summary = summary(row_sd),
  col_sd_summary = if (!is.null(col_sd)) summary(col_sd) else "constant"
)
```

## Step 4: Extract overlapping SNP clusters

**Goal:** For each discovered program, select SNPs with high-confidence non-zero loadings. A SNP is assigned to a program if:

1. **lFSR < 0.05** (local false sign rate -- the probability that the sign of the loading is wrong)
2. **|loading| > 0.10** (magnitude threshold to exclude negligibly small effects)

SNPs may belong to multiple programs, producing **overlapping clusters**.

```{r step4-extract-clusters}
ebmf_clusters <- extract_ebmf_clusters(
  flash_fit,
  lfsr_threshold = 0.05,
  magnitude_threshold = 0.10
)

list(
  n_programs = ebmf_clusters$n_programs,
  n_snps_total = ncol(Y),
  n_assigned = ebmf_clusters$n_assigned,
  n_multi_program = ebmf_clusters$n_multi_program,
  n_unassigned = ncol(Y) - ebmf_clusters$n_assigned
)
```

### Program membership summary

```{r step4-membership-summary}
program_sizes <- colSums(ebmf_clusters$membership)
data.frame(
  program = names(program_sizes),
  n_snps = as.integer(program_sizes),
  pve = round(flash_fit$pve, 4)
) |>
  dplyr::arrange(dplyr::desc(n_snps))
```

### Cluster assignments

```{r step4-cluster-table}
ebmf_clusters$clusters |>
  dplyr::group_by(program) |>
  dplyr::summarise(
    n_snps = dplyr::n(),
    mean_loading = round(mean(loading), 3),
    mean_abs_loading = round(mean(abs(loading)), 3),
    mean_lfsr = round(mean(lfsr), 4),
    .groups = "drop"
  )
```

### Multi-program SNPs

SNPs assigned to more than one program represent variants with complex pleiotropic roles spanning multiple biological axes.

```{r step4-multi-program}
multi_snps <- rownames(ebmf_clusters$membership)[rowSums(ebmf_clusters$membership) > 1]

if (length(multi_snps) > 0) {
  ebmf_clusters$clusters |>
    dplyr::filter(snp_id %in% multi_snps) |>
    dplyr::arrange(snp_id, program) |>
    dplyr::mutate(
      loading = round(loading, 3),
      lfsr = round(lfsr, 4)
    )
} else {
  cat("No SNPs assigned to multiple programs.\n")
}
```

## Step 5: Interpret biological programs

**Goal:** Identify which background traits drive each discovered program. For each program, we examine the trait loadings (L matrix) to find traits with the strongest contributions.

```{r step5-trait-loadings}
L_pm <- flash_fit$L_pm
L_lfsr <- flash_fit$L_lfsr
K <- flash_fit$n_factors

trait_ids <- rownames(Y)
trait_info <- ebmf_data$trait_info |>
  dplyr::mutate(trait_id = as.character(trait_id))

program_drivers <- lapply(seq_len(K), function(k) {
  sig <- !is.na(L_lfsr[, k]) & L_lfsr[, k] < 0.05 & abs(L_pm[, k]) > 0.05
  if (sum(sig) == 0) {
    return(NULL)
  }
  result <- data.frame(
    program = k,
    trait_id = trait_ids[sig],
    loading = L_pm[sig, k],
    lfsr = L_lfsr[sig, k],
    stringsAsFactors = FALSE
  ) |>
    dplyr::left_join(trait_info, by = "trait_id") |>
    dplyr::arrange(dplyr::desc(abs(loading)))
  return(result)
}) |>
  dplyr::bind_rows()

if (nrow(program_drivers) > 0) {
  program_drivers |>
    dplyr::slice_head(n = 10) |>
    dplyr::select(program, trait_name, loading, lfsr) |>
    dplyr::mutate(
      loading = round(loading, 3),
      lfsr = round(lfsr, 4)
    )
} else {
  cat("No significant trait drivers found.\n")
}
```

Positive loadings indicate traits that co-vary with the program's SNP factor; negative loadings indicate inverse associations.

### Heatmap of SNP factor loadings

```{r step5-loading-heatmap, fig.height = 7, fig.width = 7}
if (K > 0 && ebmf_clusters$n_assigned > 0) {
  F_pm <- flash_fit$F_pm
  assigned_snps <- rownames(ebmf_clusters$membership)[rowSums(ebmf_clusters$membership) > 0]
  F_assigned <- F_pm[assigned_snps, , drop = FALSE]
  colnames(F_assigned) <- paste0("Program ", seq_len(K))

  heatmap(
    F_assigned,
    scale = "none",
    main = "SNP factor loadings (assigned SNPs)",
    xlab = "Programs",
    ylab = "SNPs",
    cexCol = 0.8
  )
} else {
  cat("No assigned SNPs to plot.\n")
}
```

SNPs cluster along columns by their program membership. Overlapping SNPs (those assigned to multiple programs) will show non-zero loadings in multiple columns.

## Step 6: SNP membership by program

**Goal:** List the SNPs assigned to each EBMF program using readable SNP names (`display_snp`) and the genes linked to each SNP's colocalisation group.

```{r step6-snp-lookup}
snp_lookup <- ebmf_data$snp_info |>
  dplyr::transmute(
    snp_id = as.character(snp_id),
    display_snp = display_snp,
    coloc_group_id = coloc_group_id,
    chr = chr,
    bp = bp
  )

coloc_group_genes <- cg |>
  dplyr::filter(coloc_group_id %in% snp_lookup$coloc_group_id) |>
  dplyr::filter(!is.na(gene), gene != "") |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::summarise(
    genes = paste(sort(unique(gene)), collapse = ", "),
    .groups = "drop"
  )

program_snp_catalogue <- ebmf_clusters$clusters |>
  dplyr::left_join(snp_lookup, by = "snp_id") |>
  dplyr::left_join(coloc_group_genes, by = "coloc_group_id") |>
  dplyr::mutate(
    genes = dplyr::if_else(is.na(genes), "None", genes),
    loading = round(loading, 3),
    lfsr = round(lfsr, 4)
  ) |>
  dplyr::arrange(program, dplyr::desc(abs(loading)))
```

```{r step6-program-summary}
program_snp_catalogue |>
  dplyr::group_by(program) |>
  dplyr::summarise(
    n_snps = dplyr::n(),
    display_snps = paste(display_snp, collapse = ", "),
    genes = paste(unique(unlist(strsplit(genes, ", "))), collapse = ", "),
    .groups = "drop"
  )
```

### SNPs by program

```{r step6-program-tables, results='asis'}
for (prog in sort(unique(program_snp_catalogue$program))) {
  prog_table <- program_snp_catalogue |>
    dplyr::filter(program == prog) |>
    dplyr::select(display_snp, genes, loading, lfsr)

  cat("\n\n#### Program ", prog, "\n\n", sep = "")
  print(knitr::kable(
    prog_table,
    caption = paste0("Program ", prog, " (n = ", nrow(prog_table), " SNPs)")
  ))
}
```

SNPs appearing in more than one program above reflect overlapping EBMF membership (see Step 4). The `genes` column lists all distinct genes associated with that SNP's colocalisation group across the GPMap coloc data at that locus.

## Step 7: Save results

```{r step7-save-results}
ebmf_results <- list(
  flash_fit = flash_fit,
  ebmf_clusters = ebmf_clusters,
  program_drivers = program_drivers
)

results_path <- file.path(pkg_root, "vignettes", "investigation-ebmf-analysis-results.rds")
saveRDS(ebmf_results, results_path)

results_path
```

