---
title: "Investigation of multi-trait EBMF program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of multi-trait EBMF program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "prod"
  include_trans: false
  trait_ids: [1992, 1993]
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  compress_method: "asinh"
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_magnitude_threshold: 0.5
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  min_mean_internal: 0.3
  min_connectedness: 0.5
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 0
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  n_perm: 1000
  fdr_threshold: 0.05
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r trait-metadata, include=FALSE}
trait_ids <- as.character(params$trait_ids)
if (length(trait_ids) < 2) {
  stop("At least two trait_ids are required for a multi-trait comparison")
}
select_api(params$api)
trait_meta <- traits(trait_ids)
trait_name_map <- stats::setNames(
  as.character(trait_meta$traits$trait_name),
  as.character(trait_meta$traits$id)
)
trait_name_map[is.na(trait_name_map)] <- names(trait_name_map)[is.na(trait_name_map)]
```

This vignette extends the single-trait investigation
(`investigation-univariate-analysis.html`) to **multiple traits**. EBMF programs
are discovered independently for each target trait, then cross-trait program
pairs are compared. The primary similarity is the Pearson correlation of the
programs' SNP loading vectors (`F_pm`), aligned on the union of their SNPs; a
supporting Jaccard overlap is taken on each program's high-confidence
(lFSR/magnitude-gated) SNP set. A matched permutation null turns every
correlation into an empirical p-value, which is BH-corrected across all
cross-trait pairs. Credibly corresponding programs are then joined into
**multi-trait program families** (weak connected components of the significant
graph), so no one-to-one matching is forced.

Traits are supplied as an array in `params$trait_ids`. We use
`r paste0("**", unname(trait_name_map[trait_ids]), "** (trait id ", trait_ids, ")", collapse = ", ")`.

```{r params}
api <- params$api
trait_subset <- params$trait_subset
min_module_size <- as.integer(params$min_module_size)
min_snp_signals <- as.integer(params$min_snp_signals)
compress_method <- params$compress_method
compress_scale <- as.numeric(params$compress_scale)
include_trans <- isTRUE(params$include_trans)
similarity_threshold <- as.numeric(params$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(params$ebmf_lfsr_threshold)
ebmf_prior <- params$ebmf_prior
ebmf_backfit <- isTRUE(params$ebmf_backfit)
ebmf_magnitude_threshold <- if (is.na(params$ebmf_magnitude_threshold)) {
  NA_real_
} else {
  as.numeric(params$ebmf_magnitude_threshold)
}
associations <- params$associations
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
stability_threshold <- as.numeric(params$stability_threshold)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)
n_perm <- as.integer(params$n_perm)
fdr_threshold <- as.numeric(params$fdr_threshold)

knitr::kable(
  data.frame(
    api = api,
    trait_ids = paste(trait_ids, collapse = ", "),
    trait_subset = trait_subset,
    min_snp_signals = min_snp_signals,
    min_module_size = min_module_size,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    similarity_threshold = similarity_threshold,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    n_rep = n_rep,
    n_perm = n_perm,
    fdr_threshold = fdr_threshold,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

knitr::kable(
  data.frame(
    trait_id = trait_ids,
    trait_name = unname(trait_name_map[trait_ids]),
    stringsAsFactors = FALSE
  ),
  caption = "Target traits compared",
  row.names = FALSE
)
```

## Step 1: Univariate EBMF per trait

The existing single-trait pipeline is run independently for each target trait:
`run_univariate_clustering()` (EBMF) followed by `summarise_ebmf_programs()`.
Programs are kept when they pass the validation filters (`status == "valid"`).
Each program's posterior-mean SNP loadings (`F_pm`), lFSR, and high-confidence
SNP membership are then extracted, with loadings sign-oriented to the target
trait (see `extract_program_loadings()`).

```{r step1-run}
run_trait <- function(tid) {
  obj <- trait(tid, include_associations = TRUE)
  res <- run_univariate_clustering(
    obj,
    target_trait_id = tid,
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    max_snp_fraction = 0.8,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness
  )
  smry <- summarise_ebmf_programs(
    res,
    s_matrix = res$s_matrix,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    stability_threshold = stability_threshold,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = FALSE
  )
  extracted <- extract_program_loadings(res, smry, valid_only = TRUE)
  list(trait_id = tid, fit = res, summary = smry, extraction = extracted)
}

trait_results <- lapply(trait_ids, run_trait)
names(trait_results) <- trait_ids

for (tid in names(trait_results)) {
  x <- trait_results[[tid]]
  x$extraction$loadings$trait_name <- trait_name_map[[tid]]
  x$extraction$profiles$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_id <- tid
  trait_results[[tid]] <- x
}

program_meta <- dplyr::bind_rows(lapply(trait_results, function(x) {
  x$summary$programs |>
    dplyr::mutate(
      trait_id = x$trait_id,
      trait_name = trait_name_map[[x$trait_id]]
    )
}))
program_info <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$programs))
program_loadings <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$loadings))
program_data <- lapply(trait_results, function(x) x$extraction)

knitr::kable(
  program_meta |>
    dplyr::group_by(trait_id, trait_name) |>
    dplyr::summarise(
      n_programs = dplyr::n(),
      n_valid = sum(status == "valid"),
      n_valid_snps = sum(n_snps_filtered[status == "valid"]),
      .groups = "drop"
    ),
  caption = paste0(
    "Per-trait EBMF programs (valid = passes size, internal coherence, ",
    "stability, and redundancy filters; stability is not checked when ",
    "n_rep = 0)"
  ),
  row.names = FALSE
)
```

```{r step1-program-table, results='asis'}
knitr::kable(
  program_meta |>
    dplyr::select(
      trait_name, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, factor_strength, posterior_evidence, status
    ) |>
    dplyr::mutate(
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      factor_strength = round(factor_strength, 3),
      posterior_evidence = round(posterior_evidence, 2)
    ),
  caption = "All EBMF programs discovered per trait, with validation status",
  row.names = FALSE
)
```

## Step 2: Common locus universe

Programs are compared at the level of the **`coloc_group_id`** (the shared
fine-mapped locus) rather than the exact `variant_id`, because each trait carries
a single candidate SNP per coloc group. The table below counts the candidate
loci per trait and how many coloc groups the traits share.

```{r step2-universe}
trait_locus_counts <- program_loadings |>
  dplyr::group_by(trait_id, trait_name) |>
  dplyr::summarise(
    n_programs = dplyr::n_distinct(program_id),
    n_loci_any = dplyr::n_distinct(coloc_group_id),
    n_high_conf_loci = dplyr::n_distinct(coloc_group_id[high_confidence]),
    .groups = "drop"
  )

shared_loci <- program_loadings |>
  dplyr::distinct(trait_id, coloc_group_id) |>
  dplyr::count(coloc_group_id, name = "n_traits")

knitr::kable(
  trait_locus_counts,
  caption = "Candidate locus (coloc group) universe per trait (across valid programs)",
  row.names = FALSE
)
knitr::kable(
  data.frame(
    n_loci_any = dplyr::n_distinct(program_loadings$coloc_group_id),
    n_loci_shared_by_traits = sum(shared_loci$n_traits > 1),
    max_traits_per_locus = max(shared_loci$n_traits)
  ),
  caption = "Shared coloc-group universe across traits",
  row.names = FALSE
)
```

## Steps 3-4: Locus-matched, calibrated program correspondence

`compare_program_pairs()` works in three stages:

1. **Locus membership** — each program's locus set is its high-confidence
   `coloc_group_id`s; only pairs sharing at least `min_shared_loci` loci become
   candidates.
2. **Profile correlation** — the pleiotropic profiles (`L_pm`) are correlated on
   shared background traits after regressing out each trait's common (mean)
   profile.
3. **Per-program calibration** — for each program the best locus-matched partner
   is compared against a max-statistic permutation null (shuffling feature
   labels), so significance accounts for the search over partners rather than
   being spread across all 300 dependent pairs.

A pair is a `shared` edge only when it is the mutual best match of two
significant programs.

```{r step3-compare}
pairs_result <- compare_program_pairs(
  program_data,
  n_perm = n_perm,
  fdr_threshold = fdr_threshold,
  seed = 1
)
pairs_tbl <- pairs_result$pairs
matches_tbl <- pairs_result$matches

n_pairs_total <- nrow(pairs_tbl)
n_candidates <- sum(pairs_tbl$candidate, na.rm = TRUE)
n_programs_significant <- sum(matches_tbl$significant, na.rm = TRUE)
knitr::kable(
  data.frame(
    n_programs = nrow(matches_tbl),
    n_cross_trait_pairs = n_pairs_total,
    n_locus_matched = n_candidates,
    median_shared_loci = if (n_candidates > 0) {
      stats::median(pairs_tbl$n_loci_shared[pairs_tbl$candidate])
    } else {
      NA_real_
    },
    n_programs_significant = n_programs_significant,
    n_shared_edges = sum(pairs_tbl$shared, na.rm = TRUE),
    n_concordant = sum(pairs_tbl$shared & pairs_tbl$direction == "concordant", na.rm = TRUE),
    n_antagonistic = sum(pairs_tbl$shared & pairs_tbl$direction == "antagonistic", na.rm = TRUE),
    stringsAsFactors = FALSE
  ),
  caption = paste0(
    "Locus-matched pairs and per-program calibrated matches (BH FDR <= ",
    fdr_threshold, ")."
  ),
  row.names = FALSE
)
```

## Step 5: Diagnostics and strongest pairs

Do the profile correlations survive removal of each trait's common profile? PC1
of each trait's program x feature profile matrix summarises the shared global
axis; if it explains most of the variance, residualising it matters.

```{r step5-diagnostic}
pc1_var <- vapply(program_data, function(ex) {
  m <- ex$profiles |>
    dplyr::select(program_id, feature_trait_id, loading) |>
    tidyr::pivot_wider(names_from = feature_trait_id, values_from = loading)
  x <- as.matrix(m[, -1, drop = FALSE])
  x <- x[stats::complete.cases(x), , drop = FALSE]
  if (nrow(x) < 2 || ncol(x) < 2) return(NA_real_)
  s <- svd(scale(x, center = TRUE, scale = FALSE), nu = 0, nv = 0)$d
  if (length(s) < 2 || sum(s^2) == 0) return(NA_real_)
  s[1]^2 / sum(s^2)
}, numeric(1))
knitr::kable(
  data.frame(trait_id = names(pc1_var), pc1_variance = round(pc1_var, 3)),
  caption = "First PC variance of each trait's program profiles",
  row.names = FALSE
)

if (n_candidates > 0) {
  candidates_tbl <- pairs_tbl[pairs_tbl$candidate, , drop = FALSE]
  knitr::kable(
    utils::head(
      candidates_tbl[order(-abs(candidates_tbl$profile_r_resid)), ],
      25
    ) |>
      dplyr::mutate(
        jaccard_loci = round(jaccard_loci, 3),
        profile_r = round(profile_r, 3),
        profile_r_resid = round(profile_r_resid, 3),
        p_profile = signif(p_profile, 3),
        q_profile = signif(q_profile, 3)
      ) |>
      dplyr::select(
        program_id_a, program_id_b, n_loci_shared, jaccard_loci,
        profile_r, profile_r_resid, p_profile, q_profile, direction
      ),
    caption = "Strongest locus-matched pairs by residual profile correlation",
    row.names = FALSE
  )
} else {
  cat("No locus-matched program pairs.\n")
}
```

## Steps 6-7: Correspondence and direction

A pair is **shared** when it is the mutual best match of two significant
programs. `jaccard_loci` measures the overlap of the two programs'
high-confidence loci; `profile_r_resid` measures whether they point the same way
across background traits after de-confounding. `direction` (concordant /
antagonistic) is the sign of the deciding correlation.

```{r step6-correspondence}
shared_pairs <- pairs_tbl |>
  dplyr::filter(shared) |>
  dplyr::arrange(dplyr::desc(abs(profile_r_resid)))

if (nrow(shared_pairs) > 0) {
  knitr::kable(
    shared_pairs |>
      dplyr::select(
        trait_id_a, program_a, trait_id_b, program_b, n_loci_shared,
        jaccard_loci, profile_r, profile_r_resid, n_features_shared,
        p_profile, q_profile, direction
      ) |>
      dplyr::mutate(
        jaccard_loci = round(jaccard_loci, 3),
        profile_r = round(profile_r, 3),
        profile_r_resid = round(profile_r_resid, 3),
        p_profile = signif(p_profile, 3),
        q_profile = signif(q_profile, 3)
      ),
    caption = "Significant cross-trait program correspondences (mutual best matches)",
    row.names = FALSE
  )
} else {
  cat("No program pairs passed the calibrated correspondence threshold.\n")
}
```

```{r step6-heatmap, fig.height = 8, fig.width = 8}
if (n_candidates > 0) {
  program_order <- program_info |>
    dplyr::arrange(trait_id, program) |>
    dplyr::pull(program_id)
  program_order <- unique(c(program_order, pairs_tbl$program_id_a, pairs_tbl$program_id_b))

  pair_pos <- function(pairs_tbl, flip = FALSE) {
    out <- if (flip) {
      pairs_tbl |>
        dplyr::transmute(
          y = program_id_a, x = program_id_b, r = profile_r_resid, shared = shared
        )
    } else {
      pairs_tbl |>
        dplyr::transmute(
          y = program_id_b, x = program_id_a, r = profile_r_resid, shared = shared
        )
    }
    return(out)
  }
  heat_df <- dplyr::bind_rows(pair_pos(pairs_tbl), pair_pos(pairs_tbl, flip = TRUE))

  print(
    ggplot2::ggplot(
      heat_df,
      ggplot2::aes(
        x = factor(x, levels = program_order),
        y = factor(y, levels = rev(program_order)),
        fill = r
      )
    ) +
      ggplot2::geom_tile(colour = "white", linewidth = 0.2) +
      ggplot2::geom_text(
        data = heat_df |> dplyr::filter(shared),
        ggplot2::aes(label = "*"),
        colour = "black", size = 4, vjust = 0.75, na.rm = TRUE
      ) +
      ggplot2::scale_fill_gradient2(
        low = "#1b9e77", mid = "white", high = "#d95f02",
        midpoint = 0, limits = c(-1, 1), na.value = "grey90"
      ) +
      ggplot2::labs(
        title = "Cross-trait program correspondence",
        subtitle = paste0(
          "Residual pleiotropic-profile correlation; ",
          "* = calibrated mutual best match"
        ),
        x = "program", y = "program", fill = "profile_r_resid"
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, size = 7),
        axis.text.y = ggplot2::element_text(size = 7),
        panel.grid = ggplot2::element_blank()
      )
  )
} else {
  cat("No locus-matched program pairs to plot.\n")
}
```

## Step 8: Program families

Programs are nodes and *shared* pairs are undirected edges. Families are formed
from the calibrated **reciprocal best matches**: an edge is kept only when it is
the mutual best partner of two significant programs. This keeps one-to-one
correspondences and avoids the transitive "everything collapses into one giant
family" behaviour of connected components. Isolated programs form size-1
families.

```{r step8-families}
families_result <- build_program_families(
  pairs_result,
  program_info = program_info,
  fdr_threshold = fdr_threshold
)
family_nodes <- families_result$nodes
family_metrics <- families_result$families

knitr::kable(
  family_nodes |>
    dplyr::arrange(family, trait_id, program_id),
  caption = "Program family assignment",
  row.names = FALSE
)
```

```{r step8-network, fig.height = 7, fig.width = 9}
if (!is.null(families_result$graph)) {
  graph <- families_result$graph
  node_traits <- family_nodes$trait_id[
    match(igraph::V(graph)$name, family_nodes$program_id)
  ]
  trait_levels <- unique(node_traits)
  trait_pal <- stats::setNames(
    grDevices::hcl.colors(length(trait_levels), "Dark 3"),
    trait_levels
  )
  edge_sign <- igraph::E(graph)$profile_r
  igraph::plot.igraph(
    graph,
    layout = igraph::layout_with_fr(graph),
    vertex.color = unname(trait_pal[node_traits]),
    vertex.size = 13,
    vertex.frame.color = "white",
    vertex.label = igraph::V(graph)$name,
    vertex.label.cex = 0.65,
    vertex.label.color = "black",
    edge.color = ifelse(edge_sign > 0, "#d95f02", "#1b9e77"),
    edge.width = 1 + 3 * abs(edge_sign),
    main = "Multi-trait program families (orange = concordant, green = antagonistic)"
  )
  graphics::legend(
    "topleft",
    legend = trait_levels,
    fill = unname(trait_pal[trait_levels]),
    bty = "n",
    title = "trait"
  )
} else {
  cat("No significant correspondence edges; every program is a singleton family.\n")
}
```

## Step 9: Family metrics

For each family we report how many traits and programs it spans, the number of
shared edges, the mean/median locus Jaccard, the mean profile correlation, the
direction split, and the total number of shared loci.

```{r step9-metrics}
if (nrow(family_metrics) > 0) {
  knitr::kable(
    family_metrics |>
      dplyr::mutate(dplyr::across(
        c(mean_locus_concordance, median_locus_concordance, mean_profile_r),
        ~ round(.x, 3)
      )),
    caption = "Multi-trait program family metrics",
    row.names = FALSE
  )

  multi_trait_families <- family_metrics |>
    dplyr::filter(n_programs > 1, n_traits > 1)
  knitr::kable(
    data.frame(
      n_families = nrow(family_metrics),
      n_singleton_families = sum(family_metrics$n_programs == 1),
      n_multi_trait_families = nrow(multi_trait_families),
      n_programs_in_multi_trait = sum(multi_trait_families$n_programs),
      n_traits = dplyr::n_distinct(program_info$trait_id),
      stringsAsFactors = FALSE
    ),
    caption = "Family summary",
    row.names = FALSE
  )
} else {
  cat("No program families with significant correspondence edges.\n")
}
```

## Interpretation notes

- **Direction is anchored, not absolute.** EBMF factor signs are not
  identifiable and are arbitrary across independent fits.
  `extract_program_loadings()` reflects both the locus (`F_pm`) and profile
  (`L_pm`) loadings so the target-trait loading is positive, making a positive
  `profile_r_resid` a target-aligned ("concordant") architecture and a negative
  value an "antagonistic" one relative to each trait's own effect direction.
- **Locus-level membership.** Programs are compared on their high-confidence
  `coloc_group_id`s (the fine-mapped signal), not exact candidate `variant_id`s
  and not dense soft weights. This makes membership program-specific: pairs must
  share at least `min_shared_loci` high-confidence loci to be candidates, and
  the shared-locus count/Jaccard are reported. Set
  `locus_membership = "top_k"` to use the top `top_k` loci by weight instead.
- **Profiles are de-confounded.** `profile_r` is the raw correlation of the
  `L_pm` profiles; `profile_r_resid` regresses out each trait's common (mean)
  profile first. The Step 5 PC1 table shows how much global structure exists;
  large PC1 variance means the residual correlation should be preferred.
- **Per-program calibration.** Rather than testing all cross-trait pairs and
  BH-correcting across ~300 dependent tests, each program's best locus-matched
  partner is compared with a max-statistic permutation null that accounts for
  the search over partners. `matches$q` is BH-corrected across programs, and a
  pair is a `shared` edge only when it is the mutual best match of two
  significant programs.
- **`exclude_target_traits`.** By default the two target traits are excluded
  from the shared profile axis, because each is oriented positive in its own fit
  and would otherwise add a trivially concordant point.
- **Restricted to valid programs.** Only programs passing the per-trait
  validation filters are compared. Set `n_rep > 0` to include the
  trait-subsampling stability gate (slower); with `n_rep = 0` stability is not
  checked.
- **Reciprocal best match families.** Families keep only mutual best-match
  edges, so they are one-to-one correspondences rather than transitive chains.
  Set `rule = "components"` in `build_program_families()` for the older
  connected-components behaviour.
- **FDR is across programs** (one test per program), not across the full pair
  table, so it reflects the search over partners.
