---
title: "Investigation of EBMF analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of EBMF analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(flashier)
library(ebnm)
```

This vignette clusters SNPs using **Empirical Bayes Matrix Factorization (EBMF)** via `flashier`. Rather than computing pairwise SNP similarity and then clustering, EBMF discovers latent "programs" directly from the trait x SNP effect matrix. Each program captures a group of SNPs that share a common pattern of pleiotropic effects, and SNPs can belong to multiple programs (overlapping clusters).

We use **BMI** (trait id `1992`) as the target trait.

## Step 1: Construct the trait x SNP matrix (Y)

**Goal:** Build a flat matrix where each **row** is a background trait and each **column** is a target-trait SNP. Cell values are signed GWAS effect sizes (`beta`); matching standard errors are returned in `se_matrix` for use as flashier's `S` argument. No rows are collapsed, pruned, or filtered based on pathway or gene boundaries.

```{r step1-fetch}
target_trait_id <- 1992

target_trait <- trait(
  target_trait_id,
  include_associations = TRUE
)
cg <- target_trait$coloc_groups
```

### Build the matrix

`build_ebmf_matrix()` constructs the pleiotropy matrix, removes rows and columns that are entirely missing or zero (required by `flashier`), and builds annotation vectors for rows (traits) and columns (SNPs).

```{r step1-ebmf-matrix}
ebmf_data <- build_ebmf_matrix(
  trait_id = target_trait_id,
  coloc_groups = cg,
  snp_key = "variant_id",
  minimum_count_in_network = 2
)

Y <- ebmf_data$beta_matrix
S <- ebmf_data$se_matrix

list(
  target_trait = target_trait$trait$trait_name,
  n_traits = nrow(Y),
  n_snps = ncol(Y),
  fraction_non_na = round(mean(!is.na(Y)), 3)
)
```

```{r step1-dim}
dim(Y)
```

Each column of `Y` is one SNP's signed effect profile across the colocalising phenome. `NA` entries indicate no colocalisation signal for that trait at that SNP's locus. `S` holds the corresponding GWAS standard errors.

## Step 2: Inspect the annotation layer

**Goal:** Each row and column carries a concatenated metadata string that encodes its biological profile. These annotations are stored alongside the matrix for downstream interpretation.

### Trait (row) annotations

Row labels depend on `label_scheme` (default: `pathway_gene_tissue`):

- **Trait rows:** `trait_category` only
- **Molecular QTL rows:** `{PathwayOrGene}_Tissue`, `{PathwayOrGene}`, or `Tissue` depending on scheme
- **SNP columns:** `ColocGroup{ID}` only

### SNP (column) annotations

Format: `ColocGroup{ID}`.

## Step 3: Run EBMF comparison grid

**Goal:** Compare EBMF configurations by varying the EBNM prior and residual variance structure. The beta/SE matrices from Step 1 are reused across all runs -- no redundant matrix rebuilding.

Grid:

- **EBNM priors:** `point_normal`, `point_laplace`
- **Variance:** `var_type = 1` (per-row) and `var_type = c(1, 2)` (Kronecker)

```{r step3-comparison}
ebmf_comparison <- run_ebmf_comparison(
  beta_matrix = Y,
  se_matrix = S,
  label_schemes = c(
    "pathway_gene_tissue",
    "pathway_gene",
    "tissue"
  ),
  trait_info = ebmf_data$trait_info,
  ebnm_fns = c("point_normal", "point_laplace"),
  var_types = list(c(1L), c(1L, 2L)),
  greedy_Kmax = 20L,
  backfit = TRUE,
  verbose = 1L
)
```

### Comparison summary

```{r step3-summary}
ebmf_comparison$summary
```

## Step 4: Explore a specific run

Pick a run for detailed analysis. Below we use the first successful run.

```{r step4-pick-run}
successful <- ebmf_comparison$summary |>
  dplyr::filter(status == "ok")
run_id <- successful$run_id[1]
run <- ebmf_comparison$results[[run_id]]
flash_fit <- run$flash_fit
ebmf_clusters <- run$ebmf_clusters
program_drivers <- run$program_drivers

cat("Analysing run:", run_id, "\n")
list(
  n_factors = flash_fit$n_factors,
  elbo = flash_fit$elbo,
  pve = round(flash_fit$pve, 4)
)
```

### Program membership

```{r step4-membership}
program_sizes <- colSums(ebmf_clusters$membership)
data.frame(
  program = names(program_sizes),
  n_snps = as.integer(program_sizes),
  pve = round(flash_fit$pve, 4)
) |>
  dplyr::arrange(dplyr::desc(n_snps))
```

### Cluster summary

```{r step4-cluster-table}
ebmf_clusters$clusters |>
  dplyr::group_by(program) |>
  dplyr::summarise(
    n_snps = dplyr::n(),
    mean_loading = round(mean(loading), 3),
    mean_abs_loading = round(mean(abs(loading)), 3),
    mean_lfsr = round(mean(lfsr), 4),
    .groups = "drop"
  )
```

### Trait drivers per program

```{r step4-drivers}
if (nrow(program_drivers) > 0) {
  program_drivers |>
    dplyr::group_by(program) |>
    dplyr::slice_head(n = 5) |>
    dplyr::select(program, trait_name, loading, lfsr) |>
    dplyr::mutate(
      loading = round(loading, 3),
      lfsr = round(lfsr, 4)
    )
} else {
  cat("No significant trait drivers found.\n")
}
```

### SNP factor loading heatmap

```{r step4-heatmap, fig.height = 7, fig.width = 7}
K <- flash_fit$n_factors
if (K > 0 && ebmf_clusters$n_assigned > 0) {
  F_pm <- flash_fit$F_pm
  assigned_snps <- rownames(ebmf_clusters$membership)[rowSums(ebmf_clusters$membership) > 0]
  F_assigned <- F_pm[assigned_snps, , drop = FALSE]
  colnames(F_assigned) <- paste0("Program ", seq_len(K))

  heatmap(
    F_assigned,
    scale = "none",
    main = paste("SNP factor loadings -", run_id),
    xlab = "Programs",
    ylab = "SNPs",
    cexCol = 0.8
  )
} else {
  cat("No assigned SNPs to plot.\n")
}
```

### SNP catalogue by program

```{r step4-snp-catalogue}
snp_lookup <- ebmf_data$snp_info |>
  dplyr::transmute(
    snp_id = as.character(snp_id),
    display_snp = display_snp,
    coloc_group_id = coloc_group_id,
    chr = chr,
    bp = bp
  )

coloc_group_genes <- cg |>
  dplyr::filter(coloc_group_id %in% snp_lookup$coloc_group_id) |>
  dplyr::filter(!is.na(gene), gene != "") |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::summarise(
    genes = paste(sort(unique(gene)), collapse = ", "),
    .groups = "drop"
  )

program_snp_catalogue <- ebmf_clusters$clusters |>
  dplyr::left_join(snp_lookup, by = "snp_id") |>
  dplyr::left_join(coloc_group_genes, by = "coloc_group_id") |>
  dplyr::mutate(
    genes = dplyr::if_else(is.na(genes), "None", genes),
    loading = round(loading, 3),
    lfsr = round(lfsr, 4)
  ) |>
  dplyr::arrange(program, dplyr::desc(abs(loading)))

program_snp_catalogue |>
  dplyr::group_by(program) |>
  dplyr::summarise(
    n_snps = dplyr::n(),
    display_snps = paste(display_snp, collapse = ", "),
    genes = paste(unique(unlist(strsplit(genes, ", "))), collapse = ", "),
    .groups = "drop"
  )
```

## Step 5: Save results

```{r step5-save}
results_path <- file.path(pkg_root, "vignettes", "investigation-ebmf-analysis-results.rds")
saveRDS(ebmf_comparison, results_path)
results_path
```

Reload later with:

```r
comparison <- readRDS("vignettes/investigation-ebmf-analysis-results.rds")
comparison$summary
comparison$results[["ebnm_point_normal__var_1"]]$ebmf_clusters
```
