---
title: "Suzuki T2D"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Suzuki T2D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  target_guid: "c6d2e20f-6ece-d093-78ee-37534547c466"
  p_threshold: null
  include_situated_gene: false
  min_mean_internal: 0.2
  min_connectedness: 0.5
  ebmf_magnitude_threshold: 0.75
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  min_module_size: 3
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 3
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r suzuki-data, include=FALSE}
suzuki_dir <- system.file("extdata", "suzuki2024", package = "gpmapr")
suzuki_clusters <- utils::read.csv(
  file.path(suzuki_dir, "clusters.csv"),
  stringsAsFactors = FALSE
)
suzuki_profiles <- utils::read.csv(
  file.path(suzuki_dir, "cluster_trait_profiles.csv"),
  stringsAsFactors = FALSE
)
suzuki_trait_map <- utils::read.csv(
  file.path(suzuki_dir, "trait_map.csv"),
  stringsAsFactors = FALSE
)
suzuki_snvs <- utils::read.csv(
  file.path(suzuki_dir, "index_snv_clusters.csv"),
  stringsAsFactors = FALSE
)
cluster_levels <- suzuki_clusters$cluster
```

This vignette asks whether gpmapr's univariate SNP programs recover the
mechanistic clusters of type 2 diabetes (T2D) signals that Suzuki et al. (2024)
defined. The analysis runs on the European-ancestry meta-analysis from that
paper, uploaded to GPMap. The comparison is at the level of biological groups,
not a SNP-by-SNP match. The questions are:

- Do programs show a Suzuki cluster's cardiometabolic traits, judged from
  their module-specific traits and enrichments?
- Do they carry that cluster's canonical genes?
- Which programs pass validation? The program summary cards in Part 4 bring
  the three together.

## Part 1: Suzuki et al. (2024)

> Suzuki K, Hatzikotoulas K, Southam L, *et al.* Genetic drivers of heterogeneity
> in type 2 diabetes pathophysiology. *Nature* 627, 347–357 (2024).
> <https://doi.org/10.1038/s41586-024-07019-6>

The reference tables used below are CC BY 4.0 extracts of the paper's
supplementary material. They are in `inst/extdata/suzuki2024/` and are
regenerated by `data-raw/suzuki2024.R`.

### Study

- **Scale.** The Type 2 Diabetes Global Genomics Initiative meta-analysed 428,452
  T2D cases and 2,107,149 controls across six ancestry groups.
- **Discovery.** Signals come from multi-ancestry meta-regression (MR-MEGA) with
  three axes of genetic variation as covariates.
  - Index SNVs were selected at *P* < 5 × 10⁻⁸ by greedy clumping. A variant
    joined a clump if r² > 0.05 with the index SNV in *any* of the five 1000
    Genomes super-populations, within 5 Mb.
  - This gave **1,289 independent signals at 611 loci**, of which 145 loci are
    new.
- **EUR meta-analysis (the dataset used here).**
  - 36 GWAS: 242,283 cases and 1,569,734 controls (effective N ≈ 751,755,
    60.3% of the total).
  - Fixed-effects inverse-variance meta-analysis in METAL, GRCh37.
  - The largest contributors are the Million Veteran Program (90,107 cases),
    FinnGen (37,002), UK Biobank (19,119), the Estonian Biobank (12,208) and
    deCODE (11,448). 23andMe is not included.
- **Signal sets differ.** The 1,289 signals were discovered in the
  multi-ancestry analysis, so some are driven by non-European data and may be
  weak or absent in the EUR upload.

### How the clusters were built

1. **Inputs.** Each index SNV's associations with 37 cardiometabolic traits were
   taken from the largest available GWAS (Supplementary Table 5) and aligned to
   the T2D risk allele.
2. **Scaling.** Each association was scaled as $Z_{ij} = \beta_{ij} /
   (\sqrt{N_i}\, s_{ij})$, which is a standardised per-allele effect that does
   not depend on the trait GWAS's size.
3. **Clustering.** The SNVs were grouped by k-means with iterative imputation of
   missing associations (ClustImpute). The number of clusters came from the
   majority rule over 27 NbClust indices.
4. **Hard assignment.** Every SNV is assigned to exactly one cluster. This
   differs from earlier soft clustering by Bayesian non-negative matrix
   factorisation (Udler *et al.* 2018), which gives each SNV cluster weights and
   leaves some unassigned.

```{r suzuki-clusters}
knitr::kable(
  suzuki_clusters |>
    dplyr::transmute(
      cluster,
      n_signals,
      new_in_suzuki = dplyr::if_else(previously_reported, "no", "yes"),
      cardiometabolic_profile,
      example_loci,
      insulin_secretion,
      insulin_sensitivity,
      t2d_or = round(exp(t2d_log_or), 3),
      open_chromatin_enrichment
    ),
  caption = paste0(
    "The eight mechanistic clusters (main-text Table 1, Supplementary Table 9, ",
    "Fig. 2). * = trait adjusted for BMI. t2d_or is the inverse-variance ",
    "weighted mean T2D odds ratio of the cluster's index SNVs. Insulin ",
    "secretion/sensitivity give the direction of the T2D risk allele's effect."
  ),
  row.names = FALSE
)
```

```{r suzuki-profile-heatmap, fig.width = 9, fig.height = 9}
trait_order <- suzuki_trait_map$trait_acronym
suzuki_profiles |>
  dplyr::mutate(
    z = pmax(pmin(beta / se, 10), -10),
    cluster = factor(cluster, levels = cluster_levels),
    trait_acronym = factor(trait_acronym, levels = rev(trait_order))
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = cluster, y = trait_acronym, fill = z)) +
  ggplot2::geom_tile(colour = "white") +
  ggplot2::scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    limits = c(-10, 10), name = "z (capped)"
  ) +
  ggplot2::labs(
    title = "Suzuki et al. cluster profiles over the 37 clustering traits",
    subtitle = "Supplementary Table 7; beta/SE of the cluster mean, aligned to the T2D risk allele",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1))
```

### How the paper validated the clusters

- **Physiology.** The clusters form a cline in insulin-related measures
  (HOMA-B, HOMA-IR, clamp-derived sensitivity). It runs from low insulin
  secretion in the two beta-cell clusters to the most extreme insulin
  resistance in the Lipodystrophy cluster.
- **Single-cell open chromatin.** The clusters are enriched for different cell
  types:
  - beta-cell and Residual glycaemic clusters: pancreatic islets;
  - Beta cell −PI: also enterochromaffin cells;
  - Lipodystrophy: adipocytes only;
  - Metabolic syndrome: pericytes and endothelial cells;
  - Obesity: islet non-beta cells, adrenal and brain neurons;
  - Body fat: no enrichment.
- **Partitioned polygenic scores.** After adjusting for the overall T2D score:
  - the Obesity score is associated with higher risk of coronary artery
    disease, peripheral artery disease, end-stage diabetic nephropathy and
    earlier T2D onset;
  - the Beta cell +PI score is associated with *lower* coronary artery disease
    and nephropathy risk.

### Caveats for using the clusters as a benchmark

- **They are one partition, not ground truth.** They come from one algorithm on
  one trait panel.
- **Two clusters act as catch-alls.** Residual glycaemic (389 signals) and Body
  fat (273) are the largest. Body fat has no chromatin enrichment, and the three
  new clusters have weaker T2D effects than the established ones.
- **Liver/lipid metabolism has only 3 signals.** Its chromatin model was
  unstable in the paper, and it is not a realistic target for recovery.
- **Hard assignment forces membership.** SNVs with intermediate profiles still
  get a cluster, whereas gpmapr's EBMF programs are soft and overlapping.

## Part 2: The uploaded GWAS

```{r upload}
select_api(params$api)
target_guid <- params$target_guid
p_threshold <- params$p_threshold
target <- trait(target_guid, include_associations = TRUE)
target_name <- target$trait$name

upload_metadata <- tryCatch(
  jsonlite::fromJSON(target$trait$upload_metadata),
  error = function(e) return(list())
)
own_rows <- target$coloc_groups |>
  dplyr::filter(trait_id == target_guid)
n_signals_at <- function(p) {
  return(dplyr::n_distinct(own_rows$variant_id[own_rows$min_p <= p]))
}

knitr::kable(
  data.frame(
    name = target_name,
    guid = target_guid,
    sample_size = target$trait$sample_size,
    ancestry = target$trait$ancestry,
    reference_build = if (is.null(upload_metadata$reference_build)) {
      NA_character_
    } else {
      upload_metadata$reference_build
    },
    upload_p_threshold = if (is.null(upload_metadata$p_value_threshold)) {
      NA_real_
    } else {
      upload_metadata$p_value_threshold
    },
    status = target$trait$status,
    n_coloc_groups = dplyr::n_distinct(target$coloc_groups$coloc_group_id),
    n_signals_p5e6 = n_signals_at(5e-6),
    n_signals_p5e8 = n_signals_at(5e-8),
    stringsAsFactors = FALSE
  ),
  caption = paste0(
    "Upload metadata. n_signals counts the upload's own fine-mapped signals ",
    "(distinct lead variants) at each threshold. Suzuki et al. used 5e-8."
  ),
  row.names = FALSE
)
```

The analysis below uses signals at
`r if (is.null(p_threshold)) "the upload's own threshold" else format(p_threshold)`.
Re-render with `params = list(p_threshold = 5e-8)` to match the
genome-wide threshold that Suzuki et al. used.

## Part 3: Univariate analysis

The steps are the same as in
[the univariate analysis investigation](investigation-univariate-analysis.html),
except that T2D proxy traits are removed first (see below):

1. Build a pleiotropy matrix of z-scores (`beta / se`) for every background
   trait at each upload SNP, keeping phenotypic traits only and leaving out
   the T2D proxy traits.
2. Drop sparse traits (observed at fewer than `min_snp_signals` SNPs) and
   ubiquitous traits (observed at more than 80% of SNPs).
3. Orient every SNP to the T2D risk allele and compress with asinh.
4. Factorise with EBMF into overlapping programs.
5. Validate the programs.

Shared settings come from `vignettes/_ebmf_settings.yml`. After validation,
each program is annotated as in that vignette:

- module-specific traits;
- trait-category, tissue and pathway enrichment, both hard-cutoff and
  continuous;
- the genes at the program's SNPs, which this vignette adds (shown in the
  program summary cards).

Only the pleiotropy comparison of unique vs shared SNPs is left to the
univariate vignette.

```{r params}
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_specificity <- 1.25

ebmf_settings <- resolve_ebmf_settings(params)

trait_subset <- ebmf_settings$trait_subset
min_module_size <- as.integer(ebmf_settings$min_module_size)
min_snp_signals <- as.integer(ebmf_settings$min_snp_signals)
compress_method <- ebmf_settings$compress_method
compress_scale <- as.numeric(ebmf_settings$compress_scale)
include_trans <- isTRUE(ebmf_settings$include_trans)
similarity_threshold <- as.numeric(ebmf_settings$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(ebmf_settings$ebmf_lfsr_threshold)
ebmf_prior <- ebmf_settings$ebmf_prior
ebmf_backfit <- isTRUE(ebmf_settings$ebmf_backfit)
ebmf_magnitude_threshold <- as.numeric(ebmf_settings$ebmf_magnitude_threshold)
similarity_n_perm <- as.integer(ebmf_settings$similarity_n_perm)
similarity_q_threshold <- as.numeric(ebmf_settings$similarity_q)
strength_q_threshold <- as.numeric(ebmf_settings$strength_q)
ebmf_greedy_Kmax <- as.integer(ebmf_settings$ebmf_greedy_Kmax)
associations <- params$associations
include_situated_gene <- isTRUE(params$include_situated_gene)
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
stability_threshold <- as.numeric(params$stability_threshold)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)

knitr::kable(
  data.frame(
    api = params$api,
    gpmap_url = getOption("gpmap_url"),
    target_guid = target_guid,
    p_threshold = if (is.null(p_threshold)) "upload default" else format(p_threshold),
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    min_module_size = min_module_size,
    similarity_threshold = similarity_threshold,
    similarity_n_perm = similarity_n_perm,
    similarity_q = similarity_q_threshold,
    strength_q = strength_q_threshold,
    n_rep = n_rep,
    frac_traits = frac_traits,
    stability_threshold = stability_threshold,
    redundancy_threshold = redundancy_threshold,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)
```

### T2D proxy traits

GPMap holds several other GWAS of T2D itself, or of traits defined by having
diabetes. At risk-oriented T2D SNPs they all repeat the anchor trait, so they
add no information about mechanism. They can also give EBMF extra weight to
spend on small differences between near-identical rows. They are removed from
the matrix before fitting.

The list was drawn up by searching `all_traits()` and the upload's coloc
partners for diabetes-related names, then pruned by hand. It covers:

- **T2D GWAS:** GPMap's own T2D and the UK Biobank SPA/Firth versions.
- **Diagnosis:** doctor-diagnosed, self-reported and ICD10 E11.9 diabetes.
- **Treatment:** diabetes medication, metformin, gliclazide and insulin.
- **Family history:** diabetes in the mother, father or siblings.
- **Complications:** diabetes-related eye disease.

Kept on purpose:

- **Glycaemic traits** (glucose, HbA1c, insulin and proinsulin), which are
  Suzuki clustering traits.
- **Type 1 diabetes**, a distinct disease.
- **Combined medication fields**, such as "Medication for cholesterol, blood
  pressure or diabetes: Blood pressure medication".

```{r t2d-proxies}
t2d_proxy_traits <- data.frame(
  trait_id = c(
    "114", "2499", "2549",
    "2863", "3106", "4536",
    "1494", "3270", "4422", "3333", "4300", "4403",
    "3428", "3779", "3580",
    "2814"
  ),
  trait_name = c(
    "Type 2 diabetes",
    "Type 2 diabetes (Firth correction)",
    "Type 2 diabetes (SPA correction)",
    "Diabetes diagnosed by doctor",
    "Non-cancer illness code, self-reported: diabetes",
    "Diagnoses - secondary ICD10: E11.9 Without complications",
    "Medication use (drugs used in diabetes)",
    "Treatment/medication code: metformin",
    "Treatment/medication code: gliclazide",
    "Treatment/medication code: insulin product",
    "Medication for cholesterol, blood pressure or diabetes: Insulin",
    "Started insulin within one year diagnosis of diabetes",
    "Illnesses of mother: Diabetes",
    "Illnesses of father: Diabetes",
    "Illnesses of siblings: Diabetes",
    "Eye problems/disorders: Diabetes related eye disease"
  ),
  group = c(
    rep("T2D GWAS", 3), rep("Diagnosis", 3), rep("Treatment", 6),
    rep("Family history", 3), "Complication"
  ),
  stringsAsFactors = FALSE
)

proxy_counts <- target$coloc_groups |>
  dplyr::mutate(trait_id = as.character(trait_id)) |>
  dplyr::filter(trait_id %in% t2d_proxy_traits$trait_id) |>
  dplyr::group_by(trait_id) |>
  dplyr::summarise(n_coloc_groups = dplyr::n_distinct(coloc_group_id), .groups = "drop")

knitr::kable(
  t2d_proxy_traits |>
    dplyr::left_join(proxy_counts, by = "trait_id") |>
    dplyr::mutate(n_coloc_groups = tidyr::replace_na(n_coloc_groups, 0L)),
  caption = paste0(
    "T2D proxy traits removed before building the matrix. n_coloc_groups = ",
    "the upload's coloc groups that each trait shares."
  ),
  row.names = FALSE
)

target$coloc_groups <- target$coloc_groups |>
  dplyr::filter(!as.character(trait_id) %in% t2d_proxy_traits$trait_id)
```

### Fit

```{r fit}
set.seed(20240214)
ebmf_result <- run_univariate_clustering(
  target,
  target_trait_id = target_guid,
  associations = associations,
  p_threshold = p_threshold,
  trait_subset = trait_subset,
  include_trans = include_trans,
  min_snp_signals = min_snp_signals,
  max_snp_fraction = max_snp_fraction,
  compress_method = compress_method,
  compress_scale = compress_scale,
  ebmf_lfsr_threshold = ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = ebmf_magnitude_threshold,
  ebmf_greedy_Kmax = ebmf_greedy_Kmax,
  ebmf_prior = ebmf_prior,
  ebmf_backfit = ebmf_backfit,
  min_module_size = min_module_size,
  min_mean_internal = min_mean_internal,
  min_connectedness = min_connectedness
)
X <- ebmf_result$x_matrix
S <- ebmf_result$s_matrix
cg <- ebmf_result$coloc_groups

knitr::kable(
  data.frame(
    n_traits = nrow(X),
    n_phenotypic = sum(ebmf_result$trait_info$feature_type == "phenotypic", na.rm = TRUE),
    n_traits_dropped = length(ebmf_result$dropped_trait_ids),
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3),
    n_ebmf_factors = ebmf_result$cluster_details$n_programs,
    dropped_global_factors = length(ebmf_result$cluster_details$dropped_global_factors),
    greedy_Kmax_reached = isTRUE(ebmf_result$cluster_details$greedy_Kmax_reached)
  ),
  caption = "Pleiotropy matrix after filtering, and the EBMF fit",
  align = "r"
)
```

### Program validation: reported, not applied

Here the validation checks are **reported but not applied**, except for size.
Every program with at least `min_module_size` member SNPs is annotated below
and gets a summary card in Part 4, where valid programs are listed first.

What each check measures (`summarise_ebmf_programs()`). Member SNPs are those
passing the lFSR/magnitude cut. `S` is the cosine similarity between two SNPs'
oriented, compressed profiles across the background traits, with missing
cells counted as 0 and the T2D row left out (it is positive at every SNP, so
it would add the same similarity to every pair).

- **n_snps_filtered** (size gate): the number of member SNPs.
- **mean_internal_similarity** and **similarity_q** (similarity gate): the
  mean `S` between pairs of member SNPs, compared with random sets of the same
  number of T2D SNPs. **similarity_lift** is the observed mean over the null
  mean; `similarity_q` is the BH-adjusted permutation p. Because the null is
  size-matched, a tight five-SNP program is not penalised for being small.
- **connectedness** (reported): share of member pairs with
  `|S| >= similarity_threshold`.
- **factor_pve** and **strength_q** (reported, not a gate): the share of the
  matrix variance the factor explains, compared with every factor fitted to
  row-permuted copies of the matrix (the `n_null` null fits). It is not a gate
  because variance explained grows with how many SNPs and traits a program
  spans, so it penalises small, specific programs.
- **replication** (stability gate): the mean Jaccard overlap between the
  program's top SNPs and the best-matching factor when 20% of the background
  traits are held out and EBMF is refit. It is computed here for every program.
- **parent_program** (reported): when most of a program's SNPs sit at the loci
  of a larger program, that program is its parent. Sub-programs are kept.
- **posterior_evidence**: median −log10(lFSR) of the member SNPs.

```{r validation}
program_summary <- summarise_ebmf_programs(
  ebmf_result,
  s_matrix = S,
  edge_threshold = similarity_threshold,
  min_module_size = min_module_size,
  min_mean_internal = min_mean_internal,
  min_connectedness = min_connectedness,
  similarity_n_perm = similarity_n_perm,
  similarity_q_threshold = similarity_q_threshold,
  strength_q_threshold = strength_q_threshold,
  n_null = n_null,
  alpha_membership = alpha_membership,
  n_candidate_tier = n_candidate_tier,
  n_rep = n_rep,
  frac_traits = frac_traits,
  top_n = stability_top_n,
  stability_threshold = stability_threshold,
  posterior_stat = posterior_stat,
  posterior_evidence_cap = posterior_evidence_cap,
  redundancy_threshold = redundancy_threshold,
  stability_all_programs = TRUE,
  cores = cores,
  verbose = FALSE
)
# Only the size gate is applied; the other checks are reported, and the cards
# in Part 4 list valid programs first.
analysis_ids <- program_summary$programs$program[program_summary$programs$size_pass %in% TRUE]
gated_ids <- program_summary$programs$program[program_summary$programs$status == "valid"]

knitr::kable(
  program_summary$programs |>
    dplyr::transmute(
      program, n_snps_filtered,
      mean_internal_similarity = round(mean_internal_similarity, 3),
      similarity_lift = round(similarity_lift, 1),
      similarity_q = signif(similarity_q, 2),
      connectedness = round(connectedness, 3),
      factor_pve_pct = round(100 * factor_pve, 2),
      strength_q = signif(strength_q, 2),
      replication = round(replication, 2),
      posterior_evidence = round(posterior_evidence, 1),
      parent_program,
      analysed = dplyr::if_else(size_pass %in% TRUE, "yes", "no (size)"),
      current_gates = status
    ),
  caption = paste0(
    "Every EBMF program with every validation metric, valid programs first, ",
    "then by similarity. analysed = passes the size gate and is annotated ",
    "below; current_gates is what the full set of gates would say ('valid' or ",
    "the checks it fails), reported but not applied."
  ),
  row.names = FALSE
)

clusters_analysis <- program_summary$assigned |>
  dplyr::filter(program %in% analysis_ids) |>
  dplyr::transmute(snp_id = as.character(snp_id), group = as.character(program))

knitr::kable(
  data.frame(
    n_programs = nrow(program_summary$programs),
    n_passing_size = length(analysis_ids),
    n_passing_current_gates = length(gated_ids),
    n_snps = ncol(X),
    n_snps_in_analysed_programs = dplyr::n_distinct(clusters_analysis$snp_id)
  ),
  caption = "Validation summary. Programs passing size are analysed; the current gates are reported only.",
  row.names = FALSE
)
```

### Module-specific traits

For each analysed program, the background traits that are most elevated in the
program relative to the rest of the T2D SNPs
(`specificity = mean |z| in program / (mean |z| elsewhere + 0.1)`).

```{r module-traits, results = "asis"}
group_traits <- summarise_module_specific_traits(
  trait_matrix = ebmf_result$trait_matrix,
  groups = clusters_analysis,
  trait_info = ebmf_result$trait_info,
  coloc_groups = cg,
  exclude_trait_ids = target_guid,
  min_group_size = min_module_size,
  top_n = top_n_traits,
  n_categories = 3L,
  min_specificity = min_specificity
)

module_traits_overview <- dplyr::bind_rows(lapply(group_traits$by_group, function(x) {
  passing <- x$traits |>
    dplyr::filter(passes_min_specificity) |>
    utils::head(8L)
  return(data.frame(
    program = as.character(x$group),
    n_snps = sum(clusters_analysis$group == as.character(x$group)),
    module_specific_traits = if (nrow(passing) == 0) {
      "none"
    } else {
      paste0(
        passing$trait_name, " (", round(passing$specificity, 2), ")",
        collapse = "<br>"
      )
    },
    stringsAsFactors = FALSE
  ))
}))

if (nrow(module_traits_overview) == 0) {
  cat("No programs to summarise.\n")
} else {
  print(knitr::kable(
    module_traits_overview |> dplyr::arrange(dplyr::desc(n_snps)),
    caption = "Top module-specific traits per program (specificity in brackets)",
    row.names = FALSE,
    escape = FALSE
  ))
}
```

### Trait-category, tissue and pathway enrichment

Each program's member SNPs are tested against the rest of the T2D SNPs (the
trait baseline), exactly as in the univariate investigation:

- **trait categories and tissues:** hypergeometric test, BH-adjusted, with a
  bootstrap CI on the fold enrichment;
- **pathways:** KEGG and Reactome, on the colocalising QTL genes at the
  program's SNPs.

The table lists every hit at FDR ≤ 0.05.

**Tissues differ from Suzuki's.** GPMap's tissues are GTEx-style bulk tissues.
"Pancreas" is bulk GTEx pancreas, mostly exocrine, and there are no islets.
Suzuki's islet-cell enrichments therefore have no direct counterpart here.

```{r enrichment, results = "asis"}
baseline_snp_ids <- colnames(S)
has_analysis_groups <- nrow(clusters_analysis) > 0

.format_enriched_all <- function(by_group_list, value_col) {
  return(vapply(by_group_list, function(x) {
    if (is.null(x$enriched) || nrow(x$enriched) == 0) {
      return("none")
    }
    hits <- x$enriched |>
      dplyr::filter(is.finite(fold_enrichment), is.finite(fdr), fdr <= 0.05) |>
      dplyr::arrange(fdr, dplyr::desc(fold_enrichment))
    if (nrow(hits) == 0) {
      return("none")
    }
    return(paste0(
      hits[[value_col]],
      " (FE=", signif(hits$fold_enrichment, 2),
      ", FDR=", signif(hits$fdr, 3), ")",
      collapse = "<br>"
    ))
  }, character(1)))
}
.format_pathways_all <- function(by_group_list, source) {
  return(vapply(by_group_list, function(x) {
    if (is.null(x$pathways) || nrow(x$pathways) == 0) {
      return("none")
    }
    hits <- x$pathways[
      x$pathways$source == source & !is.na(x$pathways$fdr) & x$pathways$fdr <= 0.05, ,
      drop = FALSE
    ]
    if (nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[order(hits$fdr), , drop = FALSE]
    return(paste0(hits$description, " (FDR=", signif(hits$fdr, 3), ")", collapse = "<br>"))
  }, character(1)))
}

if (!has_analysis_groups) {
  cat("No programs to annotate.\n")
  group_categories <- NULL
  group_tissues <- NULL
  module_enrichment <- NULL
} else {
  group_categories <- enrich_snp_group_trait_categories(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    baseline_snp_ids = baseline_snp_ids
  )
  group_tissues <- enrich_snp_group_tissues(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    background = "trait",
    baseline_snp_ids = baseline_snp_ids
  )
  module_enrichment <- enrich_snp_group_pathways(
    groups = clusters_analysis,
    coloc_groups = cg,
    min_group_size = min_module_size,
    sources = c("KEGG", "Reactome"),
    include_situated_gene = include_situated_gene
  )
  group_ids <- function(res) {
    return(vapply(res$by_group, function(x) as.character(x$group), character(1)))
  }
  hard_overview <- dplyr::full_join(
    data.frame(
      program = group_ids(group_categories),
      trait_categories = .format_enriched_all(group_categories$by_group, "trait_category"),
      stringsAsFactors = FALSE
    ),
    data.frame(
      program = group_ids(group_tissues),
      tissues = .format_enriched_all(group_tissues$by_group, "tissue"),
      stringsAsFactors = FALSE
    ),
    by = "program"
  ) |>
    dplyr::full_join(
      data.frame(
        program = group_ids(module_enrichment),
        reactome = .format_pathways_all(module_enrichment$by_group, "Reactome"),
        kegg = .format_pathways_all(module_enrichment$by_group, "KEGG"),
        stringsAsFactors = FALSE
      ),
      by = "program"
    ) |>
    dplyr::mutate(dplyr::across(-program, function(x) return(tidyr::replace_na(x, "none")))) |>
    dplyr::left_join(
      dplyr::count(clusters_analysis, program = group, name = "n_snps"),
      by = "program"
    ) |>
    dplyr::relocate(n_snps, .after = program) |>
    dplyr::arrange(dplyr::desc(n_snps))
  print(knitr::kable(
    hard_overview,
    caption = paste0(
      "Hard-cutoff enrichment of each program's member SNPs against the ",
      "rest of the T2D SNPs: every hit at FDR <= 0.05 (FE = fold enrichment)."
    ),
    row.names = FALSE,
    escape = FALSE
  ))
}
```

### Continuous loading-based enrichment

The same annotations without the membership cutoff. Every SNP contributes to
every program, weighted by its squared EBMF loading, in the regression
`lm(loading^2 ~ membership)`. Hits need both FDR ≤ 0.05 and R² ≥ 0.05. The
FDR is computed across every program tested, separately for categories,
tissues and pathways.

```{r continuous-enrichment, results = "asis"}
min_r_squared <- 0.05
.format_continuous_all <- function(by_program_list, value_col, source = NULL) {
  return(vapply(by_program_list, function(x) {
    hits <- x$comparison
    if (is.null(hits) || nrow(hits) == 0) {
      return("none")
    }
    keep <- is.finite(hits$fdr) & hits$fdr <= 0.05 &
      is.finite(hits$r_squared) & hits$r_squared >= min_r_squared
    if (!is.null(source)) {
      keep <- keep & hits$source == source
    }
    hits <- hits[keep, , drop = FALSE]
    if (nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[order(-hits$enrichment), , drop = FALSE]
    return(paste0(
      hits[[value_col]],
      " (effect=", signif(hits$enrichment, 2),
      ", R2=", signif(hits$r_squared, 2),
      ", FDR=", signif(hits$fdr, 3), ")",
      collapse = "<br>"
    ))
  }, character(1)))
}

if (!has_analysis_groups) {
  cat("No programs to annotate.\n")
  loading_tissues <- NULL
} else {
  loading_categories <- enrich_program_loadings_trait_categories(
    clustering_result = ebmf_result,
    coloc_groups = cg,
    snp_key = "variant_id"
  )
  loading_tissues <- enrich_program_loadings_tissues(
    clustering_result = ebmf_result,
    coloc_groups = cg,
    snp_key = "variant_id"
  )
  pathway_universe <- tryCatch(
    pathway_mappings(),
    error = function(e) {
      cat("Pathway mappings endpoint unavailable (", conditionMessage(e), ").\n", sep = "")
      return(NULL)
    }
  )
  loading_pathways <- if (is.null(pathway_universe)) {
    NULL
  } else {
    enrich_program_loadings_pathways(
      clustering_result = ebmf_result,
      coloc_groups = cg,
      mappings = pathway_universe,
      snp_key = "variant_id",
      sources = c("KEGG", "Reactome"),
      include_situated_gene = include_situated_gene
    )
  }
  valid_by_program <- function(res) {
    if (is.null(res)) {
      return(list())
    }
    out <- Filter(
      function(x) return(as.character(x$program) %in% as.character(analysis_ids)),
      res$by_program
    )
    return(stats::setNames(out, vapply(out, function(x) as.character(x$program), character(1))))
  }
  cat_by <- valid_by_program(loading_categories)
  tis_by <- valid_by_program(loading_tissues)
  path_by <- valid_by_program(loading_pathways)
  programs_shown <- names(cat_by)
  continuous_overview <- data.frame(
    program = programs_shown,
    trait_categories = .format_continuous_all(cat_by, "trait_category"),
    tissues = .format_continuous_all(tis_by[programs_shown], "tissue"),
    reactome = if (length(path_by) == 0) {
      "not available"
    } else {
      .format_continuous_all(path_by[programs_shown], "description", "Reactome")
    },
    kegg = if (length(path_by) == 0) {
      "not available"
    } else {
      .format_continuous_all(path_by[programs_shown], "description", "KEGG")
    },
    stringsAsFactors = FALSE
  )
  print(knitr::kable(
    continuous_overview,
    caption = paste0(
      "Continuous loading-based enrichment for the analysed programs ",
      "(FDR <= 0.05 and R2 >= ", min_r_squared, ")."
    ),
    row.names = FALSE,
    escape = FALSE
  ))
}
```

```{r program-gene-lookup, include = FALSE}
# Genes at each analysed program's member SNPs, used by the marker-gene and
# pathway checks and the program cards in Part 4.
#
# - Positional: the nearest protein-coding gene to each SNP (0 bp if the SNP
#   falls inside the gene; all overlapping genes are listed). This is how
#   Suzuki et al. name their loci.
# - Colocalising QTL: genes whose molecular QTL colocalises with the SNP
#   (genes_at_snps()). Closer to mechanism, but can differ from the locus name
#   (the FTO signal acts through IRX3/IRX5).
gene_table <- all_genes() |>
  dplyr::filter(gene_biotype == "protein_coding") |>
  dplyr::transmute(gene_id = as.integer(id), gene, chr = as.character(chr), start, stop)

snp_positions <- cg |>
  dplyr::filter(as.character(variant_id) %in% colnames(X)) |>
  dplyr::distinct(variant_id, chr, bp) |>
  dplyr::transmute(snp_id = as.character(variant_id), chr = as.character(chr), bp) |>
  dplyr::distinct(snp_id, .keep_all = TRUE)

# Nearest protein-coding gene(s) to every T2D SNP: distance 0 inside a gene.
positional_links <- dplyr::bind_rows(lapply(split(snp_positions, snp_positions$chr), function(snps) {
  genes_chr <- gene_table[gene_table$chr == snps$chr[1], , drop = FALSE]
  if (nrow(genes_chr) == 0) {
    return(NULL)
  }
  return(dplyr::bind_rows(lapply(seq_len(nrow(snps)), function(i) {
    d <- pmax(genes_chr$start - snps$bp[i], snps$bp[i] - genes_chr$stop, 0)
    hit <- which(d == min(d))
    return(data.frame(
      snp_id = snps$snp_id[i], gene_id = genes_chr$gene_id[hit],
      gene = genes_chr$gene[hit], distance = d[hit], stringsAsFactors = FALSE
    ))
  })))
}))
qtl_links <- genes_at_snps(colnames(X), coloc_groups = cg, snp_key = "variant_id") |>
  dplyr::transmute(snp_id = as.character(snp_id), gene_id = as.integer(gene_id), gene) |>
  dplyr::distinct()

# Every gene Suzuki et al. name at an index SNV, with its cluster(s).
suzuki_gene_clusters <- dplyr::bind_rows(
  suzuki_snvs |>
    dplyr::select(cluster, gene = locus) |>
    tidyr::separate_rows(gene, sep = ",\\s*"),
  suzuki_clusters |>
    dplyr::select(cluster, gene = example_loci) |>
    tidyr::separate_rows(gene, sep = "; ") |>
    tidyr::separate_rows(gene, sep = "-")
) |>
  dplyr::distinct() |>
  dplyr::group_by(gene) |>
  dplyr::summarise(
    suzuki_clusters = paste(cluster_levels[cluster_levels %in% cluster], collapse = ", "),
    .groups = "drop"
  )
suzuki_genes <- suzuki_gene_clusters$gene

format_genes <- function(genes, max_n = 15L) {
  genes <- unique(genes)
  if (length(genes) == 0) {
    return("none")
  }
  shown <- utils::head(genes, max_n)
  shown <- ifelse(shown %in% suzuki_genes, paste0("<b>", shown, "</b>"), shown)
  more <- length(genes) - length(shown)
  return(paste0(
    paste(shown, collapse = ", "),
    if (more > 0) paste0(" (+", more, " more)") else ""
  ))
}

# Member SNPs of each analysed program, strongest absolute loading first.
F_pm <- ebmf_result$cluster_details$flash_fit$F_pm
program_members <- clusters_analysis |>
  dplyr::rename(program = group) |>
  dplyr::filter(snp_id %in% rownames(F_pm)) |>
  dplyr::mutate(
    abs_loading = abs(F_pm[cbind(match(snp_id, rownames(F_pm)), as.integer(program))])
  ) |>
  dplyr::arrange(program, dplyr::desc(abs_loading))

genes_in_order <- function(members, links) {
  ordered <- members |>
    dplyr::inner_join(links, by = "snp_id", relationship = "many-to-many") |>
    dplyr::group_by(gene) |>
    dplyr::summarise(n_snps = dplyr::n_distinct(snp_id), top = max(abs_loading), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(top), dplyr::desc(n_snps))
  return(ordered$gene)
}
program_genes <- dplyr::bind_rows(lapply(split(program_members, program_members$program), function(m) {
  positional <- genes_in_order(m, positional_links)
  qtl <- genes_in_order(m, qtl_links)
  return(data.frame(
    program = m$program[1],
    n_snps = nrow(m),
    n_positional_genes = length(positional),
    positional_genes = format_genes(positional),
    n_qtl_genes = length(qtl),
    qtl_genes = format_genes(qtl),
    stringsAsFactors = FALSE
  ))
}))

# Member SNPs of each analysed program, as a named list.
program_snps <- split(clusters_analysis$snp_id, clusters_analysis$group)
```

## Part 4: Do the programs recapture Suzuki's clusters?

### Expectations (written before the results)

- **Likely to be recovered:** Obesity, Lipodystrophy and Metabolic syndrome.
  Their defining traits (BMI, WHR, body fat %, TG, HDL, blood pressure) are
  large UK Biobank GWAS with hundreds of coloc groups in GPMap.
- **Not expected: the Beta cell +PI / −PI split.** The two clusters differ
  mainly in the sign of the proinsulin effect. GPMap's proinsulin GWAS has
  N = 10,701 and 12 coloc groups, so a joint beta-cell program would be the
  realistic outcome.
- **Glycaemic clusters lean on HbA1c.** Fasting glucose, fasting insulin and
  2-hour glucose are sparse in GPMap, so HbA1c carries most of the glycaemic
  signal.
- **Body fat vs Obesity may merge.** They differ mainly on abdominal
  subcutaneous and visceral fat, which have only 2–7 coloc groups in GPMap.
- **Liver/lipid metabolism (3 signals) is not expected.**
- **Tissues.** A Lipodystrophy-like program should be enriched in adipose, an
  Obesity-like program possibly in brain or hypothalamus (Suzuki found brain
  neurons), and a liver program in liver. Beta-cell programs are **not**
  expected to show pancreas: GPMap has no islet QTLs.
- **Pre-specified pathways (4c).** KEGG *Maturity onset diabetes of the young*,
  KEGG *Insulin secretion*, Reactome *Regulation of gene expression in beta
  cells* and Reactome *Regulation of insulin secretion* should point at the
  beta-cell programs. KEGG *PPAR signaling*, *Adipocytokine signaling* and
  *Insulin signaling* should point at Lipodystrophy or Metabolic syndrome
  programs. KEGG *Non-alcoholic fatty liver disease* should point at
  Liver/lipid.

The comparison has three parts, after a check of which Suzuki traits the
matrix contains:

- **4b:** canonical marker genes, found positionally and through colocalising
  QTLs.
- **4c:** the pre-specified pathway checks.
- **4d:** one summary card per program, with its validation status, bringing
  everything together. Judge a program's match to a Suzuki cluster from its
  card: its specific traits, tissues, pathways and Suzuki loci.

### 4a. Trait coverage

The univariate matrix already contains every phenotypic GPMap trait that
colocalises with the upload's SNPs. `trait_map.csv` records which of those rows
are Suzuki's 37 clustering traits:

- `same_gwas`: the GWAS Suzuki used;
- `proxy`: a different GWAS of the same trait;
- `unadjusted_proxy`: Suzuki used a BMI-adjusted trait and GPMap has it
  unadjusted;
- `none`: no GPMap equivalent.

Nothing extra is fetched or fitted for these traits.

```{r coverage}
suzuki_trait_map <- suzuki_trait_map |>
  dplyr::mutate(
    gpmap_trait_id = as.character(gpmap_trait_id),
    in_matrix = !is.na(gpmap_trait_id) & gpmap_trait_id %in% rownames(X),
    n_snps_observed = dplyr::if_else(
      in_matrix,
      as.integer(rowSums(!is.na(X))[gpmap_trait_id]),
      0L
    )
  )

knitr::kable(
  suzuki_trait_map |>
    dplyr::group_by(match_type) |>
    dplyr::summarise(
      n_traits = dplyr::n(),
      n_in_matrix = sum(in_matrix),
      .groups = "drop"
    ),
  caption = "Suzuki traits available as rows of the pleiotropy matrix, by match type",
  row.names = FALSE
)

knitr::kable(
  suzuki_trait_map |>
    dplyr::select(
      trait_category, trait_acronym, gpmap_trait_id, gpmap_trait_name,
      match_type, in_matrix, n_snps_observed
    ),
  caption = paste0(
    "Suzuki clustering traits and their GPMap rows. n_snps_observed = upload ",
    "SNPs at which the GPMap trait colocalises."
  ),
  row.names = FALSE
)
```

### 4b. Canonical marker genes

The example loci that Suzuki et al. give for each cluster (Table 1) are
checked against the genes at each program's member SNPs, in two ways:

- **positional:** the gene is the nearest protein-coding gene to one of the
  program's SNPs, which is how Suzuki named the locus;
- **QTL:** the gene's molecular QTL colocalises with one of the program's SNPs.

A gene found neither way is labelled "at a T2D SNP, not in a program" if
some upload SNP carries it, and "not found" otherwise. Compare the programs
listed for a cluster's markers with those programs' cards in 4d.

```{r marker-genes}
marker_genes <- suzuki_clusters |>
  dplyr::select(cluster, example_loci) |>
  tidyr::separate_rows(example_loci, sep = "; ") |>
  tidyr::separate_rows(example_loci, sep = "-") |>
  dplyr::rename(gene = example_loci) |>
  dplyr::distinct()

# Genes at each program's SNPs, from both sources.
program_gene_links <- function(links) {
  return(dplyr::bind_rows(lapply(names(program_snps), function(k) {
    genes <- links[links$snp_id %in% program_snps[[k]], , drop = FALSE]
    if (nrow(genes) == 0) {
      return(NULL)
    }
    return(dplyr::distinct(
      data.frame(program = k, gene_id = genes$gene_id, gene = genes$gene, stringsAsFactors = FALSE)
    ))
  })))
}
genes_by_program <- list(positional = program_gene_links(positional_links), QTL = program_gene_links(qtl_links))
trait_genes <- list(positional = unique(positional_links$gene), QTL = unique(qtl_links$gene))

programs_with <- function(g, source) {
  tab <- genes_by_program[[source]]
  if (is.null(tab) || nrow(tab) == 0) {
    return(character(0))
  }
  found <- unique(tab$program[tab$gene == g])
  return(paste0("P", found[order(as.integer(found))]))
}
marker_status <- marker_genes |>
  dplyr::rowwise() |>
  dplyr::mutate(
    positional = paste(programs_with(gene, "positional"), collapse = ", "),
    qtl = paste(programs_with(gene, "QTL"), collapse = ", "),
    status = dplyr::case_when(
      positional != "" | qtl != "" ~ "in a program",
      gene %in% c(trait_genes$positional, trait_genes$QTL) ~ "at a T2D SNP, not in a program",
      TRUE ~ "not found"
    )
  ) |>
  dplyr::ungroup()

knitr::kable(
  marker_status |>
    dplyr::mutate(cluster = factor(cluster, levels = cluster_levels)) |>
    dplyr::arrange(cluster) |>
    dplyr::select(cluster, gene, positional, qtl, status),
  caption = paste0(
    "Suzuki Table 1 example-locus genes. positional / qtl list the programs ",
    "in which the gene is the nearest gene to a member SNP, or a colocalising ",
    "QTL gene."
  ),
  row.names = FALSE
)

knitr::kable(
  marker_status |>
    dplyr::group_by(cluster) |>
    dplyr::summarise(
      n_markers = dplyr::n(),
      n_at_t2d_snps = sum(status != "not found"),
      n_in_any_program = sum(status == "in a program"),
      .groups = "drop"
    ) |>
    dplyr::mutate(cluster = factor(cluster, levels = cluster_levels)) |>
    dplyr::arrange(cluster),
  caption = "Marker-gene summary per Suzuki cluster",
  row.names = FALSE
)
```

### 4c. Pre-specified pathway checks

The open pathway search in Part 3 tests hundreds of pathways on small
programs. It has little power, and a top hit there is easy to over-read. The
pathways below were chosen before looking at the results, from what each
Suzuki cluster is known to involve.

- **Genes:** each program's positional and QTL genes together.
- **Background:** the genes at all T2D SNPs, so a hit means the program is
  enriched *relative to the other T2D loci*, not relative to the genome.
- **Test:** hypergeometric test for each program × pathway,
  Bonferroni-adjusted over the whole grid.
- **Reading it:** the question is whether the best program for each pathway
  looks like the expected cluster on its card (4d).

```{r prespecified-pathways, results = "asis"}
prespecified <- data.frame(
  term_id = c(
    "hsa04950", "hsa04911", "HSA-210745", "HSA-422356",
    "hsa03320", "hsa04920", "hsa04910", "hsa04932", "hsa04930"
  ),
  expected_clusters = c(
    "Beta cell +PI / -PI, Residual glycaemic", "Beta cell +PI / -PI",
    "Beta cell +PI / -PI", "Beta cell +PI / -PI",
    "Lipodystrophy, Metabolic syndrome", "Lipodystrophy, Metabolic syndrome",
    "Lipodystrophy, Metabolic syndrome", "Liver/lipid metabolism",
    "any (general T2D)"
  ),
  stringsAsFactors = FALSE
)
pathway_universe <- if (exists("pathway_universe") && !is.null(pathway_universe)) {
  pathway_universe
} else {
  tryCatch(pathway_mappings(), error = function(e) return(NULL))
}
if (is.null(pathway_universe)) {
  cat("Pathway mappings endpoint unavailable; skipping.\n")
  prespecified_grid <- data.frame()
} else {
  prespecified <- prespecified |>
    dplyr::left_join(
      pathway_universe$sizes |> dplyr::distinct(term_id, source, description),
      by = "term_id"
    )
  pathway_genes <- split(
    as.integer(pathway_universe$mappings$gene_id),
    pathway_universe$mappings$term_id
  )
  in_database <- unique(as.integer(pathway_universe$mappings$gene_id))
  all_links <- dplyr::bind_rows(positional_links, qtl_links)
  universe_ids <- intersect(unique(as.integer(all_links$gene_id)), in_database)
  gene_name <- stats::setNames(all_links$gene, all_links$gene_id)
  prespecified_grid <- dplyr::bind_rows(lapply(names(program_snps), function(k) {
    program_ids <- intersect(
      unique(as.integer(all_links$gene_id[all_links$snp_id %in% program_snps[[k]]])),
      universe_ids
    )
    return(dplyr::bind_rows(lapply(seq_len(nrow(prespecified)), function(i) {
      path_ids <- intersect(pathway_genes[[prespecified$term_id[i]]], universe_ids)
      hit <- intersect(program_ids, path_ids)
      p <- if (length(program_ids) == 0 || length(path_ids) == 0) {
        NA_real_
      } else {
        stats::phyper(
          length(hit) - 1, length(path_ids), length(universe_ids) - length(path_ids),
          length(program_ids), lower.tail = FALSE
        )
      }
      return(data.frame(
        program = k,
        term_id = prespecified$term_id[i],
        pathway = paste0(prespecified$source[i], ": ", prespecified$description[i]),
        expected_clusters = prespecified$expected_clusters[i],
        n_program_genes = length(program_ids),
        n_pathway_genes_at_t2d = length(path_ids),
        n_hit = length(hit),
        p = p,
        genes = paste(sort(unique(gene_name[as.character(hit)])), collapse = ", "),
        stringsAsFactors = FALSE
      ))
    })))
  }))
  prespecified_grid$p_bonferroni <- pmin(
    1, prespecified_grid$p * sum(is.finite(prespecified_grid$p))
  )
  prespecified_best <- prespecified_grid |>
    dplyr::filter(is.finite(p)) |>
    dplyr::group_by(pathway) |>
    dplyr::slice_min(p, n = 1, with_ties = FALSE) |>
    dplyr::ungroup()
  print(knitr::kable(
    prespecified |>
      dplyr::transmute(pathway = paste0(source, ": ", description), expected_clusters) |>
      dplyr::left_join(
        prespecified_best |>
          dplyr::transmute(
            pathway,
            best_program = paste0("P", program),
            n_hit, n_pathway_genes_at_t2d,
            p = signif(p, 3),
            p_bonferroni = signif(p_bonferroni, 3),
            genes
          ),
        by = "pathway"
      ),
    caption = paste0(
      "Pre-specified pathways: the program most enriched for each, against the ",
      "genes at all T2D SNPs. Compare best_program's card with expected_clusters. ",
      "p_bonferroni corrects over every program x pathway test."
    ),
    row.names = FALSE
  ))
}
```

### 4d. Program summary cards

One row per analysed program, valid programs first, then by how much more
similar its member SNPs are than random T2D SNPs (`similarity_lift`):

- **status:** "valid", or the checks it fails (see Program validation in
  Part 3).
- **similarity_q, replication:** the gate statistics; **strength_q** is
  reported only.
- **parent:** the larger program it sits inside, if any.
- **specific_traits:** the top three module-specific traits.
- **tissues and pathways:** hard-cutoff hits at FDR ≤ 0.05, with pre-specified
  pathway hits at Bonferroni p ≤ 0.05.
- **suzuki_loci:** positional genes named at a Suzuki index SNV, with the
  cluster Suzuki gave them.
- **genes:** the first 10 positional and QTL genes, strongest loading first.
  Genes in **bold** are named at a Suzuki et al. index SNV.

With a long gene list, some famous names will turn up by chance, so weigh the
traits and the Suzuki loci together rather than any single gene.

```{r program-cards, results = "asis"}
if (nrow(program_genes) == 0) {
  cat("No programs.\n")
} else {
  top_items <- function(res, value_col, n = 3L) {
    if (is.null(res)) {
      return(stats::setNames(character(0), character(0)))
    }
    out <- vapply(res$by_group, function(x) {
      tab <- if (value_col == "description") x$pathways else x$enriched
      if (is.null(tab) || nrow(tab) == 0) {
        return("none")
      }
      tab <- tab[is.finite(tab$fdr) & tab$fdr <= 0.05, , drop = FALSE]
      if (nrow(tab) == 0) {
        return("none")
      }
      tab <- tab[order(tab$fdr), , drop = FALSE]
      return(paste(utils::head(tab[[value_col]], n), collapse = "<br>"))
    }, character(1))
    return(stats::setNames(out, vapply(res$by_group, function(x) as.character(x$group), character(1))))
  }
  tissues_top <- top_items(group_tissues, "tissue")
  pathways_top <- top_items(module_enrichment, "description")
  traits_top <- stats::setNames(
    vapply(group_traits$by_group, function(x) {
      passing <- x$traits |> dplyr::filter(passes_min_specificity) |> utils::head(3L)
      return(if (nrow(passing) == 0) "none" else paste(passing$trait_name, collapse = "<br>"))
    }, character(1)),
    vapply(group_traits$by_group, function(x) as.character(x$group), character(1))
  )
  prespecified_hits <- if (nrow(prespecified_grid) > 0) {
    hits <- prespecified_grid[is.finite(prespecified_grid$p_bonferroni) &
                                prespecified_grid$p_bonferroni <= 0.05, , drop = FALSE]
    if (nrow(hits) == 0) {
      character(0)
    } else {
      tapply(hits$pathway, hits$program, paste, collapse = "<br>")
    }
  } else {
    character(0)
  }
  lookup <- function(x, key) {
    out <- if (length(x) == 0) NA_character_ else unname(x[key])
    return(ifelse(is.na(out), "none", out))
  }
  suzuki_loci <- program_members |>
    dplyr::inner_join(positional_links, by = "snp_id", relationship = "many-to-many") |>
    dplyr::inner_join(suzuki_gene_clusters, by = "gene") |>
    dplyr::arrange(program, dplyr::desc(abs_loading)) |>
    dplyr::distinct(program, gene, suzuki_clusters) |>
    dplyr::group_by(program) |>
    dplyr::summarise(
      suzuki_loci = paste0("<b>", gene, "</b> (", suzuki_clusters, ")", collapse = "<br>"),
      .groups = "drop"
    )
  gene_lists <- dplyr::bind_rows(lapply(split(program_members, program_members$program), function(m) {
    return(data.frame(
      program = m$program[1],
      positional_genes = format_genes(genes_in_order(m, positional_links), 10L),
      qtl_genes = format_genes(genes_in_order(m, qtl_links), 10L),
      stringsAsFactors = FALSE
    ))
  }))
  validation <- program_summary$programs |>
    dplyr::transmute(
      program = as.character(program),
      status,
      similarity_lift,
      similarity_q = signif(similarity_q, 2),
      strength_q = signif(strength_q, 2),
      replication = round(replication, 2),
      parent = dplyr::if_else(is.na(parent_program), "none", paste0("P", parent_program))
    )
  cards <- program_genes |>
    dplyr::select(program, n_snps) |>
    dplyr::left_join(validation, by = "program") |>
    dplyr::mutate(
      specific_traits = lookup(traits_top, program),
      tissues = lookup(tissues_top, program),
      pathways = lookup(pathways_top, program),
      prespecified_pathways = lookup(prespecified_hits, program)
    ) |>
    dplyr::left_join(suzuki_loci, by = "program") |>
    dplyr::mutate(suzuki_loci = tidyr::replace_na(suzuki_loci, "none")) |>
    dplyr::left_join(gene_lists, by = "program") |>
    dplyr::arrange(dplyr::desc(status == "valid"), dplyr::desc(similarity_lift), dplyr::desc(n_snps)) |>
    dplyr::select(-similarity_lift)
  print(knitr::kable(
    cards,
    caption = paste0(
      "One card per analysed program, valid programs first, then by similarity lift. ",
      "Bold genes are named at a Suzuki et al. index SNV; suzuki_loci gives the ",
      "cluster(s) Suzuki assigned them."
    ),
    row.names = FALSE,
    escape = FALSE
  ))
}
```

## Part 5: Candidate trait pairs for a multi-trait analysis

This part is documentation only; nothing here is run. Each pair runs the
upload against one partner trait in
[the multi-trait investigation](investigation-multitrait-programs.html). The
Suzuki clusters predict which shared programs the pair should show and in which
direction:

- **concordant:** the T2D risk allele raises the partner trait;
- **antagonistic:** the T2D risk allele lowers it.

| Partner trait | GPMap id | N | Coloc groups | Clusters it should separate | Expected pattern |
|---|---|---|---|---|---|
| Body mass index | 1992 | 532,396 | 992 | Obesity vs beta-cell | Obesity programs shared and concordant; Beta cell ±PI programs T2D-specific |
| Triglyceride levels | 1631 | 437,532 | 587 | Lipodystrophy, Metabolic syndrome, Liver/lipid | Lipodystrophy and Metabolic syndrome concordant; TM6SF2-type liver loci antagonistic |
| HDL cholesterol levels | 1630 | 400,754 | 735 | Lipodystrophy, Obesity, Metabolic syndrome | Antagonistic sharing (T2D risk lowers HDL) |
| Waist-to-hip ratio adjusted for BMI | 1653 | 458,349 | 478 | Lipodystrophy vs Obesity | Lipodystrophy concordant, with BMI removed |
| Waist-hip ratio | 1994 | 502,773 | 550 | Lipodystrophy, Obesity, Metabolic syndrome | Concordant for all three |
| Haemoglobin A1c levels | 1640 | 437,749 | 838 | Beta-cell, Residual glycaemic | Glycaemic clusters shared; obesity/lipid clusters weaker |
| Fasting glucose | 913 | 200,622 | 95 | Beta-cell, Residual glycaemic | As HbA1c, but sparse in GPMap |
| Body fat percentage | 2581 | 401,772 | 671 | Body fat, Obesity vs Lipodystrophy | Body fat and Obesity concordant; Lipodystrophy antagonistic |
| Alanine aminotransferase / GGT | 1642 / 1634 | ≈437,700 | 449 / 702 | Liver/lipid, Metabolic syndrome | Liver-fat loci concordant |
| Coronary artery disease | 2475 | 352,063 | 64 | Obesity vs Beta cell +PI | Obesity concordant; Beta cell +PI antagonistic (Suzuki's polygenic-score result) |
| Type 2 diabetes (GPMap native) | 114 | 655,666 | 178 | none | Replication pair: programs should correspond one to one |

**Caveat: sample overlap.** UK Biobank, and for several traits FinnGen and the
Million Veteran Program, contribute to both the T2DGGI EUR meta-analysis and
most partner GWAS. Correlated sampling error then inflates apparent sharing
between the two traits. Sharing that is weak or confined to a single locus
should be read with this in mind.

**Untested.** The multi-trait vignette has not yet been run with an upload GUID
as one of its traits. A first run would look like this:

```{r multitrait-example, eval = FALSE, echo = TRUE}
rmarkdown::render(
  "vignettes/investigation-multitrait-programs.Rmd",
  params = list(
    api = "production",
    trait_ids = c("c6d2e20f-6ece-d093-78ee-37534547c466", 1992)
  ),
  output_file = "investigation-multitrait-programs-t2dggi-bmi.html"
)
```
