---
title: "Multi-trait EBMF program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Multi-trait EBMF program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take **two (or more) traits**, each already decomposed independently into EBMF
programs by the univariate pipeline
([Univariate SNP biological clustering](investigation-snp-clustering.html)),
and ask: **which programs correspond across traits, and do the loci they share
push both traits the same way or in opposite directions?**

The runnable implementation is
[Investigation of multi-trait EBMF program correspondence](investigation-multitrait-programs.html).

This is not a search for how much biology is shared. A trait's candidate SNP
set is already restricted to that trait's own genome-wide-significant,
fine-mapped hits, so two only-moderately-correlated traits will typically share
a small fraction of that universe — that is expected, not a failure of the
method. The job of this pipeline is to resolve, cleanly and with calibrated
significance, whatever correspondence *does* exist, and to say honestly
whether it is concordant or antagonistic.

### Core correspondence preferences

The method should satisfy:

1. **Programs are discovered independently per trait.** No joint
   factorization across traits. This preserves each trait's own architecture
   (a program found in trait A is a real object even if trait B never enters
   the fit) and avoids the missing-data problem a joint approach would create:
   for two traits whose candidate loci barely overlap, most of a joint
   matrix's cells would be missing by construction.
2. **Evidence should scale with the amount of shared architecture, not
   penalise it.** A pair of programs that co-load on thirty loci should never
   score below a pair that happens to align tightly on three. This directly
   targets a defect in an earlier design of this pipeline, which correlated
   the programs' background-trait profiles: correlation is scale-free, so a
   handful of degenerately-aligned features could out-rank genuine broad
   overlap. Measured on a real trait pair, the correlation-based statistic's
   agreement with locus overlap was `Spearman(|stat|, n_loci_shared) ~ 0.14`;
   the loading-weighted replacement reaches `~0.7`. See Step 3.
3. **Two tiers of link strength.** A strict, calibrated one-to-one
   correspondence (`primary`) plus a looser, uncalibrated-for-search
   `candidate` tier, because EBMF's soft factorisation means one biological
   module can legitimately split into more than one program within a trait —
   a single true correspondence can then look like several plausible
   candidate pairs on one side.
4. **Direction is answered twice, on purpose.** Whether two programs' loading
   patterns line up is a different question from whether the traits' own
   allele effects move together, and the two can disagree. Both are reported;
   neither is allowed to silently stand in for the other.
5. **Biological corroboration is held out.** Annotations play no role in
   forming a link, so agreement between the two linked programs' independently
   computed enrichment results is genuine support for the link, not a
   restatement of the evidence that created it.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Independent EBMF per trait\\nvalid programs only\"]
  S2 [label=\"2. Shared locus (coloc-group) universe\\nper trait pair, no lFSR/magnitude gate\"]
  S3 [label=\"3. Loading-weighted locus concordance\\nsum_l w_l f_a(l) f_b(l)\"]
  S4 [label=\"4. Calibrated significance, two link tiers\\nprimary (reciprocal best match) / candidate\"]
  S5 [label=\"5. Direction: concordance sign vs module rg\\nlatent loadings vs observed allele effects\"]
  S6 [label=\"6. Biological corroboration\\ntrait categories, tissues (pathways planned)\"]
  S1 -> S2 -> S3 -> S4 -> S5 -> S6
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

### 1. Independent per-trait EBMF decomposition

Each target trait is run through the univariate pipeline exactly as described
in [Univariate SNP biological clustering](investigation-snp-clustering.html):
`run_univariate_clustering()` followed by `summarise_ebmf_programs()`. Only
programs with `status == "valid"` are carried forward — this analysis is about
correspondence between traits, not about re-litigating which univariate
programs are reliable, so invalid programs are dropped before any cross-trait
comparison is attempted.

`extract_program_loadings()` then sign-orients each program to its own target
trait, reflecting both the locus loadings (`F_pm`) and the background-trait
profile (`L_pm`) so that the target-trait loading is positive. This is what
makes a cross-trait direction call interpretable at all: EBMF's factor sign is
otherwise arbitrary and independent between the two traits' separate fits, and
without this anchoring "the same sign" would mean nothing.

Output objects, one set per trait:

- `loadings` — one row per locus x program: `coloc_group_id`, `loading`
  (sign-oriented `F_pm`), `lfsr`, `high_confidence` (the lFSR/magnitude gate,
  kept as a label — see Step 2).
- `profiles` — one row per background-feature x program (`L_pm`); used for
  the biological-corroboration step, not for the correspondence test itself.
- `programs` — per-program metadata (`program_id`, `trait_id`, `trait_name`).

### 2. Shared locus universe

Programs are compared at the level of the **`coloc_group_id`** — the shared
fine-mapped locus — rather than the exact `variant_id`, because each trait
carries one candidate SNP per coloc group.

For a pair of traits A and B, the locus axis is the **intersection of both
traits' full candidate-locus universes**: every `coloc_group_id` either
trait's EBMF fit assigned a nonzero loading to, on either side. No lFSR or
magnitude threshold is applied here. This is deliberate and is the first place
this pipeline breaks from a hard-cutoff design: restricting to only the
`high_confidence` (lFSR/magnitude-gated) loci would shrink an already-small
shared universe further, for no statistical benefit — the concordance
statistic in Step 3 down-weights low-confidence loci continuously instead.

`high_confidence` survives only as a **reporting** diagnostic:
`n_loci_shared` and `jaccard_loci` (the overlap of the two programs'
high-confidence membership sets) are computed and shown alongside every pair,
but do not gate the test.

Because each trait's candidate universe is independently restricted to that
trait's own significant, fine-mapped hits, the shared universe is typically a
small fraction of either trait's total (in one real run, 72 of 3,408 pooled
distinct loci — about 2%). This bounds how much cross-trait evidence can exist
before any statistic is applied, and is worth checking directly (Step 2 of the
runnable vignette reports it) before interpreting a low link count as a
negative result.

### 3. Loading-weighted locus concordance

For a candidate pair of programs — program \(a\) from trait A, program \(b\)
from trait B — score how much they co-load on the shared locus axis:

\[
\text{concordance}(a,b) = \sum_{l} w_l\, f_a(l)\, f_b(l),
\qquad w_l = \bigl(1-\text{lfsr}_a(l)\bigr)\bigl(1-\text{lfsr}_b(l)\bigr)
\]

where \(f_a(l)\) is program \(a\)'s sign-oriented loading at locus \(l\) (zero
if the program has no support there). This is deliberately **covariance-like,
not correlation-like**: it is left unnormalised so that evidence accumulates
with the amount of shared loading. A fully normalised (cosine) version is
reported alongside as `cosine_loci`, purely as a diagnostic of how tight the
alignment is — it is never used to rank or decide, because being scale-free it
cannot tell a pair sharing three loci from a pair sharing thirty, which is
exactly the failure this design avoids.

`n_eff_loci` — the participation ratio of the per-locus contributions,
\(\bigl(\sum_l |c_l|\bigr)^2 / \sum_l c_l^2\) where \(c_l\) is locus \(l\)'s
term in the sum above — is reported so that a large concordance built on two
or three effectively-independent loci is visibly thin, rather than
indistinguishable from one built on twenty.

Where several SNPs map to the same `coloc_group_id`, the SNP with the largest
`|loading|` represents that locus for that program (a deterministic but
somewhat arbitrary choice; a mean-based alternative has not been checked for
sensitivity).

Because loadings are sign-oriented to each trait's own target (Step 1),
`sign(concordance)` gives a first read on direction — but see Step 5: this is
a statement about loading geometry, not about the traits' own effect
directions, and the two can disagree.

### 4. Calibrated significance and two link tiers

**Per-pair test.** For each candidate pair, a permutation null is built by
shuffling the locus labels of program \(b\)'s (loading, confidence) vector —
so the null preserves \(b\)'s own loading distribution and sparsity — and
recomputing the concordance. The pair's `concordance_z` standardises the
observed value against this null's mean and SD; `p_concordance` is the
two-sided permutation p-value.

**Per-program calibration.** Testing every pair independently and correcting
across all of them would not account for the search over partners: a program
in a large trait effectively gets many chances to find a partner that looks
good by chance. Instead, each program's *single best* partner (by
`|concordance_z|`) is compared against a **max-statistic permutation null**:
at each permutation draw, the same shuffle used for the per-pair test is
standardised the same way and the maximum `|z|` over that program's candidate
partners is recorded, building a null distribution of "the best partner this
program would find by chance." The observed best match is compared against
that null to give an empirical p-value, which is then BH-corrected **across
programs** to give `matches$q` and `matches$significant`.

**Two link tiers**, both derived from the quantities above but corrected on
different axes:

- **`primary`** — the pair is the mutual best match of two `significant`
  programs (`shared = TRUE`). Because `best_partner` is single-valued per
  program, a program can be part of at most one primary edge, so weakly
  connected components of the primary-edge graph are always singletons or
  isolated pairs — never transitive chains. This tier feeds
  `build_program_families()` directly.
- **`candidate`** — the pair's own `q_concordance` (Benjamini-Hochberg,
  computed across **every** cross-trait pair tested in the run — a different,
  simpler axis than the per-program calibration above) clears
  `fdr_threshold`, without being a reciprocal best match. These are not
  corrected for the search over partners and are not used for family-building;
  they exist because a single true correspondence can legitimately appear as
  more than one plausible pair when EBMF has split a module across programs
  on one side. Treat them as leads, not conclusions.

**Program families.** `build_program_families()` (default
`rule = "reciprocal_best_match"`) builds an undirected graph with every
program as a node and `primary`-tier pairs as edges; weakly connected
components are the families. A program with no primary link is its own
family of one.

### 5. Direction: loading concordance vs module rg

`concordance_z`'s sign answers one question: do two programs' **loading
patterns** line up. It says nothing about whether the alleles involved push
the two traits the same way — that is a statement about observed effect
sizes, not about latent factor geometry, and `module_rg()` answers it
directly.

For a linked pair, across the shared locus axis (Step 2 — the full candidate
universe, not just `high_confidence` loci), `module_rg()` correlates the two
traits' **own** SNP effects, weighting each locus by how strongly the module
loads on it in both programs and by the precision of the two effect
estimates:

\[
w_l = \bigl(1-\text{lfsr}_a(l)\bigr)\bigl(1-\text{lfsr}_b(l)\bigr)\,
\frac{|f_a(l)\, f_b(l)|}{\text{se}_A(l)^2 + \text{se}_B(l)^2}
\]

so a locus the module does not load on drops out continuously rather than by
a threshold, and a locus with high measurement uncertainty in either trait
contributes less. `rg` is the resulting weighted correlation of \(\beta_A\)
against \(\beta_B\); its confidence interval is a percentile bootstrap over
loci (`n_boot`, default 1000).

- `rg > 0` — the alleles that raise trait A also raise trait B, across this
  module's loci.
- `rg < 0` — they oppose: the module is antagonistic between the two traits.
- The CI is load-bearing: **if it spans zero, direction is not established**,
  regardless of the point estimate. A pair can look strongly linked on
  `concordance_z` while its `rg` CI is uninformatively wide if the shared
  locus count is small (`n_loci_rg` reports how many loci actually
  contributed).
- `n_allele_mismatch` counts loci dropped because the two traits'
  `variant_id` disagreed at that locus (which would make a naive sign
  comparison meaningless). This should be `0` in practice; a nonzero count
  would mean effect-allele coding differs between the two traits' data at
  that locus.

`rg_direction` and `direction` (Step 3's concordance sign) are **not**
guaranteed to agree, because they are answers to different questions — one
about observed allele effects, one about latent loading structure. Where they
disagree, `rg` together with its CI is the more directly interpretable claim,
since it is anchored to actual effect sizes rather than factorisation
geometry.

**What this does not correct for.** Neither the concordance test nor
`module_rg()` accounts for sample overlap between the two traits' underlying
GWAS. If both traits were measured in largely the same cohort (as is often
the case, e.g. two UK Biobank traits), their effect-size estimation errors are
correlated, which biases `rg` away from zero even under a true null of no
shared genetic effect. There is currently no correction for this (the
GLEANR/cross-trait-LDSC approach is the standard fix); treat `rg` magnitudes
for same-cohort trait pairs as optimistic until this is addressed. Separately,
because every shared locus entered *both* traits' candidate universes by
clearing each trait's own significance and fine-mapping bar, the estimated
effects at those loci are subject to winner's curse in both directions.

### 6. Interpreting linked programs: biological corroboration

For each linked pair (both tiers), each side's high-confidence SNPs are
enriched **separately**, against that trait's own full candidate SNP universe
(not the tiny module itself, and not the other trait), using the existing
hard-cutoff enrichment functions
(`enrich_snp_group_trait_categories()`, `enrich_snp_group_tissues()`; pathway
enrichment via KEGG/Reactome is a planned extension, not yet wired in). FDR
here is the ordinary hypergeometric BH correction described in
[Univariate SNP biological clustering](investigation-snp-clustering.html#multiple-testing-correction-for-trait-category-and-tissue-enrichments) —
corrected across the categories/tissues tested **for that one program**,
independently on each side. It is not pooled across programs the way the
continuous univariate enrichment engine is; the two are separate mechanisms
operating on separate function families.

Because these annotations play no role in forming the link, agreement between
the two sides is genuine, held-out support that the correspondence is
biologically real, not a restatement of the loading evidence that created it.
Disagreement does not disprove the link — it means the link rests on the loci
alone, which is still meaningful evidence, just unconfirmed by a second,
independent axis.

One caveat: a program holding a large share of its trait's SNPs is compared
against a background that is largely itself, so "no significant enrichment"
for a dominant program is expected rather than informative — it says the
program is not distinguishable from the trait's overall composition, not that
it lacks biology.

## Variables to consider

Exposed in the multi-trait vignette today. Parameters fall into two groups:
those controlling the **correspondence test** itself, and those inherited
from the **univariate pipeline** that determine what counts as a program at
all before cross-trait comparison begins.

### Program correspondence parameters

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `n_perm` | `200`, `1000`, `3000` | Permutation replicates for both the per-pair and per-program calibration nulls | More replicates resolve finer p-values at the tail. With few replicates, several strong pairs can tie at the same BH floor (e.g. ten pairs tying at `q = 0.018` with `n_perm = 1000`); raising it separates them at the cost of runtime, since both the per-pair and per-program nulls are permutation-based. |
| `fdr_threshold` | `0.05` | BH threshold defining both program-level significance (`matches$significant`, gating the `primary` tier) and pair-level `q_concordance` (gating the `candidate` tier) | Lowering it tightens both tiers simultaneously; the two tiers are corrected on different axes (per-program max-statistic vs per-pair, across all pairs) so they will not necessarily shrink by the same amount. |
| `min_shared_loci` | `0` (default), `3` | Reporting-only threshold for the `candidate` column flag | Does **not** gate the concordance test itself (unlike the predecessor function it replaced) — every pair is tested regardless of this value; it only marks which pairs also clear a minimum *high-confidence* locus overlap for context. |
| `tiers` (in `module_rg()`) | `c("primary", "candidate")` (default), `"primary"`, `NULL` | Which link tiers get an `rg` estimate | Restricting to `"primary"` limits direction estimates to the calibrated, one-to-one tier; `NULL` estimates `rg` for every tested pair regardless of link status. |
| `n_boot` (in `module_rg()`) | `1000` | Percentile-bootstrap replicates for the `rg` confidence interval | More replicates give a smoother CI estimate; the CI's width is driven mainly by `n_loci_rg`, not `n_boot`, so increasing this rarely changes whether the CI excludes zero. |

### Univariate parameters that flow through

Everything in [Univariate SNP biological clustering](investigation-snp-clustering.html#variables-to-consider)
applies per trait before this pipeline ever runs, and matters twice over here:
`ebmf_lfsr_threshold`, `ebmf_magnitude_threshold`, `min_module_size`,
`min_mean_internal`, `min_connectedness`, and `stability_threshold` together
decide which programs are `valid` (and therefore eligible for cross-trait
comparison at all), while `trait_subset`, `include_trans`, `min_snp_signals`,
and `compress_method` shape the pleiotropy matrix each trait's programs were
discovered from in the first place. A program that is only marginally valid
in one trait's own terms is not made more reliable by finding a cross-trait
partner for it.

## Structure of Results

The multi-trait analogue of the univariate trait/program/SNP levels is
trait-pair/linked-pair/locus: the question moves from "what structure exists
within this trait" to "how much of it is shared, is it reliable, and which
direction does it point."

| Level | Primary question | Main output |
|---|---|---|
| **Trait pair** | How much shared candidate architecture exists between these two traits, and how much of it resolves into linked programs? | Size of the shared locus universe (vs. each trait's total), number of valid programs per trait, number of primary/candidate links, number of multi-trait families |
| **Linked pair / family** | Is this correspondence reliable, and which way does it point? | `concordance_z`, `q_concordance` / `matches$q`, tier (`primary`/`candidate`), `n_eff_loci`, `direction`, `rg` with CI and `rg_direction`, biological-annotation agreement between the two sides |
| **Locus** | Which loci actually carry the shared signal for this pair? | Per-locus contribution to `concordance` (which loci dominate the sum — worth checking directly when `n_eff_loci` is small), each program's loading at that locus, each trait's own `beta`/`se` there |

As with the univariate framework, the aim is not a single verdict per program
pair but a layered result: statistical evidence of correspondence and
direction, kept distinct from the biological annotation used to interpret it.
