---
title: "Simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # Simulation parameters (inputs to simulate_trait / the generative tiers)
  K: 6
  module_sizes: [60, 25, 8, 5, 3, 3]
  n_snps: 300
  n_traits: 600
  noise_sd: 1
  effect_size: 6
  effect_tail: 0
  overlap: disjoint
  nested_effect_corr: 0.8
  n_traits_per_module: [20, 15, 10, 8, 5, 3]
  background_sparsity_sd: 1.2
  # Similarity-graph realism. background_corr = 0 is the historical behaviour
  # (independent, random-signed background, which contributes nothing to
  # SNP-SNP similarity); snp_pleiotropy_sd is the per-SNP analogue of
  # background_sparsity_sd. See simulated_graph_diagnostics().
  background_corr: 0
  snp_pleiotropy_sd: 0
  background_effect_scale: 0.5
  target_pattern: module
  # Iterated generative axes: each element of these arrays becomes one input
  # version in a single run of this vignette. Paired arrays are iterated in
  # lock-step; the first element of each array is the baseline. All arrays are
  # written type-homogeneous (floats) so rmarkdown parses them as vectors
  # rather than lists of scalars.
  trait_overlap: [0.0, 0.5, 0.75, 1.0]
  trait_effect_corr: [0.0, 0.8, 0.8, 0.8]
  # p_active_background is crossed (full grid) with trait_overlap, so each
  # background density is evaluated at every trait-overlap level.
  p_active_background: [0.00, 0.02, 0.05, 0.07]
  snp_driver_groups: [1, 2, 3]
  driver_sharing: [0.0, 0.1, 0.25]
  # p_spurious and p_structural_zero are varied one at a time on a harder
  # baseline (trait_overlap 0.5, p_active_background 0.02) so contamination and
  # dropout are judged against a realistic background; see the axes below.
  p_spurious: [0.0, 0.01]
  p_structural_zero: [0.0, 0.1]

  # EBMF / clustering parameters (inputs to run_univariate_clustering)
  trait_subset: "phenotypic"
  min_snp_signals: 2
  compress_method: asinh
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_lfsr_threshold: 0.01
  ebmf_magnitude_threshold: 0.5
  ebmf_drop_global: true
  ebmf_prior: point_laplace
  ebmf_backfit: true

  # Program validation parameters (inputs to summarise_ebmf_programs)
  min_module_size: 3
  min_mean_internal: 0.3
  min_connectedness: 0.5
  # Loading-permutation coherence gate (replaces min_mean_internal /
  # min_connectedness as the gate; those two are now reported only).
  coherence_n_perm: 2000
  coherence_q: 0.05
  ebmf_greedy_Kmax: 100
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_stability_rep: 1
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5

  # Replication settings. n_sim is the Monte Carlo sample size (ADEMP);
  # override it at render time for fast test runs, e.g.
  #   rmarkdown::render(..., params = list(n_sim = 10))
  n_sim: 200
  master_seed: 1
  # Persist the ADEMP estimates/performance datasets when TRUE; output_dir
  # defaults to tempdir() when empty.
  write_outputs: false
  output_dir: ""
  # quick: fast test mode. Caps n_sim at 3, uses n_null = 1, and disables the
  # stability subsampling -- it trades reporting rigour for speed, not
  # breadth: all input versions (the full generative grid) are always run,
  # under quick or not. Use for exploring how many input parameters move the
  # recovery metrics before committing to a full-depth run; the reported
  # Monte Carlo SEs will be large under quick = TRUE. Set to FALSE for the
  # full study.
  quick: true
vignette: >
  %\VignetteIndexEntry{Simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

n_sim <- as.integer(params$n_sim)
quick <- isTRUE(params$quick)
if (quick) {
  n_sim <- min(n_sim, 3L)
}
n_null <- if (quick) 1L else as.integer(params$n_null)
n_stability_rep <- if (quick) 0L else as.integer(params$n_stability_rep)
set.seed(params$master_seed)
# One seed per replicate, reused across every input version (common random
# numbers). This makes version-to-version differences far more precise than
# independent seeds, and it makes the study exactly reproducible: parallel
# workers never touch the RNG state because every seed is passed explicitly.
sim_seeds <- sample.int(.Machine$integer.max, n_sim)

if (params$K != length(params$module_sizes)) {
  stop("K must equal length(module_sizes)")
}

clustering_settings <- list(
  min_snp_signals = params$min_snp_signals,
  trait_subset = params$trait_subset,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_greedy_Kmax = params$ebmf_greedy_Kmax,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  coherence_n_perm = params$coherence_n_perm,
  coherence_q_threshold = params$coherence_q,
  n_null = n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = n_stability_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)

simulation_defaults <- list(
  n_coloc_groups = params$n_snps,
  sign_pattern = "coherent",
  module_sizes = params$module_sizes,
  overlap = params$overlap,
  nested_effect_corr = params$nested_effect_corr,
  n_traits_per_module = params$n_traits_per_module,
  n_traits = params$n_traits,
  noise_sd = params$noise_sd,
  effect_size = params$effect_size,
  effect_tail = params$effect_tail,
  p_structural_zero = params$p_structural_zero[1],
  p_active_background = params$p_active_background[1],
  background_sparsity_sd = params$background_sparsity_sd,
  background_corr = params$background_corr,
  snp_pleiotropy_sd = params$snp_pleiotropy_sd,
  background_effect_scale = params$background_effect_scale,
  target_pattern = params$target_pattern,
  trait_overlap = params$trait_overlap[1],
  trait_effect_corr = params$trait_effect_corr[1],
  snp_driver_groups = params$snp_driver_groups[1],
  driver_sharing = params$driver_sharing[1],
  p_spurious = params$p_spurious[1]
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against synthetic data with known ground truth. It is organised around **input
versions**: the same simulation, EBMF, and program-validation pipeline is run
on versions of the input that differ in how the planted modules are structured
(background density and trait overlap, driver architecture, spurious
hits, and structural zeros), so the recovery metrics are directly comparable
across runs.

Each version is generated by `simulate_trait()`, which produces a trait object
(the target trait plus driver and background traits) with known ground truth.
The total number of SNPs is set via `n_coloc_groups`, and the total number of
traits (rows) via `n_traits`. Planted modules occupy **disjoint SNP sets** in
the clean version; overlap can then be introduced in trait space
(`trait_overlap`, `trait_effect_corr`), in SNP space (`driver_sharing`), or
within a module's own SNP set (`snp_driver_groups`).

The generative parameters are **calibrated to the messy marginals of a real
pleiotropy matrix** so the simulations are not trivially easy:

- Background association magnitudes are drawn from a standardised
  heavy-tailed distribution and scaled relative to the planted module effects
  by `background_effect_scale` (a relative effect-size parameter, not a
  significance threshold), so e.g. `background_effect_scale = 0.75` with
  `effect_size = 6` gives background hits with typical magnitude ~4.5, some
  below and some above. Noise is encoded as absence (`NA`), not as a hard
  z-score floor.
- Driver effects are heavy-tailed across traits (`effect_tail`), matching the
  across-trait heterogeneity seen in real modules.
- Background trait densities are heavy-tailed per trait
  (`background_sparsity_sd`), so most background traits are observed at only a
  handful of SNPs while a tail is dense — reproducing the real per-trait
  sparsity distribution.
- Driver traits can be split across SNP groups (`snp_driver_groups`) and can
  also act at the neighbouring module's SNP set (`driver_sharing`), creating
  within-module and cross-module SNP structure.
- Driver traits carry a low rate of background-scale spurious hits at
  off-module SNPs (`p_spurious`), and a fraction of each module's true-support
  cells is removed entirely (`p_structural_zero`), i.e. real-data dropout
  inside a module.
- The target-trait row is dense and significant at **every** SNP
  (`target_pattern = "dense"`), as real target rows are after orientation.

For every version, the identical battery is run: `run_univariate_clustering()`
fits EBMF and keeps overlapping soft programs, and
`summarise_ebmf_programs()` filters programs by lFSR, loading magnitude,
internal coherence, and trait-subsampling stability. The key output is a
per-program table in which **each validation check is its own column**, so it
is immediately visible which check failed for each program:

- `size` — at least `min_module_size` filtered SNPs (`n_snps_filtered`).
- `internal_sim` — mean internal SNP similarity `>= min_mean_internal`.
- `connectivity` — pair connectedness `>= min_connectedness`.
- `stability` — top-loading SNPs recovered with replication
  `>= stability_threshold` under trait subsampling.

A program is `valid` only when every check passes. Because the input now has
realistic marginals, **recovery is expected to be partial and noisy**: the
validation gates sit at the same levels that separate real modules from
background, so only a subset of the discovered programs should be `valid`, and
the pass rates below are the meaningful output for comparing input versions.
The analysis settings below are fixed across versions; only the generative
input changes.

## Does this simulation exercise the gates in the regime real data uses them?

The validation gates act on the SNP similarity graph, so what matters is not
whether the simulated matrix has realistic per-trait sparsity and magnitude
marginals -- it is whether the *graph* those marginals induce resembles a real
one. Those are not the same thing, and matching the marginals is not sufficient.

`simulated_graph_diagnostics()` measures the simulated graph and
`real_bmi_graph_targets()` carries the values measured on the BMI pleiotropy
matrix. The row to read first is `random_set_pass_rate`: the fraction of random
SNP sets whose mean internal similarity clears 0.3. On real BMI data that is
about 0.94 -- a random set of SNPs passes the old absolute gate almost always,
which is why that gate carried no specificity. Where the simulated value is near
0, a random set can never pass, so a null simulation study **cannot detect an
uninformative gate** and its false-discovery rate is close to un-failable by
construction.

The two generative parameters that move this are `background_corr` (background
traits sharing structure, so they contribute to SNP-SNP similarity at all --
at `0` they are independent and random-signed and contribute nothing) and
`snp_pleiotropy_sd` (per-SNP heterogeneity in how many traits a SNP is observed
on, which is what brings the median pairwise overlap down towards the real ~3
and produces the sparse SNPs that carry high cosines).

```{r graph-realism}
graph_diag <- tryCatch({
  sim_probe <- do.call(simulate_trait, c(
    simulation_defaults, list(K = length(params$module_sizes), seed = params$master_seed)
  ))
  probe_res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim_probe$trait_object), clustering_settings)
  )
  simulated_graph_diagnostics(
    probe_res$x_matrix,
    target_trait_id = probe_res$parameters$target_trait_id
  )
}, error = function(e) NULL)

if (is.null(graph_diag)) {
  cat("Graph diagnostics unavailable for this configuration.\n\n")
} else {
  targets <- real_bmi_graph_targets()
  shared_cols <- intersect(names(targets), names(graph_diag))
  comparison <- data.frame(
    statistic = shared_cols,
    simulated = signif(as.numeric(graph_diag[shared_cols]), 3),
    real_bmi = signif(as.numeric(targets[shared_cols]), 3),
    stringsAsFactors = FALSE
  )
  print(knitr::kable(
    comparison,
    row.names = FALSE,
    caption = paste0(
      "Similarity-graph diagnostics for this run's generative settings against ",
      "the values measured on the real BMI pleiotropy matrix. Read ",
      "random_set_pass_rate first: where it is near 0 the gates are not being ",
      "tested in the regime real data puts them in, whatever the recovery ",
      "metrics below say."
    )
  ))
}
```

## Simulation design (ADEMP)

This study and its null companion follow the ADEMP structure of Morris, White
and Crowther (2019, *Statistics in Medicine* 38:2074-2102,
\doi{10.1002/sim.8086}). The elements are:

**Aims.** To evaluate the EBMF program-discovery pipeline when the truth is
known: (i) how well planted modules are recovered (a **model-selection**
target), and (ii) whether the validation gates stay quiet when no structure is
planted (a **testing** target) — the latter is covered in
`investigation-univariate-null-simulations.html`. Both aim to *stretch* the
method by using realistic, messy marginals rather than a clean best case.

**Data-generating mechanisms.** Data are generated parametrically by
`simulate_trait()`. It is calibrated so that background magnitudes, per-trait
sparsity and driver heavy-tailedness match the marginal distributions of a real
pleiotropy matrix (see the bullets below). Factors are
varied as follows: `p_active_background` is **crossed** with `trait_overlap`
(full grid) so background density is assessed at every overlap level; the
driver-architecture, spurious-hit, and structural-zero axes are varied **one at
a time**. The driver axis holds the clean baseline, while the spurious-hit and
structural-zero axes are run against a harder baseline (`trait_overlap = 0.5`,
`p_active_background = 0.02`) so contamination and dropout are not judged on
trivially clean data. A full factorial over all axes was rejected because the
resulting run count is computationally infeasible; the crossed background grid
is the part where an interaction is a priori plausible. The number of SNPs
(`n_snps`), traits (`n_traits`) and the noise level are fixed across versions.

**Estimand / target.** The estimand is the true SNP-to-module partition,
`ground_truth$module_of_snp` (a selection target), with derived targets the
number of modules `K` and each module's SNP set. Recovery is scored by
`evaluate_univariate_simulation()` (see
`R/evaluate_univariate_simulation.R`).

**Methods.** A single pipeline is evaluated, with no comparators:
`run_univariate_clustering()` fits EBMF and `summarise_ebmf_programs()` applies
four gates (size, internal coherence, connectedness, stability). Thresholds are
fixed across versions and are the same as those used on real data.

**Performance measures.** With truth `module_of_snp` and predicted valid
memberships, and writing `truth`/`pred` for the labelled partitions:

- `k_hat = |distinct predicted programs|`, `k_planted = |distinct planted modules|`, `k_error = |k_hat - k_planted|`.
- `ari_structured`: adjusted Rand index over module SNPs only (background excluded), with each SNP projected to its strongest assigned program. Computed by `.adjusted_rand_index()` in `R/evaluate_univariate_simulation.R`.
- `module_recall = mean_m max_p |module_m ∩ program_p| / |module_m|` — the fraction of each planted module captured by its best predicted program; merge-fair, unlike ARI.
- `coverage = |{SNPs in any predicted program}| / n_snps`.
- `background_absorbed = |{background SNPs in any predicted program}| / n_background` — a SNP-level false-positive measure.
- Gate pass rates: per-replicate fraction of programs passing `size`, `internal_sim`, `connectedness`, `stability`, and `valid` (all gates).

**Monte Carlo uncertainty.** Each estimate is the mean over `n_sim` replicates
(default `r n_sim`; override with `params = list(n_sim = ...)` for fast test
runs). The replicate is the Monte Carlo unit, and every estimate is reported
with a Monte Carlo standard error and a 95% Monte Carlo interval. For a
continuous measure the MC SE is `sd / sqrt(n_sim)`; for a proportion it is
`sqrt(p(1-p)/n_sim)`; and `n_sim` can be chosen for a target MC SE via
`n_sim = p(1-p) / MCSE^2` (`n_sim_for_mcse()`). Common random numbers are used
(one seed vector shared across all input versions), which lowers the MC SE of
version-to-version differences. RNG states are recorded via the per-replicate
seed in the estimates dataset.

```{r settings-table}
settings_to_show <- c(
  "trait_subset", "min_snp_signals", "min_module_size", "module_sizes",
  "compress_method", "compress_scale", "similarity_threshold",
  "overlap", "nested_effect_corr",
  "n_traits_per_module", "trait_overlap", "trait_effect_corr",
  "n_traits",
  "effect_size", "effect_tail", "p_structural_zero", "p_spurious",
  "p_active_background", "background_sparsity_sd", "background_effect_scale",
  "target_pattern",
  "ebmf_lfsr_threshold", "ebmf_magnitude_threshold", "ebmf_prior",
  "min_mean_internal", "min_connectedness", "n_stability_rep",
  "frac_traits", "stability_top_n", "stability_threshold", "n_sim"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$similarity_threshold <- params$similarity_threshold
settings_values$n_stability_rep <- n_stability_rep
settings_values$module_sizes <- params$module_sizes
settings_values$overlap <- params$overlap
settings_values$nested_effect_corr <- params$nested_effect_corr
settings_values$n_traits_per_module <- params$n_traits_per_module
settings_values$trait_overlap <- params$trait_overlap[1]
settings_values$trait_effect_corr <- params$trait_effect_corr[1]
settings_values$n_traits <- params$n_traits
settings_values$effect_size <- params$effect_size
settings_values$effect_tail <- params$effect_tail
settings_values$p_structural_zero <- params$p_structural_zero[1]
settings_values$p_spurious <- params$p_spurious[1]
settings_values$p_active_background <- params$p_active_background[1]
settings_values$background_sparsity_sd <- params$background_sparsity_sd
settings_values$background_effect_scale <- params$background_effect_scale
settings_values$target_pattern <- params$target_pattern
settings_values$stability_top_n <- validation_settings$top_n
settings_values$n_sim <- n_sim
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      if (nm %in% c("module_sizes", "n_traits_per_module")) {
        return(paste(settings_values[[nm]], collapse = ", "))
      }
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "Baseline EBMF analysis and validation settings (iterated generative parameters are shown at their baseline value; each input version is listed in the tier table below)"
)
```

```{r tier-helpers}
make_tier_spec <- function(...) {
  overrides <- list(...)
  spec <- utils::modifyList(simulation_defaults, overrides)
  if (is.null(spec$n_coloc_groups)) {
    spec$n_coloc_groups <- params$n_snps
  }
  if (is.null(spec$K)) {
    spec$K <- length(spec$module_sizes)
  }
  if (spec$K == 0) {
    spec$module_sizes <- NULL
  }
  return(spec)
}

# Build one set of input versions by iterating over a generative axis. The
# named vectors are iterated in lock-step; length-1 values (including scalar
# baselines after a grid-script override) are recycled. `.transform` can
# rewrite the per-tier overrides if a parameter needs reshaping per version.
# `.label_params` restricts the version label to the named columns (useful when
# the axis also carries fixed baseline overrides that should not clutter it).
make_axis <- function(axis_name, ..., .transform = NULL, .label_params = NULL) {
  cols <- list(...)
  n <- max(lengths(cols))
  if (any(lengths(cols) != 1 & lengths(cols) != n)) {
    stop("All parameters in axis '", axis_name, "' must be length 1 or ", n)
  }
  label_params <- if (is.null(.label_params)) names(cols) else .label_params
  idx <- function(x, i) x[if (length(x) == 1) 1L else i]
  specs <- vector("list", n)
  labels <- character(n)
  for (i in seq_len(n)) {
    raw <- lapply(cols, function(x) idx(x, i))
    ov <- raw
    if (!is.null(.transform)) {
      ov <- .transform(ov)
    }
    specs[[i]] <- do.call(make_tier_spec, ov)
    parts <- vapply(label_params, function(nm) {
      paste0(nm, " = ", paste(format(raw[[nm]]), collapse = ", "))
    }, character(1))
    labels[i] <- paste0(axis_name, ": ", paste(parts, collapse = ", "))
  }
  names(specs) <- paste0(axis_name, "_", seq_len(n))
  names(labels) <- names(specs)
  return(list(labels = labels, specs = specs))
}

# Full cross of p_active_background with trait_overlap: every background
# density is evaluated at every trait-overlap level, so the background effect
# is tested over the trait-overlap versions.
make_background_cross <- function() {
  specs <- list()
  labels <- character(0)
  k <- 0L
  for (pb in params$p_active_background) {
    for (j in seq_along(params$trait_overlap)) {
      k <- k + 1L
      ov <- list(
        p_active_background = pb,
        trait_overlap = params$trait_overlap[j],
        trait_effect_corr = params$trait_effect_corr[j]
      )
      specs[[k]] <- do.call(make_tier_spec, ov)
      labels[k] <- sprintf(
        "background: p_active_background = %s, trait_overlap = %s",
        format(pb), format(params$trait_overlap[j])
      )
    }
  }
  names(specs) <- paste0("background_", seq_along(specs))
  names(labels) <- names(specs)
  return(list(labels = labels, specs = specs))
}

# For each found program, match it to the planted module whose SNP set it
# overlaps most (on the gated membership) and score recovery: module SNPs
# captured (`module_found`), module SNPs missed (`module_missed`), and program
# SNPs not in that module (`not_module`). Modules are a disjoint partition
# (`module_of_snp`; 0 = background). No-op when there are no programs or no
# planted modules.
score_module_recovery <- function(programs, assigned, module_of_snp) {
  if (nrow(programs) == 0 || is.null(module_of_snp)) {
    return(programs)
  }
  mod_ids <- sort(unique(module_of_snp[module_of_snp > 0]))
  if (length(mod_ids) == 0) {
    return(programs)
  }
  mod_snps <- lapply(mod_ids, function(m) {
    names(module_of_snp)[module_of_snp == m]
  })
  names(mod_snps) <- as.character(mod_ids)

  rows <- lapply(programs$program, function(p) {
    prog_snps <- assigned$snp_id[assigned$program == p]
    overlaps <- vapply(mod_ids, function(m) {
      sum(prog_snps %in% mod_snps[[as.character(m)]])
    }, integer(1))
    best <- mod_ids[which.max(overlaps)]
    n_module <- length(mod_snps[[as.character(best)]])
    n_found <- overlaps[best]
    data.frame(
      matched_module = best,
      n_module_snps = n_module,
      module_found = n_found,
      module_missed = n_module - n_found,
      not_module = length(prog_snps) - n_found,
      stringsAsFactors = FALSE
    )
  })
  m <- dplyr::bind_rows(rows)
  m$program <- programs$program
  return(dplyr::left_join(programs, m, by = "program"))
}

run_tier_one <- function(spec, rep) {
  dots <- spec
  # Common random numbers: the seed depends only on the replicate index, so
  # every input version sees the same simulation scenario in replicate `rep`.
  dots$seed <- sim_seeds[[rep]]
  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  # Vary the validation layer's internal randomisation by replicate too; a fixed
  # seed here would understate method-level Monte Carlo variability.
  vs <- validation_settings
  vs$seed <- sim_seeds[[rep]]
  smry <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), vs)
  )
  valid_ids <- smry$programs$program[smry$programs$status == "valid"]
  valid_assigned <- smry$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    predicted_memberships = valid_assigned
  )
  programs <- smry$programs
  programs$rep <- rep(rep, nrow(programs))
  programs <- score_module_recovery(
    programs, smry$assigned, sim$ground_truth$module_of_snp
  )
  driver_ov <- sim$ground_truth$driver_overlap
  true_driver_overlap <- if (!is.null(driver_ov) && nrow(driver_ov) > 1) {
    mean(driver_ov[upper.tri(driver_ov)])
  } else {
    NA_real_
  }
  list(
    summary = data.frame(
      sim_seed = dots$seed,
      k_planted = ev$k_planted,
      k_hat = ev$k_hat,
      k_error = ev$k_error,
      true_driver_overlap = true_driver_overlap,
      ari_all = ev$ari_all,
      ari_structured = ev$ari_structured,
      module_recall = ev$mean_module_recall,
      background_absorbed = ev$background_absorbed,
      coverage = ev$coverage,
      n_programs = nrow(programs),
      n_valid_programs = length(valid_ids),
      stringsAsFactors = FALSE
    ),
    programs = programs,
    driver_overlap = driver_ov
  )
}

run_tier <- function(name, label, spec, n_sim) {
  n_jobs <- 20L
  out <- if (n_sim > 1 && .Platform$OS.type != "windows") {
    parallel::mclapply(
      seq_len(n_sim),
      function(i) run_tier_one(spec, i),
      mc.cores = n_jobs,
      mc.set.seed = FALSE
    )
  } else {
    lapply(seq_len(n_sim), function(i) run_tier_one(spec, i))
  }
  list(
    summary = dplyr::bind_rows(lapply(out, `[[`, "summary")) |>
      dplyr::mutate(tier = label),
    programs = dplyr::bind_rows(lapply(out, `[[`, "programs")) |>
      dplyr::mutate(tier = label, tier_name = name),
    driver_overlap = out[[1]]$driver_overlap
  )
}
```

## Input versions

Each input version is produced by iterating over the generative parameter
arrays in the `params` block: every array element becomes one version, paired
arrays are iterated in lock-step, and the first element of each array is the
baseline. `p_active_background` is the exception — it is **crossed** with
`trait_overlap`, so every background density is evaluated at every
trait-overlap level. The versions fall into four groups, each stressing one
aspect of the planted structure; the driver-architecture axis holds the others
at their clean baseline, while the spurious-hit and structural-zero axes are
run on a harder baseline (`trait_overlap = 0.5`, `p_active_background = 0.02`):

- **Background density × trait overlap** (`p_active_background` ×
  `trait_overlap` / `trait_effect_corr`): the full grid of background
  association density against the fraction of each module's driver traits
  shared with the other modules (correlated, `trait_effect_corr`, but
  non-identical effects).
- **Driver architecture** (`snp_driver_groups` / `driver_sharing`): splitting
  each module's SNP set into groups (each driver trait acts on one group) and
  letting some driver traits also act at the neighbouring module's SNP set.
- **Spurious hits** (`p_spurious`): background-scale off-module hits in the
  driver rows.
- **Structural zeros** (`p_structural_zero`): a fraction of each module's
  true-support cells removed entirely (`NA`), emulating real-data dropout
  inside a module.

Null simulations (`K = 0`, no modules planted) are covered separately in
`investigation-univariate-null-simulations.html`.

```{r tiers}
# Each generative axis in the params block is iterated into input versions.
# p_active_background is crossed with trait_overlap (full grid); the driver
# axis iterates one parameter while holding the rest at their clean baseline,
# while the spurious and structural-zero axes run on a harder baseline.
axes <- list(
  background = make_background_cross(),
  drivers = make_axis(
    "drivers",
    snp_driver_groups = params$snp_driver_groups,
    driver_sharing = params$driver_sharing
  ),
  # Contamination and dropout are stress-tested on a harder baseline than the
  # clean one used by the geometry axes: moderate trait overlap (0.5, with the
  # background-grid pairing trait_effect_corr = 0.8) and some background
  # density (0.02). Only the varied parameter appears in the version label.
  spurious = make_axis(
    "spurious",
    p_spurious = params$p_spurious,
    trait_overlap = 0.5,
    trait_effect_corr = 0.8,
    p_active_background = 0.02,
    .label_params = "p_spurious"
  ),
  structural_zero = make_axis(
    "structural_zero",
    p_structural_zero = params$p_structural_zero,
    trait_overlap = 0.5,
    trait_effect_corr = 0.8,
    p_active_background = 0.02,
    .label_params = "p_structural_zero"
  )
)

tier_specs <- list()
tier_labels <- character(0)
for (a in axes) {
  tier_specs <- c(tier_specs, a$specs)
  tier_labels <- c(tier_labels, a$labels)
}

# All input versions (the full generative grid) are always run, quick or not.
# `quick` trades reporting rigour for speed (n_sim, n_null, stability, set
# above), not breadth -- the point of quick mode is to see how many input
# parameters move the recovery metrics, cheaply, before committing to a
# full-depth run.

tier_results <- lapply(names(tier_specs), function(nm) {
  run_tier(nm, tier_labels[[nm]], tier_specs[[nm]], n_sim)
})
names(tier_results) <- names(tier_specs)

tier_summary <- dplyr::bind_rows(lapply(tier_results, `[[`, "summary"))
program_checks <- dplyr::bind_rows(lapply(tier_results, `[[`, "programs"))
```

```{r tier-params-table}
tier_param_table <- do.call(rbind, lapply(names(tier_specs), function(nm) {
  s <- tier_specs[[nm]]
  data.frame(
    version = tier_labels[[nm]],
    n_snps = s$n_coloc_groups,
    K = s$K,
    module_sizes = if (s$K > 0) paste(s$module_sizes, collapse = ", ") else "-",
    n_traits_per_module = if (s$K > 0) paste(s$n_traits_per_module, collapse = ", ") else "-",
    trait_overlap = if (s$K > 0) s$trait_overlap else "-",
    trait_effect_corr = if (s$K > 0) s$trait_effect_corr else "-",
    noise_sd = s$noise_sd,
    p_structural_zero = if (s$K > 0) s$p_structural_zero else "-",
    p_spurious = s$p_spurious,
    effect_size = if (s$K > 0) s$effect_size else "-",
    effect_tail = if (s$K > 0) s$effect_tail else "-",
    background_effect_scale = if (s$K > 0) s$background_effect_scale else "-",
    p_active_background = s$p_active_background,
    background_sparsity_sd = s$background_sparsity_sd,
    target_pattern = if (s$K > 0) s$target_pattern else "-",
    driver_sharing = s$driver_sharing,
    sign_pattern = if (s$K > 0) s$sign_pattern else "-",
    snp_driver_groups = if (s$K > 0) s$snp_driver_groups else "-",
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  tier_param_table,
  row.names = FALSE,
  caption = "Generative parameters for each input version (each axis varies one feature of the planted structure; the spurious and structural-zero axes use a harder baseline, as shown in their trait_overlap / p_active_background columns)"
)

```

## Aggregate results

Across all replicates and programs of each version, the tables below report
the recovery metrics and, for every validation check, the fraction of programs
that passed it. Estimates are averages over `r n_sim` replicates and carry a
**Monte Carlo standard error** in parentheses (ADEMP; Morris, White & Crowther
2019): the replicate is the Monte Carlo unit, so every number here is an
estimate and the MC SE quantifies its uncertainty. `valid` is the fraction of
programs that passed all checks.

```{r aggregate}
# Per-replicate pass rates: the replicate is the Monte Carlo unit, so gate
# rates are computed within each replicate and then averaged across replicates.
rep_check_rates <- program_checks |>
  dplyr::group_by(tier, rep) |>
  dplyr::summarise(
    n_programs = dplyr::n(),
    rate_size = mean(size_pass),
    rate_internal_similarity = mean(internal_similarity_pass),
    rate_connectedness = mean(connectedness_pass),
    rate_stability = if (any(stability_checked)) {
      mean(stability_pass[stability_checked])
    } else {
      NA_real_
    },
    rate_valid = mean(status == "valid"),
    .groups = "drop"
  )

# Recovery metrics already have one row per replicate in tier_summary, so the
# Monte Carlo SE follows directly.
recovery_cols <- c(
  "k_hat", "n_valid_programs", "module_recall", "coverage",
  "ari_structured", "background_absorbed"
)
recovery_mc <- do.call(rbind, lapply(recovery_cols, function(cl) {
  out <- ademp_summarise(tier_summary, cl, group_cols = "tier", type = "mean")
  out$measure <- cl
  out
}))
recovery_mc$measure <- factor(
  recovery_mc$measure,
  levels = recovery_cols,
  labels = c(
    "k_hat", "valid programs", "module recall", "coverage",
    "ARI (structured)", "background absorbed"
  )
)

check_cols <- c(
  "rate_size", "rate_internal_similarity", "rate_connectedness",
  "rate_stability", "rate_valid"
)
check_mc <- do.call(rbind, lapply(check_cols, function(cl) {
  out <- ademp_summarise(
    rep_check_rates, cl, group_cols = "tier", type = "prop"
  )
  out$measure <- cl
  out
}))
check_mc$measure <- factor(
  check_mc$measure,
  levels = check_cols,
  labels = c("size", "internal sim", "connectivity", "stability", "valid")
)

format_est_mcse <- function(estimate, mcse, scale = 1, digits = 2) {
  est <- estimate * scale
  se <- mcse * scale
  ifelse(
    is.na(est), "-",
    ifelse(
      is.na(se),
      formatC(est, digits = digits, format = "f"),
      paste0(
        formatC(est, digits = digits, format = "f"), " (",
        formatC(se, digits = digits, format = "f"), ")"
      )
    )
  )
}

recovery_table <- recovery_mc |>
  dplyr::mutate(cell = format_est_mcse(estimate, mcse)) |>
  dplyr::select(tier, measure, cell) |>
  tidyr::pivot_wider(names_from = measure, values_from = cell)

check_mc_table <- check_mc |>
  dplyr::mutate(cell = format_est_mcse(estimate, mcse, scale = 100)) |>
  dplyr::select(tier, measure, cell) |>
  tidyr::pivot_wider(names_from = measure, values_from = cell)

knitr::kable(
  recovery_table,
  caption = paste0(
    "Recovery metrics by input version (mean across ", n_sim,
    " replicates; Monte Carlo SE in parentheses). module_recall and coverage ",
    "preserve overlapping valid memberships."
  )
)

knitr::kable(
  check_mc_table,
  caption = paste0(
    "Validation-gate pass rates by input version (percent; per-replicate rate ",
    "averaged across ", n_sim, " replicates, Monte Carlo SE in parentheses). ",
    "The stability rate is over programs re-checked after the size and ",
    "internal-coherence gates."
  )
)
```

The `stability` pass rate is computed only over programs that were actually
re-checked (those passing the size and internal-coherence checks).

```{r lollipop, fig.height=10, fig.width=9}
lollipop_data <- recovery_mc |>
  dplyr::filter(measure %in% c(
    "module recall", "coverage", "ARI (structured)", "background absorbed"
  ))
ggplot2::ggplot(
  lollipop_data,
  ggplot2::aes(x = estimate, y = tier)
) +
  ggplot2::geom_errorbarh(
    ggplot2::aes(xmin = conf_low, xmax = conf_high), height = 0.25
  ) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_wrap(~measure, scales = "free_x", ncol = 2) +
  ggplot2::labs(
    x = "estimate (95% Monte Carlo CI)", y = NULL,
    title = "Recovery performance with Monte Carlo uncertainty"
  ) +
  ggplot2::theme_minimal(base_size = 9)
```

## Program-level validation checks

For each version, every program found across the replicates is shown with the
four validation checks as separate columns. A check is marked `FAIL` when the
program does not meet that criterion; `n/a` under `stability` means the
program already failed the internal-coherence checks, so stability was not
re-tested.

Each program is also scored against the planted module it overlaps most (on
its gated SNPs): `% module found` is the fraction of that module's SNPs the
program recovered, `% module missed` its complement, and `% not module` is the
fraction of the program's SNPs that are not in that module (background or
other-module SNPs pulled in). `% module found` + `% module missed` always sum
to 100%. Under high `trait_overlap` a program that correctly merges two
modules matches only one, so its `% not module` rises even though it is
recovering real structure — read these alongside the merge-fair `module_recall`
in the Aggregate table.

`factor strength/snp %` is `100 * factor_strength / sqrt(n_snps_filtered)` —
the share of the observed matrix signal the program explains, normalised by the
square root of its filtered SNP count. It is reported for context and does not
gate `status`.

```{r check-helper, include=FALSE}
check_table <- function(which_tier) {
  df <- program_checks |>
    dplyr::filter(.data$tier_name == which_tier)
  if (nrow(df) == 0) {
    return(NULL)
  }
  df |>
    dplyr::mutate(
      check_size = ifelse(size_pass, "pass", "FAIL"),
      check_internal_similarity = ifelse(internal_similarity_pass, "pass", "FAIL"),
      check_connectedness = ifelse(connectedness_pass, "pass", "FAIL"),
      check_stability = dplyr::case_when(
        !stability_checked ~ "n/a",
        stability_pass ~ "pass",
        TRUE ~ "FAIL"
      ),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      factor_strength_per_snp_pct = round(100 * factor_strength_per_snp, 3),
      replication = round(replication, 3),
      matched = paste0("M", matched_module),
      pct_module_found = ifelse(
        n_module_snps > 0,
        paste0(round(100 * module_found / n_module_snps), "%"),
        "-"
      ),
      pct_module_missed = ifelse(
        n_module_snps > 0,
        paste0(round(100 * module_missed / n_module_snps), "%"),
        "-"
      ),
      pct_not_module = ifelse(
        n_snps_filtered > 0,
        paste0(round(100 * not_module / n_snps_filtered), "%"),
        "-"
      )
    ) |>
    dplyr::select(
      rep, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, factor_strength_per_snp_pct, replication, matched,
      pct_module_found, pct_module_missed, pct_not_module,
      check_size, check_internal_similarity, check_connectedness,
      check_stability, status
    ) |>
    dplyr::arrange(rep, program) |>
    knitr::kable(
      row.names = FALSE,
      col.names = c(
        "rep", "program", "n_snps", "n_snps filtered",
        "mean internal similarity", "connectedness",
        "factor strength/snp %", "replication",
        "matched module", "% module found", "% module missed", "% not module",
        "size", "internal sim", "connectivity", "stability", "status"
      )
    )
}
```

```{r tier-checks, results='asis'}
for (nm in names(tier_specs)) {
  cat("\n###", tier_labels[[nm]], "\n\n")
  tbl <- check_table(nm)
  if (is.null(tbl)) {
    cat("No EBMF programs were produced for this version.\n\n")
  } else {
    cat(tbl, sep = "\n")
    cat("\n\n")
  }
}
```

## Notes for interpreting runs

- **Fast test runs.** A production render is long: `n_sim = 200` over the 23
  input versions is thousands of EBMF fits, and the cost of a single fit is
  dominated by `summarise_ebmf_programs` (permutation calibration + stability),
  not by the repetition count alone. `quick` (default `TRUE`) trades reporting
  rigour for speed, not breadth: it always runs the full 23-version grid, but
  caps `n_sim` at 3, sets `n_null = 1`, and disables stability — about 1 min,
  and enough to see which input parameters move the recovery metrics before
  committing to a full-depth run. To iterate quickly, override `params`:
  - `params = list(quick = TRUE)` (default) — full 23-version grid, cheap
    internals (`n_sim = 3`, `n_null = 1`, no stability); about 1 min.
  - `params = list(quick = FALSE, n_sim = 20)` — full 23-version grid at
    higher fidelity (real `n_null`/stability) but reduced repetitions.
  - `params = list(quick = FALSE)` — the full study (`n_sim = 200`, full
    `n_null`/stability); long, intended for a production render.
  Lowering `n_null` and `n_stability_rep` gives the largest speed-up.
- Each input version comes from iterating the generative parameter arrays in
  the `params` block; arrays are iterated in lock-step with their paired
  parameter, and the first element of each array is the baseline.
- `p_active_background` is **crossed** with `trait_overlap` (full grid), so the
  background-density effect is evaluated at every trait-overlap level; the
  driver-architecture axis holds the others at their clean baseline, while the
  spurious-hit and structural-zero axes run on a harder baseline
  (`trait_overlap = 0.5`, `p_active_background = 0.02`).
- `module_recall` and `coverage` preserve overlapping valid EBMF memberships.
- Planted modules occupy disjoint SNP sets; module overlap in these simulations
  comes from shared trait drivers (`trait_overlap`, with correlated
  `trait_effect_corr`), or from driver traits acting at the neighbouring
  module's SNP set (`driver_sharing`).
- `snp_driver_groups` splits each module's SNP set into groups, each driver
  trait acting on only one group; with more groups a module's drivers are only
  recoverable from their own SNP subgroup, fragmenting the module's signal.
- `p_spurious` adds background-scale off-module hits to the driver rows, so
  driver traits carry contamination outside their own module.
- `p_structural_zero` removes a fraction of each module's true-support cells
  entirely (`NA`), so a module's signal is only recoverable from the cells that
  are present — real-data dropout inside a module.
- The generative marginals are calibrated to a real pleiotropy matrix: driver
  effects are heavy-tailed across traits (`effect_tail`), background densities
  are heavy-tailed per trait (`background_sparsity_sd`), and background
  magnitudes are scaled relative to the modules by `background_effect_scale`.
  With this difficulty level, recovery is expected to be partial and noisy —
  that is what makes the per-check pass rates informative.
- `ari_all` and `ari_structured` use each SNP's strongest assigned program as a
  hard projection, so they remain conventional partition metrics.
- `factor_strength` and `factor_strength_per_snp` are reported scores only;
  they do not currently filter programs.
- lFSR and magnitude thresholds determine `assigned` SNPs before coherence,
  factor-strength, and posterior-evidence calculations.
- Stability is run only for programs that pass the size and internal-coherence
  checks; the `stability` column shows `n/a` for programs that were not
  re-checked.
- Every aggregate number is a Monte Carlo estimate over `n_sim` replicates and
  is reported with its Monte Carlo SE (ADEMP). Reduce `n_sim` only for test
  runs; the MC SE in the tables will show the resulting loss of precision.
- Replicates use common random numbers: the per-replicate seed (`sim_seeds`) is
  shared across input versions and recorded in `tier_summary$sim_seed`, so the
  study is reproducible and version-to-version differences are more precise
  than independent seeds would give.
- Null simulations (`K = 0`, no modules planted) and the false-discovery rates
  of the validation checks under pure randomness are covered in
  `investigation-univariate-null-simulations.html`.

```{r write-outputs, include=FALSE}
if (isTRUE(params$write_outputs)) {
  out_dir <- if (nzchar(params$output_dir)) params$output_dir else tempdir()
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  utils::write.csv(
    tier_summary, file.path(out_dir, "module-estimates.csv"), row.names = FALSE
  )
  utils::write.csv(
    program_checks, file.path(out_dir, "module-programs.csv"), row.names = FALSE
  )
  utils::write.csv(
    recovery_mc, file.path(out_dir, "module-performance-recovery.csv"),
    row.names = FALSE
  )
  utils::write.csv(
    check_mc, file.path(out_dir, "module-performance-gates.csv"),
    row.names = FALSE
  )
  message("Wrote simulation outputs to ", out_dir)
}
```