---
title: "Real-trait univariate clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  target_trait_id: 1992
  louvain_gamma: 1.5
  similarity_threshold: 0.2
  min_module_size: 5
  n_null: 10
  alpha_membership: 0.05
  ebmf_se_mode: "unit"
  ebmf_beta_scale: "trait"
  ebmf_magnitude_threshold: 0.25
vignette: >
  %\VignetteIndexEntry{Real-trait univariate clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
```

Runs the univariate SNP clustering pipeline on real GPMap data for one target
trait, under two clustering methods: signed Louvain on the SNP similarity graph,
and empirical Bayes matrix factorization (EBMF). Louvain produces a hard
partition; EBMF produces overlapping programs. Results are descriptive
candidates, not null-calibrated discoveries.

```{r fetch}
select_api("production")
obj <- trait(params$target_trait_id, include_associations = TRUE)
cg <- obj$coloc_groups

knitr::kable(
  data.frame(
    item = c(
      "target trait id",
      "association rows",
      "SNPs (coloc groups)",
      "background traits"
    ),
    value = c(
      params$target_trait_id,
      nrow(cg),
      length(unique(cg$variant_id)),
      length(unique(cg$trait_id[cg$trait_id != params$target_trait_id]))
    )
  ),
  caption = "Dataset summary"
)
```

## Signed Louvain clustering

```{r louvain-run}
res_lv <- run_univariate_clustering(
  obj,
  louvain_gamma = params$louvain_gamma,
  similarity_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  compute_specific_traits = TRUE
)
q_lv <- res_lv$module_quality

knitr::kable(
  data.frame(
    parameter = c("Louvain gamma", "similarity threshold", "minimum module size"),
    value = c(params$louvain_gamma, params$similarity_threshold, params$min_module_size)
  ),
  caption = "Clustering settings"
)

knitr::kable(
  data.frame(
    stage = c("Louvain communities", "Reliable modules"),
    n_modules = c(nrow(q_lv), sum(q_lv$reliable))
  ),
  caption = "Module counts before and after reliability filtering"
)
```

```{r louvain-quality}
knitr::kable(
  q_lv |>
    dplyr::filter(reliable) |>
    dplyr::select(
      cluster, n_snps, mean_internal_similarity, mean_external_similarity,
      separation, connectedness
    ) |>
    dplyr::mutate(dplyr::across(
      where(is.numeric), ~ round(.x, 3)
    )),
  row.names = FALSE,
  caption = "Reliable Louvain modules"
)
```

```{r louvain-drivers}
gt <- res_lv$group_traits$by_group
driver_rows <- lapply(gt, function(g) {
  if (!g$group[1] %in% as.character(q_lv$cluster[q_lv$reliable])) {
    return(NULL)
  }
  tr <- g$traits[order(-g$traits$abs_mean_z), ]
  tr <- tr[tr$trait_id != as.character(params$target_trait_id), ]
  if (nrow(tr) == 0) {
    return(NULL)
  }
  data.frame(
    module = g$group[1],
    n_snps = g$n_snps[1],
    rank = seq_len(nrow(tr)),
    trait = paste0(tr$trait_name, " [", tr$trait_category, "]"),
    mean_z = round(tr$mean_z, 1),
    stringsAsFactors = FALSE
  )
})
driver_df <- dplyr::bind_rows(driver_rows)

knitr::kable(
  driver_df |>
    dplyr::filter(rank <= 5) |>
    tidyr::pivot_wider(
      id_cols = c(module, n_snps),
      names_from = rank,
      values_from = c(trait, mean_z),
      names_vary = "slowest"
    ),
  row.names = FALSE,
  caption = "Top background-trait drivers per reliable Louvain module (ranked by |mean z|)"
)
```

## EBMF programs

```{r ebmf-run}
res_eb <- run_univariate_clustering(
  obj,
  cluster_type = "ebmf",
  min_module_size = params$min_module_size,
  ebmf_se_mode = params$ebmf_se_mode,
  ebmf_beta_scale = params$ebmf_beta_scale,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold
)
q_eb <- res_eb$module_quality
mem <- res_eb$cluster_membership
fit <- res_eb$cluster_details$flash_fit
n_factors <- if (!is.null(fit)) as.integer(fit$n_factors) else 0L

knitr::kable(
  data.frame(
    item = c(
      "EBMF factors",
      "global mega-factors dropped",
      "programs",
      "SNPs hard-assigned",
      "SNPs in >1 program (overlap)"
    ),
    value = c(
      n_factors,
      length(res_eb$details$dropped_global_factors),
      ncol(mem),
      length(res_eb$clusters_reliable),
      sum(rowSums(mem) > 1)
    )
  ),
  caption = "EBMF fit summary"
)

if (ncol(mem) > 0) {
  knitr::kable(
    data.frame(
      program = seq_len(ncol(mem)),
      n_snps_member = as.integer(colSums(mem))
    )[order(-colSums(mem)), ],
    row.names = FALSE,
    caption = "Program sizes (SNPs passing lFSR / magnitude thresholds per factor)"
  )
}
```

```{r ebmf-quality}
if (any(q_eb$reliable)) {
  knitr::kable(
    q_eb |>
      dplyr::filter(reliable) |>
      dplyr::select(cluster, n_snps, mean_internal_similarity, separation, connectedness) |>
      dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))),
    row.names = FALSE,
    caption = "Reliable EBMF-derived programs"
  )
} else {
  cat("No EBMF-derived program passed reliability filters.\n")
}
```

```{r ebmf-drivers}
X_eb <- res_eb$x_star[, names(res_eb$clusters), drop = FALSE]
drv <- summarise_snp_modules(
  X_eb,
  res_eb$clusters,
  exclude_trait_id = params$target_trait_id,
  top_n = 4
)
info <- res_eb$trait_info
drv$trait <- info$trait_name[match(drv$trait_id, as.character(info$trait_id))]
drv$trait[is.na(drv$trait)] <- drv$trait_id[is.na(drv$trait)]
drv$n_snps_total <- as.integer(table(drv$cluster)[as.character(drv$cluster)])
drv$reliable <- drv$cluster %in%
  q_eb$cluster[q_eb$reliable]

knitr::kable(
  drv |>
    dplyr::select(cluster, n_snps_total, reliable, trait, mean_z) |>
    dplyr::arrange(cluster, dplyr::desc(abs(mean_z))),
  row.names = FALSE,
  caption = "Top drivers per EBMF program (all programs shown; reliable flag marks programs passing filters)"
)
```

## Permutation-calibrated EBMF reporting

The pipeline's hard gates are bypassed: the EBMF posterior is calibrated
against permutation nulls of the observed matrix (each trait's values shuffled
across SNPs, preserving sparsity and noise), and programs are reported only if
they retain at least `min_module_size` FDR-controlled core members.

```{r ebmf-calibration}
cal <- calibrate_ebmf_programs(
  res_eb,
  n_null = params$n_null,
  verbose = FALSE
)

knitr::kable(
  cal$programs[cal$programs$n_core > 0, ],
  row.names = FALSE,
  caption = sprintf(
    "EBMF programs after permutation calibration (%d nulls; detected = >= %d FDR-controlled members)",
    params$n_null, params$min_module_size
  )
)

core <- cal$memberships[cal$memberships$core & cal$memberships$q <= params$alpha_membership, ]
if (nrow(core) > 0) {
  cls <- setNames(core$program, core$snp_id)
  Xc <- res_eb$x_star[, names(cls), drop = FALSE]
  drv_core <- summarise_snp_modules(
    Xc, cls,
    exclude_trait_id = params$target_trait_id,
    top_n = 5
  )
  info <- res_eb$trait_info
  drv_core$trait <- info$trait_name[
    match(drv_core$trait_id, as.character(info$trait_id))
  ]
  drv_core$trait[is.na(drv_core$trait)] <-
    drv_core$trait_id[is.na(drv_core$trait)]
  knitr::kable(
    drv_core |>
      dplyr::select(cluster, trait, mean_z) |>
      dplyr::arrange(cluster, dplyr::desc(abs(mean_z))),
    row.names = FALSE,
    caption = "Top traits driving each program's FDR-controlled core SNPs"
  )
} else {
  cat("No FDR-controlled core members at alpha =", params$alpha_membership, "\n")
}
```

## Method agreement

Louvain clusters SNP *profiles* on the similarity graph; EBMF finds axes of
trait covariance with overlapping membership. Their hard assignments need not
correspond one-to-one. The table below matches each EBMF program to its best-
overlapping reliable Louvain module (Jaccard index).

```{r agreement}
rel_lv <- res_lv$clusters_reliable
lv_sets <- split(names(rel_lv), rel_lv)
agree_rows <- lapply(sort(unique(res_eb$clusters_reliable)), function(pg) {
  snps_pg <- names(res_eb$clusters_reliable)[res_eb$clusters_reliable == pg]
  jac <- vapply(lv_sets, function(ml) {
    length(intersect(snps_pg, ml)) / length(union(snps_pg, ml))
  }, numeric(1))
  data.frame(
    program = pg,
    n_snps = length(snps_pg),
    best_louvain_module = names(which.max(jac)),
    jaccard = round(max(jac), 3)
  )
})
agree_df <- dplyr::bind_rows(agree_rows)

if (nrow(agree_df) > 0) {
  knitr::kable(agree_df, row.names = FALSE,
               caption = "Best Louvain-module match per EBMF program")
} else {
  cat("No EBMF programs available for comparison.\n")
}
```

Low Jaccard is expected when EBMF programs cut across graph modules; interpret
it jointly with the driver tables rather than as error. For overlap-aware
comparison, use `res$cluster_membership` instead of hard assignments.
