---
title: "Investigation of univariate analysis"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  min_snp_signals: 5
  louvain_gamma: 2
  compress_method: "asinh"
  similarity_threshold: 0.5
  similarity_missing_method: "zero_fill"
  min_pairwise_overlap: 5
  overlap_shrinkage: 5
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
```

Single-trait SNP clustering: each column is a GWAS SNP for the target trait,
each row is a background study (phenotypic or molecular; no collapsing).
Profiles are oriented to the target, compressed, and clustered with **signed
Louvain**. Modules are then filtered by graph reliability
(internal similarity and connectedness) before interpretation.

We use **BMI** (trait id `1992`).

```{r params}
target_trait_id <- 1992
min_module_size <- 3
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_mean_internal <- 0.3
min_connectedness <- 0.5

api <- params$api
min_snp_signals <- as.integer(params$min_snp_signals)
louvain_gamma <- as.numeric(params$louvain_gamma)
compress_method <- params$compress_method
similarity_threshold <- as.numeric(params$similarity_threshold)
similarity_missing_method <- params$similarity_missing_method
min_pairwise_overlap <- as.integer(params$min_pairwise_overlap)
overlap_shrinkage <- as.numeric(params$overlap_shrinkage)

select_api(api)

knitr::kable(
  data.frame(
    api = api,
    gpmap_url = getOption("gpmap_url"),
    min_snp_signals = min_snp_signals,
    louvain_gamma = louvain_gamma,
    compress_method = compress_method,
    similarity_threshold = similarity_threshold,
    similarity_missing_method = similarity_missing_method,
    min_pairwise_overlap = min_pairwise_overlap,
    overlap_shrinkage = overlap_shrinkage,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

traits_meta <- traits(target_trait_id)
target_trait_name <- traits_meta$traits$trait_name[
  traits_meta$traits$id == target_trait_id
]
```

## Step 1: Pleiotropy matrix

`build_pleiotropy_matrix()` uses colocalisation z-scores (`beta / se`) at each
target SNP. Molecular QTL studies stay as their own rows.

```{r step1-build}
coloc_groups <- trait(target_trait_id, include_associations = TRUE)$coloc_groups
pleiotropy <- build_pleiotropy_matrix(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  snp_key = "variant_id"
)
X <- pleiotropy$x_matrix
cg <- coloc_groups

n_pheno <- sum(pleiotropy$trait_info$feature_type == "phenotypic", na.rm = TRUE)
n_mol <- sum(pleiotropy$trait_info$feature_type == "molecular", na.rm = TRUE)

knitr::kable(
  data.frame(
    trait = target_trait_name,
    n_rows = nrow(X),
    n_phenotypic = n_pheno,
    n_molecular = n_mol,
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3)
  ),
  caption = "Pleiotropy matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous traits

Drop rows with fewer than `min_snp_signals` non-`NA` values, or with non-`NA`
fraction above `max_snp_fraction` (target trait kept).

```{r step1-sparse-filter}
trait_snp_counts <- rowSums(!is.na(X))
trait_snp_frac <- trait_snp_counts / ncol(X)
sparse_trait_ids <- names(trait_snp_counts)[trait_snp_counts < min_snp_signals]
ubiquitous_trait_ids <- names(trait_snp_frac)[trait_snp_frac > max_snp_fraction]
drop_trait_ids <- setdiff(
  union(sparse_trait_ids, ubiquitous_trait_ids),
  as.character(target_trait_id)
)

X <- X[!rownames(X) %in% drop_trait_ids, , drop = FALSE]
pleiotropy$trait_info <- pleiotropy$trait_info |>
  dplyr::filter(!as.character(trait_id) %in% drop_trait_ids)

knitr::kable(
  data.frame(
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    n_sparse_traits_removed = length(setdiff(sparse_trait_ids, as.character(target_trait_id))),
    n_ubiquitous_traits_removed = length(setdiff(ubiquitous_trait_ids, as.character(target_trait_id))),
    n_traits_after = nrow(X),
    n_snps = ncol(X)
  ),
  caption = "After sparse / ubiquitous trait filter",
  align = "r"
)
```

## Step 2: Orient and compress (X\*)

```{r step2-orient}
X_star <- orient_pleiotropy_matrix(
  X,
  target_trait_id = target_trait_id
)$x_matrix
X_star <- compress_effect_matrix(X_star, method = compress_method)
trait_matrix <- t(X_star)
```

## Step 3: SNP similarity (S)

`similarity_missing_method = "zero_fill"` treats missing trait-SNP cells as
zero effects, preserving the original cosine calculation.
`"pairwise_complete"` instead calculates each SNP-pair cosine using only traits
observed for both SNPs. Pairs with fewer than `min_pairwise_overlap` shared
traits receive similarity zero. Remaining similarities are multiplied by
\(n_{\mathrm{overlap}} / (n_{\mathrm{overlap}} +
\mathrm{overlap\_shrinkage})\), so extreme similarities based on little shared
evidence are pulled toward zero.

```{r step3-similarity}
similarity <- snp_similarity_matrix(
  X_star,
  missing_method = similarity_missing_method,
  min_overlap = min_pairwise_overlap,
  overlap_shrinkage = overlap_shrinkage
)
S <- similarity$s_matrix
s_upper <- S[upper.tri(S)]
overlap_upper <- similarity$overlap_matrix[
  upper.tri(similarity$overlap_matrix)
]
eligible_upper <- similarity$eligible_matrix[
  upper.tri(similarity$eligible_matrix)
]

knitr::kable(
  data.frame(
    n_snps = ncol(S),
    median_pairwise_overlap = stats::median(overlap_upper),
    fraction_pairs_eligible = round(mean(eligible_upper), 3),
    min_similarity = round(min(s_upper, na.rm = TRUE), 2),
    max_similarity = round(max(s_upper, na.rm = TRUE), 2),
    mean_similarity = round(mean(s_upper, na.rm = TRUE), 2)
  ),
  caption = "SNP similarity summary",
  align = "r"
)
```

## Step 4: Signed Louvain clustering

Edges with `|S| < similarity_threshold` are zeroed before Louvain.

```{r step4-cluster}
clusters_louvain <- cluster_snp_profiles_louvain(
  S,
  similarity_threshold = similarity_threshold,
  gamma = louvain_gamma
)

knitr::kable(
  data.frame(
    cluster = as.integer(names(table(clusters_louvain$cluster))),
    n_snps = as.integer(table(clusters_louvain$cluster)),
    stringsAsFactors = FALSE
  ) |>
    dplyr::filter(n_snps >= min_module_size) |>
    dplyr::arrange(dplyr::desc(n_snps)),
  caption = "Signed Louvain module sizes",
  align = "rr"
)
```

## Step 4.5: Module reliability

`summarise_snp_module_quality()` scores each module on the similarity graph:

- **mean_internal_similarity** — mean off-diagonal `S` inside the module
- **separation** — internal minus external mean similarity
- **connectedness** — fraction of SNP pairs with `|S| >= similarity_threshold`
- **largest_component_frac** — share of SNPs in the largest thresholded component
- **mean_silhouette** — diagnostic only (distance `1 - S`); not used to gate
  reliability, because SNPs can participate in multiple biological programs

A module is `reliable` if it has at least `min_module_size` SNPs, internal
similarity `>= min_mean_internal`, and connectedness `>= min_connectedness`.

```{r step4-quality}
module_quality <- summarise_snp_module_quality(
  S,
  clusters_louvain$cluster,
  edge_threshold = similarity_threshold,
  min_module_size = min_module_size,
  min_mean_internal = min_mean_internal,
  min_connectedness = min_connectedness
)

knitr::kable(
  module_quality |>
    dplyr::filter(n_snps >= min_module_size) |>
    dplyr::mutate(
      dplyr::across(
        c(
          mean_internal_similarity, mean_external_similarity, separation,
          connectedness, largest_component_frac, mean_silhouette
        ),
        ~ round(.x, 3)
      )
    ),
  caption = "Module reliability metrics",
  row.names = FALSE
)

reliable_ids <- module_quality$cluster[module_quality$reliable]
clusters_analysis <- clusters_louvain$cluster[
  clusters_louvain$cluster %in% reliable_ids
]
```

```{r step4-heatmap, fig.height = 7, fig.width = 7}
if (length(clusters_analysis) >= 2) {
  snp_order <- names(clusters_analysis)[order(clusters_analysis)]
  heatmap(
    S[snp_order, snp_order, drop = FALSE],
    Rowv = NA,
    Colv = NA,
    scale = "none",
    main = "SNP similarity (reliable Louvain modules)",
    xlab = "SNPs",
    ylab = "SNPs"
  )
} else {
  cat("Fewer than 2 SNPs in reliable modules; skipping heatmap.\n")
}
```

## Step 5: Interpret reliable modules

For each reliable module, rank background traits by magnitude
(`abs_mean_z` via `summarise_snp_group_traits()`), then summarise
**trait-category and tissue enrichments vs the trait baseline**, plus
pathway hits.

Trait baseline = all SNPs attributed to this trait in the clustering analysis
(`names(clusters_louvain$cluster)`), using coloc rows for those SNPs only—not
the whole GPMap universe. Each module is compared to the **rest of the trait**
(leave-one-out): observed link share vs expected share, with fold enrichment.
Annotation tables show the top elevated categories/tissues even when FDR is not
significant; FDR is reported when it passes 0.05.

```{r step5-interpret}
baseline_snp_ids <- names(clusters_louvain$cluster)

if (length(clusters_analysis) == 0) {
  cat("No reliable modules to interpret.\n")
  group_traits <- NULL
  group_categories <- NULL
  group_tissues <- NULL
  module_enrichment <- NULL
} else {
  group_traits <- summarise_snp_group_traits(
    trait_matrix = trait_matrix,
    groups = clusters_analysis,
    trait_info = pleiotropy$trait_info,
    coloc_groups = cg,
    exclude_trait_ids = target_trait_id,
    min_group_size = min_module_size,
    top_n = top_n_traits,
    n_categories = 3L
  )
  group_categories <- enrich_snp_group_trait_categories(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    baseline_snp_ids = baseline_snp_ids
  )
  group_tissues <- enrich_snp_group_tissues(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    background = "trait",
    baseline_snp_ids = baseline_snp_ids
  )
  module_enrichment <- enrich_snp_group_pathways(
    groups = clusters_analysis,
    coloc_groups = cg,
    min_group_size = min_module_size,
    sources = c("KEGG", "Reactome")
  )
}

.format_enrichment_hits <- function(comparison, label_col, max_n = 3L) {
  if (is.null(comparison) || nrow(comparison) == 0) {
    return("none")
  }
  # Prefer elevated vs rest-of-trait; fall back to most abundant if none.
  elevated <- comparison[
    is.finite(comparison$fold_enrichment) & comparison$fold_enrichment >= 1,
    ,
    drop = FALSE
  ]
  if (nrow(elevated) == 0) {
    elevated <- comparison
  } else {
    elevated <- elevated[order(
      -elevated$fold_enrichment,
      elevated$fdr,
      elevated[[label_col]]
    ), , drop = FALSE]
  }
  top <- utils::head(elevated, max_n)
  labels <- vapply(seq_len(nrow(top)), function(i) {
    fdr_txt <- if (is.finite(top$fdr[i]) && top$fdr[i] <= 0.05) {
      paste0(", FDR=", signif(top$fdr[i], 3))
    } else {
      ""
    }
    paste0(
      top[[label_col]][i],
      " (obs=", sprintf("%.0f%%", 100 * top$frac_group[i]),
      ", exp=", sprintf("%.0f%%", 100 * top$frac_universe[i]),
      ", ratio=", signif(top$fold_enrichment[i], 2),
      fdr_txt,
      ")"
    )
  }, character(1))
  paste(labels, collapse = "<br>")
}

.format_pathway_hits <- function(pathways, max_n = 3L) {
  if (is.null(pathways) || nrow(pathways) == 0) {
    return("none")
  }
  top <- utils::head(pathways, max_n)
  labels <- paste0(top$source, ": ", top$description)
  if ("fdr" %in% names(top)) {
    labels <- paste0(labels, " (FDR=", signif(top$fdr, 3), ")")
  }
  paste(labels, collapse = "<br>")
}
```

```{r step5-module-panels, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  categories_by_group <- stats::setNames(
    group_categories$by_group,
    vapply(
      group_categories$by_group,
      function(x) as.character(x$group),
      character(1)
    )
  )
  tissues_by_group <- stats::setNames(
    group_tissues$by_group,
    vapply(group_tissues$by_group, function(x) as.character(x$group), character(1))
  )
  pathways_by_group <- stats::setNames(
    module_enrichment$by_group,
    vapply(module_enrichment$by_group, function(x) as.character(x$group), character(1))
  )

  for (x in group_traits$by_group) {
    grp <- as.character(x$group)
    cat("\n\n### Module ", grp, " (", x$n_snps, " SNPs)\n\n", sep = "")

    category_x <- categories_by_group[[grp]]
    tissue_x <- tissues_by_group[[grp]]
    pathway_x <- pathways_by_group[[grp]]
    annotation <- data.frame(
      annotation = c("trait_categories", "tissues", "pathways"),
      details = c(
        .format_enrichment_hits(
          if (is.null(category_x)) NULL else category_x$comparison,
          "trait_category"
        ),
        .format_enrichment_hits(
          if (is.null(tissue_x)) NULL else tissue_x$comparison,
          "tissue"
        ),
        .format_pathway_hits(
          if (is.null(pathway_x)) NULL else pathway_x$pathways
        )
      ),
      stringsAsFactors = FALSE
    )
    print(knitr::kable(
      annotation,
      caption = "Module annotations (enrichment vs trait baseline)",
      row.names = FALSE,
      escape = FALSE
    ))

    if (nrow(x$traits) > 0) {
      print(knitr::kable(
        x$traits |>
          dplyr::select(trait_name, trait_category, abs_mean_z, mean_z) |>
          dplyr::mutate(
            abs_mean_z = round(abs_mean_z, 3),
            mean_z = round(mean_z, 3)
          ),
        caption = "Top traits by magnitude",
        row.names = FALSE
      ))
    } else {
      cat("No traits with sufficient signal.\n\n")
    }
  }
}
```
