---
title: "Univariate SNP biological clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Univariate SNP biological clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take a **single trait** and decompose its genetic architecture by grouping SNPs
according to their genome-wide pleiotropic profiles.

The central question is: **within one trait, what distinct biological programs
are represented by its associated SNPs?**

The output is a set of SNP programs (latent factors), each annotated with
background traits, QTL studies, tissues, genes, and pathways. The runnable
implementation is
[Investigation of univariate analysis](investigation-univariate-analysis.html).
[Investigation of EBMF analysis](investigation-ebmf-analysis.html) explores the
EBMF factorisation itself (priors, variance structure, PVE).

Background traits stay at **study level**. Molecular QTL studies are not
collapsed by gene, tissue, cis/trans, or QTL type.

### Core clustering preferences

The grouping method should satisfy:

1. **Handles high sparsity** — Must natively process large, sparse matrices
   (thousands of singleton feature rows) without treating structural zeros as
   low-variance signal. EBMF does this by leaving missing cells out of the fit
   (`flashier` skips `NA`); we additionally limit sparsity with a hard threshold
   on the number of non-missing SNPs per trait (`min_snp_signals`).
2. **Signed and sign-driven clustering** — Must preserve the meaning of
   continuous positive/negative z-scores so variants group by **directional
   alignment**, not magnitude alone. We orient every SNP column by
   \(\mathrm{sign}(z_{\mathrm{target}})\) before factorization.
3. **Automated \(K\) discovery** — The engine should learn the number of
   clusters or latent factors from the data. EBMF adds factors greedily while
   the variational lower bound (ELBO) improves, then backfits; filtering the
   resulting factors by an lFSR/magnitude membership gate and by the
   validation checks below controls the final program count.
4. **Overlapping membership** — A SNP can belong to more than one biological
   program. EBMF produces soft factor loadings, so membership is per-SNP-per-
   program rather than a single hard label.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Pleiotropy matrix\\n(study-level rows x SNPs)\"]
  S2 [label=\"2. Orient by sign(z_target)\\noptional asinh compression\"]
  S3 [label=\"3. EBMF factorization Y = L F^T + E\\nlFSR + magnitude program gates\"]
  S4 [label=\"4. Validate programs\\ncoherence on SNP-SNP S, stability, factor strength\"]
  S5 [label=\"5. Interpret valid programs\\ntraits, tissues, pathways\"]
  S1 -> S2 -> S3 -> S4 -> S5
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

### 1. Construct the pleiotropy matrix

Given:

- `nsnp` = independent SNPs associated with the target trait (coloc groups)
- `ntrait` = background studies in GPMap (phenotypic and molecular)

Construct a **traits x SNPs** matrix of signed z-scores (\(z = \beta / \mathrm{SE}\)).
Conceptually this is a SNP-by-trait profile matrix; the code stores traits as
rows and SNPs as columns.

```text
              SNP1   SNP2   SNP3   SNP4   ...
Trait A        z      z      z      z
Trait B        z      z      z      z
Trait C        z      z      z      z
...
```

Which rows and columns back this matrix is controlled by three choices:

- **`include_trans`** — whether trans-QTL association rows
  (`cis_trans == "trans"`) are removed before the matrix is built. Excluding
  them (default) keeps the analysis cis-focused; including them can surface
  distributed regulatory programs at the cost of density and noise.
- **`trait_subset`** — `"all"` keeps every available background trait;
  `"phenotypic"` keeps only phenotype traits (rows whose `feature_type` is
  `"phenotypic"` or missing) plus the target trait, dropping molecular /
  gene-annotated rows. Molecular traits are much sparser, so clustering on
  phenotype traits only can recover structure that sparse molecular rows
  dilute.
- **`min_snp_signals`** — sparse rows observed in fewer than this many SNPs
  are dropped; ubiquitous rows (non-missing on more than `max_snp_fraction` of
  SNPs) are dropped too. The target trait row is always kept.
- **`compress_method`** — optional `asinh` compression of z-scores, which
  limits the leverage of extreme z-scores without a hard clip. `"none"` preserves
  effect scale but lets outliers dominate direction and norm.

### 2. Orient SNP profiles relative to the target trait

Remove arbitrary allele coding so every column represents increasing target
liability. For SNP \(j\):

\[
X^*_{tj} = \mathrm{sign}(z_j)\, X_{tj}
\]

Optional \(\mathrm{asinh}(z)\) compression (`compress_method = "asinh"`, with
`compress_scale`) then limits the leverage of extreme z-scores without a hard
clip. `"none"` preserves effect scale but lets outliers dominate direction and
norm.

### 3. Fit the EBMF latent-program model

Treat the oriented (and optionally compressed) traits x SNPs matrix as noisy
observations of latent programs:

\[
Y = LF^{\top} + E
\]

- \(Y\) = features x SNPs matrix (traits and molecular QTL studies as rows)
- \(L\) = feature loadings (which traits drive each program)
- \(F\) = SNP factors (which SNPs each program loads on)
- \(E\) = residual error

Implemented via `flashier`: factors are added greedily while the variational
lower bound improves (`greedy_Kmax`), then cyclically backfit. The number of
programs is learned, not fixed. The sparse prior on loadings and factors is
`ebmf_prior`:

- `"point_normal"` (default) — a sparse, spike-and-slab style prior; good
  general default.
- `"point_laplace"` — heavier tails; often better when true effects are large
  relative to noise.

**Extracting programs.** A SNP belongs to a program when its posterior factor
loading is both *significant* and *non-negligible*:

- `lfsr` < `ebmf_lfsr_threshold` (local false sign rate; defaults to `0.05`)
- `|loading|` > `ebmf_magnitude_threshold` (absolute posterior mean of `F_pm`;
  defaults to `0.25`)

Because EBMF is soft, one SNP can pass the gate for several programs
(overlapping membership). Before extraction, at most one *global* mega-factor
— a factor that soft-assigns nearly every SNP or explains a large share of PVE
— is dropped (`ebmf_drop_global = TRUE` via `identify_ebmf_global_factors()`);
this guards against the "one giant cluster" failure mode. Nothing assumes
factor 1 is global.

Why EBMF over community detection (Louvain / Leiden):

- SNPs can belong to multiple programs; biology is overlapping, not mutually
  exclusive.
- \(K\) is inferred by the fit (ELBO) rather than by a graph-resolution or
  edge-threshold parameter.
- The input is the dense oriented z-score matrix itself, so no information is
  lost to a pairwise similarity threshold.
- Missing data are handled probabilistically instead of being zero-filled into
  similarity edges.

The cost is that factors are not guaranteed to be disjoint SNP modules and can
capture broad technical/global effects — which is exactly why the validation
layer in Step 4 exists.

### 4. Validating programs

`run_univariate_clustering()` returns the fit and soft memberships;
`summarise_ebmf_programs()` folds everything into a per-program table with
pass/fail flags and an overall `status`. The checks split into those computed
**on the SNP x SNP similarity matrix** `S` (internal coherence) and those that
use **other means** (the EBMF fit itself and refits under trait subsampling).
A program is **`valid`** only if it passes all **filter gates**; some metrics
are reported descriptively and never gate `status`.

#### The SNP x SNP similarity matrix

EBMF never sees pairwise SNP similarity — it factorises the traits x SNPs
matrix directly — but the same oriented (and optionally compressed) SNP
profiles are used to build the SNP x SNP cosine similarity matrix `S`, which
backs the coherence checks below. For SNPs \(j\) and \(k\):

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\|\,\|x^*_k\|}
\]

- \(+1\) = concordant (parallel) profiles
- \(0\) = unrelated
- \(-1\) = discordant (anti-parallel) profiles

#### Checks on the similarity matrix (internal coherence)

Computed on `S`, restricted to the program's gated SNPs:

- **mean_internal_similarity** — mean off-diagonal `S` among member SNPs
  (diagonal excluded). High values mean oriented profiles are roughly
  parallel. Must be `>= min_mean_internal` (default `0.3`), a weak
  "same direction, not noise" bar.
- **connectedness** — the fraction of member-SNP pairs with
  \(|S| \ge\) `similarity_threshold` (the same `edge_threshold` used in the
  old graph clustering). Mean internal similarity can look decent if a few
  pairs are very similar and the rest are near zero; connectedness
  `>= min_connectedness` (default `0.5`) requires at least half of all pairs
  to be strong edges, so the program is a real connected blob on `S`, not
  one tight pair inside a diffuse background.
- **Separation** (internal minus external similarity) is also reported —
  diagnostic only, not a gate: a program can be internally coherent yet sit
  inside a similar background.

#### Checks by other means

These come from the EBMF fit itself, not from `S`:

- **Size** — at least `min_module_size` SNPs survive the lFSR/magnitude gate
  (`n_snps_filtered`, counted from the EBMF membership). This ensures the
  coherence metrics are computed on enough SNPs to be meaningful.
- **Stability** — the program's top `top_n` loading SNPs must be recovered
  with replication `>= stability_threshold` (default `0.7`) when `1 -
  frac_traits` of the trait rows are held out and flashier is refit
  (`n_rep` replicates). Because this refits EBMF, it is run only for programs
  that already pass the size and internal-coherence gates. Descriptive, but
  gated: a program that does not survive trait subsampling is rejected.

**Descriptive scores** (reported for context, but **not** used to gate
`status`):

- **factor_strength** — the proportion of the observed matrix signal
  attributable to the factor, computed from its fitted reconstruction on the
  gated (assigned) SNPs:
  \(\sum_i |L_{ik}| \sum_j |F_{jk}|\) over the total observed signal
  \(\sum |X|\). `factor_strength_per_snp = factor_strength / sqrt(n_snps_filtered)`
  normalises it by the square root of program size. These prioritise programs
  that carry real matrix signal rather than marginal memberships.
- **posterior_evidence** — aggregate (`median` or `mean`) of `-log10(lfsr)`
  across the program's assigned SNPs (capped so lFSR underflow does not blow
  up). Distinguishes programs dominated by borderline `lfsr ~ 0.05`
  memberships from programs whose memberships carry much stronger posterior
  sign evidence.
- **max_abs_factor_corr / redundant** — the maximum absolute correlation of the
  program's SNP loading vector with any other program; programs exceeding
  `redundancy_threshold` are flagged `redundant` (near-duplicate factors).

`status` is `"valid"` when all gates pass, otherwise the failed gate(s) are
listed (`size; internal; stability`). Downstream interpretation uses only the
gated memberships of `valid` programs.

**Permutation calibration (reporting layer).** `calibrate_ebmf_programs()`
additionally shuffles each trait row independently (`n_null` replicates,
preserving each trait's sparsity and effect distribution), refits the pipeline,
and pools the null loading distribution. Each SNP x program loading gets an
empirical p-value and BH q-value against this null; cells with `q <=
alpha_membership` are labelled `core`, and the top remaining loadings per
program form an explicitly uncertified `candidate` tier. This is a membership
labelling layer, not a program-level validity gate.

### 5. Interpret valid programs

The univariate vignette interprets **valid** EBMF programs as:

- **Top traits by magnitude** (`abs_mean_z` within the program).
- **Trait-category composition** (`enrich_snp_group_trait_categories()`):
  observed vs expected share against the rest of the trait SNP baseline
  (coloc rows for SNPs attributed to the analysed trait, not the whole GPMap
  universe). Top elevated categories are shown with obs/exp/FE; FDR is noted
  when significant.
- **Tissue composition** (`enrich_snp_group_tissues()`, `background = "trait"`)
  on the same leave-one-out trait baseline.
- **Pathway enrichment** on QTL genes (KEGG and Reactome), treated as
  secondary because the full-trait gene set is often empty for broad traits
  such as BMI.
- **SNP pleiotropy typology** (`summarise_snp_group_pleiotropy()`): programs
  are placed on a *breadth* axis (mean per-SNP distinct trait categories /
  distinct genes) against *tightness* (separation). Descriptive only — it
  classifies programs as specific programs (tight + narrow), regulatory hubs
  (tight + broad), or generic background (loose + broad), without acting as a
  filter or reliability criterion.

#### Multiple testing correction for trait-category and tissue enrichments

Each valid program is tested against **every** trait category and **every**
tissue for over-representation relative to the trait baseline (leave-one-out
hypergeometric test). A
single program therefore produces dozens of hypothesis tests — one per category
or tissue — and across all programs in a run the pipeline performs hundreds. If
we declared everything with raw p < 0.05 significant, we would expect ~5% of
purely null categories/tissues to pass by chance; a program testing ~50
categories would accumulate ~2-3 false hits before any real biology.

We therefore apply **Benjamini–Hochberg (BH) FDR
correction** *within each program*, across the categories (or tissues) tested
for that program. BH sorts the raw p-values, then finds the largest prefix that
can be declared significant while keeping the expected fraction of false
positives among the significant calls at or below the chosen level (0.05). The
reported `FDR` for a category/tissue is the smallest FDR level at which that
row would still be called significant (a "q-value"): a category with
`FDR = 0.03` means that if we declare it significant, we expect ≤3% of the
significant calls for that program to be false positives.

Note: **Correction is per-program, not global.** Each program's FDR is computed
over the categories/tissues tested for that program alone. Running many
programs still multiplies the opportunities for false positives across the whole
analysis, which is why the vignette insists on replication across runs before
treating a hit as real.

## Variables to consider

Exposed in the univariate vignette today (EBMF path):

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `Dataset` | `production`, `local` | Public GPMap vs new trans-QTL DB | Local data added trans-QTL studies, increasing matrix density and changing the regulatory signals that connect SNPs. This can alter both program membership and tissue/gene interpretation. Production is more reproducible but may omit those signals. |
| `include_trans` | `TRUE`, `FALSE` | Whether trans-QTL rows are retained | Including trans effects can reveal distributed regulatory programs and improve tissue/pathway interpretation, while increasing density and possible noise. Excluding them gives a more local cis-focused analysis but may miss system-level mechanisms. |
| `trait_subset` | `all`, `phenotypic` | Which trait rows back the matrix | `phenotypic` drops sparse molecular / gene-annotated rows (keeping the target trait), recovering structure that sparse molecular rows dilute but losing QTL-level signal. `all` keeps every available background trait at the cost of extra sparsity. |
| `min_snp_signals` | `1`, `2`, `3`, `5` | Minimum SNPs with a value for a background trait | Raising it removes singleton or very sparse rows, reducing unstable factors and computation. It can also remove rare, module-specific biology. Lowering it preserves those signals but makes factorization more sensitive to sparse observations. |
| `compress_method` | `asinh`, `none` | Transformation applied to z-scores | `asinh` is nearly linear around zero but compresses extreme values, so factors reflect patterns across several traits rather than one very large z-score. `none` preserves effect scale but allows outliers to dominate direction and norm. |
| `similarity_threshold` | `0`, `0.2`, `0.5` | Absolute \(\|S\|\) edge cutoff for the connectedness gate | Not an EBMF input, but used by the validation layer: `connectedness` counts SNP pairs with \(\|S\| \ge\) this threshold. Raising it makes the connectedness gate stricter and the coherence metrics more conservative. |
| `ebmf_lfsr_threshold` | `0.05`, `0.02`, `0.1` | Local false sign rate cutoff for SNP membership | Lowering it keeps only SNPs whose program loading has very strong posterior sign evidence, shrinking and sharpening programs. Raising it adds borderline `lfsr ~ threshold` memberships. |
| `ebmf_magnitude_threshold` | `0.25`, `0.5`, `0.75`, `NA` | Minimum absolute posterior loading for SNP membership | Raising it keeps only SNPs with substantial program loadings. `NA` gates membership on lFSR alone (useful when loadings are on raw-beta scale). |
| `ebmf_prior` | `point_normal`, `point_laplace` | Prior family for EBMF loadings/factors | `point_normal` is the general default; `point_laplace` has heavier tails and is often better when true effects are large relative to noise. |
| `min_module_size` | `3`, `5` | Minimum gated SNPs for the size gate | Ensures the coherence and stability metrics are computed on enough SNPs to be meaningful; also the effective minimum program size. |
| `min_mean_internal` | `0.3` | Minimum mean internal similarity | The "same direction, not noise" bar on `S` among gated member SNPs. |
| `min_connectedness` | `0.5` | Minimum pair-connectedness | Requires at least this fraction of member-SNP pairs to exceed `similarity_threshold`, so a program is a connected blob rather than one tight pair in a diffuse background. |
| `stability_threshold` | `0.5`, `0.7` | Minimum replication for stability | Fraction of a program's top-loading SNPs that must reappear when traits are held out and EBMF is refit. |

Reliability/validation cutoffs used after clustering therefore are:
`min_module_size`, `min_mean_internal = 0.3`, `min_connectedness = 0.5`,
`stability_threshold`, plus the membership gates `ebmf_lfsr_threshold` and
`ebmf_magnitude_threshold`. `factor_strength`, `posterior_evidence`, and
`max_abs_factor_corr`/`redundant` are reported descriptively but do not gate
`status`.

## What we simulate

Parameter choices are validated on synthetic data with **known ground truth**
rather than on a single real-trait sweep (see
[Simulation study of EBMF programs](investigation-univariate-simulations.html)).
`simulate_trait()` generates a trait object of the same shape as
`trait(id, include_associations = TRUE)` and plants a latent module structure
over SNPs:

- **Planted modules** — `K` modules occupy SNP sets (disjoint by default, or
  `nested` where a parent module contains the others as correlated subsets).
  Each module has `n_traits_per_module` driver traits with `effect_size`-scale
  effects on its module's SNPs, under a `sign_pattern` (`coherent`, `flipped`,
  or `random` signs).
- **Trait-space overlap** — a fraction `trait_overlap` of each module's driver
  traits is shared with the other modules; a shared trait's effects are
  correlated across the modules it drives (`trait_effect_corr`) but not
  identical. Increasing `trait_overlap` should make EBMF programs increasingly
  overlap and the planted modules harder to tell apart — the vignette runs
  versions from `0` (clean) to `1` (full overlap), plus a **null** version
  (`K = 0`, no modules) that checks for false structure.
- **Realistic noise and sparsity** — structural zeros inside true support
  (`p_structural_zero`), spurious off-support noise (`p_spurious`), and sparse
  unstructured background traits (`p_active_background`). Optional per-cell SE
  heterogeneity (`log_se_sd`) and within-module SNP sub-patterns
  (`snp_driver_groups`) break the exchangeability that would make modules
  trivially separable.

The simulation is scored against ground truth by
`evaluate_univariate_simulation()`:

- **module_recall** — for each planted module, the largest fraction of its SNPs
  contained in any single predicted program (fair to overlapping/merged
  programs).
- **coverage** — fraction of all SNPs placed into valid programs.
- **background_absorbed** — fraction of unstructured background SNPs absorbed
  into valid programs (higher = more hallucinated structure).
- **ARI** (`ari_all`, `ari_structured`) — adjusted Rand index against the
  planted partition, using each SNP's strongest program as a hard projection.

These improvements over the earlier grid sweep: the simulation now plants
overlapping modules with correlated shared-trait effects, per-cell noise, and
SNP sub-structure; the validation battery is
applied and scored per program; and the null condition directly tests the
false-discovery behaviour of the EBMF + validation pipeline.

## Structure of Results

The aim is not produce a definitive biological truth for every SNP.  Indead, the method produces a structured, evidence-based decomposition of a trait's genetic architecture, with results that can be assessed at several levels:

For practical use, the framework should provide results at the overall, program, and SNP level.

| Level       | Primary question                                                                         | Main output                                                             |
| ----------- | ---------------------------------------------------------------------------------------- | ----------------------------------------------------------------------- |
| **Trait**   | How much of this trait's genetic architecture can be decomposed into distinct structure? | Number of programs, SNP coverage, overlap, overall signal captured      |
| **Program** | Is this structure reliable, and what biology does it represent?                          | Reliability metrics plus trait, tissue, gene and pathway interpretation |
| **SNP**     | Which program or programs does this variant contribute to?                               | Soft memberships, posterior evidence, rank and molecular annotations    |

This standard report would be returned by gpmapr and as a framework for the Results section of the paper. It should keep the distinction clear between statistical evidence, which establishes whether a program is reliable, and biological annotation, which provides its interpretation.

#### Overall trait-level metrics
- Number of SNPs analysed
- Number of validated programs discovered (maybe also report candidate programs that fail validation)
- Proportion of SNPs assigned to at least one validated program
- Proportion of SNPs unassigned
- Degree of overlapping membership, for example the proportion of assigned SNPs belonging to more than one program
- Overall matrix signal explained by validated programs, if this can be meaningfully derived from the EBMF fit
- Number/proportion of programs with some evidence of biological interpretation

#### Program Level metrics
- Number of member SNPs (maybe also report number of SNPs that were filtered out by the lFSR/magnitude gate)
- Membership overlap with other programs
- Top SNPs by posterior loading or membership evidence
- Various statistical reliability metrics, including mean internal similarity, connectedness, stability, factor strength, stability 
- Biological interpretation metrics, including top traits, trait categories, tissues, genes, pathways, and pleiotropy typology

#### SNP Level metrics
- Program membership(s)
- Posterior loading in each program
- Membership lFSR
- Magnitude of membership/loading
- Core versus candidate membership status
- Rank within each program, if useful
- Whether the SNP belongs to multiple programs
- Similarity to other SNPs within its assigned program
- Variant-level pleiotropic profile, for example breadth of associated trait categories
- Molecular annotations, including associated genes and tissues where available