---
title: "Investigation of univariate analysis"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  min_snp_signals: 5
  louvain_gamma: 2
  compress_method: "none"
  compress_scale: 2
  include_trans: false
  include_situated_gene: false 
  similarity_threshold: 0
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

Single-trait SNP clustering: each column is a GWAS SNP for the target trait,
each row is a background study (phenotypic or molecular; no collapsing).
Profiles are oriented to the target, compressed, and clustered with **signed
Louvain**. Modules are then filtered by graph reliability
(internal similarity and connectedness) before interpretation.

We use **BMI** (trait id `1992`).

```{r params}
target_trait_id <- 1992
min_module_size <- 3
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_specificity <- 1.25
min_mean_internal <- 0.3
min_connectedness <- 0.5

api <- params$api
min_snp_signals <- as.integer(params$min_snp_signals)
louvain_gamma <- as.numeric(params$louvain_gamma)
compress_method <- params$compress_method
compress_scale <- as.numeric(params$compress_scale)
include_trans <- isTRUE(params$include_trans)
include_situated_gene <- isTRUE(params$include_situated_gene)
similarity_threshold <- as.numeric(params$similarity_threshold)
associations <- params$associations

select_api(api)

knitr::kable(
  data.frame(
    api = api,
    gpmap_url = getOption("gpmap_url"),
    min_snp_signals = min_snp_signals,
    louvain_gamma = louvain_gamma,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    include_situated_gene = include_situated_gene,
    min_specificity = min_specificity,
    similarity_threshold = similarity_threshold,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

traits_meta <- traits(target_trait_id)
target_trait_name <- traits_meta$traits$trait_name[
  traits_meta$traits$id == target_trait_id
]
```

## Step 1: Pleiotropy matrix

`build_pleiotropy_matrix()` builds z-scores (`beta / se`) at each target SNP.
With `associations = "coloc"` it uses colocalisation rows only; with
`associations = "full"` it uses all full associations
(`/associations-full`). Molecular QTL studies stay as their own rows.
When `include_trans = FALSE`, trans target markers are removed from the
association rows before the matrix is built.

```{r step1-build}
coloc_groups <- trait(target_trait_id, include_associations = TRUE)$coloc_groups
if (!include_trans && "cis_trans" %in% names(coloc_groups)) {
  coloc_groups <- coloc_groups |>
    dplyr::filter(is.na(cis_trans) | tolower(cis_trans) != "trans")
}
pleiotropy <- build_pleiotropy_matrix(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  snp_key = "variant_id",
  associations = associations
)
X <- pleiotropy$x_matrix
cg <- coloc_groups

n_pheno <- sum(pleiotropy$trait_info$feature_type == "phenotypic", na.rm = TRUE)
n_mol <- sum(pleiotropy$trait_info$feature_type == "molecular", na.rm = TRUE)

knitr::kable(
  data.frame(
    trait = target_trait_name,
    n_rows = nrow(X),
    n_phenotypic = n_pheno,
    n_molecular = n_mol,
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3)
  ),
  caption = "Pleiotropy matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous traits

Drop rows with fewer than `min_snp_signals` non-`NA` values, or with non-`NA`
fraction above `max_snp_fraction` (target trait kept).

```{r step1-sparse-filter}
trait_snp_counts <- rowSums(!is.na(X))
trait_snp_frac <- trait_snp_counts / ncol(X)
sparse_trait_ids <- names(trait_snp_counts)[trait_snp_counts < min_snp_signals]
ubiquitous_trait_ids <- names(trait_snp_frac)[trait_snp_frac > max_snp_fraction]
drop_trait_ids <- setdiff(
  union(sparse_trait_ids, ubiquitous_trait_ids),
  as.character(target_trait_id)
)

X <- X[!rownames(X) %in% drop_trait_ids, , drop = FALSE]
pleiotropy$trait_info <- pleiotropy$trait_info |>
  dplyr::filter(!as.character(trait_id) %in% drop_trait_ids)

knitr::kable(
  data.frame(
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    n_sparse_traits_removed = length(setdiff(sparse_trait_ids, as.character(target_trait_id))),
    n_ubiquitous_traits_removed = length(setdiff(ubiquitous_trait_ids, as.character(target_trait_id))),
    n_traits_after = nrow(X),
    n_snps = ncol(X)
  ),
  caption = "After sparse / ubiquitous trait filter",
  align = "r"
)
```

## Step 2: Orient and compress (X\*)

```{r step2-orient}
X_star <- orient_pleiotropy_matrix(
  X,
  target_trait_id = target_trait_id
)$x_matrix
X_star <- compress_effect_matrix(
  X_star,
  method = compress_method,
  asinh_scale = compress_scale
)
trait_matrix <- t(X_star)
```

## Step 3: SNP similarity (S)

Missing trait-SNP cells are treated as zero effects, preserving the original
cosine similarity calculation.

```{r step3-similarity}
similarity <- snp_similarity_matrix(X_star)
S <- similarity$s_matrix
s_upper <- S[upper.tri(S)]
overlap_upper <- similarity$overlap_matrix[
  upper.tri(similarity$overlap_matrix)
]
eligible_upper <- similarity$eligible_matrix[
  upper.tri(similarity$eligible_matrix)
]

knitr::kable(
  data.frame(
    n_snps = ncol(S),
    median_pairwise_overlap = stats::median(overlap_upper),
    fraction_pairs_eligible = round(mean(eligible_upper), 3),
    min_similarity = round(min(s_upper, na.rm = TRUE), 2),
    max_similarity = round(max(s_upper, na.rm = TRUE), 2),
    mean_similarity = round(mean(s_upper, na.rm = TRUE), 2)
  ),
  caption = "SNP similarity summary",
  align = "r"
)
```

## Step 4: Signed Louvain clustering

Edges with `|S| < similarity_threshold` are zeroed before Louvain.

```{r step4-cluster}
clusters_louvain <- cluster_snp_profiles_louvain(
  S,
  similarity_threshold = similarity_threshold,
  gamma = louvain_gamma
)
```

## Step 4.5: Signed Louvain clustering and Module reliability

`summarise_snp_module_quality()` scores each module on the similarity graph:

- **mean_internal_similarity** — mean off-diagonal `S` inside the module
- **separation** — internal minus external mean similarity
- **connectedness** — fraction of SNP pairs with `|S| >= similarity_threshold`

A module is `reliable` if it has at least `min_module_size` SNPs, internal
similarity `>= min_mean_internal`, and connectedness `>= min_connectedness`.

```{r step4-quality}
module_quality <- summarise_snp_module_quality(
  S,
  clusters_louvain$cluster,
  edge_threshold = similarity_threshold,
  min_module_size = min_module_size,
  min_mean_internal = min_mean_internal,
  min_connectedness = min_connectedness
)

knitr::kable(
  module_quality |>
    dplyr::select(-largest_component_frac, -n_components, -mean_silhouette) |>
    dplyr::filter(n_snps >= min_module_size) |>
    dplyr::mutate(
      dplyr::across(
        c(mean_internal_similarity, mean_external_similarity, separation, connectedness),
        ~ round(.x, 3)
      )
    ),
  caption = "Module reliability metrics",
  row.names = FALSE
)

reliable_ids <- module_quality$cluster[module_quality$reliable]
clusters_analysis <- clusters_louvain$cluster[
  clusters_louvain$cluster %in% reliable_ids
]
```

```{r step4-heatmap, fig.height = 7, fig.width = 7}
if (length(clusters_analysis) >= 2) {
  snp_order <- names(clusters_analysis)[order(clusters_analysis)]
  heatmap(
    S[snp_order, snp_order, drop = FALSE],
    Rowv = NA,
    Colv = NA,
    scale = "none",
    main = "SNP similarity (reliable Louvain modules)",
    xlab = "SNPs",
    ylab = "SNPs"
  )
} else {
  cat("Fewer than 2 SNPs in reliable modules; skipping heatmap.\n")
}
```

## Step 5: Interpret reliable modules

For each reliable module, rank background traits by **module specificity vs the
rest of the trait** (`summarise_module_specific_traits()`), then summarise
**trait-category and tissue enrichments vs the trait baseline**, plus
pathway hits. `specificity = mean |z| in the module / (mean |z| in the rest +
0.1)` and `abs_mean_z_delta = module - rest` highlight traits that are
specifically elevated in this module rather than merely large everywhere (which
ranking by `abs_mean_z` alone would over-emphasise). Pathway enrichment is
queried across KEGG, Reactome, and Human Phenotype Ontology (HP): KEGG/Reactome
hits are reported as **pathways** and HP terms as **phenotypes**, so the two
never compete for the same headline.

Trait baseline = all SNPs attributed to this trait in the clustering analysis
(`names(clusters_louvain$cluster)`), using coloc rows for those SNPs only—not
the whole GPMap universe. Each module is compared to the **rest of the trait**
(leave-one-out): observed link share vs expected share, with fold enrichment.

Annotation tables show the top elevated categories/tissues even when FDR is not
significant; FDR is reported when it passes 0.05. Hits that are significant by
any of the three methods at 0.05 are shown in **bold**.

Each hit is scored with three BH-corrected significance summaries, all
thresholded at 0.05:

- **FDR** — hypergeometric test of the module vs the rest of the trait.
- **perm** — permutation p-values from shuffling link labels across the
  baseline, BH-corrected within the module (`fdr_perm`).
- **vsmod** — one-sided Fisher test of the module against the pooled links of
  the *other* modules, BH-corrected within the module (`fdr_modules`).

They agree on strong signal and differ only on marginal, low-count calls. The
bootstrap `CI` for the fold enrichment shows how precisely the ratio is
estimated; because modules are small, a big ratio built on a few links carries
a wide interval that includes 1 (the null), so read the CI together with the
FDR rather than alone.

```{r step5-interpret}
baseline_snp_ids <- names(clusters_louvain$cluster)

if (length(clusters_analysis) == 0) {
  cat("No reliable modules to interpret.\n")
  group_traits <- NULL
  group_categories <- NULL
  group_tissues <- NULL
  module_enrichment <- NULL
} else {
    group_traits <- summarise_module_specific_traits(
    trait_matrix = trait_matrix,
    groups = clusters_analysis,
    trait_info = pleiotropy$trait_info,
    coloc_groups = cg,
    exclude_trait_ids = target_trait_id,
    min_group_size = min_module_size,
      top_n = top_n_traits,
      n_categories = 3L,
      min_specificity = min_specificity
  )
  group_categories <- enrich_snp_group_trait_categories(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    baseline_snp_ids = baseline_snp_ids
  )
  group_tissues <- enrich_snp_group_tissues(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    background = "trait",
    baseline_snp_ids = baseline_snp_ids
  )
  module_enrichment <- enrich_snp_group_pathways(
    groups = clusters_analysis,
    coloc_groups = cg,
    min_group_size = min_module_size,
    sources = c("KEGG", "Reactome", "HP"),
    include_situated_gene = include_situated_gene
  )
}

.format_enrichment_hits <- function(comparison, label_col, max_n = 3L) {
  if (is.null(comparison) || nrow(comparison) == 0) {
    return("none")
  }
  # Prefer elevated vs rest-of-trait; fall back to most abundant if none.
  elevated <- comparison[
    is.finite(comparison$fold_enrichment) & comparison$fold_enrichment >= 1,
    ,
    drop = FALSE
  ]
  if (nrow(elevated) == 0) {
    elevated <- comparison
  } else {
    elevated <- elevated[order(
      -elevated$fold_enrichment,
      elevated$fdr,
      elevated[[label_col]]
    ), , drop = FALSE]
  }
  top <- utils::head(elevated, max_n)
  labels <- vapply(seq_len(nrow(top)), function(i) {
    tag <- .method_tag(top$fdr[i], top$fdr_perm[i], top$fdr_modules[i])
    ci <- if (is.finite(top$fe_ci_lower[i]) && is.finite(top$fe_ci_upper[i])) {
      paste0(
        ", CI=[", signif(top$fe_ci_lower[i], 2),
        ",", signif(top$fe_ci_upper[i], 2), "]"
      )
    } else {
      ""
    }
    fdr_txt <- if (nzchar(tag)) paste0(", ", tag) else ""
    label <- top[[label_col]][i]
    if (nzchar(tag)) {
      label <- paste0("<b>", label, "</b>")
    }
    paste0(
      label,
      " (obs=", sprintf("%.0f%%", 100 * top$frac_group[i]),
      ", exp=", sprintf("%.0f%%", 100 * top$frac_universe[i]),
      ", FE=", signif(top$fold_enrichment[i], 2),
      ci,
      fdr_txt,
      ")"
    )
  }, character(1))
  paste(labels, collapse = "<br>")
}

.format_pathway_hits <- function(pathways, max_n = 3L) {
  if (is.null(pathways) || nrow(pathways) == 0) {
    return("none")
  }
  top <- utils::head(pathways, max_n)
  labels <- paste0(top$source, ": ", top$description)
  if ("fdr" %in% names(top)) {
    labels <- paste0(labels, " (FDR=", signif(top$fdr, 3), ")")
  }
  paste(labels, collapse = "<br>")
}
```

```{r step5-module-overview, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  specific_traits <- dplyr::bind_rows(lapply(group_traits$by_group, function(x) {
    passing <- x$traits |>
      dplyr::filter(passes_min_specificity) |>
      utils::head(3L)
    data.frame(
      group = x$group,
      module_specific_traits = if (nrow(passing) == 0) {
        "none"
      } else {
        paste0(
          passing$trait_name,
          " (", round(passing$specificity, 2), ")",
          collapse = "<br>"
        )
      },
      stringsAsFactors = FALSE
    )
  }))
  overview <- dplyr::full_join(
    group_traits$summary |>
      dplyr::select(group, n_snps),
    group_categories$summary |>
      dplyr::select(group, top_enriched_category_any),
    by = "group"
  ) |>
    dplyr::full_join(
    group_tissues$summary |>
      dplyr::select(group, top_enriched_tissue_any),
    by = "group"
  ) |>
    dplyr::full_join(
      module_enrichment$summary |>
        dplyr::select(group, top_enriched_pathway),
      by = "group"
    ) |>
    dplyr::full_join(
      module_enrichment$summary |>
        dplyr::select(group, top_enriched_phenotype),
      by = "group"
    ) |>
    dplyr::left_join(specific_traits, by = "group") |>
    dplyr::mutate(
      top_enriched_category_any = tidyr::replace_na(top_enriched_category_any, "none"),
      top_enriched_tissue_any = tidyr::replace_na(top_enriched_tissue_any, "none"),
      top_enriched_pathway = tidyr::replace_na(top_enriched_pathway, "none"),
      top_enriched_phenotype = tidyr::replace_na(top_enriched_phenotype, "none")
    ) |>
    dplyr::arrange(dplyr::desc(n_snps), as.integer(group)) |>
    dplyr::select(group, n_snps, module_specific_traits, dplyr::everything())

  cat("\n\n### Overview of significant annotations across modules\n\n")
  print(knitr::kable(
    overview,
    caption = "Top module-specific traits and significant category, tissue, pathway (KEGG/Reactome), and phenotype (HP) annotations; trait values are specificity scores, and enrichment method tags are shown when FDR, permutation FDR, or module-vs-modules Fisher passes 0.05",
    row.names = FALSE,
    escape = FALSE
  ))
}
```

```{r step5-module-panels, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  categories_by_group <- stats::setNames(
    group_categories$by_group,
    vapply(
      group_categories$by_group,
      function(x) as.character(x$group),
      character(1)
    )
  )
  tissues_by_group <- stats::setNames(
    group_tissues$by_group,
    vapply(group_tissues$by_group, function(x) as.character(x$group), character(1))
  )
  pathways_by_group <- stats::setNames(
    module_enrichment$by_group,
    vapply(module_enrichment$by_group, function(x) as.character(x$group), character(1))
  )

  for (x in group_traits$by_group) {
    grp <- as.character(x$group)
    cat("\n\n### Module ", grp, " (", x$n_snps, " SNPs)\n\n", sep = "")

    category_x <- categories_by_group[[grp]]
    tissue_x <- tissues_by_group[[grp]]
    pathway_x <- pathways_by_group[[grp]]
    phenotype_x <- if (is.null(pathway_x)) {
      NULL
    } else {
      pathway_x$pathways[
        pathway_x$pathways$source == "HP",
        ,
        drop = FALSE
      ]
    }
    annotation <- data.frame(
      module = paste0("Module ", grp, " (", x$n_snps, " SNPs)"),
      trait_categories = .format_enrichment_hits(
          if (is.null(category_x)) NULL else category_x$comparison,
          "trait_category"
        ),
      tissues = .format_enrichment_hits(
          if (is.null(tissue_x)) NULL else tissue_x$comparison,
          "tissue"
        ),
      pathways = .format_pathway_hits(
          if (is.null(pathway_x)) NULL else pathway_x$pathways[
            pathway_x$pathways$source != "HP", , drop = FALSE
          ]
        ),
      phenotypes = .format_pathway_hits(phenotype_x),
      stringsAsFactors = FALSE
    )
    print(knitr::kable(
      annotation,
      caption = "Module annotations (enrichment vs trait baseline)",
      row.names = FALSE,
      escape = FALSE
    ))

    traits_to_show <- x$traits |>
      dplyr::filter(passes_min_specificity) |>
      dplyr::select(
        trait_name, trait_category,
        abs_mean_z, abs_mean_z_delta, specificity, mean_z
      ) |>
      dplyr::mutate(
        abs_mean_z = round(abs_mean_z, 3),
        abs_mean_z_delta = round(abs_mean_z_delta, 3),
        specificity = round(specificity, 2),
        mean_z = round(mean_z, 3)
      )

    if (nrow(traits_to_show) > 0) {
      print(knitr::kable(
        traits_to_show,
        caption = paste0(
          "Module-specific traits (specificity >= ", min_specificity,
          "; module mean |z| / (rest mean |z| + 0.1))"
        ),
        row.names = FALSE
      ))
    } else {
      cat("No traits passed the module-specificity cutoff (specificity >= ",
          min_specificity, ").\n\n", sep = "")
    }
  }
}
```

## Step 6: Module typology by SNP pleiotropy

Each SNP in GPMap carries a pleiotropy score: the number of distinct trait
categories and the number of distinct protein-coding genes it is associated
with via coloc groups (`get_all_variant_pleiotropies()`). These describe how
**broad** a SNP's biology is, independent of its effect profile. They give
modules a second axis of description:

- **Breadth** — the mean per-SNP pleiotropy within the module (trait-category
  pleiotropy and gene pleiotropy).
- **Tightness** — separation (internal minus external similarity).

Plotting the two yields a descriptive module typology (not a filter, and not a
reliability criterion):

- **tight + narrow** — a specific biological program (e.g. a tissue module);
- **tight + broad** — a regulatory hub affecting many categories or genes;
- **loose + broad** — generic background shared across the trait;
- **loose + narrow** — a private, low-signal module.

```{r step6-pleiotropy-typology, fig.width=7, fig.height=5, out.width="100%"}
variant_pleiotropy <- get_all_variant_pleiotropies()
module_pleiotropy <- summarise_snp_group_pleiotropy(
  groups = clusters_analysis,
  variant_pleiotropies = variant_pleiotropy,
  min_group_size = min_module_size
)

module_annotation_status <- group_categories$summary |>
  dplyr::select(group, top_enriched_category_any) |>
  dplyr::full_join(
    group_tissues$summary |>
      dplyr::select(group, top_enriched_tissue_any),
    by = "group"
  ) |>
  dplyr::full_join(
    module_enrichment$summary |>
      dplyr::select(group, top_enriched_pathway, top_enriched_phenotype),
    by = "group"
  ) |>
  dplyr::mutate(
    annotated = !(is.na(top_enriched_category_any) &
      is.na(top_enriched_tissue_any) &
      is.na(top_enriched_pathway) &
      is.na(top_enriched_phenotype))
  ) |>
  dplyr::select(group, annotated)

typology <- module_pleiotropy |>
  dplyr::left_join(
    module_quality |>
      dplyr::mutate(cluster = as.character(cluster)) |>
      dplyr::select(cluster, separation, mean_internal_similarity, connectedness),
    by = c("group" = "cluster")
  ) |>
  dplyr::left_join(module_annotation_status, by = "group")

typology_long <- typology |>
  tidyr::pivot_longer(
    c(mean_trait_category_pleiotropy, mean_gene_pleiotropy),
    names_to = "metric",
    values_to = "pleiotropy"
  ) |>
  dplyr::mutate(
    metric = dplyr::if_else(
      metric == "mean_trait_category_pleiotropy",
      "trait-category pleiotropy (distinct categories per SNP)",
      "gene pleiotropy (distinct genes per SNP)"
    )
  )

if (nrow(typology_long) >= 2) {
  medians <- typology_long |>
    dplyr::group_by(metric) |>
    dplyr::summarise(mx = stats::median(separation), my = stats::median(pleiotropy))

  print(ggplot2::ggplot(
    typology_long,
    ggplot2::aes(x = separation, y = pleiotropy, size = n_snps, color = annotated)
  ) +
    ggplot2::geom_vline(
      data = medians,
      ggplot2::aes(xintercept = mx),
      linetype = "dashed",
      alpha = 0.5
    ) +
    ggplot2::geom_hline(
      data = medians,
      ggplot2::aes(yintercept = my),
      linetype = "dashed",
      alpha = 0.5
    ) +
    ggplot2::geom_point(alpha = 0.75) +
    ggplot2::scale_size_area(max_size = 8) +
    ggplot2::facet_wrap(~metric, scales = "free_y") +
    ggplot2::scale_color_manual(
      values = c(`FALSE` = "grey60", `TRUE` = "#c2571b")
    ) +
    ggplot2::labs(
      title = "Module typology: separation vs SNP pleiotropy breadth",
      subtitle = "Dashed lines at medians; colour = module has a significant annotation",
      x = "separation (internal minus external similarity)",
      y = "mean per-SNP pleiotropy",
      size = "SNPs in module",
      color = "annotated"
    ) +
    ggplot2::theme_minimal())
}
```
