---
title: "Simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # Simulation parameters (inputs to simulate_trait / the generative tiers)
  K: 3
  module_sizes: [60, 25, 8]
  n_snps: 300
  n_traits: 600
  noise_sd: 1
  effect_size: 6
  effect_tail: 0
  overlap: disjoint
  nested_effect_corr: 0.8
  n_traits_per_module: [20, 15, 10]
  p_structural_zero: 0
  background_sparsity_sd: 1.2
  background_effect_scale: 0.5
  target_pattern: module
  p_negative: 0
  n_hub_traits: 0
  hub_snp_fraction: 0
  # Iterated generative axes: each element of these arrays becomes one input
  # version in a single run of this vignette. Paired arrays are iterated in
  # lock-step; the first element of each array is the baseline. All arrays are
  # written type-homogeneous (floats) so rmarkdown parses them as vectors
  # rather than lists of scalars.
  trait_overlap: [0.0, 0.5, 0.75, 1.0]
  trait_effect_corr: [0.0, 0.8, 0.8, 0.8]
  # p_active_background is crossed (full grid) with trait_overlap, so each
  # background density is evaluated at every trait-overlap level.
  p_active_background: [0.00, 0.02, 0.05, 0.07]
  snp_driver_groups: [1, 2, 3]
  driver_sharing: [0.0, 0.1, 0.25]
  p_spurious: [0.0, 0.01]

  # EBMF / clustering parameters (inputs to run_univariate_clustering)
  cluster_type: ebmf
  trait_subset: "phenotypic"
  min_snp_signals: 2
  compress_method: asinh
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_se_mode: unit
  ebmf_beta_scale: trait
  ebmf_lfsr_threshold: 0.01
  ebmf_magnitude_threshold: 0.3
  ebmf_drop_global: true
  ebmf_prior: point_normal
  ebmf_backfit: true

  # Program validation parameters (inputs to summarise_ebmf_programs)
  min_module_size: 3
  min_mean_internal: 0.3
  min_connectedness: 0.25
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 1
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5

  # Replication settings
  n_reps: 3
vignette: >
  %\VignetteIndexEntry{Simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

n_reps <- as.integer(params$n_reps)
set.seed(1)

if (!identical(params$cluster_type, "ebmf")) {
  stop("This vignette is configured for cluster_type = 'ebmf'")
}
if (params$K != length(params$module_sizes)) {
  stop("K must equal length(module_sizes)")
}

clustering_settings <- list(
  cluster_type = params$cluster_type,
  min_snp_signals = params$min_snp_signals,
  trait_subset = params$trait_subset,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  similarity_threshold = params$similarity_threshold,
  ebmf_se_mode = params$ebmf_se_mode,
  ebmf_beta_scale = params$ebmf_beta_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  n_null = params$n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = params$n_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)

simulation_defaults <- list(
  n_coloc_groups = params$n_snps,
  sign_pattern = "coherent",
  module_sizes = params$module_sizes,
  overlap = params$overlap,
  nested_effect_corr = params$nested_effect_corr,
  n_traits_per_module = params$n_traits_per_module,
  n_traits = params$n_traits,
  noise_sd = params$noise_sd,
  effect_size = params$effect_size,
  effect_tail = params$effect_tail,
  p_structural_zero = params$p_structural_zero,
  p_active_background = params$p_active_background[1],
  background_sparsity_sd = params$background_sparsity_sd,
  background_effect_scale = params$background_effect_scale,
  target_pattern = params$target_pattern,
  p_negative = params$p_negative,
  trait_overlap = params$trait_overlap[1],
  trait_effect_corr = params$trait_effect_corr[1],
  snp_driver_groups = params$snp_driver_groups[1],
  driver_sharing = params$driver_sharing[1],
  n_hub_traits = params$n_hub_traits[1],
  hub_snp_fraction = c(params$hub_snp_fraction[1], params$hub_snp_fraction[1]),
  p_spurious = params$p_spurious[1]
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against synthetic data with known ground truth. It is organised around **input
versions**: the same simulation, EBMF, and program-validation pipeline is run
on versions of the input that differ in how the planted modules are structured
(background density and trait overlap, driver architecture, and spurious
hits), so the recovery metrics are directly comparable across runs.

Each version is generated by `simulate_trait()`, which produces a trait object
(the target trait plus driver and background traits) with known ground truth.
The total number of SNPs is set via `n_coloc_groups`, and the total number of
traits (rows) via `n_traits`. Planted modules occupy **disjoint SNP sets** in
the clean version; overlap can then be introduced in trait space
(`trait_overlap`, `trait_effect_corr`), in SNP space (`driver_sharing`), or
within a module's own SNP set (`snp_driver_groups`).

The generative parameters are **calibrated to the messy marginals of a real
pleiotropy matrix** so the simulations are not trivially easy:

- Background association magnitudes are drawn from a standardised
  heavy-tailed distribution and scaled relative to the planted module effects
  by `background_effect_scale` (a relative effect-size parameter, not a
  significance threshold), so e.g. `background_effect_scale = 0.75` with
  `effect_size = 6` gives background hits with typical magnitude ~4.5, some
  below and some above. Noise is encoded as absence (`NA`), not as a hard
  z-score floor.
- Driver effects are heavy-tailed across traits (`effect_tail`), with a
  minority of cells disagreeing with the target-trait orientation
  (`p_negative = 0.05`), matching the ~20% of negative cells seen in real
  oriented matrices.
- Background trait densities are heavy-tailed per trait
  (`background_sparsity_sd`), so most background traits are observed at only a
  handful of SNPs while a tail is dense — reproducing the real per-trait
  sparsity distribution.
- Driver traits can be split across SNP groups (`snp_driver_groups`) and can
  also act at the neighbouring module's SNP set (`driver_sharing`), creating
  within-module and cross-module SNP structure.
- Driver traits carry a low rate of background-scale spurious hits at
  off-module SNPs (`p_spurious`).
- The target-trait row is dense and significant at **every** SNP
  (`target_pattern = "dense"`), as real target rows are after orientation.

For every version, the identical battery is run: `run_univariate_clustering()`
fits EBMF and keeps overlapping soft programs, and
`summarise_ebmf_programs()` filters programs by lFSR, loading magnitude,
internal coherence, and trait-subsampling stability. The key output is a
per-program table in which **each validation check is its own column**, so it
is immediately visible which check failed for each program:

- `size` — at least `min_module_size` filtered SNPs (`n_snps_filtered`).
- `internal_sim` — mean internal SNP similarity `>= min_mean_internal`.
- `connectivity` — pair connectedness `>= min_connectedness`.
- `stability` — top-loading SNPs recovered with replication
  `>= stability_threshold` under trait subsampling.

A program is `valid` only when every check passes. Because the input now has
realistic marginals, **recovery is expected to be partial and noisy**: the
validation gates sit at the same levels that separate real modules from
background, so only a subset of the discovered programs should be `valid`, and
the pass rates below are the meaningful output for comparing input versions.
The analysis settings below are fixed across versions; only the generative
input changes.

```{r settings-table}
settings_to_show <- c(
  "cluster_type", "trait_subset", "min_snp_signals", "min_module_size", "module_sizes",
  "compress_method", "compress_scale", "similarity_threshold",
  "overlap", "nested_effect_corr",
  "n_traits_per_module", "trait_overlap", "trait_effect_corr",
  "n_traits",
  "effect_size", "effect_tail", "p_structural_zero", "p_spurious",
  "p_active_background", "background_sparsity_sd", "background_effect_scale",
  "target_pattern", "p_negative",
  "ebmf_lfsr_threshold", "ebmf_magnitude_threshold", "ebmf_prior",
  "min_mean_internal", "min_connectedness", "n_rep",
  "frac_traits", "stability_top_n", "stability_threshold"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$module_sizes <- params$module_sizes
settings_values$overlap <- params$overlap
settings_values$nested_effect_corr <- params$nested_effect_corr
settings_values$n_traits_per_module <- params$n_traits_per_module
settings_values$trait_overlap <- params$trait_overlap[1]
settings_values$trait_effect_corr <- params$trait_effect_corr[1]
settings_values$n_traits <- params$n_traits
settings_values$effect_size <- params$effect_size
settings_values$effect_tail <- params$effect_tail
settings_values$p_structural_zero <- params$p_structural_zero
settings_values$p_spurious <- params$p_spurious[1]
settings_values$p_active_background <- params$p_active_background[1]
settings_values$background_sparsity_sd <- params$background_sparsity_sd
settings_values$background_effect_scale <- params$background_effect_scale
settings_values$target_pattern <- params$target_pattern
settings_values$p_negative <- params$p_negative
settings_values$stability_top_n <- validation_settings$top_n
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      if (nm %in% c("module_sizes", "n_traits_per_module")) {
        return(paste(settings_values[[nm]], collapse = ", "))
      }
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "Baseline EBMF analysis and validation settings (iterated generative parameters are shown at their baseline value; each input version is listed in the tier table below)"
)
```

```{r tier-helpers}
make_tier_spec <- function(...) {
  overrides <- list(...)
  spec <- utils::modifyList(simulation_defaults, overrides)
  if (is.null(spec$n_coloc_groups)) {
    spec$n_coloc_groups <- params$n_snps
  }
  if (is.null(spec$K)) {
    spec$K <- length(spec$module_sizes)
  }
  if (spec$K == 0) {
    spec$module_sizes <- NULL
  }
  return(spec)
}

# Build one set of input versions by iterating over a generative axis. The
# named vectors are iterated in lock-step; length-1 values (including scalar
# baselines after a grid-script override) are recycled. `.transform` can
# rewrite the per-tier overrides if a parameter needs reshaping per version.
make_axis <- function(axis_name, ..., .transform = NULL) {
  cols <- list(...)
  n <- max(lengths(cols))
  if (any(lengths(cols) != 1 & lengths(cols) != n)) {
    stop("All parameters in axis '", axis_name, "' must be length 1 or ", n)
  }
  idx <- function(x, i) x[if (length(x) == 1) 1L else i]
  specs <- vector("list", n)
  labels <- character(n)
  for (i in seq_len(n)) {
    raw <- lapply(cols, function(x) idx(x, i))
    ov <- raw
    if (!is.null(.transform)) {
      ov <- .transform(ov)
    }
    specs[[i]] <- do.call(make_tier_spec, ov)
    parts <- vapply(seq_along(raw), function(j) {
      paste0(names(raw)[j], " = ", paste(format(raw[[j]]), collapse = ", "))
    }, character(1))
    labels[i] <- paste0(axis_name, ": ", paste(parts, collapse = ", "))
  }
  names(specs) <- paste0(axis_name, "_", seq_len(n))
  names(labels) <- names(specs)
  return(list(labels = labels, specs = specs))
}

# Full cross of p_active_background with trait_overlap: every background
# density is evaluated at every trait-overlap level, so the background effect
# is tested over the trait-overlap versions.
make_background_cross <- function() {
  specs <- list()
  labels <- character(0)
  k <- 0L
  for (pb in params$p_active_background) {
    for (j in seq_along(params$trait_overlap)) {
      k <- k + 1L
      ov <- list(
        p_active_background = pb,
        trait_overlap = params$trait_overlap[j],
        trait_effect_corr = params$trait_effect_corr[j]
      )
      specs[[k]] <- do.call(make_tier_spec, ov)
      labels[k] <- sprintf(
        "background: p_active_background = %s, trait_overlap = %s",
        format(pb), format(params$trait_overlap[j])
      )
    }
  }
  names(specs) <- paste0("background_", seq_along(specs))
  names(labels) <- names(specs)
  return(list(labels = labels, specs = specs))
}

# For each found program, match it to the planted module whose SNP set it
# overlaps most (on the gated membership) and score recovery: module SNPs
# captured (`module_found`), module SNPs missed (`module_missed`), and program
# SNPs not in that module (`not_module`). Modules are a disjoint partition
# (`module_of_snp`; 0 = background). No-op when there are no programs or no
# planted modules.
score_module_recovery <- function(programs, assigned, module_of_snp) {
  if (nrow(programs) == 0 || is.null(module_of_snp)) {
    return(programs)
  }
  mod_ids <- sort(unique(module_of_snp[module_of_snp > 0]))
  if (length(mod_ids) == 0) {
    return(programs)
  }
  mod_snps <- lapply(mod_ids, function(m) {
    names(module_of_snp)[module_of_snp == m]
  })
  names(mod_snps) <- as.character(mod_ids)

  rows <- lapply(programs$program, function(p) {
    prog_snps <- assigned$snp_id[assigned$program == p]
    overlaps <- vapply(mod_ids, function(m) {
      sum(prog_snps %in% mod_snps[[as.character(m)]])
    }, integer(1))
    best <- mod_ids[which.max(overlaps)]
    n_module <- length(mod_snps[[as.character(best)]])
    n_found <- overlaps[best]
    data.frame(
      matched_module = best,
      n_module_snps = n_module,
      module_found = n_found,
      module_missed = n_module - n_found,
      not_module = length(prog_snps) - n_found,
      stringsAsFactors = FALSE
    )
  })
  m <- dplyr::bind_rows(rows)
  m$program <- programs$program
  return(dplyr::left_join(programs, m, by = "program"))
}

run_tier_one <- function(spec, rep) {
  dots <- spec
  dots$seed <- sample.int(.Machine$integer.max, 1)
  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  smry <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), validation_settings)
  )
  valid_ids <- smry$programs$program[smry$programs$status == "valid"]
  valid_assigned <- smry$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    res,
    predicted_memberships = valid_assigned
  )
  programs <- smry$programs
  programs$rep <- rep(rep, nrow(programs))
  programs <- score_module_recovery(
    programs, smry$assigned, sim$ground_truth$module_of_snp
  )
  driver_ov <- sim$ground_truth$driver_overlap
  true_driver_overlap <- if (!is.null(driver_ov) && nrow(driver_ov) > 1) {
    mean(driver_ov[upper.tri(driver_ov)])
  } else {
    NA_real_
  }
  list(
    summary = data.frame(
      k_planted = ev$k_planted,
      k_hat = ev$k_hat,
      k_error = ev$k_error,
      true_driver_overlap = true_driver_overlap,
      ari_all = ev$ari_all,
      ari_structured = ev$ari_structured,
      module_recall = ev$mean_module_recall,
      background_absorbed = ev$background_absorbed,
      coverage = ev$coverage,
      n_programs = nrow(programs),
      n_valid_programs = length(valid_ids),
      stringsAsFactors = FALSE
    ),
    programs = programs,
    driver_overlap = driver_ov
  )
}

run_tier <- function(name, label, spec, n_reps) {
  n_jobs <- 4L
  out <- if (n_reps > 1 && .Platform$OS.type != "windows") {
    parallel::mclapply(
      seq_len(n_reps),
      function(i) run_tier_one(spec, i),
      mc.cores = min(n_jobs, n_reps)
    )
  } else {
    lapply(seq_len(n_reps), function(i) run_tier_one(spec, i))
  }
  list(
    summary = dplyr::bind_rows(lapply(out, `[[`, "summary")) |>
      dplyr::mutate(tier = label),
    programs = dplyr::bind_rows(lapply(out, `[[`, "programs")) |>
      dplyr::mutate(tier = label, tier_name = name),
    driver_overlap = out[[1]]$driver_overlap
  )
}
```

## Input versions

Each input version is produced by iterating over the generative parameter
arrays in the `params` block: every array element becomes one version, paired
arrays are iterated in lock-step, and the first element of each array is the
baseline. `p_active_background` is the exception — it is **crossed** with
`trait_overlap`, so every background density is evaluated at every
trait-overlap level. The versions fall into three groups, each stressing one
aspect of the planted structure while holding the others at their baseline:

- **Background density × trait overlap** (`p_active_background` ×
  `trait_overlap` / `trait_effect_corr`): the full grid of background
  association density against the fraction of each module's driver traits
  shared with the other modules (correlated, `trait_effect_corr`, but
  non-identical effects).
- **Driver architecture** (`snp_driver_groups` / `driver_sharing`): splitting
  each module's SNP set into groups (each driver trait acts on one group) and
  letting some driver traits also act at the neighbouring module's SNP set.
- **Spurious hits** (`p_spurious`): background-scale off-module hits in the
  driver rows.

Null simulations (`K = 0`, no modules planted) are covered separately in
`investigation-univariate-null-simulations.html`.

```{r tiers}
# Each generative axis in the params block is iterated into input versions.
# p_active_background is crossed with trait_overlap (full grid); the other
# axes iterate one parameter while holding the rest at their baseline.
axes <- list(
  background = make_background_cross(),
  drivers = make_axis(
    "drivers",
    snp_driver_groups = params$snp_driver_groups,
    driver_sharing = params$driver_sharing
  ),
  spurious = make_axis(
    "spurious",
    p_spurious = params$p_spurious
  )
)

tier_specs <- list()
tier_labels <- character(0)
for (a in axes) {
  tier_specs <- c(tier_specs, a$specs)
  tier_labels <- c(tier_labels, a$labels)
}

tier_results <- lapply(names(tier_specs), function(nm) {
  run_tier(nm, tier_labels[[nm]], tier_specs[[nm]], n_reps)
})
names(tier_results) <- names(tier_specs)

tier_summary <- dplyr::bind_rows(lapply(tier_results, `[[`, "summary"))
program_checks <- dplyr::bind_rows(lapply(tier_results, `[[`, "programs"))
```

```{r tier-params-table}
tier_param_table <- do.call(rbind, lapply(names(tier_specs), function(nm) {
  s <- tier_specs[[nm]]
  data.frame(
    version = tier_labels[[nm]],
    n_snps = s$n_coloc_groups,
    K = s$K,
    module_sizes = if (s$K > 0) paste(s$module_sizes, collapse = ", ") else "-",
    n_traits_per_module = if (s$K > 0) paste(s$n_traits_per_module, collapse = ", ") else "-",
    trait_overlap = if (s$K > 0) s$trait_overlap else "-",
    trait_effect_corr = if (s$K > 0) s$trait_effect_corr else "-",
    noise_sd = s$noise_sd,
    p_structural_zero = if (s$K > 0) s$p_structural_zero else "-",
    p_spurious = s$p_spurious,
    effect_size = if (s$K > 0) s$effect_size else "-",
    effect_tail = if (s$K > 0) s$effect_tail else "-",
    background_effect_scale = if (s$K > 0) s$background_effect_scale else "-",
    p_negative = s$p_negative,
    p_active_background = s$p_active_background,
    background_sparsity_sd = s$background_sparsity_sd,
    target_pattern = if (s$K > 0) s$target_pattern else "-",
    driver_sharing = s$driver_sharing,
    sign_pattern = if (s$K > 0) s$sign_pattern else "-",
    snp_driver_groups = if (s$K > 0) s$snp_driver_groups else "-",
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  tier_param_table,
  row.names = FALSE,
  caption = "Generative parameters for each input version (each axis varies one feature of the planted structure while holding the others at their baseline)"
)

# True driver overlap between every pair of planted modules (identical across
# replicates because it is set by the roster, not by sampling).
pairwise_overlap <- do.call(rbind, lapply(names(tier_results), function(nm) {
  ov <- tier_results[[nm]]$driver_overlap
  if (is.null(ov) || nrow(ov) <= 1) {
    return(NULL)
  }
  pairs <- utils::combn(seq_len(nrow(ov)), 2)
  data.frame(
    version = tier_labels[[nm]],
    module_pair = apply(pairs, 2, function(p) paste0("M", p[1], "-M", p[2])),
    frac_shared_first = apply(pairs, 2, function(p) round(ov[p[1], p[2]], 3)),
    frac_shared_second = apply(pairs, 2, function(p) round(ov[p[2], p[1]], 3)),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  pairwise_overlap,
  row.names = FALSE,
  caption = "True driver overlap between every pair of planted modules: frac_shared_first = fraction of the first module's driver traits shared with the second, and vice versa"
)
```

## Aggregate results

Across all replicates and programs of each version, the table below reports
the recovery metrics and, for every validation check, the fraction of programs
that passed it. `valid` is the fraction of programs that passed all checks.

```{r aggregate}
check_rates <- program_checks |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    n_programs = dplyr::n(),
    pct_size = 100 * mean(size_pass),
    pct_internal_similarity = 100 * mean(internal_similarity_pass),
    pct_connectedness = 100 * mean(connectedness_pass),
    pct_stability = if (any(stability_checked)) {
      100 * mean(stability_pass[stability_checked])
    } else {
      NA_real_
    },
    pct_valid = 100 * mean(status == "valid"),
    .groups = "drop"
  )

aggregate_results <- tier_summary |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    mean_k_hat = mean(k_hat),
    mean_valid_programs = mean(n_valid_programs),
    mean_driver_overlap = mean(true_driver_overlap),
    mean_module_recall = mean(module_recall),
    mean_coverage = mean(coverage),
    mean_ari_structured = mean(ari_structured),
    mean_background_absorbed = mean(background_absorbed),
    .groups = "drop"
  ) |>
  dplyr::left_join(check_rates, by = "tier") |>
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric),
    ~ ifelse(is.nan(.x), NA_real_, .x)
  ))

aggregate_results |>
  dplyr::mutate(dplyr::across(
    dplyr::starts_with("pct_"),
    ~ ifelse(is.na(.x), "-", paste0(round(.x), "%"))
  )) |>
  knitr::kable(
    digits = 2,
    col.names = c(
      "version", "k_hat", "valid programs", "true driver overlap",
      "module recall", "coverage",
      "ARI (structured)", "background absorbed", "n programs",
      "size", "internal sim", "connectivity", "stability", "valid"
    ),
    caption = "Aggregate recovery, true planted driver overlap, and per-check pass rates by input version"
  )
```

The `stability` pass rate is computed only over programs that were actually
re-checked (those passing the size and internal-coherence checks).

## Program-level validation checks

For each version, every program found across the replicates is shown with the
four validation checks as separate columns. A check is marked `FAIL` when the
program does not meet that criterion; `n/a` under `stability` means the
program already failed the internal-coherence checks, so stability was not
re-tested.

Each program is also scored against the planted module it overlaps most (on
its gated SNPs): `% module found` is the fraction of that module's SNPs the
program recovered, `% module missed` its complement, and `% not module` is the
fraction of the program's SNPs that are not in that module (background or
other-module SNPs pulled in). `% module found` + `% module missed` always sum
to 100%. Under high `trait_overlap` a program that correctly merges two
modules matches only one, so its `% not module` rises even though it is
recovering real structure — read these alongside the merge-fair `module_recall`
in the Aggregate table.

```{r check-helper, include=FALSE}
check_table <- function(which_tier) {
  df <- program_checks |>
    dplyr::filter(.data$tier_name == which_tier)
  if (nrow(df) == 0) {
    return(NULL)
  }
  df |>
    dplyr::mutate(
      check_size = ifelse(size_pass, "pass", "FAIL"),
      check_internal_similarity = ifelse(internal_similarity_pass, "pass", "FAIL"),
      check_connectedness = ifelse(connectedness_pass, "pass", "FAIL"),
      check_stability = dplyr::case_when(
        !stability_checked ~ "n/a",
        stability_pass ~ "pass",
        TRUE ~ "FAIL"
      ),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      replication = round(replication, 3),
      matched = paste0("M", matched_module),
      pct_module_found = ifelse(
        n_module_snps > 0,
        paste0(round(100 * module_found / n_module_snps), "%"),
        "-"
      ),
      pct_module_missed = ifelse(
        n_module_snps > 0,
        paste0(round(100 * module_missed / n_module_snps), "%"),
        "-"
      ),
      pct_not_module = ifelse(
        n_snps_filtered > 0,
        paste0(round(100 * not_module / n_snps_filtered), "%"),
        "-"
      )
    ) |>
    dplyr::select(
      rep, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, replication, matched, pct_module_found,
      pct_module_missed, pct_not_module,
      check_size, check_internal_similarity, check_connectedness,
      check_stability, status
    ) |>
    dplyr::arrange(rep, program) |>
    knitr::kable(
      row.names = FALSE,
      col.names = c(
        "rep", "program", "n_snps", "n_snps filtered",
        "mean internal similarity", "connectedness", "replication",
        "matched module", "% module found", "% module missed", "% not module",
        "size", "internal sim", "connectivity", "stability", "status"
      )
    )
}
```

```{r tier-checks, results='asis'}
for (nm in names(tier_specs)) {
  cat("\n###", tier_labels[[nm]], "\n\n")
  tbl <- check_table(nm)
  if (is.null(tbl)) {
    cat("No EBMF programs were produced for this version.\n\n")
  } else {
    cat(tbl, sep = "\n")
    cat("\n\n")
  }
}
```

## Notes for interpreting runs

- Each input version comes from iterating the generative parameter arrays in
  the `params` block; arrays are iterated in lock-step with their paired
  parameter, and the first element of each array is the baseline.
- `p_active_background` is **crossed** with `trait_overlap` (full grid), so the
  background-density effect is evaluated at every trait-overlap level; the
  other axes iterate one parameter while holding the rest at baseline.
- `module_recall` and `coverage` preserve overlapping valid EBMF memberships.
- Planted modules occupy disjoint SNP sets; module overlap in these simulations
  comes from shared trait drivers (`trait_overlap`, with correlated
  `trait_effect_corr`), or from driver traits acting at the neighbouring
  module's SNP set (`driver_sharing`).
- `snp_driver_groups` splits each module's SNP set into groups, each driver
  trait acting on only one group; with more groups a module's drivers are only
  recoverable from their own SNP subgroup, fragmenting the module's signal.
- `p_spurious` adds background-scale off-module hits to the driver rows, so
  driver traits carry contamination outside their own module.
- The generative marginals are calibrated to a real pleiotropy matrix: driver
  effects are heavy-tailed across traits (`effect_tail`), a minority of cells
  disagree with the target orientation (`p_negative`), background densities
  are heavy-tailed per trait (`background_sparsity_sd`), and background
  magnitudes are scaled relative to the modules by `background_effect_scale`.
  With this difficulty level, recovery is expected to be partial and noisy —
  that is what makes the per-check pass rates informative.
- `ari_all` and `ari_structured` use each SNP's strongest assigned program as a
  hard projection, so they remain conventional partition metrics.
- `factor_strength` and `factor_strength_per_snp` are reported scores only;
  they do not currently filter programs.
- lFSR and magnitude thresholds determine `assigned` SNPs before coherence,
  factor-strength, and posterior-evidence calculations.
- Stability is run only for programs that pass the size and internal-coherence
  checks; the `stability` column shows `n/a` for programs that were not
  re-checked.
- Null simulations (`K = 0`, no modules planted) and the false-discovery rates
  of the validation checks under pure randomness are covered in
  `investigation-univariate-null-simulations.html`.