---
title: "Univariate SNP biological clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of SNP clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take a **single trait** and decompose its genetic architecture by grouping SNPs
according to their genome-wide pleiotropic profiles.

The central question is: **within one trait, what distinct biological programs
are represented by its associated SNPs?**

The output is a set of SNP modules, each annotated with background traits,
QTL studies, tissues, genes, and pathways. The runnable implementation is
[Investigation of univariate analysis](investigation-univariate-analysis.html).
[Investigation of EBMF analysis](investigation-ebmf-analysis.html) is the
overlapping-program alternative.

Background traits stay at **study level**. Molecular QTL studies are not
collapsed by gene, tissue, cis/trans, or QTL type.

### Core clustering preferences

Any method we adopt for SNP grouping should satisfy:

1. **Handles high sparsity** — Must natively process large, sparse matrices
   (thousands of singleton feature rows) without treating structural zeros as
   low-variance signal.  Alternatively, we can limit the sparsity by a hard
   threshold on the number of non-missing SNPs per trait.
2. **Signed and sign-driven clustering** — Must preserve the meaning of
   continuous positive/negative z-scores so variants group by **directional
   alignment**, not magnitude alone.
3. **Automated \(K\) discovery** — The engine should learn the number of
   clusters or latent factors from the data (modularity, ELBO), rather than
   relying on an arbitrary cut.  Filtering the results by a minimum module size
   can help with this.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Pleiotropy matrix\\n(study-level rows x SNPs)\"]
  S2 [label=\"2. Orient by sign(z_target)\\noptional asinh compression\"]
  S3 [label=\"3. SNP-SNP cosine similarity S\"]
  S4 [label=\"4. Cluster (signed Louvain default)\"]
  S5 [label=\"5. Reliability + magnitude annotations\"]
  S6 [label=\"6. Interpret reliable modules\\ntraits, tissues, pathways\"]
  S1 -> S2 -> S3 -> S4 -> S5 -> S6
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

### 1. Construct the pleiotropy matrix

Given:

- `nsnp` = independent SNPs associated with the target trait (coloc groups)
- `ntrait` = background studies in GPMap (phenotypic and molecular)

Construct a **traits x SNPs** matrix of signed z-scores (\(z = \beta / \mathrm{SE}\)).
Conceptually this is a SNP-by-trait profile matrix; the code stores traits as
rows and SNPs as columns.

```text
              SNP1   SNP2   SNP3   SNP4   ...
Trait A        z      z      z      z
Trait B        z      z      z      z
Trait C        z      z      z      z
...
```

Sparse rows (few non-missing SNPs) and ubiquitous rows (non-missing on almost
every SNP) are dropped; the target trait row is kept.

### 2. Orient SNP profiles relative to the target trait

Remove arbitrary allele coding so every column represents increasing target
liability. For SNP \(j\):

\[
X^*_{tj} = \mathrm{sign}(z_j)\, X_{tj}
\]

Optional \(\mathrm{asinh}(z)\) compression then limits the leverage of extreme
z-scores without a hard clip.

### 3. Calculate SNP-SNP similarity

Compare every SNP with every other SNP by its oriented trajectory across
background traits. Cosine similarity:

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\|\,\|x^*_k\|}
\]

The result is SNP x SNP:

- \(+1\) = concordant (parallel) profiles
- \(0\) = unrelated
- \(-1\) = discordant (anti-parallel) profiles

### 4. Cluster the SNPs

Group SNPs into modules from \(S\). The univariate vignette uses
`cluster_snp_profiles_louvain()`. Spectral clustering is available as
`cluster_snp_profiles_spectral()`. Soft overlapping programs live in the EBMF
vignette.

The recommended trial order is below. The first method provides the current
baseline; the following methods increasingly test different assumptions.

#### 1. Signed Louvain (current baseline)

Represent \(S\) as a signed weighted graph: SNP = node, similarity = edge.
Weak edges with \(|S| <\) `similarity_threshold` can be zeroed first.
Implemented as `cluster_snp_profiles_louvain()`.

Advantages:

- Returns a data-driven number of modules for fixed graph parameters
- Handles sparse networks
- Signed Louvain uses negative edges (anti-parallel profiles) rather than
  discarding them

Problems:

- Standard (unsigned) Louvain is directionally blind; we therefore use signed
  modularity
- Very sparse graphs can produce many tiny communities
- Resolution (`gamma`) and the similarity threshold both change community size
- Local optimisation can vary across runs or become trapped in suboptimal
  partitions

Use repeated runs across seeds, `gamma`, and `similarity_threshold` as the
reference against which other methods are compared.

#### 2. Signed Leiden (highest-priority addition)

Leiden is a community-detection refinement designed to produce better-connected
and more stable communities than Louvain.

Advantages:

- Closest methodological upgrade to the current pipeline
- Retains sparse graph processing and data-driven module discovery
- Reduces disconnected or poorly connected communities

Problems:

- Standard Leiden implementations generally assume non-negative weights
- A valid implementation needs signed modularity or coupled positive and
  negative graph layers; negative \(S\) values must not simply be discarded
- Resolution and edge-threshold choices still affect the number of modules

#### 3. Empirical Bayes matrix factorization

Treat the SNP x trait matrix as noisy observations of latent programs:

\[
Y = LF^{\top} + E
\]

- \(Y\) = SNP x trait matrix
- \(L\) = SNP loadings
- \(F\) = trait factors
- \(E\) = residual error

Advantages: SNPs can belong to multiple programs; \(K\) is inferred (ELBO);
overlapping biology is the default assumption.

Problems: factors are not necessarily mutually exclusive SNP modules and can
capture broad technical or global effects. Hard module labels require a
separate loading-to-membership rule.

Prefer this if the architecture is expected to be overlapping rather than
mutually exclusive SNP groups. See
[Investigation of EBMF analysis](investigation-ebmf-analysis.html).

#### 4. Signed degree-corrected stochastic block model

Model positive and negative graph edges as arising from latent SNP blocks.
A nested model with minimum-description-length selection can infer the number
of blocks while allowing SNPs to have different overall connectivity.

Advantages:

- Strong conceptual match for sparse signed networks
- Degree correction helps distinguish modules from highly connected hub SNPs
- Model selection can estimate the number of blocks

Problems:

- Weighted signed-edge support is more specialised than Louvain or Leiden
- Considerably more implementation and validation work
- Block structure is not identical to clustering by directional profile

#### 5. HDBSCAN on a sign-preserving embedding

Apply HDBSCAN to a low-dimensional representation of the oriented SNP profiles
or signed similarity graph.

Advantages:

- Learns a data-driven number of clusters
- Can label poorly supported SNPs as noise instead of forcing membership
- Handles clusters with unequal density

Problems:

- Does not operate directly on signed edges
- Results depend on the embedding, distance, and minimum cluster-size choices
- Sparse missing values must be handled before constructing the embedding

#### 6. Spectral clustering

Uses the graph Laplacian and its eigenvectors via
`cluster_snp_profiles_spectral()`.

Advantages: compresses many background traits into major axes of variation.

Problems: fragmented similarity graphs produce many disconnected components;
usually requires selecting \(K\).

### 5. Assessing module quality

Two complementary checks. Graph diagnostics ask whether a module is a real
blob on \(S\). Trait specificity asks whether that blob has a distinct
biological driver.

#### Module reliability scores

After clustering, go back to \(S\) with `summarise_snp_module_quality()`.
These diagnostics are not Louvain-specific: they need \(S\) and hard cluster
labels from any method.

A module is marked `reliable` only if all three hold:

1. Size \(\ge 3\) SNPs
2. Mean internal similarity \(\ge 0.3\)
3. Connectedness \(\ge 0.5\)

Size, internal similarity, and connectedness ask "is this a real blob?"
Downstream interpretation uses only SNPs in modules that pass all three.

**Internal similarity.** Mean signed cosine among member SNPs (diagonal
excluded). High values mean oriented profiles are roughly parallel. The
cutoff \(0.3\) is a weak "same direction, not noise" bar.

External similarity is the mean \(S\) from members to all other SNPs.
**Separation** is internal minus external. It is diagnostic, not a cutoff: a
module can be internally coherent but still sit inside a similar background.

**Connectedness.** Graph density on the thresholded unsigned graph
(\(|S| \ge\) `similarity_threshold`). Internal similarity can look decent if a
few pairs are very similar and the rest are near zero. Connectedness \(\ge 0.5\)
requires at least half of all pairs to be strong edges.

**Silhouette.** Rousseeuw silhouette with distance \(1 - S\) is still reported,
but is not part of the `reliable` rule. Because SNPs can participate in more
than one biological program, negative silhouettes are expected under hard
clustering and should not by themselves discard a coherent module.

The table also reports the number of connected components and the fraction of
SNPs in the largest component; those are not part of the `reliable` rule.

#### Enrichment of background traits

If no traits are specifically elevated in a module relative to the rest of the
SNPs, the module is harder to trust biologically.
`summarise_module_specific_traits()` ranks traits by module-vs-rest contrast:

| Metric | Meaning |
|---|---|
| `abs_mean_z` | Mean absolute z for the trait among SNPs in the module |
| `abs_mean_z_rest` | Mean absolute z among all other SNPs |
| `specificity` | Relative enrichment of the trait inside the module |
| `abs_mean_z_delta` | Absolute lift versus the rest |

\[
\mathrm{specificity} = \frac{\overline{|z|}_{\mathrm{module}}}{\overline{|z|}_{\mathrm{rest}} + 0.1}
\]

\[
\mathrm{abs\_mean\_z\_delta} = \overline{|z|}_{\mathrm{module}} - \overline{|z|}_{\mathrm{rest}}
\]

These describe what drives each cluster without using pathway enrichment as
the clustering criterion.

### 6. Interpret biological modules

The univariate vignette currently interprets **reliable** modules as:

- **Top traits by magnitude** (`abs_mean_z` within the module).
- **Trait-category composition** (`enrich_snp_group_trait_categories()`):
  observed vs expected share against the rest of the trait SNP baseline
  (coloc rows for SNPs attributed to the analysed trait, not the whole GPMap
  universe). Top elevated categories are shown with obs/exp/FE; FDR is noted
  when significant.
- **Tissue composition** (`enrich_snp_group_tissues()`, `background = "trait"`)
  on the same leave-one-out trait baseline.
- **Pathway enrichment** on QTL genes (KEGG and Reactome), treated as
  secondary because the full-trait gene set is often empty for broad traits
  such as BMI.
- **SNP pleiotropy typology** (`summarise_snp_group_pleiotropy()`): modules are
  placed on a *breadth* axis (mean per-SNP distinct trait categories / distinct
  genes) against *tightness* (separation). Descriptive only — it classifies
  modules as specific programs (tight + narrow), regulatory hubs (tight +
  broad), or generic background (loose + broad), without acting as a filter or
  reliability criterion.

#### Multiple testing correction for trait-category and tissue enrichments

Each reliable module is tested against **every** trait category and **every**
tissue for over-representation relative to the trait baseline (leave-one-out
hypergeometric test, plus the permutation and module-vs-modules summaries). A
single module therefore produces dozens of hypothesis tests — one per category
or tissue — and across all modules in a run the pipeline performs hundreds. If
we declared everything with raw p < 0.05 significant, we would expect ~5% of
purely null categories/tissues to pass by chance; a module testing ~50
categories would accumulate ~2-3 false hits before any real biology.

We therefore apply **Benjamini–Hochberg (BH) FDR
correction** *within each module*, across the categories (or tissues) tested
for that module. BH sorts the raw p-values, then finds the largest prefix that
can be declared significant while keeping the expected fraction of false
positives among the significant calls at or below the chosen level (0.05). The
reported `FDR` for a category/tissue is the smallest FDR level at which that
row would still be called significant (a "q-value"): a category with
`FDR = 0.03` means that if we declare it significant, we expect ≤3% of the
significant calls for that module to be false positives.

Note: **Correction is per-module, not global.** Each module's FDR is computed over
the categories/tissues tested for that module alone. Running many modules
still multiplies the opportunities for false positives across the whole
analysis, which is why the vignette insists on agreement between the three
methods (`FDR` hypergeometric, `perm` permutation, `vsmod` module-vs-modules
Fisher) and on replication across grid runs before treating a hit as real.

## Variables to consider

Exposed in the univariate vignette today:

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `api` | `production`, `local` | Public GPMap vs local QTL DB | Local data can add trans-QTL studies, increasing matrix density and changing the regulatory signals that connect SNPs. This can alter both module membership and tissue/gene interpretation. Production is more reproducible but may omit those signals. |
| `min_snp_signals` | `1`, `3`, `5` | Minimum SNPs with a value for a background trait | Raising it removes singleton or very sparse rows, reducing unstable edges and computation. It can also remove rare, module-specific biology. Lowering it preserves those signals but makes similarity more sensitive to sparse observations. |
| `similarity_threshold` | `0`, `0.2`, `0.5` | Minimum \(\|S\|\) retained as a Louvain edge | Raising it removes weak positive and negative edges, producing a sparser graph that may split into more or smaller modules. Lowering it preserves gradual structure but weak or noisy links can merge otherwise distinct modules. |
| `compress_method` | `none`, `asinh` | Transformation applied to z-scores | `asinh` is nearly linear around zero but compresses extreme values, so cosine similarity reflects patterns across several traits rather than one very large z-score. `none` preserves effect scale but allows outliers to dominate direction and norm. |
| `louvain_gamma` | `2`, `3` | Signed Louvain resolution | Higher `gamma` rewards smaller communities and generally increases the number of modules; lower values merge broader profiles. It directly affects module size, separation, connectedness, and the number passing reliability filters. |

Also relevant, but not currently params of the univariate vignette:

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| Trans QTLs | include / exclude | Whether distant molecular effects are retained | Including trans effects can reveal distributed regulatory programs and improve tissue/pathway interpretation, while increasing density and possible noise. Excluding them gives a more local cis-focused analysis but may miss system-level mechanisms. |
| Clustering method | Louvain, Leiden, EBMF, SBM, HDBSCAN, spectral | Model used to define modules or programs | Louvain/Leiden/SBM produce graph communities; EBMF allows overlapping programs; HDBSCAN can leave uncertain SNPs unassigned; spectral methods usually require \(K\). Different assumptions change module count and membership, so compare stability and interpretation rather than scores alone. |

Trans QTL include/exclude is currently a manual `cis_trans` filter or an API
choice (`local` vs `production`).

Reliability cutoffs used after clustering: `min_module_size = 3`,
`min_mean_internal = 0.3`, `min_connectedness = 0.5`.

## Appendix: Choosing clustering parameters

A sweep of the three main clustering parameters was run for BMI
(`target_trait_id = 1992`, `associations = "copermloc"`) by rendering the univariate
vignette across `louvain_gamma` in {2, 3}, `min_snp_signals` in {3, 5}, and
`similarity_threshold` in {0, 0.2, 0.5} (12 runs). The table below summarises,
per run, the diagnostics over **reliable** modules only. Significance is
"significant by any method" (`sig_any` = hypergeometric FDR, permutation FDR, or
module-vs-modules Fisher, all at 0.05); `n_category_modules`, `n_tissue_modules`,
`n_pathway_modules`, and `n_phenotype_modules` count the modules with at least
one significant trait category, tissue, pathway (KEGG/Reactome), or phenotype
(HP) hit respectively.

```{r appendix-grid-comparison, echo=FALSE}
grid_summary <- data.frame(
  run = c(
    "g2 mss3 thr0", "g2 mss3 thr0.2", "g2 mss3 thr0.5",
    "g2 mss5 thr0", "g2 mss5 thr0.2", "g2 mss5 thr0.5",
    "g3 mss3 thr0", "g3 mss3 thr0.2", "g3 mss3 thr0.5",
    "g3 mss5 thr0", "g3 mss5 thr0.2", "g3 mss5 thr0.5"
  ),
  gamma = c(2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3),
  min_snp_signals = c(3, 3, 3, 5, 5, 5, 3, 3, 3, 5, 5, 5),
  similarity_threshold = c(0, 0.2, 0.5, 0, 0.2, 0.5, 0, 0.2, 0.5, 0, 0.2, 0.5),
  n_modules = c(24, 22, 18, 24, 20, 19, 22, 17, 31, 23, 16, 30),
  n_reliable = c(23, 20, 16, 23, 19, 17, 17, 16, 29, 18, 15, 27),
  reliable_snps = c(394, 437, 409, 396, 439, 423, 158, 263, 354, 160, 261, 321),
  mean_internal_similarity = c(0.604, 0.574, 0.588, 0.610, 0.574, 0.596, 0.642, 0.554, 0.560, 0.630, 0.584, 0.562),
  mean_separation = c(0.310, 0.267, 0.229, 0.313, 0.270, 0.232, 0.394, 0.297, 0.205, 0.383, 0.318, 0.210),
  mean_connectedness = c(1.000, 0.978, 0.822, 1.000, 0.966, 0.833, 1.000, 0.945, 0.843, 1.000, 0.946, 0.838),
  n_category_modules = c(5, 6, 5, 5, 7, 4, 2, 3, 5, 1, 4, 4),
  n_tissue_modules = c(3, 1, 0, 3, 1, 0, 2, 2, 0, 2, 1, 1),
  n_pathway_modules = c(5, 3, 4, 6, 4, 4, 2, 3, 2, 2, 3, 3),
  n_phenotype_modules = c(1, 1, 3, 1, 2, 2, 2, 2, 2, 2, 2, 4),
  stringsAsFactors = FALSE
)

knitr::kable(
  grid_summary |>
    dplyr::arrange(gamma, min_snp_signals, similarity_threshold),
  caption = "Parameter-grid sweep (12 runs): reliable-module diagnostics by run (BMI, coloc associations; mean internal/separation/connectedness; counts of modules with significant category, tissue, pathway, and phenotype hits)",
  row.names = FALSE,
  align = "r"
)
```

Reading across the grid:

- **`louvain_gamma` — 2 beats 3.** At `gamma = 3` the graph fragments: with
  `similarity_threshold = 0.5` it returns 30-31 modules but only ~320-355
  reliable SNPs; at `thr = 0` the reliable SNP count collapses to 158-160,
  leaving only tiny, tight cliques (mean internal similarity 0.63-0.64 but
  modules of 3-26 SNPs). `gamma = 2` keeps far more reliable SNPs (394-439),
  the same key modules at larger sizes, and the strongest fold enrichments
  (e.g. a metabolic-disease module with fold enrichment ~10).
- **`min_snp_signals` — a tie; 5 is marginally cleaner.** At `gamma = 2`,
  `mss = 5` keeps slightly more reliable SNPs (396-439 vs 394-437) and at
  `thr = 0.2` surfaces the most category hits of any run (`n_category_modules =
  7`), while the stand-out modules stay the same. Either value works.
- **`similarity_threshold` — the choice now has three regimes, and `0.2` is the
  sweet spot.** At `thr = 0` every pair is an edge, so `connectedness` is 1.000
  for every module and no longer discriminates; weak edges keep modules small
  and internally tight, and tissue-specific modules emerge that are absent
  elsewhere (Adipose Subcutaneous FE = 40, Cells Cultured fibroblasts FE = 37,
  Brain (Meta) FE = 13). But `thr = 0` also surfaces noisy viral/bacterial
  pathway hits (Virion - Herpesvirus, Salmonella infection) on weak edges. At
  `thr = 0.5` modules are fewer, denser, `connectedness` (0.82-0.84) becomes a
  meaningful gate, and specific modules appear (dyslipidemia FE = 10,
  Phototransduction, Arginine, Calmodulin) — but the tissue modules vanish
  (`n_tissue_modules` 0-1). At `thr = 0.2` `connectedness` is informative again
  (0.97-0.98) while reliable SNP counts peak (437-439 at `gamma = 2`), tissue
  hits are retained (Thyroid FE = 37), strong category hits appear (Circulatory
  FE = 8, Digestive FE = 5.2, Metabolic Disease FE = 7.8 with FDR = 0.00026),
  and most viral/bacterial noise is gone — only the large background module
  (128-171 SNPs) keeps Virion - Herpesvirus.
- **HP phenotypes add signal mostly in mid-size, weakly-defined modules.** The
  recurring hits are broad measurement terms (DHEA-sulphate measurement on
  38-50-SNP modules, Reaction time, Sodium, White matter microstructure). The
  biologically coherent ones are `Age at menarche` on the Reproductive Measures
  module (`FE = 4.6`, gamma 3), `Insulinogenic index` / `Organic aciduria` on
  gamma-3 thr-0 glycemic-like modules, and `Cerebral amyloid deposition
  measurement` on a specific lipid module that also carries the Plasma
  lipoprotein assembly pathway and a Metabolic Disease hit (see the notable
  modules table). The DHEA-sulphate term's reappearance across runs and module
  sizes is consistent with HP annotation bias — treat it as weak evidence.

**Recommended defaults: `louvain_gamma = 2`, `min_snp_signals = 5`,
`similarity_threshold = 0.2`** (run `g2 mss5 thr0.2`). It keeps the largest
reliable SNP set (439) with an informative `connectedness` gate (0.966), the
most category hits (7), a tissue hit (Thyroid FE = 37), and a specific lipid
module, while shedding the viral/bacterial noise of `thr = 0` and keeping more
signal than `thr = 0.5`. Use `thr = 0` only when tissue-level modules matter
more than a working reliability gate; `thr = 0.5` remains the conservative,
cleanest default.

### Notable modules from the sweep

Each module is described by its **significant trait-category, tissue, pathway
(KEGG/Reactome), and phenotype (HP)** hits (method tags: `FDR`/`perm`/`vsmod`;
`FE` with bootstrap `CI`). Repeated runs of the same module are given together.

| Module | Significant trait category | Significant tissue | Significant pathway | Significant phenotype (HP) |
|---|---|---|---|---|
| **Dyslipidemia / metabolic** — `g2 mss3 thr0.5` mod 191 (3 SNPs); `g2 mss5 thr0.5` mod 190 (3 SNPs) | Metabolic Disease FE=10 (FDR=0.008, perm, vsmod); Circulatory FE=9.5 | — | — | — |
| **Lipid / amyloid** — `g2 mss5 thr0.2` mod 504 (9 SNPs); `g3 mss3 thr0.2` mod 692 (6 SNPs); `g3 mss5 thr0.2` mod 695 (8 SNPs) | Metabolic Disease FE=7.8-8.9 (FDR ~ 0.0002, perm, vsmod) | — | Reactome Plasma lipoprotein assembly | HP: Cerebral amyloid deposition measurement |
| **Adipose Subcutaneous** — `g2 mss3 thr0` mod 60 (3 SNPs); `g2 mss5 thr0` mod 61 (3 SNPs) | — | Adipose Subcutaneous FE=40, CI=[33,49] (FDR, perm, vsmod) | — | — |
| **Thyroid** — `g2 mss3 thr0.2` / `g2 mss5 thr0.2` mod 470 (3 SNPs) | — | Thyroid FE=37 (FDR=0.027, perm, vsmod) | — | — |
| **CNS / Brain (Meta)** — `g2 mss3 thr0` mod 164 (7 SNPs); `g2 mss5 thr0` mod 162 (7 SNPs); `g3 mss3 thr0` mod 414 (3 SNPs); `g3 mss3 thr0.2` mod 210 (3 SNPs); `g3 mss3 thr0.2` mod 663 (7 SNPs) | Anthropometric (background) | Brain (Meta) FE=13-27 (FDR, perm, vsmod) | — | — |
| **Fibroblasts** — `g2 mss3 thr0` / `g2 mss5 thr0` mod 212 (3 SNPs) | — | Cells Cultured fibroblasts FE=37 (FDR, perm, vsmod) | — | — |
| **Testis** — `g3 mss3 thr0` mod 330 (3 SNPs); `g3 mss5 thr0` mod 333 (3 SNPs) | — | Testis FE=29 (FDR, perm, vsmod) | — | — |
| **Whole blood / immune** — `g3 mss5 thr0.5` mod 196 (3 SNPs) | — | Whole blood FE=8.9 (FDR=0.008, perm, vsmod) | — | — |
| **Socioeconomic / education** — `g3 mss3 thr0.5` mod 39 (3 SNPs); `g3 mss5 thr0.5` mod 337 (4 SNPs) | Socioeconomic Measures FE=8.8-10 (FDR, perm, vsmod) | — | — | — |
| **Reproductive** — `g3 mss3 thr0.5` / `g3 mss5 thr0.5` mod 24 (6 SNPs) | Reproductive Measures FE=4.6 (FDR=0.021, perm) | — | — | HP: Age at menarche |
| **Whole-trait "general adiposity"** — `g2 mss3 thr0` mod 101 (70 SNPs); `g2 mss5 thr0` mod 104 (70 SNPs); `g2 mss3 thr0.5` mod 249 (158 SNPs); `g2 mss5 thr0.5` mod 244 (158 SNPs); `g2 mss5 thr0.2` mod 171 (128 SNPs) | Physiological Measures FE=1.2-1.4 (FDR ~ 0.002-0.005, perm, vsmod); Anthropometric Measures FE=1.3-1.9 | — | — | — |
| **Phototransduction / melanogenesis** — `g2 mss3 thr0.5` mod 157 (9 SNPs); `g2 mss5 thr0.5` mod 218 (16 SNPs) | Anthropometric (background) | — | KEGG Phototransduction, Melanogenesis, Ras signaling (FDR ~ 0.013) | — |
| **Arginine biosynthesis** — `g2 mss5 thr0.5` mod 189 (38 SNPs) | weak (FE 1.5-2.5) | — | KEGG Arginine biosynthesis (FDR=0.02) | — |
| **Nervous-system / calcium** — `g2 mss5 thr0.5` mod 224 (15 SNPs); `g2 mss3 thr0.5` mod 228 (18 SNPs) | Nervous System Disorders (background) | — | Reactome Calmodulin induced events; G-protein mediated events, opioid signalling | — |
| **Broad measurement phenotypes (caution)** — `g2 mss3 thr0.5` mod 67 (50 SNPs); `g2 mss5 thr0.5` mod 70 (48 SNPs); `g3 mss3 thr0.5` mod 75 (38 SNPs); `g3 mss5 thr0.5` mod 77 (41 SNPs); `g2 mss3 thr0` mod 421 / `g2 mss5 thr0` mod 422 (7 SNPs) | Anthropometric (background) | — | — | HP: DHEA-sulphate measurement (recurring); HP: Reaction time measurement |

Notes on interpretation:

- The **dyslipidemia module** (Metabolic Disease FE = 10, directionally pure
  triglycerides/ApoB/HDL profile) is the most specific and reproducible
  category module at `thr = 0.5`, appearing at `gamma = 2` for both `mss = 3`
  and `mss = 5`.
- The **lipid / amyloid module is the new standout at `thr = 0.2`**: Metabolic
  Disease FE = 7.8-8.9 with FDR ~ 0.0002 and all three methods agreeing, plus a
  Reactome Plasma lipoprotein assembly pathway and an HP Cerebral amyloid
  deposition phenotype. It reproduces across `gamma = 2` and `gamma = 3` and
  both `mss` values, making it the strongest multi-source hit in the sweep.
- The **tissue-specific modules** (Adipose Subcutaneous FE = 40, fibroblasts
  FE = 37, Brain (Meta), Testis, Thyroid FE = 37) appear at `thr = 0` and
  `thr = 0.2`, where weak edges are retained. The Adipose Subcutaneous hit has
  a tight CI ([33, 49]) and is confirmed by all three significance methods.
  These are lost at `thr = 0.5`.
- The **large "general adiposity" modules** (70-171 SNPs) are only enriched
  for the ubiquitous Physiological/Anthropometric categories at small fold
  changes (FE 1.2-1.9); they capture the shared BMI background rather than a
  distinct program and are significant mainly because of their size.
- **HP phenotype hits (new in this grid).** The biologically coherent ones land
  on modules already supported by category/tissue/pathway signal: `Age at
  menarche` on the Reproductive Measures module, `Cerebral amyloid deposition`
  on the lipid module, and `Insulinogenic index` / `Organic aciduria` on
  gamma-3 thr-0 glycemic-like modules. The rest are broad measurement terms
  (DHEA-sulphate, Reaction time, White matter microstructure, Sodium,
  Intraocular pressure, Temporal) that recur on different mid-size modules
  across runs — consistent with HP annotation bias, so treat them as weak
  evidence only.
- **Likely noise to treat with caution:** the viral/bacterial pathway hits at
  `thr = 0` (Virion - Herpesvirus on an 86-SNP module, Salmonella infection),
  the residual Virion - Herpesvirus on the large background module at
  `thr = 0.2`, `gamma = 3`'s tiny 3-SNP pathway modules (Long-term potentiation,
  Hormone signaling), and sparse-count artifacts such as Cell Trait FE = 120 at
  `g3 mss5 thr0.2`. Marginal permutation-only calls (e.g. Whole blood FE = 4.1
  at `g2 mss5 thr0.5`) can flip between renders and should not be over-read.
```
