---
title: "Investigation of collapsable bivariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of collapsable bivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Consider changing this to the SCZ GWAS that I uploaded: c12cb4ba-7216-49ef-9332-b75592fac586

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
```

This vignette extends the bivariate pleiotropy workflow. Standard background **trait** rows are kept (as in the standard bivariate analysis), while sparse gene-linked molecular QTL rows are collapsed into dense **feature** rows — enriched pathways or gene-level fallbacks. The downstream similarity, clustering, and block interpretation steps are unchanged.

We use **HDL cholesterol** (trait id `1630`) and **LDL cholesterol** (trait id `1486`).

## Step 1: Construct feature pleiotropy matrices (X₁ and X₂)

**Goal:** Build two matrices whose columns are target-trait SNPs and whose rows are shared **features**: background traits plus pathway/gene features derived from molecular QTL signals.

```{r step1-fetch}
trait_id_1 <- 1630
trait_id_2 <- 1486

traits_data <- traits(
  c(trait_id_1, trait_id_2),
  include_associations = TRUE
)
cg <- traits_data$coloc_groups
```

### Gene collapse and pathway mapping

1. **Background traits:** non-molecular colocalisation rows (no `gene_id`) are kept as individual trait rows, same as standard bivariate analysis.
2. **Molecular QTL:** at each target SNP locus, merge all gene-linked rows by `gene_id` (eQTL, pQTL, etc. in different coloc groups → one z-score per gene × SNP).
3. Run pathway enrichment on the union of those genes.
4. Map each gene to enriched pathways (`feature_id = source:term_id`); aggregate gene z-scores by mean within each pathway feature.
5. Genes outside enriched pathways remain as gene-level features (`gene:<gene_id>`).

```{r step1-build}
bivariate <- build_bivariate_feature_pleiotropy_matrices(
  trait_id_1 = trait_id_1,
  trait_id_2 = trait_id_2,
  coloc_groups = cg,
  snp_key = "variant_id",
  minimum_count_in_network = 2
)

X1 <- bivariate$x1_matrix
X2 <- bivariate$x2_matrix

list(
  trait_1 = traits_data$traits$trait_name[traits_data$traits$id == trait_id_1],
  trait_2 = traits_data$traits$trait_name[traits_data$traits$id == trait_id_2],
  n_features = nrow(X1),
  n_trait_rows = sum(bivariate$feature_info$feature_type == "trait"),
  n_pathway_rows = sum(bivariate$feature_info$feature_type == "pathway"),
  n_gene_rows = sum(bivariate$feature_info$feature_type == "gene"),
  n_snps_trait_1 = ncol(X1),
  n_snps_trait_2 = ncol(X2),
  shared_rows = identical(rownames(X1), rownames(X2))
)
```

```{r step1-features}
head(bivariate$feature_info, 10)
```

```{r step1-dims}
dim(X1)
dim(X2)
```

Each column is still one target SNP's profile. Rows combine standard background traits with denser pathway- or gene-level features summarising molecular pleiotropy.

## Step 2: Orient profiles relative to each target trait (X₁\* and X₂\*)

Target-trait orientation uses the original GWAS z-scores (`z_target_1`, `z_target_2`) via `sign(z_j)` on columns — applied to trait and feature rows alike.

```{r step2-orient}
oriented_1 <- orient_pleiotropy_matrix(
  X1,
  target_trait_id = trait_id_1,
  z_target = bivariate$z_target_1
)
oriented_2 <- orient_pleiotropy_matrix(
  X2,
  target_trait_id = trait_id_2,
  z_target = bivariate$z_target_2
)

X1_star <- oriented_1$x_matrix
X2_star <- oriented_2$x_matrix
```

## Step 3: Calculate cross-trait SNP similarity matrix (S\*)

```{r step3-similarity}
S_star <- cross_snp_similarity_matrix(X1_star, X2_star)$s_matrix

list(
  dim = dim(S_star),
  min_similarity = min(S_star, na.rm = TRUE),
  max_similarity = max(S_star, na.rm = TRUE),
  mean_similarity = mean(S_star, na.rm = TRUE)
)
```

## Step 4: Identify shared biological modules

```{r step4-cluster}
cross_clusters <- cluster_cross_trait_snps(S_star, linkage = "average")
hc1 <- cross_clusters$hclust_rows
hc2 <- cross_clusters$hclust_cols
```

```{r step4-heatmap, fig.height = 7, fig.width = 9}
heatmap(
  S_star,
  Rowv = as.dendrogram(hc1),
  Colv = as.dendrogram(hc2),
  scale = "none",
  main = "Clustered cross-trait SNP similarity blocks",
  xlab = "Trait 2 SNPs",
  ylab = "Trait 1 SNPs"
)
```

## Step 5: Interpret the bivariate blocks

### Block A: Shared concordant modules (cells ≈ +0.5 to +1.0)

Parallel pleiotropy footprints across traits — a shared biological highway.

### Block B: Shared discordant modules (cells ≈ −0.5 to −1.0)

Antagonistic blocks — genetic trade-offs between the two traits.

### Block C: Trait-specific modules (cells ≈ 0)

Near-zero similarity — largely independent mechanisms.

```{r step5-cut}
k1 <- 3
k2 <- 3

cross_modules <- cluster_cross_trait_snps(
  S_star,
  linkage = "average",
  k1 = k1,
  k2 = k2
)
```

```{r step5-summarise}
module_summary <- summarise_cross_trait_modules(
  S_star,
  clusters_rows = cross_modules$clusters_rows,
  clusters_cols = cross_modules$clusters_cols
)

module_summary |>
  dplyr::mutate(
    mean_similarity = round(mean_similarity, 2),
    min_similarity = round(min_similarity, 2),
    max_similarity = round(max_similarity, 2)
  )
```

```{r step5-by-type}
module_summary |>
  dplyr::count(block_type, name = "n_blocks") |>
  dplyr::arrange(dplyr::desc(n_blocks))
```

Pathway features densify the molecular QTL portion of the profile space while background traits remain available for comparison — block structure can reflect both shared biology and trait-specific mechanisms.
