---
title: "Simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  cluster_type: ebmf
  K: 3
  module_sizes: [5, 20, 50]
  overlap: disjoint
  nested_effect_corr: 0.8
  n_traits_per_module: [12, 5, 50]
  trait_overlap: 0
  trait_effect_corr: 0.8
  n_traits: 600
  n_snps: 300
  p_molecular: 0.8
  p_molecular_drivers: 0.1
  p_active_molecular: 0.02
  min_snp_signals: 2
  trait_subset: all
  min_module_size: 3
  compress_method: asinh
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_se_mode: unit
  ebmf_beta_scale: trait
  ebmf_lfsr_threshold: 0.05
  ebmf_magnitude_threshold: 0.2
  ebmf_drop_global: true
  ebmf_prior: point_laplace
  ebmf_backfit: true
  ebmf_hard_assignment: false
  min_mean_internal: 0.3
  min_connectedness: 0.25
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 5
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  effect_size: 3
  p_structural_zero: 0.05
  p_spurious: 0.1
  p_active_background: 0.2
  driver_sharing: 0
  snp_driver_groups: 1
  n_reps: 5
vignette: >
  %\VignetteIndexEntry{Simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)


low_noise_sd <- 0.5
moderate_noise_sd <- 1
extreme_noise_sd <- 6

n_reps <- as.integer(params$n_reps)
set.seed(1)

if (!identical(params$cluster_type, "ebmf")) {
  stop("This vignette is configured for cluster_type = 'ebmf'")
}
if (params$K != length(params$module_sizes)) {
  stop("K must equal length(module_sizes)")
}

clustering_settings <- list(
  cluster_type = params$cluster_type,
  min_snp_signals = params$min_snp_signals,
  trait_subset = params$trait_subset,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  similarity_threshold = params$similarity_threshold,
  ebmf_se_mode = params$ebmf_se_mode,
  ebmf_beta_scale = params$ebmf_beta_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  ebmf_hard_assignment = params$ebmf_hard_assignment,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  n_null = params$n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = params$n_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)

simulation_defaults <- list(
  module_sizes = params$module_sizes,
  overlap = params$overlap,
  nested_effect_corr = params$nested_effect_corr,
  n_traits_per_module = params$n_traits_per_module,
  trait_overlap = params$trait_overlap,
  trait_effect_corr = params$trait_effect_corr,
  n_traits = params$n_traits,
  p_molecular = params$p_molecular,
  p_molecular_drivers = params$p_molecular_drivers,
  p_active_molecular = params$p_active_molecular,
  effect_size = params$effect_size,
  p_structural_zero = params$p_structural_zero,
  p_spurious = params$p_spurious,
  p_active_background = params$p_active_background,
  driver_sharing = params$driver_sharing,
  snp_driver_groups = params$snp_driver_groups
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against synthetic data with known ground truth. It is organised around **input
versions** rather than generative parameter axes: the same simulation, EBMF,
and program-validation pipeline is run on versions of the input that differ in
how much the planted modules share.

Each version is generated by `simulate_trait()`, which produces a trait object
(the target trait plus driver and background traits) with known ground truth.
The total number of SNPs is set via `n_coloc_groups`, and the total number of
traits (rows) via `n_traits`. Planted modules occupy **disjoint SNP sets**, so
any module overlap comes purely from the *trait* space: a fraction
`trait_overlap` of each module's driver traits is shared with the other
modules, and a shared trait's effects are correlated across the modules it
drives (`trait_effect_corr`), but not identical unless `trait_effect_corr = 1`.
Increasing `trait_overlap` should therefore make EBMF programs increasingly
overlap and the planted modules harder to tell apart.

A fraction `p_molecular` of the background traits is molecular
(gene-annotated eQTL-like traits) and observed with per-SNP probability
`p_active_molecular`, which is much sparser than the phenotype traits' density.
The molecular fraction of each module's *drivers* is set separately by
`p_molecular_drivers`, so a module can keep a dense phenotype driver core even
when the background is almost entirely molecular — if every driver in a module
were a sparse molecular trait the module would be unrecoverable.

For every version, the identical battery is run: `run_univariate_clustering()`
fits EBMF and keeps overlapping soft programs, and
`summarise_ebmf_programs()` filters programs by lFSR, loading magnitude,
internal coherence, and trait-subsampling stability. The key output is a
per-program table in which **each validation check is its own column**, so it
is immediately visible which check failed for each program:

- `size` — at least `min_module_size` filtered SNPs (`n_snps_filtered`).
- `internal_sim` — mean internal SNP similarity `>= min_mean_internal`.
- `connectivity` — pair connectedness `>= min_connectedness`.
- `stability` — top-loading SNPs recovered with replication
  `>= stability_threshold` under trait subsampling.

A program is `valid` only when every check passes. The analysis settings below
are fixed across versions; only the generative input changes.

```{r settings-table}
settings_to_show <- c(
  "cluster_type", "trait_subset", "min_snp_signals", "min_module_size", "module_sizes",
  "overlap", "nested_effect_corr",
  "n_traits_per_module", "trait_overlap", "trait_effect_corr",
  "n_traits", "p_molecular", "p_molecular_drivers", "p_active_molecular",
  "ebmf_lfsr_threshold", "ebmf_magnitude_threshold",
  "min_mean_internal", "min_connectedness", "n_rep",
  "frac_traits", "stability_top_n", "stability_threshold"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$module_sizes <- params$module_sizes
settings_values$overlap <- params$overlap
settings_values$nested_effect_corr <- params$nested_effect_corr
settings_values$n_traits_per_module <- params$n_traits_per_module
settings_values$trait_overlap <- params$trait_overlap
settings_values$trait_effect_corr <- params$trait_effect_corr
settings_values$n_traits <- params$n_traits
settings_values$p_molecular <- params$p_molecular
settings_values$p_molecular_drivers <- params$p_molecular_drivers
settings_values$p_active_molecular <- params$p_active_molecular
settings_values$stability_top_n <- validation_settings$top_n
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      if (nm %in% c("module_sizes", "n_traits_per_module")) {
        return(paste(settings_values[[nm]], collapse = ", "))
      }
      if (nm == "p_molecular_drivers" && is.null(settings_values[[nm]])) {
        return("NULL (drivers follow p_molecular)")
      }
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "Fixed EBMF analysis and validation settings"
)
```

```{r tier-helpers}
make_tier_spec <- function(...) {
  overrides <- list(...)
  spec <- utils::modifyList(simulation_defaults, overrides)
  if (is.null(spec$n_coloc_groups)) {
    spec$n_coloc_groups <- params$n_snps
  }
  if (is.null(spec$K)) {
    spec$K <- length(spec$module_sizes)
  }
  if (spec$K == 0) {
    spec$module_sizes <- NULL
  }
  return(spec)
}

run_tier_one <- function(spec, rep) {
  dots <- spec
  dots$seed <- sample.int(.Machine$integer.max, 1)
  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  smry <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), validation_settings)
  )
  valid_ids <- smry$programs$program[smry$programs$status == "valid"]
  valid_assigned <- smry$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    res,
    predicted_memberships = valid_assigned
  )
  programs <- smry$programs
  programs$rep <- rep(rep, nrow(programs))
  driver_ov <- sim$ground_truth$driver_overlap
  true_driver_overlap <- if (!is.null(driver_ov) && nrow(driver_ov) > 1) {
    mean(driver_ov[upper.tri(driver_ov)])
  } else {
    NA_real_
  }
  list(
    summary = data.frame(
      k_planted = ev$k_planted,
      k_hat = ev$k_hat,
      k_error = ev$k_error,
      true_driver_overlap = true_driver_overlap,
      ari_all = ev$ari_all,
      ari_structured = ev$ari_structured,
      module_recall = ev$mean_module_recall,
      background_absorbed = ev$background_absorbed,
      coverage = ev$coverage,
      n_programs = nrow(programs),
      n_valid_programs = length(valid_ids),
      stringsAsFactors = FALSE
    ),
    programs = programs,
    driver_overlap = driver_ov
  )
}

run_tier <- function(name, label, spec, n_reps) {
  out <- lapply(seq_len(n_reps), function(i) run_tier_one(spec, i))
  list(
    summary = dplyr::bind_rows(lapply(out, `[[`, "summary")) |>
      dplyr::mutate(tier = label),
    programs = dplyr::bind_rows(lapply(out, `[[`, "programs")) |>
      dplyr::mutate(tier = label, tier_name = name),
    driver_overlap = out[[1]]$driver_overlap
  )
}
```

## Input versions

The versions differ only in `trait_overlap` (and `trait_effect_corr`, which is
inactive at zero overlap); every other generative parameter is identical. The
**null** version keeps only `n_coloc_groups` (SNPs) and `n_traits`; `K = 0`
means no modules are planted and the driver/module parameters are unused.

```{r tiers}
tier_base <- list(
  n_coloc_groups = params$n_snps,
  sign_pattern = "coherent",
  noise_sd = moderate_noise_sd,
  p_structural_zero = params$p_structural_zero,
  p_spurious = params$p_spurious,
  p_molecular = params$p_molecular,
  p_molecular_drivers = params$p_molecular_drivers,
  p_active_molecular = params$p_active_molecular,
  p_active_background = params$p_active_background,
  driver_sharing = params$driver_sharing,
  n_traits_per_module = params$n_traits_per_module,
  snp_driver_groups = params$snp_driver_groups
)

tier_specs <- list(
  clean = do.call(
    make_tier_spec,
    c(tier_base, list(trait_overlap = 0, trait_effect_corr = 0))
  ),
  moderate_trait_overlap = do.call(
    make_tier_spec,
    c(tier_base, list(trait_overlap = 0.5, trait_effect_corr = 0.8))
  ),
  strong_trait_overlap = do.call(
    make_tier_spec,
    c(tier_base, list(trait_overlap = 0.75, trait_effect_corr = 0.8))
  ),
  very_strong_trait_overlap = do.call(
    make_tier_spec,
    c(tier_base, list(trait_overlap = 0.9, trait_effect_corr = 0.8))
  ),
  full_trait_overlap = do.call(
    make_tier_spec,
    c(tier_base, list(trait_overlap = 1, trait_effect_corr = 0.8))
  ),
  full_trait_overlap_very_noisy = do.call(
    make_tier_spec,
    c(tier_base, list(trait_overlap = 1, trait_effect_corr = 0.8, noise_sd = extreme_noise_sd))
  ),
  null = make_tier_spec(
    n_coloc_groups = params$n_snps,
    K = 0,
    noise_sd = moderate_noise_sd,
    p_molecular = params$p_molecular,
    p_active_molecular = params$p_active_molecular,
    p_active_background = params$p_active_background
  )
)

tier_labels <- c(
  clean = "clean (no trait overlap)",
  moderate_trait_overlap = "moderate trait overlap",
  strong_trait_overlap = "strong trait overlap",
  very_strong_trait_overlap = "very strong trait overlap",
  full_trait_overlap = "full trait overlap",
  full_trait_overlap_very_noisy = "full trait overlap, very noisy",
  null = "null / no modules"
)

tier_results <- lapply(names(tier_specs), function(nm) {
  run_tier(nm, tier_labels[[nm]], tier_specs[[nm]], n_reps)
})
names(tier_results) <- names(tier_specs)

tier_summary <- dplyr::bind_rows(lapply(tier_results, `[[`, "summary"))
program_checks <- dplyr::bind_rows(lapply(tier_results, `[[`, "programs"))
```

```{r tier-params-table}
tier_param_table <- do.call(rbind, lapply(names(tier_specs), function(nm) {
  s <- tier_specs[[nm]]
  data.frame(
    version = tier_labels[[nm]],
    n_snps = s$n_coloc_groups,
    K = s$K,
    module_sizes = if (s$K > 0) paste(s$module_sizes, collapse = ", ") else "-",
    n_traits_per_module = if (s$K > 0) paste(s$n_traits_per_module, collapse = ", ") else "-",
    trait_overlap = if (s$K > 0) s$trait_overlap else "-",
    trait_effect_corr = if (s$K > 0) s$trait_effect_corr else "-",
    noise_sd = s$noise_sd,
    p_structural_zero = if (s$K > 0) s$p_structural_zero else "-",
    p_spurious = s$p_spurious,
    p_molecular = s$p_molecular,
    p_molecular_drivers = if (s$K > 0) s$p_molecular_drivers else "-",
    p_active_molecular = s$p_active_molecular,
    p_active_background = s$p_active_background,
    driver_sharing = s$driver_sharing,
    sign_pattern = if (s$K > 0) s$sign_pattern else "-",
    snp_driver_groups = if (s$K > 0) s$snp_driver_groups else "-",
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  tier_param_table,
  row.names = FALSE,
  caption = "Generative parameters for each input version (trait-overlap scenarios share all parameters except trait_overlap / trait_effect_corr)"
)

# True driver overlap between every pair of planted modules (identical across
# replicates because it is set by the roster, not by sampling).
pairwise_overlap <- do.call(rbind, lapply(names(tier_results), function(nm) {
  ov <- tier_results[[nm]]$driver_overlap
  if (is.null(ov) || nrow(ov) <= 1) {
    return(NULL)
  }
  pairs <- utils::combn(seq_len(nrow(ov)), 2)
  data.frame(
    version = tier_labels[[nm]],
    module_pair = apply(pairs, 2, function(p) paste0("M", p[1], "-M", p[2])),
    frac_shared_first = apply(pairs, 2, function(p) round(ov[p[1], p[2]], 3)),
    frac_shared_second = apply(pairs, 2, function(p) round(ov[p[2], p[1]], 3)),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  pairwise_overlap,
  row.names = FALSE,
  caption = "True driver overlap between every pair of planted modules: frac_shared_first = fraction of the first module's driver traits shared with the second, and vice versa"
)
```

## Aggregate results

Across all replicates and programs of each version, the table below reports
the recovery metrics and, for every validation check, the fraction of programs
that passed it. `valid` is the fraction of programs that passed all checks.

```{r aggregate}
check_rates <- program_checks |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    n_programs = dplyr::n(),
    pct_size = 100 * mean(size_pass),
    pct_internal_similarity = 100 * mean(internal_similarity_pass),
    pct_connectedness = 100 * mean(connectedness_pass),
    pct_stability = if (any(stability_checked)) {
      100 * mean(stability_pass[stability_checked])
    } else {
      NA_real_
    },
    pct_valid = 100 * mean(status == "valid"),
    .groups = "drop"
  )

aggregate_results <- tier_summary |>
  dplyr::group_by(tier) |>
  dplyr::summarise(
    mean_k_hat = mean(k_hat),
    mean_valid_programs = mean(n_valid_programs),
    mean_driver_overlap = mean(true_driver_overlap),
    mean_module_recall = mean(module_recall),
    mean_coverage = mean(coverage),
    mean_ari_structured = mean(ari_structured),
    mean_background_absorbed = mean(background_absorbed),
    .groups = "drop"
  ) |>
  dplyr::left_join(check_rates, by = "tier") |>
  dplyr::mutate(dplyr::across(
    dplyr::where(is.numeric),
    ~ ifelse(is.nan(.x), NA_real_, .x)
  ))

aggregate_results |>
  dplyr::mutate(dplyr::across(
    dplyr::starts_with("pct_"),
    ~ ifelse(is.na(.x), "-", paste0(round(.x), "%"))
  )) |>
  knitr::kable(
    digits = 2,
    col.names = c(
      "version", "k_hat", "valid programs", "true driver overlap",
      "module recall", "coverage",
      "ARI (structured)", "background absorbed", "n programs",
      "size", "internal sim", "connectivity", "stability", "valid"
    ),
    caption = "Aggregate recovery, true planted driver overlap, and per-check pass rates by input version"
  )
```

The `stability` pass rate is computed only over programs that were actually
re-checked (those passing the size and internal-coherence checks).

## Program-level validation checks

For each version, every program found across the replicates is shown with the
four validation checks as separate columns. A check is marked `FAIL` when the
program does not meet that criterion; `n/a` under `stability` means the
program already failed the internal-coherence checks, so stability was not
re-tested.

```{r check-helper, include=FALSE}
check_table <- function(which_tier) {
  df <- program_checks |>
    dplyr::filter(.data$tier_name == which_tier)
  if (nrow(df) == 0) {
    return(NULL)
  }
  df |>
    dplyr::mutate(
      check_size = ifelse(size_pass, "pass", "FAIL"),
      check_internal_similarity = ifelse(internal_similarity_pass, "pass", "FAIL"),
      check_connectedness = ifelse(connectedness_pass, "pass", "FAIL"),
      check_stability = dplyr::case_when(
        !stability_checked ~ "n/a",
        stability_pass ~ "pass",
        TRUE ~ "FAIL"
      ),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      replication = round(replication, 3)
    ) |>
    dplyr::select(
      rep, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, replication,
      check_size, check_internal_similarity, check_connectedness,
      check_stability, status
    ) |>
    dplyr::arrange(rep, program) |>
    knitr::kable(
      row.names = FALSE,
      col.names = c(
        "rep", "program", "n_snps", "n_snps filtered",
        "mean internal similarity", "connectedness", "replication",
        "size", "internal sim", "connectivity", "stability", "status"
      )
    )
}
```

### Clean (no trait overlap)

`trait_overlap = 0`: every module has its own exclusive driver set, so the
planted modules have independent trait signatures and disjoint SNPs. Recovery
should be near-perfect.

```{r clean-checks}
clean_checks <- check_table("clean")
if (is.null(clean_checks)) {
  "No EBMF programs were produced for the clean version."
} else {
  clean_checks
}
```

### Moderate trait overlap

`trait_overlap = 0.5`: half of each module's driver traits are shared with the
other modules, with correlated (`trait_effect_corr = 0.8`) but non-identical
effects. EBMF programs should start to overlap, making the modules harder to
separate.

```{r moderate-trait-overlap-checks}
moderate_trait_overlap_checks <- check_table("moderate_trait_overlap")
if (is.null(moderate_trait_overlap_checks)) {
  "No EBMF programs were produced for the moderate trait overlap version."
} else {
  moderate_trait_overlap_checks
}
```

### Strong trait overlap

`trait_overlap = 0.75`: three quarters of each module's drivers are shared.
Strongly overlapping trait signatures should make program separation
increasingly difficult.

```{r strong-trait-overlap-checks}
strong_trait_overlap_checks <- check_table("strong_trait_overlap")
if (is.null(strong_trait_overlap_checks)) {
  "No EBMF programs were produced for the strong trait overlap version."
} else {
  strong_trait_overlap_checks
}
```

### Very strong trait overlap

`trait_overlap = 0.9`: almost every driver is shared, so the modules have
almost identical trait signatures and are expected to merge or lose
resolution.

```{r very-strong-trait-overlap-checks}
very_strong_trait_overlap_checks <- check_table("very_strong_trait_overlap")
if (is.null(very_strong_trait_overlap_checks)) {
  "No EBMF programs were produced for the very strong trait overlap version."
} else {
  very_strong_trait_overlap_checks
}
```

### Full trait overlap

`trait_overlap = 1`: every driver is shared across the modules, so the modules
have fully overlapping trait signatures (effects still correlated at
`trait_effect_corr = 0.8`, not identical) and are expected to collapse into a
single program or be recovered only as merged, near-identical factors.

```{r full-trait-overlap-checks}
full_trait_overlap_checks <- check_table("full_trait_overlap")
if (is.null(full_trait_overlap_checks)) {
  "No EBMF programs were produced for the full trait overlap version."
} else {
  full_trait_overlap_checks
}
```

### Full trait overlap, very noisy

Same as the previous version, but with `noise_sd = 6` instead of `1`
```{r full-trait-very-noisy-overlap-checks}
full_trait_overlap_checks <- check_table("full_trait_overlap_very_noisy")
if (is.null(full_trait_overlap_checks)) {
  "No EBMF programs were produced for the full trait overlap version."
} else {
  full_trait_overlap_checks
}
```

### Null / no modules

`K = 0`: only the number of SNPs (`n_coloc_groups`) and traits (`n_traits`)
are used; all traits are unstructured background. EBMF should not produce any
program that passes the checks, i.e. no false structure.

```{r null-checks}
null_checks <- check_table("null")
if (is.null(null_checks)) {
  "No EBMF programs were produced for the null version."
} else {
  null_checks
}
```

## Notes for interpreting runs

- `module_recall` and `coverage` preserve overlapping valid EBMF memberships.
- Planted modules occupy disjoint SNP sets; all module overlap in these
  simulations comes from shared trait drivers (`trait_overlap`), with
  correlated (`trait_effect_corr`) but non-identical effects, so the modules
  read as "different SNPs + overlapping trait signatures".
- `ari_all` and `ari_structured` use each SNP's strongest assigned program as a
  hard projection, so they remain conventional partition metrics.
- `factor_strength` and `factor_strength_per_snp` are reported scores only;
  they do not currently filter programs.
- lFSR and magnitude thresholds determine `assigned` SNPs before coherence,
  factor-strength, and posterior-evidence calculations.
- Stability is run only for programs that pass the size and internal-coherence
  checks; the `stability` column shows `n/a` for programs that were not
  re-checked.
- Molecular traits are gene-annotated and much sparser than phenotype traits.
  Sparse molecular background traits frequently fall below `min_snp_signals`
  and are dropped before matrix construction, so `n_traits` is a nominal
  matrix size rather than the number of traits that survive to clustering.