---
title: "Investigation of univariate analysis"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  cluster_type: "ebmf"
  louvain_gamma: 1.5
  compress_method: "asinh"
  compress_scale: 5
  include_trans: false
  include_situated_gene: false 
  similarity_threshold: 0.2
  ebmf_se_mode: "unit"
  ebmf_beta_scale: "trait"
  ebmf_magnitude_threshold: 0.75
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  min_mean_internal: 0.3
  min_connectedness: 0.5
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 3
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

Single-trait SNP clustering: each column is a GWAS SNP for the target trait,
each row is a background study (phenotypic or molecular; no collapsing).
Profiles are oriented, compressed, and clustered with **`r params$cluster_type`**
(Louvain: signed modularity on the SNP similarity graph; EBMF: empirical Bayes
matrix factorization with overlapping programs, hard-assigned to their
strongest factor). Modules/programs are then validated with a filtering
framework — internal coherence (mean internal similarity and connectedness),
factor strength normalised by the square root of the number of filtered SNPs,
and trait-subsampling stability —
plus descriptive evidence scores (aggregate posterior evidence and factor
distinctiveness) before interpretation.

We use **BMI** (trait id `1992`).

```{r params}
target_trait_id <- 1992
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_specificity <- 1.25

api <- params$api
trait_subset <- params$trait_subset
min_module_size <- as.integer(params$min_module_size)
min_snp_signals <- as.integer(params$min_snp_signals)
cluster_type <- params$cluster_type
louvain_gamma <- as.numeric(params$louvain_gamma)
compress_method <- params$compress_method
compress_scale <- as.numeric(params$compress_scale)
include_trans <- isTRUE(params$include_trans)
include_situated_gene <- isTRUE(params$include_situated_gene)
similarity_threshold <- as.numeric(params$similarity_threshold)
ebmf_se_mode <- params$ebmf_se_mode
ebmf_beta_scale <- params$ebmf_beta_scale
ebmf_lfsr_threshold <- as.numeric(params$ebmf_lfsr_threshold)
ebmf_prior <- params$ebmf_prior
ebmf_backfit <- isTRUE(params$ebmf_backfit)
ebmf_magnitude_threshold <- if (is.na(params$ebmf_magnitude_threshold)) {
  NA_real_
} else {
  as.numeric(params$ebmf_magnitude_threshold)
}
associations <- params$associations
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
stability_threshold <- as.numeric(params$stability_threshold)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)

select_api(api)

knitr::kable(
  data.frame(
    api = api,
    gpmap_url = getOption("gpmap_url"),
    min_snp_signals = min_snp_signals,
    trait_subset = trait_subset,
    cluster_type = cluster_type,
    louvain_gamma = louvain_gamma,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    include_situated_gene = include_situated_gene,
    min_specificity = min_specificity,
    similarity_threshold = similarity_threshold,
    ebmf_se_mode = if (cluster_type == "ebmf") ebmf_se_mode else NULL,
    ebmf_beta_scale = if (cluster_type == "ebmf") ebmf_beta_scale else NULL,
    ebmf_lfsr_threshold = if (cluster_type == "ebmf") ebmf_lfsr_threshold else NULL,
    ebmf_prior = if (cluster_type == "ebmf") ebmf_prior else NULL,
    ebmf_backfit = if (cluster_type == "ebmf") ebmf_backfit else NULL,
    ebmf_magnitude_threshold = if (cluster_type == "ebmf") ebmf_magnitude_threshold else NULL,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = if (cluster_type == "ebmf") n_null else NULL,
    alpha_membership = if (cluster_type == "ebmf") alpha_membership else NULL,
    n_candidate_tier = if (cluster_type == "ebmf") n_candidate_tier else NULL,
    n_rep = if (cluster_type == "ebmf") n_rep else NULL,
    frac_traits = if (cluster_type == "ebmf") frac_traits else NULL,
    stability_top_n = if (cluster_type == "ebmf") stability_top_n else NULL,
    stability_threshold = if (cluster_type == "ebmf") stability_threshold else NULL,
    posterior_stat = if (cluster_type == "ebmf") posterior_stat else NULL,
    posterior_evidence_cap = if (cluster_type == "ebmf") posterior_evidence_cap else NULL,
    redundancy_threshold = if (cluster_type == "ebmf") redundancy_threshold else NULL,
    cores = if (cluster_type == "ebmf") cores else NULL,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

traits_meta <- traits(target_trait_id)
target_trait_name <- traits_meta$traits$trait_name[
  traits_meta$traits$id == target_trait_id
]
```

## Step 1: Pleiotropy matrix

`build_pleiotropy_matrix()` builds z-scores (`beta / se`) at each target SNP.
With `associations = "coloc"` it uses colocalisation rows only; with
`associations = "full"` it uses all full associations
(`/associations-full`). Molecular QTL studies stay as their own rows.
When `include_trans = FALSE`, trans target markers are removed from the
association rows before the matrix is built.

```{r step1-build}
coloc_groups <- trait(target_trait_id, include_associations = TRUE)$coloc_groups
if (!include_trans && "cis_trans" %in% names(coloc_groups)) {
  coloc_groups <- coloc_groups |>
    dplyr::filter(is.na(cis_trans) | tolower(cis_trans) != "trans")
}
pleiotropy <- build_pleiotropy_matrix(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  snp_key = "variant_id",
  associations = associations
)
X <- pleiotropy$x_matrix
cg <- coloc_groups

n_pheno <- sum(pleiotropy$trait_info$feature_type == "phenotypic", na.rm = TRUE)
n_mol <- sum(pleiotropy$trait_info$feature_type == "molecular", na.rm = TRUE)

knitr::kable(
  data.frame(
    trait = target_trait_name,
    n_rows = nrow(X),
    n_phenotypic = n_pheno,
    n_molecular = n_mol,
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3)
  ),
  caption = "Pleiotropy matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous traits

Drop rows with fewer than `min_snp_signals` non-`NA` values, or with non-`NA`
fraction above `max_snp_fraction` (target trait kept).

```{r step1-sparse-filter}
trait_snp_counts <- rowSums(!is.na(X))
trait_snp_frac <- trait_snp_counts / ncol(X)
sparse_trait_ids <- names(trait_snp_counts)[trait_snp_counts < min_snp_signals]
ubiquitous_trait_ids <- names(trait_snp_frac)[trait_snp_frac > max_snp_fraction]
drop_trait_ids <- setdiff(
  union(sparse_trait_ids, ubiquitous_trait_ids),
  as.character(target_trait_id)
)

X <- X[!rownames(X) %in% drop_trait_ids, , drop = FALSE]
pleiotropy$trait_info <- pleiotropy$trait_info |>
  dplyr::filter(!as.character(trait_id) %in% drop_trait_ids)

knitr::kable(
  data.frame(
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    n_sparse_traits_removed = length(setdiff(sparse_trait_ids, as.character(target_trait_id))),
    n_ubiquitous_traits_removed = length(setdiff(ubiquitous_trait_ids, as.character(target_trait_id))),
    n_traits_after = nrow(X),
    n_snps = ncol(X)
  ),
  caption = "After sparse / ubiquitous trait filter",
  align = "r"
)
```

## Step 2: Orient and compress (X\*)

```{r step2-orient}
X_star <- orient_pleiotropy_matrix(
  X,
  target_trait_id = target_trait_id
)$x_matrix
X_star <- compress_effect_matrix(
  X_star,
  method = compress_method,
  asinh_scale = compress_scale
)
trait_matrix <- t(X_star)
```

## Step 3: SNP similarity (S)

Missing trait-SNP cells are treated as zero effects, preserving the original
cosine similarity calculation.

```{r step3-similarity}
similarity <- snp_similarity_matrix(X_star)
S <- similarity$s_matrix
s_upper <- S[upper.tri(S)]
overlap_upper <- similarity$overlap_matrix[
  upper.tri(similarity$overlap_matrix)
]
eligible_upper <- similarity$eligible_matrix[
  upper.tri(similarity$eligible_matrix)
]

knitr::kable(
  data.frame(
    n_snps = ncol(S),
    median_pairwise_overlap = stats::median(overlap_upper),
    fraction_pairs_eligible = round(mean(eligible_upper), 3),
    min_similarity = round(min(s_upper, na.rm = TRUE), 2),
    max_similarity = round(max(s_upper, na.rm = TRUE), 2),
    mean_similarity = round(mean(s_upper, na.rm = TRUE), 2)
  ),
  caption = "SNP similarity summary",
  align = "r"
)
```

## Step 4: Clustering

Edges with `|S| < similarity_threshold` are zeroed before Louvain. For EBMF,
the oriented compressed matrix `X_star` is factorised directly (features x
SNPs), a global mega-factor is dropped if detected, and each SNP is assigned to
the factor with the largest absolute posterior loading among the factors it
significantly belongs to.

```{r step4-cluster}
ebmf_result <- NULL
if (cluster_type == "louvain") {
  clusters_fit <- cluster_snp_profiles_louvain(
    S,
    similarity_threshold = similarity_threshold,
    gamma = louvain_gamma
  )
} else if (cluster_type == "ebmf") {
  ebmf_result <- run_univariate_clustering(
    list(
      trait = list(id = target_trait_id),
      coloc_groups = cg
    ),
    target_trait_id = target_trait_id,
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    compress_method = compress_method,
    compress_scale = compress_scale,
    similarity_threshold = similarity_threshold,
    cluster_type = "ebmf",
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    ebmf_se_mode = ebmf_se_mode,
    ebmf_beta_scale = ebmf_beta_scale,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness
  )
  clusters_fit <- list(
    cluster = ebmf_result$clusters,
    details = ebmf_result$cluster_details
  )
} else {
  stop("Unknown cluster_type: ", cluster_type)
}
```

## Step 4.5: Module / program validation

Both clustering paths share the same internal-coherence metrics, computed on
the SNP similarity graph:

- **mean_internal_similarity** — mean off-diagonal `S` inside the module
- **separation** — internal minus external mean similarity
- **connectedness** — fraction of SNP pairs with `|S| >= similarity_threshold`

For EBMF, `summarise_ebmf_programs()` folds the graph-reliability metrics
together with trait-subsampling **stability** and the descriptive
**additional scores** into a single per-program table with pass/fail flags and
an overall `status`:

- **factor_strength / factor_strength_per_snp** — proportion of observed
  matrix signal attributable to the factor (from its fitted reconstruction),
  computed on the lFSR/magnitude-filtered subset of EBMF-supported SNPs;
  `factor_strength_per_snp = factor_strength / sqrt(n_snps_filtered)`. Reported for
  context only; it does not gate `status`.
- **replication** — trait-subsampling stability: how often the top
  `stability_top_n` loading SNPs are recovered when `1 - frac_traits` of the
  traits are held out and flashier is refit (`n_rep` reruns, run in parallel
  on `cores` cores). Because this is expensive, it is only computed for
  programs that pass the internal-coherence checks; those programs must reach
  `replication >= stability_threshold`. Stability should use the same filtered
  SNP reference sets as the other program-level calculations, and the
  refitted programs are filtered with the same lFSR/magnitude thresholds.
- **posterior_evidence** — aggregate (`posterior_stat`) of `-log10(lfsr)`
  across assigned SNPs. Distinguishes programs dominated by borderline
  `lfsr ~ 0.05` memberships from programs whose memberships carry much stronger
  posterior sign evidence.
- **max_abs_factor_corr / redundant** — maximum absolute correlation of the
  program's SNP loading vector with any other program; programs exceeding
  `redundancy_threshold` are flagged `redundant`.

A program is `valid` if it passes all three filters (`size`, `internal`,
`stability`); the failure reason is recorded in `status`.

```{r step4-quality, results='asis'}
# if (cluster_type == "louvain") {
#   module_quality <- summarise_snp_module_quality(
#     S,
#     clusters_fit$cluster,
#     edge_threshold = similarity_threshold,
#     min_module_size = min_module_size,
#     min_mean_internal = min_mean_internal,
#     min_connectedness = min_connectedness
#   )
#   program_summary <- NULL

#   knitr::kable(
#     module_quality |>
#       dplyr::select(
#         cluster, n_snps, mean_internal_similarity, mean_external_similarity,
#         separation, connectedness
#       ) |>
#       dplyr::filter(n_snps >= min_module_size) |>
#       dplyr::mutate(dplyr::across(
#         c(
#           mean_internal_similarity, mean_external_similarity,
#           separation, connectedness
#         ),
#         ~ round(.x, 3)
#       )),
#     caption = "Louvain module reliability metrics",
#     row.names = FALSE
#   )

#   reliable_ids <- module_quality$cluster[module_quality$reliable]
#   clusters_analysis <- clusters_fit$cluster[
#     clusters_fit$cluster %in% reliable_ids
#   ]
# } else {
  program_summary <- summarise_ebmf_programs(
    ebmf_result,
    s_matrix = S,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    stability_threshold = stability_threshold,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = TRUE
  )

  valid_ids <- program_summary$programs$program[
    program_summary$programs$status == "valid"
  ]

  print(
    knitr::kable(
      program_summary$programs |>
        dplyr::select(
          program, n_snps, n_snps_filtered,
          mean_internal_similarity, connectedness,
          factor_strength, factor_strength_per_snp, replication,
          posterior_evidence, max_abs_factor_corr, redundant,
          size_pass, internal_pass,
          stability_pass, stability_checked, status
        ) |>
        dplyr::mutate(
          factor_strength_pct = round(100 * factor_strength, 2),
          factor_strength_per_snp_pct = round(100 * factor_strength_per_snp, 3),
          mean_internal_similarity = round(mean_internal_similarity, 3),
          connectedness = round(connectedness, 3),
          replication = round(replication, 3),
          posterior_evidence = round(posterior_evidence, 2),
          max_abs_factor_corr = round(max_abs_factor_corr, 3),
          replication = dplyr::if_else(
            is.na(replication), "--", as.character(replication)
          ),
          redundant = dplyr::if_else(redundant, "yes", "no"),
          size_pass = dplyr::if_else(size_pass, "pass", "fail"),
          internal_pass = dplyr::if_else(internal_pass, "pass", "fail"),
          stability_pass = dplyr::case_when(
            !stability_checked ~ "n/a (not checked)",
            stability_pass ~ "pass",
            TRUE ~ "fail"
          )
        ) |>
        dplyr::select(
          program, n_snps, n_snps_filtered,
          mean_internal_similarity, connectedness,
          factor_strength_pct, factor_strength_per_snp_pct, replication, posterior_evidence,
          max_abs_factor_corr, redundant,
          size_pass, internal_pass, stability_pass, status
        ),
      row.names = FALSE,
      caption = paste0(
        "All EBMF programs returned by flash, with each filter check. n_snps ",
        "is the number of SNPs with nonzero EBMF support and n_snps_filtered is ",
        "the subset surviving the lFSR/magnitude gate; ",
        "factor_strength (as % of observed matrix signal) and ",
        "factor_strength/sqrt(n_snps_filtered) are reported for context but do not gate ",
        "status; replication is only checked for programs passing internal ",
        "coherence (otherwise --); status is 'valid' or lists the failed checks ",
        "(size; internal; stability)."
      )
    )
  )

  print(
    knitr::kable(
      program_summary$programs |>
        dplyr::summarise(
          n_programs = dplyr::n(),
          n_valid = sum(status == "valid"),
          n_fail_size = sum(!size_pass),
          n_fail_internal = sum(!internal_pass),
          n_fail_stability = sum(!stability_pass),
          n_redundant = sum(redundant, na.rm = TRUE)
        ),
      row.names = FALSE,
      caption = "EBMF program validation summary (factor strength is reported but does not gate status)"
    )
  )

  soft_memberships <- program_summary$assigned |>
    dplyr::filter(program %in% valid_ids)
  clusters_analysis <- soft_memberships |>
    dplyr::transmute(
      snp_id = snp_id,
      group = as.character(program)
    )
# }
```

```{r step4-heatmap, fig.height = 7, fig.width = 7}
if (is.data.frame(clusters_analysis) && nrow(clusters_analysis) >= 2) {
  if (is.data.frame(clusters_analysis)) {
    heatmap_groups <- stats::setNames(
      clusters_analysis$group[!duplicated(clusters_analysis$snp_id)],
      clusters_analysis$snp_id[!duplicated(clusters_analysis$snp_id)]
    )
  } else {
    heatmap_groups <- clusters_analysis
  }
  snp_order <- names(heatmap_groups)[order(heatmap_groups)]
  snp_order <- intersect(snp_order, rownames(S))
  if (length(snp_order) < 2) {
    cat("Fewer than 2 reported SNPs are present in the similarity matrix; skipping heatmap.\n")
  } else {
    heatmap(
      S[snp_order, snp_order, drop = FALSE],
      Rowv = NA,
      Colv = NA,
      scale = "none",
      main = sprintf(
        "SNP similarity (%s)",
        if (cluster_type == "ebmf") "valid EBMF programs" else
          paste("reliable", cluster_type, "modules")
      ),
      xlab = "SNPs",
      ylab = "SNPs"
    )
  }
} else {
  cat("Fewer than 2 SNPs in reliable modules; skipping heatmap.\n")
}
```

### Program overlap

Because EBMF programs are soft, the same SNP can belong to several programs.
The UpSet plot below shows, for every observed intersection of programs, how
many SNPs are shared by exactly that combination (a Venn diagram generalised to
any number of programs). The left-hand bar chart is each program's size; the
dot matrix marks which programs each intersection involves.

```{r step4-overlap, fig.height = 5.5, fig.width = 9}
if (is.data.frame(clusters_analysis) && nrow(clusters_analysis) >= 2) {
  members_df <- if (is.data.frame(clusters_analysis)) {
    clusters_analysis |>
      dplyr::select(snp_id, group) |>
      dplyr::distinct()
  } else {
    data.frame(
      snp_id = names(clusters_analysis),
      group = as.character(clusters_analysis),
      stringsAsFactors = FALSE
    )
  }
  n_programs <- dplyr::n_distinct(members_df$group)
  if (n_programs >= 2) {
    membership <- members_df |>
      dplyr::mutate(value = 1L) |>
      tidyr::pivot_wider(
        id_cols = "snp_id",
        names_from = "group",
        names_prefix = "P",
        values_from = "value",
        values_fill = 0L
      )
    set_order <- names(sort(colSums(membership[, -1, drop = FALSE]),
                            decreasing = TRUE))
    if (n_programs > 12) {
      cat("More than 12 programs; showing the 12 largest only.\n")
      set_order <- set_order[seq_len(12)]
    }
    plot_ok <- tryCatch({
      membership_plot <- as.data.frame(membership[, c("snp_id", set_order), drop = FALSE])
      rownames(membership_plot) <- membership_plot$snp_id
      membership_plot$snp_id <- NULL
      print(UpSetR::upset(
        membership_plot,
        sets = set_order,
        order.by = "freq",
        nsets = length(set_order),
        nintersects = 40,
        empty.intersections = "off",
        main.bar.color = "#1b9e77",
        sets.bar.color = "#7570b3",
        point.size = 2.6,
        line.size = 0.6
      ))
      TRUE
    }, error = function(e) {
      cat("UpSet plot failed (", conditionMessage(e),
          "); showing pairwise SNP-overlap counts instead.\n")
      FALSE
    })
    if (!plot_ok) {
      sizes <- colSums(membership[, set_order, drop = FALSE])
      overlaps <- outer(set_order, set_order, Vectorize(function(a, b) {
        sum(membership[[a]] & membership[[b]])
      }))
      dimnames(overlaps) <- list(set_order, set_order)
      diag(overlaps) <- sizes[set_order]
      print(knitr::kable(
        as.data.frame(overlaps),
        row.names = TRUE,
        caption = "Pairwise SNP overlap between programs (diagonal = program size)"
      ))
    }
  } else {
    cat("Only one program; skipping program-overlap plot.\n")
  }
} else {
  cat("Fewer than 2 SNPs in reliable modules; skipping program-overlap plot.\n")
}
```

## Step 5: Interpret reliable modules

For each reliable module, rank background traits by **module specificity vs the
rest of the trait** (`summarise_module_specific_traits()`), then summarise
**trait-category and tissue enrichments vs the trait baseline**, plus
pathway hits. `specificity = mean |z| in the module / (mean |z| in the rest +
0.1)` and `abs_mean_z_delta = module - rest` highlight traits that are
specifically elevated in this module rather than merely large everywhere (which
ranking by `abs_mean_z` alone would over-emphasise). Pathway enrichment is
queried across Reactome, KEGG, and Human Phenotype Ontology (HP) with a
separate call per source: Reactome and KEGG hits are reported side by side as
**pathways**, and HP terms as **phenotypes**, so the three never compete for
the same headline.

Trait baseline = all SNPs attributed to this trait in the clustering analysis
(`colnames(S)`; for Louvain this equals `names(clusters$cluster)`, and for EBMF
it is kept identical so both methods are annotated against the same baseline),
using coloc rows for those SNPs only—not
the whole GPMap universe. Each module is compared to the **rest of the trait**
(leave-one-out): observed link share vs expected share, with fold enrichment.

Annotation tables show the top elevated categories/tissues even when FDR is not
significant; FDR is reported when it passes 0.05. Hits that are significant by
any of the three methods at 0.05 are shown in **bold**.

Each hit is scored with three BH-corrected significance summaries, all
thresholded at 0.05:

- **FDR** — hypergeometric test of the module vs the rest of the trait.
- **perm** — permutation p-values from shuffling link labels across the
  baseline, BH-corrected within the module (`fdr_perm`).
- **vsmod** — one-sided Fisher test of the module against the pooled links of
  the *other* modules, BH-corrected within the module (`fdr_modules`).

They agree on strong signal and differ only on marginal, low-count calls. The
bootstrap `CI` for the fold enrichment shows how precisely the ratio is
estimated; because modules are small, a big ratio built on a few links carries
a wide interval that includes 1 (the null), so read the CI together with the
FDR rather than alone.

```{r step5-interpret}
baseline_snp_ids <- colnames(S)

has_analysis_groups <- if (is.data.frame(clusters_analysis)) {
  nrow(clusters_analysis) > 0
} else {
  length(clusters_analysis) > 0
}

if (!has_analysis_groups) {
  cat("No reliable modules to interpret.\n")
  group_traits <- NULL
  group_categories <- NULL
  group_tissues <- NULL
  module_enrichment <- NULL
} else {
    group_traits <- summarise_module_specific_traits(
    trait_matrix = trait_matrix,
    groups = clusters_analysis,
    trait_info = pleiotropy$trait_info,
    coloc_groups = cg,
    exclude_trait_ids = target_trait_id,
    min_group_size = min_module_size,
      top_n = top_n_traits,
      n_categories = 3L,
      min_specificity = min_specificity
  )
  group_categories <- enrich_snp_group_trait_categories(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    baseline_snp_ids = baseline_snp_ids
  )
  group_tissues <- enrich_snp_group_tissues(
    groups = clusters_analysis,
    coloc_groups = cg,
    snp_key = "variant_id",
    min_group_size = min_module_size,
    background = "trait",
    baseline_snp_ids = baseline_snp_ids
  )
  module_enrichment <- enrich_snp_group_pathways(
    groups = clusters_analysis,
    coloc_groups = cg,
    min_group_size = min_module_size,
    sources = c("KEGG", "Reactome", "HP"),
    include_situated_gene = include_situated_gene
  )
}

.format_enrichment_hits <- function(comparison, label_col, max_n = 3L) {
  if (is.null(comparison) || nrow(comparison) == 0) {
    return("none")
  }
  # Prefer elevated vs rest-of-trait; fall back to most abundant if none.
  elevated <- comparison[
    is.finite(comparison$fold_enrichment) & comparison$fold_enrichment >= 1,
    ,
    drop = FALSE
  ]
  if (nrow(elevated) == 0) {
    elevated <- comparison
  } else {
    elevated <- elevated[order(
      -elevated$fold_enrichment,
      elevated$fdr,
      elevated[[label_col]]
    ), , drop = FALSE]
  }
  top <- utils::head(elevated, max_n)
  labels <- vapply(seq_len(nrow(top)), function(i) {
    tag <- .method_tag(top$fdr[i], top$fdr_perm[i], top$fdr_modules[i])
    ci <- if (is.finite(top$fe_ci_lower[i]) && is.finite(top$fe_ci_upper[i])) {
      paste0(
        ", CI=[", signif(top$fe_ci_lower[i], 2),
        ",", signif(top$fe_ci_upper[i], 2), "]"
      )
    } else {
      ""
    }
    fdr_txt <- if (nzchar(tag)) paste0(", ", tag) else ""
    label <- top[[label_col]][i]
    if (nzchar(tag)) {
      label <- paste0("<b>", label, "</b>")
    }
    paste0(
      label,
      " (obs=", sprintf("%.0f%%", 100 * top$frac_group[i]),
      ", exp=", sprintf("%.0f%%", 100 * top$frac_universe[i]),
      ", FE=", signif(top$fold_enrichment[i], 2),
      ci,
      fdr_txt,
      ")"
    )
  }, character(1))
  paste(labels, collapse = "<br>")
}

.format_pathway_hits <- function(pathways, max_n = 3L, with_source = TRUE) {
  if (is.null(pathways) || nrow(pathways) == 0) {
    return("none")
  }
  top <- utils::head(pathways, max_n)
  labels <- if (with_source) {
    paste0(top$source, ": ", top$description)
  } else {
    top$description
  }
  if ("fdr" %in% names(top)) {
    labels <- paste0(labels, " (FDR=", signif(top$fdr, 3), ")")
  }
  paste(labels, collapse = "<br>")
}
```

```{r step5-module-overview, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  specific_traits <- dplyr::bind_rows(lapply(group_traits$by_group, function(x) {
    passing <- x$traits |>
      dplyr::filter(passes_min_specificity) |>
      utils::head(3L)
    data.frame(
      group = x$group,
      module_specific_traits = if (nrow(passing) == 0) {
        "none"
      } else {
        paste0(
          passing$trait_name,
          " (", round(passing$specificity, 2), ")",
          collapse = "<br>"
        )
      },
      stringsAsFactors = FALSE
    )
  }))

  # SNPs that belong to exactly one program: not shared with any other program.
  members_df <- if (is.data.frame(clusters_analysis)) {
    clusters_analysis |>
      dplyr::select(snp_id, group) |>
      dplyr::distinct()
  } else {
    data.frame(
      snp_id = names(clusters_analysis),
      group = as.character(clusters_analysis),
      stringsAsFactors = FALSE
    )
  }
  snp_group_counts <- members_df |>
    dplyr::count(snp_id, name = "n_groups")
  unique_by_group <- members_df |>
    dplyr::filter(
      snp_id %in% snp_group_counts$snp_id[snp_group_counts$n_groups == 1]
    ) |>
    dplyr::count(group, name = "n_unique_snps")

  overview <- dplyr::full_join(
    group_traits$summary |>
      dplyr::select(group, n_snps),
    unique_by_group,
    by = "group"
  ) |>
    dplyr::full_join(
      group_categories$summary |>
        dplyr::select(group, top_enriched_category_any),
      by = "group"
    ) |>
    dplyr::full_join(
    group_tissues$summary |>
      dplyr::select(group, top_enriched_tissue_any),
    by = "group"
  ) |>
    dplyr::full_join(
      module_enrichment$summary |>
        dplyr::select(group, top_enriched_reactome, top_enriched_kegg),
      by = "group"
    ) |>
    dplyr::full_join(
      module_enrichment$summary |>
        dplyr::select(group, top_enriched_phenotype),
      by = "group"
    ) |>
    dplyr::left_join(specific_traits, by = "group") |>
    dplyr::mutate(
      n_unique_snps = as.integer(tidyr::replace_na(n_unique_snps, 0L)),
      top_enriched_category_any = tidyr::replace_na(top_enriched_category_any, "none"),
      top_enriched_tissue_any = tidyr::replace_na(top_enriched_tissue_any, "none"),
      top_enriched_reactome = tidyr::replace_na(top_enriched_reactome, "none"),
      top_enriched_kegg = tidyr::replace_na(top_enriched_kegg, "none"),
      top_enriched_phenotype = tidyr::replace_na(top_enriched_phenotype, "none")
    ) |>
    dplyr::arrange(dplyr::desc(n_snps), as.integer(group)) |>
    dplyr::select(
      group, n_snps, n_unique_snps, module_specific_traits, dplyr::everything()
    )

  cat("\n\n### Overview of significant annotations across modules\n\n")
  print(knitr::kable(
    overview,
    caption = "Top module-specific traits and significant category, tissue, pathway (Reactome and KEGG separately) and phenotype (HP) annotations; n_snps = SNPs in the program, n_unique_snps = SNPs not present in any other program; trait values are specificity scores, and enrichment method tags are shown when FDR, permutation FDR, or module-vs-modules Fisher passes 0.05",
    row.names = FALSE,
    escape = FALSE
  ))
}
```

```{r step5-module-panels, echo=FALSE, results='asis'}
if (!is.null(group_traits) && length(group_traits$by_group) > 0) {
  categories_by_group <- stats::setNames(
    group_categories$by_group,
    vapply(
      group_categories$by_group,
      function(x) as.character(x$group),
      character(1)
    )
  )
  tissues_by_group <- stats::setNames(
    group_tissues$by_group,
    vapply(group_tissues$by_group, function(x) as.character(x$group), character(1))
  )
  pathways_by_group <- stats::setNames(
    module_enrichment$by_group,
    vapply(module_enrichment$by_group, function(x) as.character(x$group), character(1))
  )

  for (x in group_traits$by_group) {
    grp <- as.character(x$group)
    cat("\n\n### Module ", grp, " (", x$n_snps, " SNPs)\n\n", sep = "")

    category_x <- categories_by_group[[grp]]
    tissue_x <- tissues_by_group[[grp]]
    pathway_x <- pathways_by_group[[grp]]
    pathway_sub <- function(src) {
      if (is.null(pathway_x)) {
        NULL
      } else {
        pathway_x$pathways[
          pathway_x$pathways$source == src,
          ,
          drop = FALSE
        ]
      }
    }
    annotation <- data.frame(
      module = paste0("Module ", grp, " (", x$n_snps, " SNPs)"),
      trait_categories = .format_enrichment_hits(
          if (is.null(category_x)) NULL else category_x$comparison,
          "trait_category"
        ),
      tissues = .format_enrichment_hits(
          if (is.null(tissue_x)) NULL else tissue_x$comparison,
          "tissue"
        ),
      reactome = .format_pathway_hits(pathway_sub("Reactome"), with_source = FALSE),
      kegg = .format_pathway_hits(pathway_sub("KEGG"), with_source = FALSE),
      phenotypes = .format_pathway_hits(pathway_sub("HP"), with_source = FALSE),
      stringsAsFactors = FALSE
    )
    print(knitr::kable(
      annotation,
      caption = "Module annotations (enrichment vs trait baseline)",
      row.names = FALSE,
      escape = FALSE
    ))

    traits_to_show <- x$traits |>
      dplyr::filter(passes_min_specificity) |>
      dplyr::select(
        trait_name, trait_category,
        abs_mean_z, abs_mean_z_delta, specificity, mean_z
      ) |>
      dplyr::mutate(
        abs_mean_z = round(abs_mean_z, 3),
        abs_mean_z_delta = round(abs_mean_z_delta, 3),
        specificity = round(specificity, 2),
        mean_z = round(mean_z, 3)
      )

    if (nrow(traits_to_show) > 0) {
      print(knitr::kable(
        traits_to_show,
        caption = paste0(
          "Module-specific traits (specificity >= ", min_specificity,
          "; module mean |z| / (rest mean |z| + 0.1))"
        ),
        row.names = FALSE
      ))
    } else {
      cat("No traits passed the module-specificity cutoff (specificity >= ",
          min_specificity, ").\n\n", sep = "")
    }
  }
}
```

## Step 6: Module typology by SNP pleiotropy

Each SNP in GPMap carries a pleiotropy score: the number of distinct trait
categories and the number of distinct protein-coding genes it is associated
with via coloc groups (`get_all_variant_pleiotropies()`). These describe how
**broad** a SNP's biology is, independent of its effect profile. They give
modules a second axis of description:

- **Breadth** — the mean per-SNP pleiotropy within the module (trait-category
  pleiotropy and gene pleiotropy).
- **Tightness** — separation (internal minus external similarity).

Plotting the two yields a descriptive module typology (not a filter, and not a
reliability criterion):

- **tight + narrow** — a specific biological program (e.g. a tissue module);
- **tight + broad** — a regulatory hub affecting many categories or genes;
- **loose + broad** — generic background shared across the trait;
- **loose + narrow** — a private, low-signal module.

```{r step6-pleiotropy-typology, fig.width=7, fig.height=5, out.width="100%"}
variant_pleiotropy <- get_all_variant_pleiotropies()
module_pleiotropy <- summarise_snp_group_pleiotropy(
  groups = clusters_analysis,
  variant_pleiotropies = variant_pleiotropy,
  min_group_size = min_module_size
)

module_annotation_status <- group_categories$summary |>
  dplyr::select(group, top_enriched_category_any) |>
  dplyr::full_join(
    group_tissues$summary |>
      dplyr::select(group, top_enriched_tissue_any),
    by = "group"
  ) |>
  dplyr::full_join(
    module_enrichment$summary |>
      dplyr::select(
        group, top_enriched_reactome, top_enriched_kegg, top_enriched_phenotype
      ),
    by = "group"
  ) |>
  dplyr::mutate(
    annotated = !(is.na(top_enriched_category_any) &
      is.na(top_enriched_tissue_any) &
      is.na(top_enriched_reactome) &
      is.na(top_enriched_kegg) &
      is.na(top_enriched_phenotype))
  ) |>
  dplyr::select(group, annotated)

quality_for_typology <- if (cluster_type == "louvain") {
  module_quality |>
    dplyr::mutate(cluster = as.character(cluster)) |>
    dplyr::select(cluster, separation, mean_internal_similarity, connectedness)
} else {
  program_summary$programs |>
    dplyr::mutate(cluster = as.character(program)) |>
    dplyr::select(cluster, separation, mean_internal_similarity, connectedness)
}

typology <- module_pleiotropy |>
  dplyr::left_join(quality_for_typology, by = c("group" = "cluster")) |>
  dplyr::left_join(module_annotation_status, by = "group")

typology_long <- typology |>
  tidyr::pivot_longer(
    c(mean_trait_category_pleiotropy, mean_gene_pleiotropy),
    names_to = "metric",
    values_to = "pleiotropy"
  ) |>
  dplyr::mutate(
    metric = dplyr::if_else(
      metric == "mean_trait_category_pleiotropy",
      "trait-category pleiotropy (distinct categories per SNP)",
      "gene pleiotropy (distinct genes per SNP)"
    )
  )

if (nrow(typology_long) >= 2) {
  medians <- typology_long |>
    dplyr::group_by(metric) |>
    dplyr::summarise(mx = stats::median(separation), my = stats::median(pleiotropy))

  print(ggplot2::ggplot(
    typology_long,
    ggplot2::aes(x = separation, y = pleiotropy, size = n_snps, color = annotated)
  ) +
    ggplot2::geom_vline(
      data = medians,
      ggplot2::aes(xintercept = mx),
      linetype = "dashed",
      alpha = 0.5
    ) +
    ggplot2::geom_hline(
      data = medians,
      ggplot2::aes(yintercept = my),
      linetype = "dashed",
      alpha = 0.5
    ) +
    ggplot2::geom_point(alpha = 0.75) +
    ggplot2::scale_size_area(max_size = 8) +
    ggplot2::facet_wrap(~metric, scales = "free_y") +
    ggplot2::scale_color_manual(
      values = c(`FALSE` = "grey60", `TRUE` = "#c2571b")
    ) +
    ggplot2::labs(
      title = "Module typology: separation vs SNP pleiotropy breadth",
      subtitle = "Dashed lines at medians; colour = module has a significant annotation",
      x = "separation (internal minus external similarity)",
      y = "mean per-SNP pleiotropy",
      size = "SNPs in module",
      color = "annotated"
    ) +
    ggplot2::theme_minimal())
}
```
