---
title: "Investigation of univariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(igraph)
library(mclust)
```

This vignette builds a bivariate/univariate pleiotropy analysis from GPMap colocalisation data.

## Step 1: Construct the pleiotropy matrix (X)

**Goal:** Build a matrix where each **column** is a target-trait SNP and each **row** is a background trait. Values are signed z-scores (`beta / se`) describing that trait's association at the SNP's colocalisation locus — a phenome-wide fingerprint per SNP.

We use **HDL cholesterol** (trait id `1630`) as the target trait.

```{r step1-fetch}
target_trait_id <- 1630

target_trait <- trait(
  target_trait_id,
  include_associations = TRUE
)
cg <- target_trait$coloc_groups
```

### Target SNPs

SNPs are taken from loci for the target trait (for now, all loci).

```{r step1-target-snps}
target_snps <- cg |>
  dplyr::filter(
    trait_id == target_trait_id,
    !is.na(beta),
    !is.na(se),
    se > 0
  ) |>
  dplyr::distinct(variant_id, display_snp, coloc_group_id, chr, bp)

```

### Build matrix X (traits x SNPs)

`build_pleiotropy_matrix()` extracts z-scores for every trait that colocalises at each target SNP locus and arranges them into a traits-by-SNPs matrix.

```{r step1-pleiotropy-matrix}
pleiotropy <- build_pleiotropy_matrix(
  trait_id = target_trait_id,
  coloc_groups = cg,
  snp_key = "variant_id"
)

X <- pleiotropy$x_matrix

list(
  target_trait = target_trait$trait$trait_name,
  n_traits = nrow(X),
  n_snps = ncol(X),
  fraction_non_na = mean(!is.na(X))
)
```

Rows are background traits (`n_trait`); columns are target SNPs (`n_snp`). Entries are `NA` when a trait has no colocalisation signal at that SNP's locus.

```{r step1-dim}
dim(X)
```

Each column of `X` is one SNP's signed association profile across the colocalising phenome.

## Step 2: Orient profiles relative to the target trait (X\*)

**Goal:** Remove arbitrary allele-swapping noise so each SNP profile reflects what happens to the phenome when that variant **increases liability** for the target trait.

For each SNP column \(j\):

\[
X^*_{tj} = \mathrm{sign}(z_j) \cdot X_{tj}
\]

where \(z_j\) is the target trait's signed z-score at SNP \(j\).

```{r step2-z-target}
z_target <- X[as.character(target_trait_id), ]

head(z_target)
```

```{r step2-signs}
target_signs <- sign(z_target)

table(target_signs, useNA = "ifany")
```

```{r step2-orient}
X_star <- sweep(X, 2, target_signs, `*`)

# equivalent helper:
oriented <- orient_pleiotropy_matrix(X, target_trait_id = target_trait_id)
stopifnot(all.equal(X_star, oriented$x_matrix, check.attributes = FALSE))
```

After orientation, the target trait row is non-negative (magnitude preserved, sign unified). Background trait effects are flipped where needed so they align with the target's direction of effect.

```{r step2-check}
target_row <- X_star[as.character(target_trait_id), ]

list(
  n_negative_target_z = sum(target_row < 0, na.rm = TRUE),
  n_positive_target_z = sum(target_row > 0, na.rm = TRUE),
  mean_abs_target_z = mean(abs(target_row), na.rm = TRUE)
)
```

## Step 3: Calculate the SNP-SNP similarity matrix (S)

**Goal:** Measure how parallel the multi-trait trajectories are between every pair of SNPs. Cosine similarity scales each profile to unit length so highly pleiotropic variants do not dominate.

For SNPs \(j\) and \(k\):

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\| \|x^*_k\|}
\]

```{r step3-col-norms}
X_star_filled <- X_star
X_star_filled[is.na(X_star_filled)] <- 0

col_norms <- sqrt(colSums(X_star_filled^2))

summary(col_norms)
```

```{r step3-normalize}
zero_cols <- col_norms == 0
col_norms_div <- col_norms
col_norms_div[zero_cols] <- 1

X_star_norm <- sweep(X_star_filled, 2, col_norms_div, "/")
X_star_norm[, zero_cols] <- 0
```

```{r step3-similarity}
S <- crossprod(X_star_norm)

# equivalent helper:
similarity <- snp_similarity_matrix(X_star)
stopifnot(all.equal(S, similarity$s_matrix, check.attributes = FALSE))
```

`S` is symmetric with values in \([-1, 1]\). The diagonal is 1 (each SNP is identical to itself).

```{r step3-dim}
dim(S)
```

```{r step3-summary}
list(
  diag_all_one = all(abs(diag(S) - 1) < 1e-10),
  min_similarity = min(S[upper.tri(S)], na.rm = TRUE),
  max_similarity = max(S[upper.tri(S)], na.rm = TRUE),
  mean_similarity = mean(S[upper.tri(S)], na.rm = TRUE)
)
```

Interpretation:

- `diag_all_one`: the diagonal of `S` should be 1
- `min_similarity`: strong anti-parallelism: pairs of SNPs where one variant drives background traits in the exact opposite direction of the other, even though they both ultimately cause the target trait to increase.
- `max_similarity`: If high, there are independent SNPs in your dataset that possess near identical pleiotropic footprints
- `mean_similarity`: On average, how any two random SNPs chosen from your total pool are similar to each other.

If there is high max similarity and low mean similarity, this suggests that there is mechanistic heterogeneity in the SNPs.

## Step 4: Cluster the SNPs

**Goal:** Group variants into distinct functional classes based on the geometric parallelism of their pleiotropy profiles.

We compare four clustering approaches:

1. **Hierarchical clustering** on distance `1 - S` (average linkage)
2. **Spectral clustering** on the similarity graph
3. **Community detection** (Louvain) on a weighted similarity graph
4. **Gaussian mixture models** in latent space (PCA of unit-normalised profiles + `mclust`)

### Hierarchical clustering

```{r step4-hierarchical}
k <- 3
D <- 1 - S
dist_snp <- as.dist(D)

hc <- hclust(dist_snp, method = "average")
clusters_hierarchical <- cutree(hc, k = k)

table(clusters_hierarchical)

# equivalent helper:
hclust_result <- cluster_snp_profiles(S, k = k, method = "hierarchical")
stopifnot(identical(clusters_hierarchical, hclust_result$cluster))
```

```{r step4-dendrogram, fig.height = 6, fig.width = 10}
plot(hc, labels = names(clusters_hierarchical), cex = 0.4, main = "SNP hierarchical clustering")
rect.hclust(hc, k = k, border = 2:4)
```

### Spectral clustering

```{r step4-spectral}
clusters_spectral <- cluster_snp_profiles(S, k = k, method = "spectral")

table(clusters_spectral$cluster)
```

### Community detection

Louvain finds a data-driven number of modules (does not take a fixed `k`). Leading-eigenvector clustering performs `k - 1` recursive graph splits (`steps = k - 1` in **igraph**); the resulting module count may be fewer than `k` if a split is not supported.

```{r step4-community}
clusters_louvain <- cluster_snp_profiles(
  S,
  method = "community",
  community_algorithm = "louvain"
)

clusters_leading_eigen <- cluster_snp_profiles(
  S,
  k = k,
  method = "community",
  community_algorithm = "leading_eigen"
)

list(
  louvain_n_modules = clusters_louvain$n_clusters,
  louvain_sizes = table(clusters_louvain$cluster),
  leading_eigen_sizes = table(clusters_leading_eigen$cluster)
)
```

### Gaussian mixture models in latent space

SNP coordinates are taken from a PCA of the unit-normalised oriented profiles (`traits × SNPs` transposed to `SNPs × traits`), then `mclust` fits a `k`-component GMM.

```{r step4-gmm}
clusters_gmm <- cluster_snp_profiles(
  S,
  k = k,
  method = "gmm",
  x_matrix = X_star
)

table(clusters_gmm$cluster)
```

### Compare methods

```{r step4-compare}
comparison <- data.frame(
  snp_id = names(clusters_hierarchical),
  hierarchical = clusters_hierarchical,
  spectral = clusters_spectral$cluster,
  leading_eigen = clusters_leading_eigen$cluster,
  louvain = clusters_louvain$cluster[names(clusters_hierarchical)],
  gmm = clusters_gmm$cluster,
  stringsAsFactors = FALSE
)

head(comparison, 10)
```

```{r step4-agreement}
# pairwise agreement (identical assignment rate)
methods <- c("hierarchical", "spectral", "leading_eigen", "gmm")
agreement <- outer(methods, methods, Vectorize(function(a, b) {
  mean(comparison[[a]] == comparison[[b]])
}))

rownames(agreement) <- methods
colnames(agreement) <- methods

round(agreement, 2)
```

High agreement between methods suggests robust functional modules; disagreement may reflect SNPs on cluster boundaries or methods with different assumptions (e.g. Louvain's variable module count).

For interpretation below we use the hierarchical clusters (`clusters_hierarchical`) and the dendrogram `hc` from Step 4.

## Step 5: Interpret biological modules

**Goal:** Identify which background pathways or clinical traits drive each clustered SNP module.

For each cluster, we take the mean oriented profile across member SNPs (rows of `X_star`) and rank traits by absolute magnitude.

```{r step5-summarise}
module_drivers <- summarise_snp_modules(
  x_matrix = X_star,
  cluster = clusters_hierarchical,
  trait_info = pleiotropy$trait_info,
  exclude_trait_id = target_trait_id,
  top_n = 10
)

module_drivers
```

```{r step5-by-cluster}
module_drivers |>
  dplyr::group_by(cluster) |>
  dplyr::slice_head(n = 5) |>
  dplyr::select(cluster, trait_name, mean_z, abs_mean_z, n_snps) |>
  dplyr::mutate(
    mean_z = round(mean_z, 2),
    abs_mean_z = round(abs_mean_z, 2)
  )
```

Positive `mean_z` means traits tend to increase when the module's SNPs increase target-trait liability; negative values indicate inverse associations.

### Heatmap of SNP similarity

```{r step5-heatmap, fig.height = 7, fig.width = 7}
heatmap(
  S,
  Rowv = as.dendrogram(hc),
  Colv = as.dendrogram(hc),
  scale = "none",
  main = "SNP functional clustering",
  xlab = "SNPs",
  ylab = "SNPs"
)
```

The dendrogram ordering groups SNPs with parallel pleiotropy footprints; block structure along the diagonal indicates distinct functional modules.