---
title: "Investigation of collapsable bivariate analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of collapsable bivariate analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
```

Collapsable bivariate analysis for two target traits: molecular QTL rows are collapsed into pathway/gene **features**, SNPs are oriented per trait, cross-trait similarity `S^*` is computed, and **signed Louvain** groups SNPs on each side. Rectangular blocks are classified as concordant, discordant, or trait-specific. Compare with [Investigation of bivariate analysis](investigation-bivariate-analysis.html) (trait rows only, same traits and parameters).

We use **BMI** (trait id `1992`) and **Waist-to-hip ratio adjusted for BMI** (trait id `1653`).

```{r params}
trait_id_1 <- 1992
trait_id_2 <- 1653
min_module_size <- 3
min_group_size <- 5
min_snp_signals <- 5L
max_snp_fraction <- 0.8
similarity_threshold <- 0.5
louvain_gamma <- 3

traits_data <- traits(c(trait_id_1, trait_id_2), include_associations = TRUE)
cg <- traits_data$coloc_groups
trait_1_name <- traits_data$traits$trait_name[traits_data$traits$id == trait_id_1]
trait_2_name <- traits_data$traits$trait_name[traits_data$traits$id == trait_id_2]
```

## Verification 1: baseline pathway enrichment

Before clustering, enrich molecular QTL genes at each trait's GWAS SNPs (KEGG + Reactome).

```{r verify1-baseline}
baseline_1 <- enrich_trait_pathways(
  trait_id = trait_id_1,
  coloc_groups = cg,
  sources = c("KEGG", "Reactome")
)
baseline_2 <- enrich_trait_pathways(
  trait_id = trait_id_2,
  coloc_groups = cg,
  sources = c("KEGG", "Reactome")
)

knitr::kable(
  dplyr::bind_rows(baseline_1$summary, baseline_2$summary),
  caption = "Baseline trait pathway enrichment (KEGG + Reactome)",
  na = "",
  align = c("r", "r", "r", "r", "l")
)

if (nrow(baseline_1$pathways) > 0) {
  knitr::kable(
    baseline_1$pathways |>
      dplyr::mutate(p_value = signif(p_value, 3), fdr = signif(fdr, 3)),
    caption = paste0(trait_1_name, ": enriched pathways"),
    row.names = FALSE
  )
}
if (nrow(baseline_2$pathways) > 0) {
  knitr::kable(
    baseline_2$pathways |>
      dplyr::mutate(p_value = signif(p_value, 3), fdr = signif(fdr, 3)),
    caption = paste0(trait_2_name, ": enriched pathways"),
    row.names = FALSE
  )
}
```

## Step 1: Feature pleiotropy matrices

`build_bivariate_feature_pleiotropy_matrices()` uses `trait(..., include_full_associations = TRUE)` for dense betas (`z = beta / se`), pathway collapse (size 5–200, no FDR cut), and aligned feature rows across both traits.

```{r step1-build}
bivariate <- build_bivariate_feature_pleiotropy_matrices(
  trait_id_1 = trait_id_1,
  trait_id_2 = trait_id_2,
  coloc_groups = cg,
  snp_key = "variant_id",
  pathway_min_size = 5,
  pathway_max_size = 200,
  minimum_count_in_network = 2
)

X1 <- bivariate$x1_matrix
X2 <- bivariate$x2_matrix

knitr::kable(
  data.frame(
    trait_1 = trait_1_name,
    trait_2 = trait_2_name,
    n_features = nrow(X1),
    n_trait_rows = sum(bivariate$feature_info$feature_type == "trait"),
    n_pathway_rows = sum(bivariate$feature_info$feature_type == "pathway"),
    n_gene_rows = sum(bivariate$feature_info$feature_type == "gene"),
    n_snps_trait_1 = ncol(X1),
    n_snps_trait_2 = ncol(X2)
  ),
  caption = "Feature matrix summary",
  align = "r"
)
```

## Step 1.5: Filter sparse and ubiquitous features

Drop shared feature rows with fewer than `min_snp_signals` non-`NA` values in **both** matrices, or with non-`NA` fraction above `max_snp_fraction` in **either** matrix (target trait rows kept).

```{r step1-sparse-filter}
protected_ids <- c(as.character(trait_id_1), as.character(trait_id_2))
feature_snp_counts_1 <- rowSums(!is.na(X1))
feature_snp_counts_2 <- rowSums(!is.na(X2))
feature_snp_frac_1 <- feature_snp_counts_1 / ncol(X1)
feature_snp_frac_2 <- feature_snp_counts_2 / ncol(X2)

sparse_feature_ids <- names(feature_snp_counts_1)[
  (feature_snp_counts_1 < min_snp_signals) &
    (feature_snp_counts_2 < min_snp_signals)
]
ubiquitous_feature_ids <- names(feature_snp_frac_1)[
  (feature_snp_frac_1 > max_snp_fraction) |
    (feature_snp_frac_2 > max_snp_fraction)
]
drop_feature_ids <- setdiff(union(sparse_feature_ids, ubiquitous_feature_ids), protected_ids)

keep_rows <- !rownames(X1) %in% drop_feature_ids
X1 <- X1[keep_rows, , drop = FALSE]
X2 <- X2[keep_rows, , drop = FALSE]
bivariate$feature_info <- bivariate$feature_info |>
  dplyr::filter(!feature_id %in% drop_feature_ids)

list(
  min_snp_signals = min_snp_signals,
  max_snp_fraction = max_snp_fraction,
  n_sparse_features_removed = length(setdiff(sparse_feature_ids, protected_ids)),
  n_ubiquitous_features_removed = length(setdiff(ubiquitous_feature_ids, protected_ids)),
  n_features_after = nrow(X1),
  n_snps_trait_1 = ncol(X1),
  n_snps_trait_2 = ncol(X2)
)
```

## Step 2: Orient profiles (X₁\*, X₂\*)

```{r step2-orient}
X1_star <- orient_pleiotropy_matrix(
  X1,
  target_trait_id = trait_id_1,
  z_target = bivariate$z_target_1
)$x_matrix

X2_star <- orient_pleiotropy_matrix(
  X2,
  target_trait_id = trait_id_2,
  z_target = bivariate$z_target_2
)$x_matrix
```

## Step 3: Cross-trait similarity (S\*)

```{r step3-similarity}
S_star <- cross_snp_similarity_matrix(X1_star, X2_star)$s_matrix

list(
  dim = paste(nrow(S_star), "x", ncol(S_star)),
  min_similarity = round(min(S_star, na.rm = TRUE), 2),
  max_similarity = round(max(S_star, na.rm = TRUE), 2),
  mean_similarity = round(mean(S_star, na.rm = TRUE), 2)
)
```

## Step 4: Signed Louvain modules

Louvain runs on cosine similarity of cross-trait **profiles** (rows and columns of `S^*`). Edges with `|similarity| < similarity_threshold` are zeroed before clustering.

```{r step4-cluster}
cross_modules <- cluster_cross_trait_snps(
  S_star,
  method = "louvain",
  similarity_threshold = similarity_threshold,
  gamma = louvain_gamma
)

knitr::kable(
  dplyr::bind_rows(
    data.frame(
      trait = trait_1_name,
      cluster = as.integer(names(table(cross_modules$clusters_rows))),
      n_snps = as.integer(table(cross_modules$clusters_rows)),
      stringsAsFactors = FALSE
    ),
    data.frame(
      trait = trait_2_name,
      cluster = as.integer(names(table(cross_modules$clusters_cols))),
      n_snps = as.integer(table(cross_modules$clusters_cols)),
      stringsAsFactors = FALSE
    )
  ) |>
    dplyr::arrange(trait, dplyr::desc(n_snps)),
  caption = "Signed Louvain module sizes",
  align = c("l", "r", "r")
)
```

```{r step4-heatmap, fig.height = 7, fig.width = 9}
heatmap(
  S_star,
  Rowv = as.dendrogram(cross_modules$hclust_rows),
  Colv = as.dendrogram(cross_modules$hclust_cols),
  scale = "none",
  main = "Cross-trait SNP similarity (Louvain ordering)",
  xlab = paste(trait_2_name, "SNPs"),
  ylab = paste(trait_1_name, "SNPs")
)
```

## Step 5: Bivariate blocks

Each heatmap cell block pairs one Louvain module of trait-1 SNPs with one module of trait-2 SNPs. Mean block similarity classifies the relationship:

- **Concordant** — shared same-direction pleiotropy
- **Discordant** — opposing pleiotropy (trade-off)
- **Trait-specific** — weak cross-trait alignment

Block counts and “strongest” examples use only blocks where **both** sides have `> min_module_size` SNPs. Within each type, blocks are ranked by
`|mean_similarity| * sqrt(n1 * n2)` so tiny extreme cells do not dominate.

```{r step5-blocks}
module_summary <- summarise_cross_trait_modules(
  S_star,
  clusters_rows = cross_modules$clusters_rows,
  clusters_cols = cross_modules$clusters_cols
)

large_blocks <- module_summary |>
  dplyr::filter(
    n_snps_trait_1 > min_module_size,
    n_snps_trait_2 > min_module_size
  ) |>
  dplyr::mutate(
    block_score = abs(mean_similarity) * sqrt(n_snps_trait_1 * n_snps_trait_2)
  )

knitr::kable(
  large_blocks |>
    dplyr::count(block_type, name = "n_blocks") |>
    dplyr::arrange(dplyr::desc(n_blocks)),
  caption = paste0(
    "Block counts by type (both sides > ", min_module_size, " SNPs)"
  ),
  align = "lr"
)

blocks_of_interest <- large_blocks |>
  dplyr::filter(block_type %in% c("concordant", "discordant", "trait_specific")) |>
  dplyr::group_by(block_type) |>
  dplyr::slice_max(block_score, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

top_feature_drivers <- function(x_oriented, snp_ids, feature_info, feature_types, n = 8) {
  snp_ids <- intersect(as.character(snp_ids), colnames(x_oriented))
  feature_ids <- feature_info |>
    dplyr::filter(feature_type %in% feature_types) |>
    dplyr::pull(feature_id)
  feature_ids <- intersect(feature_ids, rownames(x_oriented))
  if (length(snp_ids) == 0 || length(feature_ids) == 0) {
    return(data.frame())
  }
  feat_mat <- x_oriented[feature_ids, snp_ids, drop = FALSE]
  data.frame(
    feature_id = rownames(feat_mat),
    mean_abs_z = rowMeans(abs(feat_mat), na.rm = TRUE),
    stringsAsFactors = FALSE
  ) |>
    dplyr::filter(is.finite(mean_abs_z)) |>
    dplyr::left_join(feature_info, by = "feature_id") |>
    dplyr::arrange(dplyr::desc(mean_abs_z)) |>
    dplyr::slice_head(n = n) |>
    dplyr::select(feature_id, feature_name, feature_type, source, mean_abs_z)
}

if (nrow(blocks_of_interest) == 0) {
  block_details <- list()
  cat(
    "No blocks with both sides > ", min_module_size,
    " SNPs for concordant / discordant / trait_specific.\n",
    sep = ""
  )
} else {
  block_details <- lapply(seq_len(nrow(blocks_of_interest)), function(i) {
    block <- blocks_of_interest[i, , drop = FALSE]
    row_snps <- names(cross_modules$clusters_rows)[
      cross_modules$clusters_rows == block$row_cluster
    ]
    col_snps <- names(cross_modules$clusters_cols)[
      cross_modules$clusters_cols == block$col_cluster
    ]
    list(
      block_type = block$block_type,
      mean_similarity = block$mean_similarity,
      block_score = block$block_score,
      n_snps_trait_1 = length(row_snps),
      n_snps_trait_2 = length(col_snps),
      top_pathways_trait_1 = top_feature_drivers(
        X1_star, row_snps, bivariate$feature_info, c("pathway", "gene")
      ),
      top_traits_trait_1 = top_feature_drivers(
        X1_star, row_snps, bivariate$feature_info, "trait"
      ),
      top_pathways_trait_2 = top_feature_drivers(
        X2_star, col_snps, bivariate$feature_info, c("pathway", "gene")
      ),
      top_traits_trait_2 = top_feature_drivers(
        X2_star, col_snps, bivariate$feature_info, "trait"
      )
    )
  })
  names(block_details) <- blocks_of_interest$block_type

  knitr::kable(
    dplyr::bind_rows(lapply(block_details, function(b) {
      data.frame(
        block_type = b$block_type,
        mean_similarity = round(b$mean_similarity, 2),
        block_score = round(b$block_score, 2),
        n_snps_trait_1 = b$n_snps_trait_1,
        n_snps_trait_2 = b$n_snps_trait_2,
        stringsAsFactors = FALSE
      )
    })),
    caption = "Top-scoring large block of each type",
    align = "r"
  )
}
```

```{r step5-block-details, echo=FALSE, results='asis'}
for (block_type in names(block_details)) {
  b <- block_details[[block_type]]
  cat("\n\n### ", tools::toTitleCase(gsub("_", " ", block_type)), " block\n\n", sep = "")
  cat(
    "Mean similarity: ", round(b$mean_similarity, 2),
    "; score: ", round(b$block_score, 2),
    "; ", trait_1_name, " SNPs: ", b$n_snps_trait_1,
    "; ", trait_2_name, " SNPs: ", b$n_snps_trait_2, "\n\n",
    sep = ""
  )
  if (nrow(b$top_pathways_trait_1) > 0) {
    cat("**Top pathway/gene drivers (", trait_1_name, "):**\n\n", sep = "")
    knitr::kable(b$top_pathways_trait_1, row.names = FALSE)
  } else {
    cat("No finite pathway/gene drivers for ", trait_1_name, ".\n\n", sep = "")
  }
  if (nrow(b$top_traits_trait_1) > 0) {
    cat("\n\n**Top trait drivers (", trait_1_name, "):**\n\n", sep = "")
    knitr::kable(b$top_traits_trait_1, row.names = FALSE)
  }
  if (nrow(b$top_pathways_trait_2) > 0) {
    cat("\n\n**Top pathway/gene drivers (", trait_2_name, "):**\n\n", sep = "")
    knitr::kable(b$top_pathways_trait_2, row.names = FALSE)
  } else {
    cat("\n\nNo finite pathway/gene drivers for ", trait_2_name, ".\n\n", sep = "")
  }
  if (nrow(b$top_traits_trait_2) > 0) {
    cat("\n\n**Top trait drivers (", trait_2_name, "):**\n\n", sep = "")
    knitr::kable(b$top_traits_trait_2, row.names = FALSE)
  }
}
```

## Verification 3: pathway enrichment per Louvain module

For modules with more than `min_group_size` SNPs on each trait side, re-run KEGG + Reactome enrichment and compare to that trait's baseline: do the same pathways come back, and do they split cleanly across modules?

```{r verify3-enrichment}
trait_1_enrichment <- enrich_snp_group_pathways(
  groups = cross_modules$clusters_rows,
  coloc_groups = cg,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)
trait_2_enrichment <- enrich_snp_group_pathways(
  groups = cross_modules$clusters_cols,
  coloc_groups = cg,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)
cmp_1 <- compare_group_pathways(baseline_1, trait_1_enrichment)
cmp_2 <- compare_group_pathways(baseline_2, trait_2_enrichment)

knitr::kable(
  trait_1_enrichment$summary,
  caption = paste0(trait_1_name, ": module pathway enrichment"),
  na = "",
  align = c("l", "r", "r", "r", "l")
)
knitr::kable(
  cmp_1$pathway_status,
  caption = paste0(trait_1_name, ": baseline vs module pathway status"),
  na = "",
  align = c("l", "l", "r", "r", "l", "l")
)

knitr::kable(
  trait_2_enrichment$summary,
  caption = paste0(trait_2_name, ": module pathway enrichment"),
  na = "",
  align = c("l", "r", "r", "r", "l")
)
knitr::kable(
  cmp_2$pathway_status,
  caption = paste0(trait_2_name, ": baseline vs module pathway status"),
  na = "",
  align = c("l", "l", "r", "r", "l", "l")
)
```

```{r verify3-details, echo=FALSE, results='asis'}
print_group_enrichment <- function(result, trait_label) {
  if (length(result$by_group) == 0) {
    cat("\n\n### ", trait_label, "\n\n", sep = "")
    cat("No modules with more than ", min_group_size, " SNPs.\n\n", sep = "")
    return(invisible(NULL))
  }
  cat("\n\n### ", trait_label, "\n\n", sep = "")
  for (x in result$by_group) {
    cat("\n\n#### Module ", x$group, "\n\n", sep = "")
    cat("SNPs: ", x$n_snps, "; genes: ", nrow(x$genes), "\n\n", sep = "")
    if (nrow(x$genes) > 0) {
      print(knitr::kable(x$genes, caption = "Genes at module SNPs", row.names = FALSE))
    }
    if (nrow(x$pathways) > 0) {
      print(knitr::kable(
        x$pathways |>
          dplyr::mutate(p_value = signif(p_value, 3), fdr = signif(fdr, 3)),
        caption = "Enriched pathways (KEGG + Reactome, FDR <= 0.05)",
        row.names = FALSE
      ))
    }
  }
}

print_group_enrichment(trait_1_enrichment, trait_1_name)
print_group_enrichment(trait_2_enrichment, trait_2_name)
```

Distinct Louvain modules should ideally show **different enriched pathways**, supporting separation of biological mechanisms between SNP groups at linked loci.
