---
title: "Univariate SNP biological clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Univariate SNP biological clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take a **single trait** and decompose its genetic architecture by grouping SNPs
according to their phenome-wide pleiotropic profiles.

The central question is: **within one trait, what distinct biological programs
are represented by its associated SNPs?**

The output is a set of SNP programs (latent factors), each annotated with
background traits, QTL studies, tissues, genes, and pathways. The runnable
implementation is
[Investigation of univariate analysis](investigation-univariate-analysis.html).

Background traits stay at **study level**. Molecular QTL studies are not
collapsed by gene, tissue, cis/trans, or QTL type.

### Core clustering preferences

The grouping method should satisfy:

1. **Handles high sparsity** — Must natively process large, sparse matrices
   (thousands of singleton feature rows) without treating structural zeros as
   low-variance signal. EBMF does this by leaving missing cells out of the fit
   (`flashier` skips `NA`); we additionally limit sparsity with a hard threshold
   on the number of non-missing SNPs per trait (`min_snp_signals`).
2. **Signed and sign-driven clustering** — Must preserve the meaning of
   continuous positive/negative z-scores so variants group by **directional
   alignment**, not magnitude alone. We orient every SNP column by
   \(\mathrm{sign}(z_{\mathrm{target}})\) before factorization.
3. **Automated \(K\) discovery** — The engine should learn the number of
   clusters or latent factors from the data. EBMF adds factors greedily while
   the variational lower bound (ELBO) improves, then backfits; filtering the
   resulting factors by an lFSR/magnitude membership gate and by the
   validation checks below controls the final program count.
4. **Overlapping membership** — A SNP can belong to more than one biological
   program. EBMF produces soft factor loadings, so membership is per-SNP-per-
   program rather than a single hard label.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Pleiotropy matrix\\n(study-level rows x SNPs)\"]
  S2 [label=\"2. Orient by sign(z_target)\\noptional asinh compression\"]
  S3 [label=\"3. EBMF factorization Y = L F^T + E\\nlFSR + magnitude program gates\"]
  S4 [label=\"4. Validate programs\\ncoherence on SNP-SNP S, stability, factor strength\"]
  S5 [label=\"5. Interpret valid programs\\ntraits, tissues, pathways\"]
  S1 -> S2 -> S3 -> S4 -> S5
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

### 1. Construct the pleiotropy matrix

Given:

- `nsnp` = independent SNPs associated with the target trait (coloc groups)
- `ntrait` = background studies in GPMap (phenotypic and molecular)

Construct a **traits x SNPs** matrix of signed z-scores (\(z = \beta / \mathrm{SE}\)).
Conceptually this is a SNP-by-trait profile matrix; the code stores traits as
rows and SNPs as columns.

```text
              SNP1   SNP2   SNP3   SNP4   ...
Trait A        z      z      z      z
Trait B        z      z      z      z
Trait C        z      z      z      z
...
```

Which rows and columns back this matrix is controlled by three choices:

- **`include_trans`** — whether trans-QTL association rows
  (`cis_trans == "trans"`) are removed before the matrix is built. Excluding
  them (default) keeps the analysis cis-focused; including them can surface
  distributed regulatory programs at the cost of density and noise.
- **`trait_subset`** — `"all"` keeps every available background trait;
  `"phenotypic"` keeps only phenotype traits (rows whose `feature_type` is
  `"phenotypic"` or missing) plus the target trait, dropping molecular /
  gene-annotated rows. Molecular traits are much sparser, so clustering on
  phenotype traits only can recover structure that sparse molecular rows
  dilute.
- **`min_snp_signals`** — sparse rows observed in fewer than this many SNPs
  are dropped; ubiquitous rows (non-missing on more than `max_snp_fraction` of
  SNPs) are dropped too. The target trait row is always kept.
- **`compress_method`** — optional `asinh` compression of z-scores, which
  limits the leverage of extreme z-scores without a hard clip. `"none"` preserves
  effect scale but lets outliers dominate direction and norm.

### 2. Orient SNP profiles relative to the target trait

Remove arbitrary allele coding so every column represents increasing target
liability. For SNP \(j\):

\[
X^*_{tj} = \mathrm{sign}(z_j)\, X_{tj}
\]

Optional \(\mathrm{asinh}(z)\) compression (`compress_method = "asinh"`, with
`compress_scale`) then limits the leverage of extreme z-scores without a hard
clip. `"none"` preserves effect scale but lets outliers dominate direction and
norm.

### 3. Fit the EBMF latent-program model

Treat the oriented (and optionally compressed) traits x SNPs matrix as noisy
observations of latent programs:

\[
Y = LF^{\top} + E
\]

- \(Y\) = features x SNPs matrix (traits and molecular QTL studies as rows)
- \(L\) = feature loadings (which traits drive each program)
- \(F\) = SNP factors (which SNPs each program loads on)
- \(E\) = residual error

**Two loading matrices, two kinds of output each.** `flashier` returns
posterior summaries for both \(L\) and \(F\), and for each matrix it
separately reports a point estimate and an uncertainty measure. These are
easy to conflate — both are called "loadings" — but they answer different
questions about different axes of the model:

| | SNP-loading matrix `F` | trait-loading matrix `L` |
|---|---|---|
| point estimate | `F_pm` — posterior mean loading of each SNP on the program | `L_pm` — posterior mean loading of each background trait on the program |
| uncertainty | `F_lfsr` — local false sign rate: how confident EBMF is in that SNP's sign/nonzero-ness | `L_lfsr` — same, for the trait loadings |

`F_pm`/`F_lfsr` are what the program-extraction gate just below uses
(`|loading| > ebmf_magnitude_threshold` and `lfsr < ebmf_lfsr_threshold`),
and what the SNP-similarity matrix `S` in Step 4 is built from. `L_pm`/`L_lfsr`
describe which *background traits* drive a program and are used downstream
for biological interpretation (Step 5's "top traits by magnitude"), not for
deciding SNP membership.

Magnitude and lfsr are not the same axis, even within one matrix: magnitude
(\(|F_{pm}|\)) is how big the point estimate is; lfsr is how much to trust it
— specifically, the posterior probability that the true sign is the opposite
of what was estimated. A locus can have a large magnitude and still a high
lfsr if it is weakly powered: the point estimate looks big, but the posterior
around it is wide enough that it could easily be near zero or flipped. That
is why program extraction gates on both, not magnitude alone.

Implemented via `flashier`: factors are added greedily while the variational
lower bound improves (`greedy_Kmax`), then cyclically backfit. The number of
programs is learned, not fixed. The sparse prior on loadings and factors is
`ebmf_prior`:

- `"point_normal"` (default) — a sparse, spike-and-slab style prior; good
  general default.
- `"point_laplace"` — heavier tails; often better when true effects are large
  relative to noise.

**Extracting programs.** A SNP belongs to a program when its posterior factor
loading is both *significant* and *non-negligible*:

- `lfsr` < `ebmf_lfsr_threshold` (local false sign rate; defaults to `0.05`)
- `|loading|` > `ebmf_magnitude_threshold` (absolute posterior mean of `F_pm`;
  defaults to `0.25`)

Because EBMF is soft, one SNP can pass the gate for several programs
(overlapping membership). Before extraction, at most one *global* mega-factor
— a factor that soft-assigns nearly every SNP or explains a large share of PVE
— is dropped (`ebmf_drop_global = TRUE` via `identify_ebmf_global_factors()`);
this guards against the "one giant cluster" failure mode. Nothing assumes
factor 1 is global.

Why EBMF:

- SNPs can belong to multiple programs; biology is overlapping, not mutually
  exclusive.
- \(K\) is inferred by the fit (ELBO) rather than by a graph-resolution or
  edge-threshold parameter.
- The input is the dense oriented z-score matrix itself, so no information is
  lost to a pairwise similarity threshold.
- Missing data are handled probabilistically instead of being zero-filled into
  similarity edges.

The cost is that factors are not guaranteed to be disjoint SNP modules and can
capture broad technical/global effects — which is exactly why the validation
layer in Step 4 exists.

### 4. Validating programs

`run_univariate_clustering()` returns the fit and soft memberships;
`summarise_ebmf_programs()` folds everything into a per-program table with
pass/fail flags and an overall `status`. The checks split into those computed
**on the SNP x SNP similarity matrix** `S` (internal coherence) and those that
use **other means** (the EBMF fit itself and refits under trait subsampling).
A program is **`valid`** only if it passes all **filter gates**; some metrics
are reported descriptively and never gate `status`.

#### The SNP x SNP similarity matrix

EBMF never sees pairwise SNP similarity — it factorises the traits x SNPs
matrix directly — but the same oriented (and optionally compressed) SNP
profiles are used to build the SNP x SNP cosine similarity matrix `S`, which
backs the coherence checks below. For SNPs \(j\) and \(k\):

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\|\,\|x^*_k\|}
\]

- \(+1\) = concordant (parallel) profiles
- \(0\) = unrelated
- \(-1\) = discordant (anti-parallel) profiles

#### Checks on the similarity matrix (internal coherence)

Computed on `S`, restricted to the program's gated SNPs:

- **mean_internal_similarity** — mean off-diagonal `S` among member SNPs
  (diagonal excluded). High values mean oriented profiles are roughly
  parallel. Must be `>= min_mean_internal` (default `0.3`), a weak
  "same direction, not noise" bar.
- **connectedness** — the fraction of member-SNP pairs with
  \(|S| \ge\) `similarity_threshold` (the same `edge_threshold` used in
  program validation). Mean internal similarity can look decent if a few
  pairs are very similar and the rest are near zero; connectedness
  `>= min_connectedness` (default `0.5`) requires at least half of all pairs
  to be strong edges, so the program is a real connected blob on `S`, not
  one tight pair inside a diffuse background.
- **Separation** (internal minus external similarity) is also reported —
  diagnostic only, not a gate: a program can be internally coherent yet sit
  inside a similar background.

#### Checks by other means

These come from the EBMF fit itself, not from `S`:

- **Size** — at least `min_module_size` SNPs survive the lFSR/magnitude gate
  (`n_snps_filtered`, counted from the EBMF membership). This ensures the
  coherence metrics are computed on enough SNPs to be meaningful.
- **Stability** — the program's top `top_n` loading SNPs must be recovered
  with replication `>= stability_threshold` (default `0.7`) when `1 -
  frac_traits` of the trait rows are held out and flashier is refit
  (`n_rep` replicates). Because this refits EBMF, it is run only for programs
  that already pass the size and internal-coherence gates. Descriptive, but
  gated: a program that does not survive trait subsampling is rejected.

**Descriptive scores** (reported for context, but **not** used to gate
`status`):

- **factor_strength** — the proportion of the observed matrix signal
  attributable to the factor, computed from its fitted reconstruction on the
  gated (assigned) SNPs:
  \(\sum_i |L_{ik}| \sum_j |F_{jk}|\) over the total observed signal
  \(\sum |X|\). `factor_strength_per_snp = factor_strength / sqrt(n_snps_filtered)`
  normalises it by the square root of program size. These prioritise programs
  that carry real matrix signal rather than marginal memberships.
- **posterior_evidence** — aggregate (`median` or `mean`) of `-log10(lfsr)`
  across the program's assigned SNPs (capped so lFSR underflow does not blow
  up). Distinguishes programs dominated by borderline `lfsr ~ 0.05`
  memberships from programs whose memberships carry much stronger posterior
  sign evidence.
- **max_abs_factor_corr / redundant** — the maximum absolute correlation of the
  program's SNP loading vector with any other program; programs exceeding
  `redundancy_threshold` are flagged `redundant` (near-duplicate factors).

`status` is `"valid"` when all gates pass, otherwise the failed gate(s) are
listed (`size; internal; stability`). Downstream interpretation uses only the
gated memberships of `valid` programs.

**Permutation calibration (reporting layer).** `calibrate_ebmf_programs()`
additionally shuffles each trait row independently (`n_null` replicates,
preserving each trait's sparsity and effect distribution), refits the pipeline,
and pools the null loading distribution. Each SNP x program loading gets an
empirical p-value and BH q-value against this null; cells with `q <=
alpha_membership` are labelled `core`, and the top remaining loadings per
program form an explicitly uncertified `candidate` tier. This is a membership
labelling layer, not a program-level validity gate.

### 5. Interpret valid programs

Programs are interpreted two ways in parallel — a **hard-cutoff** approach
(gate SNPs into a program, then test the gated set for enrichment) and a
**continuous** approach (weight every SNP by its loading, no gate). Both are
run on the same `valid` programs so their results can be compared side by
side; the continuous approach does not replace the hard-cutoff one.

**Hard-cutoff enrichment.** The univariate vignette interprets **valid** EBMF
programs as:

- **Top traits by magnitude** (`abs_mean_z` within the program).
- **Trait-category composition** (`enrich_snp_group_trait_categories()`):
  observed vs expected share against the rest of the trait SNP baseline
  (coloc rows for SNPs attributed to the analysed trait, not the whole GPMap
  universe). Top elevated categories are shown with obs/exp/FE; FDR is noted
  when significant.
- **Tissue composition** (`enrich_snp_group_tissues()`, `background = "trait"`)
  on the same leave-one-out trait baseline.
- **Pathway enrichment** on QTL genes (KEGG and Reactome; `enrich_snp_group_pathways()`),
  treated as secondary because the full-trait gene set is often empty for
  broad polygenic traits.
- **SNP pleiotropy typology** (`summarise_snp_group_pleiotropy()`): programs
  are placed on a *breadth* axis (mean per-SNP distinct trait categories /
  distinct genes) against *tightness* (separation). Descriptive only — it
  classifies programs as specific programs (tight + narrow), regulatory hubs
  (tight + broad), or generic background (loose + broad), without acting as a
  filter or reliability criterion.

#### Multiple testing correction for the hard-cutoff enrichments

Each valid program is tested against **every** trait category and **every**
tissue for over-representation relative to the trait baseline (leave-one-out
hypergeometric test). A
single program therefore produces dozens of hypothesis tests — one per category
or tissue — and across all programs in a run the pipeline performs hundreds. If
we declared everything with raw p < 0.05 significant, we would expect ~5% of
purely null categories/tissues to pass by chance; a program testing ~50
categories would accumulate ~2-3 false hits before any real biology.

We therefore apply **Benjamini–Hochberg (BH) FDR
correction** *within each program*, across the categories (or tissues) tested
for that program. BH sorts the raw p-values, then finds the largest prefix that
can be declared significant while keeping the expected fraction of false
positives among the significant calls at or below the chosen level (0.05). The
reported `FDR` for a category/tissue is the smallest FDR level at which that
row would still be called significant (a "q-value"): a category with
`FDR = 0.03` means that if we declare it significant, we expect ≤3% of the
significant calls for that program to be false positives.

Note: **Correction is per-program, not global.** Each program's FDR is computed
over the categories/tissues tested for that program alone. Running many
programs still multiplies the opportunities for false positives across the whole
analysis, which is why the vignette insists on replication across runs before
treating a hit as real.

#### Continuous loading-based enrichment (parallel alternative)

`enrich_program_loadings_trait_categories()`, `enrich_program_loadings_tissues()`,
and `enrich_program_loadings_pathways()` test the same three annotation types
without the lFSR/magnitude gate: every SNP with a finite EBMF loading
contributes to every program's test, weighted by its squared loading, rather
than being included or excluded by a threshold. For membership indicator
\(x \in \{0, 1\}\) (does this SNP link to this category/tissue/pathway term)
and squared loading \(y = F_{jk}^2\), an ordinary least squares regression

\[
y = \beta_0 + \beta_1 x + \varepsilon
\]

(`lm(loading^2 ~ membership)`) is fit per program per candidate term. The
fitted \(\beta_1\) — the
difference in mean squared loading between member and non-member SNPs — is
the reported `enrichment`; its `p` comes from the usual OLS \(t\)-test.

**Effect size.** `enrichment` is a raw mean-difference in squared-loading
units, so its scale depends on that program's own loadings and is not
comparable between programs. `r_squared` (the fraction of a program's
squared-loading variance the term explains) is bounded in \([0, 1]\) and is
comparable across programs, which is what makes a single fixed threshold
meaningful — the univariate vignette's comparison step filters on
`r_squared >= 0.05`, between Cohen's (1988) conventional "small" (0.01) and
"medium" (0.09) benchmarks, on top of `FDR <= 0.05`, and orders surviving hits
by `enrichment` rather than by significance.
* FDR alone is not a sufficient
filter here: with hundreds of SNPs feeding every regression, a small,
practically negligible difference in mean squared loading can still reach
significance, so an effect-size floor matters more than it does for the
hard-cutoff hypergeometric test's fold-enrichment ratio. One caveat: because
the number of SNPs tested is roughly constant across programs within one
trait, `r_squared` here tracks \(p\) closely rather than supplying a fully
independent axis — it is a bounded, human-interpretable rescaling of the same
evidence, not a second, unrelated filter.

**Multiple testing correction is pooled across programs, not per-program.**
Unlike the hypergeometric functions above, each of the three continuous
functions runs one regression per program per category/tissue/pathway term,
then Benjamini-Hochberg-corrects across *every* program's tests for that
enrichment type in a single family — so `fdr` reflects the true number of
regressions run for the trait, not one program's slice of them. For pathways,
KEGG and Reactome are pooled into that same family rather than corrected
separately, so a KEGG hit competes against the combined KEGG+Reactome,
all-programs test count; `source` remains a label column, not a correction
boundary. Trait-category, tissue, and pathway enrichment each stay their own
separate family — none of the three is pooled with either of the others.

**SNPs are treated as independent observations** in every regression above.

The two approaches are run side by side (not merged into one report) so their
results can be compared program by program; see the "continuous loading-based
enrichment (comparison)" step of
[Investigation of univariate analysis](investigation-univariate-analysis.html).

## Variables to consider

Exposed in the univariate vignette today (EBMF path). Parameters fall into two
groups: the **EBMF clustering parameters** that control matrix construction and
factor extraction, and the **program validation parameters** that decide whether
a program is called `valid`.

### EBMF clustering parameters

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `include_trans` | `TRUE`, `FALSE` | Whether trans-QTL rows are retained | Including trans effects can reveal distributed regulatory programs and improve tissue/pathway interpretation, while increasing density and possible noise. Excluding them gives a more local cis-focused analysis but may miss system-level mechanisms. |
| `trait_subset` | `all`, `phenotypic` | Which trait rows back the matrix | `phenotypic` drops sparse molecular / gene-annotated rows (keeping the target trait), recovering structure that sparse molecular rows dilute but losing QTL-level signal. `all` keeps every available background trait at the cost of extra sparsity. |
| `min_snp_signals` | `1`, `2`, `3`, `5` | Minimum SNPs with a value for a background trait | Raising it removes singleton or very sparse rows, reducing unstable factors and computation. It can also remove rare, module-specific biology. Lowering it preserves those signals but makes factorization more sensitive to sparse observations. |
| `compress_method` | `asinh`, `none` | Transformation applied to z-scores | `asinh` is nearly linear around zero but compresses extreme values, so factors reflect patterns across several traits rather than one very large z-score. `none` preserves effect scale but allows outliers to dominate direction and norm. |
| `ebmf_lfsr_threshold` | `0.05`, `0.02`, `0.1` | Local false sign rate cutoff for SNP membership | Lowering it keeps only SNPs whose program loading has very strong posterior sign evidence, shrinking and sharpening programs. Raising it adds borderline `lfsr ~ threshold` memberships. |
| `ebmf_magnitude_threshold` | `0.25`, `0.5`, `0.75`, `NA` | Minimum absolute posterior loading for SNP membership | Raising it keeps only SNPs with substantial program loadings. `NA` gates membership on lFSR alone (useful when loadings are on raw-beta scale). |
| `ebmf_prior` | `point_normal`, `point_laplace` | Prior family for EBMF loadings/factors | `point_normal` is the general default; `point_laplace` has heavier tails and is often better when true effects are large relative to noise. |

### Program validation parameters

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `similarity_threshold` | `0`, `0.2`, `0.5` | Absolute \(\|S\|\) edge cutoff for the connectedness gate | Not an EBMF input, but used by the validation layer: `connectedness` counts SNP pairs with \(\|S\| \ge\) this threshold. Raising it makes the connectedness gate stricter and the coherence metrics more conservative. |
| `min_module_size` | `3`, `5` | Minimum gated SNPs for the size gate | Ensures the coherence and stability metrics are computed on enough SNPs to be meaningful; also the effective minimum program size. |
| `min_mean_internal` | `0.3` | Minimum mean internal similarity | The "same direction, not noise" bar on `S` among gated member SNPs. |
| `min_connectedness` | `0.5` | Minimum pair-connectedness | Requires at least this fraction of member-SNP pairs to exceed `similarity_threshold`, so a program is a connected blob rather than one tight pair in a diffuse background. |
| `stability_threshold` | `0.5`, `0.7` | Minimum replication for stability | Fraction of a program's top-loading SNPs that must reappear when traits are held out and EBMF is refit. |
| `redundancy_threshold` | `0.8`, `0.9` | Threshold for redundancy program filtering | If a program shares the same set of SNPs as another program, it is considered redundant and filtered out. |

Reliability/validation cutoffs used after clustering therefore are:
`min_module_size`, `min_mean_internal = 0.3`, `min_connectedness = 0.5`,
`stability_threshold`, plus the membership gates `ebmf_lfsr_threshold` and
`ebmf_magnitude_threshold`. `factor_strength`, `posterior_evidence`, and
`max_abs_factor_corr`/`redundant` are reported descriptively but do not gate
`status`.

### Continuous enrichment parameters

Inputs to `enrich_program_loadings_trait_categories()`,
`enrich_program_loadings_tissues()`, and `enrich_program_loadings_pathways()`.
`min_category_size` and `min_loading_magnitude` are function arguments;
`r_squared` filtering is applied by the comparison step in the vignette, not
built into the functions themselves.

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `min_category_size` | `5` | Minimum linked SNPs for a category/tissue/pathway term to be tested | Below this, a term is skipped entirely (not tested, not reported) rather than tested on too few SNPs to fit a stable regression. |
| `min_loading_magnitude` | `0`, `0.02` | Soft membership gate on `x = 1` | `0` disables gating (every linked SNP counts as a member regardless of its loading on this program). Raising it relabels linked-but-negligible-loading SNPs to `x = 0` for that program's test only, sharpening the member/non-member contrast without a hard lFSR/magnitude cutoff. SNPs are never dropped from the analysis, only relabelled. |
| `r_squared` floor (vignette-level, not a function argument) | `0.01` (Cohen small), `0.05`, `0.09` (Cohen medium) | Minimum variance-explained for a hit to be reported, on top of `FDR <= 0.05` | Because `enrichment`'s scale is not comparable across programs, a fixed cutoff must be applied to `r_squared` instead. `0.05` (used in the univariate vignette's comparison step) sits between Cohen's (1988) "small" and "medium" benchmarks. |

## Simulation Strategy

Simulations will help with 2 levels of parameter choice:
1. Choosing what goes into the matrix (data, z-score attenuation) and LFSR/magnitude membership gates.
2. Choosing the validation cutoffs that define a program as `valid` or not.

`simulate_trait()` generates a trait
object of the same shape as `trait(id, include_associations = TRUE)` and
plants structure over SNPs. The simulations come in two families — **null**
(`K = 0`, no modules) and **module** (`K > 0`, planted modules) — so the same
EBMF + validation pipeline can be tested for both *hallucination* (does it
invent structure from randomness?) and *recovery* (does it find the structure
we planted, and how well?).

### Null simulations: testing hallucination

The null family keeps `K = 0`: no modules and no driver traits. The matrix is
the dense target row plus unstructured background traits. Because there is no
planted structure, any program that passes all four validation checks is a
**false discovery**, so the null runs measure how often the EBMF + validation
battery invents structure from randomness.

- The meaningful controls here are the **background effect** parameters
  (below) and `noise_sd`. The driver/module parameters (`p_structural_zero`,
  `p_spurious`, `effect_tail`, `trait_overlap`, `snp_driver_groups`, ...) are
  inert when `K = 0` because they only act on driver rows.
- Key outputs: `n_programs` (how many factors EBMF proposes),
  `n_valid_programs` (how many pass every check — the false-positive rate),
  and `background_absorbed` (the fraction of background SNPs pulled into valid
  programs — the SNP-level false-discovery rate).

The null simulations, and how the pass rates change as the background is made
denser, sparser, or stronger, are documented in
[Null simulation study of EBMF programs](investigation-univariate-null-simulations.html).

### Module simulations: testing recovery

The module family plants `K` modules over SNPs and runs the identical battery
on versions that differ along several axes of control. Each axis stresses a
different way real data can depart from a clean block structure; the versions
are generated by iterating the generative parameter arrays in the simulations
vignette
([Simulation study of EBMF programs](investigation-univariate-simulations.html)).

**Planted modules.** `K` modules occupy SNP sets (disjoint by default). Each
module has `n_traits_per_module` driver traits with `effect_size`-scale effects
on its module's SNPs, under a `sign_pattern` (`coherent`, `flipped`, or
`random`). A driver trait is a row whose associations are **localised to one
module's SNP block** — this is the "signal" the EBMF programs are supposed to
recover.

**How we control the simulations**. The parameters shape the input, and they differ in 
where the structure (or mess) they add lives:

1. **Background effect parameters** (`p_active_background`,
`background_sparsity_sd`, `background_effect_scale`) — the unstructured
noise floor. Every background SNP association is an independent draw:
`p_active_background` sets how often a background trait is observed at a SNP
(density), `background_sparsity_sd` makes per-trait density heavy-tailed
(most traits observed at a handful of SNPs, with a tail of denser traits),
and `background_effect_scale` sets the magnitude of the hits relative to
`effect_size`. Unlike modules, the background has no shared structure;
it makes recovery harder by reducing the contrast between coherent module
signal and unrelated associations.

2. **SNP driver groups and driver sharing** (`snp_driver_groups`,
`driver_sharing`) — the internal and cross-module SNP architecture of the
planted modules. These parameters introduce additional structured
complexity, rather than unstructured noise, but they act in different ways.

    snp_driver_groups partitions each module's SNPs into groups and assigns
    each driver trait to only one corresponding SNP group. Thus, increasing
    snp_driver_groups progressively changes a planted module from a single
    coherent SNP-by-trait pattern into several distinct subpatterns. This is
    therefore a relatively strong form of within-module heterogeneity: at high
    values, the simulated data may contain multiple statistically distinguishable
    latent patterns even though all of their SNPs are labelled as belonging to
    the same ground-truth module. The parameter consequently tests how many
    distinct subpatterns can be embedded within one nominal biological programme
    before EBMF and subsequent clustering resolve them as separate factors or
    programs.

    driver_sharing acts differently. It does not control similarity between
    SNP groups within the same module. Instead, the first
    floor(n_traits_per_module * driver_sharing) driver traits of module m
    are also active across the SNP set of the previous module. It therefore
    creates structured overlap in SNP usage between adjacent modules: a driver
    trait nominally associated with one module can also generate signal among
    SNPs belonging to another module. This differs from both trait_overlap,
    which makes modules share driver traits themselves, and the background
    parameters, which generate associations without coherent shared structure.

3. **Module trait overlap** (`trait_overlap`, `trait_effect_corr`) — the
overlap of trait signatures between modules. `trait_overlap` determines
the proportion of each module's driver traits that are shared with other
modules. As trait overlap increases, different modules are defined by
increasingly similar sets of traits, reducing the distinction between their
pleiotropic signatures. `trait_effect_corr` determines how similar the
corresponding effects are when traits are shared: shared drivers can have
correlated but non-identical effects across modules.

    This manipulation is distinct from driver_sharing. trait_overlap changes
    which traits define each module, whereas driver_sharing changes which
    SNPs are affected by particular driver traits. At high trait overlap,
    modules may become difficult to separate because their trait signatures are
    genuinely similar, even when their SNP sets remain distinct. The parameter
    therefore tests the resolution limit of the method when biologically
    distinct SNP groups have increasingly overlapping pleiotropic profiles.

4. **Spurious hits** (`p_spurious`) — contamination of the driver signal.
`p_spurious` adds background-scale off-module associations to driver traits,
introducing associations outside their intended structured SNP patterns.
Unlike `snp_driver_groups` or `driver_sharing`, these hits do not represent
additional coherent latent structure. They instead test robustness to
isolated or weak associations that partially obscure the intended module
boundaries.

The module simulations are scored against ground truth by
`evaluate_univariate_simulation()`:

- **module_recall** — for each planted module, the largest fraction of its SNPs
  contained in any single predicted program (fair to overlapping/merged
  programs).
- **coverage** — fraction of all SNPs placed into valid programs.
- **background_absorbed** — fraction of unstructured background SNPs absorbed
  into valid programs (higher = more hallucinated structure).
- **ARI** (`ari_all`, `ari_structured`) — adjusted Rand index against the
  planted partition, using each SNP's strongest program as a hard projection.

The module and null runs are read against each other: the module runs show how
much planted structure is recovered, and the null runs supply
`background_absorbed` as a false-discovery rate — how much of what is recovered
under no planted structure is invention.

## Structure of Results

The aim is not produce a definitive biological truth for every SNP.  Indead, the method produces a structured, evidence-based decomposition of a trait's genetic architecture, with results that can be assessed at several levels:

For practical use, the framework should provide results at the overall, program, and SNP level.

| Level       | Primary question                                                                         | Main output                                                             |
| ----------- | ---------------------------------------------------------------------------------------- | ----------------------------------------------------------------------- |
| **Trait**   | How much of this trait's genetic architecture can be decomposed into distinct structure? | Number of programs, SNP coverage, overlap, overall signal captured      |
| **Program** | Is this structure reliable, and what biology does it represent?                          | Reliability metrics plus trait, tissue, gene and pathway interpretation |
| **SNP**     | Which program or programs does this variant contribute to?                               | Soft memberships, posterior evidence, rank and molecular annotations    |

This standard report would be returned by gpmapr and as a framework for the Results section of the paper. It should keep the distinction clear between statistical evidence, which establishes whether a program is reliable, and biological annotation, which provides its interpretation.

#### Overall trait-level metrics
- Number of SNPs analysed
- Number of validated programs discovered (maybe also report candidate programs that fail validation)
- Proportion of SNPs assigned to at least one validated program
- Proportion of SNPs unassigned
- Degree of overlapping membership, for example the proportion of assigned SNPs belonging to more than one program
- Overall matrix signal explained by validated programs, if this can be meaningfully derived from the EBMF fit
- Number/proportion of programs with some evidence of biological interpretation

#### Program Level metrics
- Number of member SNPs (maybe also report number of SNPs that were filtered out by the lFSR/magnitude gate)
- Membership overlap with other programs
- Top SNPs by posterior loading or membership evidence
- Various statistical reliability metrics, including mean internal similarity, connectedness, stability, factor strength, stability 
- Biological interpretation metrics, including top traits, trait categories, tissues, genes, pathways, and pleiotropy typology, from both the hard-cutoff (fold enrichment, FDR) and continuous loading-based (effect, r_squared, FDR) enrichment routes

#### SNP Level metrics
- Program membership(s)
- Posterior loading in each program
- Membership lFSR
- Magnitude of membership/loading
- Core versus candidate membership status
- Rank within each program, if useful
- Whether the SNP belongs to multiple programs
- Similarity to other SNPs within its assigned program
- Variant-level pleiotropic profile, for example breadth of associated trait categories
- Molecular annotations, including associated genes and tissues where available