---
title: "Investigation of SNP clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of SNP clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Overview

This note is an **overview** of how we cluster SNPs from GPMap colocalisation and pleiotropy data. The goal is to group variants that share **concordant**, **discordant**, or **trait-specific** pleiotropy structure — without manually guessing the number of clusters.

The detailed, runnable workflows live in the investigation vignettes linked below. Each vignette implements one slice of the method comparison outlined here.

## Outline

1. [Three core clustering preferences](#three-core-clustering-preferences)
2. [Statistical methods evaluated](#statistical-methods-evaluated)
3. [Dimensionality reduction: genes and tissues](#dimensionality-reduction-genes-and-tissues)
4. [Pathway verification](#pathway-verification)
5. [Workflow diagram](#workflow-diagram)
6. [Investigation vignettes](#investigation-vignettes)

## Summary so far

| Preference | EBMF | Signed Leiden / Louvain | hclust | Leading eigen |
|------------|------|-------------------------|--------|---------------|
| High sparsity | Strong | Moderate (graph sparsification helps) | Weak | Weak |
| Sign-aware | Yes | Yes | Only with custom metrics | Hacky |
| Automated *K* | Yes (ELBO) | Yes (modularity) | No | No |
| Soft clusters | Yes | No | No | No |
| Results | Improving: drop F1 + stricter membership | Ok: distinct but large clusters | NA | NA |

**Recommended path:** EBMF for discovery and soft program assignment; signed graph clustering (Louvain today, Leiden target) on trait (optionally gene x tissue) matrices for hard SNP modules and bivariate concordance / discordance blocks. Use the linked investigation vignettes for reproducible examples on GPMap data.

## Investigation vignettes

| Workflow | Vignette | What it demonstrates |
|----------|----------|----------------------|
| EBMF (matrix factorization) | [Investigation of EBMF analysis](investigation-ebmf-analysis.html) | Soft programs via `flashier`; automated *K*; sign-aware sparse priors |
| Univariate | [Investigation of univariate analysis](investigation-univariate-analysis.html) | Trait-only matrix; signed Louvain on `S` (BMI) |
| Bivariate | [Investigation of bivariate analysis](investigation-bivariate-analysis.html) | Trait matrices with optional gene x tissue QTL collapse; Louvain on `S*` (BMI + WHR) |

## Three core clustering preferences

Any method we adopt for SNP (or SNP-block) grouping should satisfy:

1. **Handles high sparsity** — Must natively process large, sparse matrices (e.g. thousands of singleton feature rows) without numerical gridlock or treating structural zeros as low-variance signal.

2. **Signed and sign-driven clustering** — Must preserve the biological meaning of continuous positive/negative values (z-scores / betas) so variants group by **directional alignment** (concordant vs. discordant effects), not magnitude alone.

3. **Automated *K* discovery** — The engine should learn the effective number of clusters or latent factors from the data (ELBO, modularity, or similar), rather than relying on arbitrary manual cuts.

## Statistical methods evaluated

### EBMF (`flashier`) — best fit (current champion)

Native sparsity via empirical Bayes priors; fully sign-aware; automatically chooses *K* through ELBO optimisation; supports **soft clustering** (pleiotropic SNPs can load on multiple programs).

**Tunable parameters**

- **`ebnm_fn`:** Switch from `point_normal` to `point_laplace` for sharper shrinkage and to break up dense macro-clusters.
- **`var_type`:** Set to `1` with `row.group.assignment` vectors for fast, robust background noise models instead of slow iterative loops.
- **`greedy_Kmax` and `tol`:** Raise discovery ceilings (e.g. 30–50) so the model searches beyond a global intercept factor.
- **Input architecture (`Y` and `S`):** Pass separate beta (`Y`) and SE (`S`) matrices for a weighted meta-analytic fit; impute missing cells explicitly (`Y = 0`, `S = 1e6`) to stabilise optimisation.

See [Investigation of EBMF analysis](investigation-ebmf-analysis.html).

### Signed Leiden — excellent graph alternative

Addresses structural flaws of signed Louvain (fewer disconnected or trapped communities). Fully sign-aware. **Hard clustering** only (one SNP, one group).

**Tunable parameters:** `louvain_gamma` (resolution; scales community size) and upstream `similarity_threshold` (drop weak signed edges before building the adjacency matrix).

*Not yet implemented in gpmapr vignettes; listed here as the preferred graph upgrade over Louvain.*

### Signed Louvain — graph method in current vignettes

Fast and sign-aware, but can fragment in high-sparsity regions (local modularity traps). Hard clustering.

**Tunable parameters:** Same as Leiden — `gamma` and `similarity_threshold` in `cluster_snp_profiles()` and `cluster_cross_trait_snps()`.

See [Investigation of univariate analysis](investigation-univariate-analysis.html) and [Investigation of bivariate analysis](investigation-bivariate-analysis.html).

### Hierarchical clustering (`hclust`) — poor fit

Traditional distance-based clustering. Does not handle extreme sparsity cleanly, cannot model overlapping soft clusters, and does not choose *K* automatically (requires arbitrary tree cuts). Standard distance metrics also collapse directional sign unless custom signed similarity matrices are used.

Used in baseline investigation vignettes for comparison and heatmap ordering only.

### Leading eigenvector — poor fit

Top-down spectral bisection; background modularity is structurally dense (*O*(*n*³) scaling). Requires manual handling of signed edges and forces hard clustering.

## Dimensionality reduction: genes and tissues

Molecular QTL studies are sparse at the study level. The bivariate workflow can
optionally collapse them with `collapse_gene_tissue = TRUE`, merging cis/trans and
QTL types within each gene x tissue while keeping tissue labels for annotation.

Pathway enrichment remains a **post-hoc verification** step on module genes
(`enrich_trait_pathways()` / `enrich_snp_group_pathways()`), not a matrix-row
collapse for Louvain.

## Pathway and alternative verification

Investigation vignettes use shared verification helpers:

1. **Baseline** — before clustering, summarise recoverable signal for the trait SNP set.
2. **Analysis** — run Louvain / EBMF as usual.
3. **Per-group** — for modules / programs with more than 5 SNPs, repeat the check and compare to baseline.

Primary checks when pathway enrichment is empty (common for broad traits like BMI):

- `summarise_baseline_traits()` / `summarise_snp_group_traits()` / `compare_group_traits()`
- `summarise_baseline_tissues()` / `enrich_snp_group_tissues()`

Pathway checks remain available via `enrich_trait_pathways()`, `enrich_snp_group_pathways()`, and `compare_group_pathways()`.

## Workflow diagram

The flowchart below matches the Mermaid design used in method planning. It is rendered with **DiagrammeR** (`grViz`, Graphviz DOT) so it builds offline inside package vignettes.

```{r workflow-diagram, echo=FALSE, fig.height=14, fig.width=12}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.4 ranksep=0.5]

  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.15]

  subgraph cluster_prefs {
    label=\"Our 3 Clustering Preferences\"
    style=filled color=lightgrey
    P1 [label=\"1. Handles High Sparsity\" fillcolor=\"#e1f5fe\" color=\"#01579b\"]
    P2 [label=\"2. Signed & Sign-Driven\" fillcolor=\"#e1f5fe\" color=\"#01579b\"]
    P3 [label=\"3. Automated K Detection\" fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  }

  subgraph cluster_prep {
    label=\"Dimensionality Reduction & Data Prep\"
    style=filled color=lightgrey
    D1 [label=\"Raw Studies: Beta & SE\" fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D2 [label=\"Filter Pathway Size\" shape=diamond fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D3 [label=\"15 to 200 Genes\\n(Optimize Resolution)\" fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D4 [label=\">500 Genes\\n(Prune Umbrella Folders)\" fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D5 [label=\"Multi-Pathway Gene?\" shape=diamond fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D6 [label=\"Concatenate Labels\\n(1-3 pathways)\" fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D7 [label=\"Collapse to Parent\\nOR Duplicate Rows\" fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D8 [label=\"Impute Gaps: Y=0, S=1e6\" fillcolor=\"#fff3e0\" color=\"#ef6c00\"]
    D1 -> D2
    D2 -> D3 [label=\"15-200\"]
    D2 -> D4 [label=\">500\"]
    D3 -> D5
    D5 -> D6 [label=\"1-3\"]
    D5 -> D7 [label=\">3\"]
    D6 -> D8
    D7 -> D8
    D4 -> D8
  }

  M [label=\"Select Statistical Method\" shape=diamond fillcolor=white]

  subgraph cluster_methods {
    label=\"Method Evaluation\"
    style=filled color=lightgrey
    M1 [label=\"EBMF (flashier)\\nMatrix / Soft Clusters\" fillcolor=\"#c8e6c9\" color=\"#2e7d32\" penwidth=2]
    M2 [label=\"Signed Leiden\\nGraph / Hard Clusters\" fillcolor=\"#fff9c4\" color=\"#fbc02d\"]
    M3 [label=\"Signed Louvain\\n(local traps)\" fillcolor=\"#efebe9\" color=\"#4e342e\" style=\"filled,dashed\"]
    M4 [label=\"Leading Eigenvector\\n(dense bottleneck)\" fillcolor=\"#efebe9\" color=\"#4e342e\" style=\"filled,dashed\"]
  }

  subgraph cluster_tune {
    label=\"EBMF Optimisation Loops\"
    style=filled color=lightgrey
    T1 [label=\"Adjust Parameters\" fillcolor=\"#e8f5e9\" color=\"#2e7d32\"]
    T2 [label=\"Prior: ebnm_point_laplace\" fillcolor=\"#e8f5e9\" color=\"#2e7d32\"]
    T3 [label=\"Noise: var_type=1 + Labels\" fillcolor=\"#e8f5e9\" color=\"#2e7d32\"]
    T4 [label=\"Search: higher greedy_Kmax\" fillcolor=\"#e8f5e9\" color=\"#2e7d32\"]
    T5 [label=\"Isolate global intercept\" fillcolor=\"#e8f5e9\" color=\"#2e7d32\"]
    T1 -> T2
    T1 -> T3
    T1 -> T4
    T1 -> T5
  }

  O [label=\"Output:\\nSparse, Decoupled\\nBiological SNP Clusters\" fillcolor=\"#c8e6c9\" color=\"#2e7d32\" penwidth=2]

  P1 -> D1 [style=invis]
  D8 -> M
  M -> M1 [label=\"soft\"]
  M -> M2 [label=\"hard\"]
  M -> M3 [style=dashed label=\"legacy\"]
  M -> M4 [style=dashed label=\"legacy\"]
  M1 -> T1
  T2 -> O
  T3 -> O
  T4 -> O
  T5 -> O
  M2 -> O [style=dashed]
}
"

DiagrammeR::grViz(workflow_dot)
```

