---
title: "Investigation of EBMF analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Investigation of EBMF analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(flashier)
library(ebnm)
```

This vignette clusters SNPs using **Empirical Bayes Matrix Factorization (EBMF)** via `flashier`. Rather than computing pairwise SNP similarity and then clustering, EBMF discovers latent programs from a features x SNP z-score matrix. SNPs can belong to multiple programs.

Pipeline (aligned with the SNP-perturbation investigation):

1. Build / filter / orient the EBMF matrix (phenotypic traits + Stouffer-collapsed genes)
2. **Verification (baseline):** top associated traits + tissue composition
3. Fit an EBNM x `var_type` comparison grid; pick a run by total PVE
4. **Verification (programs):** top traits and tissue enrichment per program
5. Pathway enrichment as a secondary check (often empty for BMI)

We use **BMI** (trait id `1992`).

```{r params}
target_trait_id <- 1992
magnitude_threshold <- 0.25
lfsr_threshold <- 0.05
min_group_size <- 5
min_snp_signals <- 5L
compress_method <- "asinh"
min_total_pve <- 0.01
top_n_traits <- 10L
```

## Step 1: Construct the features x SNP matrix

- **Phenotypic rows:** signed \(z = \beta / \mathrm{SE}\) (best \(p\) per trait-SNP).
- **Gene rows:** molecular QTLs Stouffer-collapsed to genes, then kept only if seen in
  at least `min_snp_signals` SNPs (same sparsity rule as phenotypic traits).
- SNP columns are oriented by \(\mathrm{sign}(z_{\mathrm{target}})\).
- Soft \(\mathrm{asinh}(z)\) compression limits leverage of extreme / corrupt values
  without a hard clip (same helper as SNP perturbation clustering).
- Default flash fits use unit \(S = 1\) with `NA` missingness.

```{r step1-ebmf-matrix}
coloc_groups <- trait(target_trait_id, include_associations = TRUE)$coloc_groups
ebmf_data <- build_ebmf_matrix(
  trait_id = target_trait_id,
  snp_key = "variant_id",
  min_snp_signals = min_snp_signals,
  compress_method = compress_method
)

Y <- ebmf_data$beta_matrix
S <- ebmf_data$se_matrix
trait_matrix <- ebmf_data$verification_trait_matrix

knitr::kable(
  ebmf_data$prep_summary |>
    dplyr::mutate(fraction_observed = round(fraction_observed, 3)),
  caption = "EBMF matrix prep summary",
  align = "r"
)
```

## Verification 1: baseline traits and tissues

Before clustering, summarise the pooled phenotypic profile (same helpers as the
perturbation vignette). Pathway enrichment is deferred to a secondary check.

```{r verify1-baseline-traits}
baseline_traits <- summarise_baseline_traits(
  trait_matrix = trait_matrix,
  trait_info = ebmf_data$verification_trait_info,
  coloc_groups = coloc_groups,
  exclude_trait_ids = target_trait_id,
  top_n = top_n_traits,
  n_categories = 2L
)

knitr::kable(
  baseline_traits$traits |>
    dplyr::mutate(
      mean_z = round(mean_z, 3),
      abs_mean_z = round(abs_mean_z, 3)
    ),
  caption = "Baseline top associated traits",
  row.names = FALSE
)

knitr::kable(
  baseline_traits$categories |>
    dplyr::mutate(score = round(score, 2)),
  caption = "Baseline top trait categories",
  row.names = FALSE
)
```

```{r verify1-baseline-tissues}
baseline_tissues <- summarise_baseline_tissues(
  snp_ids = rownames(trait_matrix),
  coloc_groups = coloc_groups,
  snp_key = "variant_id"
)

knitr::kable(
  utils::head(baseline_tissues, 15) |>
    dplyr::mutate(frac_links = round(frac_links, 3)),
  caption = "Baseline QTL tissue composition (top 15)",
  row.names = FALSE
)
```

## Step 2: Run EBMF comparison grid

Compare EBNM priors and residual variance on the same clipped matrix.
Defaults: `se_mode = "unit"` with `var_type` in `{NULL, 1}` (no Kronecker).

```{r step2-comparison}
ebmf_comparison <- run_ebmf_comparison(
  trait_id = target_trait_id,
  snp_key = "variant_id",
  min_snp_signals = min_snp_signals,
  compress_method = compress_method,
  ebnm_fns = c("point_normal", "point_laplace"),
  se_modes = c("unit"),
  var_types = list(NULL, 1L),
  greedy_Kmax = 50L,
  backfit = TRUE,
  lfsr_threshold = lfsr_threshold,
  magnitude_threshold = magnitude_threshold,
  verbose = 1L
)
```

```{r step2-summary}
ebmf_comparison$summary |>
  dplyr::mutate(
    total_pve = round(total_pve, 4),
    elbo = round(elbo, 1)
  )
```

## Step 3: Explore the selected run

Only accept runs with `total_pve >= min_total_pve`. Near-null fits are rejected
rather than ranked by tiny differences like `1e-8` vs `1e-9`.

```{r step3-pick-run}
run_id <- select_ebmf_comparison_run(
  ebmf_comparison,
  min_total_pve = min_total_pve
)
if (is.null(run_id)) {
  stop(
    "No EBMF run reached min_total_pve = ", min_total_pve,
    ". Inspect the comparison summary before interpreting soft membership."
  )
}
run <- ebmf_comparison$results[[run_id]]
flash_fit <- run$flash_fit

cat("Analysing run:", run_id, "\n")
list(
  n_factors = flash_fit$n_factors,
  elbo = flash_fit$elbo,
  pve = round(flash_fit$pve, 4),
  total_pve = round(sum(flash_fit$pve), 4),
  magnitude_threshold = magnitude_threshold,
  lfsr_threshold = lfsr_threshold
)
```

```{r step3-pve}
knitr::kable(
  data.frame(
    program = seq_along(flash_fit$pve),
    pve = round(flash_fit$pve, 6),
    stringsAsFactors = FALSE
  ),
  caption = "Per-factor PVE (full fit)",
  align = "rr"
)
```

### All programs (strict membership)

```{r step3-membership-all}
ebmf_clusters_all <- extract_ebmf_clusters(
  flash_fit,
  lfsr_threshold = lfsr_threshold,
  magnitude_threshold = magnitude_threshold
)

program_sizes_all <- colSums(ebmf_clusters_all$membership)
data.frame(
  program = names(program_sizes_all),
  n_snps = as.integer(program_sizes_all),
  pve = round(flash_fit$pve, 4)
) |>
  dplyr::arrange(dplyr::desc(n_snps))
```

### Drop global-looking factors (PVE-aware)

Do **not** assume factor 1 is global. Drop factors only when total PVE clears the
floor and a factor looks global by soft coverage / PVE share.

```{r step3-drop-global}
drop_kset <- identify_ebmf_global_factors(
  flash_fit = flash_fit,
  membership = ebmf_clusters_all$membership,
  min_total_pve = min_total_pve
)
flash_secondary <- remove_ebmf_factors(flash_fit, kset = drop_kset)

ebmf_clusters <- extract_ebmf_clusters(
  flash_secondary,
  lfsr_threshold = lfsr_threshold,
  magnitude_threshold = magnitude_threshold
)
program_drivers <- summarise_ebmf_program_drivers(
  flash_fit = flash_secondary,
  trait_info = ebmf_data$trait_info
)

list(
  dropped_factors = if (length(drop_kset) == 0) {
    "none"
  } else {
    paste(drop_kset, collapse = ", ")
  },
  n_factors_full = flash_fit$n_factors,
  n_factors_kept = flash_secondary$n_factors,
  pve_kept = round(flash_secondary$pve, 4),
  n_assigned = ebmf_clusters$n_assigned,
  n_multi_program = ebmf_clusters$n_multi_program
)
```

```{r step3-membership}
program_sizes <- colSums(ebmf_clusters$membership)
data.frame(
  program = names(program_sizes),
  n_snps = as.integer(program_sizes),
  pve = round(flash_secondary$pve, 4)
) |>
  dplyr::arrange(dplyr::desc(n_snps))
```

### Feature loadings (L)

```{r step3-drivers}
if (nrow(program_drivers) > 0) {
  program_drivers |>
    dplyr::group_by(program) |>
    dplyr::slice_head(n = 5) |>
    dplyr::select(
      program, feature_type, trait_label, trait_name, loading, lfsr
    ) |>
    dplyr::mutate(
      loading = round(loading, 3),
      lfsr = round(lfsr, 4)
    )
} else {
  cat("No significant feature drivers found.\n")
}
```

### SNP factor loading heatmap

```{r step3-heatmap, fig.height = 7, fig.width = 7}
K <- flash_secondary$n_factors
if (K > 0 && ebmf_clusters$n_assigned > 0) {
  F_pm <- flash_secondary$F_pm
  assigned_snps <- rownames(ebmf_clusters$membership)[
    rowSums(ebmf_clusters$membership) > 0
  ]
  F_assigned <- F_pm[assigned_snps, , drop = FALSE]
  colnames(F_assigned) <- paste0("Program ", seq_len(K))

  heatmap(
    F_assigned,
    scale = "none",
    main = paste("SNP loadings -", run_id),
    xlab = "Programs",
    ylab = "SNPs",
    cexCol = 0.8
  )
} else {
  cat("No assigned SNPs to plot.\n")
}
```

## Verification 2: module-specific traits and tissues

Ask whether programs recover drivers elevated vs other SNPs (trait/category
specificity) and tissues enriched vs leave-one-out rest.

```{r verify2-traits}
group_traits <- summarise_module_specific_traits(
  trait_matrix = trait_matrix,
  groups = ebmf_clusters$clusters,
  trait_info = ebmf_data$verification_trait_info,
  coloc_groups = coloc_groups,
  exclude_trait_ids = target_trait_id,
  min_group_size = min_group_size,
  top_n = top_n_traits,
  n_categories = 2L
)
trait_comparison <- compare_group_traits(
  baseline_traits = baseline_traits,
  group_traits = group_traits,
  top_n = top_n_traits
)

knitr::kable(
  group_traits$summary |>
    dplyr::mutate(
      top_specificity = round(top_specificity, 2),
      top_abs_mean_z = round(top_abs_mean_z, 3)
    ),
  caption = paste0(
    "Module-specific traits/categories per program (> ",
    min_group_size,
    " SNPs)"
  ),
  na = "",
  row.names = FALSE
)

knitr::kable(
  trait_comparison$group_overlap,
  caption = "Per-program overlap with baseline top traits",
  row.names = FALSE
)

knitr::kable(
  trait_comparison$trait_status |>
    dplyr::filter(status != "baseline_only" | n_groups > 0),
  caption = "Baseline vs program trait-driver status",
  na = "",
  row.names = FALSE
)
```

```{r verify2-trait-details, echo=FALSE, results='asis'}
for (x in group_traits$by_group) {
  cat("\n\n### Program ", x$group, " traits\n\n", sep = "")
  cat("SNPs: ", x$n_snps, "\n\n", sep = "")
  cat("Leading specific categories: ", x$top_categories, "\n\n", sep = "")
  if (nrow(x$categories) > 0) {
    print(knitr::kable(
      x$categories |>
        dplyr::mutate(
          score = round(score, 2),
          mean_specificity = round(mean_specificity, 2)
        ),
      caption = "Module-specific trait categories",
      row.names = FALSE
    ))
  }
  if (nrow(x$traits) > 0) {
    print(knitr::kable(
      x$traits |>
        dplyr::mutate(
          mean_z = round(mean_z, 3),
          abs_mean_z = round(abs_mean_z, 3),
          abs_mean_z_rest = round(abs_mean_z_rest, 3),
          specificity = round(specificity, 2),
          abs_mean_z_delta = round(abs_mean_z_delta, 3)
        ),
      caption = "Module-specific traits",
      row.names = FALSE
    ))
  }
}
```

```{r verify2-tissues}
group_tissues <- enrich_snp_group_tissues(
  groups = ebmf_clusters$clusters,
  coloc_groups = coloc_groups,
  snp_key = "variant_id",
  min_group_size = min_group_size,
  background = "rest"
)

knitr::kable(
  group_tissues$summary,
  caption = paste0(
    "Module-specific tissue enrichment (> ",
    min_group_size,
    " SNPs; vs rest)"
  ),
  na = "",
  row.names = FALSE
)
```

```{r verify2-tissue-details, echo=FALSE, results='asis'}
for (x in group_tissues$by_group) {
  cat("\n\n### Program ", x$group, " tissues\n\n", sep = "")
  cat("SNPs: ", x$n_snps, "\n\n", sep = "")
  if (nrow(x$enriched) > 0) {
    print(knitr::kable(
      x$enriched |>
        dplyr::mutate(
          fold_enrichment = round(fold_enrichment, 2),
          p_value = signif(p_value, 3),
          fdr = signif(fdr, 3)
        ),
      caption = "Enriched tissues vs rest",
      row.names = FALSE
    ))
  } else {
    cat("No significantly enriched tissues.\n\n")
  }
}
```

## Verification 3 (secondary): pathways per program

```{r verify3-pathways}
baseline_pathways <- enrich_trait_pathways(
  trait_id = target_trait_id,
  coloc_groups = coloc_groups,
  sources = c("KEGG", "Reactome")
)
program_enrichment <- enrich_snp_group_pathways(
  groups = ebmf_clusters$clusters,
  coloc_groups = coloc_groups,
  min_group_size = min_group_size,
  sources = c("KEGG", "Reactome")
)
pathway_comparison <- compare_group_pathways(baseline_pathways, program_enrichment)

knitr::kable(
  baseline_pathways$summary,
  caption = "Baseline trait pathway enrichment (KEGG + Reactome)",
  na = "",
  align = c("r", "r", "r", "r", "l")
)

knitr::kable(
  program_enrichment$summary,
  caption = paste0(
    "Pathway enrichment per secondary EBMF program (> ",
    min_group_size,
    " SNPs)"
  ),
  na = "",
  align = c("l", "r", "r", "r", "l")
)

knitr::kable(
  pathway_comparison$group_overlap,
  caption = "Per-program overlap with baseline trait pathways",
  na = "",
  align = c("l", "r", "r", "r")
)
```

## Save results

```{r step4-save}
results_path <- file.path(
  pkg_root, "vignettes", "investigation-ebmf-analysis-results.rds"
)
saveRDS(
  list(
    comparison = ebmf_comparison,
    run_id = run_id,
    drop_kset = drop_kset,
    compress_method = compress_method,
    min_total_pve = min_total_pve,
    magnitude_threshold = magnitude_threshold,
    lfsr_threshold = lfsr_threshold,
    flash_fit = flash_fit,
    flash_secondary = flash_secondary,
    ebmf_clusters_all = ebmf_clusters_all,
    ebmf_clusters = ebmf_clusters,
    program_drivers = program_drivers,
    baseline_traits = baseline_traits,
    group_traits = group_traits
  ),
  results_path
)
results_path
```

Reload later with:

```r
saved <- readRDS("vignettes/investigation-ebmf-analysis-results.rds")
saved$comparison$summary
saved$ebmf_clusters
```
