---
title: "Simulation study of multi-trait program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # --- generative structure -------------------------------------------------
  n_loci_per_trait: 200
  n_shared_loci: 80
  K_shared: 3
  K_specific: 2
  module_size: 15
  n_drivers_per_program: 8
  n_background_traits: 300
  effect_size: 6
  noise_sd: 1
  p_structural_zero: 0
  # --- background realism ---------------------------------------------------
  p_active_background: 0.02
  background_sparsity_sd: 1.2
  background_effect_scale: 0.5
  background_corr: 0
  snp_pleiotropy_sd: 0
  # --- the two axes this study exists to separate ---------------------------
  # Crossed, not varied one at a time: the whole question is how the two
  # statistics behave as locus overlap falls while profile overlap is held up.
  locus_overlap: [0.0, 0.25, 0.5, 1.0]
  profile_overlap: [0.4, 0.8]
  # --- cross-trait comparison ----------------------------------------------
  n_perm: 1000
  fdr_threshold: 0.05
  # --- replication ----------------------------------------------------------
  n_sim: 50
  master_seed: 1
  # Replicates within a version run in parallel via parallel::mclapply
  # (fork-based; falls back to serial on Windows or when cores = 1).
  cores: 5
  # quick: fast test mode. Caps n_sim at 2 and lowers the permutation counts.
  # All input versions are always run. Set FALSE for the full study.
  quick: false
vignette: >
  %\VignetteIndexEntry{Simulation study of multi-trait program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, out.width = "100%", fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)

quick <- isTRUE(params$quick)
n_sim <- if (quick) min(as.integer(params$n_sim), 2L) else as.integer(params$n_sim)
n_perm <- if (quick) 300L else as.integer(params$n_perm)
coherence_perm <- if (quick) 300L else 2000L

# Shared EBMF settings, from the same file the univariate and real-data studies
# read, so the per-trait discovery step here is the same discovery step used
# everywhere else.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

set.seed(params$master_seed)
sim_seeds <- sample.int(.Machine$integer.max, n_sim)

generative_base <- list(
  n_loci_per_trait = as.integer(params$n_loci_per_trait),
  n_shared_loci = as.integer(params$n_shared_loci),
  K_shared = as.integer(params$K_shared),
  K_specific = as.integer(params$K_specific),
  module_size = as.integer(params$module_size),
  n_drivers_per_program = as.integer(params$n_drivers_per_program),
  n_background_traits = as.integer(params$n_background_traits),
  effect_size = params$effect_size,
  noise_sd = params$noise_sd,
  p_structural_zero = params$p_structural_zero,
  p_active_background = params$p_active_background,
  background_sparsity_sd = params$background_sparsity_sd,
  background_effect_scale = params$background_effect_scale,
  background_corr = params$background_corr,
  snp_pleiotropy_sd = params$snp_pleiotropy_sd
)
```

This vignette asks one question with known ground truth: **when two traits share a
biological program, which cross-trait statistic finds it?**

There are two candidates, and they make different assumptions.
`compare_program_pairs_loadings()` scores a pair of programs as an inner product
over the loci both traits carry. `compare_program_pairs_profiles()` scores them
over the background studies both traits load on. The first requires the shared
program to act at the *same variants* in both traits; the second does not.

That difference is not testable on real data, because there the truth is unknown
and the two axes have very different sizes (for BMI and height, 73 shared loci
against 656 shared background studies) so any disagreement is uninterpretable.
Here both axes are planted deliberately and independently:
`simulate_trait_pair()` controls **locus overlap** (what fraction of a shared
program's SNPs are the same locus in both traits) separately from **profile
overlap** (what fraction of its driver studies are the same).

The decisive cell of the design is **low locus overlap with high profile
overlap**: a program genuinely shared between the traits, acting at different
variants in each. If the profile axis recovers it there and the locus axis does
not, the profile axis earns its place. If it does not, the locus axis is
sufficient and the extra statistic should be dropped.

## Simulation design (ADEMP)

Following Morris, White and Crowther (2019, *Statistics in Medicine* 38:2074-2102),
as the univariate studies do.

**Aims.** (i) Establish whether the trait-profile axis recovers shared programs
that the locus axis cannot, as a function of locus overlap. (ii) Measure the
false-discovery rate of each axis when nothing is shared.

**Data-generating mechanism.** `simulate_trait_pair()`. Two target traits each
carry `n_loci_per_trait` loci, of which `n_shared_loci` sit in a pool both
traits carry; `K_shared` programs are planted in both traits and `K_specific` in
only one. For a shared program, `locus_overlap` sets the fraction of its SNPs
drawn from the shared pool (the rest are private to each trait) and
`profile_overlap` the fraction of its driver studies common to both. Half the
shared programs are planted antagonistic (the second trait's driver effects
flipped) so direction is scoreable in both senses. Background studies are drawn
from one pool shared by both traits, so the feature axis intersects.

**Estimand / target.** The planted cross-trait correspondence
(`ground_truth$correspondence`): which program in trait A is the same program as
which in trait B, and whether it is concordant or antagonistic. This is a
selection-plus-sign target, not a parameter.

**Methods.** Two comparators on identical data, which is the point of the study:
`compare_program_pairs_profiles()` and `compare_program_pairs_loadings()`, each
after the same per-trait discovery (`run_univariate_clustering()` then
`summarise_ebmf_programs()`, at the shared settings). Only `primary`-tier links
are scored, that being the calibrated one-to-one claim.

**Performance measures.** Discovered programs are mapped onto planted programs by
`map_programs_to_planted()` -- on SNP membership for the locus axis and on
loading-weighted driver overlap for the feature axis, because the feature table
is dense and a set overlap would reject a perfect match. Then
`evaluate_multitrait_simulation()` reports:

- `precision` = true-positive links / discovered links. A method that links
  everything is penalised.
- `recall` = planted shared programs recovered / `K_shared`. Credited once per
  planted program, so EBMF factor splitting cannot inflate it.
- `sign_accuracy` = fraction of true-positive links whose reported direction
  matches the planted direction. Scored on true positives only: a link that is
  not real has no meaningful direction.
- Under the null (`K_shared = 0`) every link is a false discovery, so `n_links`
  is the FDR numerator directly.

**Monte Carlo uncertainty.** Every estimate is a mean over `n_sim` replicates
with a Monte Carlo standard error; the replicate is the MC unit. Common random
numbers are used -- one seed vector shared across all input versions -- so
version-to-version differences are estimated more precisely than with
independent seeds.

```{r settings-table}
knitr::kable(
  data.frame(
    parameter = c(names(generative_base), "locus_overlap", "profile_overlap",
                  "n_sim", "n_perm", "fdr_threshold", "quick",
                  "ebmf_settings_signature"),
    value = c(
      vapply(generative_base, function(x) paste(x, collapse = ", "), character(1)),
      paste(params$locus_overlap, collapse = ", "),
      paste(params$profile_overlap, collapse = ", "),
      n_sim, n_perm, params$fdr_threshold, quick, settings_signature
    ),
    stringsAsFactors = FALSE
  ),
  row.names = FALSE,
  caption = paste0(
    "Active settings. ebmf_settings_signature identifies the shared analysis ",
    "settings this run resolved from _ebmf_settings.yml; two studies carrying ",
    "the same signature were run on the same discovery settings."
  )
)
```

```{r run-helpers}
discover_one_trait <- function(trait_object, trait_id) {
  res <- run_univariate_clustering(
    trait_object,
    target_trait_id = as.integer(trait_id),
    trait_subset = ebmf_settings$trait_subset,
    min_snp_signals = as.integer(ebmf_settings$min_snp_signals),
    compress_method = ebmf_settings$compress_method,
    compress_scale = as.numeric(ebmf_settings$compress_scale),
    ebmf_lfsr_threshold = as.numeric(ebmf_settings$ebmf_lfsr_threshold),
    ebmf_magnitude_threshold = as.numeric(ebmf_settings$ebmf_magnitude_threshold),
    ebmf_greedy_Kmax = as.integer(ebmf_settings$ebmf_greedy_Kmax),
    ebmf_prior = ebmf_settings$ebmf_prior,
    ebmf_backfit = isTRUE(ebmf_settings$ebmf_backfit),
    min_module_size = as.integer(ebmf_settings$min_module_size)
  )
  smry <- summarise_ebmf_programs(
    res, s_matrix = res$s_matrix,
    edge_threshold = as.numeric(ebmf_settings$similarity_threshold),
    min_module_size = as.integer(ebmf_settings$min_module_size),
    coherence_n_perm = coherence_perm,
    coherence_q_threshold = as.numeric(ebmf_settings$coherence_q),
    n_null = 2, n_rep = 0, verbose = FALSE
  )
  return(extract_program_loadings(res, smry, valid_only = TRUE))
}

run_one_replicate <- function(spec, seed) {
  # modifyList, not c(): the null version overrides K_shared, and c() would pass
  # it twice.
  sp <- do.call(
    simulate_trait_pair,
    c(utils::modifyList(generative_base, spec), list(seed = seed))
  )
  extraction <- lapply(names(sp$traits), function(tid) {
    discover_one_trait(sp$traits[[tid]], tid)
  })
  names(extraction) <- names(sp$traits)

  loadings <- dplyr::bind_rows(lapply(extraction, function(x) x$loadings))
  profiles <- dplyr::bind_rows(lapply(extraction, function(x) x$profiles))
  if (nrow(loadings) == 0) {
    return(NULL)
  }

  prof <- compare_program_pairs_profiles(
    extraction, n_perm = n_perm, fdr_threshold = params$fdr_threshold, seed = 1
  )
  loci <- compare_program_pairs_loadings(
    extraction, n_perm = n_perm, fdr_threshold = params$fdr_threshold, seed = 1
  )

  map_profile <- map_programs_to_planted(profiles, sp$ground_truth, by = "drivers")
  map_locus <- map_programs_to_planted(loadings, sp$ground_truth, by = "snps")

  axis_rows <- dplyr::bind_rows(
    cbind(
      axis = "profile",
      evaluate_multitrait_simulation(prof$pairs, map_profile, sp$ground_truth)
    ),
    cbind(
      axis = "locus",
      evaluate_multitrait_simulation(loci$pairs, map_locus, sp$ground_truth)
    )
  )
  axis_rows$n_valid_a <- length(unique(extraction[[1]]$loadings$program_id))
  axis_rows$n_valid_b <- length(unique(extraction[[2]]$loadings$program_id))
  axis_rows$n_features_shared <- {
    u <- shared_feature_universe(extraction)
    if (nrow(u) > 0) u$n_features_shared[1] else NA_integer_
  }
  return(axis_rows)
}

mc_summary <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) {
    return(list(est = NA_real_, se = NA_real_, n = 0L))
  }
  return(list(
    est = mean(x),
    se = if (length(x) > 1) stats::sd(x) / sqrt(length(x)) else NA_real_,
    n = length(x)
  ))
}

fmt_mc <- function(x, digits = 2) {
  s <- mc_summary(x)
  if (!is.finite(s$est)) {
    return("--")
  }
  if (!is.finite(s$se)) {
    return(formatC(s$est, format = "f", digits = digits))
  }
  return(sprintf(
    "%s (%s)", formatC(s$est, format = "f", digits = digits),
    formatC(s$se, format = "f", digits = digits)
  ))
}
```

```{r build-versions}
versions <- list()
for (po in as.numeric(params$profile_overlap)) {
  for (lo in as.numeric(params$locus_overlap)) {
    versions[[length(versions) + 1L]] <- list(
      label = sprintf("locus_overlap = %.2f, profile_overlap = %.2f", lo, po),
      family = "shared",
      spec = list(locus_overlap = lo, profile_overlap = po)
    )
  }
}
# The null: two traits with programs of their own and no correspondence. Any
# link is a false discovery, so this is the FDR arm.
versions[[length(versions) + 1L]] <- list(
  label = "null: K_shared = 0 (no shared programs)",
  family = "null",
  spec = list(locus_overlap = 0, profile_overlap = 0, K_shared = 0L)
)
```

```{r run-study}
# Runs one replicate and reports its outcome via the return value rather than
# a `<<-` side effect, so error tracking survives being called from a forked
# mclapply worker (a child process's `<<-` never reaches the parent).
run_replicate_safe <- function(spec, r, seed) {
  err_message <- NULL
  out <- tryCatch(
    run_one_replicate(spec, seed),
    error = function(e) {
      err_message <<- conditionMessage(e)
      return(NULL)
    }
  )
  if (is.null(out)) {
    return(list(row = NULL, error = err_message, rep = r))
  }
  out$rep <- r
  return(list(row = out, error = NULL, rep = r))
}

n_jobs <- max(1L, min(as.integer(params$cores), n_sim))

replicate_errors <- list()
results <- dplyr::bind_rows(lapply(seq_along(versions), function(v) {
  spec <- versions[[v]]$spec
  rep_out <- if (n_jobs > 1 && .Platform$OS.type != "windows") {
    parallel::mclapply(
      seq_len(n_sim),
      function(r) run_replicate_safe(spec, r, sim_seeds[[r]]),
      mc.cores = n_jobs,
      mc.set.seed = FALSE
    )
  } else {
    lapply(seq_len(n_sim), function(r) run_replicate_safe(spec, r, sim_seeds[[r]]))
  }

  # Recorded rather than swallowed: a version that silently drops every
  # replicate would otherwise render as an empty row.
  errs <- dplyr::bind_rows(lapply(rep_out, function(x) {
    if (is.null(x$error)) {
      return(NULL)
    }
    return(data.frame(
      version = versions[[v]]$label, rep = x$rep,
      message = x$error, stringsAsFactors = FALSE
    ))
  }))
  if (nrow(errs) > 0) {
    replicate_errors[[length(replicate_errors) + 1L]] <<- errs
  }

  rows <- dplyr::bind_rows(lapply(rep_out, `[[`, "row"))
  if (nrow(rows) == 0) {
    return(NULL)
  }
  rows$version <- versions[[v]]$label
  rows$family <- versions[[v]]$family
  return(rows)
}))
```

```{r replicate-errors, results='asis'}
if (length(replicate_errors) > 0) {
  err <- dplyr::bind_rows(replicate_errors)
  print(knitr::kable(
    err |>
      dplyr::group_by(version, message) |>
      dplyr::summarise(n_failed = dplyr::n(), .groups = "drop"),
    row.names = FALSE,
    caption = paste0(
      "Replicates that failed. Any row here means the version below is ",
      "reported on fewer replicates than n_sim, or not at all."
    )
  ))
}
```

## Does the profile axis recover programs the locus axis cannot?

This is the headline table. Read the `locus_overlap = 0` rows first: there the
two traits share a program but **no loci**, so the locus axis has nothing to form
an inner product over, while the same program is still visible in the driver
profile.

```{r headline}
shared_res <- results[results$family == "shared", , drop = FALSE]
if (nrow(shared_res) == 0) {
  cat("No replicates completed.\n\n")
} else {
  tbl <- shared_res |>
    dplyr::group_by(version, axis) |>
    dplyr::summarise(
      links = fmt_mc(n_links, 1),
      recall = fmt_mc(recall),
      precision = fmt_mc(precision),
      sign_accuracy = fmt_mc(sign_accuracy),
      .groups = "drop"
    ) |>
    tidyr::pivot_wider(
      names_from = axis,
      values_from = c(links, recall, precision, sign_accuracy)
    ) |>
    dplyr::arrange(version)
  print(knitr::kable(
    tbl, row.names = FALSE,
    caption = paste0(
      "Link recovery by axis and input version (mean over ", n_sim,
      " replicates, Monte Carlo SE in parentheses). recall is over planted ",
      "shared programs, precision over discovered links, and sign_accuracy is ",
      "computed on true-positive links only."
    )
  ))
}
```

```{r headline-plot, fig.height=3.6}
if (requireNamespace("ggplot2", quietly = TRUE) && nrow(shared_res) > 0) {
  plot_df <- shared_res |>
    dplyr::group_by(version, axis) |>
    dplyr::summarise(
      recall = mean(recall, na.rm = TRUE),
      se = stats::sd(recall, na.rm = TRUE) / sqrt(dplyr::n()),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      locus_overlap = as.numeric(sub(".*locus_overlap = ([0-9.]+).*", "\\1", version)),
      profile_overlap = sub(".*profile_overlap = ", "", version)
    )
  print(
    ggplot2::ggplot(
      plot_df,
      ggplot2::aes(locus_overlap, recall, colour = axis, group = axis)
    ) +
      ggplot2::geom_line() +
      ggplot2::geom_point(size = 2) +
      ggplot2::facet_wrap(~ paste("profile overlap", profile_overlap)) +
      ggplot2::scale_y_continuous(limits = c(0, 1)) +
      ggplot2::labs(
        x = "locus overlap (fraction of a shared program's SNPs common to both traits)",
        y = "recall of planted shared programs", colour = "axis"
      ) +
      ggplot2::theme_minimal()
  )
}
```

## False discoveries when nothing is shared

Under `K_shared = 0` the two traits have programs of their own and no
correspondence, so every reported link is a false discovery and `links` is the
FDR numerator directly.

```{r null-arm}
null_res <- results[results$family == "null", , drop = FALSE]
if (nrow(null_res) == 0) {
  cat("The null arm produced no completed replicates.\n\n")
} else {
  print(knitr::kable(
    null_res |>
      dplyr::group_by(axis) |>
      dplyr::summarise(
        false_links = fmt_mc(n_links, 2),
        reps_with_any_link = sprintf(
          "%d of %d", sum(n_links > 0), dplyr::n()
        ),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "False-discovery output with no shared programs planted. false_links is ",
      "the mean number of primary-tier links per replicate (Monte Carlo SE in ",
      "parentheses); any value above 0 is a false discovery."
    )
  ))
}
```

## Scale of each comparison

Context for the tables above: how many programs survived per-trait validation,
and how wide each matching axis was.

```{r scale}
if (nrow(results) > 0) {
  print(knitr::kable(
    results |>
      dplyr::filter(axis == "profile") |>
      dplyr::group_by(version) |>
      dplyr::summarise(
        valid_programs_trait_a = fmt_mc(n_valid_a, 1),
        valid_programs_trait_b = fmt_mc(n_valid_b, 1),
        shared_features = fmt_mc(n_features_shared, 0),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "Programs surviving per-trait validation and the size of the shared ",
      "feature axis, per input version. A version where few programs survive ",
      "validation cannot support either comparison, whatever the recall ",
      "columns say."
    )
  ))
}
```

## Notes for interpreting runs

- **The two axes are scored on identical data.** Any difference between them is
  a property of the statistics, not of the simulation, which is why this
  comparison is meaningful even where the absolute recall numbers depend on the
  generative settings.
- **Recall is credited per planted program, not per link.** EBMF splits one
  planted module into several factors, so counting links would reward splitting.
- **`sign_accuracy` is conditional on a true positive.** A high value alongside a
  low recall means the few links found were directionally right, not that
  direction is well recovered overall.
- **The mapping differs by axis by necessity.** The SNP table is sparse and maps
  by set overlap; the feature table is dense (every background study carries some
  loading on every program) so it maps by loading-weighted share against the
  evenly-spread expectation. Using a set overlap on the dense table rejects even
  a perfect match.
- **Background realism is off by default.** `background_corr` and
  `snp_pleiotropy_sd` default to 0, reproducing the historical independent,
  random-signed background. The headline contrast between the two axes is
  structural and largely insensitive to that, but the **null arm is not**: an FDR
  measured in an unrealistic noise regime is as weak here as it is in the
  univariate null study. Run the realism arms of the grid script before quoting
  the FDR.
- **Sample overlap is not simulated.** Two real target traits from the same
  cohort share participants, which biases cross-trait statistics away from the
  null. Nothing here reproduces that, so these FDRs are optimistic for
  same-cohort pairs.
