---
title: "Investigation of multi-trait EBMF program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of multi-trait EBMF program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "local"
  include_trans: true
  trait_ids: [1992, 1993]
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  compress_method: "asinh"
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_magnitude_threshold: 0.5
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  min_mean_internal: 0.3
  min_connectedness: 0.5
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 2
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  n_perm: 3000
  fdr_threshold: 0.05
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r trait-metadata, include=FALSE}
trait_ids <- as.character(params$trait_ids)
if (length(trait_ids) < 2) {
  stop("At least two trait_ids are required for a multi-trait comparison")
}
select_api(params$api)
trait_meta <- traits(trait_ids)
trait_name_map <- stats::setNames(
  as.character(trait_meta$traits$trait_name),
  as.character(trait_meta$traits$id)
)
trait_name_map[is.na(trait_name_map)] <- names(trait_name_map)[is.na(trait_name_map)]
```

This vignette extends the single-trait investigation
(`investigation-univariate-analysis.html`) to **multiple traits**, and asks one
question: **which programs are shared between traits, and do they push the two
traits the same way or opposite ways?**

EBMF programs are discovered independently for each target trait, so each trait's
architecture is described in its own terms first. Every cross-trait pair of
programs is then scored by **loading-weighted locus concordance**: across the loci
both traits carry, how much of the same signal do the two programs load on, and
with what sign. No hard locus cutoff is applied — every shared locus contributes
in proportion to how strongly both programs load on it, so the evidence grows with
the amount of shared architecture rather than with how tightly a handful of loci
happen to align. A permutation null calibrated per program turns that into
significance, linked programs are joined into **multi-trait families**, and each
link is then given a direction two independent ways: the sign of the concordance,
and `module_rg()`, the correlation of the two traits' own SNP effects across the
module's loci.

Traits are supplied as an array in `params$trait_ids`. We use
`r paste0("**", unname(trait_name_map[trait_ids]), "** (trait id ", trait_ids, ")", collapse = ", ")`.

```{r params}
api <- params$api
# Resolved from vignettes/_ebmf_settings.yml, the same file the univariate
# study reads. Previously this vignette carried its own ebmf_magnitude_threshold
# (0.5 vs 0.75) and include_trans (FALSE vs TRUE), so it factorised a different
# matrix and its "programs" were not the univariate programs at all -- which is
# why the two studies did not line up.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

trait_subset <- ebmf_settings$trait_subset
min_module_size <- as.integer(ebmf_settings$min_module_size)
min_snp_signals <- as.integer(ebmf_settings$min_snp_signals)
compress_method <- ebmf_settings$compress_method
compress_scale <- as.numeric(ebmf_settings$compress_scale)
include_trans <- isTRUE(ebmf_settings$include_trans)
similarity_threshold <- as.numeric(ebmf_settings$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(ebmf_settings$ebmf_lfsr_threshold)
ebmf_prior <- ebmf_settings$ebmf_prior
ebmf_backfit <- isTRUE(ebmf_settings$ebmf_backfit)
ebmf_magnitude_threshold <- as.numeric(ebmf_settings$ebmf_magnitude_threshold)
ebmf_greedy_Kmax <- as.integer(ebmf_settings$ebmf_greedy_Kmax)
coherence_n_perm <- as.integer(ebmf_settings$coherence_n_perm)
coherence_q_threshold <- as.numeric(ebmf_settings$coherence_q)
associations <- params$associations
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
stability_threshold <- as.numeric(params$stability_threshold)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)
n_perm <- as.integer(params$n_perm)
fdr_threshold <- as.numeric(params$fdr_threshold)

knitr::kable(
  data.frame(
    api = api,
    trait_ids = paste(trait_ids, collapse = ", "),
    trait_subset = trait_subset,
    min_snp_signals = min_snp_signals,
    min_module_size = min_module_size,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    similarity_threshold = similarity_threshold,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    n_rep = n_rep,
    n_perm = n_perm,
    fdr_threshold = fdr_threshold,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

knitr::kable(
  data.frame(
    trait_id = trait_ids,
    trait_name = unname(trait_name_map[trait_ids]),
    stringsAsFactors = FALSE
  ),
  caption = "Target traits compared",
  row.names = FALSE
)
```

## Step 1: Univariate EBMF per trait

The existing single-trait pipeline is run independently for each target trait:
`run_univariate_clustering()` (EBMF) followed by `summarise_ebmf_programs()`.
Programs are kept when they pass the validation filters (`status == "valid"`).
Each program's posterior-mean SNP loadings (`F_pm`), lFSR, and high-confidence
SNP membership are then extracted, with loadings sign-oriented to the target
trait (see `extract_program_loadings()`).

```{r step1-run}
run_trait <- function(tid) {
  obj <- trait(tid, include_associations = TRUE)
  res <- run_univariate_clustering(
    obj,
    target_trait_id = tid,
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    max_snp_fraction = 0.8,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness
  )
  smry <- summarise_ebmf_programs(
    res,
    s_matrix = res$s_matrix,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    coherence_n_perm = coherence_n_perm,
    coherence_q_threshold = coherence_q_threshold,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    stability_threshold = stability_threshold,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = FALSE
  )
  extracted <- extract_program_loadings(res, smry, valid_only = TRUE)
  list(trait_id = tid, fit = res, summary = smry, extraction = extracted)
}

trait_results <- lapply(trait_ids, run_trait)
names(trait_results) <- trait_ids

for (tid in names(trait_results)) {
  x <- trait_results[[tid]]
  x$extraction$loadings$trait_name <- trait_name_map[[tid]]
  x$extraction$profiles$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_id <- tid
  trait_results[[tid]] <- x
}

program_meta <- dplyr::bind_rows(lapply(trait_results, function(x) {
  x$summary$programs |>
    dplyr::mutate(
      trait_id = x$trait_id,
      trait_name = trait_name_map[[x$trait_id]]
    )
}))
program_info <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$programs))
program_loadings <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$loadings))
program_data <- lapply(trait_results, function(x) x$extraction)

knitr::kable(
  program_meta |>
    dplyr::group_by(trait_id, trait_name) |>
    dplyr::summarise(
      n_programs = dplyr::n(),
      n_valid = sum(status == "valid"),
      n_valid_snps = sum(n_snps_filtered[status == "valid"]),
      .groups = "drop"
    ),
  caption = paste0(
    "Per-trait EBMF programs (valid = passes size, internal coherence, ",
    "stability, and redundancy filters; stability is not checked when ",
    "n_rep = 0)"
  ),
  row.names = FALSE
)
```

```{r step1-program-table, results='asis'}
knitr::kable(
  program_meta |>
    dplyr::filter(status == "valid") |>
    dplyr::select(
      trait_name, program, n_snps_filtered, mean_internal_similarity,
      connectedness, factor_strength, posterior_evidence
    ) |>
    dplyr::arrange(trait_name, dplyr::desc(n_snps_filtered)) |>
    dplyr::mutate(
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      factor_strength = round(factor_strength, 3),
      posterior_evidence = round(posterior_evidence, 2)
    ),
  caption = paste0(
    "Valid programs only -- everything downstream uses these. ",
    "`n_snps_filtered` is how many SNPs pass the program's lFSR/magnitude gate; ",
    "`mean_internal_similarity` and `connectedness` say how tightly those SNPs ",
    "cohere on the SNP-similarity graph; `factor_strength` is the share of ",
    "matrix signal the factor carries (it scales with program size, so small ",
    "programs are not worse for having a small value); `posterior_evidence` is ",
    "the median -log10(lFSR) of its members. Programs that failed validation are ",
    "not shown -- see the univariate vignette for that diagnostic view."
  ),
  row.names = FALSE
)
```

## Step 2: Common locus universe

Programs are compared at the level of the **`coloc_group_id`** (the shared
fine-mapped locus) rather than the exact `variant_id`, because each trait carries
a single candidate SNP per coloc group. The table below counts the candidate
loci per trait and how many coloc groups the traits share.

```{r step2-universe}
trait_locus_counts <- program_loadings |>
  dplyr::group_by(trait_id, trait_name) |>
  dplyr::summarise(
    n_programs = dplyr::n_distinct(program_id),
    n_loci_any = dplyr::n_distinct(coloc_group_id),
    n_high_conf_loci = dplyr::n_distinct(coloc_group_id[high_confidence]),
    .groups = "drop"
  )

shared_loci <- program_loadings |>
  dplyr::distinct(trait_id, coloc_group_id) |>
  dplyr::count(coloc_group_id, name = "n_traits")

knitr::kable(
  trait_locus_counts,
  caption = "Candidate locus (coloc group) universe per trait (across valid programs)",
  row.names = FALSE
)
knitr::kable(
  data.frame(
    n_loci_any = dplyr::n_distinct(program_loadings$coloc_group_id),
    n_loci_shared_by_traits = sum(shared_loci$n_traits > 1),
    max_traits_per_locus = max(shared_loci$n_traits)
  ),
  caption = "Shared coloc-group universe across traits",
  row.names = FALSE
)
```


## Step 3: Loading-weighted locus concordance

Every cross-trait pair of programs is scored on the **locus axis**: the loci that
both traits carry a loading for (`n_loci_axis`). No lFSR or magnitude threshold is
applied — each locus contributes in proportion to how strongly it loads in *both*
programs:

$$\text{locus\_concordance} = \sum_l w_l \, f_a(l) \, f_b(l), \qquad
w_l = (1 - \text{lfsr}_a(l))(1 - \text{lfsr}_b(l))$$

Two programs that load on many of the same loci, in the same direction, score
highly; a locus that matters to only one of them contributes almost nothing on its
own. Because `extract_program_loadings()` orients each program so its own target
trait loads positively, the **sign** is meaningful: positive means the two
programs rise and fall together, negative means they oppose.

## Step 2.5: Which axis can these traits even be compared on?

Two programs can only be matched on evidence both traits carry. There are two
candidate axes, and they are very different in size.

The **locus** axis is the set of `coloc_group_id`s both traits carry. Joining on
coloc group requires both traits to have been fine-mapped to the *same* signal,
which is far stricter than positional overlap, so this axis can be small even
for two well-powered traits.

The **feature** axis is the set of background studies both traits' fits load on.
Because every target trait is profiled against the same GPMap background
universe, this axis does not require the two traits to share any loci at all.

This matters because `locus_concordance` is an inner product *over shared loci*:
two programs describing the same biology at different loci score near zero by
construction. If the belief is that a shared biological program need not act at
identical variants, the locus axis cannot express it, and the feature axis is
where the comparison has to happen.

```{r step2b-axes}
feature_universe <- shared_feature_universe(program_data)
knitr::kable(
  feature_universe,
  row.names = FALSE,
  caption = paste0(
    "Size of each candidate matching axis per trait pair. n_features_shared is ",
    "the feature (background-study) axis; n_loci_shared is the locus axis. The ",
    "premise of profile matching is that the first is much larger than the ",
    "second -- where it is not, the profile axis has no more support than the ",
    "locus axis and its links should be read with the same caution."
  )
)
```

## Step 3a: Profile concordance (primary axis)

`compare_program_pairs_profiles()` scores every cross-trait pair on how
similarly the two programs load across the shared background studies, using the
same confidence weighting, the same max-statistic permutation null, and the same
primary/candidate tiering as the locus statistic. Sign is meaningful for the
same reason: each program is oriented so its own target trait loads positively.

Read `n_eff_features` the way `n_eff_loci` is read below -- a large
`concordance_z` resting on two or three effective features is thin.

```{r step3a-profiles}
prof_result <- compare_program_pairs_profiles(
  program_data,
  n_perm = n_perm,
  fdr_threshold = fdr_threshold,
  seed = 1
)
prof_pairs <- prof_result$pairs
prof_links <- prof_pairs[!is.na(prof_pairs$link_tier), , drop = FALSE]
prof_links <- prof_links[order(
  factor(prof_links$link_tier, levels = c("primary", "candidate")),
  -abs(prof_links$concordance_z)
), , drop = FALSE]

if (nrow(prof_links) == 0) {
  cat("No program pairs linked on the profile axis.\n\n")
} else {
  print(knitr::kable(
    prof_links |>
      dplyr::mutate(
        concordance_z = round(concordance_z, 2),
        q_concordance = signif(q_concordance, 3),
        profile_concordance = round(profile_concordance, 2),
        cosine_features = round(cosine_features, 3),
        n_eff_features = round(n_eff_features, 1)
      ) |>
      dplyr::select(
        program_id_a, program_id_b, link_tier, direction, concordance_z,
        q_concordance, n_features_axis, n_eff_features, profile_concordance,
        cosine_features
      ),
    row.names = FALSE,
    caption = paste0(
      "Program pairs linked on the trait-profile axis, primary tier first. ",
      "These links do not require the two programs to share any loci."
    )
  ))
}
```

```{r step3-concordance}
coloc_by_trait <- stats::setNames(
  lapply(trait_ids, function(tid) {
    return(trait(tid, include_associations = TRUE)$coloc_groups)
  }),
  trait_ids
)

conc_result <- compare_program_pairs_loadings(
  program_data,
  n_perm = n_perm,
  fdr_threshold = fdr_threshold,
  seed = 1,
  coloc_groups = coloc_by_trait
)
conc_pairs <- conc_result$pairs
conc_matches <- conc_result$matches

knitr::kable(
  data.frame(
    n_programs = nrow(conc_matches),
    n_cross_trait_pairs = nrow(conc_pairs),
    n_loci_axis = stats::median(conc_pairs$n_loci_axis, na.rm = TRUE),
    n_programs_significant = sum(conc_matches$significant, na.rm = TRUE),
    n_primary_links = sum(conc_pairs$link_tier == "primary", na.rm = TRUE),
    n_candidate_links = sum(conc_pairs$link_tier == "candidate", na.rm = TRUE),
    n_concordant = sum(
      !is.na(conc_pairs$link_tier) & conc_pairs$direction == "concordant",
      na.rm = TRUE
    ),
    n_antagonistic = sum(
      !is.na(conc_pairs$link_tier) & conc_pairs$direction == "antagonistic",
      na.rm = TRUE
    ),
    stringsAsFactors = FALSE
  ),
  caption = paste0(
    "Scale of the comparison. `n_loci_axis` is how many loci each pair is scored ",
    "on (the two traits' shared coloc-group universe). `n_programs_significant` ",
    "counts programs whose best partner beats what they would have found by ",
    "chance. The two link tiers are explained in Step 4."
  ),
  row.names = FALSE
)
```

## Step 4: What counts as a link

A pair is called a link in one of two tiers.

- **`primary`** — the two programs are each other's *best* partner, **and** both
  are individually significant. Program-level significance comes from a
  max-statistic permutation null: it asks whether a program's best partner is
  better than the best partner it would have found by chance, so it is not
  inflated by searching over many candidates. This is the strict, one-to-one
  reading and is what `shared = TRUE` marks.
- **`candidate`** — the pair's own `q_concordance` clears the FDR threshold but it
  is not a reciprocal best match. These are worth looking at because one program
  can genuinely correspond to two in the other trait (EBMF often splits one
  biological module into several factors), but they are not calibrated for the
  search over partners, so treat them as leads rather than conclusions.

The columns to read, and what each is telling you:

| column | what it means |
|---|---|
| `concordance_z` | the headline statistic: how far the pair's shared loading is from its own permutation null, in SD units. Sign gives the direction. This is what ranks and links. |
| `n_loci_shared` | how many *high-confidence* loci the two programs both claim. Context only — it no longer gates anything. |
| `n_eff_loci` | how many loci are effectively carrying the score. A value near 2–3 means a couple of loci are doing all the work, however large `concordance_z` looks. |
| `locus_concordance` | the raw weighted sum. Scales with both the number of co-loading loci and their magnitude, so it is not comparable across pairs on its own. |
| `cosine_loci` | the same quantity fully normalised. Deliberately *not* used to decide links: being scale-free it cannot tell a pair sharing 3 loci from one sharing 30. Useful as a "how tight is the alignment" readout. |
| `q_concordance` | BH-adjusted pair-level p. Defines the `candidate` tier; it does **not** decide `primary` links. |

```{r step4-links, results='asis'}
linked <- conc_pairs[!is.na(conc_pairs$link_tier), , drop = FALSE]
linked <- linked[order(linked$link_tier != "primary", -abs(linked$concordance_z)), ]

if (nrow(linked) > 0) {
  print(knitr::kable(
    linked |>
      dplyr::mutate(
        concordance_z = round(concordance_z, 2),
        locus_concordance = round(locus_concordance, 2),
        cosine_loci = round(cosine_loci, 3),
        n_eff_loci = round(n_eff_loci, 1),
        q_concordance = signif(q_concordance, 3)
      ) |>
      dplyr::select(
        program_id_a, program_id_b, link_tier, direction, concordance_z,
        q_concordance, n_loci_shared, n_eff_loci, locus_concordance, cosine_loci
      ),
    caption = paste0(
      "Linked program pairs, primary tier first then candidates, each ordered by ",
      "|concordance_z|. Read `n_eff_loci` alongside `concordance_z`: a large z ",
      "resting on 2-3 effective loci is a thin result."
    ),
    row.names = FALSE
  ))
} else {
  cat("No program pairs reached either link tier.\n")
}
```

## Step 5: Program families

Programs are nodes and **primary** links are edges. Because primary links are
reciprocal best matches, families are one-to-one correspondences rather than
transitive chains that collapse everything into one component. A program with no
primary link forms its own family of one.

```{r step5-families}
conc_families <- build_program_families(
  conc_result,
  program_info = program_info,
  fdr_threshold = fdr_threshold
)
multi_trait <- conc_families$families$family[conc_families$families$n_traits > 1]

if (length(multi_trait) > 0) {
  print(knitr::kable(
    conc_families$nodes |>
      dplyr::filter(family %in% multi_trait) |>
      dplyr::arrange(family, trait_id, program_id),
    caption = paste0(
      "Programs joined into multi-trait families. Only families spanning more ",
      "than one trait are shown; single-program families are omitted."
    ),
    row.names = FALSE
  ))
  print(knitr::kable(
    conc_families$families |>
      dplyr::filter(family %in% multi_trait) |>
      dplyr::mutate(
        mean_locus_concordance = round(mean_locus_concordance, 2),
        median_locus_concordance = round(median_locus_concordance, 2),
        mean_concordance_z = round(mean_concordance_z, 2)
      ) |>
      dplyr::select(
        family, n_programs, n_traits, n_edges, mean_concordance_z,
        mean_locus_concordance, n_concordant, n_antagonistic, n_shared_loci
      ),
    caption = paste0(
      "Per-family metrics. `n_concordant` / `n_antagonistic` count edges by the ",
      "sign of `concordance_z`, so a family carrying both is one whose programs ",
      "do not agree on direction."
    ),
    row.names = FALSE
  ))
} else {
  cat("No families span more than one trait.\n")
}
```

```{r step5-network, fig.height = 7, fig.width = 9, echo=FALSE}
if (!is.null(conc_families$graph)) {
  graph <- conc_families$graph
  node_traits <- conc_families$nodes$trait_id[
    match(igraph::V(graph)$name, conc_families$nodes$program_id)
  ]
  trait_levels <- unique(node_traits)
  trait_pal <- stats::setNames(
    grDevices::hcl.colors(length(trait_levels), "Dark 3"), trait_levels
  )
  edge_sign <- igraph::E(graph)$concordance_z
  igraph::plot.igraph(
    graph,
    layout = igraph::layout_with_fr(graph),
    vertex.color = unname(trait_pal[node_traits]),
    vertex.size = 13,
    vertex.frame.color = "white",
    vertex.label = igraph::V(graph)$name,
    vertex.label.cex = 0.65,
    vertex.label.color = "black",
    edge.color = ifelse(edge_sign > 0, "#d95f02", "#1b9e77"),
    edge.width = 1 + abs(edge_sign) / 2,
    main = "Multi-trait program families (orange = concordant, green = antagonistic)"
  )
  graphics::legend(
    "topleft", legend = trait_levels, fill = unname(trait_pal[trait_levels]),
    bty = "n", title = "trait"
  )
} else {
  cat("No primary links, so every program is a family of one.\n")
}
```

## Step 6: Direction of the shared effect (module rg)

`concordance_z` says whether two programs' *loading patterns* line up. It does not
say whether the alleles involved push the two traits the same way. `module_rg()`
answers that directly: across the module's loci it correlates the two traits'
**own** SNP effects, weighting each locus by how strongly the module loads on it
and by the precision of the two estimates.

- **`rg` > 0** — alleles that raise trait A also raise trait B across this module.
- **`rg` < 0** — they oppose: the module is antagonistic between the two traits.
- **`rg_ci_lower` / `rg_ci_upper`** — percentile bootstrap over loci. **If the
  interval spans 0, the direction is not established**, whatever the point
  estimate says.
- **`n_allele_mismatch`** — loci dropped because the two traits' `variant_id` did
  not agree, which would make the sign meaningless. Should be 0.

`rg_direction` and `direction` need not agree: the first is about observed allele
effects, the second about latent factor structure. Where they disagree, `rg` with
its CI is the more directly interpretable claim.

```{r step6-module-rg}
if (nrow(conc_result$module_rg) > 0) {
  knitr::kable(
    conc_result$module_rg |>
      dplyr::arrange(link_tier != "primary", dplyr::desc(abs(rg))) |>
      dplyr::mutate(
        rg = round(rg, 3),
        rg_ci_lower = round(rg_ci_lower, 3),
        rg_ci_upper = round(rg_ci_upper, 3),
        ci_excludes_zero = is.finite(rg_ci_lower) & is.finite(rg_ci_upper) &
          (rg_ci_lower > 0 | rg_ci_upper < 0)
      ),
    caption = paste0(
      "Cross-trait effect correlation across each linked module's loci. ",
      "`ci_excludes_zero` is the column to filter on before believing a direction."
    ),
    row.names = FALSE
  )
} else {
  cat("No linked pairs to estimate module rg for.\n")
}
```

## Step 7: Do linked programs recover the same biology?

**Both link tiers are checked here, not just `primary`.** A pair only needs to
have cleared *some* significance threshold in Step 4 (mutual best match for
`primary`, or its own `q_concordance <= fdr_threshold` for `candidate`) to be
included below — the `tier` column in the table says which. Candidates are
included deliberately: they are not corrected for the search over partners
the way `primary` is, so this biology check is one of the best ways to tell
whether a given candidate is a real lead or noise — held-out annotation
agreement it wasn't calibrated to produce is stronger support than the
loading evidence alone gives it.

Annotations play no part in forming a link, so agreement here is **independent
corroboration** rather than a restatement of the same evidence. Each program's
high-confidence SNPs are enriched separately, against its own trait's full
candidate universe, for trait categories and for tissues.

Read it as: if the two sides of a pair independently surface the same category or
tissue, the link has support beyond the loading geometry that created it. If they
surface nothing in common, the link rests on the loci alone.

One caveat: a program holding a large share of its trait's SNPs is being compared
against a background that is largely itself, so "none" for a dominant program is
expected rather than informative.

```{r step7-biology, results='asis'}
program_enrichment <- function(program_id, what = c("category", "tissue")) {
  what <- match.arg(what)
  tid <- as.character(program_loadings$trait_id[
    program_loadings$program_id == program_id
  ][1])
  snps <- unique(program_loadings$snp_id[
    program_loadings$program_id == program_id & program_loadings$high_confidence
  ])
  if (length(snps) < 5) {
    return("too few high-confidence SNPs")
  }
  baseline <- unique(program_loadings$snp_id[program_loadings$trait_id == tid])
  groups <- data.frame(
    snp_id = snps, group = program_id, stringsAsFactors = FALSE
  )
  res <- if (what == "category") {
    enrich_snp_group_trait_categories(
      groups = groups, coloc_groups = coloc_by_trait[[tid]],
      snp_key = "variant_id", min_group_size = 5L, baseline_snp_ids = baseline
    )
  } else {
    enrich_snp_group_tissues(
      groups = groups, coloc_groups = coloc_by_trait[[tid]],
      snp_key = "variant_id", min_group_size = 5L, background = "trait",
      baseline_snp_ids = baseline
    )
  }
  if (length(res$by_group) == 0 || nrow(res$by_group[[1]]$enriched) == 0) {
    return("none")
  }
  hits <- utils::head(res$by_group[[1]]$enriched, 4)
  label <- if (what == "category") hits$trait_category else hits$tissue
  return(paste0(label, " (FDR=", signif(hits$fdr, 2), ")", collapse = "<br>"))
}

if (nrow(linked) > 0) {
  cat(sprintf(
    "\nChecking %d linked pairs: %d primary, %d candidate.\n\n",
    nrow(linked),
    sum(linked$link_tier == "primary"),
    sum(linked$link_tier == "candidate")
  ))
  biology <- dplyr::bind_rows(lapply(seq_len(nrow(linked)), function(i) {
    pa <- linked$program_id_a[i]
    pb <- linked$program_id_b[i]
    return(data.frame(
      pair = paste(pa, pb, sep = " ~ "),
      tier = linked$link_tier[i],
      direction = linked$direction[i],
      categories_a = program_enrichment(pa, "category"),
      categories_b = program_enrichment(pb, "category"),
      tissues_a = program_enrichment(pa, "tissue"),
      tissues_b = program_enrichment(pb, "tissue"),
      stringsAsFactors = FALSE
    ))
  }))
  print(knitr::kable(
    biology,
    caption = paste0(
      "Trait-category and tissue enrichment of each side of a linked pair, ",
      "both primary and candidate tiers, computed independently per trait ",
      "against that trait's own SNP universe. Matching entries between the ",
      "`_a` and `_b` columns are held-out support for the link."
    ),
    row.names = FALSE,
    escape = FALSE
  ))
} else {
  cat("No linked pairs to annotate.\n")
}
```

## Interpretation notes

- **Direction is anchored, not absolute.** EBMF factor signs are not identifiable
  and are arbitrary across independent fits. `extract_program_loadings()` reflects
  both `F_pm` and `L_pm` so each program's target-trait loading is positive, which
  is what makes a positive `concordance_z` readable as "target-aligned".
- **Evidence scales with shared loading, not with tightness.** The statistic is
  deliberately covariance-like. A fully normalised version (`cosine_loci`) cannot
  distinguish a pair sharing 3 loci from one sharing 30, which is why it is
  reported but not used to decide links.
- **Always read `n_eff_loci`.** A large `concordance_z` resting on 2–3 effective
  loci is a thin result, and the permutation null cannot tell you that on its own.
- **Two tiers, two strengths of claim.** `primary` is calibrated for the search
  over partners and is one-to-one. `candidate` is not, and exists because EBMF
  factor splitting means a single module can appear as several programs.
- **Sample overlap is not corrected.** When the two target traits come from the
  same cohort, their effect estimates share participants, which biases `module_rg`
  away from the null. Treat rg magnitudes from same-cohort pairs as optimistic.
- **Winner's curse.** Loci enter each trait's universe only by clearing that
  trait's own significance and fine-mapping bar, so the shared set is doubly
  selected and effect sizes are inflated.
- **Restricted to valid programs.** Only programs passing the per-trait validation
  filters are compared; Step 1 shows exactly which survived.
