---
title: "Null simulation study of EBMF programs"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # Simulation parameters (K is always 0: no modules are ever planted)
  n_snps: 300
  n_traits: 600
  noise_sd: 1
  effect_size: 6
  p_active_background: 0.1
  background_sparsity_sd: 1.2
  # Similarity-graph realism. background_corr = 0 is the historical behaviour
  # (independent, random-signed background, which contributes nothing to
  # SNP-SNP similarity); snp_pleiotropy_sd is the per-SNP analogue of
  # background_sparsity_sd. See simulated_graph_diagnostics().
  background_corr: 0
  snp_pleiotropy_sd: 0
  background_effect_scale: 0.5
  target_pattern: module

  # EBMF / clustering parameters (inputs to run_univariate_clustering)
  trait_subset: "phenotypic"
  min_snp_signals: 2
  compress_method: asinh
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_lfsr_threshold: 0.05
  ebmf_magnitude_threshold: 0.3
  ebmf_drop_global: true
  ebmf_prior: point_normal
  ebmf_backfit: true

  # Program validation parameters (inputs to summarise_ebmf_programs)
  min_module_size: 3
  min_mean_internal: 0.3
  min_connectedness: 0.25
  # Loading-permutation coherence gate (replaces min_mean_internal /
  # min_connectedness as the gate; those two are now reported only).
  coherence_n_perm: 2000
  coherence_q: 0.05
  ebmf_greedy_Kmax: 100
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_stability_rep: 5
  frac_traits: 0.8
  stability_top_n: 20
  stability_threshold: 0.5
  posterior_stat: median
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5

  # Replication settings. n_sim is the Monte Carlo sample size (ADEMP);
  # override it at render time for fast test runs, e.g.
  #   rmarkdown::render(..., params = list(n_sim = 10))
  n_sim: 200
  master_seed: 1
  # Persist the ADEMP estimates/performance datasets when TRUE; output_dir
  # defaults to tempdir() when empty.
  write_outputs: false
  output_dir: ""
  # quick: fast test mode. Keeps the full null-version grid but caps n_sim at
  # 3, uses n_null = 1, and disables the stability subsampling. Use for smoke
  # tests only; the reported Monte Carlo SEs will be large.
  quick: true
vignette: >
  %\VignetteIndexEntry{Null simulation study of EBMF programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

n_sim <- as.integer(params$n_sim)
quick <- isTRUE(params$quick)
if (quick) {
  n_sim <- min(n_sim, 3L)
}
n_null <- if (quick) 1L else as.integer(params$n_null)
n_stability_rep <- if (quick) 0L else as.integer(params$n_stability_rep)
# A null fit with 600 traits is dominated by EBMF; shrink the row count in quick
# mode so smoke tests stay fast. This is a size change, not a design change.
n_traits <- if (quick) 100L else as.integer(params$n_traits)
set.seed(params$master_seed)
# One seed per replicate, reused across every null version (common random
# numbers), and passed explicitly so parallel workers never touch the RNG state.
sim_seeds <- sample.int(.Machine$integer.max, n_sim)

clustering_settings <- list(
  min_snp_signals = params$min_snp_signals,
  trait_subset = params$trait_subset,
  compress_method = params$compress_method,
  compress_scale = params$compress_scale,
  ebmf_lfsr_threshold = params$ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = params$ebmf_magnitude_threshold,
  ebmf_greedy_Kmax = params$ebmf_greedy_Kmax,
  ebmf_drop_global = params$ebmf_drop_global,
  ebmf_prior = params$ebmf_prior,
  ebmf_backfit = params$ebmf_backfit,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness
)

validation_settings <- list(
  edge_threshold = params$similarity_threshold,
  min_module_size = params$min_module_size,
  min_mean_internal = params$min_mean_internal,
  min_connectedness = params$min_connectedness,
  coherence_n_perm = params$coherence_n_perm,
  coherence_q_threshold = params$coherence_q,
  n_null = n_null,
  alpha_membership = params$alpha_membership,
  n_candidate_tier = params$n_candidate_tier,
  n_rep = n_stability_rep,
  frac_traits = params$frac_traits,
  top_n = params$stability_top_n,
  stability_threshold = params$stability_threshold,
  posterior_stat = params$posterior_stat,
  posterior_evidence_cap = params$posterior_evidence_cap,
  redundancy_threshold = params$redundancy_threshold,
  cores = params$cores,
  verbose = FALSE
)
```

This vignette evaluates empirical Bayes matrix factorisation (EBMF) programs
against **pure-null** synthetic data: `K = 0`, so no modules, driver SNPs, or
driver traits are ever planted. Each simulation is generated by
`simulate_trait()` with only the target-trait row plus unstructured background
traits. The pipeline — `run_univariate_clustering()` followed by
`summarise_ebmf_programs()` — is byte-for-byte the same battery used in the
module simulations
(`investigation-univariate-simulations.html`); the analysis settings are also
identical. The only difference is that EBMF is now trying to discover programs
in data where the only structure is randomness.

The four validation checks are exactly the same gates used on the real pipeline:

- `size` — at least `min_module_size` filtered SNPs (`n_snps_filtered`).
- `internal_sim` — mean internal SNP similarity `>= min_mean_internal`.
- `connectivity` — pair connectedness `>= min_connectedness`.
- `stability` — top-loading SNPs recovered with replication
  `>= stability_threshold` under trait subsampling.

A program is `valid` only when every check passes. Under a true null, **any**
program that passes all four checks is a false discovery, so the per-check pass
rates below are false-discovery rates: they show how often each gate (and the
pipeline as a whole) is fooled by random background. `background_absorbed` is
the SNP-level analogue — the fraction of background SNPs pulled into valid
programs.

Because `K = 0`, the module-only generative parameters (`module_sizes`,
`overlap`, `n_traits_per_module`, `trait_overlap`, `driver_sharing`,
`p_structural_zero`, `p_spurious`, `effect_tail`) are inert and are not exposed
here. The active axes of background realism are `p_active_background`
(how often a background trait is observed at a SNP), `background_sparsity_sd`
(heavy-tailed per-trait density variation), `background_effect_scale`
(background magnitude relative to `effect_size`), and `noise_sd` (per-cell
noise). Each version below holds everything else at its realistic baseline and
varies one axis.

## Does this simulation exercise the gates in the regime real data uses them?

The validation gates act on the SNP similarity graph, so what matters is not
whether the simulated matrix has realistic per-trait sparsity and magnitude
marginals -- it is whether the *graph* those marginals induce resembles a real
one. Those are not the same thing, and matching the marginals is not sufficient.

`simulated_graph_diagnostics()` measures the simulated graph and
`real_bmi_graph_targets()` carries the values measured on the BMI pleiotropy
matrix. The row to read first is `random_set_pass_rate`: the fraction of random
SNP sets whose mean internal similarity clears 0.3. On real BMI data that is
about 0.94 -- a random set of SNPs passes the old absolute gate almost always,
which is why that gate carried no specificity. Where the simulated value is near
0, a random set can never pass, so a null simulation study **cannot detect an
uninformative gate** and its false-discovery rate is close to un-failable by
construction.

The two generative parameters that move this are `background_corr` (background
traits sharing structure, so they contribute to SNP-SNP similarity at all --
at `0` they are independent and random-signed and contribute nothing) and
`snp_pleiotropy_sd` (per-SNP heterogeneity in how many traits a SNP is observed
on, which is what brings the median pairwise overlap down towards the real ~3
and produces the sparse SNPs that carry high cosines).

```{r graph-realism}
graph_diag <- tryCatch({
  sim_probe <- simulate_trait(
    n_coloc_groups = params$n_snps,
    K = 0L,
    n_traits = n_traits,
    noise_sd = params$noise_sd,
    effect_size = params$effect_size,
    p_active_background = params$p_active_background,
    background_sparsity_sd = params$background_sparsity_sd,
    background_corr = params$background_corr,
    snp_pleiotropy_sd = params$snp_pleiotropy_sd,
    background_effect_scale = params$background_effect_scale,
    target_pattern = params$target_pattern,
    seed = params$master_seed
  )
  probe_res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim_probe$trait_object), clustering_settings)
  )
  simulated_graph_diagnostics(
    probe_res$x_matrix,
    target_trait_id = probe_res$parameters$target_trait_id
  )
}, error = function(e) NULL)

if (is.null(graph_diag)) {
  cat("Graph diagnostics unavailable for this configuration.\n\n")
} else {
  targets <- real_bmi_graph_targets()
  shared_cols <- intersect(names(targets), names(graph_diag))
  comparison <- data.frame(
    statistic = shared_cols,
    simulated = signif(as.numeric(graph_diag[shared_cols]), 3),
    real_bmi = signif(as.numeric(targets[shared_cols]), 3),
    stringsAsFactors = FALSE
  )
  print(knitr::kable(
    comparison,
    row.names = FALSE,
    caption = paste0(
      "Similarity-graph diagnostics for this run's generative settings against ",
      "the values measured on the real BMI pleiotropy matrix. Read ",
      "random_set_pass_rate first: where it is near 0 the gates are not being ",
      "tested in the regime real data puts them in, whatever the recovery ",
      "metrics below say."
    )
  ))
}
```

## Simulation design (ADEMP)

This null study follows the ADEMP structure of Morris, White and Crowther
(2019, *Statistics in Medicine* 38:2074-2102, \doi{10.1002/sim.8086}); it is
the testing-target companion to `investigation-univariate-simulations.html`.

**Aims.** To estimate the **type I error / false-discovery rate** of the EBMF
program pipeline under a true null, and how that rate depends on background
realism. This is a *stretch-or-break* aim: find where the gates are fooled.

**Data-generating mechanism.** `simulate_trait()` with `K = 0`, so only the
oriented target row and unstructured background rows exist. The varied factors
are `p_active_background`, `background_sparsity_sd`, `background_effect_scale`,
and `noise_sd`, each varied one at a time from the realistic baseline.

**Estimand / target.** The target is the null hypothesis of no planted module;
under it, any program passing all four gates is a false discovery, so the
`valid` rate is the pipeline-level false-discovery rate.

**Methods.** Byte-for-byte the same pipeline and thresholds as the module
study: `run_univariate_clustering()` then `summarise_ebmf_programs()` with the
four gates (size, internal coherence, connectedness, stability).

**Performance measures.**

- `n programs`: EBMF programs surviving the lFSR/magnitude membership gate.
- `valid programs`: programs passing all four gates.
- `valid rate = valid programs / n programs` (program-level false-discovery rate).
- `background_absorbed = |{background SNPs in valid programs}| / n_background` (SNP-level false-discovery rate).
- Per-gate pass rates: `size`, `internal_sim`, `connectedness`, `stability`.

**Monte Carlo uncertainty.** Each rate is the mean over `n_sim` replicates
(default `r n_sim`; override with `params = list(n_sim = ...)` for fast test
runs), with the replicate as the Monte Carlo unit. Proportions carry MC SE
`sqrt(p(1-p)/n_sim)` and a 95% Monte Carlo interval; continuous measures carry
`sd / sqrt(n_sim)`. Common random numbers are used across null versions, and
the per-replicate seed is recorded in the estimates dataset.

```{r settings-table}
settings_to_show <- c(
  "trait_subset", "min_snp_signals", "min_module_size",
  "compress_method", "compress_scale", "similarity_threshold",
  "ebmf_lfsr_threshold",
  "ebmf_magnitude_threshold", "ebmf_drop_global", "ebmf_prior",
  "ebmf_backfit",
  "min_mean_internal", "min_connectedness",
  "n_null", "alpha_membership", "n_candidate_tier", "n_stability_rep",
  "frac_traits", "stability_top_n", "stability_threshold",
  "posterior_stat", "posterior_evidence_cap", "redundancy_threshold", "cores",
  "n_sim"
)
settings_values <- c(clustering_settings, validation_settings)
settings_values$similarity_threshold <- params$similarity_threshold
settings_values$n_stability_rep <- n_stability_rep
settings_values$stability_top_n <- validation_settings$top_n
settings_values$n_sim <- n_sim
knitr::kable(
  data.frame(
    parameter = settings_to_show,
    value = vapply(settings_to_show, function(nm) {
      as.character(settings_values[[nm]])
    }, character(1))
  ),
  caption = "Fixed EBMF analysis and validation settings (identical across all null versions)"
)
```

```{r tier-helpers}
null_base <- list(
  n_coloc_groups = params$n_snps,
  K = 0L,
  n_traits = n_traits,
  noise_sd = params$noise_sd,
  effect_size = params$effect_size,
  p_active_background = params$p_active_background,
  background_sparsity_sd = params$background_sparsity_sd,
  background_corr = params$background_corr,
  snp_pleiotropy_sd = params$snp_pleiotropy_sd,
  background_effect_scale = params$background_effect_scale,
  target_pattern = params$target_pattern
)

# K is pinned to 0 no matter what: these are null simulations by construction.
make_tier_spec <- function(...) {
  overrides <- list(...)
  spec <- utils::modifyList(null_base, overrides)
  spec$K <- 0L
  spec$module_sizes <- NULL
  spec$n_coloc_groups <- params$n_snps
  return(spec)
}

run_tier_one <- function(spec, rep) {
  dots <- spec
  # Common random numbers: the seed depends only on the replicate index, so
  # every null version sees the same scenario in replicate `rep`.
  dots$seed <- sim_seeds[[rep]]
  sim <- do.call(simulate_trait, dots)
  res <- do.call(
    run_univariate_clustering,
    c(list(trait_object = sim$trait_object), clustering_settings)
  )
  # Vary the validation layer's internal randomisation by replicate.
  vs <- validation_settings
  vs$seed <- sim_seeds[[rep]]
  smry <- do.call(
    summarise_ebmf_programs,
    c(list(clustering_result = res, s_matrix = res$s_matrix), vs)
  )
  valid_ids <- smry$programs$program[smry$programs$status == "valid"]
  valid_assigned <- smry$assigned |>
    dplyr::filter(program %in% valid_ids)
  ev <- evaluate_univariate_simulation(
    sim,
    predicted_memberships = valid_assigned
  )
  programs <- smry$programs
  programs$rep <- rep(rep, nrow(programs))
  list(
    summary = data.frame(
      sim_seed = dots$seed,
      n_programs = nrow(programs),
      n_valid_programs = length(valid_ids),
      background_absorbed = ev$background_absorbed,
      stringsAsFactors = FALSE
    ),
    programs = programs
  )
}

run_tier <- function(name, label, spec, n_sim) {
  n_jobs <- 40L
  out <- if (n_sim > 1 && .Platform$OS.type != "windows") {
    parallel::mclapply(
      seq_len(n_sim),
      function(i) run_tier_one(spec, i),
      mc.cores = min(n_jobs, n_sim),
      mc.set.seed = FALSE
    )
  } else {
    lapply(seq_len(n_sim), function(i) run_tier_one(spec, i))
  }
  list(
    summary = dplyr::bind_rows(lapply(out, `[[`, "summary")) |>
      dplyr::mutate(tier = label),
    programs = dplyr::bind_rows(lapply(out, `[[`, "programs")) |>
      dplyr::mutate(tier = label, tier_name = name)
  )
}
```

## Null versions

Every version plants no modules (`K = 0`); the versions differ only in the
background realism parameter being stressed, with everything else at its
realistic baseline.

```{r tiers}
tier_specs <- list(
  baseline = make_tier_spec(),
  background_absent = make_tier_spec(p_active_background = 0),
  background_sparse = make_tier_spec(p_active_background = 0.02),
  background_dense = make_tier_spec(p_active_background = 0.3),
  sparsity_uniform = make_tier_spec(background_sparsity_sd = 0),
  sparsity_heavy = make_tier_spec(background_sparsity_sd = 1.8),
  weak_background_effects = make_tier_spec(background_effect_scale = 0.25),
  strong_background_effects = make_tier_spec(background_effect_scale = 1.5),
  low_noise = make_tier_spec(noise_sd = 0.25),
  high_noise = make_tier_spec(noise_sd = 4)
)

tier_labels <- c(
  baseline = "baseline (default background)",
  background_absent = "no background associations",
  background_sparse = "sparse background (p_active_background = 0.02)",
  background_dense = "dense background (p_active_background = 0.3)",
  sparsity_uniform = "uniform background density (background_sparsity_sd = 0)",
  sparsity_heavy = "heavy-tailed background density (background_sparsity_sd = 1.8)",
  weak_background_effects = "weak background magnitudes (background_effect_scale = 0.25)",
  strong_background_effects = "strong background magnitudes (background_effect_scale = 1.5)",
  low_noise = "low noise (noise_sd = 0.25)",
  high_noise = "high noise (noise_sd = 4)"
)

# Quick test mode runs a representative subset of the null versions; the full
# set is used when quick = FALSE.
if (quick) {
  keep <- intersect(
    c(
      "baseline", "background_dense", "strong_background_effects",
      "high_noise"
    ),
    names(tier_specs)
  )
  tier_specs <- tier_specs[keep]
  tier_labels <- tier_labels[keep]
}

tier_results <- lapply(names(tier_specs), function(nm) {
  run_tier(nm, tier_labels[[nm]], tier_specs[[nm]], n_sim)
})
names(tier_results) <- names(tier_specs)

tier_summary <- dplyr::bind_rows(lapply(tier_results, `[[`, "summary"))
program_checks <- dplyr::bind_rows(lapply(tier_results, `[[`, "programs"))
```

```{r tier-params-table}
tier_param_table <- do.call(rbind, lapply(names(tier_specs), function(nm) {
  s <- tier_specs[[nm]]
  data.frame(
    version = tier_labels[[nm]],
    K = 0L,
    n_snps = s$n_coloc_groups,
    n_traits = s$n_traits,
    noise_sd = s$noise_sd,
    effect_size = s$effect_size,
    p_active_background = s$p_active_background,
    background_sparsity_sd = s$background_sparsity_sd,
    background_effect_scale = s$background_effect_scale,
    target_pattern = s$target_pattern,
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  tier_param_table,
  row.names = FALSE,
  caption = "Generative parameters for each null version (all versions plant no modules, K = 0)"
)
```

## Aggregate results

Across all replicates and programs of each version, the tables below report
the number of programs EBMF produced before validation (`n programs`), how many
survived every check (`valid programs`), the false-discovery rates
(`valid rate` at program level, `background absorbed` at SNP level), and the
per-gate pass rates. Estimates are averages over `r n_sim` replicates and carry
a **Monte Carlo standard error** in parentheses (ADEMP; Morris, White &
Crowther 2019): the replicate is the Monte Carlo unit, so every number is an
estimate with quantified uncertainty.

```{r aggregate}
# Per-replicate rates: the replicate is the Monte Carlo unit for both the
# false-discovery rate and the individual gates.
rep_check_rates <- program_checks |>
  dplyr::group_by(tier, rep) |>
  dplyr::summarise(
    n_programs = dplyr::n(),
    rate_size = mean(size_pass),
    rate_internal_similarity = mean(internal_similarity_pass),
    rate_connectedness = mean(connectedness_pass),
    rate_stability = if (any(stability_checked)) {
      mean(stability_pass[stability_checked])
    } else {
      NA_real_
    },
    rate_valid = mean(status == "valid"),
    .groups = "drop"
  )

count_cols <- c("n_programs", "n_valid_programs", "background_absorbed")
count_mc <- do.call(rbind, lapply(count_cols, function(cl) {
  out <- ademp_summarise(tier_summary, cl, group_cols = "tier", type = "mean")
  out$measure <- cl
  out
}))
count_mc$measure <- factor(
  count_mc$measure,
  levels = count_cols,
  labels = c("n programs", "valid programs", "background absorbed")
)

check_cols <- c(
  "rate_size", "rate_internal_similarity", "rate_connectedness",
  "rate_stability", "rate_valid"
)
check_mc <- do.call(rbind, lapply(check_cols, function(cl) {
  out <- ademp_summarise(
    rep_check_rates, cl, group_cols = "tier", type = "prop"
  )
  out$measure <- cl
  out
}))
check_mc$measure <- factor(
  check_mc$measure,
  levels = check_cols,
  labels = c("size", "internal sim", "connectivity", "stability", "valid")
)

format_est_mcse <- function(estimate, mcse, scale = 1, digits = 2) {
  est <- estimate * scale
  se <- mcse * scale
  ifelse(
    is.na(est), "-",
    ifelse(
      is.na(se),
      formatC(est, digits = digits, format = "f"),
      paste0(
        formatC(est, digits = digits, format = "f"), " (",
        formatC(se, digits = digits, format = "f"), ")"
      )
    )
  )
}

count_table <- count_mc |>
  dplyr::mutate(cell = format_est_mcse(estimate, mcse)) |>
  dplyr::select(tier, measure, cell) |>
  tidyr::pivot_wider(names_from = measure, values_from = cell)

check_mc_table <- check_mc |>
  dplyr::mutate(cell = format_est_mcse(estimate, mcse, scale = 100)) |>
  dplyr::select(tier, measure, cell) |>
  tidyr::pivot_wider(names_from = measure, values_from = cell)

knitr::kable(
  count_table,
  caption = paste0(
    "Null-version false-discovery output (mean across ", n_sim,
    " replicates; Monte Carlo SE in parentheses). Under a true null, valid ",
    "programs are false discoveries and background absorbed is the SNP-level ",
    "false-discovery rate."
  )
)

knitr::kable(
  check_mc_table,
  caption = paste0(
    "Per-gate pass rates by null version (percent; per-replicate rate averaged ",
    "across ", n_sim, " replicates, Monte Carlo SE in parentheses). The ",
    "stability rate is over programs re-checked after the size and ",
    "internal-coherence gates."
  )
)
```

The `stability` pass rate is computed only over programs that were actually
re-checked (those passing the size and internal-coherence checks).

```{r pass-rate-plot, fig.height=8, fig.width=9}
gate_plot_data <- check_mc |>
  dplyr::filter(measure != "size")
ggplot2::ggplot(
  gate_plot_data,
  ggplot2::aes(x = 100 * estimate, y = tier)
) +
  ggplot2::geom_errorbarh(
    ggplot2::aes(xmin = 100 * conf_low, xmax = 100 * conf_high), height = 0.25
  ) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_wrap(~measure, scales = "free_x", ncol = 2) +
  ggplot2::labs(
    x = "pass rate (%, 95% Monte Carlo CI)", y = NULL,
    title = "Validation-gate pass rates under the null"
  ) +
  ggplot2::theme_minimal(base_size = 9)
```

## Program-level validation checks

For each version, every program found across the replicates is shown with the
four validation checks as separate columns. A check is marked `FAIL` when the
program does not meet that criterion; `n/a` under `stability` means the
program already failed the internal-coherence checks, so stability was not
re-tested.

```{r check-helper, include=FALSE}
check_table <- function(which_tier) {
  df <- program_checks |>
    dplyr::filter(.data$tier_name == which_tier)
  if (nrow(df) == 0) {
    return(NULL)
  }
  df |>
    dplyr::mutate(
      check_size = ifelse(size_pass, "pass", "FAIL"),
      check_internal_similarity = ifelse(internal_similarity_pass, "pass", "FAIL"),
      check_connectedness = ifelse(connectedness_pass, "pass", "FAIL"),
      check_stability = dplyr::case_when(
        !stability_checked ~ "n/a",
        stability_pass ~ "pass",
        TRUE ~ "FAIL"
      ),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      connectedness = round(connectedness, 3),
      replication = round(replication, 3)
    ) |>
    dplyr::select(
      rep, program, n_snps, n_snps_filtered, mean_internal_similarity,
      connectedness, replication,
      check_size, check_internal_similarity, check_connectedness,
      check_stability, status
    ) |>
    dplyr::arrange(rep, program) |>
    knitr::kable(
      row.names = FALSE,
      col.names = c(
        "rep", "program", "n_snps", "n_snps filtered",
        "mean internal similarity", "connectedness", "replication",
        "size", "internal sim", "connectivity", "stability", "status"
      )
    )
}
```

```{r tier-checks, results='asis'}
for (nm in names(tier_specs)) {
  cat("\n###", tier_labels[[nm]], "\n\n")
  tbl <- check_table(nm)
  if (is.null(tbl)) {
    cat("No EBMF programs were produced for this version.\n\n")
  } else {
    cat(tbl, sep = "\n")
    cat("\n\n")
  }
}
```

## Notes for interpreting runs

- **Fast test runs.** A production render is long: `n_sim = 200` over the 10
  null versions is thousands of EBMF fits, and a single null fit is expensive
  because there are 600 trait rows. To iterate quickly, override `params`:
  - `params = list(quick = TRUE)` — smoke test (4 representative versions,
    `n_sim = 3`, `n_null = 1`, no stability, `n_traits = 100`); about 90 s.
  - `params = list(n_sim = 5, n_null = 1, n_stability_rep = 0, n_traits = 100)`
    — full null-version grid with cheap internals and a smaller row count;
    about 1-2 min.
  - `params = list(n_sim = 20)` — full-fidelity settings at reduced repetitions
    (still slow, because of the 600 traits).
  Lowering `n_null`, `n_stability_rep` and `n_traits` gives the largest
  speed-up.
- Under a true null (`K = 0`) any program passing all four
  validation checks is a **false discovery**; the `valid` column is the
  pipeline-level false-positive rate.
- `n programs` is the number of EBMF programs surviving the lFSR/magnitude
  membership gate, before any validation check is applied. `valid programs` is
  how many of those pass every check.
- `background_absorbed` is the fraction of background SNPs placed into valid
  programs. Because there is no planted structure, every background SNP is
  genuinely background, so this is the SNP-level false-discovery rate.
- `p_structural_zero`, `p_spurious`, `effect_tail`, and the module/driver
  parameters are inert when `K = 0` (they only act on driver rows), so they
  are not exposed here.
- With `p_active_background = 0` every background trait is dropped by the
  `min_snp_signals` filter and the EBMF input reduces to the oriented target
  row; any factor is then the global one and is dropped by `ebmf_drop_global`.
- The four checks use the same thresholds as the module simulations
  (`investigation-univariate-simulations.html`), so pass rates here are directly
  comparable to how well the gates separate real modules from background there.
- Every aggregate number is a Monte Carlo estimate over `n_sim` replicates and
  is reported with its Monte Carlo SE (ADEMP). Reduce `n_sim` only for test
  runs; the MC SE in the tables will show the resulting loss of precision.
- Replicates use common random numbers: the per-replicate seed (`sim_seeds`) is
  shared across null versions and recorded in `tier_summary$sim_seed`, so the
  study is reproducible and version-to-version differences are more precise
  than independent seeds would give.

```{r write-outputs, include=FALSE}
if (isTRUE(params$write_outputs)) {
  out_dir <- if (nzchar(params$output_dir)) params$output_dir else tempdir()
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  utils::write.csv(
    tier_summary, file.path(out_dir, "null-estimates.csv"), row.names = FALSE
  )
  utils::write.csv(
    program_checks, file.path(out_dir, "null-programs.csv"), row.names = FALSE
  )
  utils::write.csv(
    count_mc, file.path(out_dir, "null-performance-counts.csv"),
    row.names = FALSE
  )
  utils::write.csv(
    check_mc, file.path(out_dir, "null-performance-gates.csv"),
    row.names = FALSE
  )
  message("Wrote simulation outputs to ", out_dir)
}
```