---
title: "Investigation of multi-trait EBMF program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of multi-trait EBMF program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  include_trans: true
  trait_ids: ["c6d2e20f-6ece-d093-78ee-37534547c466", "1631", "1630", "1640"]
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  compress_method: "asinh"
  compress_scale: 5
  similarity_threshold: 0.2
  ebmf_magnitude_threshold: 0.5
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  min_mean_internal: 0.3
  min_connectedness: 0.5
  n_null: 5
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 2
  frac_traits: 0.8
  stability_top_n: 20
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  n_perm: 3000
  fdr_threshold: 0.05
  phi_high: 0.85
  phi_low: 0.70
  min_alignment: 0.5
  min_eff_loci: 5
  min_shared_loci: 3
  mutual_best: false
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r trait-metadata, include=FALSE}
trait_ids <- as.character(params$trait_ids)
if (length(trait_ids) < 2) {
  stop("At least two trait_ids are required for a multi-trait comparison")
}
select_api(params$api)
# One call per trait, not the plural traits(): a GWAS upload's own numeric
# database id (trait()$trait$id) is not its GUID, so a lookup keyed by `id`
# silently drops upload rows. `trait_name` is set for both a public trait and
# an upload (see .decorate_gwas_upload()), so it is read directly against the
# id each trait is actually looked up by everywhere else in this vignette.
trait_name_map <- stats::setNames(
  vapply(trait_ids, function(tid) {
    nm <- trait(tid)$trait$trait_name
    return(if (is.null(nm) || is.na(nm) || !nzchar(nm)) tid else nm)
  }, character(1)),
  trait_ids
)
```

This vignette extends the single-trait investigation
(`investigation-univariate-analysis.html`) to **multiple traits**, and asks one
question: **which programs are shared between traits, which SNPs carry them,
and do those SNPs push the traits the same way or opposite ways?**

The default set runs this on real data with a strong, independently
published prior to check against: the T2D GWAS upload from
[Suzuki T2D](investigation-suzuki-t2d.html) (GUID
`c6d2e20f-6ece-d093-78ee-37534547c466`), whose high/medium programs turned out
unusually clean, alongside three more specific metabolic traits rather than
BMI -- whose own background is dominated by dozens of body-size traits that
move with it almost everywhere (see
[Investigation of univariate analysis](investigation-univariate-analysis.html))
and so needs more care than a first cross-trait check calls for. Every
cross-trait pair of programs is scored, from however many traits are
supplied: triglyceride levels (1631), HDL cholesterol (1630) and HbA1c (1640),
each paired with T2D and with each other. Suzuki et al.'s own cluster profiles
give concrete expectations to check the links against
([Suzuki T2D, Part 5](investigation-suzuki-t2d.html)):

- **Triglycerides** should link to T2D's Lipodystrophy and Metabolic syndrome
  programs (P2, P6, P10, P24, P31, P35, P47 in the T2D-only fit) and be mostly
  **concordant**, but with a possible **antagonistic** TM6SF2-type liver
  locus -- the one place a single partner trait is expected to show both
  directions at once.
- **HDL cholesterol** should link to the same Lipodystrophy/Metabolic
  syndrome programs, **antagonistically** (the T2D risk allele lowers HDL).
- **HbA1c** should link to T2D's beta-cell and Residual glycaemic signal
  instead -- a different part of the partition from the lipid traits, and the
  only one of the three expected to touch T2D's glycaemic core directly.

EBMF programs are discovered independently for each target trait, so each
trait's architecture is described in its own terms first. Every cross-trait
pair of programs is then asked two questions. **Same biology?** -- are the two
programs driven by the same background traits in the same proportions
(profile congruence of their trait loadings, `L_pm`)? **Same SNPs?** -- where
the two traits carry the same fine-mapped loci, do the programs load on them
alike (alignment of their SNP loadings, `F_pm`)? The answers are combined into
high / medium / low tiers; high- and medium-tier pairs are **links**, and links
are grouped into **multi-trait families**, each with the set of loci its
programs claim. The **direction** of a link -- whether the alleles involved
raise both traits or raise one and lower the other -- comes from the sign of
the profile congruence and, where enough loci are shared, from `module_rg()`,
the correlation of the traits' own allele effects.

Traits are supplied as an array in `params$trait_ids`. We use
`r paste0("**", unname(trait_name_map[trait_ids]), "** (trait id ", trait_ids, ")", collapse = ", ")`.

```{r params}
api <- params$api
# Resolved from vignettes/_ebmf_settings.yml, the same file the univariate
# study reads. Previously this vignette carried its own ebmf_magnitude_threshold
# (0.5 vs 0.75) and include_trans (FALSE vs TRUE), so it factorised a different
# matrix and its "programs" were not the univariate programs at all -- which is
# why the two studies did not line up.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

trait_subset <- ebmf_settings$trait_subset
min_module_size <- as.integer(ebmf_settings$min_module_size)
min_snp_signals <- as.integer(ebmf_settings$min_snp_signals)
compress_method <- ebmf_settings$compress_method
compress_scale <- as.numeric(ebmf_settings$compress_scale)
include_trans <- isTRUE(ebmf_settings$include_trans)
similarity_threshold <- as.numeric(ebmf_settings$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(ebmf_settings$ebmf_lfsr_threshold)
ebmf_prior <- ebmf_settings$ebmf_prior
ebmf_backfit <- isTRUE(ebmf_settings$ebmf_backfit)
ebmf_magnitude_threshold <- as.numeric(ebmf_settings$ebmf_magnitude_threshold)
ebmf_greedy_Kmax <- as.integer(ebmf_settings$ebmf_greedy_Kmax)
similarity_n_perm <- as.integer(ebmf_settings$similarity_n_perm)
similarity_q_threshold <- as.numeric(ebmf_settings$similarity_q)
high_similarity_q <- as.numeric(ebmf_settings$high_similarity_q)
high_connectedness <- as.numeric(ebmf_settings$high_connectedness)
min_effective_traits <- as.numeric(ebmf_settings$min_effective_traits)
associations <- params$associations
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)
n_perm <- as.integer(params$n_perm)
fdr_threshold <- as.numeric(params$fdr_threshold)
phi_high <- as.numeric(params$phi_high)
phi_low <- as.numeric(params$phi_low)
min_alignment <- as.numeric(params$min_alignment)
min_eff_loci <- as.numeric(params$min_eff_loci)
min_shared_loci_link <- as.integer(params$min_shared_loci)
mutual_best <- isTRUE(params$mutual_best)

knitr::kable(
  data.frame(
    api = api,
    trait_ids = paste(trait_ids, collapse = ", "),
    trait_subset = trait_subset,
    min_snp_signals = min_snp_signals,
    min_module_size = min_module_size,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    similarity_threshold = similarity_threshold,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness,
    n_null = n_null,
    n_rep = n_rep,
    n_perm = n_perm,
    fdr_threshold = fdr_threshold,
    phi_high = phi_high,
    phi_low = phi_low,
    min_alignment = min_alignment,
    min_eff_loci = min_eff_loci,
    min_shared_loci = min_shared_loci_link,
    mutual_best = mutual_best,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)

knitr::kable(
  data.frame(
    trait_id = trait_ids,
    trait_name = unname(trait_name_map[trait_ids]),
    stringsAsFactors = FALSE
  ),
  caption = "Target traits compared",
  row.names = FALSE
)
```

## Step 1: Univariate EBMF per trait

The existing single-trait pipeline is run independently for each target trait:
`run_univariate_clustering()` (EBMF) followed by `summarise_ebmf_programs()`.
Programs of high or medium confidence are kept (not low or single_trait).
Each program's posterior-mean SNP loadings (`F_pm`), lFSR, and high-confidence
SNP membership are then extracted, with loadings sign-oriented to the target
trait (see `extract_program_loadings()`).

Each anchor trait has its own list of **duplicate traits**, removed before
fitting: other GWAS of the same measurement (and, for T2D, diagnosis,
treatment, family-history and complication traits that repeat the T2D row
at every SNP, the same list as in
[Suzuki T2D](univariate-analysis-suzuki-t2d.html)). A duplicate of the anchor
is positive at almost every one of its SNPs after orientation, so it adds the
same component to every SNP's profile and hides the programs that actually
differ between SNPs (see [BMI](univariate-analysis-bmi.html) for what that
costs).

The same lists are passed to the cross-trait comparison as
`exclude_features`: a trait listed for either anchor is left out of every
profile comparison involving that anchor, including the partner's fit, where
it is an ordinary background trait. Without that, a TG duplicate sitting in
the HDL fit could carry a TG ~ HDL profile match almost on its own.

Only measurements of the anchor itself are listed. Related but different
measurements are kept on purpose: NMR lipoprotein-subclass traits, apolipoprotein
A1, glucose traits, and haemoglobin / red-cell traits (which run opposite to
HbA1c and are what separates a red-cell route to HbA1c from a glycaemic one).
The table after the fit shows, for each listed trait, how closely it tracks
its anchor at the anchor's own loci.

```{r proxy-traits}
# Duplicate traits per anchor, keyed by the trait_id used below. The T2D list
# is the same as univariate-analysis-suzuki-t2d.Rmd's; the TG, HDL and HbA1c
# lists are other GWAS of the same measurement (see the evidence table after
# Step 1's fit).
t2d_id <- "c6d2e20f-6ece-d093-78ee-37534547c466"
duplicate_traits <- data.frame(
  anchor_id = c(rep(t2d_id, 16), rep("1631", 3), rep("1630", 2), rep("1640", 2)),
  trait_id = c(
    "114", "2499", "2549",
    "2863", "3106", "4536",
    "1494", "3270", "4422", "3333", "4300", "4403",
    "3428", "3779", "3580",
    "2814",
    "2617", "1490", "2236",
    "2612", "1485",
    "2611", "915"
  ),
  trait_name = c(
    "Type 2 diabetes",
    "Type 2 diabetes (Firth correction)",
    "Type 2 diabetes (SPA correction)",
    "Diabetes diagnosed by doctor",
    "Non-cancer illness code, self-reported: diabetes",
    "Diagnoses - secondary ICD10: E11.9 Without complications",
    "Medication use (drugs used in diabetes)",
    "Treatment/medication code: metformin",
    "Treatment/medication code: gliclazide",
    "Treatment/medication code: insulin product",
    "Medication for cholesterol, blood pressure or diabetes: Insulin",
    "Started insulin within one year diagnosis of diabetes",
    "Illnesses of mother: Diabetes",
    "Illnesses of father: Diabetes",
    "Illnesses of siblings: Diabetes",
    "Eye problems/disorders: Diabetes related eye disease",
    "Triglyceride levels (UKB data field 30870)",
    "Triglycerides",
    "Total triglycerides levels",
    "High density lipoprotein cholesterol levels (UKB data field 30760)",
    "HDL cholesterol",
    "Glycated haemoglobin HbA1c levels (UKB data field 30750)",
    "Glycated haemoglobin levels"
  ),
  group = c(
    rep("T2D GWAS", 3), rep("Diagnosis", 3), rep("Treatment", 6),
    rep("Family history", 3), "Complication",
    rep("Other GWAS of the anchor", 7)
  ),
  stringsAsFactors = FALSE
)
proxy_trait_ids <- split(duplicate_traits$trait_id, duplicate_traits$anchor_id)

knitr::kable(
  duplicate_traits |>
    dplyr::mutate(anchor = unname(trait_name_map[anchor_id])) |>
    dplyr::select(anchor, trait_id, trait_name, group),
  caption = paste0(
    "Duplicate traits removed from each anchor's own matrix before fitting, ",
    "and left out of every profile comparison involving that anchor."
  ),
  row.names = FALSE
)
```

```{r step1-run}
# How closely each listed duplicate tracks its anchor at the anchor's own
# loci: n_loci (loci where the duplicate is observed), coverage (share of the
# anchor's loci), pct_same_sign (effect in the same direction after orienting
# to the anchor's risk allele) and r_z (uncentred correlation of the oriented
# z-scores with the anchor's |z|; 1 = the same effects up to scale).
duplicate_evidence <- function(coloc_groups, tid, ids) {
  cg <- coloc_groups |>
    dplyr::mutate(trait_id = as.character(trait_id)) |>
    dplyr::filter(!is.na(beta), !is.na(se), se > 0)
  anchor <- cg |>
    dplyr::filter(trait_id == tid) |>
    dplyr::group_by(coloc_group_id) |>
    dplyr::slice_min(min_p, n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::transmute(coloc_group_id, z_anchor = beta / se)
  hits <- cg |>
    dplyr::filter(trait_id %in% ids) |>
    dplyr::group_by(trait_id, coloc_group_id) |>
    dplyr::slice_min(min_p, n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::inner_join(anchor, by = "coloc_group_id") |>
    dplyr::mutate(z = beta / se * sign(z_anchor), za = abs(z_anchor)) |>
    dplyr::group_by(trait_id) |>
    dplyr::summarise(
      n_loci = dplyr::n(),
      pct_same_sign = round(100 * mean(z > 0)),
      r_z = round(sum(z * za) / sqrt(sum(z^2) * sum(za^2)), 2),
      .groups = "drop"
    )
  out <- data.frame(anchor_id = tid, trait_id = ids, stringsAsFactors = FALSE) |>
    dplyr::left_join(hits, by = "trait_id") |>
    dplyr::mutate(
      n_loci = dplyr::coalesce(n_loci, 0L),
      coverage = round(n_loci / max(nrow(anchor), 1), 2)
    )
  return(out)
}

run_trait <- function(tid) {
  obj <- trait(tid, include_associations = TRUE)
  exclude_ids <- proxy_trait_ids[[tid]]
  dup_evidence <- NULL
  if (!is.null(exclude_ids)) {
    dup_evidence <- duplicate_evidence(obj$coloc_groups, tid, exclude_ids)
    obj$coloc_groups <- obj$coloc_groups |>
      dplyr::filter(!as.character(trait_id) %in% exclude_ids)
  }
  res <- run_univariate_clustering(
    obj,
    target_trait_id = tid,
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    max_snp_fraction = 0.8,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_module_size = min_module_size,
    min_mean_internal = min_mean_internal,
    min_connectedness = min_connectedness
  )
  smry <- summarise_ebmf_programs(
    res,
    s_matrix = res$s_matrix,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    similarity_n_perm = similarity_n_perm,
    similarity_q_threshold = similarity_q_threshold,
    high_similarity_q = high_similarity_q,
    high_min_connectedness = high_connectedness,
    min_effective_traits = min_effective_traits,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = FALSE
  )
  extracted <- extract_program_loadings(res, smry, valid_only = TRUE)
  list(
    trait_id = tid, fit = res, summary = smry, extraction = extracted,
    duplicate_evidence = dup_evidence
  )
}

trait_results <- lapply(trait_ids, run_trait)
names(trait_results) <- trait_ids

for (tid in names(trait_results)) {
  x <- trait_results[[tid]]
  x$extraction$loadings$trait_name <- trait_name_map[[tid]]
  x$extraction$profiles$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_name <- trait_name_map[[tid]]
  x$extraction$programs$trait_id <- tid
  trait_results[[tid]] <- x
}

program_meta <- dplyr::bind_rows(lapply(trait_results, function(x) {
  x$summary$programs |>
    dplyr::mutate(
      trait_id = x$trait_id,
      trait_name = trait_name_map[[x$trait_id]]
    )
}))
program_info <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$programs))
program_loadings <- dplyr::bind_rows(lapply(trait_results, function(x) x$extraction$loadings))
program_data <- lapply(trait_results, function(x) x$extraction)

knitr::kable(
  program_meta |>
    dplyr::group_by(trait_id, trait_name) |>
    dplyr::summarise(
      n_programs = dplyr::n(),
      n_high = sum(confidence_tier == "high"),
      n_medium = sum(confidence_tier == "medium"),
      n_single_trait = sum(confidence_tier == "single_trait"),
      n_kept_snps = sum(n_snps_filtered[confidence_tier %in% c("high", "medium")]),
      .groups = "drop"
    ),
  caption = paste0(
    "Per-trait EBMF programs by confidence tier; high- and medium-confidence ",
    "programs are kept for the cross-trait comparison"
  ),
  row.names = FALSE
)

knitr::kable(
  dplyr::bind_rows(lapply(trait_results, function(x) x$duplicate_evidence)) |>
    dplyr::left_join(
      duplicate_traits[, c("anchor_id", "trait_id", "trait_name")],
      by = c("anchor_id", "trait_id")
    ) |>
    dplyr::mutate(
      anchor = unname(trait_name_map[anchor_id]),
      note = ifelse(n_loci == 0, "not present: no effect", "")
    ) |>
    dplyr::select(anchor, trait_id, trait_name, n_loci, coverage, pct_same_sign, r_z, note),
  caption = paste0(
    "How closely each removed duplicate tracks its anchor at the anchor's own ",
    "loci. `coverage` is the share of the anchor's loci where it is observed, ",
    "`pct_same_sign` the share with an effect in the same direction after ",
    "orienting to the anchor's risk allele, and `r_z` the uncentred ",
    "correlation of its oriented z-scores with the anchor's |z| (1 = the same ",
    "effects up to scale). A true duplicate is near 100% and near 1; the ",
    "higher its coverage, the more it would have distorted the fit."
  ),
  row.names = FALSE
)
```

```{r step1-program-table, results='asis'}
kept_programs <- program_meta |>
  dplyr::filter(confidence_tier %in% c("high", "medium"))
kept_programs <- dplyr::bind_rows(lapply(
  split(kept_programs, kept_programs$trait_name),
  order_programs_by_confidence
))
kable_confidence(
  kept_programs |>
    dplyr::transmute(
      trait_name,
      program = program_label,
      confidence_tier = as.character(confidence_tier),
      n_snps_filtered,
      effective_n_traits = round(effective_n_traits, 2),
      similarity_q = signif(similarity_q, 2),
      connectedness = round(connectedness, 2),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      factor_pve_pct = round(100 * factor_pve, 2)
    ),
  caption = paste0(
    "High- and medium-confidence programs only (green high, orange medium), ",
    "a program nested inside a larger one indented under it -- everything ",
    "downstream uses these. The tier is set by `similarity_q` (are the ",
    "program's SNPs more alike than random SNP sets of the same size? below ",
    similarity_q_threshold, " to pass, below ", high_similarity_q, " for high), ",
    "`connectedness` (share of member pairs with similarity >= ",
    similarity_threshold, "; at least ", high_connectedness, " for high) and ",
    "`effective_n_traits` (at least ", min_effective_traits, ", otherwise ",
    "single_trait). `n_snps_filtered` is how many SNPs pass the program's ",
    "lFSR/magnitude gate; `mean_internal_similarity` and `factor_pve_pct` (the ",
    "share of matrix variance the factor explains) are descriptive. ",
    "Low-confidence and single-trait programs are not shown -- see the ",
    "univariate vignette for that diagnostic view."
  )
)
```

```{r coloc-by-trait}
# Each trait's own coloc_groups, keyed by trait id -- needed both for
# module_rg() (Step 3) and, for any pair involving an uploaded GWAS
# trait, to bridge that trait's own (private) coloc_group_id numbering onto its
# partner's (Steps 2-4); see compare_program_pairs_loadings()'s docs. Fetched
# once here, ahead of where it is first needed, rather than refetched per use.
coloc_by_trait <- stats::setNames(
  lapply(trait_ids, function(tid) {
    return(trait(tid, include_associations = TRUE)$coloc_groups)
  }),
  trait_ids
)

# Short display labels: "<trait name> P<k>".
program_label <- function(program_id) {
  tid <- sub(":[^:]*$", "", program_id)
  k <- sub("^.*:", "", program_id)
  return(paste0(unname(trait_name_map[tid]), " P", k))
}
```

## Step 2: Common locus universe

Programs are compared at the level of the **`coloc_group_id`** (the shared
fine-mapped locus) rather than the exact `variant_id`, because each trait carries
a single candidate SNP per coloc group. Two traits can only be matched where
they were fine-mapped to the *same* signal, which is far stricter than
positional overlap, so this universe bounds how much cross-trait evidence can
exist before any statistic is applied.

An uploaded GWAS numbers its coloc groups privately, so the T2D upload's loci
are first translated onto GPMap's shared numbering through a locus bridge
(the upload's own coloc record of each partner trait, then exact `variant_id`,
then LD). `compare_program_pairs_loadings()` builds those bridges, so it is
run here and its scores are read in Step 3.

```{r step2-universe}
conc_result <- compare_program_pairs_loadings(
  program_data,
  n_perm = n_perm,
  fdr_threshold = fdr_threshold,
  min_shared_loci = min_shared_loci_link,
  exclude_features = proxy_trait_ids,
  phi_high = phi_high,
  phi_low = phi_low,
  min_alignment = min_alignment,
  min_eff_loci = min_eff_loci,
  seed = 1,
  coloc_groups = coloc_by_trait
)
conc_pairs <- conc_result$pairs

# Every locus on GPMap's shared numbering (upload loci through their bridge).
canonical_loci <- program_loadings |>
  dplyr::filter(!is.na(coloc_group_id)) |>
  dplyr::distinct(trait_id, trait_name, coloc_group_id, high_confidence) |>
  dplyr::group_by(trait_id) |>
  dplyr::mutate(locus = gpmapr:::.canonical_locus_ids(
    coloc_group_id, trait_id[1], conc_result$bridges
  )) |>
  dplyr::ungroup()

trait_locus_counts <- canonical_loci |>
  dplyr::group_by(trait_id, trait_name) |>
  dplyr::summarise(
    n_loci_any = dplyr::n_distinct(locus),
    n_high_conf_loci = dplyr::n_distinct(locus[high_confidence]),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    program_loadings |>
      dplyr::group_by(trait_id) |>
      dplyr::summarise(n_programs = dplyr::n_distinct(program_id), .groups = "drop"),
    by = "trait_id"
  ) |>
  dplyr::select(trait_name, n_programs, n_loci_any, n_high_conf_loci)

shared_loci <- canonical_loci |>
  dplyr::distinct(trait_id, locus) |>
  dplyr::count(locus, name = "n_traits")

loci_by_trait <- split(canonical_loci$locus, canonical_loci$trait_id)
trait_pair_loci <- dplyr::bind_rows(lapply(
  utils::combn(trait_ids, 2, simplify = FALSE),
  function(tp) {
    return(data.frame(
      trait_a = trait_name_map[[tp[1]]],
      trait_b = trait_name_map[[tp[2]]],
      n_loci_shared = length(intersect(loci_by_trait[[tp[1]]], loci_by_trait[[tp[2]]])),
      stringsAsFactors = FALSE
    ))
  }
))

knitr::kable(
  trait_locus_counts,
  caption = "Candidate locus (coloc group) universe per trait (across valid programs)",
  row.names = FALSE
)
knitr::kable(
  trait_pair_loci,
  caption = paste0(
    "Loci each pair of traits shares -- the most any two of their programs ",
    "could be matched on."
  ),
  row.names = FALSE
)
knitr::kable(
  data.frame(
    n_loci_any = nrow(shared_loci),
    n_loci_shared_by_traits = sum(shared_loci$n_traits > 1),
    max_traits_per_locus = max(shared_loci$n_traits)
  ),
  caption = "Shared coloc-group universe across traits",
  row.names = FALSE
)
```

## Step 3: Tiers -- strength and direction

Every cross-trait pair of programs is asked two questions: do the same
background traits drive both programs (profile congruence, `phi_traits`), and
do the same SNPs carry them (the locus measures)? What each measure checks is
explained in
[Multi-trait EBMF program correspondence](multitrait-overview.html#what-each-measure-checks);
here they are combined.

`profile_strength` is **strong** when |φ| ≥ `r phi_high`, **moderate** when
`r phi_low` ≤ |φ| < `r phi_high`, and **weak** otherwise. `locus_status` has
three outcomes:

| `locus_status` | rule | reading |
|---|---|---|
| **too_few_loci** | `n_loci_shared` < `r min_shared_loci_link` **or** `n_eff_loci` < `r min_eff_loci` | the loci cannot settle it either way |
| **agree** | testable, `alignment` ≥ `r min_alignment` **and** `q_concordance` ≤ `r fdr_threshold` | the same SNPs carry both programs |
| **disagree** | testable, but `alignment` or `q_concordance` fails | the shared SNPs do not carry both programs |

The two are combined into a tier:

| tier | profile_strength | locus_status | reading |
|---|---|---|---|
| **high** | strong | agree | same biology, carried by the same SNPs |
| **medium** | strong | too_few_loci | same biology; too few shared loci to check the SNPs |
| **medium** | moderate | agree | a slightly weaker profile match, confirmed by the SNPs |
| **low** | strong | disagree | same background biology, but the shared SNPs do not carry it |
| **low** | moderate | too_few_loci | a weaker profile match the loci cannot check |
| **low** | weak | agree | **shared loci with different biology** -- typical of large pleiotropic loci that every lipid or glycaemic program touches |
| **none** | anything else | | including a moderate profile whose loci disagree |

High and medium pairs are **links** and form the families of Step 4. The φ
cutoffs follow Tucker's conventions for factor similarity (0.85 "fair
similarity", 0.95 "equal"; Lorenzo-Seva & ten Berge, 2006), applied to
genetic loadings they were not calibrated on: they are parameters of this
vignette (`phi_high`, `phi_low`, `min_shared_loci`, `min_eff_loci`,
`min_alignment`, `fdr_threshold`).

The **direction** of a link is the sign of `phi_traits`
(`direction_profile`). `rg` (`module_rg()`) checks it with the two traits' own
allele effects across the loci either program claims, calling a direction
only when its bootstrap interval excludes 0; it needs at least 10 usable
shared loci. `direction_check` compares the two: `agree`, `conflict` (worth
inspecting) or `no call`. `rg` never changes a tier.

| column | what it means |
|---|---|
| `tier`, `profile_strength`, `locus_status` | as above |
| `phi_traits` | profile congruence, signed; its sign is the direction |
| `n_eff_traits` | how many background traits effectively carry the profile agreement |
| `top_traits` | the background traits carrying it, with their share (negative = works against φ's sign) |
| `n_loci_shared`, `n_eff_loci` | shared high-confidence loci, and loci effectively carrying the locus score |
| `alignment` | R² of the SNP loadings over the shared loci |
| `comparability` | share of both programs' loading on the shared loci |
| `q_concordance` | BH-adjusted permutation p-value of the locus concordance |
| `direction_profile` | sign of `phi_traits` |
| `rg`, `rg_ci`, `direction_check` | allele-effect correlation, its 95% bootstrap interval, and whether its direction agrees with `direction_profile` (links only) |

For a pair involving the uploaded T2D GWAS, `rg` reads both traits' effects
from the **upload's own coloc table**, where the partner trait's row in each
upload coloc group sits at the upload's lead variant and on the same allele;
GPMap reports an upload's and an existing trait's associations on opposite
alleles at about half of all variants, so the partner's own table cannot be
used. `n_allele_mismatch` counts loci dropped because they could not be
aligned.

```{r step3-tiers, results='asis'}
tier_levels <- c("high", "medium", "low", "none")
print(knitr::kable(
  conc_pairs |>
    dplyr::mutate(
      trait_pair = paste(trait_name_map[trait_id_a], "~", trait_name_map[trait_id_b]),
      tier = factor(tier, levels = tier_levels)
    ) |>
    dplyr::count(trait_pair, tier, .drop = FALSE) |>
    tidyr::pivot_wider(names_from = tier, values_from = n, values_fill = 0),
  caption = "Program pairs per trait pair, by tier. High and medium are links.",
  row.names = FALSE
))

tiered <- conc_pairs[conc_pairs$tier != "none", , drop = FALSE] |>
  dplyr::left_join(
    conc_result$module_rg |>
      dplyr::select(
        program_id_a, program_id_b, rg, rg_ci_lower, rg_ci_upper,
        n_loci_rg, n_allele_mismatch, rg_direction, direction_check
      ),
    by = c("program_id_a", "program_id_b")
  ) |>
  dplyr::left_join(
    profile_contributions(conc_result, program_data, pairs = conc_pairs[conc_pairs$tier != "none", ])$pairs,
    by = c("program_id_a", "program_id_b")
  )
tiered <- tiered[order(
  factor(tiered$tier, levels = tier_levels), -abs(tiered$phi_traits)
), , drop = FALSE]

if (nrow(tiered) > 0) {
  print(kable_confidence(
    tiered |>
      dplyr::transmute(
        program_a = program_label(program_id_a),
        program_b = program_label(program_id_b),
        tier = as.character(tier), profile_strength,
        phi_traits = round(phi_traits, 3),
        n_eff_traits = round(n_eff_traits, 1),
        top_traits,
        locus_status,
        n_loci_shared,
        n_eff_loci = round(n_eff_loci, 1),
        alignment = round(alignment, 3),
        comparability = round(comparability, 3),
        q_concordance = signif(q_concordance, 3),
        direction_profile,
        rg = round(rg, 2),
        rg_ci = ifelse(
          is.finite(rg_ci_lower),
          sprintf("[%.2f, %.2f]", rg_ci_lower, rg_ci_upper), ""
        ),
        direction_check
      ),
    caption = paste0(
      "Every pair of tier low or above, by tier then |phi_traits| (green high, ",
      "orange medium, red low). Links are high and medium. `locus_status` is ",
      "`too_few_loci` when n_loci_shared < ", min_shared_loci_link,
      " or n_eff_loci < ", min_eff_loci, "; otherwise `agree` when alignment >= ",
      min_alignment, " and q_concordance <= ", fdr_threshold, ", else `disagree`. ",
      "`top_traits` are the background traits carrying phi. `rg` and ",
      "`direction_check` come from the traits' own allele effects (links only)."
    ),
    tier_col = "tier"
  ))
  linked_pairs <- tiered[tiered$linked, , drop = FALSE]
  if (nrow(linked_pairs) > 0) {
    print(knitr::kable(
      linked_pairs |>
        dplyr::mutate(
          trait_pair = paste(trait_name_map[trait_id_a], "~", trait_name_map[trait_id_b]),
          profile_vs_rg = dplyr::coalesce(direction_check, "no call")
        ) |>
        dplyr::count(trait_pair, direction_profile, profile_vs_rg) |>
        tidyr::pivot_wider(names_from = profile_vs_rg, values_from = n, values_fill = 0),
      caption = paste0(
        "Links per trait pair by profile direction, and whether `rg` (where it ",
        "calls a direction) agrees."
      ),
      row.names = FALSE
    ))
  }
} else {
  cat("No program pair reached the low tier.\n")
}
```

## Step 4: Program families

Links are grouped into **multi-trait families**: programs are nodes, links are
edges weighted by `|phi_traits|`, and families are the graph's communities
(Louvain modularity, `build_program_families()`). A program with no link
belongs to no family. With `mutual_best: true`, a link is kept only when each
of its programs is the other's closest partner (highest `|phi_traits|`) in the
other's trait, so a family tends to hold one program per trait (this run:
`mutual_best = ` `r mutual_best`).

Each family is described by the loci its programs claim as high-confidence
members, on GPMap's shared locus numbering: `n_core_loci` are claimed by every
member and `n_union_loci` by any. Family direction is counted from its links
(`n_rg_*`).

```{r step4-families, results='asis'}
conc_families <- build_program_families(
  conc_result,
  program_info = program_info,
  program_data = program_data,
  mutual_best = mutual_best
)
if (nrow(conc_families$families) > 0) {
  cat(sprintf(
    "\n%d families from %d linked programs (%d programs have no link); modularity %.2f.\n\n",
    nrow(conc_families$families), nrow(conc_families$nodes),
    conc_families$n_unlinked, conc_families$modularity
  ))
  print(knitr::kable(
    conc_families$families |>
      dplyr::mutate(mean_phi = round(mean_phi, 2)),
    caption = paste0(
      "Per-family metrics. `mean_phi` is the mean profile congruence of the ",
      "family's links and `n_tier_*` their tiers; `n_rg_*` count links by ",
      "allele-effect direction (links with no `rg` estimate count in none); ",
      "`n_core_loci` / `n_union_loci` count loci claimed by every member and by ",
      "any member."
    ),
    row.names = FALSE
  ))
} else {
  cat("No program pairs were linked, so there are no families.\n")
}
```

```{r step4-network, fig.height = 8, fig.width = 10, echo=FALSE}
if (!is.null(conc_families$graph)) {
  graph <- conc_families$graph
  node_traits <- igraph::V(graph)$trait_id
  trait_pal <- stats::setNames(
    grDevices::hcl.colors(length(trait_ids), "Dark 3"), trait_ids
  )
  # Direction from rg where it calls one, otherwise from the profile sign
  # (dashed).
  rg_dir <- igraph::E(graph)$rg_direction
  rg_called <- !is.na(rg_dir) & rg_dir %in% c("concordant", "antagonistic")
  edge_dir <- ifelse(rg_called, rg_dir, igraph::E(graph)$direction_profile)
  edge_col <- c(concordant = "#2a78d6", antagonistic = "#eb6834")[edge_dir]
  edge_col[is.na(edge_col)] <- "#9e9e9e"
  phi <- abs(igraph::E(graph)$phi_traits)
  fam_ids <- igraph::V(graph)$family
  groups <- split(seq_along(fam_ids), fam_ids)
  groups <- groups[lengths(groups) > 1]
  set.seed(1)
  igraph::plot.igraph(
    graph,
    layout = igraph::layout_with_fr(graph),
    vertex.color = unname(trait_pal[node_traits]),
    vertex.size = 9,
    vertex.frame.color = "white",
    vertex.label = sub("^.*:", "P", igraph::V(graph)$name),
    vertex.label.cex = 0.6,
    vertex.label.color = "white",
    edge.color = unname(edge_col),
    edge.lty = ifelse(rg_called, 1, 2),
    edge.width = 0.5 + 4 * (phi - min(phi)) / max(1e-9, max(phi) - min(phi)),
    mark.groups = groups,
    mark.col = grDevices::adjustcolor("grey80", alpha.f = 0.25),
    mark.border = "grey70",
    main = "Multi-trait program families"
  )
  graphics::legend(
    "topleft", legend = unname(trait_name_map[trait_ids]),
    fill = unname(trait_pal[trait_ids]), bty = "n", title = "trait", cex = 0.8
  )
  graphics::legend(
    "bottomleft",
    legend = c("concordant", "antagonistic", "solid: from rg", "dashed: from profile sign"),
    col = c("#2a78d6", "#eb6834", "grey30", "grey30"),
    lty = c(1, 1, 1, 2), lwd = 2, bty = "n", title = "link direction", cex = 0.8
  )
}
```

Each node is one program, coloured by trait and labelled with its program
number; shaded hulls are the families. Edge width is `|phi_traits|`, the
strength of the shared profile; edge colour is the link's direction -- from
`rg` where it calls one (solid), otherwise from the sign of the profile
congruence (dashed). Programs with no link are not drawn.

## Step 5: Do families recover the same biology?

Annotations play no part in forming a link, so agreement here is
**independent corroboration** rather than a restatement of the same evidence.
For each family, the high-confidence SNPs of its programs in each trait are
pooled and enriched separately, against that trait's full candidate universe,
for trait categories, tissues, and Reactome and KEGG pathways (through the
SNPs' colocalising QTL genes, `enrich_snp_group_pathways()`). If the traits of
one family independently surface the same category, tissue or pathway, the
family has support beyond the loadings that created it; if they surface
nothing in common, it rests on the loadings alone.

One caveat: a SNP set holding a large share of its trait's SNPs is being
compared against a background that is largely itself, so "none" for a dominant
program is expected rather than informative.

```{r step5-biology, results='asis'}
snp_set_enrichment <- function(snps, tid, label, what = c("category", "tissue")) {
  what <- match.arg(what)
  if (length(snps) < 5) {
    return("too few high-confidence SNPs")
  }
  baseline <- unique(program_loadings$snp_id[program_loadings$trait_id == tid])
  groups <- data.frame(snp_id = snps, group = label, stringsAsFactors = FALSE)
  res <- if (what == "category") {
    enrich_snp_group_trait_categories(
      groups = groups, coloc_groups = coloc_by_trait[[tid]],
      snp_key = "variant_id", min_group_size = 5L, baseline_snp_ids = baseline
    )
  } else {
    enrich_snp_group_tissues(
      groups = groups, coloc_groups = coloc_by_trait[[tid]],
      snp_key = "variant_id", min_group_size = 5L, background = "trait",
      baseline_snp_ids = baseline
    )
  }
  if (length(res$by_group) == 0 || nrow(res$by_group[[1]]$enriched) == 0) {
    return("none")
  }
  hits <- utils::head(res$by_group[[1]]$enriched, 4)
  hit_label <- if (what == "category") hits$trait_category else hits$tissue
  return(paste0(hit_label, " (FDR=", signif(hits$fdr, 2), ")", collapse = "<br>"))
}

# Top Reactome and KEGG terms of one SNP set, from its colocalising QTL genes.
snp_set_pathways <- function(snps, tid, label) {
  if (length(snps) < 5) {
    return(c(reactome = "too few high-confidence SNPs", kegg = "too few high-confidence SNPs"))
  }
  res <- tryCatch(
    enrich_snp_group_pathways(
      groups = data.frame(snp_id = snps, group = label, stringsAsFactors = FALSE),
      coloc_groups = coloc_by_trait[[tid]],
      min_group_size = 5L, snp_key = "variant_id",
      sources = c("KEGG", "Reactome")
    ),
    error = function(e) return(NULL)
  )
  if (is.null(res)) {
    return(c(reactome = "unavailable", kegg = "unavailable"))
  }
  if (nrow(res$summary) == 0) {
    return(c(reactome = "too few genes", kegg = "too few genes"))
  }
  label_of <- function(x) {
    x <- as.character(x[1])
    return(if (is.na(x) || !nzchar(x)) "none" else x)
  }
  return(c(
    reactome = label_of(res$summary$top_enriched_reactome),
    kegg = label_of(res$summary$top_enriched_kegg)
  ))
}

if (nrow(conc_families$nodes) > 0) {
  members <- conc_families$nodes
  biology <- dplyr::bind_rows(lapply(
    split(members, list(members$family, members$trait_id), drop = TRUE),
    function(m) {
      tid <- as.character(m$trait_id[1])
      snps <- unique(program_loadings$snp_id[
        program_loadings$program_id %in% m$program_id & program_loadings$high_confidence
      ])
      label <- paste(m$family[1], tid, sep = "|")
      pw <- snp_set_pathways(snps, tid, label)
      return(data.frame(
        family = m$family[1],
        trait = trait_name_map[[tid]],
        programs = paste(sub("^.*:", "P", m$program_id), collapse = ", "),
        n_snps = length(snps),
        categories = snp_set_enrichment(snps, tid, label, "category"),
        tissues = snp_set_enrichment(snps, tid, label, "tissue"),
        reactome = pw[["reactome"]],
        kegg = pw[["kegg"]],
        stringsAsFactors = FALSE
      ))
    }
  ))
  biology <- biology[order(as.integer(sub("^family_", "", biology$family)), biology$trait), ]
  print(knitr::kable(
    biology,
    caption = paste0(
      "Trait-category, tissue and pathway enrichment of each family's ",
      "high-confidence SNPs, pooled per trait and computed against that trait's ",
      "own SNP universe (pathways through the SNPs' colocalising QTL genes; top ",
      "Reactome and KEGG term). Matching entries across a family's traits are ",
      "held-out support for the family."
    ),
    row.names = FALSE,
    escape = FALSE
  ))
} else {
  cat("No families to annotate.\n")
}
```

## Interpretation notes

- **Profile first, loci second.** A link says two programs are driven by the
  same background traits in the same proportions (`phi_traits`); `alignment`
  adds whether the same SNPs carry it, where the traits share loci at all.
  Two traits that share few fine-mapped loci (T2D shares 36–70 with each
  partner here) can still share programs; their links are usually
  `too_few_loci`.
- **`too_few_loci` and `disagree` are different.** The first means the loci
  cannot check the profile match; the second means they were checked and the
  shared SNPs do not carry both programs, so a strong profile match is
  demoted to low.
- **Read `n_eff_traits` and `top_traits`.** A high congruence carried by one
  or two traits is a narrower claim -- often "these programs both carry the
  same lipid fraction" -- than one carried by dozens. Near-duplicate traits
  each count separately, so check the names, not just the number. Neither is
  used to gate.
- **Shared loci with different biology** (weak profile, loci agree, tier low)
  usually means large pleiotropic loci that many programs touch, not a shared
  program.
- **The profile sign is a direction.** Each trait's SNPs are oriented to its
  own risk allele, so antagonism mirrors the profile (\(\phi < 0\)). `rg`, from
  the traits' own allele effects, checks it where enough loci are shared; a
  `direction_check` of `conflict` is worth inspecting.
- **Cutoffs are judgements.** 0.85 and 0.70 come from factor-analysis
  conventions, not from calibration on genetic loadings.
- **Sample overlap is not corrected.** When the two target traits come from the
  same cohort, their effect estimates share participants, which biases `rg`
  away from the null. Treat rg magnitudes from same-cohort pairs as optimistic.
- **Winner's curse.** Loci enter each trait's universe only by clearing that
  trait's own significance and fine-mapping bar, so the shared set is doubly
  selected and effect sizes are inflated.
- **Restricted to confident programs.** Only programs of high or medium
  confidence in their own trait are compared; Step 1 shows exactly which.
