---
title: "Multi-trait EBMF program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Multi-trait EBMF program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take **two (or more) traits**, each already decomposed independently into EBMF
programs by the univariate pipeline
([SNP biological clustering](univariate-overview.html)),
and ask: **which programs describe the same biology across traits, which SNPs
carry it, and do those SNPs push both traits the same way or in opposite
directions?** Two programs are compared on both axes of their EBMF fits: their
**trait profiles** (`L_pm` -- which background traits drive them), which says
whether they describe the same biology, and their **SNP loadings** (`F_pm`) at
the loci the two traits share, which says whether the same SNPs carry it.

The runnable implementation is
[Investigation of multi-trait EBMF program correspondence](multitrait-analysis.html).

This is not a search for how much biology is shared. A trait's candidate SNP
set is already restricted to that trait's own genome-wide-significant,
fine-mapped hits, so two only-moderately-correlated traits will typically share
a small fraction of that universe — that is expected, not a failure of the
method. The job of this pipeline is to resolve, cleanly and with calibrated
significance, whatever correspondence *does* exist, and to say honestly
whether it is concordant or antagonistic.

### Core correspondence preferences

Current Methodology:

1. **Programs are discovered independently per trait.** No joint
   factorization across traits. This preserves each trait's own architecture
   (a program found in trait A is a real object even if trait B never enters
   the fit).
2. **Same biology is judged on the trait profile; same SNPs on the shared loci.**
   Two traits share only a small fraction of their fine-mapped loci,
   so a comparison that requires shared loci can only ever see a sliver of
   each program, and one that counts unshared loci as disagreement mostly
   measures how many loci the traits happen to share. The trait profile is
   defined on background traits both fits carry, so it can be compared for
   every pair; the SNP loadings are compared only on the loci both traits
   carry, and how much of each program those loci cover is reported
   separately.
3. **Effect sizes, not only significance.** :
   EBMF programs overlap heavily within a trait (on real data each
   triglyceride locus belonged to about four programs), so most programs share
   a few loci with many programs of another trait, and significance alone
   links them into a dense, many-to-many graph. Links require a strong
   profile congruence, graded into tiers by whether the shared loci agree,
   disagree or are too few to check (Step 4). There is no
   one-to-one restriction: EBMF's soft factorisation can split one module
   into several programs within a trait, so one program may correspond to
   several in another trait. Links are grouped into multi-trait **families**
   by community detection on the link graph.
4. **Direction is read from the profile sign and from observed allele
   effects.** Each trait is oriented to its own risk allele before fitting,
   so antagonism mirrors a program's trait profile (Step 3) -- the sign of the
   profile congruence is a direction -- while the SNP loadings of a shared
   module stay positive in both traits whichever way they move. Where enough
   loci are shared, `module_rg()` checks that direction against the traits'
   own allele effects; it never changes a tier.
5. **Biological annotation is held out.** Annotations play no role in
   forming a link, so agreement between the two linked programs' independently
   computed enrichment results is genuine support for the link, not a
   restatement of the evidence that created it.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Independent EBMF per trait\\nvalid programs only, biological annotation\"]
  S2 [label=\"2. Comparable traits and shared loci\\nper trait pair, uploads bridged by LD\"]
  S3 [label=\"3. Measures: phi (same traits?), alignment (same SNPs?)\\nshared / effective loci, q, rg (direction check)\"]
  S4 [label=\"4. profile_strength x locus_status -> tier\\nlinks = high + medium; direction = sign of phi\"]
  S5 [label=\"5. Families = communities of links\\nfamily-level rg\"]
  S1 -> S2 -> S3 -> S4 -> S5
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

### 1. Independent per-trait EBMF decomposition

Each target trait is run through the univariate pipeline exactly as described
in [SNP biological clustering](univariate-overview.html):
`run_univariate_clustering()` followed by `summarise_ebmf_programs()`.

`extract_program_loadings()` then sign-orients each program to its own target
trait, reflecting both the locus loadings (`F_pm`) and the background-trait
profile (`L_pm`) so that the target-trait loading is positive. EBMF's factor
sign is otherwise arbitrary and independent between the two traits' separate
fits; with it, the sign of a cross-trait *profile* comparison is a direction
(Step 3).

Output objects, one set per trait:

- `loadings` — one row per locus x program: `coloc_group_id`, `loading`
  (sign-oriented `F_pm`, the **SNP**-loading matrix), `lfsr` (the matching
  `F_lfsr`), `high_confidence` (the lFSR/magnitude gate that defines which
  loci a program *claims* — see Step 2). The locus side of the comparison.
- `profiles` — one row per background-feature x program (`L_pm`/`L_lfsr`, the
  **trait**-loading matrix — which background traits drive a program, not
  which SNPs). The profile side of the comparison. See [SNP biological clustering](univariate-overview.html#fit-the-ebmf-latent-program-model)
  for what `F`/`L`, `_pm`/`_lfsr` mean.
- `programs` — per-program metadata (`program_id`, `trait_id`, `trait_name`).

### 2. Comparable traits and shared loci

**Comparable traits.** Each trait's matrix has its own set of background
traits (rows): its own anchor, its own proxy removals, and the rows that pass
its own sparsity filters. Two programs' profiles are compared over the
background traits **both** fits carry, leaving out both anchor traits — each
is positive at every SNP in its own fit but an ordinary row in the other —
and any trait removed from either fit as a proxy (`exclude_features`).

For a pair of traits A and B, the **shared universe** is every
`coloc_group_id` both traits' fits carry. A given pair of programs is scored
on a narrower **axis** within it (`n_loci_axis`): the shared loci that at
least one of the two programs **claims** as a `high_confidence` member
(lFSR/magnitude gate), or all of a program's loci if it claims none. Loci
neither program claims are left out because every program carries a small
loading at every locus, and those near-zero tails can be faintly correlated
between two traits — over the whole shared universe a permutation null would
turn that into a large `concordance_z` for loadings of no real size. Within
the axis there is no further cutoff: Step 3's statistic down-weights
low-confidence loci continuously.

`high_confidence` membership also decides whether the loci can be tested at
all: below `min_shared_loci` (default 3) shared high-confidence loci
(`n_loci_shared`) a pair's `locus_status` is `too_few_loci` (Step 4).
`jaccard_loci` and `p_locus` (overlap of the two membership sets) are
reported alongside.

**Uploaded GWAS.** An uploaded GWAS numbers its coloc groups privately, so its
`coloc_group_id`s never equal an existing trait's. For any pair involving an
upload, a **locus bridge** translates the upload's loci onto the partner's:
first by the upload's own coloc record of that trait
(`existing_study_extraction_id`), then by exact `variant_id`, then by LD
(`ld_proxies()`, `|r| >= min_ld_r`). Every locus comparison for that pair —
eligibility, the score, `module_rg()` and the family locus sets — uses the
same bridge.

Because each trait's candidate universe is independently restricted to that
trait's own significant, fine-mapped hits, the shared universe is typically a
small fraction of either trait's total (on real data 36–143 loci per trait
pair, against 440–620 per trait). This bounds how much locus evidence can
exist, and is why the profile, not the loci, is the primary comparison.

### 3. What each measure checks

Every cross-trait pair of programs is described by a handful of measures. Each
answers one narrow question; Step 4 combines them. In short: **φ** asks
whether the same background traits drive both programs, **alignment** whether
the same SNPs carry them, and **rg** whether the two traits' own allele
effects point the same way.

#### φ vs alignment vs rg

| | φ | alignment | rg |
|---|---|---|---|
| **data** | the EBMF fits: trait loadings (`L`) | the EBMF fits: SNP loadings (`F`) | the two traits' own GWAS effects, outside EBMF |
| **question** | do the same background traits drive both programs? | do the same SNPs carry both programs? | do the module's alleles move the two traits together or apart? |
| **the two anchor traits** | left out | only through their effect on the fit | the only measure that uses them directly |
| **direction** | its sign (mirrored profile = antagonistic) | none | its sign, in a common allele frame |
| **needs** | comparable traits (always 500+) | at least `min_shared_loci` shared and `min_eff_loci` effective loci | at least 10 usable loci and a CI that excludes 0 |
| **role** | gate | gate (with `q_concordance`) | check only |

φ and alignment both describe the decomposition; rg is the only check on
data the factorisation never saw.

#### φ (`phi_traits`): do the same background traits drive both programs?

A program's trait profile (`L_pm`) lists how strongly each background trait
loads on it. φ compares two programs' profiles over the comparable traits
(Step 2) with **Tucker's congruence coefficient**.

The easiest way to read it is as **Pearson's correlation without the
mean-centring**. Pearson's r subtracts each vector's mean and then takes the
cosine of the angle between them; Tucker's φ takes the cosine directly:

\[
\phi = \frac{\sum_t l_a(t)\, l_b(t)}{\sqrt{\sum_t l_a(t)^2 \, \sum_t l_b(t)^2}}
\]

(each loading weighted by its confidence, \((1 - 2\,\text{lfsr})_+\): 0 for a
sign that is a coin flip, 1 for a certain one). It is the same quantity as
the correlation in a regression through the origin. Skipping the centring
matters here because a loading of zero has a meaning -- "this trait does not
drive the program" -- and most of a program's 550+ loadings are near zero.
Centring would move every one of them away from zero by the mean, so two
programs would look alike or unlike partly because of the hundreds of traits
neither of them uses.

- **Range and sign.** Like r, φ runs from −1 to 1, and its sign is a
  direction. Each trait's SNPs are oriented to its own risk allele, so a
  module that raises trait A and lowers trait B appears in B's fit with every
  background trait mirrored: φ near −1 is an antagonistic program, near +1 a
  concordant one.
- **An effect size, not a test.** φ has no p-value. The cutoffs used (0.85
  "fair similarity", 0.95 "equal"; Lorenzo-Seva & ten Berge, 2006) are
  conventions from factor analysis, much as 0.1 / 0.3 / 0.5 are Cohen's
  conventions for r. They were set for psychometric factors, not genetic
  loadings, and are not calibrated here.

#### `n_eff_traits` and `top_traits`: how many traits carry φ?

Every comparable trait contributes the product of its two loadings to φ's
numerator. `n_eff_traits` is the effective number of traits making those
contributions: 1 when a single trait carries the whole agreement, *k* when *k*
traits contribute equally. `top_traits` names the largest contributors and
their share (`profile_contributions()`).

A low value is not a bad sign in itself -- EBMF profiles are sparse, and most
pairs rest on two or three traits -- but it is a narrower claim: "both programs
are mostly about trait X". It needs a closer look when trait X is a
measurement of one of the two anchors, which is why anchor duplicates are
removed (below). Near-duplicate traits push the number the other way: five
GWAS of one lipid fraction count as five traits, so a match can look broad
when it is one trait measured several times. Read the names, not just the
number. Neither column is used to gate.

#### alignment and comparability: do the same SNPs carry both programs?

`alignment` applies the same uncentred correlation to the programs' **SNP
loadings** (`F_pm`) over the loci both traits carry, and squares it -- so it
reads like an R² from a regression through the origin: 1 when the two
programs load on the same loci in the same proportions, near 0 when they load
on different loci. Loci only one trait carries cannot agree or disagree, so
they are left out; `comparability` reports how much of each program they
leave out (the share of each program's loading on the shared loci,
multiplied). A pair with `alignment` 0.9 and `comparability` 0.05 agrees fully
on a sliver of each program. alignment says nothing about direction: a shared
module loads positively in both fits whichever way the traits move.

#### Shared and effective loci: is there enough to test?

`n_loci_shared` counts loci both programs **claim**: lFSR below
`ebmf_lfsr_threshold` and |loading| above `ebmf_magnitude_threshold` in each
program, at a locus both traits carry. `n_eff_loci` is the effective number of
loci carrying the locus score (the same participation ratio as
`n_eff_traits`), so a few loci with huge loadings cannot stand in for many.
Two traits share only a small fraction of their loci, so these are often
small.

#### `q_concordance`: is the SNP agreement more than chance?

A permutation test: shuffle which locus is which in one program, many times,
and ask how often the shuffled loadings agree with the other program as well
as the real ones do. `q_concordance` is that p-value with a Benjamini-Hochberg
correction across **all** scored pairs (correcting over only the pairs with
enough shared loci would be anti-conservative, because having enough shared
loci is itself related to the statistic).

#### rg: do the two traits' own allele effects agree?

`module_rg()` takes each trait's **own GWAS effects** (β, se) at the loci
either program claims and computes their correlation through the origin,
weighting each locus by how strongly the module loads on it and by the
precision of the two estimates. It is similar in spirit to a local genetic
correlation restricted to the module's loci. `rg > 0` means the alleles that
raise trait A also raise trait B across the module; a direction is called
(`rg_direction`) only when its bootstrap interval excludes 0, and at least
`min_loci_rg` (10) usable loci are needed.

Two practical points. For a pair involving an uploaded GWAS, both effects are
read from the upload's own coloc table, because GPMap reports an upload's and
an existing trait's associations on opposite alleles at about half of all
variants (`n_allele_mismatch` counts loci that could not be aligned). And
neither rg nor the locus test corrects for **sample overlap** between the two
GWAS, which biases rg away from zero for same-cohort traits, or for
**winner's curse** at loci selected by both traits.


#### Duplicate traits

Duplicates matter in three different places:

1. **A duplicate of the anchor inside its own fit** (UKB triglycerides in the
   triglycerides fit). It is positive at almost every SNP after orientation,
   so it adds the same component to every SNP and hides the programs that
   differ between them -- the problem the BMI and T2D proxy lists solve.
   Each anchor's duplicates are removed before its fit.
2. **A duplicate of the anchor inside the partner's fit, during comparison**
   (UKB triglycerides in the HDL fit, in a TG ~ HDL comparison). There it is
   an ordinary background trait, but it stands in for an anchor, and could
   carry a high φ almost on its own -- the reason the anchors themselves are
   left out. A trait listed for either anchor is therefore left out of every
   comparison involving that anchor (`exclude_features`), while staying in
   the other fits.
3. **Near-duplicate background traits** (NMR lipoprotein subclasses, the
   several T2D diagnosis traits inside the lipid fits). These are kept: they
   measure related but different things, and may be exactly what separates
   two programs. Their cost is that one biological axis is counted several
   times, which shows up in `top_traits`.

Only true duplicates are listed: other GWAS of the same measurement, and for
T2D the diagnosis, treatment and family-history traits defined by it. The
runnable vignette shows, for each listed trait, how closely it tracks its
anchor at the anchor's own loci.

### 4. How the measures are used together

**What gates and what is reported.** A pair's tier depends only on φ and on
the locus measures (`n_loci_shared`, `n_eff_loci`, `alignment`,
`q_concordance`). `n_eff_traits`, `top_traits`, `comparability` and rg are
reported alongside to help read a link, but never change its tier.

**Profile strength** (`profile_strength`): **strong** when
|φ| ≥ `phi_high` (0.85), **moderate** when `phi_low` (0.70) ≤ |φ| < `phi_high`,
**weak** otherwise.

**Locus status** (`locus_status`) separates "could not check" from "checked
and disagreed":

| `locus_status` | rule |
|---|---|
| **too_few_loci** | `n_loci_shared` < `min_shared_loci` (3) **or** `n_eff_loci` < `min_eff_loci` (5) |
| **agree** | testable, **and** `alignment` ≥ `min_alignment` (0.5) **and** `q_concordance` ≤ `fdr_threshold` (0.05) |
| **disagree** | testable, but `alignment` or `q_concordance` fails |

**Tiers.**

| tier | profile_strength | locus_status | reading |
|---|---|---|---|
| **high** | strong | agree | same biology, carried by the same SNPs |
| **medium** | strong | too_few_loci | same biology; too few shared loci to check the SNPs |
| **medium** | moderate | agree | a slightly weaker profile match, confirmed by the SNPs |
| **low** | strong | disagree | same background biology, but the shared SNPs do not carry it |
| **low** | moderate | too_few_loci | a weaker profile match the loci cannot check |
| **low** | weak | agree | shared loci, different biology -- typical of large pleiotropic loci |
| **none** | anything else | | including a moderate profile whose loci disagree |

**Links** are high- and medium-tier pairs. There is no one-to-one
restriction: EBMF can split one module into several programs within a trait,
so a program may link to several programs of another trait.

**Direction.** A link's direction is the sign of φ (`direction_profile`).
Where rg calls a direction, `direction_check` compares the two: `agree`,
`conflict` or `no call`. A conflict is a flag to inspect -- for example, a
program whose target-trait loading is small has an uncertain orientation,
which makes φ's sign uncertain too -- not something rg decides on its own.

**Rules considered and not used.** Requiring the loci to agree for every link
(high tier only) was stricter, but drops pairs that simply share too few loci
to check, including some whose few shared loci agree almost perfectly.
Rescuing strong-φ pairs whose loci disagree when rg confirms the direction
was looser, but would let rg -- with its sample-overlap bias -- decide links.

### 5. Program families

`build_program_families()` builds an undirected graph with every linked
program as a node and every link as an edge weighted by `|phi_traits|`, and
takes its **communities** (Louvain modularity, seeded) as the families.
`rule = "components"` uses plain connected components instead;
`mutual_best = TRUE` keeps only links where each program is the other's
closest partner (highest `|phi_traits|`) in the other's trait. A program with
no link belongs to no family (`n_unlinked`).

Each family is described by the loci its programs claim, on one locus id space
across traits (an upload's loci translated through its bridge):
`n_core_loci` are claimed by every member and `n_union_loci` by any;
`family_loci` lists them with their SNPs.

**Family direction.** A single link often rests on too few loci for an rg
estimate. `family_module_rg()` pools, for each family and each pair of traits
it spans, the loci claimed by all of that family's programs in those two
traits (each trait's loading at a locus taken from whichever of its programs
loads there most), and runs the same estimate.

### 6. Interpreting families: biological corroboration

For each family, the high-confidence SNPs of its programs in each trait are
pooled and enriched **separately**, against that trait's own full candidate
SNP universe (not the tiny module itself, and not the other trait), using the
existing hard-cutoff enrichment functions
(`enrich_snp_group_trait_categories()`, `enrich_snp_group_tissues()`, and
`enrich_snp_group_pathways()` for Reactome and KEGG through the SNPs'
colocalising QTL genes). FDR here is the ordinary hypergeometric BH
correction described in
[SNP biological clustering](univariate-overview.html) —
corrected across the categories/tissues tested **for that one SNP set**,
independently for each trait. It is not pooled across programs the way the
continuous univariate enrichment engine is; the two are separate mechanisms
operating on separate function families.

Because these annotations play no role in forming the links, agreement between
a family's traits is genuine, held-out support that the correspondence is
biologically real, not a restatement of the loading evidence that created it.
Disagreement does not disprove the link — it means the link rests on the
loadings alone, which is still meaningful evidence, just unconfirmed by a
second, independent axis.

One caveat: a program holding a large share of its trait's SNPs is compared
against a background that is largely itself, so "no significant enrichment"
for a dominant program is expected rather than informative — it says the
program is not distinguishable from the trait's overall composition, not that
it lacks biology.

## Variables to consider

Exposed in the multi-trait vignette today. Parameters fall into two groups:
those controlling the **correspondence test** itself, and those inherited
from the **univariate pipeline** that determine what counts as a program at
all before cross-trait comparison begins.

### Program correspondence parameters

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `phi_high` / `phi_low` | `0.85` / `0.70` | `profile_strength` cutoffs: strong / moderate | The main lever on which pairs link. Tucker's conventions, not calibrated on genetic loadings; the multi-trait vignette shows the distribution of \|φ\| they cut. |
| `min_shared_loci` | `3` (default), `0` | Below it, `locus_status` is `too_few_loci` | Shared *high-confidence* loci. Every pair is still scored, reported and included in the BH correction. |
| `min_eff_loci` | `5` | Below it, `locus_status` is `too_few_loci` | Stops a few loci with very large loadings from standing in for many. |
| `min_alignment` | `0.5` | A testable pair needs it for `agree` | Corresponds to a congruence of about 0.71 on the shared loci. |
| `fdr_threshold` | `0.05` | A testable pair needs `q_concordance` at or below it for `agree` | Lowering it turns more testable pairs into `disagree`. |
| `n_perm` | `200`, `1000`, `3000` | Permutation replicates for the locus null | The smallest attainable p-value is `1/(n_perm + 1)`, so with few replicates many strong pairs tie at the same BH floor. |
| `exclude_features` | duplicate lists per anchor | Traits removed from that anchor's fit and left out of every profile comparison involving it | See "Duplicate traits" in Step 3. |
| `options(gpmapr.lfsr_weight)` | `"two_sided"` (default), `"one_sided"` | Confidence weight of a loading: \((1 - 2\,\text{lfsr})_+\) or \(1 - \text{lfsr}\) | `"two_sided"` gives a coin-flip sign no weight; `"one_sided"` still gives it half. Applies to the profile, the locus measures and the `rg` weights. |
| `mutual_best` (in `build_program_families()`) | `FALSE` (default), `TRUE` | Keep only links that are each program's closest link (highest \|φ\|) to the other's trait | `TRUE` gives at most one partner per program per other trait, so families tend to hold one program per trait; programs that split one module lose their weaker links. |
| `min_ld_r` | `0.5` | Minimum `\|r\|` for the LD tier of an upload's locus bridge | GPMap's proxy table only stores `r^2 >= 0.8`, so values below about `0.89` have no effect. |
| `rule` (in `build_program_families()`) | `"communities"` (default), `"components"` | How links are grouped into families | Communities (Louvain) keep densely linked groups apart; components join anything reachable through a chain of links. |
| `seed` (in `build_program_families()`) | `1` | Seed for the community detection | Louvain is randomised; a different seed can move a program on the boundary between two families. |
| `linked_only` (in `module_rg()`) | `TRUE` (default), `FALSE` | Which pairs get an `rg` estimate | `FALSE` estimates `rg` for every pair supplied, linked or not. |
| `n_boot` (in `module_rg()`, `family_module_rg()`) | `1000` | Percentile-bootstrap replicates for the `rg` confidence interval | More replicates give a smoother CI estimate; the CI's width is driven mainly by `n_loci_rg`, not `n_boot`, so increasing this rarely changes whether the CI excludes zero. |
| `min_loci_rg` (in `module_rg()`, `family_module_rg()`) | `10` | Minimum usable loci for an `rg` estimate | Below it the link has no direction estimate; the family-level estimate pools more loci and often clears it where single links do not. |

### Univariate parameters that flow through

Everything in [SNP biological clustering](univariate-overview.html#variables-to-consider)
applies per trait before this pipeline ever runs, and matters twice over here:
`ebmf_lfsr_threshold`, `ebmf_magnitude_threshold`, `min_module_size`,
`similarity_q`, `high_similarity_q`, `high_connectedness` and
`min_effective_traits` together decide
each program's confidence tier (and therefore whether it is eligible for
cross-trait comparison at all), while `trait_subset`, `include_trans`, `min_snp_signals`,
and `compress_method` shape the pleiotropy matrix each trait's programs were
discovered from in the first place. A program that is only marginally valid
in one trait's own terms is not made more reliable by finding a cross-trait
partner for it.

## Verification Strategy

The simulation study that checks this pipeline where the answer is known is
described in
[Simulation overview](simulations-overview.html#multi-trait-simulations-testing-correspondence).

## Structure of Results

The multi-trait analogue of the univariate trait/program/SNP levels is
trait-pair/link/family/locus: the question moves from "what structure exists
within this trait" to "how much of it is shared, is it reliable, and which
direction does it point."

| Level | Primary question | Main output |
|---|---|---|
| **Trait pair** | How much shared candidate architecture exists between these two traits, and how much of it resolves into linked programs? | Size of the shared locus universe (vs. each trait's total), number of valid programs per trait, number of links and their directions |
| **Link** | Do the two programs describe the same biology, do the same SNPs carry it, and which way does it point? | `tier`, `profile_strength`, `locus_status`, `phi_traits` with `n_eff_traits` / `top_traits`, `alignment`, `comparability`, `n_loci_shared`, `n_eff_loci`, `q_concordance`, `direction_profile`, `rg` with CI and `direction_check` |
| **Family** | Which programs, across all traits, describe the same biology, and which SNPs hold them together? | Members per trait, `n_core_loci` / `n_union_loci`, `family_loci`, family-level `rg` per trait pair, biological-annotation agreement across the family's traits |
| **Locus** | Which loci actually carry the shared signal for this pair? | Each program's loading at the shared loci (which dominate — worth checking when `n_eff_loci` is small), each trait's own `beta`/`se` there |

As with the univariate framework, the aim is not a single verdict per program
pair but a layered result: statistical evidence of correspondence and
direction, kept distinct from the biological annotation used to interpret it.
