---
title: "Simulation study of multi-trait program correspondence"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
params:
  # --- generative structure -------------------------------------------------
  # 240 loci per trait: at SNP overlap 0 the private pool must hold
  # (K_shared + K_specific) x module_size = 144 loci.
  n_loci_per_trait: 240
  n_shared_loci: 80
  # Two shared programs per replicate, one concordant and one antagonistic, so
  # direction is scored in every replicate rather than being a grid axis.
  K_shared: 2
  directions: !r c(1, -1)
  K_specific: 2
  module_size: 36
  n_drivers_per_program: 8
  n_background_traits: 300
  effect_size: 6
  locus_effect_sd: 0.5
  noise_sd: 1
  p_structural_zero: 0
  # --- background realism ---------------------------------------------------
  p_active_background: 0.02
  background_sparsity_sd: 1.2
  background_effect_scale: 0.5
  background_corr: 0
  snp_pleiotropy_sd: 0
  # An unshared driver is still a row in the other trait's matrix, as an
  # ordinary background study, so profile congruence is penalised for it.
  unshared_driver_rows: "background"
  # --- the grid -------------------------------------------------------------
  # Shared loci of 36: 0, 4, 9, 18, 36. Shared drivers of 8: 0, 2, 4, 6, 8.
  snp_overlap: !r c(0, 0.1, 0.25, 0.5, 1.0)
  driver_overlap: !r c(0, 0.25, 0.5, 0.75, 1.0)
  # --- linking, at the defaults of compare_program_pairs_loadings() ---------
  phi_high: 0.85
  phi_low: 0.70
  min_alignment: 0.5
  min_eff_loci: 5
  min_shared_loci: 3
  fdr_threshold: 0.05
  min_loci_rg: 10
  n_perm: 1000
  # --- calibration targets (Q3) ---------------------------------------------
  # A cutoff set is acceptable when false links per null replicate stay at or
  # below max_false_links and pooled precision in the shared runs is at least
  # min_precision; the recommended set is the acceptable one with the highest
  # mean recall over the grid.
  max_false_links: 0.05
  min_precision: 0.95
  # rg calls a direction reliably from the smallest n_loci_rg at which every
  # larger bin has sign accuracy of at least this.
  min_rg_accuracy: 0.95
  # --- replication ----------------------------------------------------------
  n_sim: 20
  master_seed: 1
  # Replicates within a version run in parallel via parallel::mclapply
  # (fork-based; falls back to serial on Windows or when cores = 1).
  cores: 5
  # quick: fast test mode. Caps n_sim at 2 and lowers the permutation counts.
  # Every grid cell is always run. Set FALSE for the full study.
  quick: true
vignette: >
  %\VignetteIndexEntry{Simulation study of multi-trait program correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, out.width = "100%", fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(ggplot2)

quick <- isTRUE(params$quick)
n_sim <- if (quick) min(as.integer(params$n_sim), 2L) else as.integer(params$n_sim)
n_perm <- if (quick) 300L else as.integer(params$n_perm)
similarity_perm <- if (quick) 300L else 10000L

# Shared EBMF settings, from the same file the univariate and real-data studies
# read, so the per-trait discovery step here is the one used everywhere else.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

# One seed vector for every version: common random numbers.
set.seed(params$master_seed)
sim_seeds <- sample.int(.Machine$integer.max, n_sim)

generative_base <- list(
  n_loci_per_trait = as.integer(params$n_loci_per_trait),
  n_shared_loci = as.integer(params$n_shared_loci),
  K_specific = as.integer(params$K_specific),
  module_size = as.integer(params$module_size),
  n_drivers_per_program = as.integer(params$n_drivers_per_program),
  n_background_traits = as.integer(params$n_background_traits),
  effect_size = params$effect_size,
  locus_effect_sd = params$locus_effect_sd,
  noise_sd = params$noise_sd,
  p_structural_zero = params$p_structural_zero,
  p_active_background = params$p_active_background,
  background_sparsity_sd = params$background_sparsity_sd,
  background_effect_scale = params$background_effect_scale,
  background_corr = params$background_corr,
  snp_pleiotropy_sd = params$snp_pleiotropy_sd,
  unshared_driver_rows = params$unshared_driver_rows
)

# The linking cutoffs the pipeline runs at, and the ones every re-tiering in
# Q3 is compared against.
current_cutoffs <- list(
  phi_high = as.numeric(params$phi_high),
  phi_low = as.numeric(params$phi_low),
  min_alignment = as.numeric(params$min_alignment),
  min_eff_loci = as.numeric(params$min_eff_loci),
  min_shared_loci = as.integer(params$min_shared_loci),
  fdr_threshold = as.numeric(params$fdr_threshold)
)

palette_direction <- c(concordant = "#2a78d6", antagonistic = "#eb6834")
palette_tier <- c(
  high = "#1a6b3c", medium = "#74b98a", low = "#e3b448", none = "#d0d0d0",
  `no valid pair` = "#8c8c8c"
)
```

## How to read this document

gpmapr finds *programs* -- groups of loci that behave alike across hundreds of
background GWAS -- separately in each target trait, then links two traits'
programs with `compare_program_pairs_loadings()`, as the real-data analysis
([multi-trait analysis](multitrait-analysis.html)) does. This document runs
that pipeline on simulated pairs of traits where the answer is known, and asks
three questions:

1. **Are planted programs recovered?** Is each planted program found in its own
   trait, and is a program planted in both traits linked across them, in the
   right direction?
2. **How much do the two copies need to have in common?** The two copies of a
   shared program are planted with only part of their driver studies and only
   part of their SNPs in common. Where does linking stop finding them?
3. **What cutoffs?** Which values of `phi_high`/`phi_low`, `min_alignment`
   and `min_loci_rg` keep false links rare while linking as much as possible?

**Glossary.**

- **Program**: one EBMF factor -- its SNP loadings and the background studies
  (trait loadings) that drive it.
- **φ (`phi_traits`)**: congruence of two programs' trait loadings over the
  background studies both fits carry. Its sign is the link's direction.
- **alignment**: R² of two programs' SNP loadings over the loci both traits
  carry; tested only when they share enough loci (`n_loci_shared`,
  `n_eff_loci`), otherwise the loci are `too_few_loci`.
- **Tier / link**: φ strength (strong ≥ `phi_high`, moderate ≥ `phi_low`)
  combined with the locus status (agree / disagree / too few). High and medium
  are links.
- **rg** (`module_rg()`): correlation of the two targets' own allele effects
  over the program's shared loci; a check on direction, called only when its
  bootstrap CI excludes 0.
- **Driver / SNP overlap**: the fraction of the shared program's 8 driver
  studies, and of its 36 loci, that are the same in both traits.

```{r settings-table}
knitr::kable(
  data.frame(
    parameter = c(
      names(generative_base), "K_shared", "directions", "snp_overlap",
      "driver_overlap", names(current_cutoffs), "min_loci_rg", "n_perm",
      "max_false_links", "min_precision", "n_sim", "quick",
      "ebmf_settings_signature"
    ),
    value = c(
      vapply(generative_base, function(x) paste(x, collapse = ", "), character(1)),
      params$K_shared, paste(params$directions, collapse = ", "),
      paste(params$snp_overlap, collapse = ", "),
      paste(params$driver_overlap, collapse = ", "),
      vapply(current_cutoffs, as.character, character(1)),
      params$min_loci_rg, n_perm, params$max_false_links, params$min_precision,
      n_sim, quick, settings_signature
    ),
    stringsAsFactors = FALSE
  ),
  row.names = FALSE,
  caption = paste0(
    "Active settings. ebmf_settings_signature identifies the shared discovery ",
    "settings resolved from _ebmf_settings.yml."
  )
)
```

## Design

Following Morris, White and Crowther (2019, *Statistics in Medicine*
38:2074-2102).

**Data.** `simulate_trait_pair()` builds two target traits, each a matrix of
background studies × loci. Two programs are planted in both traits, one
**concordant** (raises both) and one **antagonistic** (raises one, lowers the
other), and `K_specific` programs in each trait alone. Each copy of a shared
program has 36 loci and 8 driver studies; a fraction `snp_overlap` of the loci
and a fraction `driver_overlap` of the drivers are the same in both traits,
the rest are the trait's own. A driver that is not shared is still a row in the
other trait's matrix, where it behaves like any background study. The grid
crosses the two overlaps (25 cells), plus a **null** in which nothing is
shared.

**Method.** The real-data pipeline: per trait, `run_univariate_clustering()`
and `summarise_ebmf_programs()` at the shared settings; then
`compare_program_pairs_loadings()` on the valid (high- or medium-confidence)
programs of the two traits, and `module_rg()` on every pair. Every cross-trait
pair is stored with its raw φ, alignment, q-value and rg, so Q3 can re-tier
them at any cutoffs without refitting.

**Estimands.** Discovered programs are mapped to planted ones by SNP
membership (`map_programs_to_planted()`, Jaccard ≥ 0.1). A pair is **true**
when both of its programs map to the same shared program.

- *Recall*: per shared program, whether any true pair is linked. A shared
  program lost by either trait's validation counts as not recovered.
- *False links*: linked pairs that are not true, per replicate.
- *Precision*: true links over all links, pooled over the shared runs.
- *Direction accuracy*: on true pairs, the φ sign and the rg call against the
  planted direction.

**Monte Carlo uncertainty.** Each estimate is a mean over `n_sim` replicates;
tables give the Monte Carlo SE in parentheses. All versions share one seed
vector.

```{r versions}
versions <- list()
for (dov in as.numeric(params$driver_overlap)) {
  for (sov in as.numeric(params$snp_overlap)) {
    versions[[length(versions) + 1L]] <- list(
      label = sprintf("snp %.2f, drivers %.2f", sov, dov),
      family = "shared",
      snp_overlap = sov,
      driver_overlap = dov,
      spec = list(
        K_shared = as.integer(params$K_shared), locus_overlap = sov,
        profile_overlap = dov, directions = as.integer(params$directions)
      )
    )
  }
}
# The null: programs of their own in each trait and nothing shared, so every
# link is a false discovery.
versions[[length(versions) + 1L]] <- list(
  label = "null: nothing shared",
  family = "null",
  snp_overlap = NA_real_,
  driver_overlap = NA_real_,
  spec = list(K_shared = 0L, locus_overlap = 0, profile_overlap = 0)
)
```

```{r run-helpers}
# Fit one trait and validate every program. Returns the extraction of every
# fitted program and each program's status ("valid" = high or medium
# confidence, else its tier).
discover_one_trait <- function(trait_object, trait_id) {
  res <- run_univariate_clustering(
    trait_object,
    target_trait_id = as.integer(trait_id),
    trait_subset = ebmf_settings$trait_subset,
    min_snp_signals = as.integer(ebmf_settings$min_snp_signals),
    compress_method = ebmf_settings$compress_method,
    compress_scale = as.numeric(ebmf_settings$compress_scale),
    ebmf_lfsr_threshold = as.numeric(ebmf_settings$ebmf_lfsr_threshold),
    ebmf_magnitude_threshold = as.numeric(ebmf_settings$ebmf_magnitude_threshold),
    ebmf_greedy_Kmax = as.integer(ebmf_settings$ebmf_greedy_Kmax),
    ebmf_prior = ebmf_settings$ebmf_prior,
    ebmf_backfit = isTRUE(ebmf_settings$ebmf_backfit),
    min_module_size = as.integer(ebmf_settings$min_module_size)
  )
  # n_rep = 0: replication refits answer none of this study's questions.
  smry <- summarise_ebmf_programs(
    res, s_matrix = res$s_matrix,
    edge_threshold = as.numeric(ebmf_settings$similarity_threshold),
    min_module_size = as.integer(ebmf_settings$min_module_size),
    similarity_n_perm = similarity_perm,
    similarity_q_threshold = as.numeric(ebmf_settings$similarity_q),
    high_similarity_q = as.numeric(ebmf_settings$high_similarity_q),
    high_min_connectedness = as.numeric(ebmf_settings$high_connectedness),
    min_effective_traits = as.numeric(ebmf_settings$min_effective_traits),
    n_null = 2,
    n_rep = 0,
    verbose = FALSE
  )
  program_tier <- as.character(smry$programs$confidence_tier)
  return(list(
    all = extract_program_loadings(res, smry, valid_only = FALSE),
    status = data.frame(
      program_id = paste0(trait_id, ":", smry$programs$program),
      status = ifelse(program_tier %in% c("high", "medium"), "valid", program_tier),
      stringsAsFactors = FALSE
    )
  ))
}

# Keep only the valid programs of an extraction.
valid_extraction <- function(extraction, valid_ids) {
  keep <- function(tab) {
    return(tab[tab$program_id %in% valid_ids, , drop = FALSE])
  }
  return(list(
    loadings = keep(extraction$loadings),
    profiles = keep(extraction$profiles),
    programs = keep(extraction$programs)
  ))
}

# For each planted program in one trait: found valid, found but filtered by
# validation, or not found; and, for the program that represents it (best
# valid match, else best match of any status), how closely its high-confidence
# SNPs and its loadings match the planted ones.
planted_outcomes <- function(mapping, status, gt, trait_id, loadings) {
  labels <- gt$programs$program_label[gt$programs$trait_id == as.integer(trait_id)]
  m <- dplyr::left_join(mapping, status, by = "program_id")
  trait_loadings <- loadings[loadings$trait_id == as.character(trait_id), , drop = FALSE]
  rows <- lapply(labels, function(lab) {
    # Planted labels are common to both traits ("shared1"), so a hit must also
    # be one of this trait's programs.
    hits <- m[
      !is.na(m$planted_label) & m$planted_label == lab &
        m$trait_id == as.character(trait_id), ,
      drop = FALSE
    ]
    valid_hits <- hits[hits$status %in% "valid", , drop = FALSE]
    outcome <- if (nrow(hits) == 0) {
      "not found"
    } else if (nrow(valid_hits) > 0) {
      "valid"
    } else {
      "filtered"
    }
    quality <- data.frame(jaccard = NA_real_, loading_cor = NA_real_)
    if (nrow(hits) > 0) {
      pool <- if (nrow(valid_hits) > 0) valid_hits else hits
      rep_id <- pool$program_id[which.max(pool$score)]
      prog <- trait_loadings[trait_loadings$program_id == rep_id, , drop = FALSE]
      planted <- gt$program_snps[[paste0(trait_id, ":", lab)]]
      resolved <- unique(prog$snp_id[prog$high_confidence %in% TRUE])
      effect <- gt$program_effects[[paste0(trait_id, ":", lab)]]
      truth <- ifelse(prog$snp_id %in% names(effect), effect[prog$snp_id], 0)
      quality <- data.frame(
        jaccard = length(intersect(resolved, planted)) /
          length(union(resolved, planted)),
        loading_cor = if (nrow(prog) > 2 && stats::sd(prog$loading) > 0) {
          stats::cor(prog$loading, truth)
        } else {
          NA_real_
        }
      )
    }
    return(data.frame(
      trait_id = as.character(trait_id), planted_label = lab,
      outcome = outcome, quality, stringsAsFactors = FALSE
    ))
  })
  return(dplyr::bind_rows(rows))
}

pair_columns <- c(
  "program_id_a", "program_id_b", "phi_traits", "n_eff_traits",
  "n_loci_shared", "n_eff_loci", "alignment", "q_concordance",
  "direction_profile", "tier", "linked"
)

run_replicate <- function(spec, seed) {
  sp <- do.call(
    simulate_trait_pair,
    c(utils::modifyList(generative_base, spec), list(seed = seed))
  )
  gt <- sp$ground_truth
  tids <- names(sp$traits)
  fits <- lapply(tids, function(tid) discover_one_trait(sp$traits[[tid]], tid))
  names(fits) <- tids

  status <- dplyr::bind_rows(lapply(fits, function(x) x$status))
  valid_ids <- status$program_id[status$status == "valid"]
  loadings_all <- dplyr::bind_rows(lapply(fits, function(x) x$all$loadings))
  mapping <- map_programs_to_planted(loadings_all, gt, by = "snps")

  direction_of <- stats::setNames(
    gt$correspondence$direction, gt$correspondence$program_label
  )
  discovery <- dplyr::bind_rows(lapply(tids, function(tid) {
    return(planted_outcomes(mapping, status, gt, tid, loadings_all))
  }))
  discovery$kind <- dplyr::coalesce(
    unname(direction_of[discovery$planted_label]), "trait-specific"
  )

  # Link the valid programs, as the real analysis does, and run module_rg() on
  # every pair (not only the linked ones) so Q3 can study rg on its own.
  # min_loci_rg = 2 keeps an estimate wherever one is possible; the cutoff is
  # applied afterwards.
  extraction <- lapply(fits, function(x) valid_extraction(x$all, valid_ids))
  n_valid <- vapply(extraction, function(x) {
    return(length(unique(x$loadings$program_id)))
  }, integer(1))
  pairs <- data.frame()
  if (all(n_valid > 0)) {
    cmp <- do.call(
      compare_program_pairs_loadings,
      c(
        list(program_data = extraction, n_perm = n_perm, seed = 1),
        current_cutoffs
      )
    )
    pairs <- cmp$pairs[, pair_columns, drop = FALSE]
    if (nrow(pairs) > 0) {
      rg <- module_rg(
        cmp, program_data = extraction,
        coloc_groups = stats::setNames(
          lapply(tids, function(tid) sp$traits[[tid]]$coloc_groups), tids
        ),
        linked_only = FALSE, min_loci_rg = 2L, seed = 1
      )
      pairs <- dplyr::left_join(
        pairs,
        rg[, c(
          "program_id_a", "program_id_b", "rg", "rg_ci_lower", "rg_ci_upper",
          "n_loci_rg", "rg_direction"
        )],
        by = c("program_id_a", "program_id_b")
      )
      label_of <- stats::setNames(mapping$planted_label, mapping$program_id)
      la <- unname(label_of[as.character(pairs$program_id_a)])
      lb <- unname(label_of[as.character(pairs$program_id_b)])
      pairs$true_pair <- !is.na(la) & !is.na(lb) & la == lb & la %in% names(direction_of)
      pairs$planted_label <- ifelse(pairs$true_pair, la, NA_character_)
      pairs$direction_planted <- unname(direction_of[pairs$planted_label])
    }
  }
  # One row per planted shared program, so a replicate in which no true pair
  # exists (a copy was not valid in its trait) still counts towards recall.
  shared_programs <- data.frame(
    planted_label = gt$correspondence$program_label,
    direction_planted = gt$correspondence$direction,
    stringsAsFactors = FALSE
  )
  return(list(
    discovery = discovery, pairs = pairs, shared_programs = shared_programs
  ))
}

# Run every version. The runner returns a named list of tables; each is stacked
# across replicates and versions and tagged with both. Errors are returned
# rather than recorded with `<<-`, so they survive a forked mclapply worker.
run_versions <- function(versions, runner) {
  n_jobs <- max(1L, min(as.integer(params$cores), n_sim))
  one <- function(spec, r) {
    err <- NULL
    out <- tryCatch(
      runner(spec, sim_seeds[[r]]),
      error = function(e) {
        err <<- conditionMessage(e)
        return(NULL)
      }
    )
    return(list(out = out, error = err, rep = r))
  }
  tables <- list()
  errors <- list()
  for (v in seq_along(versions)) {
    spec <- versions[[v]]$spec
    rep_out <- if (n_jobs > 1 && .Platform$OS.type != "windows") {
      parallel::mclapply(
        seq_len(n_sim), function(r) one(spec, r),
        mc.cores = n_jobs, mc.set.seed = FALSE
      )
    } else {
      lapply(seq_len(n_sim), function(r) one(spec, r))
    }
    for (x in rep_out) {
      if (!is.null(x$error)) {
        errors[[length(errors) + 1L]] <- data.frame(
          version = versions[[v]]$label, rep = x$rep, message = x$error,
          stringsAsFactors = FALSE
        )
      }
      for (nm in names(x$out)) {
        tab <- x$out[[nm]]
        if (is.null(tab) || nrow(tab) == 0) {
          next
        }
        tab$rep <- x$rep
        tab$version <- versions[[v]]$label
        tab$family <- versions[[v]]$family
        tab$snp_overlap <- versions[[v]]$snp_overlap
        tab$driver_overlap <- versions[[v]]$driver_overlap
        tables[[nm]][[length(tables[[nm]]) + 1L]] <- tab
      }
    }
  }
  out <- lapply(tables, dplyr::bind_rows)
  out$errors <- dplyr::bind_rows(errors)
  return(out)
}

mc_summary <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) {
    return(list(est = NA_real_, se = NA_real_, n = 0L))
  }
  return(list(
    est = mean(x),
    se = if (length(x) > 1) stats::sd(x) / sqrt(length(x)) else NA_real_,
    n = length(x)
  ))
}

fmt_mc <- function(x, digits = 2) {
  s <- mc_summary(x)
  if (!is.finite(s$est)) {
    return("--")
  }
  if (!is.finite(s$se)) {
    return(formatC(s$est, format = "f", digits = digits))
  }
  return(sprintf(
    "%s (%s)", formatC(s$est, format = "f", digits = digits),
    formatC(s$se, format = "f", digits = digits)
  ))
}

pct <- function(x) {
  return(sprintf("%.0f%%", 100 * mean(x %in% TRUE)))
}

print_errors <- function(err) {
  if (is.null(err) || nrow(err) == 0) {
    return(invisible(NULL))
  }
  print(knitr::kable(
    err |>
      dplyr::group_by(version, message) |>
      dplyr::summarise(n_failed = dplyr::n(), .groups = "drop"),
    row.names = FALSE,
    caption = paste0(
      "Replicates that failed. Any row here means the version is reported on ",
      "fewer replicates than n_sim, or not at all."
    )
  ))
  return(invisible(NULL))
}
```

```{r scoring-helpers}
# Re-tier every stored pair at a set of cutoffs, with the pipeline's own tier
# rule, so any cutoff can be scored without refitting.
retier <- function(pairs, cutoffs) {
  if (nrow(pairs) == 0) {
    return(pairs)
  }
  return(gpmapr:::.assign_link_tiers(
    pairs,
    phi_high = cutoffs$phi_high, phi_low = cutoffs$phi_low,
    min_shared_loci = cutoffs$min_shared_loci,
    min_eff_loci = cutoffs$min_eff_loci,
    min_alignment = cutoffs$min_alignment,
    fdr_threshold = cutoffs$fdr_threshold
  ))
}

# The best true pair of each planted shared program in each replicate (highest
# tier, then largest |phi|), with "no valid pair" when no true pair exists.
tier_levels <- c("high", "medium", "low", "none")
best_true_pairs <- function(pairs, shared_programs) {
  keys <- c("version", "rep", "planted_label")
  best <- pairs[pairs$true_pair, , drop = FALSE] |>
    dplyr::mutate(tier_rank = match(tier, tier_levels)) |>
    dplyr::group_by(dplyr::across(dplyr::all_of(keys))) |>
    dplyr::arrange(tier_rank, dplyr::desc(abs(phi_traits)), .by_group = TRUE) |>
    dplyr::slice(1) |>
    dplyr::ungroup() |>
    dplyr::select(-tier_rank, -direction_planted, -snp_overlap, -driver_overlap, -family)
  out <- dplyr::left_join(shared_programs, best, by = keys)
  out$tier <- dplyr::coalesce(out$tier, "no valid pair")
  out$linked <- out$tier %in% c("high", "medium")
  return(out)
}

# Recall per grid cell and planted direction, false links per replicate in the
# null and shared runs, and pooled precision, at one set of cutoffs.
score_cutoffs <- function(pairs, shared_programs, cutoffs) {
  tiered <- retier(pairs, cutoffs)
  best <- best_true_pairs(tiered, shared_programs)
  linked <- tiered[tiered$linked, , drop = FALSE]
  false_per_rep <- function(fam) {
    reps <- unique(all_reps[all_reps$family == fam, c("version", "rep")])
    n_false <- linked[linked$family == fam & !linked$true_pair, , drop = FALSE] |>
      dplyr::count(version, rep, name = "n_false")
    counts <- dplyr::left_join(reps, n_false, by = c("version", "rep"))
    return(dplyr::coalesce(counts$n_false, 0L))
  }
  shared_linked <- linked[linked$family == "shared", , drop = FALSE]
  return(list(
    best = best,
    recall = mean(best$linked),
    false_null = false_per_rep("null"),
    false_shared = false_per_rep("shared"),
    precision = if (nrow(shared_linked) > 0) mean(shared_linked$true_pair) else NA_real_
  ))
}
```

```{r run}
t_start <- Sys.time()
sim <- run_versions(versions, run_replicate)
run_minutes <- as.numeric(difftime(Sys.time(), t_start, units = "mins"))
empty_if_null <- function(x) {
  return(if (is.null(x)) data.frame() else x)
}
discovery <- empty_if_null(sim$discovery)
pairs <- empty_if_null(sim$pairs)
shared_programs <- empty_if_null(sim$shared_programs)
# Every completed replicate, including those that produced no pair at all.
all_reps <- dplyr::distinct(discovery, version, rep, family)
if (nrow(pairs) > 0) {
  # The offline re-tiering must reproduce the pipeline's own links exactly, or
  # nothing in Q3 can be trusted.
  stopifnot(identical(retier(pairs, current_cutoffs)$linked, pairs$linked))
}
current <- if (nrow(pairs) > 0) {
  score_cutoffs(pairs, shared_programs, current_cutoffs)
} else {
  NULL
}
```

```{r errors, results='asis'}
print_errors(sim$errors)
cat(sprintf(
  "\nRun time: %.1f minutes for %d versions x %d replicates (%.1f s per replicate).\n\n",
  run_minutes, length(versions), n_sim,
  60 * run_minutes / max(1, length(versions) * n_sim)
))
```

## Q1. Are planted programs recovered?

**Within each trait.** For every planted program, the share of
trait-replicates in which a **valid** program represents it, in which the best
match was **filtered** out by validation, and in which no fitted program
matches it at all (**not found**). `jaccard` and `loading_cor` describe the
representing program: overlap of its high-confidence SNPs with the planted
SNPs, and correlation of its loadings with the planted per-locus effects.

```{r q1-discovery, results='asis'}
if (nrow(discovery) == 0) {
  cat("No replicates completed.\n\n")
} else {
  kind_levels <- c("concordant", "antagonistic", "trait-specific")
  print(knitr::kable(
    discovery |>
      dplyr::mutate(kind = factor(kind, levels = kind_levels)) |>
      dplyr::group_by(planted = kind) |>
      dplyr::summarise(
        n = dplyr::n(),
        valid = pct(outcome == "valid"),
        filtered = pct(outcome == "filtered"),
        not_found = pct(outcome == "not found"),
        jaccard = fmt_mc(jaccard),
        loading_cor = fmt_mc(loading_cor),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "Recovery of each planted program in its own trait, pooled over the ",
      "grid, both traits and replicates (n = program-trait-replicates). ",
      "Jaccard and loading_cor are means (Monte Carlo SE) over replicates in ",
      "which a fitted program matched."
    )
  ))
  by_cell <- discovery[discovery$family == "shared" & discovery$kind != "trait-specific", ] |>
    dplyr::group_by(version) |>
    dplyr::summarise(valid = mean(outcome == "valid"), .groups = "drop")
  cat(sprintf(
    "\nThe shared programs are valid in %.0f%%-%.0f%% of trait-replicates across the %d grid cells.\n\n",
    100 * min(by_cell$valid), 100 * max(by_cell$valid), nrow(by_cell)
  ))
}
```

Each trait is fitted on its own matrix, so recovery here should not depend on
the overlap with the partner trait: a wide range across grid cells points to a
problem in the simulation, not the fit. A shared program that is not valid in
both traits cannot be linked, so this is the ceiling on recall below.

**Across traits, when the copies are identical.** At full overlap (all 36 loci
and all 8 drivers shared), the share of replicates in which the shared
program is linked, the tier of its best true pair, and whether the link has
the right direction -- from the sign of φ, and from rg where rg calls one
(at least `min_loci_rg` = `r params$min_loci_rg` loci and a CI excluding 0).

```{r q1-links, results='asis'}
full_label <- sprintf("snp %.2f, drivers %.2f", 1, 1)
if (is.null(current) || !full_label %in% current$best$version) {
  cat("No completed replicates at full overlap.\n\n")
} else {
  full <- current$best[current$best$version == full_label, , drop = FALSE] |>
    dplyr::mutate(
      phi_ok = direction_profile == direction_planted,
      rg_call = is.finite(n_loci_rg) & n_loci_rg >= params$min_loci_rg &
        rg_direction %in% c("concordant", "antagonistic"),
      rg_ok = rg_direction == direction_planted
    )
  print(knitr::kable(
    full |>
      dplyr::group_by(direction = direction_planted) |>
      dplyr::summarise(
        recall = fmt_mc(as.numeric(linked)),
        high = pct(tier == "high"),
        medium = pct(tier == "medium"),
        `low/none` = pct(tier %in% c("low", "none")),
        no_valid_pair = pct(tier == "no valid pair"),
        phi_sign_correct = fmt_mc(as.numeric(phi_ok[linked])),
        rg_called = pct(rg_call[linked]),
        rg_correct = fmt_mc(as.numeric(rg_ok[rg_call & linked])),
        .groups = "drop"
      ),
    row.names = FALSE,
    caption = paste0(
      "The shared programs at full overlap, at the current cutoffs. Tier ",
      "columns are the share of replicates whose best true pair is in each ",
      "tier. Direction accuracy is over linked true pairs; rg_called is the ",
      "share of them in which rg calls a direction."
    )
  ))
  shared_full_links <- retier(pairs, current_cutoffs) |>
    dplyr::filter(version == full_label, linked)
  cat(sprintf(
    paste0(
      "\nPrecision at full overlap: %s of %d links are true. False links per ",
      "replicate in the null: %s, in %d of %d replicates with any.\n\n"
    ),
    pct(shared_full_links$true_pair), nrow(shared_full_links),
    fmt_mc(current$false_null), sum(current$false_null > 0),
    length(current$false_null)
  ))
}
```

**How to read it.** At full overlap the true pair should be **high** tier
(strong φ, loci agree) in nearly every replicate, with φ's sign and rg both
reading the planted direction. Any **no valid pair** goes back to the
within-trait table above.

## Q2. How much do the copies need to have in common?

**What these figures show.** Recall of the shared program at the current
cutoffs over every combination of driver overlap and SNP overlap, then which
tier the true pair lands in, then the two scores that decide the tier.

```{r q2-recall-heatmap, fig.width = 9, fig.height = 4.5}
recall_heatmap <- function(best, title) {
  cell <- best |>
    dplyr::filter(family == "shared") |>
    dplyr::group_by(snp_overlap, driver_overlap, direction_planted) |>
    dplyr::summarise(recall = mean(linked), .groups = "drop") |>
    dplyr::mutate(direction_planted = factor(direction_planted, levels = names(palette_direction)))
  return(
    ggplot(cell, aes(factor(snp_overlap), factor(driver_overlap), fill = recall)) +
      geom_tile(colour = "white") +
      geom_text(aes(label = sprintf("%.2f", recall), colour = recall > 0.6), size = 3.2) +
      facet_wrap(~direction_planted) +
      scale_fill_gradient(low = "#f4f1ea", high = "#1a6b3c", limits = c(0, 1)) +
      scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20"), guide = "none") +
      labs(
        title = title,
        x = "SNP overlap (share of the 36 loci common to both copies)",
        y = "driver overlap (share of 8 drivers)", fill = "recall"
      ) +
      theme_minimal() +
      theme(panel.grid = element_blank())
  )
}
if (!is.null(current)) {
  print(recall_heatmap(current$best, "Recall of the shared program, current cutoffs"))
}
```

```{r q2-tiers, fig.width = 10, fig.height = 7}
if (!is.null(current)) {
  tiers <- current$best |>
    dplyr::filter(family == "shared") |>
    dplyr::count(snp_overlap, driver_overlap, direction_planted, tier) |>
    dplyr::group_by(snp_overlap, driver_overlap, direction_planted) |>
    dplyr::mutate(share = n / sum(n)) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      tier = factor(tier, levels = names(palette_tier)),
      driver_overlap = paste("drivers", driver_overlap)
    )
  print(
    ggplot(tiers, aes(factor(snp_overlap), share, fill = tier)) +
      geom_col(width = 0.8) +
      facet_grid(direction_planted ~ driver_overlap) +
      scale_fill_manual(values = palette_tier) +
      labs(
        title = "Tier of the shared program's best true pair",
        x = "SNP overlap", y = "share of replicates", fill = NULL
      ) +
      theme_minimal() +
      theme(legend.position = "top", panel.grid.major.x = element_blank())
  )
}
```

```{r q2-scores, fig.width = 9, fig.height = 3.5}
if (!is.null(current)) {
  scores <- current$best |>
    dplyr::filter(family == "shared", tier != "no valid pair")
  score_plot <- function(df, x, colour, y, cuts, title, x_lab, colour_lab) {
    med <- df |>
      dplyr::group_by(x = .data[[x]], colour = factor(.data[[colour]])) |>
      dplyr::summarise(value = stats::median(.data[[y]], na.rm = TRUE), .groups = "drop") |>
      dplyr::filter(is.finite(value))
    return(
      ggplot(med, aes(x, value, colour = colour)) +
        geom_hline(yintercept = cuts, linetype = "dashed", colour = "#b2182b") +
        geom_line(linewidth = 0.7) +
        geom_point(size = 1.8) +
        scale_colour_viridis_d(end = 0.9) +
        scale_y_continuous(limits = c(0, 1)) +
        labs(title = title, x = x_lab, y = NULL, colour = colour_lab) +
        theme_minimal()
    )
  }
  scores$abs_phi <- abs(scores$phi_traits)
  print(score_plot(
    scores, "driver_overlap", "snp_overlap", "abs_phi",
    c(current_cutoffs$phi_high, current_cutoffs$phi_low),
    "Median |phi| of the best true pair (dashed: phi_high, phi_low)",
    "driver overlap", "SNP overlap"
  ))
  print(score_plot(
    scores, "snp_overlap", "driver_overlap", "alignment",
    current_cutoffs$min_alignment,
    "Median alignment of the best true pair (dashed: min_alignment; missing where too few loci are shared)",
    "SNP overlap", "driver overlap"
  ))
}
```

```{r q2-breakpoints, results='asis'}
if (!is.null(current)) {
  cell <- current$best |>
    dplyr::filter(family == "shared") |>
    dplyr::group_by(snp_overlap, driver_overlap) |>
    dplyr::summarise(recall = mean(linked), .groups = "drop")
  first_at <- function(x, ok) {
    return(if (any(ok)) as.character(min(x[ok])) else "never")
  }
  print(knitr::kable(
    dplyr::bind_rows(
      cell |>
        dplyr::group_by(fixed = "driver overlap", at = driver_overlap) |>
        dplyr::summarise(
          varied = "SNP overlap", needed = first_at(snp_overlap, recall >= 0.8),
          .groups = "drop"
        ),
      cell |>
        dplyr::group_by(fixed = "SNP overlap", at = snp_overlap) |>
        dplyr::summarise(
          varied = "driver overlap", needed = first_at(driver_overlap, recall >= 0.8),
          .groups = "drop"
        )
    ),
    row.names = FALSE,
    caption = paste0(
      "Holding one overlap fixed at `at`, the smallest value of the other ",
      "(`varied`) at which recall reaches 0.8, both directions pooled; ",
      "'never' when no value in the grid does."
    )
  ))
}
```

**How to read it.**

- **The φ axis can link on its own.** With no loci in common the loci are
  `too_few_loci`, so a strong φ alone gives a **medium** link; recall along the
  SNP overlap = 0 column is driver overlap working alone.
- **Loci alone cannot.** With no drivers in common φ is weak, and a weak φ
  whose loci agree is only **low** -- not a link -- however many loci are
  shared. The drivers 0 row should be near 0 at every SNP overlap.
- **Partial SNP overlap can hurt.** A few shared loci make the loci testable;
  if they then fail `min_alignment` or the permutation test, a strong-φ pair
  falls from medium (`too_few_loci`) to low (`disagree`). A dip in the middle
  of a row is this.
- The breakpoint table is the summary answer to Q2.

## Q3. What cutoffs?

**What these figures show.** The distribution of |φ| and of alignment for
true pairs (the two copies of a shared program, over the whole grid) and false
pairs (everything else, including every pair in the null), with the current
cutoffs dashed. A good cutoff sits where the false pairs end.

```{r q3-distributions, fig.width = 10, fig.height = 4.5}
if (nrow(pairs) > 0) {
  dist_df <- dplyr::bind_rows(
    pairs |>
      dplyr::transmute(metric = "|phi|", value = abs(phi_traits), true_pair),
    pairs |>
      dplyr::filter(
        n_loci_shared >= current_cutoffs$min_shared_loci,
        n_eff_loci >= current_cutoffs$min_eff_loci
      ) |>
      dplyr::transmute(metric = "alignment (testable pairs)", value = alignment, true_pair)
  ) |>
    dplyr::filter(is.finite(value)) |>
    dplyr::mutate(pair = ifelse(true_pair, "true pair", "false pair"))
  cut_df <- data.frame(
    metric = c("|phi|", "|phi|", "alignment (testable pairs)"),
    value = c(current_cutoffs$phi_high, current_cutoffs$phi_low, current_cutoffs$min_alignment)
  )
  print(
    ggplot(dist_df, aes(value, colour = pair)) +
      stat_ecdf(linewidth = 0.8) +
      geom_vline(data = cut_df, aes(xintercept = value), linetype = "dashed", colour = "#b2182b") +
      facet_wrap(~metric) +
      scale_colour_manual(values = c(`true pair` = "#1a6b3c", `false pair` = "#8c8c8c")) +
      labs(
        title = "Scores of true and false pairs",
        subtitle = sprintf(
          "%d true and %d false pairs, pooled over the grid and the null.",
          sum(pairs$true_pair), sum(!pairs$true_pair)
        ),
        x = NULL, y = "cumulative share of pairs", colour = NULL
      ) +
      theme_minimal() +
      theme(legend.position = "top")
  )
}
```

**The sweep.** Every stored pair is re-tiered at each combination of
`phi_high` (0.50 to 0.95), `phi_low` (`phi_high` minus 0.10, 0.15 or 0.20)
and `min_alignment` (0.1 to 0.9). A combination is **acceptable** when false
links per null replicate average at most `r params$max_false_links` and
precision in the shared runs is at least `r params$min_precision`. The
recommended combination is the acceptable one with the highest mean recall
over the grid; ties go to the stricter one.

```{r q3-sweep, results='asis'}
sweep <- data.frame()
recommended <- NULL
if (nrow(pairs) > 0) {
  grid <- expand.grid(
    phi_high = seq(0.50, 0.95, by = 0.05),
    phi_gap = c(0.10, 0.15, 0.20),
    min_alignment = seq(0.1, 0.9, by = 0.1)
  )
  sweep <- dplyr::bind_rows(lapply(seq_len(nrow(grid)), function(i) {
    cutoffs <- utils::modifyList(current_cutoffs, list(
      phi_high = grid$phi_high[i],
      phi_low = grid$phi_high[i] - grid$phi_gap[i],
      min_alignment = grid$min_alignment[i]
    ))
    s <- score_cutoffs(pairs, shared_programs, cutoffs)
    return(data.frame(
      phi_high = cutoffs$phi_high, phi_low = cutoffs$phi_low,
      min_alignment = cutoffs$min_alignment,
      recall = s$recall,
      false_null = mean(s$false_null),
      false_shared = mean(s$false_shared),
      precision = s$precision
    ))
  }))
  sweep$acceptable <- sweep$false_null <= params$max_false_links &
    dplyr::coalesce(sweep$precision, 1) >= params$min_precision
  # Highest recall first; among equal recall, the stricter setting first.
  sweep <- sweep[order(
    -sweep$recall, -sweep$phi_high, -sweep$phi_low, -sweep$min_alignment
  ), , drop = FALSE]
  is_current <- abs(sweep$phi_high - current_cutoffs$phi_high) < 1e-9 &
    abs(sweep$phi_low - current_cutoffs$phi_low) < 1e-9 &
    abs(sweep$min_alignment - current_cutoffs$min_alignment) < 1e-9
  ok <- sweep[sweep$acceptable, , drop = FALSE]
  if (nrow(ok) > 0) {
    recommended <- utils::modifyList(current_cutoffs, as.list(
      ok[1, c("phi_high", "phi_low", "min_alignment")]
    ))
  }
  is_recommended <- if (is.null(recommended)) {
    rep(FALSE, nrow(sweep))
  } else {
    abs(sweep$phi_high - recommended$phi_high) < 1e-9 &
      abs(sweep$phi_low - recommended$phi_low) < 1e-9 &
      abs(sweep$min_alignment - recommended$min_alignment) < 1e-9
  }
  show <- sweep |>
    dplyr::mutate(setting = dplyr::case_when(
      is_current & is_recommended ~ "current = recommended",
      is_current ~ "current",
      is_recommended ~ "recommended",
      TRUE ~ ""
    )) |>
    dplyr::filter(acceptable | setting != "") |>
    dplyr::filter(dplyr::row_number() <= 15 | setting != "") |>
    dplyr::mutate(dplyr::across(c(recall, false_null, false_shared, precision), ~ round(.x, 3)))
  print(knitr::kable(
    show, row.names = FALSE,
    caption = paste0(
      "Top acceptable cutoff combinations by mean recall over the grid (up to ",
      "15; equal recall ordered strictest first), plus the current setting. false_null and false_shared are mean ",
      "false links per replicate in the null and shared runs. ",
      sum(sweep$acceptable), " of ", nrow(sweep), " combinations are acceptable."
    )
  ))
  if (is.null(recommended)) {
    cat("\n**No combination meets both targets**; no cutoffs are recommended.\n\n")
  }
}
```

```{r q3-sweep-plot, fig.width = 10, fig.height = 4.5}
if (nrow(sweep) > 0) {
  best_gap <- sweep |>
    dplyr::group_by(phi_high, min_alignment) |>
    dplyr::slice_max(recall, n = 1, with_ties = FALSE) |>
    dplyr::ungroup()
  print(
    ggplot(best_gap, aes(factor(phi_high), factor(min_alignment), fill = recall)) +
      geom_tile(colour = "white") +
      geom_point(data = best_gap[!best_gap$acceptable, ], shape = 4, size = 2, colour = "#b2182b") +
      scale_fill_gradient(low = "#f4f1ea", high = "#1a6b3c", limits = c(0, 1)) +
      labs(
        title = "Mean recall over the grid by phi_high and min_alignment",
        subtitle = "Best phi_low gap per cell. x: fails the false-link or precision target.",
        x = "phi_high", y = "min_alignment", fill = "recall"
      ) +
      theme_minimal() +
      theme(panel.grid = element_blank())
  )
}
```

**rg.** rg never forms a link; it checks the direction of one. The question is
how many shared loci it needs before its call can be trusted. For every true
pair, by the number of loci rg was estimated over: how often its CI excludes 0
(a call), and how often that call is the planted direction.

```{r q3-rg, results='asis'}
recommended_rg <- NA_integer_
if (nrow(pairs) > 0) {
  rg_true <- pairs[pairs$true_pair & is.finite(pairs$rg), , drop = FALSE]
  if (nrow(rg_true) > 0) {
    breaks <- c(2, 5, 10, 15, 20, 30, Inf)
    rg_true$bin <- cut(rg_true$n_loci_rg, breaks = breaks, right = FALSE)
    rg_true$called <- rg_true$rg_direction %in% c("concordant", "antagonistic")
    rg_true$correct <- rg_true$rg_direction == rg_true$direction_planted
    rg_tbl <- rg_true |>
      dplyr::group_by(n_loci_rg = bin) |>
      dplyr::summarise(
        n_pairs = dplyr::n(),
        call_rate = pct(called),
        accuracy = if (any(called)) mean(correct[called]) else NA_real_,
        accuracy_concordant = if (any(called & direction_planted == "concordant")) {
          mean(correct[called & direction_planted == "concordant"])
        } else {
          NA_real_
        },
        accuracy_antagonistic = if (any(called & direction_planted == "antagonistic")) {
          mean(correct[called & direction_planted == "antagonistic"])
        } else {
          NA_real_
        },
        .groups = "drop"
      )
    # Smallest bin from which every larger bin is accurate enough.
    good <- rev(cumprod(rev(dplyr::coalesce(rg_tbl$accuracy, 0) >= params$min_rg_accuracy)))
    if (any(good == 1)) {
      first_bin <- as.character(rg_tbl$n_loci_rg[which(good == 1)[1]])
      recommended_rg <- as.integer(breaks[match(first_bin, levels(rg_true$bin))])
    }
    print(knitr::kable(
      rg_tbl |>
        dplyr::mutate(dplyr::across(dplyr::starts_with("accuracy"), ~ round(.x, 3))),
      row.names = FALSE,
      caption = paste0(
        "rg on true pairs, by usable loci (n_loci_rg). called = CI excludes 0; ",
        "accuracy = share of calls in the planted direction."
      )
    ))
  }
  rg_false <- pairs[
    !pairs$true_pair & is.finite(pairs$rg) & pairs$n_loci_rg >= params$min_loci_rg, ,
    drop = FALSE
  ]
  linked_true <- pairs[pairs$true_pair & pairs$linked, , drop = FALSE]
  conflict <- linked_true$rg_direction %in% c("concordant", "antagonistic") &
    linked_true$n_loci_rg >= params$min_loci_rg &
    linked_true$rg_direction != linked_true$direction_profile
  direction_msg <- paste0(
    "\nOn linked true pairs, the sign of phi gives the planted direction in %s ",
    "(%d pairs); rg (at least %d loci) contradicts phi in %s. ",
    "%d false pairs had an rg estimate over at least %d loci.\n\n"
  )
  cat(sprintf(
    direction_msg,
    pct(linked_true$direction_profile == linked_true$direction_planted),
    nrow(linked_true), as.integer(params$min_loci_rg), pct(conflict),
    nrow(rg_false), as.integer(params$min_loci_rg)
  ))
}
```

**Recommended against current.**

```{r q3-recommended, results='asis'}
if (!is.null(current)) {
  rows <- list(current = current_cutoffs)
  if (!is.null(recommended)) {
    rows$recommended <- recommended
  }
  print(knitr::kable(
    dplyr::bind_rows(lapply(names(rows), function(nm) {
      s <- score_cutoffs(pairs, shared_programs, rows[[nm]])
      return(data.frame(
        setting = nm,
        phi_high = rows[[nm]]$phi_high,
        phi_low = rows[[nm]]$phi_low,
        min_alignment = rows[[nm]]$min_alignment,
        min_loci_rg = if (nm == "current") as.integer(params$min_loci_rg) else recommended_rg,
        recall = fmt_mc(as.numeric(s$best$linked)),
        false_null = fmt_mc(s$false_null),
        false_shared = fmt_mc(s$false_shared),
        precision = round(s$precision, 3)
      ))
    })),
    row.names = FALSE,
    caption = paste0(
      "Current and recommended cutoffs. recall is over every shared program in ",
      "the grid; false_null / false_shared are false links per replicate. ",
      "The recommended min_loci_rg is the smallest n_loci_rg bin from which rg's ",
      "calls are at least ", params$min_rg_accuracy, " accurate (NA when none is)."
    )
  ))
}
```

```{r q3-recommended-heatmap, fig.width = 9, fig.height = 4.5}
if (!is.null(recommended)) {
  rec <- score_cutoffs(pairs, shared_programs, recommended)
  print(recall_heatmap(rec$best, sprintf(
    "Recall at the recommended cutoffs (phi_high %.2f, phi_low %.2f, min_alignment %.1f)",
    recommended$phi_high, recommended$phi_low, recommended$min_alignment
  )))
}
```

**How to read it.** The recommendation is only as good as its negatives.
Compare the recommended heatmap with the one in Q2: the cells that gain
recall are the overlaps the current cutoffs are too strict for. A
recommended `phi_high` well below the Tucker conventions behind the current
0.85 is a statement about these simulated negatives, which are easy (see
Notes); treat it as the least strict cutoff that could be justified, not the
one to adopt unexamined.

## Notes

- **The false pairs are easy.** Different planted programs share neither
  drivers nor loci here, so the false-link rate is optimistic. Real data has
  pleiotropic hub loci and correlated background studies that many programs
  touch; run the realism arm of the grid script before quoting a false-link
  rate.
- **rg is a direction check only.** False pairs here rarely share enough loci
  for an rg estimate (the count is given under Q3), so rg is not calibrated as
  a link filter.
- **Sample overlap is not simulated.** Two target traits from the same cohort
  share participants, which biases `module_rg()` away from the null.
- **Background realism is off by default** (`background_corr` and
  `snp_pleiotropy_sd` are 0).
- **Not comparable with earlier renders.** Earlier versions linked on loci
  only, planted one shared program per replicate with every driver shared, and
  left unshared drivers out of the partner's matrix.
