---
title: "Colocalisation consistency, latent blocks, and pathway enrichment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colocalisation consistency, latent blocks, and pathway enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
library(gpmapr)
library(dplyr)
library(tidyr)
```

This vignette walks through three complementary layers of colocalisation analysis in **gpmapr**:

1. **Pleiotropic Consistency Score (PCS)** and **Net Directionality Score (NDS)** — quantify locus overlap and effect concordance between traits.
2. **Latent causal blocks** — cluster loci with similar multi-trait effect profiles using Latent Profile Analysis (LPA).
3. **Pathway enrichment** — test whether genes mapped to those loci are enriched for biological pathways, and compare LPA-refined gene sets with broader colocalisation blocks.

We use **LDL cholesterol** (trait id `1486`) and **HDL cholesterol levels** (trait id `1630`) as an illustrative pair of lipid traits that share some, but not all, causal loci.

## Fetch colocalisation data

We fetch both traits in a single call and retain association statistics (`beta`, `se`) for directionality and z-score construction.

```{r fetch-data}
ldl_id <- 1486
hdl_id <- 1630

lipid_traits <- traits(
  c(ldl_id, hdl_id),
  include_associations = TRUE
)
cg <- lipid_traits$coloc_groups
```

## Pleiotropic consistency (PCS)

`pleiotropic_consistency_score()` measures how much two traits colocalise at the same LD blocks versus having trait-specific signals. The score is a Jaccard-style ratio:

$$\text{PCS} = \frac{n}{n + m_1 + m_2}$$

where $n$ is the number of shared loci, $m_1$ are loci unique to trait 1, and $m_2$ are loci unique to trait 2.

```{r pcs}
pcs_result <- pleiotropic_consistency_score(
  ldl_id, hdl_id,
  coloc_groups = cg
)

pcs_result
```

LDL and HDL share a modest fraction of colocalising loci: most signals are trait-specific. This low PCS motivates splitting loci into blocks before interpreting shared biology.

`pcs_matrix()` extends this to all pairs when comparing more than two traits:

```{r pcs-matrix}
pcs_mat <- pcs_matrix(c(ldl_id, hdl_id), coloc_groups = cg)
pcs_mat
```

## Net directionality (NDS)

At loci where both traits colocalise, `net_directionality_score()` asks whether effect directions agree:

$$\text{NDS} = \frac{n_+ - n_-}{n_+ + n_-}$$

```{r nds}
nds_result <- net_directionality_score(
  ldl_id, hdl_id,
  coloc_groups = cg
)

nds_result
```

A positive NDS means concordant effect directions predominate at shared loci (here, slightly more concordant than discordant pairs).

## Latent causal blocks

PCS splits loci by *presence* of colocalisation per trait. **Latent Profile Analysis** instead groups loci by their *multi-trait effect profile* — the pattern of z-scores across traits at each locus.

### Build the z-score matrix

`build_locus_zscore_matrix()` constructs a loci x traits matrix of z-scores (`beta / se`). Loci without a genome-wide signal for a trait are set to 0. Only loci with signals in at least two traits are retained.

```{r z-matrix}
z_data <- build_locus_zscore_matrix(
  c(ldl_id, hdl_id),
  coloc_groups = cg
)

dim(z_data$z_matrix)
```

### Fit LPA profiles

`latent_causal_blocks()` fits LPA models via **tidyLPA** and returns profile assignments for each locus. We search over 2-4 profiles and use model 1 (equal variances, zero covariances) for speed.

```{r lpa}
lpa_result <- latent_causal_blocks(
  z_data$z_matrix
)

lpa_result$best_model
table(lpa_result$assignments$profile)
```

### Summarise profile biology

`summarise_latent_blocks()` reports mean z-scores per trait within each profile — the signature of each latent block.

```{r summarise-blocks}
block_summary <- summarise_latent_blocks(
  z_data$z_matrix,
  lpa_result$assignments,
  trait_info = z_data$trait_info
)

knitr::kable(
  block_summary |>
    dplyr::select(profile, trait_name, mean_z, sd_z, n_loci) |>
    dplyr::arrange(profile, trait_name),
  digits = 2
)
```

Profile 1 loci tend toward positive LDL and HDL z-scores; profile 2 loci tend toward negative values for both traits. Profile 3 contains only a handful of loci with strong negative effects.

## Pathway enrichment comparison

`pathway_enrichment()` tests a gene set against Reactome, KEGG, and HP pathway databases via the GPMap API. We use numeric `gene_id` values here, but the API also accepts gene names.

Why can this be useful?

* We are asking slightly different questions here.
  * Among my input genes, is overlap with this pathway larger than expected by chance, given pathway size and background?
* it can provide different results due to:
  * a. how many genes you are testing against
  * b. which genes you are testing
* In a big mixed list, many genes hit many pathways weakly
* In a small coherent LPA list, overlap with the "right" pathway can be a higher fraction of the input set
  * That can produce stronger p-values for those specific pathways

```{r gene-sets}
locus_genes <- cg |>
  dplyr::distinct(coloc_group_id, gene_id, gene) |>
  dplyr::filter(!is.na(gene_id))

assignments <- lpa_result$assignments |>
  dplyr::mutate(coloc_group_id = as.integer(locus_id))

genes_for_loci <- function(locus_ids) {
  unique(locus_genes$gene_id[locus_genes$coloc_group_id %in% locus_ids])
}

ldl_loci <- cg |>
  dplyr::filter(trait_id == ldl_id, min_p <= 5e-8) |>
  dplyr::distinct(coloc_group_id) |>
  dplyr::pull(coloc_group_id)

hdl_loci <- cg |>
  dplyr::filter(trait_id == hdl_id, min_p <= 5e-8) |>
  dplyr::distinct(coloc_group_id) |>
  dplyr::pull(coloc_group_id)

ldl_only_loci <- setdiff(ldl_loci, hdl_loci)
hdl_only_loci <- setdiff(hdl_loci, ldl_loci)

lpa_genes <- genes_for_loci(assignments$coloc_group_id)
ldl_only_genes <- genes_for_loci(ldl_only_loci)
hdl_only_genes <- genes_for_loci(hdl_only_loci)

gene_sets <- list(
  all_ldl = genes_for_loci(ldl_loci),
  all_hdl = genes_for_loci(hdl_loci),
  ldl_only_minus_lpa = setdiff(ldl_only_genes, lpa_genes),
  hdl_only_minus_lpa = setdiff(hdl_only_genes, lpa_genes)
)

for (profile_id in sort(unique(assignments$profile))) {
  loci <- assignments$coloc_group_id[assignments$profile == profile_id]
  gene_sets[[paste0("lpa_profile_", profile_id)]] <- genes_for_loci(loci)
}

gene_set_sizes <- data.frame(
  gene_set = names(gene_sets),
  n_genes = vapply(gene_sets, length, integer(1)),
  row.names = NULL
)
knitr::kable(gene_set_sizes)
```

```{r pathway-enrichment}
run_pathway_enrichment <- function(gene_ids, label) {
  if (length(gene_ids) == 0) {
    return(NULL)
  }
  n_input_genes <- length(gene_ids)
  pe <- pathway_enrichment(
    gene_ids,
    p_value_threshold = 0.05,
    minimum_count_in_network = 2
  )
  results <- pe$results
  if (is.null(results) || !is.data.frame(results) || nrow(results) == 0) {
    return(NULL)
  }
  results |>
    dplyr::mutate(
      gene_set = label,
      n_genes = n_input_genes,
      n_pathway_genes = vapply(pathway_gene_ids, length, integer(1)),
      pathway_key = paste(source, term_id, sep = ":")
    )
}

pathway_results_full <- dplyr::bind_rows(
  lapply(names(gene_sets), function(nm) {
    run_pathway_enrichment(gene_sets[[nm]], nm)
  })
)

if (nrow(pathway_results_full) == 0) {
  stop(
    "No pathway enrichment results were returned. ",
    "Reinstall or reload gpmapr (devtools::load_all()) so the API uses the 'genes' parameter."
  )
}

pathway_comparison <- pathway_results_full |>
  dplyr::group_by(gene_set) |>
  dplyr::slice_min(fdr, n = 5, with_ties = FALSE) |>
  dplyr::ungroup()

knitr::kable(
  pathway_comparison |>
    dplyr::arrange(gene_set, fdr) |>
    dplyr::mutate(fdr = sprintf("%.1e", fdr)) |>
    dplyr::select(gene_set, n_genes, source, description, overlap, n_pathway_genes, fdr)
)
```

### All LDL vs all HDL, and LPA novelty

Finally, we enrich all genome-wide colocalisation genes for each trait and ask whether LPA profiles recover pathways that the full trait gene sets do not.

```{r pathway-novelty}
lpa_pathways <- pathway_results_full |>
  dplyr::filter(grepl("^lpa_profile_", gene_set)) |>
  dplyr::distinct(pathway_key, source, term_id, description)

all_ldl_pathways <- pathway_results_full |>
  dplyr::filter(gene_set == "all_ldl") |>
  dplyr::distinct(pathway_key, source, term_id, description, fdr)

all_hdl_pathways <- pathway_results_full |>
  dplyr::filter(gene_set == "all_hdl") |>
  dplyr::distinct(pathway_key, source, term_id, description, fdr)

full_trait_pathways <- dplyr::bind_rows(all_ldl_pathways, all_hdl_pathways) |>
  dplyr::distinct(pathway_key, source, term_id, description)

lpa_novel_pathways <- lpa_pathways |>
  dplyr::filter(!(pathway_key %in% full_trait_pathways$pathway_key))

ldl_not_hdl <- all_ldl_pathways |>
  dplyr::filter(!(pathway_key %in% all_hdl_pathways$pathway_key))

hdl_not_ldl <- all_hdl_pathways |>
  dplyr::filter(!(pathway_key %in% all_ldl_pathways$pathway_key))

pathway_novelty_summary <- data.frame(
  comparison = c(
    "All LDL pathways",
    "All HDL pathways",
    "LDL pathways not in all HDL",
    "HDL pathways not in all LDL",
    "LPA pathways not in all LDL or all HDL"
  ),
  n_pathways = c(
    nrow(all_ldl_pathways),
    nrow(all_hdl_pathways),
    nrow(ldl_not_hdl),
    nrow(hdl_not_ldl),
    nrow(lpa_novel_pathways)
  ),
  row.names = NULL
)

knitr::kable(pathway_novelty_summary)

if (nrow(lpa_novel_pathways) > 0) {
  lpa_novel_with_fdr <- pathway_results_full |>
    dplyr::filter(
      grepl("^lpa_profile_", gene_set),
      pathway_key %in% lpa_novel_pathways$pathway_key
    ) |>
    dplyr::group_by(pathway_key) |>
    dplyr::slice_min(fdr, n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(fdr) |>
    dplyr::slice_head(n = 10)

  knitr::kable(
    lpa_novel_with_fdr |>
      dplyr::mutate(fdr = sprintf("%.1e", fdr)) |>
      dplyr::select(gene_set, source, description, fdr),
    caption = "Pathways enriched in LPA profiles but not in full LDL or HDL gene sets"
  )
}
```

### Interpreting the comparison

Several patterns emerge from the LDL vs HDL comparison:

- **LPA profiles** enrich for pathways driven by loci with shared multi-trait effect directions (e.g. lipid metabolism in concordant-negative blocks, xenobiotic metabolism in concordant-positive blocks).
- **LDL-only / HDL-only minus LPA** reveal trait-specific biology in colocalisation loci *not* tagged by any latent profile. Pathways shared with the LPA union are consistent with pleiotropic blocks; pathways **not** in the LPA union are candidates for trait-specific mechanisms.
- **All LDL vs all HDL** show the broad trait-level pathway landscape. Differences between these sets reflect trait-specific colocalisation architecture at genome-wide significance.
- **LPA novelty** (pathways in LPA profiles but not in either full trait set) indicates that splitting by directional profile can surface enriched terms that are diluted when all colocalisation genes are pooled — the main justification for the LPA step.

On gene-level agreement between LPA and pathways:

- **Partially similar:** genes driving LPA profile enrichment tend to sit in the same profile loci — LPA and pathway annotation agree on modular biology.
- **Meaningfully different:** pathway overlap genes are a small fraction of each profile's gene list (low Jaccard), and `pathway_gene_ids` are much larger than any LPA gene set.
- **Complementary:** LPA groups loci by effect profile; pathway enrichment maps those genes to curated modules; subtracting LPA genes from trait-specific loci separates shared from remainder biology.

## Session summary

```{r summary}
summary_table <- data.frame(
  metric = c(
    "PCS",
    "Shared loci (n)",
    "LDL-only loci (m1)",
    "HDL-only loci (m2)",
    "NDS",
    "LPA profiles (best BIC)",
    "Loci in z-score matrix"
  ),
  value = c(
    round(pcs_result$pcs, 3),
    pcs_result$n_shared,
    pcs_result$m1,
    pcs_result$m2,
    round(nds_result$nds, 3),
    lpa_result$best_model$Classes,
    nrow(z_data$z_matrix)
  ),
  row.names = NULL
)
knitr::kable(summary_table)
```

## Further reading

- `?pleiotropic_consistency_score`, `?net_directionality_score`
- `?build_locus_zscore_matrix`, `?latent_causal_blocks`, `?summarise_latent_blocks`
- `?pathway_enrichment`
