---
title: "GWAS upload tutorial"
output:
  rmarkdown::html_vignette:
    css: vignette-wide.css
vignette: >
  %\VignetteIndexEntry{GWAS upload}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Introduction

The GPMap API allows you to upload your own GWAS summary statistics and receive colocalisation results against the existing genotype–phenotype map. This vignette covers how to upload a GWAS, retrieve your results, interpret them, and compare your upload with another trait (including another upload) to find insights.

# Part 1: How to use it and get the data

## Uploading a GWAS

Use `upload_gwas()` to submit your summary statistics. You need a file with standard GWAS columns and metadata such as sample size and ancestry.

```{r upload-example, eval = FALSE}
# Example: upload a GWAS file
result <- gpmapr::upload_gwas(
  file = "path/to/your_gwas.tsv.gz",
  name = "Specific Triat",
  p_value_threshold = 5e-8,
  column_names = list(
    CHR = "chr",
    BP = "pos",
    P = "pval",
    EA = "effect_allele",
    OA = "other_allele",
    EAF = "eaf",
    BETA = "beta",
    SE = "se"
  ),
  email = "your@email.com",
  category = "continuous",
  ancestry = "EUR",
  sample_size = 50000,
  reference_build = "GRCh38"
)
```

The response includes a **GUID** (a UUID such as `a1b2c3d4-e5f6-7890-abcd-ef1234567890`). Save this GUID; you use it to fetch your results once they are ready.

```{r guid-example, eval = FALSE}
# The GUID is returned in the response
my_guid <- result$id
```

## Fetching your results

Processing can take some time. Use the GUID with `trait()` to check status and retrieve results:

```{r fetch-results, eval = FALSE}
# Using trait() - works with both numeric trait IDs and GUIDs
my_results <- gpmapr::trait(my_guid, include_associations = TRUE, include_coloc_pairs = TRUE)
```

When processing is complete, the result includes `trait`, `coloc_groups`, `study_extractions`, `upload_study_extractions`, and optionally `associations` and `coloc_pairs`.



# Part 2: Interpreting your results

Your upload results use the same structure as traits in the database. Here are the main components and what they mean.

**Coloc groups** identify genomic regions (LD blocks) where your trait colocalises with other studies in the map. Each row is a study extraction that shares a colocalisation signal with your trait at that region.

If you are interested in the relationship between your trait and another specific trait, you can filter the coloc groups by trait id.

```{r coloc-groups-demo}
# Filter the coloc groups by trait of interest
trait_of_interest <- 931
coloc_groups <- my_results$coloc_groups |>
  dplyr::filter(trait_id == trait_of_interest)

compare_coloc_groups <- my_results$coloc_groups |>
  dplyr::filter(coloc_group_id %in% coloc_groups$coloc_group_id & (trait_id == trait_of_interest | gwas_upload_id == my_results$trait$id))

head(compare_coloc_groups)
```

**Coloc pairs** give pairwise colocalisation between study extractions, with probabilities for shared vs distinct causal variants:

- **h4**: Probability that both traits share **one common causal variant**
- **h3**: Probability that both traits have association at the region but through **two distinct causal variants**

High h4 (e.g. > 0.8) suggests strong evidence for a shared causal variant. The `h4_threshold` parameter filters pairs by this probability.

```{r coloc-pairs-demo}
study_extractions_of_gwas_upload <- my_results$study_extractions

coloc_pairs_of_interest <- my_results$coloc_pairs |>
  dplyr::filter(
    study_extraction_id_a == trait_of_interest | study_extraction_id_b == trait_of_interest |
    existing_study_extraction_id_a == trait_of_interest | existing_study_extraction_id_b == trait_of_interest
  )

# Coloc pairs from my_results (upload + trait of interest)
# Column names vary: study_extraction_id_a/b or existing_study_extraction_id_a/b for GWAS uploads
pairs <- my_results$coloc_pairs
if (!is.null(pairs) && nrow(pairs) > 0) {
  cols <- intersect(names(pairs), c(
    "study_extraction_id_a", "study_extraction_id_b",
    "existing_study_extraction_id_a", "existing_study_extraction_id_b",
    "ld_block_id", "h3", "h4", "false_positive", "spurious"
  ))
  head(pairs[, cols, drop = FALSE])
} else {
  "No coloc pairs for these traits"
}
```

## Study extractions

**Study extractions** are the individual finemapped regions (LD blocks) that contribute to coloc groups. They link your trait to specific studies, SNPs, and genes.

Key columns include `study`, `snp`, `gene`, `chr`, `bp`, `min_p`, and trait metadata (`trait_name`, `data_type`, `tissue`).

```{r study-extractions-demo}
# Study extractions (structure may vary - list of dataframes or single dataframe)
se <- demo$study_extractions
df <- if (is.data.frame(se) && nrow(se) > 0) se else if (is.list(se) && length(se) > 0 && is.data.frame(se[[1]])) se[[1]] else NULL
if (!is.null(df) && nrow(df) > 0) {
  cols <- intersect(names(df), c("study", "snp", "gene", "trait_name", "chr", "bp"))
  if (length(cols) > 0) head(df[, cols, drop = FALSE]) else head(df)
}
```

## Rare results

If present, **rare_results** are rare-disease associations that colocalise with your trait at the same regions. They use a similar structure to coloc groups, with `rare_result_group_id` and study/trait metadata.

# Part 3: Comparing with another GUID and filtering for insights

## Fetching multiple traits (including GUIDs)

Use `traits()` to fetch several traits or GUIDs in one call. This returns combined `coloc_groups`, `study_extractions`, and `rare_results` for all of them.

```{r compare-fetch, eval = FALSE}
# Compare your upload with another upload or a published trait
both <- gpmapr::traits(c(my_guid, "another-guid-or-numeric-id"),
  include_associations = TRUE,
  include_coloc_pairs = TRUE
)
```

```{r compare-fetch-demo}
# Demo with two published traits (same pattern as GUIDs)
both <- gpmapr::traits(c(931, 932), include_associations = TRUE, include_coloc_pairs = TRUE)
```

## Filtering to focus on the comparison

### Shared coloc groups

Find regions where both traits colocalise with the same studies:

```{r filter-shared}
# Get coloc groups for both traits
cg <- both$coloc_groups
if (is.null(cg)) cg <- data.frame()
shared_groups <- cg

# Find coloc_group_ids that appear for both traits (when trait_id column exists)
if (nrow(cg) > 0 && "trait_id" %in% names(cg)) {
  shared_groups <- cg |>
    dplyr::group_by(coloc_group_id) |>
    dplyr::filter(dplyr::n_distinct(trait_id) >= 2) |>
    dplyr::ungroup()

  # Regions (genes) shared between your traits
  shared_groups |>
    dplyr::distinct(coloc_group_id, gene, display_snp, trait_name) |>
    head(10)
} else {
  head(cg, 10)
}
```

### Filter by trait

Restrict results to one of your traits:

```{r filter-by-trait}
# Focus on one trait's coloc groups
if ("trait_id" %in% names(cg)) {
  trait_931_only <- cg |> dplyr::filter(trait_id == 931)
  head(trait_931_only)
} else {
  head(cg)
}
```

### Filter by tissue or data type

Narrow down to specific molecular layers (e.g. eQTL, pQTL) or tissues:

```{r filter-tissue}
# eQTL colocalisations only (when data_type column exists)
if ("data_type" %in% names(cg)) {
  eqtl_only <- cg |> dplyr::filter(data_type == "eQTL")
  nrow(eqtl_only)
}
# Specific tissue
if ("tissue" %in% names(cg)) {
  liver_only <- cg |> dplyr::filter(tissue == "Liver")
}
```

### High-confidence coloc pairs

Keep only pairs with strong evidence for a shared causal variant:

```{r filter-h4}
pairs <- both$coloc_pairs
if (!is.null(pairs) && nrow(pairs) > 0 && "h4" %in% names(pairs)) {
  high_h4 <- pairs |> dplyr::filter(h4 >= 0.8)
  if ("false_positive" %in% names(high_h4)) high_h4 <- high_h4 |> dplyr::filter(!false_positive)
  else if ("spurious" %in% names(high_h4)) high_h4 <- high_h4 |> dplyr::filter(!spurious)
  head(high_h4)
}
```

## Finding insights

1. **Shared biology**: Use shared coloc groups to find genes/regions where both traits likely share causal mechanisms.
2. **Tissue specificity**: Filter by `tissue` or `data_type` to see which molecular contexts drive colocalisation.
3. **Prioritise by h4**: Use `coloc_pairs` with high h4 to rank the strongest shared-signal pairs.
4. **Rare disease links**: Inspect `rare_results` for potential rare-variant overlap with your trait.

```{r insights-summary}
# Example: genes with coloc evidence for both traits
df <- if (nrow(shared_groups) > 0) shared_groups else cg
if (nrow(df) > 0 && "gene" %in% names(df)) {
  df |> dplyr::distinct(gene, display_snp, trait_name) |> head(15)
}
```

