---
title: "Colocalisation consistency, latent blocks, and pathway enrichment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colocalisation consistency, latent blocks, and pathway enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
library(gpmapr)
library(dplyr)
library(tidyr)
```

This vignette walks through three complementary layers of colocalisation analysis in **gpmapr**:

1. **Pleiotropic Consistency Score (PCS)** and **Net Directionality Score (NDS)** — quantify locus overlap and effect concordance between traits.
2. **Latent causal blocks** — cluster loci with similar multi-trait effect profiles using Latent Profile Analysis (LPA).
3. **Pathway enrichment** — test whether genes mapped to those loci are enriched for biological pathways, and compare LPA-refined gene sets with broader colocalisation blocks.

We use **LDL cholesterol** (trait id `1486`) and **HDL cholesterol levels** (trait id `1630`) as an illustrative pair of lipid traits that share some, but not all, causal loci.

## Fetch colocalisation data

We fetch both traits in a single call and retain association statistics (`beta`, `se`) for directionality and z-score construction.

```{r fetch-data}
ldl_id <- 1486
hdl_id <- 1630

lipid_traits <- traits(
  c(ldl_id, hdl_id),
  include_associations = TRUE
)
cg <- lipid_traits$coloc_groups
```

## Pleiotropic consistency (PCS)

`pleiotropic_consistency_score()` measures how much two traits colocalise at the same LD blocks versus having trait-specific signals. The score is a Jaccard-style ratio:

$$\text{PCS} = \frac{n}{n + m_1 + m_2}$$

where $n$ is the number of shared loci, $m_1$ are loci unique to trait 1, and $m_2$ are loci unique to trait 2.

```{r pcs}
pcs_result <- pleiotropic_consistency_score(
  ldl_id, hdl_id,
  coloc_groups = cg
)

pcs_result
```

LDL and HDL share a modest fraction of colocalising loci: most signals are trait-specific. This low PCS motivates splitting loci into blocks before interpreting shared biology.

`pcs_matrix()` extends this to all pairs when comparing more than two traits:

```{r pcs-matrix}
pcs_mat <- pcs_matrix(c(ldl_id, hdl_id), coloc_groups = cg)
pcs_mat
```

## Net directionality (NDS)

At loci where both traits colocalise, `net_directionality_score()` asks whether effect directions agree:

$$\text{NDS} = \frac{n_+ - n_-}{n_+ + n_-}$$

```{r nds}
nds_result <- net_directionality_score(
  ldl_id, hdl_id,
  coloc_groups = cg
)

nds_result
```

A positive NDS means concordant effect directions predominate at shared loci (here, slightly more concordant than discordant pairs).

## Latent causal blocks

PCS splits loci by *presence* of colocalisation per trait. **Latent Profile Analysis** instead groups loci by their *multi-trait effect profile* — the pattern of z-scores across traits at each locus.

### Build the z-score matrix

`build_locus_zscore_matrix()` constructs a loci x traits matrix of z-scores (`beta / se`). Loci without a genome-wide signal for a trait are set to 0. Only loci with signals in at least two traits are retained.

```{r z-matrix}
z_data <- build_locus_zscore_matrix(
  c(ldl_id, hdl_id),
  coloc_groups = cg
)

dim(z_data$z_matrix)
head(z_data$locus_info)
```

### Fit LPA profiles

`latent_causal_blocks()` fits LPA models via **tidyLPA** and returns profile assignments for each locus. We search over 2-4 profiles and use model 1 (equal variances, zero covariances) for speed.

```{r lpa}
lpa_result <- latent_causal_blocks(
  z_data$z_matrix,
  n_profiles = 2:4,
  models = 1
)

lpa_result$best_model
table(lpa_result$assignments$profile)
```

### Summarise profile biology

`summarise_latent_blocks()` reports mean z-scores per trait within each profile — the signature of each latent block.

```{r summarise-blocks}
block_summary <- summarise_latent_blocks(
  z_data$z_matrix,
  lpa_result$assignments,
  trait_info = z_data$trait_info
)

knitr::kable(
  block_summary |>
    dplyr::select(profile, trait_name, mean_z, sd_z, n_loci) |>
    dplyr::arrange(profile, trait_name),
  digits = 2
)
```

Profile 1 loci tend toward positive LDL and HDL z-scores; profile 2 loci tend toward negative values for both traits. Profile 3 contains only a handful of loci with strong negative effects.

## Pathway enrichment comparison

`pathway_enrichment()` tests a gene set against Reactome, KEGG, and HP pathway databases via the GPMap API. We use numeric `gene_id` values here, but the API also accepts gene names.

We ask four related questions:

1. Which pathways are enriched in each **LPA profile**?
2. In trait-specific loci, what remains after **removing LPA genes** from LDL-only and HDL-only blocks?
3. What pathways appear when enriching **all LDL** or **all HDL** colocalisation genes?
4. Does LPA surface **novel** pathways not found in the full trait gene sets?

```{r gene-sets}
locus_genes <- cg |>
  dplyr::distinct(coloc_group_id, gene_id, gene) |>
  dplyr::filter(!is.na(gene_id))

assignments <- lpa_result$assignments |>
  dplyr::mutate(coloc_group_id = as.integer(locus_id))

genes_for_loci <- function(locus_ids) {
  unique(locus_genes$gene_id[locus_genes$coloc_group_id %in% locus_ids])
}

ldl_loci <- cg |>
  dplyr::filter(trait_id == ldl_id, min_p <= 5e-8) |>
  dplyr::distinct(coloc_group_id) |>
  dplyr::pull(coloc_group_id)

hdl_loci <- cg |>
  dplyr::filter(trait_id == hdl_id, min_p <= 5e-8) |>
  dplyr::distinct(coloc_group_id) |>
  dplyr::pull(coloc_group_id)

ldl_only_loci <- setdiff(ldl_loci, hdl_loci)
hdl_only_loci <- setdiff(hdl_loci, ldl_loci)

lpa_genes <- genes_for_loci(assignments$coloc_group_id)
ldl_only_genes <- genes_for_loci(ldl_only_loci)
hdl_only_genes <- genes_for_loci(hdl_only_loci)

gene_sets <- list(
  all_ldl = genes_for_loci(ldl_loci),
  all_hdl = genes_for_loci(hdl_loci),
  ldl_only_minus_lpa = setdiff(ldl_only_genes, lpa_genes),
  hdl_only_minus_lpa = setdiff(hdl_only_genes, lpa_genes)
)

for (profile_id in sort(unique(assignments$profile))) {
  loci <- assignments$coloc_group_id[assignments$profile == profile_id]
  gene_sets[[paste0("lpa_profile_", profile_id)]] <- genes_for_loci(loci)
}

gene_set_sizes <- data.frame(
  gene_set = names(gene_sets),
  n_genes = vapply(gene_sets, length, integer(1)),
  row.names = NULL
)
knitr::kable(gene_set_sizes)
```

```{r pathway-enrichment}
run_pathway_enrichment <- function(gene_ids, label) {
  if (length(gene_ids) == 0) {
    return(NULL)
  }
  n_input_genes <- length(gene_ids)
  pe <- pathway_enrichment(
    gene_ids,
    p_value_threshold = 0.05,
    minimum_count_in_network = 1
  )
  results <- pe$results
  if (is.null(results) || !is.data.frame(results) || nrow(results) == 0) {
    return(NULL)
  }
  results |>
    dplyr::mutate(
      gene_set = label,
      n_genes = n_input_genes,
      n_pathway_genes = vapply(pathway_gene_ids, length, integer(1)),
      pathway_key = paste(source, term_id, sep = ":")
    )
}

pathway_results_full <- dplyr::bind_rows(
  lapply(names(gene_sets), function(nm) {
    run_pathway_enrichment(gene_sets[[nm]], nm)
  })
)

if (nrow(pathway_results_full) == 0) {
  stop(
    "No pathway enrichment results were returned. ",
    "Reinstall or reload gpmapr (devtools::load_all()) so the API uses the 'genes' parameter."
  )
}

pathway_comparison <- pathway_results_full |>
  dplyr::group_by(gene_set) |>
  dplyr::slice_min(fdr, n = 5, with_ties = FALSE) |>
  dplyr::ungroup()

knitr::kable(
  pathway_comparison |>
    dplyr::arrange(gene_set, fdr) |>
    dplyr::mutate(fdr = sprintf("%.1e", fdr)) |>
    dplyr::select(gene_set, n_genes, source, description, overlap, n_pathway_genes, fdr)
)
```

### Shared vs trait-specific pathways after removing LPA genes

Pathways enriched in any LPA profile are compared with pathways found in LDL-only or HDL-only gene sets **after LPA genes are removed**. Terms in the remainder sets that do not appear in the LPA union are trait-specific signal not captured by the latent blocks.

```{r pathway-overlap}
lpa_pathways <- pathway_results_full |>
  dplyr::filter(grepl("^lpa_profile_", gene_set)) |>
  dplyr::distinct(pathway_key, source, term_id, description)

ldl_remainder_pathways <- pathway_results_full |>
  dplyr::filter(gene_set == "ldl_only_minus_lpa") |>
  dplyr::distinct(pathway_key, source, term_id, description, fdr)

hdl_remainder_pathways <- pathway_results_full |>
  dplyr::filter(gene_set == "hdl_only_minus_lpa") |>
  dplyr::distinct(pathway_key, source, term_id, description, fdr)

ldl_shared_with_lpa <- ldl_remainder_pathways |>
  dplyr::filter(pathway_key %in% lpa_pathways$pathway_key)

ldl_not_in_lpa <- ldl_remainder_pathways |>
  dplyr::filter(!(pathway_key %in% lpa_pathways$pathway_key))

hdl_shared_with_lpa <- hdl_remainder_pathways |>
  dplyr::filter(pathway_key %in% lpa_pathways$pathway_key)

hdl_not_in_lpa <- hdl_remainder_pathways |>
  dplyr::filter(!(pathway_key %in% lpa_pathways$pathway_key))

pathway_overlap_summary <- data.frame(
  comparison = c(
    "LPA profile pathways (union)",
    "LDL-only minus LPA (total)",
    "LDL-only minus LPA (also in LPA)",
    "LDL-only minus LPA (not in LPA)",
    "HDL-only minus LPA (total)",
    "HDL-only minus LPA (also in LPA)",
    "HDL-only minus LPA (not in LPA)"
  ),
  n_pathways = c(
    nrow(lpa_pathways),
    nrow(ldl_remainder_pathways),
    nrow(ldl_shared_with_lpa),
    nrow(ldl_not_in_lpa),
    nrow(hdl_remainder_pathways),
    nrow(hdl_shared_with_lpa),
    nrow(hdl_not_in_lpa)
  ),
  row.names = NULL
)

knitr::kable(pathway_overlap_summary)

if (nrow(ldl_not_in_lpa) > 0) {
  knitr::kable(
    ldl_not_in_lpa |>
      dplyr::arrange(fdr) |>
      dplyr::slice_head(n = 5) |>
      dplyr::mutate(fdr = sprintf("%.1e", fdr)) |>
      dplyr::select(source, description, fdr),
    caption = "Top LDL-only pathways not enriched in any LPA profile"
  )
}

if (nrow(hdl_not_in_lpa) > 0) {
  knitr::kable(
    hdl_not_in_lpa |>
      dplyr::arrange(fdr) |>
      dplyr::slice_head(n = 5) |>
      dplyr::mutate(fdr = sprintf("%.1e", fdr)) |>
      dplyr::select(source, description, fdr),
    caption = "Top HDL-only pathways not enriched in any LPA profile"
  )
}
```

### All LDL vs all HDL, and LPA novelty

Finally, we enrich all genome-wide colocalisation genes for each trait and ask whether LPA profiles recover pathways that the full trait gene sets do not.

```{r pathway-novelty}
all_ldl_pathways <- pathway_results_full |>
  dplyr::filter(gene_set == "all_ldl") |>
  dplyr::distinct(pathway_key, source, term_id, description, fdr)

all_hdl_pathways <- pathway_results_full |>
  dplyr::filter(gene_set == "all_hdl") |>
  dplyr::distinct(pathway_key, source, term_id, description, fdr)

full_trait_pathways <- dplyr::bind_rows(all_ldl_pathways, all_hdl_pathways) |>
  dplyr::distinct(pathway_key, source, term_id, description)

lpa_novel_pathways <- lpa_pathways |>
  dplyr::filter(!(pathway_key %in% full_trait_pathways$pathway_key))

ldl_not_hdl <- all_ldl_pathways |>
  dplyr::filter(!(pathway_key %in% all_hdl_pathways$pathway_key))

hdl_not_ldl <- all_hdl_pathways |>
  dplyr::filter(!(pathway_key %in% all_ldl_pathways$pathway_key))

pathway_novelty_summary <- data.frame(
  comparison = c(
    "All LDL pathways",
    "All HDL pathways",
    "LDL pathways not in all HDL",
    "HDL pathways not in all LDL",
    "LPA pathways not in all LDL or all HDL"
  ),
  n_pathways = c(
    nrow(all_ldl_pathways),
    nrow(all_hdl_pathways),
    nrow(ldl_not_hdl),
    nrow(hdl_not_ldl),
    nrow(lpa_novel_pathways)
  ),
  row.names = NULL
)

knitr::kable(pathway_novelty_summary)

if (nrow(lpa_novel_pathways) > 0) {
  lpa_novel_with_fdr <- pathway_results_full |>
    dplyr::filter(
      grepl("^lpa_profile_", gene_set),
      pathway_key %in% lpa_novel_pathways$pathway_key
    ) |>
    dplyr::group_by(pathway_key) |>
    dplyr::slice_min(fdr, n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(fdr) |>
    dplyr::slice_head(n = 10)

  knitr::kable(
    lpa_novel_with_fdr |>
      dplyr::mutate(fdr = sprintf("%.1e", fdr)) |>
      dplyr::select(gene_set, source, description, fdr),
    caption = "Pathways enriched in LPA profiles but not in full LDL or HDL gene sets"
  )
}
```

### Do LPA gene groups match pathway gene groups?

LPA groups genes by locus-level effect *profiles*; pathway enrichment groups genes by curated pathway membership. They answer different questions, so we should not expect identical gene lists. We can still ask how much they agree.

First, for each LPA profile, compare the full profile gene set with the union of input genes (`gene_ids`) that drove enrichment:

```{r lpa-pathway-gene-overlap}
jaccard <- function(a, b) {
  intersection <- length(intersect(a, b))
  union_size <- length(union(a, b))
  if (union_size == 0) 0 else intersection / union_size
}

lpa_pathway_overlap <- pathway_comparison |>
  dplyr::filter(grepl("^lpa_profile_", gene_set)) |>
  dplyr::group_by(gene_set) |>
  dplyr::summarise(
    pathway_overlap_genes = list(unique(unlist(gene_ids))),
    .groups = "drop"
  ) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    n_lpa_genes = length(gene_sets[[gene_set]]),
    n_pathway_overlap_genes = length(pathway_overlap_genes),
    n_shared_genes = length(intersect(
      gene_sets[[gene_set]],
      pathway_overlap_genes
    )),
    jaccard = jaccard(gene_sets[[gene_set]], pathway_overlap_genes)
  ) |>
  dplyr::ungroup() |>
  dplyr::select(
    gene_set, n_lpa_genes, n_pathway_overlap_genes, n_shared_genes, jaccard
  )

knitr::kable(lpa_pathway_overlap, digits = 3)
```

Low Jaccard scores are expected: most genes at loci in a profile are not members of any significantly enriched pathway term. Enrichment is typically driven by a small subset of pathway-mapped genes.

Next, for all genes that appear in enriched pathways, check which LPA profile their colocalisation loci were assigned to:

```{r enriched-gene-profiles}
gene_to_profile <- assignments |>
  dplyr::inner_join(
    locus_genes |> dplyr::distinct(coloc_group_id, gene_id, gene),
    by = "coloc_group_id",
    relationship = "many-to-many"
  ) |>
  dplyr::distinct(gene_id, gene, profile)

enriched_genes <- unique(unlist(
  pathway_comparison$gene_ids[grepl("^lpa_profile_", pathway_comparison$gene_set)]
))

enriched_gene_profiles <- gene_to_profile |>
  dplyr::filter(gene_id %in% enriched_genes) |>
  dplyr::count(profile, name = "n_enriched_genes")

knitr::kable(enriched_gene_profiles)

if (length(enriched_genes) > 0) {
  knitr::kable(
    gene_to_profile |>
      dplyr::filter(gene_id %in% enriched_genes) |>
      dplyr::arrange(profile, gene),
    col.names = c("Gene ID", "Gene", "LPA profile")
  )
}
```

### Interpreting the comparison

Several patterns emerge from the LDL vs HDL comparison:

- **LPA profiles** enrich for pathways driven by loci with shared multi-trait effect directions (e.g. lipid metabolism in concordant-negative blocks, xenobiotic metabolism in concordant-positive blocks).
- **LDL-only / HDL-only minus LPA** reveal trait-specific biology in colocalisation loci *not* tagged by any latent profile. Pathways shared with the LPA union are consistent with pleiotropic blocks; pathways **not** in the LPA union are candidates for trait-specific mechanisms.
- **All LDL vs all HDL** show the broad trait-level pathway landscape. Differences between these sets reflect trait-specific colocalisation architecture at genome-wide significance.
- **LPA novelty** (pathways in LPA profiles but not in either full trait set) indicates that splitting by directional profile can surface enriched terms that are diluted when all colocalisation genes are pooled — the main justification for the LPA step.

On gene-level agreement between LPA and pathways:

- **Partially similar:** genes driving LPA profile enrichment tend to sit in the same profile loci — LPA and pathway annotation agree on modular biology.
- **Meaningfully different:** pathway overlap genes are a small fraction of each profile's gene list (low Jaccard), and `pathway_gene_ids` are much larger than any LPA gene set.
- **Complementary:** LPA groups loci by effect profile; pathway enrichment maps those genes to curated modules; subtracting LPA genes from trait-specific loci separates shared from remainder biology.

## Session summary

```{r summary}
summary_table <- data.frame(
  metric = c(
    "PCS",
    "Shared loci (n)",
    "LDL-only loci (m1)",
    "HDL-only loci (m2)",
    "NDS",
    "LPA profiles (best BIC)",
    "Loci in z-score matrix"
  ),
  value = c(
    round(pcs_result$pcs, 3),
    pcs_result$n_shared,
    pcs_result$m1,
    pcs_result$m2,
    round(nds_result$nds, 3),
    lpa_result$best_model$Classes,
    nrow(z_data$z_matrix)
  ),
  row.names = NULL
)
knitr::kable(summary_table)
```

## Further reading

- `?pleiotropic_consistency_score`, `?net_directionality_score`
- `?build_locus_zscore_matrix`, `?latent_causal_blocks`, `?summarise_latent_blocks`
- `?pathway_enrichment`
