---
title: "Investigation of univariate analysis: BMI"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Investigation of univariate analysis: BMI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "local"
  include_trans: true
  trait_subset: "phenotypic"
  min_snp_signals: 2
  min_module_size: 3
  compress_method: "asinh"
  compress_scale: 5
  include_situated_gene: false
  similarity_threshold: 0.2
  ebmf_magnitude_threshold: 0.75
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  ebmf_backfit: true
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 3
  frac_traits: 0.8
  stability_top_n: 20
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r trait-metadata, include=FALSE}
api <- params$api
target_trait_id <- 1992L
select_api(api)
target_trait <- trait(target_trait_id, include_associations = TRUE)
target_trait_name <- target_trait$trait$trait_name
```

Single-trait SNP program discovery for **`r target_trait_name`** (trait id
`r target_trait_id`). Each column of the matrix is a BMI GWAS SNP and each row
a background study. Profiles are oriented to the BMI-raising allele,
compressed, and factorised with empirical Bayes matrix factorization (EBMF)
into overlapping programs. Each program then gets a confidence tier (high,
medium or low) from a size-matched similarity test and connectedness.

BMI loci have a strong shared direction: almost every BMI-raising allele also
raises weight, fat mass, fat-free mass and other body-size measures. That makes
two random BMI SNPs look alike, which hides the structure programs are meant to
capture. This vignette fits and filters the programs two ways to see what
removing that background changes:

- **Run A: all traits.** Only other GWAS of BMI and obesity diagnoses are
  removed.
- **Run B: body-size traits removed.** Run A minus the traits that move with
  BMI at almost every SNP.

Step 4 compares the two runs' continuous enrichment of trait categories,
tissues and pathways, and Steps 5–7 interpret both runs' programs side by
side.

```{r params}
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_specificity <- 1.25

# Shared analysis settings come from vignettes/_ebmf_settings.yml so the
# simulation, null-simulation, univariate and multi-trait studies cannot drift
# apart; anything supplied in params (e.g. by the HPC grid scripts) wins.
ebmf_settings <- resolve_ebmf_settings(params)
settings_signature <- ebmf_settings_signature(ebmf_settings)

trait_subset <- ebmf_settings$trait_subset
min_module_size <- as.integer(ebmf_settings$min_module_size)
min_snp_signals <- as.integer(ebmf_settings$min_snp_signals)
compress_method <- ebmf_settings$compress_method
compress_scale <- as.numeric(ebmf_settings$compress_scale)
include_trans <- isTRUE(ebmf_settings$include_trans)
include_situated_gene <- isTRUE(params$include_situated_gene)
similarity_threshold <- as.numeric(ebmf_settings$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(ebmf_settings$ebmf_lfsr_threshold)
ebmf_prior <- ebmf_settings$ebmf_prior
ebmf_backfit <- isTRUE(ebmf_settings$ebmf_backfit)
ebmf_magnitude_threshold <- as.numeric(
  ebmf_settings$ebmf_magnitude_threshold
)
similarity_n_perm <- as.integer(ebmf_settings$similarity_n_perm)
similarity_q_threshold <- as.numeric(ebmf_settings$similarity_q)
high_similarity_q <- as.numeric(ebmf_settings$high_similarity_q)
high_connectedness <- as.numeric(ebmf_settings$high_connectedness)
min_effective_traits <- as.numeric(ebmf_settings$min_effective_traits)
ebmf_greedy_Kmax <- as.integer(ebmf_settings$ebmf_greedy_Kmax)
associations <- params$associations
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)

knitr::kable(
  data.frame(
    api = api,
    gpmap_url = getOption("gpmap_url"),
    min_snp_signals = min_snp_signals,
    trait_subset = trait_subset,
    compress_method = compress_method,
    compress_scale = compress_scale,
    include_trans = include_trans,
    include_situated_gene = include_situated_gene,
    min_specificity = min_specificity,
    similarity_threshold = similarity_threshold,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    stability_top_n = stability_top_n,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    similarity_n_perm = similarity_n_perm,
    similarity_q = similarity_q_threshold,
    high_similarity_q = high_similarity_q,
    high_connectedness = high_connectedness,
    min_effective_traits = min_effective_traits,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    associations = associations,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)
```

## Step 1: Traits removed

### BMI proxies (both runs)

Other GWAS of BMI, and diagnoses defined by BMI, repeat the anchor trait at
every SNP. They add no information about mechanism and are removed in both
runs.

```{r bmi-proxies}
bmi_proxy_traits <- data.frame(
  trait_id = c("3748", "2477", "2333", "3338"),
  trait_name = c(
    "Body mass index (BMI)",
    "Body mass index (UKB data field 21001)",
    "Obesity",
    "Diagnoses - secondary ICD10: E66.9 Obesity, unspecified"
  ),
  reason = c("BMI GWAS", "BMI GWAS", "Defined by BMI", "Defined by BMI"),
  stringsAsFactors = FALSE
)

proxy_counts <- target_trait$coloc_groups |>
  dplyr::mutate(trait_id = as.character(trait_id)) |>
  dplyr::filter(trait_id %in% bmi_proxy_traits$trait_id) |>
  dplyr::group_by(trait_id) |>
  dplyr::summarise(n_coloc_groups = dplyr::n_distinct(coloc_group_id), .groups = "drop")

knitr::kable(
  bmi_proxy_traits |>
    dplyr::left_join(proxy_counts, by = "trait_id") |>
    dplyr::mutate(n_coloc_groups = tidyr::replace_na(n_coloc_groups, 0L)),
  caption = "BMI proxy traits removed in every run. n_coloc_groups = BMI coloc groups each trait shares.",
  row.names = FALSE
)

coloc_groups_all <- target_trait$coloc_groups |>
  dplyr::filter(!as.character(trait_id) %in% bmi_proxy_traits$trait_id)
if (!include_trans && "cis_trans" %in% names(coloc_groups_all)) {
  coloc_groups_all <- coloc_groups_all |>
    dplyr::filter(is.na(cis_trans) | tolower(cis_trans) != "trans")
}
```

### Body-size traits (removed in Run B)

The list below was drawn up from the BMI coloc partners with a rule:
anthropometric or body-composition traits observed at 10 or more BMI SNPs
whose effect has the same sign at 80% or more of them after orienting to the
BMI-raising allele (positive, or negative as for impedance). These traits move
with BMI almost everywhere. Waist-hip ratio and waist-to-hip ratio adjusted for
BMI also meet the rule but are kept, because fat distribution is what separates
obesity sub-programs. Height, sitting height and appendicular lean mass do not
meet the rule and are kept. The table shows each listed trait's concordance in
the current data.

```{r body-size-traits}
body_size_trait_ids <- c(
  "4241", "4400", "4286", "3584", "3699", "3090", "3775", "3759", "4503",
  "3342", "2581", "2623", "3613", "3814", "2006", "2624", "1481", "3395",
  "3499", "3424", "3084", "3242", "3744", "4472", "3709", "3265", "3143",
  "3448", "3743", "3219", "4303", "3693", "3516", "4532", "4353", "4053",
  "3992", "4423", "3827", "2465"
)

# Concordance of each listed trait with BMI in the current data: the share of
# its observed effects that are positive at the BMI-raising allele.
bmi_rows <- coloc_groups_all |>
  dplyr::filter(as.character(trait_id) == as.character(target_trait_id)) |>
  dplyr::filter(!is.na(beta), !is.na(se), se > 0) |>
  dplyr::group_by(coloc_group_id) |>
  dplyr::slice_min(min_p, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::transmute(coloc_group_id, bmi_sign = sign(beta))
body_size_concordance <- coloc_groups_all |>
  dplyr::mutate(trait_id = as.character(trait_id)) |>
  dplyr::filter(trait_id %in% body_size_trait_ids, !is.na(beta)) |>
  dplyr::inner_join(bmi_rows, by = "coloc_group_id") |>
  dplyr::group_by(trait_id, trait_name) |>
  dplyr::summarise(
    n_coloc_groups = dplyr::n_distinct(coloc_group_id),
    pct_positive_at_bmi_raising = round(100 * mean(sign(beta) == bmi_sign)),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(n_coloc_groups))

knitr::kable(
  body_size_concordance,
  caption = paste0(
    "Body-size traits removed in Run B (", length(body_size_trait_ids), " traits). ",
    "pct_positive_at_bmi_raising near 100 (or near 0) means the trait moves with ",
    "(or against) BMI at almost every shared locus."
  ),
  row.names = FALSE
)

coloc_groups_cut <- coloc_groups_all |>
  dplyr::filter(!as.character(trait_id) %in% body_size_trait_ids)
```

## Step 2: Two runs

Each run builds the pleiotropy matrix from its trait set, factorises it with
EBMF (`run_univariate_clustering()`), and scores the programs
(`summarise_ebmf_programs()`) at the shared settings. Programs with fewer than
`min_module_size` member SNPs are dropped before any check.

A program's **confidence tier** comes from a similarity test and
connectedness. `S` is the cosine similarity between two SNPs' oriented,
compressed trait profiles (missing cells counted as 0, the BMI row left out).

- **similarity_q** — the mean `S` between member SNPs is compared with
  `similarity_n_perm` random groups of the same number of SNPs from the same
  run; `similarity_q` is the BH-adjusted share of random groups at least as
  similar. It is a significance test, so larger programs pass more easily, and
  the run's background similarity sets how hard it is to pass.
- **connectedness** — share of member pairs with `S >= similarity_threshold`.

- **low** — `similarity_q >= similarity_q`: the member SNPs are no more alike
  than random SNP groups of the same size. This is checked first, so a program
  that fails it is low whatever its number of traits, and can be set aside.
- **single_trait** — passes the similarity test, but the program's trait
  loadings rest on fewer than `min_effective_traits` effective traits
  (`effective_n_traits`). A program is defined as multi-trait: a factor
  carried by one trait is just that trait's hits, which noise produces too,
  so it needs a closer look rather than being discarded. Near-duplicate traits
  each count, so a program spread over several body-size traits is not
  flagged.
- **high** — `similarity_q < high_similarity_q` and `connectedness >=
  high_connectedness`.
- **medium** — `similarity_q < similarity_q`, but not high.

`excess_similarity` and `excess_connectedness` (each minus its value in the
random groups: effect sizes that do not grow with program size), factor strength
(`factor_pve`) and trait-subsampling `replication` are reported but do not
affect the tier. A program whose SNPs are at least 90% contained in a larger
program's loci is listed under it as a child.

```{r two-runs}
run_bmi_analysis <- function(label, coloc_groups) {
  set.seed(20240214)
  fit <- run_univariate_clustering(
    list(trait = list(id = target_trait_id), coloc_groups = coloc_groups),
    target_trait_id = target_trait_id,
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    max_snp_fraction = max_snp_fraction,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    min_module_size = min_module_size
  )
  summary <- summarise_ebmf_programs(
    fit,
    edge_threshold = similarity_threshold,
    min_module_size = min_module_size,
    similarity_n_perm = similarity_n_perm,
    similarity_q_threshold = similarity_q_threshold,
    high_similarity_q = high_similarity_q,
    high_min_connectedness = high_connectedness,
    min_effective_traits = min_effective_traits,
    n_null = n_null,
    alpha_membership = alpha_membership,
    n_candidate_tier = n_candidate_tier,
    n_rep = n_rep,
    frac_traits = frac_traits,
    top_n = stability_top_n,
    posterior_stat = posterior_stat,
    posterior_evidence_cap = posterior_evidence_cap,
    redundancy_threshold = redundancy_threshold,
    cores = cores,
    verbose = FALSE
  )
  return(list(
    label = label,
    coloc_groups = fit$coloc_groups,
    fit = fit,
    summary = summary
  ))
}

runs <- list(
  A = run_bmi_analysis("A: all traits", coloc_groups_all),
  B = run_bmi_analysis("B: body-size traits removed", coloc_groups_cut)
)
```

### Comparison

The **background** columns describe each run's matrix before any program is
found. `background_mean_S` is the mean similarity over all pairs of BMI SNPs and
`background_edge_rate` the share of pairs with `S >= similarity_threshold`: how
alike two random BMI SNPs already are. `median_shared_traits` is how many
traits two SNPs are typically both observed on, which is how much evidence each
similarity rests on. A high background makes every program look coherent and
shrinks the excess that real structure can show.

```{r comparison}
comparison <- dplyr::bind_rows(lapply(runs, function(r) {
  p <- r$summary$programs
  bg <- r$summary$background
  X_run <- r$fit$x_matrix
  return(data.frame(
    run = r$label,
    n_traits = nrow(X_run),
    n_snps = ncol(X_run),
    fraction_observed = round(mean(!is.na(X_run)), 4),
    background_mean_S = round(bg$mean_similarity, 3),
    background_edge_rate = round(bg$frac_pairs_edge, 3),
    median_shared_traits = bg$median_shared_traits,
    n_programs = nrow(p),
    n_dropped_small = nrow(r$summary$dropped_programs),
    n_high = sum(p$confidence_tier == "high"),
    n_medium = sum(p$confidence_tier == "medium"),
    n_single_trait = sum(p$confidence_tier == "single_trait"),
    n_low = sum(p$confidence_tier == "low"),
    median_excess_similarity = round(stats::median(p$excess_similarity, na.rm = TRUE), 3),
    median_excess_connectedness = round(stats::median(p$excess_connectedness, na.rm = TRUE), 3),
    stringsAsFactors = FALSE
  ))
}))
knitr::kable(
  comparison,
  caption = paste0(
    "The two runs side by side. n_programs counts programs with at least ",
    "min_module_size member SNPs; n_dropped_small those removed below it."
  ),
  row.names = FALSE
)
```

## Step 3: Programs per run

Programs are listed by tier (green high, orange medium, grey single trait,
red low), each child indented under its parent.

```{r program-tables, results = "asis"}
for (run_id in names(runs)) {
  r <- runs[[run_id]]
  cat("\n\n### Run ", r$label, "\n\n", sep = "")
  print(kable_confidence(
    r$summary$programs |>
      dplyr::transmute(
        program = program_label,
        confidence_tier = as.character(confidence_tier),
        n_snps_filtered,
        effective_n_traits = round(effective_n_traits, 2),
        mean_internal_similarity = round(mean_internal_similarity, 3),
        excess_similarity = round(excess_similarity, 3),
        connectedness = round(connectedness, 3),
        excess_connectedness = round(excess_connectedness, 3),
        similarity_q = signif(similarity_q, 3),
        factor_pve_pct = round(100 * factor_pve, 2),
        replication = round(replication, 2),
        parent_program
      ),
    caption = paste0(
      "Run ", r$label, ": programs by confidence tier. factor_pve_pct is the % ",
      "of matrix variance the factor explains; similarity_q, factor_pve_pct and ",
      "replication are reported, not used for the tier."
    )
  ))
}
```

**Program distinctness.** How alike each run's kept programs are, on both
axes of the fit, with the same confidence weighting as the cross-trait links
(`program_distinctness()`): the largest Tucker congruence between two
programs' trait loadings (BMI row left out; 0.85 is fair similarity, 0.95
equal) and the number of pairs at or above 0.7, the largest squared uncentred
correlation between their SNP loadings, and the effective number of distinct
programs (equal to `n_programs` when every pair is orthogonal).

```{r distinctness}
knitr::kable(
  dplyr::bind_rows(lapply(runs, function(r) {
    out <- program_distinctness(r$fit, r$summary$programs$program)$summary
    return(data.frame(run = r$label, out, check.names = FALSE))
  })) |>
    dplyr::mutate(
      max_abs_phi = round(max_abs_phi, 2),
      max_alignment = round(max_alignment, 2),
      effective_rank = round(effective_rank, 1)
    ),
  caption = "Distinctness of each run's kept programs",
  row.names = FALSE
)
```

## Step 4: Continuous enrichment, Run A vs Run B

Each run's high- and medium-confidence programs are annotated with
**continuous loading-based enrichment** of trait categories, tissues and
pathways (KEGG and Reactome, on the colocalising QTL genes). There is no
membership cutoff: every SNP with a finite EBMF loading contributes to every
program's test, weighted by its squared loading, in the regression
`lm(loading^2 ~ membership)` against each annotation's 0/1 indicator
(`enrich_program_loadings_trait_categories()`,
`enrich_program_loadings_tissues()`, `enrich_program_loadings_pathways()`).
A hit needs both FDR ≤ 0.05 and R² ≥ 0.05. The FDR is computed within each
run, across every program tested, separately for categories, tissues and
pathways (KEGG and Reactome pooled). R² is the share of a program's
squared-loading variance the annotation explains; 0.05 sits between Cohen's
"small" (0.01) and "medium" (0.09).

Each run is a separate factorisation, so program numbers do not carry over:
A3 and B3 are unrelated. The heatmap first matches programs across runs by the
overlap of their member SNPs.

```{r run-enrichment}
min_r_squared <- 0.05
enrichment_types <- c("trait category", "tissue", "Reactome", "KEGG")

pathway_mapping_universe <- tryCatch(
  pathway_mappings(),
  error = function(e) {
    cat("Pathway mappings endpoint unavailable (", conditionMessage(e), "); ",
        "skipping continuous pathway enrichment.\n", sep = "")
    return(NULL)
  }
)

valid_programs_of <- function(r) {
  p <- r$summary$programs
  return(as.character(p$program[as.character(p$confidence_tier) %in% c("high", "medium")]))
}

run_continuous_enrichment <- function(r) {
  return(list(
    categories = enrich_program_loadings_trait_categories(
      clustering_result = r$fit,
      coloc_groups = r$coloc_groups,
      snp_key = "variant_id"
    ),
    tissues = enrich_program_loadings_tissues(
      clustering_result = r$fit,
      coloc_groups = r$coloc_groups,
      snp_key = "variant_id"
    ),
    pathways = if (is.null(pathway_mapping_universe)) {
      NULL
    } else {
      enrich_program_loadings_pathways(
        clustering_result = r$fit,
        coloc_groups = r$coloc_groups,
        mappings = pathway_mapping_universe,
        snp_key = "variant_id",
        sources = c("KEGG", "Reactome"),
        include_situated_gene = include_situated_gene
      )
    }
  ))
}
for (run_id in names(runs)) {
  runs[[run_id]]$enrichment <- run_continuous_enrichment(runs[[run_id]])
}

# Significant hits (FDR <= 0.05 and R2 >= min_r_squared) of one enrichment
# result, for the run's high- and medium-confidence programs.
significant_hits <- function(res, run_id, valid, type, value_col, source = NULL) {
  if (is.null(res)) {
    return(NULL)
  }
  return(dplyr::bind_rows(lapply(res$by_program, function(x) {
    hits <- x$comparison
    if (!as.character(x$program) %in% valid || is.null(hits) || nrow(hits) == 0) {
      return(NULL)
    }
    keep <- is.finite(hits$fdr) & hits$fdr <= 0.05 &
      is.finite(hits$r_squared) & hits$r_squared >= min_r_squared
    if (!is.null(source)) {
      keep <- keep & hits$source == source
    }
    hits <- hits[keep, , drop = FALSE]
    if (nrow(hits) == 0) {
      return(NULL)
    }
    return(data.frame(
      run = run_id,
      program = as.character(x$program),
      type = type,
      annotation = hits[[value_col]],
      effect = hits$enrichment,
      r_squared = hits$r_squared,
      fdr = hits$fdr,
      stringsAsFactors = FALSE
    ))
  })))
}
enrichment_hits <- dplyr::bind_rows(lapply(names(runs), function(run_id) {
  r <- runs[[run_id]]
  valid <- valid_programs_of(r)
  e <- r$enrichment
  return(dplyr::bind_rows(
    significant_hits(e$categories, run_id, valid, "trait category", "trait_category"),
    significant_hits(e$tissues, run_id, valid, "tissue", "tissue"),
    significant_hits(e$pathways, run_id, valid, "Reactome", "description", "Reactome"),
    significant_hits(e$pathways, run_id, valid, "KEGG", "description", "KEGG")
  ))
}))
```

```{r run-program-match, fig.width = 8, fig.height = 7}
# Jaccard overlap of member SNPs between every pair of high/medium programs.
member_sets <- function(run_id) {
  r <- runs[[run_id]]
  valid <- valid_programs_of(r)
  ordered <- order_programs_by_confidence(
    r$summary$programs[as.character(r$summary$programs$program) %in% valid, , drop = FALSE]
  )
  a <- r$summary$assigned
  sets <- lapply(as.character(ordered$program), function(k) {
    return(unique(as.character(a$snp_id[as.character(a$program) == k])))
  })
  return(stats::setNames(sets, paste0(run_id, ordered$program)))
}
sets_a <- member_sets("A")
sets_b <- member_sets("B")
if (length(sets_a) == 0 || length(sets_b) == 0) {
  cat("A run has no high- or medium-confidence programs; skipping the match.\n")
} else {
  program_jaccard <- expand.grid(
    program_a = names(sets_a), program_b = names(sets_b), stringsAsFactors = FALSE
  )
  program_jaccard$jaccard <- mapply(function(a, b) {
    return(length(intersect(sets_a[[a]], sets_b[[b]])) / length(union(sets_a[[a]], sets_b[[b]])))
  }, program_jaccard$program_a, program_jaccard$program_b)
  print(
    program_jaccard |>
      dplyr::mutate(
        program_a = factor(program_a, levels = rev(names(sets_a))),
        program_b = factor(program_b, levels = names(sets_b))
      ) |>
      ggplot2::ggplot(ggplot2::aes(x = program_b, y = program_a, fill = jaccard)) +
      ggplot2::geom_tile(colour = "white") +
      ggplot2::geom_text(
        ggplot2::aes(label = ifelse(jaccard >= 0.1, sprintf("%.2f", jaccard), "")),
        size = 2.6
      ) +
      ggplot2::scale_fill_gradient(low = "white", high = "#b2182b", limits = c(0, 1), name = "Jaccard") +
      ggplot2::labs(
        title = "Matching programs across runs",
        subtitle = "Jaccard overlap of member SNPs; high/medium programs in confidence order; labels >= 0.1",
        x = "Run B program", y = "Run A program"
      ) +
      ggplot2::theme_minimal(base_size = 9) +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5))
  )
}
```

The first table counts, for each enrichment type, the annotations that are
significant in at least one high- or medium-confidence program of both runs,
of Run A only, or of Run B only. The second lists every such annotation with
the programs it is significant in, as `program (effect, R²)`.

```{r run-enrichment-comparison, results = "asis"}
if (nrow(enrichment_hits) == 0) {
  cat("No significant continuous enrichment in either run.\n")
} else {
  # One run's hits for an annotation, as "A3 (effect, R2)", largest effect first.
  format_run_hits <- function(run, program, effect, r_squared, which_run) {
    keep <- run == which_run
    if (!any(keep)) {
      return("")
    }
    o <- order(-effect[keep])
    return(paste0(
      which_run, program[keep][o],
      " (", signif(effect[keep][o], 2), ", ", signif(r_squared[keep][o], 2), ")",
      collapse = "<br>"
    ))
  }
  annotation_comparison <- enrichment_hits |>
    dplyr::group_by(type, annotation) |>
    dplyr::summarise(
      run_A = format_run_hits(run, program, effect, r_squared, "A"),
      run_B = format_run_hits(run, program, effect, r_squared, "B"),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      found_in = dplyr::case_when(
        run_A != "" & run_B != "" ~ "both",
        run_A != "" ~ "A only",
        TRUE ~ "B only"
      ),
      found_in = factor(found_in, levels = c("both", "A only", "B only")),
      type = factor(type, levels = enrichment_types)
    ) |>
    dplyr::arrange(type, found_in, annotation)

  enrichment_summary <- annotation_comparison |>
    dplyr::count(type, found_in, .drop = FALSE) |>
    tidyr::pivot_wider(names_from = found_in, values_from = n, values_fill = 0) |>
    dplyr::left_join(
      enrichment_hits |>
        dplyr::group_by(type = factor(type, levels = enrichment_types), run) |>
        dplyr::summarise(n = dplyr::n_distinct(program), .groups = "drop") |>
        tidyr::pivot_wider(names_from = run, values_from = n, names_prefix = "programs_with_hit_", values_fill = 0),
      by = "type"
    ) |>
    dplyr::mutate(dplyr::across(dplyr::starts_with("programs_with_hit_"), function(x) return(tidyr::replace_na(x, 0L))))
  print(knitr::kable(
    enrichment_summary,
    caption = paste0(
      "Significant continuous-enrichment annotations (FDR <= 0.05, R2 >= ",
      min_r_squared, ") by run. both / A only / B only count annotations; ",
      "programs_with_hit_A/B count high/medium programs with at least one hit. ",
      "Run A has ", length(sets_a), " high/medium programs and Run B ", length(sets_b), "."
    ),
    row.names = FALSE
  ))

  comparison_table <- knitr::kable(
    annotation_comparison |>
      dplyr::select(type, annotation, found_in, run_A, run_B),
    format = "html",
    escape = FALSE,
    row.names = FALSE,
    caption = "Every significant annotation and the programs it is significant in, as program (effect, R2)"
  )
  if (nrow(annotation_comparison) > 40) {
    cat("<details><summary>All ", nrow(annotation_comparison),
        " significant annotations (click to expand)</summary>\n\n", sep = "")
    print(comparison_table)
    cat("\n\n</details>\n")
  } else {
    print(comparison_table)
  }
}
```

## Step 5: Programs of each run

Steps 5–7 interpret the high- and medium-confidence programs of both runs, one
after the other, so the runs can be compared directly. The table below
summarises each run's programs; each run then gets a SNP-similarity heatmap
and an overlap plot.

```{r interpret-setup}
# High/medium programs of a run, as one row per (SNP, program) membership.
run_members <- function(run_id) {
  r <- runs[[run_id]]
  valid <- valid_programs_of(r)
  return(
    r$summary$assigned |>
      dplyr::filter(as.character(program) %in% valid) |>
      dplyr::transmute(snp_id = as.character(snp_id), group = as.character(program)) |>
      dplyr::distinct()
  )
}
members_by_run <- lapply(stats::setNames(names(runs), names(runs)), run_members)

# Run A's filtered matrix is written out, as before.
data.table::fwrite(runs$A$fit$x_matrix, file = "pleiotropy_matrix_filtered.csv", row.names = TRUE)

knitr::kable(
  dplyr::bind_rows(lapply(names(runs), function(run_id) {
    r <- runs[[run_id]]
    p <- r$summary$programs
    n_snps_total <- ncol(r$fit$x_matrix)
    n_unique_snps_all <- dplyr::n_distinct(r$summary$assigned$snp_id)
    n_unique_snps_valid <- dplyr::n_distinct(members_by_run[[run_id]]$snp_id)
    return(data.frame(
      run = r$label,
      n_snps = n_snps_total,
      n_programs = nrow(p),
      n_high = sum(p$confidence_tier == "high"),
      n_medium = sum(p$confidence_tier == "medium"),
      n_children = sum(!is.na(p$parent_program)),
      n_unique_snps_all = n_unique_snps_all,
      n_unique_snps_valid = n_unique_snps_valid,
      pct_of_total_snps_all = round(100 * n_unique_snps_all / n_snps_total, 1),
      pct_of_total_snps_valid = round(100 * n_unique_snps_valid / n_snps_total, 1),
      stringsAsFactors = FALSE
    ))
  })),
  caption = paste0(
    "Programs of each run. n_unique_snps counts each SNP once even when it ",
    "belongs to several programs (valid = high- and medium-confidence ",
    "programs); pct_of_total_snps is relative to the run's n_snps."
  ),
  row.names = FALSE
)
```

**SNP similarity.** `S` between the member SNPs of the valid programs, with
SNPs ordered by program, so each program should show as a block along the
diagonal.

**Program overlap.** Because EBMF programs are soft, the same SNP can belong
to several programs. The UpSet plot shows all valid programs, and for every
observed intersection how many SNPs are shared by exactly that combination (a
Venn diagram generalised to any number of programs). The left-hand bar chart
is each program's size; the dot matrix marks which programs each intersection
involves.

```{r step5-per-run, results = "asis", fig.height = 7, fig.width = 8}
plot_similarity_heatmap <- function(run_id) {
  members <- members_by_run[[run_id]]
  S_run <- runs[[run_id]]$fit$s_matrix
  if (nrow(members) < 2) {
    cat("Fewer than 2 SNPs in valid programs; skipping heatmap.\n\n")
    return(invisible(NULL))
  }
  first_program <- members[!duplicated(members$snp_id), , drop = FALSE]
  snp_order <- first_program$snp_id[order(as.integer(first_program$group))]
  snp_order <- intersect(snp_order, rownames(S_run))
  if (length(snp_order) < 2) {
    cat("Fewer than 2 valid-program SNPs are in the similarity matrix; skipping heatmap.\n\n")
    return(invisible(NULL))
  }
  heatmap(
    S_run[snp_order, snp_order, drop = FALSE],
    Rowv = NA,
    Colv = NA,
    scale = "none",
    main = paste0("SNP similarity, Run ", runs[[run_id]]$label),
    xlab = "SNPs",
    ylab = "SNPs"
  )
  return(invisible(NULL))
}

plot_program_overlap <- function(run_id) {
  members <- members_by_run[[run_id]]
  n_programs <- dplyr::n_distinct(members$group)
  if (n_programs < 2) {
    cat("Fewer than 2 valid programs; skipping the overlap plot.\n\n")
    return(invisible(NULL))
  }
  membership <- members |>
    dplyr::mutate(value = 1L) |>
    tidyr::pivot_wider(
      id_cols = "snp_id",
      names_from = "group",
      names_prefix = run_id,
      values_from = "value",
      values_fill = 0L
    )
  set_order <- names(sort(colSums(membership[, -1, drop = FALSE]), decreasing = TRUE))
  if (n_programs > 40) {
    cat("More than 40 programs; showing the 40 largest only.\n\n")
    set_order <- set_order[seq_len(40)]
  }
  plot_ok <- tryCatch({
    membership_plot <- as.data.frame(membership[, c("snp_id", set_order), drop = FALSE])
    rownames(membership_plot) <- membership_plot$snp_id
    membership_plot$snp_id <- NULL
    print(UpSetR::upset(
      membership_plot,
      sets = set_order,
      order.by = "freq",
      nsets = length(set_order),
      nintersects = 100,
      main.bar.color = "#1b9e77",
      sets.bar.color = "#7570b3",
      point.size = 2.6,
      line.size = 0.6
    ))
    TRUE
  }, error = function(e) {
    cat("UpSet plot failed (", conditionMessage(e),
        "); showing pairwise SNP-overlap counts instead.\n\n")
    return(FALSE)
  })
  if (!plot_ok) {
    sizes <- colSums(membership[, set_order, drop = FALSE])
    overlaps <- outer(set_order, set_order, Vectorize(function(a, b) {
      return(sum(membership[[a]] & membership[[b]]))
    }))
    dimnames(overlaps) <- list(set_order, set_order)
    diag(overlaps) <- sizes[set_order]
    print(knitr::kable(
      as.data.frame(overlaps),
      row.names = TRUE,
      caption = "Pairwise SNP overlap between programs (diagonal = program size)"
    ))
  }
  return(invisible(NULL))
}

for (run_id in names(runs)) {
  cat("\n\n### Run ", runs[[run_id]]$label, "\n\n", sep = "")
  plot_similarity_heatmap(run_id)
  cat("\n\n")
  plot_program_overlap(run_id)
  cat("\n\n")
}
```

## Step 6: Module-specific traits and continuous enrichment

One table per run, with one row per high- or medium-confidence program in
confidence order (green high, orange medium) and each child indented under
its parent. Program numbers are prefixed with the run (A3, B3), as in Step 4.

- **module_specific_traits:** the background traits most elevated in the
  program relative to the rest of the BMI SNPs
  (`summarise_module_specific_traits()`), with
  `specificity = mean |z| in the program / (mean |z| in the rest + 0.1)` in
  brackets. Traits need a specificity of at least `r min_specificity`, and
  the top 6 are listed. This highlights traits specifically elevated in the
  program rather than merely large everywhere.
- **n_unique_snps:** member SNPs not in any other high/medium program of the
  same run.
- **trait_categories, tissues, reactome, kegg:** the continuous loading-based
  hits from Step 4 (FDR ≤ 0.05 and R² ≥ 0.05), largest effect first, as
  `(effect, R², FDR)`.

```{r step6-module-overview, results = "asis"}
# Step 4 hits of one run and type, collapsed to one cell per program.
format_hits <- function(run_id, type_name) {
  hits <- enrichment_hits[enrichment_hits$run == run_id & enrichment_hits$type == type_name, , drop = FALSE]
  if (nrow(hits) == 0) {
    return(data.frame(program = character(0), value = character(0)))
  }
  return(
    hits |>
      dplyr::arrange(program, dplyr::desc(effect)) |>
      dplyr::group_by(program) |>
      dplyr::summarise(
        value = paste0(
          annotation, " (", signif(effect, 2), ", ", signif(r_squared, 2), ", ",
          signif(fdr, 3), ")",
          collapse = "<br>"
        ),
        .groups = "drop"
      )
  )
}

module_overview <- function(run_id) {
  r <- runs[[run_id]]
  members <- members_by_run[[run_id]]
  group_traits <- summarise_module_specific_traits(
    trait_matrix = r$fit$trait_matrix,
    groups = members,
    trait_info = r$fit$trait_info,
    coloc_groups = r$coloc_groups,
    exclude_trait_ids = target_trait_id,
    min_group_size = min_module_size,
    top_n = top_n_traits,
    n_categories = 3L,
    min_specificity = min_specificity
  )
  specific_traits <- dplyr::bind_rows(lapply(group_traits$by_group, function(x) {
    passing <- x$traits |>
      dplyr::filter(passes_min_specificity) |>
      utils::head(6L)
    return(data.frame(
      program = as.character(x$group),
      module_specific_traits = if (nrow(passing) == 0) {
        "none"
      } else {
        paste0(passing$trait_name, " (", round(passing$specificity, 2), ")", collapse = "<br>")
      },
      stringsAsFactors = FALSE
    ))
  }))
  # SNPs that belong to exactly one program: not shared with any other program.
  snp_group_counts <- dplyr::count(members, snp_id, name = "n_groups")
  unique_by_group <- members |>
    dplyr::filter(snp_id %in% snp_group_counts$snp_id[snp_group_counts$n_groups == 1]) |>
    dplyr::count(program = group, name = "n_unique_snps")

  return(
    r$summary$programs |>
      dplyr::filter(as.character(program) %in% valid_programs_of(r)) |>
      dplyr::transmute(
        program = as.character(program),
        confidence_tier = as.character(confidence_tier),
        similarity_z,
        n_snps_filtered,
        parent_program
      ) |>
      dplyr::left_join(dplyr::count(members, program = group, name = "n_snps"), by = "program") |>
      dplyr::left_join(unique_by_group, by = "program") |>
      dplyr::left_join(specific_traits, by = "program") |>
      dplyr::left_join(dplyr::rename(format_hits(run_id, "trait category"), trait_categories = value), by = "program") |>
      dplyr::left_join(dplyr::rename(format_hits(run_id, "tissue"), tissues = value), by = "program") |>
      dplyr::left_join(dplyr::rename(format_hits(run_id, "Reactome"), reactome = value), by = "program") |>
      dplyr::left_join(dplyr::rename(format_hits(run_id, "KEGG"), kegg = value), by = "program") |>
      dplyr::mutate(
        n_unique_snps = as.integer(tidyr::replace_na(n_unique_snps, 0L)),
        dplyr::across(
          c(module_specific_traits, trait_categories, tissues, reactome, kegg),
          function(x) return(tidyr::replace_na(x, "none"))
        )
      ) |>
      order_programs_by_confidence() |>
      dplyr::transmute(
        program = sub("([0-9]+)$", paste0(run_id, "\\1"), program_label),
        confidence_tier, n_snps, n_unique_snps, module_specific_traits,
        trait_categories, tissues, reactome, kegg
      )
  )
}

for (run_id in names(runs)) {
  cat("\n\n### Run ", runs[[run_id]]$label, "\n\n", sep = "")
  if (nrow(members_by_run[[run_id]]) == 0) {
    cat("No high- or medium-confidence programs to interpret.\n\n")
    next
  }
  print(kable_confidence(
    module_overview(run_id),
    drop_tier_col = TRUE,
    caption = paste0(
      "Run ", runs[[run_id]]$label, ": top 6 module-specific traits ",
      "(specificity) and continuous enrichment hits (effect, R2, FDR) per ",
      "high/medium program. n_unique_snps = member SNPs in no other program."
    )
  ))
}
```

## Step 7: Pleiotropy of program-unique vs shared SNPs

Because EBMF programs are soft, a SNP can be assigned to exactly one valid
program (**unique**) or to several (**shared**). If programs capture distinct
biological programmes, SNPs that uniquely define a single program should be
more specific, i.e. carry a lower pleiotropy score (fewer distinct trait
categories and protein-coding genes via coloc groups) than SNPs whose signal is
shared across programs. This is tested in each run with
`get_all_variant_pleiotropies()`.

Gene pleiotropy has a long tail: most SNPs reach 1–4 genes, while a few reach
hundreds. The y-axis is therefore on a **pseudo-log scale**, which is linear
near 0 and logarithmic above it, so the bulk of SNPs stays readable and the
tail is squashed rather than cut off. Boxes and medians are unaffected by the
scale; the diamond marks each group's mean.

```{r step7-unique-vs-shared-pleiotropy, results = "asis", fig.width = 9, fig.height = 7, out.width = "100%"}
variant_pleiotropy <- get_all_variant_pleiotropies() |>
  dplyr::mutate(variant_id = as.character(variant_id)) |>
  dplyr::select(variant_id, distinct_trait_categories, distinct_protein_coding_genes)

pleio <- dplyr::bind_rows(lapply(names(runs), function(run_id) {
  members <- members_by_run[[run_id]]
  if (nrow(members) == 0) {
    return(NULL)
  }
  return(
    members |>
      dplyr::count(snp_id, name = "n_programs") |>
      dplyr::mutate(
        run = runs[[run_id]]$label,
        membership_class = dplyr::if_else(n_programs == 1, "unique", "shared")
      )
  )
})) |>
  dplyr::left_join(variant_pleiotropy, by = c("snp_id" = "variant_id")) |>
  tidyr::pivot_longer(
    c(distinct_trait_categories, distinct_protein_coding_genes),
    names_to = "metric",
    values_to = "pleiotropy"
  ) |>
  dplyr::mutate(
    metric = dplyr::if_else(
      metric == "distinct_trait_categories",
      "trait-category pleiotropy (distinct categories per SNP)",
      "gene pleiotropy (distinct genes per SNP)"
    ),
    membership_class = factor(membership_class, levels = c("unique", "shared"))
  )

if (nrow(pleio) == 0) {
  cat("No valid programs to classify SNPs.\n")
} else {
  tests <- pleio |>
    dplyr::group_by(run, metric) |>
    dplyr::summarise(
      wilcoxon_p = tryCatch(
        signif(stats::wilcox.test(pleiotropy ~ membership_class)$p.value, 3),
        error = function(e) return(NA_real_)
      ),
      .groups = "drop"
    )
  print(knitr::kable(
    pleio |>
      dplyr::group_by(run, metric, membership_class) |>
      dplyr::summarise(
        n_snps = sum(!is.na(pleiotropy)),
        mean_pleiotropy = round(mean(pleiotropy, na.rm = TRUE), 2),
        median_pleiotropy = round(stats::median(pleiotropy, na.rm = TRUE), 2),
        max_pleiotropy = max(pleiotropy, na.rm = TRUE),
        .groups = "drop"
      ) |>
      dplyr::left_join(tests, by = c("run", "metric")),
    row.names = FALSE,
    caption = paste0(
      "Pleiotropy of program-unique vs shared SNPs of the valid programs, by ",
      "run. wilcoxon_p compares unique with shared SNPs within each run and ",
      "metric (Wilcoxon rank-sum test)."
    )
  ))

  # Means on the original scale: stat_summary() would average the
  # pseudo-log-transformed values instead.
  pleio_means <- pleio |>
    dplyr::group_by(run, metric, membership_class) |>
    dplyr::summarise(pleiotropy = mean(pleiotropy, na.rm = TRUE), .groups = "drop")
  print(
    ggplot2::ggplot(
      pleio |> dplyr::filter(!is.na(pleiotropy)),
      ggplot2::aes(x = membership_class, y = pleiotropy, fill = membership_class)
    ) +
      ggplot2::geom_boxplot(alpha = 0.6, outlier.shape = NA) +
      ggplot2::geom_jitter(width = 0.15, height = 0, alpha = 0.35, size = 0.8) +
      ggplot2::geom_point(data = pleio_means, shape = 23, size = 3, fill = "white") +
      ggplot2::facet_grid(metric ~ run, scales = "free_y") +
      ggplot2::scale_y_continuous(
        transform = "pseudo_log",
        breaks = c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000)
      ) +
      ggplot2::scale_fill_manual(values = c(unique = "#1b9e77", shared = "#7570b3")) +
      ggplot2::labs(
        title = "Pleiotropy of program-unique vs shared SNPs",
        subtitle = paste0(
          "unique = in exactly one valid program; shared = in >= 2. ",
          "Pseudo-log y-axis; diamond = mean"
        ),
        x = NULL,
        y = "pleiotropy score"
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(legend.position = "none")
  )
}
```

## Why the body-size traits change the results so much

Run B drops `r length(body_size_trait_ids)` of Run A's
`r nrow(runs$A$fit$x_matrix) - 1L` background traits, yet the two runs
disagree on almost everything: how alike two random BMI SNPs look,
which programs pass, and whether any program has tissue or pathway
enrichment. That other BMI GWASes have to go is expected, since they repeat
the anchor trait at nearly every SNP. That the body-size traits matter as much
is less obvious, because weight, fat mass or hip circumference each have GWAS
hits that BMI does not share.

In short, at BMI's own SNPs the body-size traits are close to being BMI
measured again, and there are forty of them. Three things act together, and
the table puts numbers on each.

```{r body-size-weight}
# Run A's oriented, compressed matrix (BMI row left out), split into the
# body-size traits and the rest.
x_a <- runs$A$fit$x_star
bmi_id <- as.character(target_trait_id)
z_bmi <- runs$A$fit$x_matrix[bmi_id, ]
background_ids <- setdiff(rownames(x_a), bmi_id)
body_ids <- intersect(background_ids, body_size_trait_ids)
other_ids <- setdiff(background_ids, body_ids)
# The other traits observed at the most BMI SNPs, as many as there are
# body-size traits, for a like-for-like comparison of the two set-level checks.
dense_other_ids <- other_ids[
  order(-rowSums(!is.na(x_a[other_ids, , drop = FALSE])))
][seq_along(body_ids)]
trait_names <- stats::setNames(
  as.character(runs$A$fit$trait_info$trait_name),
  as.character(runs$A$fit$trait_info$trait_id)
)

n_observed_per_snp <- function(ids) {
  return(colSums(!is.na(x_a[ids, , drop = FALSE])))
}
# Share of a trait set's signal on its first component, missing cells as 0
# (as in S).
first_component_share <- function(ids) {
  m <- x_a[ids, , drop = FALSE]
  m[is.na(m)] <- 0
  d <- svd(m, nu = 0, nv = 0)$d
  return(d[1]^2 / sum(d^2))
}
# Median correlation between two traits over the BMI SNPs both are observed at.
median_trait_correlation <- function(ids) {
  r <- suppressWarnings(
    stats::cor(t(x_a[ids, , drop = FALSE]), use = "pairwise.complete.obs")
  )
  return(stats::median(r[upper.tri(r)], na.rm = TRUE))
}
# Mean S between BMI SNPs on these traits alone, computed as in Step 2.
mean_similarity_on <- function(ids) {
  s <- withCallingHandlers(
    snp_similarity_matrix(x_a[ids, , drop = FALSE]),
    warning = function(w) {
      if (grepl("zero norm", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
  return(.similarity_background(
    s$s_matrix, s$overlap_matrix, similarity_threshold
  )$mean_similarity)
}

n_body_per_snp <- n_observed_per_snp(body_ids)
n_other_per_snp <- n_observed_per_snp(other_ids)
cells <- rowSums(!is.na(x_a[background_ids, , drop = FALSE]))
sum_sq <- rowSums(x_a[background_ids, , drop = FALSE]^2, na.rm = TRUE)
height_ids <- intersect(
  names(trait_names)[trait_names %in% c("Height", "Standing height")],
  background_ids
)
height_pct_positive <- vapply(height_ids, function(i) {
  return(100 * mean(x_a[i, ] > 0, na.rm = TRUE))
}, numeric(1))

# How much of Run A's factorisation rests on the body-size traits.
flash_a <- runs$A$fit$cluster_details$flash_fit
loadings_a <- flash_a$L_pm[background_ids, , drop = FALSE]
body_share_of <- function(k) {
  l2 <- loadings_a[, k]^2
  return(sum(l2[body_ids]) / sum(l2))
}
body_factor <- vapply(seq_len(flash_a$n_factors), body_share_of, numeric(1)) > 0.5
programs_a <- runs$A$summary$programs |>
  dplyr::filter(as.character(program) %in% valid_programs_of(runs$A)) |>
  order_programs_by_confidence()
body_size_programs <- dplyr::bind_rows(lapply(seq_len(nrow(programs_a)), function(i) {
  k <- as.integer(as.character(programs_a$program[i]))
  snps <- members_by_run$A$snp_id[members_by_run$A$group == as.character(k)]
  top <- names(sort(abs(loadings_a[, k]), decreasing = TRUE))[1:3]
  return(data.frame(
    program = sub("([0-9]+)$", "A\\1", programs_a$program_label[i]),
    confidence_tier = as.character(programs_a$confidence_tier[i]),
    n_snps = length(snps),
    body_size_share = 100 * body_share_of(k),
    top_traits = paste(trait_names[top], collapse = "<br>"),
    body_size_traits_per_snp = stats::median(n_body_per_snp[snps], na.rm = TRUE),
    bmi_abs_z = round(stats::median(abs(z_bmi[snps]), na.rm = TRUE), 1),
    stringsAsFactors = FALSE
  ))
}))

n_hits <- function(run_id, types) {
  if (nrow(enrichment_hits) == 0) {
    return(0L)
  }
  return(dplyr::n_distinct(
    enrichment_hits$annotation[enrichment_hits$run == run_id & enrichment_hits$type %in% types]
  ))
}
coverage <- function(run_id) {
  return(100 * dplyr::n_distinct(members_by_run[[run_id]]$snp_id) / ncol(runs[[run_id]]$fit$x_matrix))
}
concordance_pct <- body_size_concordance$pct_positive_at_bmi_raising
comparison_a <- comparison[comparison$run == runs$A$label, , drop = FALSE]

bs <- list(
  n_body = length(body_ids),
  n_background = length(background_ids),
  n_snps = ncol(x_a),
  pct_rows = 100 * length(body_ids) / length(background_ids),
  pct_cells = 100 * sum(cells[body_ids]) / sum(cells),
  pct_sum_sq = 100 * sum(sum_sq[body_ids]) / sum(sum_sq),
  snps_with_body = sum(n_body_per_snp > 0),
  median_body_per_snp = stats::median(n_body_per_snp),
  min_concordance = min(pmax(concordance_pct, 100 - concordance_pct)),
  rho_body = stats::cor(n_body_per_snp, abs(z_bmi), method = "spearman"),
  rho_other = stats::cor(n_other_per_snp, abs(z_bmi), method = "spearman"),
  cor_body = median_trait_correlation(body_ids),
  pc1_body = 100 * first_component_share(body_ids),
  pc1_dense_other = 100 * first_component_share(dense_other_ids),
  s_body = mean_similarity_on(body_ids),
  s_all = mean_similarity_on(background_ids),
  s_other = mean_similarity_on(other_ids),
  excess_a = comparison_a$median_excess_similarity,
  n_low_a = comparison_a$n_low,
  n_programs_a = comparison_a$n_programs,
  n_factors = flash_a$n_factors,
  n_body_factors = sum(body_factor),
  pve_body = 100 * sum(flash_a$pve[body_factor]),
  pve_all = 100 * sum(flash_a$pve),
  n_valid_a = nrow(body_size_programs),
  n_valid_a_body = sum(body_size_programs$body_size_share >= 90),
  n_tissue_pathway_a = n_hits("A", c("tissue", "Reactome", "KEGG")),
  n_reactome_b = n_hits("B", "Reactome"),
  n_kegg_b = n_hits("B", "KEGG"),
  height_pct_min = min(height_pct_positive),
  height_pct_max = max(height_pct_positive),
  empty_without_body = sum(n_other_per_snp == 0),
  coverage_a = coverage("A"),
  coverage_b = coverage("B")
)

knitr::kable(
  data.frame(
    measure = c(
      "Background traits",
      "Share of the matrix's observed cells",
      "Share of the matrix's sum of squares",
      "Mean |value| of an observed cell",
      "BMI SNPs where at least one is observed",
      "Traits observed at a typical BMI SNP (median)",
      "Rank correlation of that number with BMI's own |z|",
      "Median correlation between two of the traits, over SNPs both are observed at *",
      "Share of the traits' signal in their first component *",
      "Mean S between BMI SNPs, on these traits alone"
    ),
    body_size_traits = c(
      bs$n_body,
      sprintf("%.0f%%", bs$pct_cells),
      sprintf("%.0f%%", bs$pct_sum_sq),
      sprintf("%.1f", mean(abs(x_a[body_ids, ]), na.rm = TRUE)),
      bs$snps_with_body,
      bs$median_body_per_snp,
      sprintf("%.2f", bs$rho_body),
      sprintf("%.2f", bs$cor_body),
      sprintf("%.0f%%", bs$pc1_body),
      sprintf("%.3f", bs$s_body)
    ),
    other_traits = c(
      length(other_ids),
      sprintf("%.0f%%", 100 - bs$pct_cells),
      sprintf("%.0f%%", 100 - bs$pct_sum_sq),
      sprintf("%.1f", mean(abs(x_a[other_ids, ]), na.rm = TRUE)),
      sum(n_other_per_snp > 0),
      stats::median(n_other_per_snp),
      sprintf("%.2f", bs$rho_other),
      sprintf("%.2f", median_trait_correlation(dense_other_ids)),
      sprintf("%.0f%%", bs$pc1_dense_other),
      sprintf("%.3f", bs$s_other)
    ),
    stringsAsFactors = FALSE
  ),
  caption = paste0(
    "The ", bs$n_body, " body-size traits against the rest of Run A's background ",
    "traits, on Run A's oriented, compressed matrix (BMI row left out). ",
    "* other traits: the ", bs$n_body, " observed at the most BMI SNPs, for a ",
    "like-for-like comparison."
  ),
  row.names = FALSE
)
```

**The matrix only sees BMI's SNPs.** Its columns are BMI SNPs, so a
background trait enters only through its effects at those SNPs. Whether a
body-size trait shares BMI's hits genome-wide does not matter: its hits
elsewhere never reach the matrix. At BMI's SNPs the body-size traits are
almost everywhere. At least one of them is observed at
`r bs$snps_with_body` of the `r bs$n_snps` SNPs, and
`r bs$median_body_per_snp` at a typical one. Step 1 shows each moving with
BMI at `r bs$min_concordance`–100% of the SNPs it shares (impedance moves
against it, just as consistently).

**At those SNPs they are BMI measured again.** BMI is weight over height
squared, weight is fat mass plus fat-free mass, and waist and hip are girths
of the same body. A variant that raises BMI raises these as part of the same
change in the body, not through a separate route. The data say the same:

- Two body-size traits typically correlate at
  `r sprintf("%.2f", bs$cor_body)` across the BMI SNPs they share.
- One component carries `r round(bs$pc1_body)`% of their combined signal,
  against `r round(bs$pc1_dense_other)`% for a matched set of other traits.
- How many of them reach significance at a SNP follows the strength of BMI's
  own signal there: the rank correlation with BMI's |z| is
  `r sprintf("%.2f", bs$rho_body)`, against
  `r sprintf("%.2f", bs$rho_other)` for the other traits.

Many are literally one measurement. 29 of the 40 come from UK Biobank's
bioimpedance measurement in the same 455,000 or so people: fat mass, fat
percentage, fat-free mass, predicted mass and impedance of each arm, each leg,
the trunk and the whole body, with left and right limbs as separate traits.
Waist and hip circumference each appear twice, from two analyses of the same
UK Biobank measurement.

**And there are forty of them.** One extra BMI-like trait would add little.
Forty count the same measurement forty times, both in `S` and in
the factorisation. In Run A the body-size traits are
`r round(bs$pct_rows, 1)`% of the background traits but hold
`r round(bs$pct_cells)`% of the observed cells and
`r round(bs$pct_sum_sq)`% of the sum of squares. That is not because their
values are larger (an observed cell is no larger on average than another
trait's) but because they are observed at so many SNPs. EBMF fits only the
observed cells, so about two-thirds of what it fits in Run A is body size.
The BMI GWASes removed from both runs (Step 1) are the same problem in its
purest form; the body-size traits are forty slightly weaker copies of it.

**What this does to Run A.**

- **Random BMI SNPs already look alike.** `S` is a cosine over traits, so
  two SNPs that both carry a few body-size traits share several large entries
  of the same sign and look alike whatever else they do. On the body-size
  traits alone, the mean `S` between two BMI SNPs is
  `r sprintf("%.3f", bs$s_body)`, against `r sprintf("%.3f", bs$s_all)` on
  all of Run A's traits and `r sprintf("%.3f", bs$s_other)` without them.
  The confidence tier's similarity test compares each program with random
  groups of BMI SNPs, and in Run A those random groups already look alike, so
  real structure has little room to stand out. Run A's median excess similarity is
  `r bs$excess_a`, and `r bs$n_low_a` of its `r bs$n_programs_a` programs are
  low.
- **The factorisation spends itself on body size.** Of Run A's
  `r bs$n_factors` factors, the `r bs$n_body_factors` that put most of their
  trait loading on body-size traits explain `r round(bs$pve_body, 1)`% of the
  matrix variance, out of `r round(bs$pve_all, 1)`% for all factors together.
  In the table below, `r bs$n_valid_a_body` of Run A's `r bs$n_valid_a` high-
  and medium-confidence programs put 90% or more of their trait loading on
  body-size traits. Most draw on SNPs that carry more body-size traits than a
  typical BMI SNP, which are the stronger BMI signals.

```{r body-size-programs, results = "asis"}
print(kable_confidence(
  dplyr::bind_rows(
    data.frame(
      program = "all BMI SNPs", confidence_tier = "background", n_snps = bs$n_snps,
      body_size_share = "", top_traits = "",
      body_size_traits_per_snp = stats::median(n_body_per_snp),
      bmi_abs_z = round(stats::median(abs(z_bmi)), 1),
      stringsAsFactors = FALSE
    ),
    body_size_programs |>
      dplyr::mutate(body_size_share = sprintf("%.0f%%", body_size_share))
  ),
  drop_tier_col = TRUE,
  caption = paste0(
    "Run A's high- and medium-confidence programs. body_size_share = share of ",
    "the program's squared trait loadings (BMI row left out) on body-size ",
    "traits; top_traits = its three largest trait loadings; ",
    "body_size_traits_per_snp and bmi_abs_z = medians over its member SNPs. ",
    "The first row is the same over all BMI SNPs."
  )
))
```

These programs are not noise. They split BMI SNPs by which body measures move
most (`top_traits`), which is real biology. But it is variation within body
size, carried by forty near-copies of one measurement, and it says little
about mechanism. Run A's programs have `r bs$n_tissue_pathway_a` tissue or
pathway hits between them; Run B's have `r bs$n_reactome_b` Reactome and
`r bs$n_kegg_b` KEGG hits. Traits that BMI itself moves, such as cholesterol
and glucose, probably act the same way in weaker form: they reach
significance more often at the strongest BMI SNPs, which would explain why
they top Step 6's specific traits for Run A's two largest programs.

**It is not body size as such.** Height, sitting height and appendicular lean
mass are body measurements too, and were kept. At BMI SNPs they go both ways:
the two height GWASes, for example, are positive at the BMI-raising allele at
`r round(bs$height_pct_min)`–`r round(bs$height_pct_max)`% of the SNPs they
share, close to an even split. So they separate BMI SNPs rather than make them alike, and height
shows up among the specific traits of Run B's programs. The problem is a
trait that moves with the anchor trait at almost every SNP; for BMI, the
body-size traits are the large family that does.

**What removing them costs.** `r bs$empty_without_body` of the
`r bs$n_snps` BMI SNPs have no observation outside the body-size traits. In
Run B their profile is empty and they cannot join any program. That is why
Run B's valid programs cover `r round(bs$coverage_b, 1)`% of the SNPs (Run A:
`r round(bs$coverage_a, 1)`%), and why two SNPs in Run B typically share no
observed trait at all. So Run B is not a clean answer either: it describes the
SNPs that have something to say beyond body size, and is silent about the
rest. It also gives up any body-composition biology (fat against lean mass,
fat distribution) that only the removed traits carried.

**What this says about the method.** The similarity tests and the
factorisation both weigh a phenotype by how many times the database happens
to measure it. Neither can tell a re-measurement of the anchor trait from an
independent trait that shares its SNPs. Any anchor with a large family of
close, consistently signed measurements will behave this way: glycaemic
traits and diabetes definitions for type 2 diabetes, or other lipid fractions
for a lipid trait. Which traits count as the anchor measured again is
therefore a decision made before the factorisation, here by the rule in
Step 1, and this vignette shows how much rests on it.
