---
title: "SNP biological clustering"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{SNP biological clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

## Goal

Take a **single trait** and decompose its genetic architecture by grouping SNPs
according to their phenome-wide pleiotropic profiles.

The central question is: **within one trait, what distinct biological programs
are represented by its associated SNPs?**

The output is a set of SNP programs (latent factors), each annotated with
background traits, trait categories, tissues and pathways. The runnable
implementation is
[Investigation of univariate analysis](investigation-univariate-analysis.html),
and [Suzuki T2D](investigation-suzuki-t2d.html) benchmarks the programs against
a published clustering of type 2 diabetes loci. The same single-trait programs
are also the input to a **multi-trait** analysis that links the programs of
different traits when they are driven by the same background traits and,
where the traits share loci, carried by the same SNPs (Step 6 below, and
[Multi-trait programs](investigation-multitrait-programs.html)).

Background traits stay at **study level**. Molecular QTL studies are not
collapsed by gene, tissue, cis/trans, or QTL type.

### Core clustering preferences

The grouping method should satisfy:

1. **Handles high sparsity** — Must natively process large, sparse matrices
   (thousands of singleton feature rows) without treating structural zeros as
   low-variance signal. EBMF does this by leaving missing cells out of the fit
   (`flashier` skips `NA`); we additionally limit sparsity with a hard threshold
   on the number of non-missing SNPs per trait (`min_snp_signals`).
2. **Signed and sign-driven clustering** — Must preserve the meaning of
   continuous positive/negative z-scores so variants group by **directional
   alignment**, not magnitude alone. We orient every SNP column by
   \(\mathrm{sign}(z_{\mathrm{target}})\) before factorization.
3. **Automated \(K\) discovery** — The engine should learn the number of
   clusters or latent factors from the data. EBMF adds factors greedily while
   the variational lower bound (ELBO) improves, then backfits; filtering the
   resulting factors by an lFSR/magnitude membership gate and by the
   validation checks below controls the final program count.
4. **Overlapping membership** — A SNP can belong to more than one biological
   program. EBMF produces soft factor loadings, so membership is per-SNP-per-
   program rather than a single hard label.

```{r workflow-diagram, echo=FALSE, fig.height=8, fig.width=8}
workflow_dot <- "
digraph workflow {
  graph [rankdir=TB bgcolor=transparent fontname=Helvetica nodesep=0.35 ranksep=0.45]
  node [shape=box style=filled fontname=Helvetica fontsize=10 margin=0.12
        fillcolor=\"#e1f5fe\" color=\"#01579b\"]
  S1 [label=\"1. Pleiotropy matrix\\n(study-level rows x SNPs)\\nremove target proxies and near-duplicate traits\"]
  S2 [label=\"2. Orient by sign(z_target)\\noptional asinh compression\"]
  S3 [label=\"3. EBMF factorization Y = L F^T + E\\nlFSR + magnitude membership gate\"]
  S4 [label=\"4. Confidence tier\\nsimilarity vs random SNP sets, then trait breadth, then connectedness\"]
  S5 [label=\"5. Interpret high/medium programs\\nmodule-specific traits, continuous enrichment\"]
  S1 -> S2 -> S3 -> S4 -> S5
}
"
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(workflow_dot)
}
```

## Steps

The investigation vignettes take their settings from
`vignettes/_ebmf_settings.yml`, so the simulation, null-simulation, univariate
and multi-trait studies cannot drift apart. The values quoted below are the
ones in that file; a vignette parameter supplied at render time overrides them.
The package functions have their own, older defaults, which differ in places.

### 1. Construct the pleiotropy matrix

Given:

- `nsnp` = independent SNPs associated with the target trait (coloc groups)
- `ntrait` = background studies in GPMap (phenotypic and molecular)

Construct a **traits x SNPs** matrix of signed z-scores (\(z = \beta / \mathrm{SE}\)).
Conceptually this is a SNP-by-trait profile matrix; the code stores traits as
rows and SNPs as columns.

```text
              SNP1   SNP2   SNP3   SNP4   ...
Trait A        z      z      z      z
Trait B        z      z      z      z
Trait C        z      z      z      z
...
```

Which rows and columns back this matrix is controlled by these choices:

- **`include_trans`** — whether trans-QTL association rows
  (`cis_trans == "trans"`) are kept. Including them (canonical setting) can
  surface distributed regulatory programs at the cost of density and noise;
  excluding them keeps the analysis cis-focused.
- **`trait_subset`** — `"all"` keeps every available background trait;
  `"phenotypic"` (canonical) keeps only phenotype traits (rows whose
  `feature_type` is `"phenotypic"` or missing) plus the target trait, dropping
  molecular / gene-annotated rows. Molecular traits are much sparser, so
  clustering on phenotype traits only can recover structure that sparse
  molecular rows dilute.
- **`min_snp_signals`** — rows observed in fewer than this many SNPs are
  dropped (canonical 2); ubiquitous rows (non-missing on more than
  `max_snp_fraction` of SNPs) are dropped too. The target trait row is always
  kept.
- **Duplicate traits** — other GWAS of the target trait, and background traits
  that repeat one another, are removed by hand before the matrix is built. This
  is the most consequential choice of the five; it has its own section below.

#### Removing duplicate traits

GPMap often holds several GWAS of the same trait, and many traits that are
near-copies of each other at a given trait's loci. Two kinds matter.

**Duplicates of the anchor trait.** Other GWAS of the target trait, and traits
defined by it, repeat the target row at every SNP. For BMI these are the other
BMI GWAS and obesity diagnoses; for T2D, the other T2D GWAS, diabetes
diagnoses, diabetes medication, family history and complications (see the
"T2D proxy traits" section of [Suzuki T2D](investigation-suzuki-t2d.html)).
After orientation each of them is positive at almost every SNP, so it adds the
same component to every SNP's profile. The target row itself is left out of
`S` for exactly this reason (Step 4), and a proxy left in the matrix brings the
problem back. Proxies also give EBMF extra weight to spend on small
differences between near-identical rows.

**Duplicates within the background.** Two cases:

- **The same trait measured several ways.** For example, a consortium GWAS and
  the UK Biobank data field of the same trait, or the SPA- and Firth-corrected
  versions of one UK Biobank binary trait. In the Suzuki T2D analysis the top
  module-specific traits of one program were "High blood pressure (UKB data
  field 6150)" twice, once with each correction.
- **Traits that move with the anchor almost everywhere.** For BMI, body-size
  and body-composition measures (weight, fat mass, fat-free mass, impedance,
  hip circumference, ...) have the same sign at nearly every BMI SNP.

Either kind makes two random SNPs look alike, because they share the same block
of near-identical, same-signed rows. That background similarity hides the
structure programs are meant to capture, and it raises the bar for the
similarity test, which compares each program with random SNP sets from the
same matrix (Step 4). Near-duplicates also inflate `effective_n_traits`: each
copy counts as a separate trait, so a program carried by five versions of one
trait does not look like a single-trait program.

**Evidence from BMI.**
[Investigation of univariate analysis](investigation-univariate-analysis.html)
fits BMI twice. Run A removes only the 4 BMI proxies. Run B also removes 40
body-size traits, chosen by a rule: anthropometric or body-composition traits
observed at 10 or more BMI SNPs whose effect has the same sign at 80% or more
of them after orientation. Waist-hip ratio meets the rule but is kept, because
fat distribution is what separates obesity sub-programs. At the time of
writing (local GPMap build):

| | Run A: proxies removed | Run B: + 40 body-size traits removed |
|---|---|---|
| Background traits | 719 | 679 |
| Mean `S` over all SNP pairs | 0.246 | 0.011 |
| Share of SNP pairs with `S >= 0.2` | 0.51 | 0.035 |
| High / medium programs | 3 / 2 | 7 / 6 |
| Low programs | 40 | 17 |
| Annotations with continuous enrichment (categories / Reactome / KEGG) | 2 / 0 / 0 | 7 / 104 / 48 |

Removing 40 of 719 rows (6%) cut the background similarity more than
twentyfold, more than doubled the number of usable programs, and is the
difference between no pathway signal and 152 pathway hits. In Run A, half of
all random SNP pairs were already "connected", so almost no program could show
that its SNPs were more alike than chance.

**The cost.** Removing traits also removes observations. In Run B the matrix
is 1.0% observed against 2.5% in Run A, and the median pair of SNPs shares no
observed trait (2 in Run A), so many similarities rest on few traits. A
removed trait can also carry real biology. The aim is to remove the axis every
SNP shares, and to keep traits that differ between SNPs, such as fat
distribution.

**Current practice.** The lists are curated by hand for each target trait: a
search of the target's coloc partners for proxies, and for background
duplicates a concordance rule like the one above. Neither is automated yet.
Check the background similarity of every run (`summary$background` from
`summarise_ebmf_programs()`, shown as `background_mean_S` and
`background_edge_rate` in the univariate vignette). A high background is the
signal that duplicates are still in the matrix.

### 2. Orient SNP profiles relative to the target trait

Remove arbitrary allele coding so every column represents increasing target
liability. For SNP \(j\):

\[
X^*_{tj} = \mathrm{sign}(z_j)\, X_{tj}
\]

Optional \(\mathrm{asinh}(z)\) compression (`compress_method = "asinh"`, with
`compress_scale`; canonical `asinh` at scale 5) then limits the leverage of
extreme z-scores without a hard clip. `"none"` preserves effect scale but lets
outliers dominate direction and norm.

The oriented, compressed matrix is what EBMF factorises (stored as
`ebmf_input`). The returned `x_matrix` holds the raw, un-oriented z-scores;
use `ebmf_input` whenever profiles must be aligned to the risk allele.

### 3. Fit the EBMF latent-program model

Treat the oriented (and optionally compressed) traits x SNPs matrix as noisy
observations of latent programs:

\[
Y = LF^{\top} + E
\]

- \(Y\) = features x SNPs matrix (traits and molecular QTL studies as rows)
- \(L\) = feature loadings (which traits drive each program)
- \(F\) = SNP factors (which SNPs each program loads on)
- \(E\) = residual error

**Two loading matrices, two kinds of output each.** `flashier` returns
posterior summaries for both \(L\) and \(F\), and for each matrix it
separately reports a point estimate and an uncertainty measure. These are
easy to conflate — both are called "loadings" — but they answer different
questions about different axes of the model:

| | SNP-loading matrix `F` | trait-loading matrix `L` |
|---|---|---|
| point estimate | `F_pm` — posterior mean loading of each SNP on the program | `L_pm` — posterior mean loading of each background trait on the program |
| uncertainty | `F_lfsr` — local false sign rate: how confident EBMF is in that SNP's sign/nonzero-ness | `L_lfsr` — same, for the trait loadings |

`F_pm`/`F_lfsr` decide SNP membership (the gate just below), and squared
`F_pm` is the weight each SNP carries in the continuous enrichment of Step 5.
`L_pm` describes which background traits drive a program; it gives the
program's `effective_n_traits` (Step 4). Neither feeds the SNP-similarity
matrix `S`, which is built from the oriented SNP profiles themselves.

Magnitude and lfsr are not the same axis, even within one matrix: magnitude
(\(|F_{pm}|\)) is how big the point estimate is; lfsr is how much to trust it
— specifically, the posterior probability that the true sign is the opposite
of what was estimated. A locus can have a large magnitude and still a high
lfsr if it is weakly powered: the point estimate looks big, but the posterior
around it is wide enough that it could easily be near zero or flipped. That
is why membership gates on both, not magnitude alone.

Implemented via `flashier`: factors are added greedily while the variational
lower bound improves (up to `ebmf_greedy_Kmax`, canonical 100), then
cyclically backfit. The number of programs is learned, not fixed. The sparse
prior on loadings and factors is `ebmf_prior`:

- `"point_laplace"` (canonical) — heavier tails; often better when true
  effects are large relative to noise.
- `"point_normal"` — a sparse, spike-and-slab style prior.

**Extracting programs.** A SNP belongs to a program when its posterior factor
loading is both *significant* and *non-negligible*:

- `lfsr` < `ebmf_lfsr_threshold` (canonical 0.05)
- `|loading|` > `ebmf_magnitude_threshold` (absolute posterior mean of `F_pm`;
  canonical 0.5)

Because EBMF is soft, one SNP can pass the gate for several programs
(overlapping membership). Before extraction, at most one *global* mega-factor
— a factor that soft-assigns nearly every SNP or explains a large share of PVE
— is dropped (`ebmf_drop_global = TRUE` via `identify_ebmf_global_factors()`);
this guards against the "one giant cluster" failure mode. Nothing assumes
factor 1 is global.

Why EBMF:

- SNPs can belong to multiple programs; biology is overlapping, not mutually
  exclusive.
- \(K\) is inferred by the fit (ELBO) rather than by a graph-resolution or
  edge-threshold parameter.
- The input is the dense oriented z-score matrix itself, so no information is
  lost to a pairwise similarity threshold.
- Missing data are handled probabilistically instead of being zero-filled into
  similarity edges.

The cost is that factors are not guaranteed to be disjoint SNP modules and can
capture broad technical/global effects — which is exactly why the validation
layer in Step 4 exists.

### 4. Validating programs

`run_univariate_clustering()` returns the fit and soft memberships;
`summarise_ebmf_programs()` folds everything into a per-program table and gives
each program a **confidence tier**: high, medium, single_trait or low.
Programs with fewer than `min_module_size` member SNPs (canonical 3) are
dropped before any check. Programs of high or medium confidence are the
**valid** programs interpreted downstream.

#### The SNP x SNP similarity matrix

EBMF never sees pairwise SNP similarity — it factorises the traits x SNPs
matrix directly — but the same oriented (and optionally compressed) SNP
profiles are used to build the SNP x SNP cosine similarity matrix `S`, which
backs the similarity test below. Missing cells count as 0. The target row is
left out: after orientation it is positive at every SNP, so it would add the
same similarity to every pair. For SNPs \(j\) and \(k\):

\[
S_{jk} = \frac{{x^*_j}^{\top} x^*_k}{\|x^*_j\|\,\|x^*_k\|}
\]

- \(+1\) = concordant (parallel) profiles
- \(0\) = unrelated, or no trait observed at both SNPs
- \(-1\) = discordant (anti-parallel) profiles

#### What `similarity_q` measures

`similarity_q` asks: **are this program's SNPs more alike than the same number
of SNPs picked at random from this trait?**

1. **Observed.** `mean_internal_similarity` is the mean of `S` over every pair
   of the program's \(n\) member SNPs.
2. **Random sets.** `similarity_n_perm` (canonical 10,000) sets of \(n\) SNPs
   are drawn at random from the same trait (from the SNPs with a non-zero
   similarity to at least one other SNP), and the same mean is computed for
   each. Programs of the same size share one set of draws.
3. **Empirical p.** `similarity_emp_p` = (1 + number of random sets at least
   as similar as the program) / (1 + `similarity_n_perm`). With 10,000 sets the
   smallest possible p is about 0.0001.
4. **q.** `similarity_q` is the Benjamini-Hochberg adjustment of
   `similarity_emp_p` across all the trait's kept programs.

Read it as a significance test, not an effect size:

- **It favours larger programs.** The spread of the random-set means shrinks
  as \(n\) grows, so a 40-SNP program passes with a small margin over random,
  while a 3-SNP program needs a large one.
- **It is relative to the trait's own background.** The random sets come from
  the same matrix, so in a trait where every SNP already looks like every other
  (BMI with body-size traits left in), random sets are already similar and
  programs struggle to pass. The test is calibrated within each trait, which is
  what lets one threshold work across traits of different size, density and
  background.

#### Confidence tier

The tier is assigned in this order:

1. **low** — `similarity_q >= similarity_q` (canonical 0.05), or no estimate.
   The member SNPs are no more alike than random SNP sets of the same size.
   This is checked first, so a program that fails it is low whatever its
   number of traits, and can be set aside.
2. **single_trait** — passes the similarity test, but
   `effective_n_traits < min_effective_traits` (canonical 1.5). A program is
   defined as multi-trait: a factor carried by one trait is just that trait's
   hits, which noise produces too, and SNPs that share one trait look coherent
   to the similarity test whether or not they are real. It needs a closer look
   rather than being discarded or accepted, so it sorts above low.
3. **high** — `similarity_q < high_similarity_q` (canonical 0.01) and
   `connectedness >= high_connectedness` (canonical 0.4).
4. **medium** — passes the similarity test, but not high.

The quantities behind it:

- **Trait breadth** — `effective_n_traits` is \(1 / \sum_t p_t^2\), where
  \(p_t\) is trait \(t\)'s squared loading on the program (`L_pm`) as a share
  of the total (target row left out): 1 for a program on one trait, \(n\) for
  \(n\) equally loaded traits. Near-duplicate traits each count, which is one
  more reason to remove them (Step 1).
- **Connectedness** — the share of member pairs with
  \(S \ge\) `similarity_threshold` (canonical 0.2; positive similarity only).
  It asks whether the similarity is spread across the program or carried by a
  few very similar pairs.

Under a true null, almost every EBMF program rests on one trait (98.5% of null
programs in a quick run of the
[null simulation study](investigation-univariate-null-simulations.html)), but
most fail the similarity test and are therefore `low`. The `single_trait` share
under the null is how often noise lands in the tier that needs a closer look.

**Reported alongside** (not used for the tier):

- **excess_similarity / excess_connectedness** — `mean_internal_similarity` and
  `connectedness` minus their means in the random sets. Because the random sets
  are drawn from the same pool whatever the program's size, their expected mean
  is the trait's background similarity, so within one run these are the
  observed values minus a constant. In the Suzuki T2D run the difference was
  0.030 for every one of 44 programs (SD 0.0002), equal to the background mean
  `S`, and the two columns correlate perfectly with the unadjusted ones. They add
  nothing within a run; across runs, the background itself
  (`summary$background`) says the same thing more directly.
- **similarity_z** — the program's mean similarity in standard deviations of
  the random-set means; used to order programs within a tier.
- **factor_pve** — the share of the matrix variance the factor explains.
- **replication** — the mean Jaccard overlap between the program's top
  `top_n` SNPs and the best-matching factor when `1 - frac_traits` of the
  trait rows are held out and flashier is refit (`n_rep` replicates).
- **parent_program** — when at least 90% of a program's SNPs sit within
  500 kb of the SNPs of a larger program, that program is its parent. Nested
  sub-programs are kept; `order_programs_by_confidence()` lists them under
  their parent.
- **posterior_evidence** — aggregate (`median` or `mean`) of `-log10(lfsr)`
  across the program's assigned SNPs (capped so lFSR underflow does not blow
  up).
- **max_abs_factor_corr / redundant** — the maximum absolute correlation of the
  program's SNP loading vector with any other program; programs exceeding
  `redundancy_threshold` are flagged `redundant` (near-duplicate factors).

`kable_confidence()` colours program tables by tier (green high, orange
medium, grey single_trait, red low).

**Permutation calibration (reporting layer).** `calibrate_ebmf_programs()`
additionally shuffles each trait row independently (`n_null` replicates,
preserving each trait's sparsity and effect distribution), refits the pipeline,
and pools the null loading distribution. Each SNP x program loading gets an
empirical p-value and BH q-value against this null; cells with `q <=
alpha_membership` are labelled `core`, and the top remaining loadings per
program form an explicitly uncertified `candidate` tier. This is a membership
labelling layer, not a program-level validity gate.

### 5. Interpret valid programs

Valid (high or medium) programs are interpreted with two kinds of annotation:
the background traits that set the program apart, and a **continuous**
loading-based enrichment of trait categories, tissues and pathways. Both are
shown together in one table per run in the univariate and Suzuki vignettes.

**Module-specific traits** (`summarise_module_specific_traits()`). For each
background trait, `specificity = mean |z| in the program / (mean |z| in the
rest of the trait's SNPs + 0.1)`. Traits are listed if their specificity is at
least `min_specificity` (1.25 in the vignettes). This picks out traits that
are elevated in the program in particular, rather than traits that are large
at every SNP, which ranking by `|z|` alone would over-emphasise.

#### Continuous loading-based enrichment

`enrich_program_loadings_trait_categories()`, `enrich_program_loadings_tissues()`
and `enrich_program_loadings_pathways()` ask whether a program's SNP loadings
are larger at SNPs linked to an annotation. There is no membership gate: every
SNP with a finite EBMF loading contributes to every program's test. For
program \(k\) and annotation term \(a\) (a trait category, a tissue, or a KEGG
or Reactome pathway):

- \(y_j = F_{jk}^2\), SNP \(j\)'s squared loading on the program;
- \(x_j = 1\) if SNP \(j\) is linked to the term, else 0. A SNP is linked to a
  category or tissue through its coloc rows, and to a pathway through its
  colocalising QTL genes.

An ordinary least squares regression

\[
y = \beta_0 + \beta_1 x + \varepsilon
\]

(`lm(loading^2 ~ membership)`) is fitted for each program and term. The fitted
\(\beta_1\), the difference in mean squared loading between linked and
unlinked SNPs, is reported as `enrichment`; its `p` comes from the usual OLS
\(t\)-test.

Why this replaced the hard-cutoff approach (gate SNPs into a program, then
test the gated set with a hypergeometric test):

- **No threshold on membership.** The hard-cutoff test depends on the
  lFSR/magnitude gate: a SNP just below it counts for nothing, one just above
  counts fully, and a program of 5 gated SNPs has little power whatever its
  loadings say. The continuous test uses the whole loading vector, weighting
  each SNP by how strongly it loads.
- **One correction for the whole trait.** See below.

Details:

- **Soft membership gate** (`min_loading_magnitude`, default 0.02). A linked SNP
  counts as \(x = 1\) for a program only if its absolute loading on that program
  exceeds this value; otherwise it is treated as \(x = 0\) for that program's
  test. This keeps near-zero, shrinkage-only loadings from diluting the
  contrast. SNPs are never dropped, only relabelled.
- **Minimum term size** (`min_category_size`, default 5). Terms linked to fewer
  SNPs are not tested.
- **Effect size.** `enrichment` is in squared-loading units, so it is not
  comparable between programs. `r_squared`, the share of the program's
  squared-loading variance the term explains, is bounded in \([0, 1]\) and
  comparable. The vignettes require **FDR ≤ 0.05 and R² ≥ 0.05** (between
  Cohen's "small", 0.01, and "medium", 0.09), and order hits by `enrichment`.
  FDR alone is not enough: with hundreds of SNPs in every regression, a
  negligible difference can still be significant. Because the number of SNPs
  is about the same for every program of a trait, `r_squared` tracks `p`
  closely; it is a bounded, readable rescaling of the same evidence, not an
  independent second filter.
- **Multiple testing.** Each of the three functions Benjamini-Hochberg-corrects
  across *every* program's tests for that annotation type at once, so `fdr`
  reflects the number of regressions actually run for the trait. Every
  fitted EBMF factor is tested, including those below `min_module_size` and
  those of low confidence, and the vignettes then report the high and medium
  ones. KEGG and Reactome are pooled into one pathway family; categories,
  tissues and pathways are three separate families.
- **SNPs are treated as independent observations.** Nearby SNPs in LD are not
  independent, so the p-values are optimistic. A formal analysis would
  aggregate SNPs to genes, or use cluster-robust standard errors or a
  permutation within LD blocks.

The hard-cutoff functions (`enrich_snp_group_trait_categories()`,
`enrich_snp_group_tissues()`, `enrich_snp_group_pathways()`) remain in the
package; the trait-category and tissue versions are still used by
[Multi-trait programs](investigation-multitrait-programs.html).

#### Comparing runs and benchmarks

- **Runs with different trait sets.** EBMF is refitted for each trait set, so
  program numbers do not carry over between runs. The univariate vignette
  matches programs across runs by the Jaccard overlap of their member SNPs,
  then compares which annotations are significant in both runs or in one only.
- **Unique vs shared SNPs.** A SNP in exactly one valid program should be more
  specific (fewer distinct trait categories and genes) than a SNP shared by
  several. The univariate vignette tests this with
  `get_all_variant_pleiotropies()`. Gene pleiotropy has a long tail (a median
  of 1 gene per SNP across GPMap, but a maximum in the thousands), so it is
  plotted on a pseudo-log axis and read through medians and a rank test rather
  than means.
- **External benchmark.** [Suzuki T2D](investigation-suzuki-t2d.html) matches
  each Suzuki et al. (2024) index SNV to the GPMap SNP in its LD block and
  compares program and cluster trait profiles, to check whether the programs
  recover a published clustering.

### 6. Linking programs across traits

Each target trait is decomposed on its own, so a program found for one trait
and a program found for another are separate objects with arbitrary numbers.
The multi-trait analysis
([Multi-trait programs](investigation-multitrait-programs.html), methods in
[Multi-trait correspondence](investigation-multitrait-correspondence.html))
asks of every pair of programs from two traits whether they describe the
same biology, using both matrices of Step 3.

**Same biology? — profile congruence.** A program's trait profile (`L_pm`)
says which background traits drive it and how strongly. The two programs'
profiles are compared with Tucker's congruence coefficient, an uncentred
correlation:

\[
\phi = \frac{\sum_t \tilde l_a(t)\, \tilde l_b(t)}
{\sqrt{\sum_t \tilde l_a(t)^2 \sum_t \tilde l_b(t)^2}},
\qquad \tilde l = l \cdot (1 - 2\,\text{lfsr})_+
\]

- **Which traits are compared.** Each trait's matrix has its own rows, so the
  sum runs over the background traits **both** fits carry, leaving out the
  two anchor traits (each is positive at every SNP in its own fit but an
  ordinary row in the other) and any trait removed from either fit as a proxy
  (Step 1). On real data this left 550–660 traits per pair of traits.
- **No magnitude cutoff.** Every trait loading counts, in proportion to its
  size and its confidence: the weight \((1 - 2\,\text{lfsr})_+\) is 0 for a
  loading whose sign is a coin flip and 1 for a certain one.
- **The sign is a direction.** Each trait's SNPs are oriented to its own risk
  allele (Step 2), so a module that raises one trait and lowers the other
  appears in the second trait's fit with every background trait mirrored:
  \(\phi < 0\) is an antagonistic program, \(\phi > 0\) a concordant one.
- **How many traits carry it** (`n_eff_traits`, the participation ratio of
  the per-trait contributions) is reported: a congruence carried by one or two
  traits is a narrower claim than one carried by dozens.

**Same SNPs? — locus alignment.** Where the two traits were fine-mapped to the
same loci (`coloc_group_id`; an uploaded GWAS is bridged by its own coloc
record, variant and LD), the programs' SNP loadings (`F_pm`) are compared
directly: `alignment` is the R² between the two loading vectors over the
shared loci, with the same confidence weights and no magnitude cutoff. Loci
only one trait carries cannot agree or disagree, so they are left out, and
`comparability` reports how much of each program they leave out. Two traits
typically share only 5–30% of their loci, so the locus check confirms a
profile match where it can rather than deciding it. Its result
(`locus_status`) is `too_few_loci` when the pair shares fewer than 3
high-confidence loci or fewer than 5 loci effectively carry it; otherwise
`agree` when `alignment >= 0.5` and a permutation test of the
loading-weighted concordance passes (BH across all pairs), and `disagree`
when either fails.

**Tiers.** The two are combined: a strong profile (\(\lvert\phi\rvert \ge
0.85\)) is **high** when the loci agree, **medium** when there are too few
loci to check, and **low** when the loci disagree; a moderate profile
(\(0.70 \le \lvert\phi\rvert < 0.85\)) is medium only when the loci
agree. The full table, and what each measure checks, is in
[Multi-trait EBMF program correspondence](multitrait-overview.html).

High and medium pairs are **links**; links are grouped into multi-trait
**families** by community detection, and each family's SNPs, pooled per
trait, are annotated with trait categories, tissues and pathways as in
Step 5, so that agreement across traits is independent support. The cutoffs
follow Tucker's conventions for factor similarity (0.85 "fair similarity",
0.95 "equal"), which were set for psychometric factors, not genetic loadings,
so they are parameters to tune against the observed distributions. A link's
direction is the sign of \(\phi\); where enough loci are shared,
`module_rg()`, the correlation of the two traits' own allele effects across
the program's loci, checks it (`direction_check`) without changing the tier.

## Variables to consider

Parameters fall into three groups: the **EBMF clustering parameters** that
control matrix construction and membership, the **program validation
parameters** that set the confidence tier, and the **continuous enrichment
parameters**. "Canonical" values are those in `vignettes/_ebmf_settings.yml`.

### EBMF clustering parameters

| Parameter | Canonical | Values | Explanation |
|---|---|---|---|
| Duplicate-trait removal | by hand, per trait | proxies only; proxies + near-duplicates | The most consequential choice (Step 1). Leaving duplicates in inflates background similarity, which hides programs and makes the similarity test hard to pass; removing them also removes observations. |
| `include_trans` | `TRUE` | `TRUE`, `FALSE` | Including trans effects can reveal distributed regulatory programs and improve tissue/pathway interpretation, while increasing density and possible noise. Excluding them gives a more local cis-focused analysis but may miss system-level mechanisms. |
| `trait_subset` | `phenotypic` | `all`, `phenotypic` | `phenotypic` drops sparse molecular / gene-annotated rows (keeping the target trait), recovering structure that sparse molecular rows dilute but losing QTL-level signal. `all` keeps every available background trait at the cost of extra sparsity. |
| `min_snp_signals` | `2` | `1`, `2`, `3`, `5` | Raising it removes singleton or very sparse rows, reducing unstable factors and computation. It can also remove rare, module-specific biology. Lowering it preserves those signals but makes factorization more sensitive to sparse observations. |
| `compress_method` / `compress_scale` | `asinh` / `5` | `asinh`, `none` | `asinh` is nearly linear around zero but compresses extreme values, so factors reflect patterns across several traits rather than one very large z-score. `none` preserves effect scale but allows outliers to dominate direction and norm. |
| `ebmf_lfsr_threshold` | `0.05` | `0.02`, `0.05`, `0.1` | Lowering it keeps only SNPs whose program loading has very strong posterior sign evidence, shrinking and sharpening programs. Raising it adds borderline memberships. |
| `ebmf_magnitude_threshold` | `0.5` | `0.25`, `0.5`, `0.75`, `NA` | Raising it keeps only SNPs with substantial program loadings. `NA` gates membership on lFSR alone. |
| `ebmf_prior` | `point_laplace` | `point_normal`, `point_laplace` | `point_laplace` has heavier tails and is often better when true effects are large relative to noise; `point_normal` is a spike-and-slab style prior. |

### Program validation parameters

| Parameter | Canonical | Meaning | Explanation |
|---|---|---|---|
| `min_module_size` | `3` | Minimum member SNPs | Programs below it are dropped before any check: a 2-SNP program cannot be annotated. |
| `similarity_n_perm` | `10000` | Random same-size SNP sets for the similarity test | Sets the smallest possible p (about 1 / `similarity_n_perm`), so it must be large enough for `high_similarity_q` to be reachable. |
| `similarity_q` | `0.05` | BH q below which a program passes the similarity test | Failing it makes a program `low`, whatever its traits. |
| `min_effective_traits` | `1.5` | Effective number of traits below which a program that passes the similarity test is `single_trait` | A program resting on one trait's hits needs a closer look before being treated as a program. |
| `high_similarity_q` | `0.01` | BH q threshold for high confidence | Together with `high_connectedness`. |
| `high_connectedness` | `0.4` | Minimum connectedness for high confidence | Share of member pairs with `S` at or above `similarity_threshold`. |
| `similarity_threshold` | `0.2` | Edge cutoff for connectedness | Connectedness counts member pairs with `S >= similarity_threshold` (positive similarity only). Also used for the background edge rate. |
| `redundancy_threshold` | `0.9` | Factor-loading correlation above which a program is flagged `redundant` | A flag for near-duplicate factors; redundant programs are not removed. |

The confidence tier therefore rests on `min_module_size`, `similarity_q`,
`min_effective_traits`, `high_similarity_q` and `high_connectedness`, on top of
the membership gates `ebmf_lfsr_threshold` and `ebmf_magnitude_threshold`.
`excess_*`, `similarity_z`, `factor_pve`, `replication`, `parent_program`,
`posterior_evidence` and `max_abs_factor_corr`/`redundant` are reported
descriptively.

### Continuous enrichment parameters

| Parameter | Values | Meaning | Explanation |
|---|---|---|---|
| `min_category_size` | `5` (default) | Minimum linked SNPs for a term to be tested | Below this, a term is skipped rather than tested on too few SNPs to fit a stable regression. |
| `min_loading_magnitude` | `0.02` (default), `0` | Soft membership gate on `x = 1` | `0` counts every linked SNP as a member. Raising it relabels linked SNPs with negligible loadings on this program to `x = 0` for that program's test only. SNPs are never dropped. |
| `r_squared` floor (applied in the vignettes) | `0.01` (Cohen small), `0.05`, `0.09` (Cohen medium) | Minimum variance explained for a hit, on top of `FDR <= 0.05` | `enrichment` is not comparable across programs, so the fixed cutoff is applied to `r_squared`. The vignettes use `0.05`. |

## Verification Strategy

Two complementary checks verify that the pipeline produces reliable, meaningful
programs: **simulation**, where the ground truth is planted and exactly known,
and an **external benchmark**, where real GPMap data is compared against an
independent, published clustering whose "ground truth" is itself one
algorithm's answer on one trait panel, not certain. Simulation controls what
real data cannot — exactly which SNPs belong to which module; the external
benchmark supplies what simulation cannot — real trait duplication, real
sample-size imbalance between studies, real LD, and a result from independent
data and an independent method to agree or disagree with. A pipeline that only
passes simulations might still fail on real data's mess; a pipeline that only
matches one published partition might be overfit to it. The two checks are
read together, not as a single pass/fail.

### Simulation

The simulation plan — null simulations (hallucination), module simulations
(recovery) and multi-trait simulations (correspondence) — is described in
[Simulation overview](simulations-overview.html).

### External benchmark: Suzuki et al. (2024) T2D clusters

[Suzuki T2D](investigation-suzuki-t2d.html) runs the full pipeline above on a
type 2 diabetes GWAS and compares the resulting programs with a published,
independently derived partition of T2D loci: Suzuki *et al.* (2024) clustered
1,289 T2D index SNVs into 8 mechanistic groups (Obesity, Lipodystrophy,
Metabolic syndrome, two beta-cell clusters, Residual glycaemic, Body fat,
Liver/lipid metabolism) by k-means over each SNV's standardised associations
with 37 cardiometabolic traits. Nothing about this partition comes from
gpmapr: different trait panel (their 37 traits vs. whatever colocalises with
the T2D upload in GPMap), different population (their multi-ancestry
meta-regression vs. GPMap's EUR fine-mapping), different algorithm (hard
k-means vs. soft EBMF), different unit (a single index SNV vs. a GPMap SNP
that may represent a different, correlated variant at the same locus). Getting
the comparison right is as much about **bridging two separate ID spaces**
as it is about the statistics on either side of the bridge.

**Why this can't be one clustering-comparison statistic.** The module
simulations ([Simulation overview](simulations-overview.html)) score recovery with the adjusted Rand index because the
planted partition and the recovered programs live over the *same* set of
SNPs, so every item can be looked up in both at once. Suzuki's SNVs and
GPMap's SNPs are not the same set: they come from independent fine-mapping of
different data, so the two partitions share no common index to compare
directly. The comparison therefore has to go through several independent
lines of evidence, each of which turns one half of the mismatch into
something comparable, rather than through a single number.

**Bridging SNPs to SNVs.** Suzuki's index SNV and GPMap's lead variant at the
same locus are usually different variants, but GPMap already treats
colocalisation within an **LD block** as "the same signal", so that is the
unit the match is made in: each Suzuki SNV is placed in the LD block of its
GRCh38 position, and matched to the nearest pleiotropy-matrix SNP in that
block. A SNV with no upload signal in its block, or one that colocalises with
too few traits to enter the matrix at all, cannot be matched and says so
explicitly rather than being silently dropped. Several SNVs, sometimes from
different clusters, can land on the same matrix SNP (one locus held more than
one Suzuki signal); each matched SNP then contributes a weight of 1, split
evenly across the SNVs it matches, so one promiscuous SNP cannot inflate every
cluster it touches.

**Is a program's cluster mix more than chance? (the SNV enrichment test.)**
Once a program's member SNPs are matched to Suzuki SNVs, the question is
whether their cluster composition differs from what matched SNPs look like
overall — the same *is this more than a random draw* logic as
`similarity_q` (Step 4), applied to cluster labels instead of SNP similarity:

1. **Observed.** For program \(k\) and cluster \(c\), the weighted count of
   \(k\)'s matched member SNPs labelled \(c\).
2. **Background.** The same count's share over *every* matched SNP in the
   matrix — what a program with no preference for any cluster would show.
3. **Enrichment.** The program's share of cluster \(c\) divided by the
   background share: a fold change, the effect size. It says *which* cluster
   and *how much*, but not how sure.
4. **Permutation p.** 2,000 random sets of matched SNPs, of the program's
   size, drawn from the same matched pool; `p` is the share of random sets
   whose count for cluster \(c\) is at least the program's. This asks *how
   sure*, holding the program's size and the pool fixed, exactly as the
   similarity-test null does.
5. **q.** Benjamini-Hochberg over every program × cluster cell, because a
   program with 8 clusters and dozens of programs tested means hundreds of
   tests, most of which should show nothing.

Enrichment and q are the same two-axis split used throughout this document
(`similarity_lift`/`similarity_z` vs `similarity_q`; `enrichment` vs `fdr` in
continuous enrichment): a large fold change with a weak q is not evidence, and
a significant q on a tiny fold change is not a strong claim either. A
`top_cluster` is only named when the program has at least 2 matched SNPs — below
that the permutation test has essentially no power to say anything.

**Does a program's trait profile look like a cluster's? (profile cosine.)**
Suzuki also publish each cluster's own mean profile over the 37 clustering
traits (`beta / se`, their Supplementary Table 7). A program's profile is the
mean of its member SNPs' rows in the same (oriented, asinh-compressed) matrix
EBMF factorises, restricted to the Suzuki traits GPMap actually has (4a). The
two profiles are **not on the same scale** — compressed colocalisation
z-scores against the z of a published cluster mean — so only their *pattern*
across traits is compared, not the size of either: the cosine between them,
1 for the same pattern of signs and relative sizes, 0 for unrelated, -1 for
opposite. This is the same reasoning as the SNP-similarity matrix `S` in Step
4 — cosine is invariant to overall scale, so it measures shape, not
magnitude, which is exactly what's comparable between two quantities on
different units.

A resolution ceiling has to be checked before reading any program's cosine:
the cosine **between Suzuki's own cluster profiles**, over the same 37 traits.
Three pairs come out close (Beta cell +PI vs −PI: 0.92; Metabolic syndrome vs
Lipodystrophy: 0.81; Body fat vs Obesity: 0.73) — these clusters already look
almost identical on the traits GPMap has, so no program, however good, can be
expected to separate them on this axis. A program's low cosine with one of a
close pair and high cosine with the other would be more surprising than
useful; the honest reading credits the program with the pair, not a false
precision.

**Combining the evidence (4f).** The SNV enrichment test and the profile
cosine measure different things — discrete locus overlap against continuous
trait pattern — and can disagree: a program with a clean profile match can
have too few matched SNVs for the enrichment test to say anything, and vice
versa. Gene-level anchors (4b: Suzuki's own worked-example loci, checked
positionally and through colocalising QTLs) and the full SNV-to-program
table (4c) sit alongside both as further, qualitative evidence. Because no
single statistic resolves every case, the final call is a hand-written,
three-level judgement — **clear** (SNV enrichment and profile agree on the
cluster), **partial** (only one agrees, or the program sits between two
indistinguishable clusters), **none** — with the reasoning for each program
written down next to it, rather than either a black-box composite score or an
unaudited claim. A machine check re-derives each call's two quantitative
conditions against the current run and flags any program whose numbers no
longer support what the text says, so the hand-written reading cannot drift
silently from a re-render's results.

## Structure of Results

The aim is not to produce a definitive biological truth for every SNP. Instead,
the method produces a structured, evidence-based decomposition of a trait's
genetic architecture, with results that can be assessed at several levels.

For practical use, the framework should provide results at the overall, program, and SNP level.

| Level       | Primary question                                                                         | Main output                                                             |
| ----------- | ---------------------------------------------------------------------------------------- | ----------------------------------------------------------------------- |
| **Trait**   | How much of this trait's genetic architecture can be decomposed into distinct structure? | Number of programs, SNP coverage, overlap, overall signal captured      |
| **Program** | Is this structure reliable, and what biology does it represent?                          | Reliability metrics plus trait, tissue, gene and pathway interpretation |
| **SNP**     | Which program or programs does this variant contribute to?                               | Soft memberships, posterior evidence, rank and molecular annotations    |

This standard report would be returned by gpmapr and as a framework for the Results section of the paper. It should keep the distinction clear between statistical evidence, which establishes whether a program is reliable, and biological annotation, which provides its interpretation.

#### Overall trait-level metrics
- Number of SNPs analysed, and the traits removed as proxies or near-duplicates
- Background similarity of the matrix (`background_mean_S`, `background_edge_rate`)
- Number of programs in each confidence tier (high, medium, single_trait, low)
- Proportion of SNPs assigned to at least one valid program
- Proportion of SNPs unassigned
- Degree of overlapping membership, for example the proportion of assigned SNPs belonging to more than one program
- Overall matrix signal explained by validated programs, if this can be meaningfully derived from the EBMF fit
- Number/proportion of programs with some evidence of biological interpretation

#### Program Level metrics
- Number of member SNPs (maybe also report number of SNPs that were filtered out by the lFSR/magnitude gate)
- Membership overlap with other programs
- Top SNPs by posterior loading or membership evidence
- Statistical reliability: confidence tier, `similarity_q`, `mean_internal_similarity`, connectedness, `effective_n_traits`, factor variance explained, replication, parent program
- Biological interpretation: module-specific traits, and continuous loading-based enrichment of trait categories, tissues and pathways (effect, r_squared, FDR)

#### SNP Level metrics
- Program membership(s)
- Posterior loading in each program
- Membership lFSR
- Magnitude of membership/loading
- Core versus candidate membership status
- Rank within each program, if useful
- Whether the SNP belongs to multiple programs
- Similarity to other SNPs within its assigned program
- Variant-level pleiotropic profile, for example breadth of associated trait categories
- Molecular annotations, including associated genes and tissues where available
