---
title: "Suzuki T2D"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: investigation-wide.html
vignette: >
  %\VignetteIndexEntry{Suzuki T2D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  api: "production"
  target_guid: "c6d2e20f-6ece-d093-78ee-37534547c466"
  p_threshold: null
  include_situated_gene: false
  min_mean_internal: 0.2
  min_connectedness: 0.5
  ebmf_magnitude_threshold: 0.75
  ebmf_lfsr_threshold: 0.02
  ebmf_prior: "point_laplace"
  min_module_size: 3
  n_null: 10
  alpha_membership: 0.05
  n_candidate_tier: 25
  n_rep: 3
  frac_traits: 0.8
  stability_top_n: 20
  posterior_stat: "median"
  posterior_evidence_cap: 50
  redundancy_threshold: 0.9
  cores: 5
  associations: "coloc"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  out.width = "100%",
  fig.align = "center"
)
```

```{r libraries}
pkg_root <- if (file.exists("../DESCRIPTION")) ".." else "."
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(gpmapr)
}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r suzuki-data, include=FALSE}
suzuki_dir <- system.file("extdata", "suzuki2024", package = "gpmapr")
suzuki_clusters <- utils::read.csv(
  file.path(suzuki_dir, "clusters.csv"),
  stringsAsFactors = FALSE
)
suzuki_profiles <- utils::read.csv(
  file.path(suzuki_dir, "cluster_trait_profiles.csv"),
  stringsAsFactors = FALSE
)
suzuki_trait_map <- utils::read.csv(
  file.path(suzuki_dir, "trait_map.csv"),
  stringsAsFactors = FALSE
)
suzuki_snvs <- utils::read.csv(
  file.path(suzuki_dir, "index_snv_clusters.csv"),
  stringsAsFactors = FALSE
)
cluster_levels <- suzuki_clusters$cluster
```

This vignette asks whether gpmapr's univariate SNP programs recover the
mechanistic clusters of type 2 diabetes (T2D) signals that Suzuki et al. (2024)
defined. The analysis runs on the European-ancestry meta-analysis from that
paper, uploaded to GPMap. The questions are:

- Which programs pass validation? Only high and medium confidence programs are
  compared with Suzuki's clusters.
- Do a program's member SNPs match the index SNVs of a Suzuki cluster?
- Does its profile over Suzuki's cardiometabolic traits match the cluster's?
- Does it carry that cluster's canonical genes?

The program summary cards in Part 4 bring these together, and Part 4 ends
with a hand-written match of each program to Suzuki's clusters.

## Part 1: Suzuki et al. (2024)

> Suzuki K, Hatzikotoulas K, Southam L, *et al.* Genetic drivers of heterogeneity
> in type 2 diabetes pathophysiology. *Nature* 627, 347–357 (2024).
> <https://doi.org/10.1038/s41586-024-07019-6>

The reference tables used below are CC BY 4.0 extracts of the paper's
supplementary material. They are in `inst/extdata/suzuki2024/` and are
regenerated by `data-raw/suzuki2024.R`.

### Study

- **Scale.** The Type 2 Diabetes Global Genomics Initiative meta-analysed 428,452
  T2D cases and 2,107,149 controls across six ancestry groups.
- **Discovery.** Signals come from multi-ancestry meta-regression (MR-MEGA) with
  three axes of genetic variation as covariates.
  - Index SNVs were selected at *P* < 5 × 10⁻⁸ by greedy clumping. A variant
    joined a clump if r² > 0.05 with the index SNV in *any* of the five 1000
    Genomes super-populations, within 5 Mb.
  - This gave **1,289 independent signals at 611 loci**, of which 145 loci are
    new.
- **EUR meta-analysis (the dataset used here).**
  - 36 GWAS: 242,283 cases and 1,569,734 controls (effective N ≈ 751,755,
    60.3% of the total).
  - Fixed-effects inverse-variance meta-analysis in METAL, GRCh37.
  - The largest contributors are the Million Veteran Program (90,107 cases),
    FinnGen (37,002), UK Biobank (19,119), the Estonian Biobank (12,208) and
    deCODE (11,448). 23andMe is not included.
- **Signal sets differ.** The 1,289 signals were discovered in the
  multi-ancestry analysis, so some are driven by non-European data and may be
  weak or absent in the EUR upload.

### How the clusters were built

1. **Inputs.** Each index SNV's associations with 37 cardiometabolic traits were
   taken from the largest available GWAS (Supplementary Table 5) and aligned to
   the T2D risk allele.
2. **Scaling.** Each association was scaled as $Z_{ij} = \beta_{ij} /
   (\sqrt{N_i}\, s_{ij})$, which is a standardised per-allele effect that does
   not depend on the trait GWAS's size.
3. **Clustering.** The SNVs were grouped by k-means with iterative imputation of
   missing associations (ClustImpute). The number of clusters came from the
   majority rule over 27 NbClust indices.
4. **Hard assignment.** Every SNV is assigned to exactly one cluster. This
   differs from earlier soft clustering by Bayesian non-negative matrix
   factorisation (Udler *et al.* 2018), which gives each SNV cluster weights and
   leaves some unassigned.

```{r suzuki-clusters}
knitr::kable(
  suzuki_clusters |>
    dplyr::transmute(
      cluster,
      n_signals,
      new_in_suzuki = dplyr::if_else(previously_reported, "no", "yes"),
      cardiometabolic_profile,
      example_loci,
      insulin_secretion,
      insulin_sensitivity,
      t2d_or = round(exp(t2d_log_or), 3),
      open_chromatin_enrichment
    ),
  caption = paste0(
    "The eight mechanistic clusters (main-text Table 1, Supplementary Table 9, ",
    "Fig. 2). * = trait adjusted for BMI. t2d_or is the inverse-variance ",
    "weighted mean T2D odds ratio of the cluster's index SNVs. Insulin ",
    "secretion/sensitivity give the direction of the T2D risk allele's effect."
  ),
  row.names = FALSE
)
```

```{r suzuki-profile-heatmap, fig.width = 9, fig.height = 9}
trait_order <- suzuki_trait_map$trait_acronym
suzuki_profiles |>
  dplyr::mutate(
    z = pmax(pmin(beta / se, 10), -10),
    cluster = factor(cluster, levels = cluster_levels),
    trait_acronym = factor(trait_acronym, levels = rev(trait_order))
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = cluster, y = trait_acronym, fill = z)) +
  ggplot2::geom_tile(colour = "white") +
  ggplot2::scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    limits = c(-10, 10), name = "z (capped)"
  ) +
  ggplot2::labs(
    title = "Suzuki et al. cluster profiles over the 37 clustering traits",
    subtitle = "Supplementary Table 7; beta/SE of the cluster mean, aligned to the T2D risk allele",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1))
```

### How the paper validated the clusters

- **Physiology.** The clusters form a cline in insulin-related measures
  (HOMA-B, HOMA-IR, clamp-derived sensitivity). It runs from low insulin
  secretion in the two beta-cell clusters to the most extreme insulin
  resistance in the Lipodystrophy cluster.
- **Single-cell open chromatin.** The clusters are enriched for different cell
  types:
  - beta-cell and Residual glycaemic clusters: pancreatic islets;
  - Beta cell −PI: also enterochromaffin cells;
  - Lipodystrophy: adipocytes only;
  - Metabolic syndrome: pericytes and endothelial cells;
  - Obesity: islet non-beta cells, adrenal and brain neurons;
  - Body fat: no enrichment.
- **Partitioned polygenic scores.** After adjusting for the overall T2D score:
  - the Obesity score is associated with higher risk of coronary artery
    disease, peripheral artery disease, end-stage diabetic nephropathy and
    earlier T2D onset;
  - the Beta cell +PI score is associated with *lower* coronary artery disease
    and nephropathy risk.

### Caveats for using the clusters as a benchmark

- **They are one partition, not ground truth.** They come from one algorithm on
  one trait panel.
- **Two clusters act as catch-alls.** Residual glycaemic (389 signals) and Body
  fat (273) are the largest. Body fat has no chromatin enrichment, and the three
  new clusters have weaker T2D effects than the established ones.
- **Liver/lipid metabolism has only 3 signals.** Its chromatin model was
  unstable in the paper, and it is not a realistic target for recovery.
- **Hard assignment forces membership.** SNVs with intermediate profiles still
  get a cluster, whereas gpmapr's EBMF programs are soft and overlapping.

## Part 2: The uploaded GWAS

```{r upload}
select_api(params$api)
target_guid <- params$target_guid
p_threshold <- params$p_threshold
target <- trait(target_guid, include_associations = TRUE)
target_name <- target$trait$name

upload_metadata <- tryCatch(
  jsonlite::fromJSON(target$trait$upload_metadata),
  error = function(e) return(list())
)
own_rows <- target$coloc_groups |>
  dplyr::filter(trait_id == target_guid)
n_signals_at <- function(p) {
  return(dplyr::n_distinct(own_rows$variant_id[own_rows$min_p <= p]))
}

knitr::kable(
  data.frame(
    name = target_name,
    guid = target_guid,
    sample_size = target$trait$sample_size,
    ancestry = target$trait$ancestry,
    reference_build = if (is.null(upload_metadata$reference_build)) {
      NA_character_
    } else {
      upload_metadata$reference_build
    },
    upload_p_threshold = if (is.null(upload_metadata$p_value_threshold)) {
      NA_real_
    } else {
      upload_metadata$p_value_threshold
    },
    status = target$trait$status,
    n_coloc_groups = dplyr::n_distinct(target$coloc_groups$coloc_group_id),
    n_signals_p5e6 = n_signals_at(5e-6),
    n_signals_p5e8 = n_signals_at(5e-8),
    stringsAsFactors = FALSE
  ),
  caption = paste0(
    "Upload metadata. n_signals counts the upload's own fine-mapped signals ",
    "(distinct lead variants) at each threshold. Suzuki et al. used 5e-8."
  ),
  row.names = FALSE
)
```

The analysis below uses signals at
`r if (is.null(p_threshold)) "the upload's own threshold" else format(p_threshold)`.
Re-render with `params = list(p_threshold = 5e-8)` to match the
genome-wide threshold that Suzuki et al. used.

## Part 3: Univariate analysis

The steps are the same as in
[the univariate analysis investigation](investigation-univariate-analysis.html),
except that T2D proxy traits are removed first (see below):

1. Build a pleiotropy matrix of z-scores (`beta / se`) for every background
   trait at each upload SNP, keeping phenotypic traits only and leaving out
   the T2D proxy traits.
2. Drop sparse traits (observed at fewer than `min_snp_signals` SNPs) and
   ubiquitous traits (observed at more than 80% of SNPs).
3. Orient every SNP to the T2D risk allele and compress with asinh.
4. Factorise with EBMF into overlapping programs.
5. Validate the programs.

Shared settings come from `vignettes/_ebmf_settings.yml`. After validation,
each program is annotated as in that vignette, in one table:

- module-specific traits;
- loading-based (continuous) enrichment of trait categories, tissues and
  pathways;
- the genes at the program's SNPs, which this vignette adds (shown in the
  program summary cards).

Only the pleiotropy comparison of unique vs shared SNPs is left to the
univariate vignette.

```{r params}
min_group_size <- 5
max_snp_fraction <- 0.8
top_n_traits <- 10L
min_specificity <- 1.25

ebmf_settings <- resolve_ebmf_settings(params)

trait_subset <- ebmf_settings$trait_subset
min_module_size <- as.integer(ebmf_settings$min_module_size)
min_snp_signals <- as.integer(ebmf_settings$min_snp_signals)
compress_method <- ebmf_settings$compress_method
compress_scale <- as.numeric(ebmf_settings$compress_scale)
include_trans <- isTRUE(ebmf_settings$include_trans)
similarity_threshold <- as.numeric(ebmf_settings$similarity_threshold)
ebmf_lfsr_threshold <- as.numeric(ebmf_settings$ebmf_lfsr_threshold)
ebmf_prior <- ebmf_settings$ebmf_prior
ebmf_backfit <- isTRUE(ebmf_settings$ebmf_backfit)
ebmf_magnitude_threshold <- as.numeric(ebmf_settings$ebmf_magnitude_threshold)
similarity_n_perm <- as.integer(ebmf_settings$similarity_n_perm)
similarity_q_threshold <- as.numeric(ebmf_settings$similarity_q)
high_similarity_q <- as.numeric(ebmf_settings$high_similarity_q)
high_connectedness <- as.numeric(ebmf_settings$high_connectedness)
min_effective_traits <- as.numeric(ebmf_settings$min_effective_traits)
ebmf_greedy_Kmax <- as.integer(ebmf_settings$ebmf_greedy_Kmax)
associations <- params$associations
include_situated_gene <- isTRUE(params$include_situated_gene)
min_mean_internal <- as.numeric(params$min_mean_internal)
min_connectedness <- as.numeric(params$min_connectedness)
n_null <- as.integer(params$n_null)
alpha_membership <- as.numeric(params$alpha_membership)
n_candidate_tier <- as.integer(params$n_candidate_tier)
n_rep <- as.integer(params$n_rep)
frac_traits <- as.numeric(params$frac_traits)
stability_top_n <- as.integer(params$stability_top_n)
posterior_stat <- params$posterior_stat
posterior_evidence_cap <- as.numeric(params$posterior_evidence_cap)
redundancy_threshold <- as.numeric(params$redundancy_threshold)
cores <- as.integer(params$cores)

knitr::kable(
  data.frame(
    api = params$api,
    gpmap_url = getOption("gpmap_url"),
    target_guid = target_guid,
    p_threshold = if (is.null(p_threshold)) "upload default" else format(p_threshold),
    associations = associations,
    trait_subset = trait_subset,
    include_trans = include_trans,
    min_snp_signals = min_snp_signals,
    compress_method = compress_method,
    compress_scale = compress_scale,
    ebmf_prior = ebmf_prior,
    ebmf_backfit = ebmf_backfit,
    ebmf_lfsr_threshold = ebmf_lfsr_threshold,
    ebmf_magnitude_threshold = ebmf_magnitude_threshold,
    ebmf_greedy_Kmax = ebmf_greedy_Kmax,
    min_module_size = min_module_size,
    similarity_threshold = similarity_threshold,
    similarity_n_perm = similarity_n_perm,
    similarity_q = similarity_q_threshold,
    high_similarity_q = high_similarity_q,
    high_connectedness = high_connectedness,
    min_effective_traits = min_effective_traits,
    n_rep = n_rep,
    frac_traits = frac_traits,
    redundancy_threshold = redundancy_threshold,
    stringsAsFactors = FALSE
  ),
  caption = "Active run parameters",
  row.names = FALSE
)
```

### T2D proxy traits

GPMap holds several other GWAS of T2D itself, or of traits defined by having
diabetes. At risk-oriented T2D SNPs they all repeat the anchor trait, so they
add no information about mechanism. They can also give EBMF extra weight to
spend on small differences between near-identical rows. They are removed from
the matrix before fitting.

The list was drawn up by searching `all_traits()` and the upload's coloc
partners for diabetes-related names, then pruned by hand. It covers:

- **T2D GWAS:** GPMap's own T2D and the UK Biobank SPA/Firth versions.
- **Diagnosis:** doctor-diagnosed, self-reported and ICD10 E11.9 diabetes.
- **Treatment:** diabetes medication, metformin, gliclazide and insulin.
- **Family history:** diabetes in the mother, father or siblings.
- **Complications:** diabetes-related eye disease.

Kept on purpose:

- **Glycaemic traits** (glucose, HbA1c, insulin and proinsulin), which are
  Suzuki clustering traits.
- **Type 1 diabetes**, a distinct disease.
- **Combined medication fields**, such as "Medication for cholesterol, blood
  pressure or diabetes: Blood pressure medication".

```{r t2d-proxies}
t2d_proxy_traits <- data.frame(
  trait_id = c(
    "114", "2499", "2549",
    "2863", "3106", "4536",
    "1494", "3270", "4422", "3333", "4300", "4403",
    "3428", "3779", "3580",
    "2814"
  ),
  trait_name = c(
    "Type 2 diabetes",
    "Type 2 diabetes (Firth correction)",
    "Type 2 diabetes (SPA correction)",
    "Diabetes diagnosed by doctor",
    "Non-cancer illness code, self-reported: diabetes",
    "Diagnoses - secondary ICD10: E11.9 Without complications",
    "Medication use (drugs used in diabetes)",
    "Treatment/medication code: metformin",
    "Treatment/medication code: gliclazide",
    "Treatment/medication code: insulin product",
    "Medication for cholesterol, blood pressure or diabetes: Insulin",
    "Started insulin within one year diagnosis of diabetes",
    "Illnesses of mother: Diabetes",
    "Illnesses of father: Diabetes",
    "Illnesses of siblings: Diabetes",
    "Eye problems/disorders: Diabetes related eye disease"
  ),
  group = c(
    rep("T2D GWAS", 3), rep("Diagnosis", 3), rep("Treatment", 6),
    rep("Family history", 3), "Complication"
  ),
  stringsAsFactors = FALSE
)

proxy_counts <- target$coloc_groups |>
  dplyr::mutate(trait_id = as.character(trait_id)) |>
  dplyr::filter(trait_id %in% t2d_proxy_traits$trait_id) |>
  dplyr::group_by(trait_id) |>
  dplyr::summarise(n_coloc_groups = dplyr::n_distinct(coloc_group_id), .groups = "drop")

knitr::kable(
  t2d_proxy_traits |>
    dplyr::left_join(proxy_counts, by = "trait_id") |>
    dplyr::mutate(n_coloc_groups = tidyr::replace_na(n_coloc_groups, 0L)),
  caption = paste0(
    "T2D proxy traits removed before building the matrix. n_coloc_groups = ",
    "the upload's coloc groups that each trait shares."
  ),
  row.names = FALSE
)

target$coloc_groups <- target$coloc_groups |>
  dplyr::filter(!as.character(trait_id) %in% t2d_proxy_traits$trait_id)
```

### Fit

```{r fit}
set.seed(20240214)
ebmf_result <- run_univariate_clustering(
  target,
  target_trait_id = target_guid,
  associations = associations,
  p_threshold = p_threshold,
  trait_subset = trait_subset,
  include_trans = include_trans,
  min_snp_signals = min_snp_signals,
  max_snp_fraction = max_snp_fraction,
  compress_method = compress_method,
  compress_scale = compress_scale,
  ebmf_lfsr_threshold = ebmf_lfsr_threshold,
  ebmf_magnitude_threshold = ebmf_magnitude_threshold,
  ebmf_greedy_Kmax = ebmf_greedy_Kmax,
  ebmf_prior = ebmf_prior,
  ebmf_backfit = ebmf_backfit,
  min_module_size = min_module_size,
  min_mean_internal = min_mean_internal,
  min_connectedness = min_connectedness
)
X <- ebmf_result$x_matrix
S <- ebmf_result$s_matrix
cg <- ebmf_result$coloc_groups

knitr::kable(
  data.frame(
    n_traits = nrow(X),
    n_phenotypic = sum(ebmf_result$trait_info$feature_type == "phenotypic", na.rm = TRUE),
    n_traits_dropped = length(ebmf_result$dropped_trait_ids),
    n_snps = ncol(X),
    fraction_non_na = round(mean(!is.na(X)), 3),
    n_ebmf_factors = ebmf_result$cluster_details$n_programs,
    dropped_global_factors = length(ebmf_result$cluster_details$dropped_global_factors),
    greedy_Kmax_reached = isTRUE(ebmf_result$cluster_details$greedy_Kmax_reached)
  ),
  caption = "Pleiotropy matrix after filtering, and the EBMF fit",
  align = "r"
)
```

### Program confidence

`summarise_ebmf_programs()` drops programs with fewer than `min_module_size`
member SNPs and gives every other program a **confidence tier**. Every kept
program is annotated in Part 3, whatever its tier; Part 4 compares only the
high and medium programs with Suzuki's clusters.

Member SNPs are those passing the lFSR/magnitude cut. `S` is the cosine
similarity between two SNPs' oriented, compressed profiles across the
background traits, with missing cells counted as 0 and the T2D row left out
(it is positive at every SNP, so it would add the same similarity to every
pair).

- **low** — `similarity_q >= similarity_q`: the member SNPs are no more alike
  than random SNP groups of the same size. This is checked first, so a program
  that fails it is low whatever its number of traits, and can be set aside.
- **single_trait** — passes the similarity test, but the program's trait
  loadings rest on fewer than `min_effective_traits` effective traits. A
  program is defined as multi-trait: a factor carried by one trait is just
  that trait's hits, which noise produces too, so it needs a closer look
  rather than being discarded.
- **high** — `similarity_q < high_similarity_q` and `connectedness >=
  high_connectedness`.
- **medium** — `similarity_q < similarity_q`, but not high.

The columns:

- **effective_n_traits**: `1 / sum(p^2)`, where `p` are the program's squared
  trait loadings as shares of their total (T2D row left out): 1 for a program
  on one trait, `n` for `n` equally loaded traits.

- **mean_internal_similarity** and **similarity_q**: the mean `S` between pairs
  of member SNPs, and the BH-adjusted share of random groups of the same number
  of T2D SNPs that are at least as similar. A significance test, so larger
  programs pass more easily.
- **connectedness**: the share of member pairs with `S >= similarity_threshold`.
- **excess_similarity** and **excess_connectedness** (reported): each of the
  above minus its value in the random groups; effect sizes that do not grow
  with program size.
- **factor_pve_pct** (reported): the share of the matrix variance the factor
  explains.
- **replication** (reported): the mean Jaccard overlap between the program's
  top SNPs and the best-matching factor when 20% of the background traits are
  held out and EBMF is refit.
- **parent_program**: when at least 90% of a program's SNPs sit at the loci of
  a larger program, that program is its parent. Sub-programs are kept and listed under
  their parent.

```{r validation, results = "asis"}
program_summary <- summarise_ebmf_programs(
  ebmf_result,
  s_matrix = S,
  edge_threshold = similarity_threshold,
  min_module_size = min_module_size,
  similarity_n_perm = similarity_n_perm,
  similarity_q_threshold = similarity_q_threshold,
  high_similarity_q = high_similarity_q,
  high_min_connectedness = high_connectedness,
  min_effective_traits = min_effective_traits,
  n_null = n_null,
  alpha_membership = alpha_membership,
  n_candidate_tier = n_candidate_tier,
  n_rep = n_rep,
  frac_traits = frac_traits,
  top_n = stability_top_n,
  posterior_stat = posterior_stat,
  posterior_evidence_cap = posterior_evidence_cap,
  redundancy_threshold = redundancy_threshold,
  cores = cores,
  verbose = FALSE
)
# Every kept program (at least min_module_size SNPs) is annotated; the tier is
# reported, not applied.
analysis_ids <- program_summary$programs$program

print(kable_confidence(
  program_summary$programs |>
    dplyr::transmute(
      program = program_label,
      confidence_tier = as.character(confidence_tier),
      n_snps_filtered,
      effective_n_traits = round(effective_n_traits, 2),
      mean_internal_similarity = round(mean_internal_similarity, 3),
      excess_similarity = round(excess_similarity, 3),
      connectedness = round(connectedness, 3),
      excess_connectedness = round(excess_connectedness, 3),
      similarity_q = signif(similarity_q, 2),
      factor_pve_pct = round(100 * factor_pve, 2),
      replication = round(replication, 2),
      parent_program
    ),
  caption = paste0(
    "Every EBMF program by confidence tier (green high, orange medium, grey ",
    "single trait, red low), each child program indented under its parent."
  )
))

clusters_analysis <- program_summary$assigned |>
  dplyr::filter(program %in% analysis_ids) |>
  dplyr::transmute(snp_id = as.character(snp_id), group = as.character(program))

print(knitr::kable(
  data.frame(
    n_programs = nrow(program_summary$programs),
    n_high = sum(program_summary$programs$confidence_tier == "high"),
    n_medium = sum(program_summary$programs$confidence_tier == "medium"),
    n_single_trait = sum(program_summary$programs$confidence_tier == "single_trait"),
    n_low = sum(program_summary$programs$confidence_tier == "low"),
    n_dropped_small = nrow(program_summary$dropped_programs),
    n_snps = ncol(X),
    n_snps_in_analysed_programs = dplyr::n_distinct(clusters_analysis$snp_id)
  ),
  caption = paste0(
    "Program confidence summary. n_dropped_small = programs below ",
    "min_module_size, removed before any check. T2D background: mean S = ",
    round(program_summary$background$mean_similarity, 3), ", share of SNP pairs ",
    "with S >= ", similarity_threshold, " = ",
    round(program_summary$background$frac_pairs_edge, 3), "."
  ),
  row.names = FALSE
))
```

**Program distinctness.** How alike the kept programs are, on both axes of the
fit, with the same confidence weighting as the cross-trait links
(`program_distinctness()`):

- **max_abs_phi** and **n_pairs_phi_0.7**: the largest Tucker congruence
  between two programs' trait loadings (T2D row left out), and the number of
  pairs at or above 0.7. Tucker's conventions call 0.85 fair similarity and
  0.95 equal.
- **max_alignment**: the largest squared uncentred correlation between two
  programs' SNP loadings.
- **effective_rank**: the effective number of distinct programs, from the
  eigenvalues of the programs' Gram matrix; it equals `n_programs` when every
  pair is orthogonal and falls as programs overlap.

```{r distinctness}
distinctness <- program_distinctness(ebmf_result, analysis_ids)
knitr::kable(
  distinctness$summary |>
    dplyr::mutate(
      max_abs_phi = round(max_abs_phi, 2),
      max_alignment = round(max_alignment, 2),
      effective_rank = round(effective_rank, 1)
    ),
  caption = "Distinctness of the kept programs",
  row.names = FALSE
)
```

Across the `r distinctness$summary$n_programs` kept programs, the largest
|φ| is `r round(distinctness$summary$max_abs_phi, 2)`
(`r distinctness$summary$n_pairs_phi_0.7` pairs at or above 0.7), the largest
SNP alignment is `r round(distinctness$summary$max_alignment, 2)`, and the
effective rank is `r round(distinctness$summary$effective_rank, 1)`.

### Module-specific traits and loading-based enrichment

One row per analysed program, in confidence order (green high, orange medium,
grey single trait, red low), with each child program indented under its
parent. The columns come from two analyses.

**Module-specific traits.** The background traits most elevated in the
program relative to the rest of the T2D SNPs:
`specificity = mean |z| in program / (mean |z| elsewhere + 0.1)`, shown in
brackets. Traits need a specificity of at least `r min_specificity`, and the
top 8 are listed.

**Loading-based enrichment** of trait categories, tissues and pathways
(KEGG and Reactome, on the colocalising QTL genes). There is no membership
cutoff: every SNP contributes to every program, weighted by its squared EBMF
loading, in the regression `lm(loading^2 ~ membership)`. A hit needs both
FDR ≤ 0.05 and R² ≥ 0.05. The FDR is computed across every program tested,
separately for categories, tissues and pathways.

**Tissues differ from Suzuki's.** GPMap's tissues are GTEx-style bulk tissues.
"Pancreas" is bulk GTEx pancreas, mostly exocrine, and there are no islets.
Suzuki's islet-cell enrichments therefore have no direct counterpart here.

```{r program-annotation, results = "asis"}
has_analysis_groups <- nrow(clusters_analysis) > 0
min_r_squared <- 0.05

group_traits <- summarise_module_specific_traits(
  trait_matrix = ebmf_result$trait_matrix,
  groups = clusters_analysis,
  trait_info = ebmf_result$trait_info,
  coloc_groups = cg,
  exclude_trait_ids = target_guid,
  min_group_size = min_module_size,
  top_n = top_n_traits,
  n_categories = 3L,
  min_specificity = min_specificity
)
module_traits_overview <- dplyr::bind_rows(lapply(group_traits$by_group, function(x) {
  passing <- x$traits |>
    dplyr::filter(passes_min_specificity) |>
    utils::head(8L)
  return(data.frame(
    program = as.character(x$group),
    module_specific_traits = if (nrow(passing) == 0) {
      "none"
    } else {
      paste0(
        passing$trait_name, " (", round(passing$specificity, 2), ")",
        collapse = "<br>"
      )
    },
    stringsAsFactors = FALSE
  ))
}))

.format_continuous_all <- function(by_program_list, value_col, source = NULL) {
  return(vapply(by_program_list, function(x) {
    hits <- x$comparison
    if (is.null(hits) || nrow(hits) == 0) {
      return("none")
    }
    keep <- is.finite(hits$fdr) & hits$fdr <= 0.05 &
      is.finite(hits$r_squared) & hits$r_squared >= min_r_squared
    if (!is.null(source)) {
      keep <- keep & hits$source == source
    }
    hits <- hits[keep, , drop = FALSE]
    if (nrow(hits) == 0) {
      return("none")
    }
    hits <- hits[order(-hits$enrichment), , drop = FALSE]
    return(paste0(
      hits[[value_col]],
      " (effect=", signif(hits$enrichment, 2),
      ", R2=", signif(hits$r_squared, 2),
      ", FDR=", signif(hits$fdr, 3), ")",
      collapse = "<br>"
    ))
  }, character(1)))
}
valid_by_program <- function(res) {
  if (is.null(res)) {
    return(list())
  }
  out <- Filter(
    function(x) return(as.character(x$program) %in% as.character(analysis_ids)),
    res$by_program
  )
  return(stats::setNames(out, vapply(out, function(x) as.character(x$program), character(1))))
}

if (!has_analysis_groups) {
  cat("No programs to annotate.\n")
  cat_by <- list()
  tis_by <- list()
  path_by <- list()
} else {
  loading_categories <- enrich_program_loadings_trait_categories(
    clustering_result = ebmf_result,
    coloc_groups = cg,
    snp_key = "variant_id"
  )
  loading_tissues <- enrich_program_loadings_tissues(
    clustering_result = ebmf_result,
    coloc_groups = cg,
    snp_key = "variant_id"
  )
  pathway_universe <- tryCatch(
    pathway_mappings(),
    error = function(e) {
      cat("Pathway mappings endpoint unavailable (", conditionMessage(e), ").\n", sep = "")
      return(NULL)
    }
  )
  loading_pathways <- if (is.null(pathway_universe)) {
    NULL
  } else {
    enrich_program_loadings_pathways(
      clustering_result = ebmf_result,
      coloc_groups = cg,
      mappings = pathway_universe,
      snp_key = "variant_id",
      sources = c("KEGG", "Reactome"),
      include_situated_gene = include_situated_gene
    )
  }
  cat_by <- valid_by_program(loading_categories)
  tis_by <- valid_by_program(loading_tissues)
  path_by <- valid_by_program(loading_pathways)
  programs_shown <- names(cat_by)
  continuous_overview <- data.frame(
    program = programs_shown,
    trait_categories = .format_continuous_all(cat_by, "trait_category"),
    tissues = .format_continuous_all(tis_by[programs_shown], "tissue"),
    reactome = if (length(path_by) == 0) {
      "not available"
    } else {
      .format_continuous_all(path_by[programs_shown], "description", "Reactome")
    },
    kegg = if (length(path_by) == 0) {
      "not available"
    } else {
      .format_continuous_all(path_by[programs_shown], "description", "KEGG")
    },
    stringsAsFactors = FALSE
  )

  program_annotation <- program_summary$programs |>
    dplyr::transmute(
      program = as.character(program),
      confidence_tier = as.character(confidence_tier),
      similarity_z,
      n_snps_filtered,
      parent_program
    ) |>
    dplyr::left_join(dplyr::count(clusters_analysis, program = group, name = "n_snps"), by = "program") |>
    dplyr::left_join(module_traits_overview, by = "program") |>
    dplyr::left_join(continuous_overview, by = "program") |>
    dplyr::mutate(dplyr::across(
      c(module_specific_traits, trait_categories, tissues, reactome, kegg),
      function(x) return(tidyr::replace_na(x, "none"))
    )) |>
    order_programs_by_confidence() |>
    dplyr::transmute(
      program = program_label, confidence_tier, n_snps, module_specific_traits,
      trait_categories, tissues, reactome, kegg
    )
  print(kable_confidence(
    program_annotation,
    drop_tier_col = TRUE,
    caption = paste0(
      "Module-specific traits (specificity in brackets) and loading-based ",
      "enrichment (FDR <= 0.05 and R2 >= ", min_r_squared, ") for every ",
      "analysed program, by confidence tier."
    )
  ))
}
```

```{r program-gene-lookup, include = FALSE}
# Genes at each analysed program's member SNPs, used by the marker-gene check
# and the program cards in Part 4.
#
# - Positional: the nearest protein-coding gene to each SNP (0 bp if the SNP
#   falls inside the gene; all overlapping genes are listed). This is how
#   Suzuki et al. name their loci.
# - Colocalising QTL: genes whose molecular QTL colocalises with the SNP
#   (genes_at_snps()). Closer to mechanism, but can differ from the locus name
#   (the FTO signal acts through IRX3/IRX5).
gene_table <- all_genes() |>
  dplyr::filter(gene_biotype == "protein_coding") |>
  dplyr::transmute(gene_id = as.integer(id), gene, chr = as.character(chr), start, stop)

snp_positions <- cg |>
  dplyr::filter(as.character(variant_id) %in% colnames(X)) |>
  dplyr::distinct(variant_id, chr, bp) |>
  dplyr::transmute(snp_id = as.character(variant_id), chr = as.character(chr), bp) |>
  dplyr::distinct(snp_id, .keep_all = TRUE)

# Nearest protein-coding gene(s) to every T2D SNP: distance 0 inside a gene.
positional_links <- dplyr::bind_rows(lapply(split(snp_positions, snp_positions$chr), function(snps) {
  genes_chr <- gene_table[gene_table$chr == snps$chr[1], , drop = FALSE]
  if (nrow(genes_chr) == 0) {
    return(NULL)
  }
  return(dplyr::bind_rows(lapply(seq_len(nrow(snps)), function(i) {
    d <- pmax(genes_chr$start - snps$bp[i], snps$bp[i] - genes_chr$stop, 0)
    hit <- which(d == min(d))
    return(data.frame(
      snp_id = snps$snp_id[i], gene_id = genes_chr$gene_id[hit],
      gene = genes_chr$gene[hit], distance = d[hit], stringsAsFactors = FALSE
    ))
  })))
}))
qtl_links <- genes_at_snps(colnames(X), coloc_groups = cg, snp_key = "variant_id") |>
  dplyr::transmute(snp_id = as.character(snp_id), gene_id = as.integer(gene_id), gene) |>
  dplyr::distinct()

# Every gene Suzuki et al. name at an index SNV, with its cluster(s).
suzuki_gene_clusters <- dplyr::bind_rows(
  suzuki_snvs |>
    dplyr::select(cluster, gene = locus) |>
    tidyr::separate_rows(gene, sep = ",\\s*"),
  suzuki_clusters |>
    dplyr::select(cluster, gene = example_loci) |>
    tidyr::separate_rows(gene, sep = "; ") |>
    tidyr::separate_rows(gene, sep = "-")
) |>
  dplyr::distinct() |>
  dplyr::group_by(gene) |>
  dplyr::summarise(
    suzuki_clusters = paste(cluster_levels[cluster_levels %in% cluster], collapse = ", "),
    .groups = "drop"
  )
suzuki_genes <- suzuki_gene_clusters$gene

format_genes <- function(genes, max_n = 15L) {
  genes <- unique(genes)
  if (length(genes) == 0) {
    return("none")
  }
  shown <- utils::head(genes, max_n)
  shown <- ifelse(shown %in% suzuki_genes, paste0("<b>", shown, "</b>"), shown)
  more <- length(genes) - length(shown)
  return(paste0(
    paste(shown, collapse = ", "),
    if (more > 0) paste0(" (+", more, " more)") else ""
  ))
}

# Member SNPs of each analysed program, strongest absolute loading first.
F_pm <- ebmf_result$cluster_details$flash_fit$F_pm
program_members <- clusters_analysis |>
  dplyr::rename(program = group) |>
  dplyr::filter(snp_id %in% rownames(F_pm)) |>
  dplyr::mutate(
    abs_loading = abs(F_pm[cbind(match(snp_id, rownames(F_pm)), as.integer(program))])
  ) |>
  dplyr::arrange(program, dplyr::desc(abs_loading))

genes_in_order <- function(members, links) {
  ordered <- members |>
    dplyr::inner_join(links, by = "snp_id", relationship = "many-to-many") |>
    dplyr::group_by(gene) |>
    dplyr::summarise(n_snps = dplyr::n_distinct(snp_id), top = max(abs_loading), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(top), dplyr::desc(n_snps))
  return(ordered$gene)
}
program_genes <- dplyr::bind_rows(lapply(split(program_members, program_members$program), function(m) {
  positional <- genes_in_order(m, positional_links)
  qtl <- genes_in_order(m, qtl_links)
  return(data.frame(
    program = m$program[1],
    n_snps = nrow(m),
    n_positional_genes = length(positional),
    positional_genes = format_genes(positional),
    n_qtl_genes = length(qtl),
    qtl_genes = format_genes(qtl),
    stringsAsFactors = FALSE
  ))
}))

# Member SNPs of each analysed program, as a named list.
program_snps <- split(clusters_analysis$snp_id, clusters_analysis$group)

# Part 4 compares only the high and medium confidence programs with Suzuki.
strong_programs <- as.character(program_summary$programs$program[
  program_summary$programs$confidence_tier %in% c("high", "medium")
])
```

## Part 4: Do the programs recapture Suzuki's clusters?

### Expectations (written before the results)

- **Likely to be recovered:** Obesity, Lipodystrophy and Metabolic syndrome.
  Their defining traits (BMI, WHR, body fat %, TG, HDL, blood pressure) are
  large UK Biobank GWAS with hundreds of coloc groups in GPMap.
- **Not expected: the Beta cell +PI / −PI split.** The two clusters differ
  mainly in the sign of the proinsulin effect. GPMap's proinsulin GWAS has
  N = 10,701 and 12 coloc groups, so a joint beta-cell program would be the
  realistic outcome.
- **Glycaemic clusters lean on HbA1c.** Fasting glucose, fasting insulin and
  2-hour glucose are sparse in GPMap, so HbA1c carries most of the glycaemic
  signal.
- **Body fat vs Obesity may merge.** They differ mainly on abdominal
  subcutaneous and visceral fat, which have only 2–7 coloc groups in GPMap.
- **Liver/lipid metabolism (3 signals) is not expected.**
- **Tissues.** A Lipodystrophy-like program should be enriched in adipose, an
  Obesity-like program possibly in brain or hypothalamus (Suzuki found brain
  neurons), and a liver program in liver. Beta-cell programs are **not**
  expected to show pancreas: GPMap has no islet QTLs.

Part 4 compares only the **high and medium confidence** programs with
Suzuki's clusters. Low and single-trait programs are listed in Part 3 but are
not interpreted here. The comparison has five parts, after a check of which
Suzuki traits the matrix contains:

- **4b:** Suzuki's canonical marker genes, found positionally and through
  colocalising QTLs.
- **4c:** each Suzuki index SNV, matched to the GPMap SNP in its LD block, and
  the programs that SNP ended up in.
- **4d:** each program's mean profile over the Suzuki clustering traits,
  compared with each cluster's profile.
- **4e:** one summary card per program.
- **4f:** a hand-written reading of which program matches which cluster,
  using 4c–4e together.

### 4a. Trait coverage

The univariate matrix already contains every phenotypic GPMap trait that
colocalises with the upload's SNPs. `trait_map.csv` records which of those rows
are Suzuki's 37 clustering traits:

- `same_gwas`: the GWAS Suzuki used;
- `proxy`: a different GWAS of the same trait;
- `unadjusted_proxy`: Suzuki used a BMI-adjusted trait and GPMap has it
  unadjusted;
- `none`: no GPMap equivalent.

Nothing extra is fetched or fitted for these traits.

```{r coverage}
suzuki_trait_map <- suzuki_trait_map |>
  dplyr::mutate(
    gpmap_trait_id = as.character(gpmap_trait_id),
    in_matrix = !is.na(gpmap_trait_id) & gpmap_trait_id %in% rownames(X),
    n_snps_observed = dplyr::if_else(
      in_matrix,
      as.integer(rowSums(!is.na(X))[gpmap_trait_id]),
      0L
    )
  )

knitr::kable(
  suzuki_trait_map |>
    dplyr::group_by(match_type) |>
    dplyr::summarise(
      n_traits = dplyr::n(),
      n_in_matrix = sum(in_matrix),
      .groups = "drop"
    ),
  caption = "Suzuki traits available as rows of the pleiotropy matrix, by match type",
  row.names = FALSE
)

knitr::kable(
  suzuki_trait_map |>
    dplyr::select(
      trait_category, trait_acronym, gpmap_trait_id, gpmap_trait_name,
      match_type, in_matrix, n_snps_observed
    ),
  caption = paste0(
    "Suzuki clustering traits and their GPMap rows. n_snps_observed = upload ",
    "SNPs at which the GPMap trait colocalises."
  ),
  row.names = FALSE
)
```

### 4b. Canonical marker genes

The example loci that Suzuki et al. give for each cluster (Table 1) are
checked against the genes at the member SNPs of each high or medium program,
in two ways:

- **positional:** the gene is the nearest protein-coding gene to one of the
  program's SNPs, which is how Suzuki named the locus;
- **QTL:** the gene's molecular QTL colocalises with one of the program's SNPs.

A gene found neither way is labelled "low / single-trait program only" if it
is at a member SNP of a program outside those tiers, "at a matrix SNP, not in
a program" if some SNP in the pleiotropy matrix carries it, and "not found"
otherwise. "Not found" usually means the locus has no upload signal that
colocalises with enough traits to enter the matrix (see 4c).

```{r marker-genes}
marker_genes <- suzuki_clusters |>
  dplyr::select(cluster, example_loci) |>
  tidyr::separate_rows(example_loci, sep = "; ") |>
  tidyr::separate_rows(example_loci, sep = "-") |>
  dplyr::rename(gene = example_loci) |>
  dplyr::distinct()

# Genes at each program's SNPs, from both sources.
program_gene_links <- function(links) {
  return(dplyr::bind_rows(lapply(names(program_snps), function(k) {
    genes <- links[links$snp_id %in% program_snps[[k]], , drop = FALSE]
    if (nrow(genes) == 0) {
      return(NULL)
    }
    return(dplyr::distinct(
      data.frame(program = k, gene_id = genes$gene_id, gene = genes$gene, stringsAsFactors = FALSE)
    ))
  })))
}
genes_by_program <- list(positional = program_gene_links(positional_links), QTL = program_gene_links(qtl_links))
trait_genes <- list(positional = unique(positional_links$gene), QTL = unique(qtl_links$gene))

# The programs in `ids` at which gene g is found, as "P1, P3"; "" if none.
programs_with <- function(g, source, ids = strong_programs) {
  tab <- genes_by_program[[source]]
  if (is.null(tab) || nrow(tab) == 0) {
    return("")
  }
  found <- unique(tab$program[tab$gene == g & tab$program %in% ids])
  if (length(found) == 0) {
    return("")
  }
  return(paste(paste0("P", found[order(as.integer(found))]), collapse = ", "))
}
marker_status <- marker_genes |>
  dplyr::rowwise() |>
  dplyr::mutate(
    positional = programs_with(gene, "positional"),
    qtl = programs_with(gene, "QTL"),
    in_other_program = programs_with(gene, "positional", names(program_snps)) != "" |
      programs_with(gene, "QTL", names(program_snps)) != "",
    status = dplyr::case_when(
      positional != "" | qtl != "" ~ "in a high/medium program",
      in_other_program ~ "low / single-trait program only",
      gene %in% c(trait_genes$positional, trait_genes$QTL) ~ "at a matrix SNP, not in a program",
      TRUE ~ "not found"
    )
  ) |>
  dplyr::ungroup()

knitr::kable(
  marker_status |>
    dplyr::mutate(cluster = factor(cluster, levels = cluster_levels)) |>
    dplyr::arrange(cluster) |>
    dplyr::select(cluster, gene, positional, qtl, status),
  caption = paste0(
    "Suzuki Table 1 example-locus genes. positional / qtl list the high or ",
    "medium programs in which the gene is the nearest gene to a member SNP, ",
    "or a colocalising QTL gene."
  ),
  row.names = FALSE
)

knitr::kable(
  marker_status |>
    dplyr::group_by(cluster) |>
    dplyr::summarise(
      n_markers = dplyr::n(),
      n_at_matrix_snps = sum(status != "not found"),
      n_in_any_program = sum(status %in% c("in a high/medium program", "low / single-trait program only")),
      n_in_high_medium_program = sum(status == "in a high/medium program"),
      .groups = "drop"
    ) |>
    dplyr::mutate(cluster = factor(cluster, levels = cluster_levels)) |>
    dplyr::arrange(cluster),
  caption = "Marker-gene summary per Suzuki cluster",
  row.names = FALSE
)
```

### 4c. Suzuki index SNVs in the programs

Suzuki's clusters are sets of index SNVs, so the most direct comparison is at
the SNP level. Gene names are not enough for this: Suzuki's 1,289 signals sit
at 611 named loci, and 201 of those loci have signals in more than one cluster
(TCF7L2 has 32 signals across four clusters).

Each Suzuki index SNV is matched to the upload as follows:

- **Does the region have any upload signal at all?** The SNV's GRCh38
  position (`pos_b38` in `index_snv_clusters.csv`, looked up by rsid in
  Ensembl) is checked against the LD blocks of the upload's own signals and
  of the pleiotropy matrix's SNPs. This is a coarse first gate, not the
  match itself -- GPMap's LD blocks span whole regions (tens of Mb in places),
  so sharing a block does not mean two variants are in LD.
- **Match.** The SNV's rsid is resolved to a GPMap variant id (`variants()`),
  then matched to the pleiotropy-matrix SNP with the strongest linkage
  disequilibrium to it (`ld_proxies()`), taking an exact match (`r` = 1)
  first. The two are usually different variants even when they tag the same
  signal: GPMap's lead variant comes from its own fine-mapping of the EUR
  upload, while Suzuki's index SNV comes from the multi-ancestry
  meta-regression. LD is looked up from both sides, because GPMap only
  stores proxy lists for some variants -- almost none of Suzuki's SNVs have
  one, but the matrix SNPs mostly do. GPMap's proxy table stops at
  `r`^2^ = 0.8, so a SNV whose rsid does not resolve, or that has no matrix
  SNP within `r`^2^ ≥ 0.8, has no match. A negative `ld_r` means the two
  variants' alleles are in opposite phase.
- **Status.** Each SNV then ends up in one of five places: no upload signal
  in its LD block; an upload signal that is not in the matrix (it colocalises
  with too few traits, or no matrix SNP is in strong LD with it); in the
  matrix but in no program; in low or single-trait programs only; or in at
  least one high or medium program.

Several Suzuki SNVs, sometimes from different clusters, can match the same
GPMap SNP. In the per-program counts below, each matched GPMap SNP counts once,
split evenly across the SNVs it matches: a SNP matched by two Obesity SNVs and
one Body fat SNV adds 2/3 to Obesity and 1/3 to Body fat.

```{r suzuki-snv-matching, results = "asis"}
# LD block of every upload signal, and of every pleiotropy-matrix SNP.
# Blocks are named "<population>/<chr>/<start>-<end>" on GRCh38.
parse_blocks <- function(blocks) {
  blocks <- unique(stats::na.omit(as.character(blocks)))
  parts <- do.call(rbind, strsplit(sub("^[^/]+/", "", blocks), "[/-]"))
  return(data.frame(
    ld_block = blocks,
    chr = parts[, 1],
    start = as.numeric(parts[, 2]),
    end = as.numeric(parts[, 3]),
    stringsAsFactors = FALSE
  ))
}
upload_blocks <- parse_blocks(target$upload_study_extractions$ld_block)
matrix_snps <- cg |>
  dplyr::filter(as.character(variant_id) %in% colnames(X)) |>
  dplyr::transmute(
    snp_id = as.character(variant_id), gpmap_rsid = rsid,
    chr = as.character(chr), bp, ld_block
  ) |>
  dplyr::distinct(snp_id, .keep_all = TRUE)
all_blocks <- parse_blocks(c(upload_blocks$ld_block, matrix_snps$ld_block))

# Coarse first gate only (see the prose above): does the region have any
# upload signal at all.
snv_block <- vapply(seq_len(nrow(suzuki_snvs)), function(i) {
  hit <- which(
    all_blocks$chr == as.character(suzuki_snvs$chr[i]) &
      all_blocks$start <= suzuki_snvs$pos_b38[i] &
      all_blocks$end > suzuki_snvs$pos_b38[i]
  )
  return(if (length(hit) == 0) NA_character_ else all_blocks$ld_block[hit[1]])
}, character(1))

# The actual match: each SNV's rsid resolved to a GPMap variant id
# (variants(), <= 50 per call), then matched to the pleiotropy-matrix SNP
# with the best linkage disequilibrium to it (.ld_best_match(), the same
# primitive the multi-trait comparison uses to bridge an uploaded trait's
# own locus numbering to another trait's). ld_proxies()'s own rsid/string
# input does not resolve rsids -- only a GPMap variant id works, which is
# exactly what variants() looks up.
resolve_rsids <- function(rsids) {
  empty <- data.frame(rsid = character(0), gpmap_variant_id = character(0), stringsAsFactors = FALSE)
  rsids <- unique(stats::na.omit(as.character(rsids)))
  if (length(rsids) == 0) {
    return(empty)
  }
  chunks <- split(rsids, ceiling(seq_along(rsids) / 50))
  out <- dplyr::bind_rows(lapply(chunks, function(ids) {
    res <- tryCatch(variants(variants = ids, expand = FALSE), error = function(e) NULL)
    if (is.null(res) || !is.data.frame(res$variants) || nrow(res$variants) == 0) {
      return(NULL)
    }
    return(data.frame(
      rsid = as.character(res$variants$rsid),
      gpmap_variant_id = as.character(res$variants$id),
      stringsAsFactors = FALSE
    ))
  }))
  if (is.null(out) || nrow(out) == 0) {
    return(empty)
  }
  return(out[!duplicated(out$rsid), , drop = FALSE])
}
resolved_rsids <- resolve_rsids(suzuki_snvs$rsid_current)
ld_matches <- .ld_best_match(
  resolved_rsids$gpmap_variant_id, matrix_snps$snp_id, min_r = 0.5
)

snv_match <- suzuki_snvs |>
  dplyr::transmute(.row = dplyr::row_number(), rsid_current) |>
  dplyr::left_join(resolved_rsids, by = c("rsid_current" = "rsid")) |>
  dplyr::left_join(
    ld_matches,
    by = c("gpmap_variant_id" = "query_variant_id")
  ) |>
  dplyr::left_join(
    matrix_snps, by = c("target_variant_id" = "snp_id")
  ) |>
  dplyr::arrange(.row) |>
  dplyr::transmute(
    snp_id = target_variant_id, gpmap_rsid, r,
    distance_kb = round(abs(bp - suzuki_snvs$pos_b38) / 1000, 1)
  )

# Programs each matrix SNP belongs to, split into high/medium and the rest.
membership <- program_summary$assigned |>
  dplyr::transmute(snp_id = as.character(snp_id), program = as.character(program), loading) |>
  dplyr::filter(program %in% names(program_snps))
programs_of <- function(snp, ids) {
  m <- membership[membership$snp_id == snp & membership$program %in% ids, , drop = FALSE]
  m <- m[order(-abs(m$loading)), , drop = FALSE]
  return(if (nrow(m) == 0) "" else paste0("P", m$program, collapse = ", "))
}
other_programs <- setdiff(names(program_snps), strong_programs)

status_levels <- c(
  "no upload signal in LD block", "upload signal, not in matrix",
  "in matrix, no program", "low / single-trait program only",
  "high/medium program"
)
snv_status <- suzuki_snvs |>
  dplyr::mutate(ld_block = snv_block) |>
  dplyr::bind_cols(snv_match) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    high_medium_programs = if (is.na(snp_id)) "" else programs_of(snp_id, strong_programs),
    other_programs = if (is.na(snp_id)) "" else programs_of(snp_id, other_programs)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    status = dplyr::case_when(
      is.na(ld_block) ~ status_levels[1],
      is.na(snp_id) ~ status_levels[2],
      high_medium_programs != "" ~ status_levels[5],
      other_programs != "" ~ status_levels[4],
      TRUE ~ status_levels[3]
    ),
    status = factor(status, levels = status_levels),
    cluster = factor(cluster, levels = cluster_levels)
  )

funnel <- snv_status |>
  dplyr::count(cluster, status, .drop = FALSE) |>
  tidyr::pivot_wider(names_from = status, values_from = n, values_fill = 0)
funnel <- dplyr::bind_rows(
  funnel |> dplyr::mutate(cluster = as.character(cluster)),
  funnel |>
    dplyr::summarise(dplyr::across(-cluster, sum)) |>
    dplyr::mutate(cluster = "All clusters")
)
funnel$n_signals <- rowSums(funnel[, status_levels])
print(knitr::kable(
  funnel[, c("cluster", "n_signals", status_levels)],
  caption = paste0(
    "Where each cluster's Suzuki index SNVs ended up. Each SNV is counted in ",
    "the last step it reached."
  ),
  row.names = FALSE
))
```

The table below counts, for each high or medium program, the Suzuki clusters
of the index SNVs matched to its member SNPs (split as described above). The
first row is the same count over every matched SNP in the matrix, which is
what a program with no preference for any cluster would look like.

- **enrichment:** a cluster's share of the program's matched SNPs divided by
  its share of all matched SNPs.
- **q:** the share of 2,000 random sets of matched SNPs, of the program's
  size, with at least the program's count for that cluster; BH-adjusted over
  every program × cluster cell.
- **top_cluster:** the cluster with the smallest p, among clusters with at
  least two matched SNPs in the program. A top cluster with q above 0.05 is
  no evidence of a preference.

```{r suzuki-snv-crosstab, results = "asis"}
# Each matched matrix SNP carries weight 1, split evenly across its SNVs.
snp_cluster_weights <- snv_status |>
  dplyr::filter(!is.na(snp_id)) |>
  dplyr::group_by(snp_id) |>
  dplyr::mutate(weight = 1 / dplyr::n()) |>
  dplyr::group_by(snp_id, cluster) |>
  dplyr::summarise(weight = sum(weight), .groups = "drop")

# Matched matrix SNPs x clusters: each row sums to 1.
weight_matrix <- snp_cluster_weights |>
  dplyr::mutate(cluster = factor(cluster, levels = cluster_levels)) |>
  tidyr::pivot_wider(
    names_from = cluster, values_from = weight, values_fill = 0,
    names_expand = TRUE
  )
matched_snps <- weight_matrix$snp_id
weight_matrix <- as.matrix(weight_matrix[, cluster_levels])
rownames(weight_matrix) <- matched_snps
background_weights <- colSums(weight_matrix)
background_share <- background_weights / sum(background_weights)

# Permutation null: random sets of matched SNPs of the program's size.
set.seed(20240214)
n_perm_snv <- 2000L
program_tiers <- program_summary$programs |>
  dplyr::transmute(
    program = as.character(program), confidence_tier = as.character(confidence_tier),
    similarity_z, n_snps_filtered, parent_program
  )
snv_cells <- dplyr::bind_rows(lapply(strong_programs, function(k) {
  members <- intersect(program_snps[[k]], matched_snps)
  observed <- colSums(weight_matrix[members, , drop = FALSE])
  null <- if (length(members) == 0) {
    matrix(0, length(cluster_levels), n_perm_snv)
  } else {
    replicate(n_perm_snv, colSums(weight_matrix[sample(matched_snps, length(members)), , drop = FALSE]))
  }
  return(data.frame(
    program = k,
    cluster = cluster_levels,
    n_member_snps = length(program_snps[[k]]),
    n_matched = length(members),
    weight = observed,
    enrichment = if (length(members) == 0) NA_real_ else (observed / length(members)) / background_share,
    p = (1 + rowSums(null >= observed)) / (n_perm_snv + 1),
    stringsAsFactors = FALSE
  ))
}))
snv_cells$q <- stats::p.adjust(snv_cells$p, method = "BH")

snv_top <- snv_cells |>
  dplyr::filter(weight >= 2) |>
  dplyr::group_by(program) |>
  dplyr::arrange(p, dplyr::desc(enrichment), .by_group = TRUE) |>
  dplyr::slice_head(n = 1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    program, top_cluster = cluster, enrichment = round(enrichment, 2), q = signif(q, 2)
  )
snv_crosstab <- snv_cells |>
  dplyr::mutate(weight = round(weight, 1)) |>
  dplyr::select(program, n_member_snps, n_matched, cluster, weight) |>
  tidyr::pivot_wider(names_from = cluster, values_from = weight) |>
  dplyr::left_join(snv_top, by = "program") |>
  dplyr::mutate(top_cluster = tidyr::replace_na(top_cluster, "none")) |>
  dplyr::left_join(program_tiers, by = "program") |>
  order_programs_by_confidence()

background_row <- data.frame(
  program = "all matched SNPs",
  n_member_snps = ncol(X),
  n_matched = length(matched_snps),
  t(round(background_weights, 1)),
  top_cluster = "",
  enrichment = NA_real_,
  q = NA_real_,
  confidence_tier = "background",
  check.names = FALSE,
  stringsAsFactors = FALSE
)
print(kable_confidence(
  dplyr::bind_rows(
    background_row,
    snv_crosstab |>
      dplyr::mutate(program = program_label) |>
      dplyr::select(dplyr::all_of(names(background_row)))
  ),
  drop_tier_col = TRUE,
  caption = paste0(
    "Suzuki clusters of the index SNVs matched to each high (green) or ",
    "medium (orange) program's member SNPs. n_matched = member SNPs matched ",
    "by at least one Suzuki SNV; cluster columns are weighted counts."
  )
))
```

By tier, the programs whose top cluster is enriched (q ≤ 0.05):

```{r suzuki-tier-enrichment}
suzuki_by_tier <- snv_crosstab |>
  dplyr::filter(confidence_tier %in% c("high", "medium")) |>
  dplyr::mutate(enriched = !is.na(q) & q <= 0.05) |>
  dplyr::group_by(confidence_tier) |>
  dplyr::summarise(
    n_programs = dplyr::n(),
    n_enriched = sum(enriched),
    enriched_clusters = {
      counts <- table(top_cluster[enriched])
      paste0(names(counts), " (", counts, ")", collapse = ", ")
    },
    .groups = "drop"
  ) |>
  dplyr::arrange(factor(confidence_tier, levels = c("high", "medium")))
tier_enriched <- function(tier) {
  row <- suzuki_by_tier[suzuki_by_tier$confidence_tier == tier, ]
  return(if (nrow(row) == 0) "0 of 0" else paste(row$n_enriched, "of", row$n_programs))
}
knitr::kable(
  suzuki_by_tier,
  caption = paste0(
    "Programs per tier whose top Suzuki cluster is enriched at q <= 0.05, ",
    "and those clusters (number of programs)."
  ),
  row.names = FALSE
)
```

`r tier_enriched("high")` high programs and `r tier_enriched("medium")`
medium programs have an enriched top cluster.

The full SNV-level table is below: every Suzuki index SNV, the GPMap SNP it
was matched to, and the programs that SNP is in (strongest loading first).

```{r suzuki-snv-table, results = "asis"}
cat("<details><summary>All 1,289 Suzuki index SNVs (click to expand)</summary>\n\n")
print(knitr::kable(
  snv_status |>
    dplyr::arrange(cluster, chr, pos_b38) |>
    dplyr::mutate(ld_r = round(r, 3)) |>
    dplyr::transmute(
      cluster, locus, rsid, chr, pos_b38, gpmap_rsid, ld_r, distance_kb,
      status, high_medium_programs, other_programs
    ),
  format = "html",
  row.names = FALSE,
  caption = paste0(
    "Suzuki index SNVs and the GPMap SNP each was matched to. ld_r = linkage ",
    "disequilibrium r between the SNV and the matched SNP (1 = the same ",
    "variant; negative = alleles in opposite phase); distance_kb = their ",
    "physical distance."
  )
))
cat("\n\n</details>\n")
```

### 4d. Trait-profile similarity

Suzuki's clusters are defined by their profiles over 37 cardiometabolic
traits. This compares those profiles directly with each program's.

- **Program profile:** the mean of the EBMF input matrix over the program's
  member SNPs, on the Suzuki traits that are rows of the matrix (4a). This is
  the matrix EBMF factorises: each SNP is aligned to its T2D risk allele, as
  in Suzuki's profiles, and the z-scores are asinh-compressed. A missing cell
  (no colocalisation) counts as 0. (, the raw matrix, is not aligned, so
  its column means would mix risk and protective alleles.)
- **Cluster profile:** Suzuki's cluster mean, as `beta / se` (Supplementary
  Table 7).
- **Similarity:** the cosine between the two profiles over the shared traits:
  1 for the same pattern of signs and relative sizes, 0 for unrelated, −1 for
  opposite.

The two profiles are on different scales (asinh-compressed colocalisation z
against the z of a cluster mean), so only the pattern is compared, not the
size. A program that loads on few of the Suzuki traits gets a noisy cosine.
Some Suzuki clusters have similar profiles over these traits, so a program
cannot separate them; the first table gives the cosine between the clusters
themselves as a reference.

```{r trait-profile-similarity, fig.width = 9, fig.height = 7}
profile_traits <- suzuki_trait_map |>
  dplyr::filter(in_matrix) |>
  dplyr::select(trait_acronym, gpmap_trait_id)
cluster_profiles <- suzuki_profiles |>
  dplyr::filter(trait_acronym %in% profile_traits$trait_acronym) |>
  dplyr::mutate(z = beta / se) |>
  dplyr::select(cluster, trait_acronym, z) |>
  tidyr::pivot_wider(names_from = cluster, values_from = z)
cluster_profiles <- as.matrix(cluster_profiles[
  match(profile_traits$trait_acronym, cluster_profiles$trait_acronym),
  cluster_levels
])

x_profile <- ebmf_result$ebmf_input[profile_traits$gpmap_trait_id, , drop = FALSE]
x_profile[is.na(x_profile)] <- 0
cosine <- function(a, b) {
  denom <- sqrt(sum(a^2)) * sqrt(sum(b^2))
  return(if (denom == 0) NA_real_ else sum(a * b) / denom)
}
cluster_cosine <- outer(
  cluster_levels, cluster_levels,
  Vectorize(function(a, b) return(round(cosine(cluster_profiles[, a], cluster_profiles[, b]), 2)))
)
dimnames(cluster_cosine) <- list(cluster_levels, cluster_levels)
knitr::kable(
  cluster_cosine,
  caption = paste0(
    "Cosine between the Suzuki cluster profiles themselves, over the same ",
    nrow(profile_traits), " traits"
  )
)

profile_similarity <- dplyr::bind_rows(lapply(strong_programs, function(k) {
  member_mean <- rowMeans(x_profile[, intersect(program_snps[[k]], colnames(x_profile)), drop = FALSE])
  return(data.frame(
    program = k,
    cluster = cluster_levels,
    cosine = vapply(cluster_levels, function(cl) cosine(member_mean, cluster_profiles[, cl]), numeric(1)),
    stringsAsFactors = FALSE
  ))
}))
profile_order <- program_tiers |>
  dplyr::filter(program %in% strong_programs) |>
  order_programs_by_confidence()
profile_similarity |>
  dplyr::mutate(
    program = factor(paste0("P", program), levels = rev(paste0("P", profile_order$program))),
    cluster = factor(cluster, levels = cluster_levels)
  ) |>
  ggplot2::ggplot(ggplot2::aes(x = cluster, y = program, fill = cosine)) +
  ggplot2::geom_tile(colour = "white") +
  ggplot2::geom_text(ggplot2::aes(label = sprintf("%.2f", cosine)), size = 3) +
  ggplot2::scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    limits = c(-1, 1), name = "cosine"
  ) +
  ggplot2::labs(
    title = "Program vs Suzuki cluster trait profiles",
    subtitle = paste0(
      "Cosine over the ", nrow(profile_traits),
      " Suzuki traits in the matrix; programs in confidence order"
    ),
    x = NULL, y = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1))

profile_best <- profile_similarity |>
  dplyr::filter(is.finite(cosine)) |>
  dplyr::group_by(program) |>
  dplyr::arrange(dplyr::desc(cosine), .by_group = TRUE) |>
  dplyr::summarise(
    best_profile_cluster = cluster[1],
    best_cosine = round(cosine[1], 2),
    runner_up = paste0(cluster[2], " (", round(cosine[2], 2), ")"),
    .groups = "drop"
  )
```

### 4e. Program summary cards

One row per high (green) or medium (orange) program, by confidence tier and
then by similarity, with each child program indented under its parent:

- **effective_n_traits:** how many traits the program's loadings are spread
  over (see Program confidence in Part 3).
- **specific_traits:** the top three module-specific traits.
- **tissues and pathways:** the top three loading-based hits (FDR ≤ 0.05 and
  R² ≥ 0.05, largest effect first; Part 3).
- **suzuki_snvs:** the Suzuki clusters of the index SNVs matched to the
  program's member SNPs (4c), largest first.
- **suzuki_loci:** those SNVs' loci (Suzuki's names), with the clusters of
  the SNVs at each (×n for several SNVs), strongest loading first. Several
  clusters at one locus means Suzuki had several signals there that GPMap
  resolves as one.
- **profile_match:** the Suzuki cluster whose trait profile is closest to the
  program's (4d), with the cosine.
- **genes:** the first 10 positional and QTL genes, strongest loading first.
  Genes in **bold** are named at a Suzuki et al. index SNV.

```{r program-cards, results = "asis"}
if (length(strong_programs) == 0) {
  cat("No high or medium programs.\n")
} else {
  # Top loading-based hits per program, as used in Part 3.
  top_items <- function(by_program_list, value_col, n = 3L) {
    return(vapply(by_program_list, function(x) {
      tab <- x$comparison
      if (is.null(tab) || nrow(tab) == 0) {
        return("none")
      }
      tab <- tab[is.finite(tab$fdr) & tab$fdr <= 0.05 &
                   is.finite(tab$r_squared) & tab$r_squared >= min_r_squared, , drop = FALSE]
      if (nrow(tab) == 0) {
        return("none")
      }
      tab <- tab[order(-tab$enrichment), , drop = FALSE]
      return(paste(utils::head(tab[[value_col]], n), collapse = "<br>"))
    }, character(1)))
  }
  tissues_top <- top_items(tis_by, "tissue")
  pathways_top <- top_items(path_by, "description")
  traits_top <- stats::setNames(
    vapply(group_traits$by_group, function(x) {
      passing <- x$traits |> dplyr::filter(passes_min_specificity) |> utils::head(3L)
      return(if (nrow(passing) == 0) "none" else paste(passing$trait_name, collapse = "<br>"))
    }, character(1)),
    vapply(group_traits$by_group, function(x) as.character(x$group), character(1))
  )
  lookup <- function(x, key) {
    out <- if (length(x) == 0) NA_character_ else unname(x[key])
    return(ifelse(is.na(out), "none", out))
  }
  suzuki_snv_summary <- snv_crosstab |>
    dplyr::rowwise() |>
    dplyr::mutate(suzuki_snvs = {
      w <- unlist(dplyr::pick(dplyr::all_of(cluster_levels)))
      w <- sort(w[w > 0], decreasing = TRUE)
      if (length(w) == 0) "none" else paste0(names(w), " ", w, collapse = "<br>")
    }) |>
    dplyr::ungroup() |>
    dplyr::select(program, suzuki_snvs)
  suzuki_loci <- program_members |>
    dplyr::filter(program %in% strong_programs) |>
    dplyr::inner_join(
      snv_status |>
        dplyr::filter(!is.na(snp_id)) |>
        dplyr::select(snp_id, locus, cluster),
      by = "snp_id",
      relationship = "many-to-many"
    ) |>
    dplyr::distinct(program, snp_id, abs_loading, locus, cluster) |>
    dplyr::group_by(program, locus) |>
    dplyr::summarise(
      top = max(abs_loading),
      clusters = {
        n <- table(factor(cluster, levels = cluster_levels))
        n <- n[n > 0]
        paste0(names(n), ifelse(n > 1, paste0(" ×", n), ""), collapse = ", ")
      },
      .groups = "drop"
    ) |>
    dplyr::arrange(program, dplyr::desc(top)) |>
    dplyr::group_by(program) |>
    dplyr::summarise(
      suzuki_loci = paste0(
        paste0("<i>", utils::head(locus, 8L), "</i>: ", utils::head(clusters, 8L), collapse = "<br>"),
        if (dplyr::n() > 8L) paste0("<br>(+", dplyr::n() - 8L, " more loci)") else ""
      ),
      .groups = "drop"
    )
  gene_lists <- dplyr::bind_rows(lapply(split(program_members, program_members$program), function(m) {
    return(data.frame(
      program = m$program[1],
      positional_genes = format_genes(genes_in_order(m, positional_links), 10L),
      qtl_genes = format_genes(genes_in_order(m, qtl_links), 10L),
      stringsAsFactors = FALSE
    ))
  }))
  cards <- program_summary$programs |>
    dplyr::transmute(
      program = as.character(program),
      confidence_tier = as.character(confidence_tier),
      similarity_z,
      n_snps_filtered,
      parent_program,
      effective_n_traits = round(effective_n_traits, 2)
    ) |>
    dplyr::filter(program %in% strong_programs) |>
    dplyr::left_join(program_genes |> dplyr::select(program, n_snps), by = "program") |>
    dplyr::mutate(
      specific_traits = lookup(traits_top, program),
      tissues = lookup(tissues_top, program),
      pathways = lookup(pathways_top, program)
    ) |>
    dplyr::left_join(suzuki_snv_summary, by = "program") |>
    dplyr::left_join(suzuki_loci, by = "program") |>
    dplyr::left_join(profile_best, by = "program") |>
    dplyr::mutate(
      suzuki_snvs = tidyr::replace_na(suzuki_snvs, "none"),
      suzuki_loci = tidyr::replace_na(suzuki_loci, "none"),
      profile_match = paste0(best_profile_cluster, " (", best_cosine, ")")
    ) |>
    dplyr::left_join(gene_lists, by = "program") |>
    order_programs_by_confidence() |>
    dplyr::mutate(program = program_label) |>
    dplyr::select(
      program, confidence_tier, n_snps, effective_n_traits, specific_traits,
      tissues, pathways, suzuki_snvs, suzuki_loci, profile_match,
      positional_genes, qtl_genes
    )
  print(kable_confidence(
    cards,
    drop_tier_col = TRUE,
    caption = paste0(
      "One card per high (green) or medium (orange) program, each child ",
      "indented under its parent. Bold genes are named at a Suzuki et al. ",
      "index SNV."
    )
  ))
}
```

### 4f. Matching programs to Suzuki clusters

This section is written by hand from the results above, reading each
program's card (4e) with its Suzuki SNV counts (4c) and trait-profile
similarity (4d). The calls are:

- **clear:** the SNV counts favour the cluster (q ≤ 0.05) *and* the trait
  profile is closest to it or to a cluster it cannot be told apart from;
- **partial:** only one of the two supports the cluster, or the program sits
  between two similar clusters;
- **none:** no consistent match.

Three pairs of clusters have similar profiles over the traits GPMap has (4d):
Beta cell +PI and −PI (cosine 0.92), Metabolic syndrome and Lipodystrophy
(0.81), and Body fat and Obesity (0.73). A program is not expected to
separate the members of a pair, so a call can name both.

The program numbers and statistics below come from one render. The next chunk
checks each call against the current results and prints a warning if a
program has changed, so the text cannot silently go stale.

```{r manual-matches, results = "asis"}
manual_matches <- dplyr::tribble(
  ~program, ~suzuki_cluster, ~call, ~reading,
  "1", "Obesity", "clear",
  "Largest program. Most of its matched SNVs are Obesity (SEC16B, TFAP2B, GNPDA2, BDNF, FTO) and its profile is closest to Obesity. Its most specific traits (education, intelligence, apolipoprotein B) are not Suzuki traits.",
  "49", "Obesity", "clear",
  "A compact BMI core: BMI, body size at age 10 and impedance. Most of its SNPs are also in P1, including SEC16B, GNPDA2 and TFAP2B.",
  "44", "Obesity", "clear",
  "Obesity SNVs and an Obesity profile, with C-reactive protein as its most specific trait: an inflammation-flavoured obesity program.",
  "11", "Beta cell +PI / Beta cell -PI", "clear",
  "The only beta-cell program. HbA1c is its most specific glycaemic trait, it carries CDC123/CAMK1D, CDKN2A/B and KCNJ11/ABCC8, and its SNVs lean to Beta cell -PI while its profile is closest to Beta cell +PI. The two clusters are not separable here, as expected.",
  "6", "Lipodystrophy / Metabolic syndrome", "clear",
  "GRB14/COBLL1, INSR and ARL15; SHBG, heel bone density and lean mass. Lipodystrophy SNVs are the most over-represented and Metabolic syndrome SNVs the most numerous; the profile is closest to Lipodystrophy.",
  "2", "Lipodystrophy", "clear",
  "A lipid (VLDL, triglyceride) program at the insulin-resistance loci (INSR, GRB14/COBLL1, ANKRD55, CMIP). P33 and P36 sit inside it.",
  "33", "Lipodystrophy / Metabolic syndrome", "partial",
  "Sub-program of P2 with 4 SNPs: triglycerides and HDL. Too few SNPs for the SNV counts; the profile matches Lipodystrophy.",
  "36", "Lipodystrophy / Metabolic syndrome", "partial",
  "Sub-program of P33 with 3 SNPs. Too few SNPs for the SNV counts; the profile matches Lipodystrophy.",
  "54", "Lipodystrophy / Metabolic syndrome", "partial",
  "HDL and triglycerides with KLF14; the profile matches Lipodystrophy but its 6 SNPs show no cluster preference.",
  "40", "none", "none",
  "Three SNPs in the MHC region with immune pathways; not a Suzuki-like program.",
  "22", "Obesity / Body fat", "partial",
  "CRP and limb fat percentage; the profile is closest to Obesity, but the SNV counts show no preference.",
  "50", "Beta cell +PI / Beta cell -PI", "partial",
  "MTNR1B, a fasting-glucose locus, with triglycerides and GGT. Its profile is closest to Beta cell +PI, but the SNV preference is not significant.",
  "16", "Obesity", "clear",
  "Obesity SNVs (SEC16B, TFAP2B, GNPDA2, PCSK1) and an Obesity profile; shares many SNPs with P1.",
  "4", "Obesity", "clear",
  "Second large Obesity program (education, sitting height, triglycerides); shares many SNPs with P1.",
  "13", "none", "none",
  "Mixed: alcohol intake, LDL and pulse pressure. The profile sits between Lipodystrophy and Metabolic syndrome, but the SNVs show no preference.",
  "31", "Lipodystrophy / Metabolic syndrome", "clear",
  "Blood pressure and impedance at GRB14/COBLL1, EBF1 and CYTL1; Lipodystrophy SNVs over-represented and a Lipodystrophy profile.",
  "10", "Metabolic syndrome / Lipodystrophy", "clear",
  "Red-cell traits and triglycerides at INSR, ABO and GRB14/COBLL1; Metabolic syndrome SNVs over-represented, Lipodystrophy profile.",
  "26", "none", "none",
  "Six SNPs, HDL and triglycerides; no SNV preference and no clear profile.",
  "35", "Metabolic syndrome / Lipodystrophy", "clear",
  "Platelet and red-cell traits with SHBG; Metabolic syndrome SNVs over-represented, Lipodystrophy profile.",
  "24", "Metabolic syndrome", "clear",
  "Lean mass and heel bone density; the strongest Metabolic syndrome SNV enrichment, and its profile is closest to Metabolic syndrome.",
  "53", "Lipodystrophy / Metabolic syndrome", "partial",
  "HDL cholesterol; a Lipodystrophy profile but no significant SNV preference.",
  "17", "Lipodystrophy / Metabolic syndrome", "partial",
  "Three SNPs, triglycerides; a Lipodystrophy profile, too few SNPs for the SNV counts.",
  "45", "none", "none",
  "Lung function and lean mass; no SNV preference, and the profile is only weakly Obesity-like.",
  "47", "Lipodystrophy / Metabolic syndrome", "clear",
  "IGF-1, AST and red-cell traits at PNPLA3 and INS/IGF2; Lipodystrophy SNVs over-represented, Metabolic syndrome profile."
)

# Check each call against the current results: the program must still be high
# or medium, and a clear call needs its cluster(s) to be the program's
# significant top SNV cluster or among its two closest profiles.
profile_top2 <- profile_similarity |>
  dplyr::group_by(program) |>
  dplyr::slice_max(cosine, n = 2, with_ties = FALSE) |>
  dplyr::summarise(profile_top2 = list(cluster), .groups = "drop")
manual_check <- manual_matches |>
  dplyr::left_join(snv_top, by = "program") |>
  dplyr::left_join(profile_top2, by = "program") |>
  dplyr::rowwise() |>
  dplyr::mutate(
    named = list(strsplit(suzuki_cluster, " / ", fixed = TRUE)[[1]]),
    still_strong = program %in% strong_programs,
    snv_supports = !is.na(q) && q <= 0.05 && top_cluster %in% named,
    profile_supports = any(named %in% profile_top2),
    consistent = still_strong && (call != "clear" || (snv_supports && profile_supports))
  ) |>
  dplyr::ungroup()
stale <- manual_check$program[!manual_check$consistent]
missing <- setdiff(strong_programs, manual_matches$program)
if (length(stale) > 0 || length(missing) > 0) {
  cat(
    "<div style=\"border:2px solid #b2182b;padding:8px;margin:8px 0;\">",
    "<b>Warning: the hand-written matches below no longer fit the results.</b> ",
    if (length(stale) > 0) paste0("Calls that no longer hold: P", paste(stale, collapse = ", P"), ". "),
    if (length(missing) > 0) paste0("High/medium programs with no call: P", paste(missing, collapse = ", P"), ". "),
    "Re-read 4c–4e and update this section.</div>\n\n",
    sep = ""
  )
}

print(kable_confidence(
  manual_matches |>
    dplyr::left_join(program_tiers, by = "program") |>
    dplyr::mutate(confidence_tier = tidyr::replace_na(confidence_tier, "")) |>
    order_programs_by_confidence() |>
    dplyr::transmute(program = program_label, confidence_tier, suzuki_cluster, call, reading),
  drop_tier_col = TRUE,
  caption = "Hand-written match of each high (green) or medium (orange) program to Suzuki's clusters"
))
```

**Against the expectations.**

- **Obesity is recovered clearly and repeatedly** (P1, P49, P44, P16, P4).
  P1, P4 and P16 share many SNPs, and P49 is a BMI core that sits mostly
  inside P1.
- **Lipodystrophy and Metabolic syndrome are recovered together.** Many
  programs sit at the insulin-resistance loci (GRB14/COBLL1, INSR, ANKRD55,
  CMIP, KLF14). The SNV counts pull some towards Lipodystrophy (P2, P6, P31,
  P47) and others towards Metabolic syndrome (P10, P24, P35). The trait
  profiles cannot separate the two, because Suzuki's own profiles are close
  over these traits.
- **Beta cell: one joint program (P11), as expected.** It carries markers of
  both beta-cell clusters, and neither the SNVs nor the profile split them.
- **Body fat merges into Obesity, as expected.** No program prefers Body fat
  SNVs; they are spread over the Obesity programs.
- **Residual glycaemic, the largest cluster, has no program of its own.** Its
  SNVs are spread across programs (4c), which fits its role as a catch-all.
- **Liver/lipid metabolism is not recovered**, as expected with 3 signals.
- **Tissues give little support to the expectations.** Only four high or
  medium programs have a loading-based tissue hit (4e). P35 (Metabolic
  syndrome / Lipodystrophy) is enriched in liver, P22 and P17 in plasma, and
  P26 (no match) in kidney cortex and brain amygdala. No Lipodystrophy-like
  program is enriched in adipose, and no Obesity program in brain.

## Part 5: Candidate trait pairs for a multi-trait analysis

This part is documentation only; nothing here is run. Each pair runs the
upload against one partner trait in
[the multi-trait investigation](investigation-multitrait-programs.html). The
Suzuki clusters predict which shared programs the pair should show and in which
direction:

- **concordant:** the T2D risk allele raises the partner trait;
- **antagonistic:** the T2D risk allele lowers it.

| Partner trait | GPMap id | N | Coloc groups | Clusters it should separate | Expected pattern |
|---|---|---|---|---|---|
| Body mass index | 1992 | 532,396 | 992 | Obesity vs beta-cell | Obesity programs shared and concordant; Beta cell ±PI programs T2D-specific |
| Triglyceride levels | 1631 | 437,532 | 587 | Lipodystrophy, Metabolic syndrome, Liver/lipid | Lipodystrophy and Metabolic syndrome concordant; TM6SF2-type liver loci antagonistic |
| HDL cholesterol levels | 1630 | 400,754 | 735 | Lipodystrophy, Obesity, Metabolic syndrome | Antagonistic sharing (T2D risk lowers HDL) |
| Waist-to-hip ratio adjusted for BMI | 1653 | 458,349 | 478 | Lipodystrophy vs Obesity | Lipodystrophy concordant, with BMI removed |
| Waist-hip ratio | 1994 | 502,773 | 550 | Lipodystrophy, Obesity, Metabolic syndrome | Concordant for all three |
| Haemoglobin A1c levels | 1640 | 437,749 | 838 | Beta-cell, Residual glycaemic | Glycaemic clusters shared; obesity/lipid clusters weaker |
| Fasting glucose | 913 | 200,622 | 95 | Beta-cell, Residual glycaemic | As HbA1c, but sparse in GPMap |
| Body fat percentage | 2581 | 401,772 | 671 | Body fat, Obesity vs Lipodystrophy | Body fat and Obesity concordant; Lipodystrophy antagonistic |
| Alanine aminotransferase / GGT | 1642 / 1634 | ≈437,700 | 449 / 702 | Liver/lipid, Metabolic syndrome | Liver-fat loci concordant |
| Coronary artery disease | 2475 | 352,063 | 64 | Obesity vs Beta cell +PI | Obesity concordant; Beta cell +PI antagonistic (Suzuki's polygenic-score result) |
| Type 2 diabetes (GPMap native) | 114 | 655,666 | 178 | none | Replication pair: programs should correspond one to one |

**Caveat: sample overlap.** UK Biobank, and for several traits FinnGen and the
Million Veteran Program, contribute to both the T2DGGI EUR meta-analysis and
most partner GWAS. Correlated sampling error then inflates apparent sharing
between the two traits. Sharing that is weak or confined to a single locus
should be read with this in mind.

**Untested.** The multi-trait vignette has not yet been run with an upload GUID
as one of its traits. A first run would look like this:

```{r multitrait-example, eval = FALSE, echo = TRUE}
rmarkdown::render(
  "vignettes/investigation-multitrait-programs.Rmd",
  params = list(
    api = "production",
    trait_ids = c("c6d2e20f-6ece-d093-78ee-37534547c466", 1992)
  ),
  output_file = "investigation-multitrait-programs-t2dggi-bmi.html"
)
```
